### 2016-10-12 main.tex
Add proposal from OneNote.

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}

\title{6.830 Project Proposal \\
  \large Data Imputation as a Query Optimization}
\author{Jose Cambronero \& John Feser \& Micah Smith}

\begin{document}
\maketitle

\section{Problem Statement}
Data imputation is a well studied problem in the machine learning literature, but some aspects related to the implementation of imputation in database systems have not been sufficiently examined. Traditionally studies have viewed imputation as a pre-processing step on a table, rather than as an active step in each query. Particularly, existing work has not incorporated imputation into query optimization, nor has it investigated the relative impact of early versus late stage imputation on the quality of query results.

\section{Approach}
In order to study the placement of an imputation step, we propose to create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process in SimpleDB.  We plan to introduce measures for uncertainty and completeness of query results as well as the cost of imputation, which outline the main tradeoffs in the imputation placement. We introduce these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning. Finally, we plan to produce empirical results that show the tradeoffs between efficiency and accuracy in the context of simple data imputation models in queries that perform aggregate operations or return a significant number of tuples (in contrast to simple point-queries) and compare to existing approaches.  These types of queries are more likely to provide reasonable cases for imputation of missing values and reflecting an increasingly large portion of workloads in modern OLAP warehouses.

\section{Example}

\section{Evaluation}
We are considering various possible data sources to experiment with. Existing literature has used the American Community Survey, so this may be a natural choice in our case as well. The ACS data is a good test dataset because of its scale and relative simplicity. Additionally, we may be able to compare to existing work that used ACS data.

In order to measure accuracy of our imputation algorithm, we plan to compute results with and without missing values where possible, and use standard error measures to compare to our results. The best-case scenario for data imputation is missing data where the data is removed uniformly at random. It should be easy to construct a test dataset from an existing dataset simply by removing data from it; artificially degrading an existing dataset also means that we will have ground truth to compare our results with.

In contrast to pre-processing-based imputation, we plan to make a case for imputation at the level of intermediate results. There is an analogy to be made between this approach and views, as neither alter the original table and provide logical independence. Many modern database servers have a variety of clients, with different needs, making one imputation strategy unlikely to satisfy everyone. Consider a financial data warehouse, with clients such as business strategy and actuaries. It is unlikely that the queries issued by these two groups will require or benefit from the same imputation strategy. If a salary value is missing for a client, the business strategy query may not require a costly regression model estimate and a simple heuristic could be enough to return a reasonable answer.

Furthermore, we plan to show that there is an opportunity for more efficient imputation on (smaller) intermediate results. In very large databases it may not be feasible to perform imputation on the base tables, particularly if the imputation method of choice is expensive. However, queries often return results that are smaller than the base tables, so it may be more reasonable to perform imputation on these intermediate results. The intermediate results could then be stored as a materialized view if they are expected to be reused or recomputed as necessary.

\section{Project Milestones}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2016-10-12 main.tex
Add milestones.

@@
 \section{Project Milestones}
+
+We outline a series of milestones to be completed:
+\begin{enumerate}
+\item Initial data collection and query workload for evaluation.
+\item Consolidate set of imputation strategies.
+\item Formalize a measure of uncertainty, completeness and cost in imputation as related to 2.
+\item Formalize the possible query rewrites when using an imputation operator.
+\item Incorporate imputation-related measures into cost estimation for SimpleDB.
+\item Incorporate imputation-related rewrites for SimpleDB.
+\item Evaluate performance.
+\end{enumerate}
 \end{document}

### 2016-10-12 main.tex
proposal: added example table and diagrams

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{qtree} % for diagram
+\usepackage{listings} % for code excerpt
+\usepackage{color} % for code excerpt
+\lstset{
+  basicstyle=\ttfamily\footnotesize,
+  language=SQL,
+  keywordstyle=\color{blue},
+  otherkeywords={IMPUTE, impute}
+}
+\usepackage{booktabs} % for table
 
@@
 \section{Example}
+We consider a simple example that shows the potential impact of placing an imputation operator at different points in our query plan.
+
+We assume the following relation \lstinline{Employees(id, salary, dept)}, with tuples shown in Table~\ref{example-table}.
+We impute by averaging over \lstinline{salary}, and we consider the query \lstinline{SELECT AVG(salary) FROM Employees GROUP BY dept}.
+
+\begin{table}[h]
+\centering
+\caption{Example table with missing data}
+\label{example-table}
+\begin{tabular}{@{}lll@{}}
+\toprule
+id & salary & dept \\ \midrule
+1  & 10     & 1    \\
+2  & 15     & 1    \\
+3  & NA     & 1    \\
+4  & 5      & 2    \\ \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{figure}
+ \begin{minipage}[b]{0.5\textwidth}
+	\Tree [.\lstinline{AVG(salary)} [.\lstinline{GROUP BY (dept)} [.\lstinline{IMPUTE(Salary)} \lstinline{Employees} ] ] ]
+    \caption{Early placement of impute operator}
+    \label{early-fig}
+  \end{minipage}
+\hfill
+ \begin{minipage}[b]{0.5\textwidth}
+	\Tree [.\lstinline{AVG(salary)} [.\lstinline{IMPUTE(Salary)} [.\lstinline{GROUP BY (dept)} \lstinline{Employees} ] ] ]
+    \caption{Late placement of impute operator}
+    \label{late-fig}
+  \end{minipage}
+\end{figure}
+
+In the early case (Figure~\ref{early-fig}), we impute over all salaries, and obtain as a final result \lstinline{[(1, 11.67), (2, 5)]}, while the late case (Figure~\ref{late-fig}) results in \lstinline{[(1, 12.5), (2, 5)]}.
+
+This example shows that the imputation operator, even in its simplest form, can have a significant impact on the query results.
+Furthermore, this example fails to capture other important dimensions such as cost of imputation.
 

### 2016-10-12 main.tex
proposal: minor typos

@@
 Finally, we plan to produce empirical results that show the tradeoffs between efficiency and accuracy in the context of simple data imputation models in queries that perform aggregate operations or return a significant number of tuples (in contrast to simple point-queries) and compare to existing approaches.
-These types of queries are more likely to provide reasonable cases for imputation of missing values and reflecting an increasingly large portion of workloads in modern OLAP warehouses.
+These types of queries are more likely to provide reasonable cases for imputation of missing values and reflect an increasingly large portion of workloads in modern OLAP warehouses.
 
@@
 
-This example shows that the imputation operator, even in its simplest form, can have a significant impact on the query results.
+This example shows that the imputation operator, even in its simplest form, can have a significant impact on query results.
 Furthermore, this example fails to capture other important dimensions such as cost of imputation.

### 2016-10-12 main.tex
Spelling and phrasing.

@@
 Data imputation is a well studied problem in the machine learning literature, but some aspects related to the implementation of imputation in database systems have not been sufficiently examined.
-Traditionally studies have viewed imputation as a pre-processing step on a table, rather than as an active step in each query.
+Previous work has viewed imputation as a pre-processing step on a table, rather than as an active step in each query.
 Particularly, existing work has not incorporated imputation into query optimization, nor has it investigated the relative impact of early versus late stage imputation on the quality of query results.
@@
 In order to study the placement of an imputation step, we propose to create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process in SimpleDB.
-We plan to introduce measures for uncertainty and completeness of query results as well as the cost of imputation, which outline the main tradeoffs in the imputation placement.
-We introduce these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning.
-Finally, we plan to produce empirical results that show the tradeoffs between efficiency and accuracy in the context of simple data imputation models in queries that perform aggregate operations or return a significant number of tuples (in contrast to simple point-queries) and compare to existing approaches.
-These types of queries are more likely to provide reasonable cases for imputation of missing values and reflect an increasingly large portion of workloads in modern OLAP warehouses.
+We will introduce measures of uncertainty and completeness of query results as well as the cost of imputation, which outline the main trade-offs in the imputation placement.
+We add these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning.
+Finally, we will produce empirical results that show the trade-offs between efficiency and accuracy for simple data imputation models.
+We will focus on queries that perform aggregate operations or return a significant number of tuples (in contrast to simple point-queries).
+These types of queries are likely to provide opportunities for imputation of missing values and represent an increasingly large portion of workloads in modern OLAP warehouses.
+Finally, we will compare our results to existing approaches.
 
 \section{Example}
-We consider a simple example that shows the potential impact of placing an imputation operator at different points in our query plan.
+In this section, we use a simple example to show that placing the imputation operator at different points in the query plan has a significant impact on the query results.
 
-We assume the following relation \lstinline{Employees(id, salary, dept)}, with tuples shown in Table~\ref{example-table}.
-We impute by averaging over \lstinline{salary}, and we consider the query \lstinline{SELECT AVG(salary) FROM Employees GROUP BY dept}.
+We start with the relation \lstinline{Employees(id, salary, dept)} shown in Table~\ref{example-table}.
+We impute  by averaging over \lstinline{salary}, and we compare two plans for the query \lstinline{SELECT AVG(salary) FROM Employees GROUP BY dept}.
 
@@
 
-In the early case (Figure~\ref{early-fig}), we impute over all salaries, and obtain as a final result \lstinline{[(1, 11.67), (2, 5)]}, while the late case (Figure~\ref{late-fig}) results in \lstinline{[(1, 12.5), (2, 5)]}.
+In the early case (Figure~\ref{early-fig}), we impute over all salaries, and obtain as a final result \lstinline{[(1, 11.67), (2, 5)]}.
+The late case (Figure~\ref{late-fig}) averages over salaries within a department and results in \lstinline{[(1, 12.5), (2, 5)]}.
 
@@
 \section{Evaluation}
-We are considering various possible data sources to experiment with.
-Existing literature has used the American Community Survey, so this may be a natural choice in our case as well.
+We are considering several data sources for our evaluation.
+Existing literature uses the American Community Survey, so we think this is a good choice for our experiments as well.
 The ACS data is a good test dataset because of its scale and relative simplicity.
-Additionally, we may be able to compare to existing work that used ACS data.
+Additionally, we may be able to compare directly to existing work that uses ACS data.
 
-In order to measure accuracy of our imputation algorithm, we plan to compute results with and without missing values where possible, and use standard error measures to compare to our results.
-The best-case scenario for data imputation is missing data where the data is removed uniformly at random.
-It should be easy to construct a test dataset from an existing dataset simply by removing data from it; artificially degrading an existing dataset also means that we will have ground truth to compare our results with.
+In order to measure the accuracy of our imputation algorithm, we plan to compute results with and without missing values where possible, and use standard error measures to compare to our results.
+The best-case scenario for imputation is a dataset where data is removed uniformly at random.
+It should be easy to construct a test dataset  simply by removing data from an existing dataset; artificially degrading an existing dataset also means that we will have ground truth with which to compare our results.
 
-In contrast to pre-processing-based imputation, we plan to make a case for imputation at the level of intermediate results.
-There is an analogy to be made between this approach and views, as neither alter the original table and provide logical independence.
-Many modern database servers have a variety of clients, with different needs, making one imputation strategy unlikely to satisfy everyone.
-Consider a financial data warehouse, with clients such as business strategy and actuaries.
+In contrast to pre-processing-based imputation, we plan to make a case for imputation on intermediate query results.
+This approach is similar to a view, as neither alter the original table and both provide logical independence.
+Many modern database servers have a variety of clients with different needs, so one imputation strategy is unlikely to satisfy everyone.
+Consider a financial data warehouse with a business strategy client and an actuarial client.
 It is unlikely that the queries issued by these two groups will require or benefit from the same imputation strategy.
-If a salary value is missing for a client, the business strategy query may not require a costly regression model estimate and a simple heuristic could be enough to return a reasonable answer.
+If a salary value is missing for a customer, the business strategy query may not require a costly regression model estimate and a simple heuristic could be enough to return a reasonable answer.
+It may be cheaper to do imputation on the fly than to maintain a separate copy of the data for each imputation strategy.
 
-Furthermore, we plan to show that there is an opportunity for more efficient imputation on (smaller) intermediate results.
+Furthermore, we believe that there is an opportunity for more efficient imputation on (smaller) intermediate results.
 In very large databases it may not be feasible to perform imputation on the base tables, particularly if the imputation method of choice is expensive.
-However, queries often return results that are smaller than the base tables, so it may be more reasonable to perform imputation on these intermediate results.
-The intermediate results could then be stored as a materialized view if they are expected to be reused or recomputed as necessary.
+However, queries often return results that are smaller than the base tables, so it may be faster to perform imputation on intermediate results than on the base tables.
+The intermediate results could then be stored as a materialized view if they are expected to be reused or they could be recomputed as necessary.
 

### 2016-11-28 paper/algorithm.tex
add paper skeleton

--- 2016-10-12 main.tex
@@
-\documentclass{article}
-\usepackage[utf8]{inputenc}
-\usepackage[margin=1in]{geometry}
-\usepackage{qtree} % for diagram
-\usepackage{listings} % for code excerpt
-\usepackage{color} % for code excerpt
-\lstset{
-  basicstyle=\ttfamily\footnotesize,
-  language=SQL,
-  keywordstyle=\color{blue},
-  otherkeywords={IMPUTE, impute}
-}
-\usepackage{booktabs} % for table
-
-\title{6.830 Project Proposal \\
-  \large Data Imputation as a Query Optimization}
-\author{Jose Cambronero \& John Feser \& Micah Smith}
-
-\begin{document}
-\maketitle
-
-\section{Problem Statement}
-Data imputation is a well studied problem in the machine learning literature, but some aspects related to the implementation of imputation in database systems have not been sufficiently examined.
-Traditionally studies have viewed imputation as a pre-processing step on a table, rather than as an active step in each query.
-Particularly, existing work has not incorporated imputation into query optimization, nor has it investigated the relative impact of early versus late stage imputation on the quality of query results.
-
-\section{Approach}
-In order to study the placement of an imputation step, we propose to create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process in SimpleDB.
-We plan to introduce measures for uncertainty and completeness of query results as well as the cost of imputation, which outline the main tradeoffs in the imputation placement.
-We introduce these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning.
-Finally, we plan to produce empirical results that show the tradeoffs between efficiency and accuracy in the context of simple data imputation models in queries that perform aggregate operations or return a significant number of tuples (in contrast to simple point-queries) and compare to existing approaches.
-These types of queries are more likely to provide reasonable cases for imputation of missing values and reflect an increasingly large portion of workloads in modern OLAP warehouses.
-
-\section{Example}
-We consider a simple example that shows the potential impact of placing an imputation operator at different points in our query plan.
-
-We assume the following relation \lstinline{Employees(id, salary, dept)}, with tuples shown in Table~\ref{example-table}.
-We impute by averaging over \lstinline{salary}, and we consider the query \lstinline{SELECT AVG(salary) FROM Employees GROUP BY dept}.
-
-\begin{table}[h]
-\centering
-\caption{Example table with missing data}
-\label{example-table}
-\begin{tabular}{@{}lll@{}}
-\toprule
-id & salary & dept \\ \midrule
-1  & 10     & 1    \\
-2  & 15     & 1    \\
-3  & NA     & 1    \\
-4  & 5      & 2    \\ \bottomrule
-\end{tabular}
-\end{table}
-
-\begin{figure}
- \begin{minipage}[b]{0.5\textwidth}
-	\Tree [.\lstinline{AVG(salary)} [.\lstinline{GROUP BY (dept)} [.\lstinline{IMPUTE(Salary)} \lstinline{Employees} ] ] ]
-    \caption{Early placement of impute operator}
-    \label{early-fig}
-  \end{minipage}
-\hfill
- \begin{minipage}[b]{0.5\textwidth}
-	\Tree [.\lstinline{AVG(salary)} [.\lstinline{IMPUTE(Salary)} [.\lstinline{GROUP BY (dept)} \lstinline{Employees} ] ] ]
-    \caption{Late placement of impute operator}
-    \label{late-fig}
-  \end{minipage}
-\end{figure}
-
-In the early case (Figure~\ref{early-fig}), we impute over all salaries, and obtain as a final result \lstinline{[(1, 11.67), (2, 5)]}, while the late case (Figure~\ref{late-fig}) results in \lstinline{[(1, 12.5), (2, 5)]}.
-
-This example shows that the imputation operator, even in its simplest form, can have a significant impact on query results.
-Furthermore, this example fails to capture other important dimensions such as cost of imputation.
-
-\section{Evaluation}
-We are considering various possible data sources to experiment with.
-Existing literature has used the American Community Survey, so this may be a natural choice in our case as well.
-The ACS data is a good test dataset because of its scale and relative simplicity.
-Additionally, we may be able to compare to existing work that used ACS data.
-
-In order to measure accuracy of our imputation algorithm, we plan to compute results with and without missing values where possible, and use standard error measures to compare to our results.
-The best-case scenario for data imputation is missing data where the data is removed uniformly at random.
-It should be easy to construct a test dataset from an existing dataset simply by removing data from it; artificially degrading an existing dataset also means that we will have ground truth to compare our results with.
-
-In contrast to pre-processing-based imputation, we plan to make a case for imputation at the level of intermediate results.
-There is an analogy to be made between this approach and views, as neither alter the original table and provide logical independence.
-Many modern database servers have a variety of clients, with different needs, making one imputation strategy unlikely to satisfy everyone.
-Consider a financial data warehouse, with clients such as business strategy and actuaries.
-It is unlikely that the queries issued by these two groups will require or benefit from the same imputation strategy.
-If a salary value is missing for a client, the business strategy query may not require a costly regression model estimate and a simple heuristic could be enough to return a reasonable answer.
-
-Furthermore, we plan to show that there is an opportunity for more efficient imputation on (smaller) intermediate results.
-In very large databases it may not be feasible to perform imputation on the base tables, particularly if the imputation method of choice is expensive.
-However, queries often return results that are smaller than the base tables, so it may be more reasonable to perform imputation on these intermediate results.
-The intermediate results could then be stored as a materialized view if they are expected to be reused or recomputed as necessary.
-
-\section{Project Milestones}
-
-We outline a series of milestones to be completed:
-\begin{enumerate}
-\item Initial data collection and query workload for evaluation.
-\item Consolidate set of imputation strategies.
-\item Formalize a measure of uncertainty, completeness and cost in imputation as related to 2.
-\item Formalize the possible query rewrites when using an imputation operator.
-\item Incorporate imputation-related measures into cost estimation for SimpleDB.
-\item Incorporate imputation-related rewrites for SimpleDB.
-\item Evaluate performance.
-\end{enumerate}
-\end{document}
+\section{Algorithm}
 
@@
 %%% mode: latex
-%%% TeX-master: t
+%%% TeX-master: "main"
 %%% End:

### 2016-11-28 paper/main.tex
add paper skeleton

\documentclass{article}

\title{Data Imputation as a Query Optimization}

\author{
  Jose Cambronero \\
  \and
  Micah Cohen
  \and
  John K. Feser}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

\begin{document}

\maketitle

\begin{abstract}
In order to study the placement of an imputation step, we create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process in SimpleDB.  We introduce measures of uncertainty and completeness of query results as well as the cost of imputation, which outline the main trade-offs in the imputation placement. We add these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning. We show the trade-offs between efficiency and accuracy for simple data imputation models. \end{abstract}

\input{introduction}
\input{examples}
\input{algorithm}
\input{experiments}
\input{relatedwork}
\input{conclusion}

\bibliography{references}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2016-11-30 paper/algorithm.tex
start filling out algorithm section from onenote

@@
 \section{Algorithm}
+
+At a high level, our goal is to select a query plan that minimizes a metric which combines the cost of the query and the quality of the results.
+Precisely, we will find a query plan $Q$ that minimizes $\text{Cost}(Q) = \alpha \times \text{Time}(Q) + (1 - \alpha) \times \text{Loss}(Q)$.
+$\alpha$ is a parameter to the query optimizer that controls the emphasis on quality versus performance.
+$\alpha = 1.0$ means that the query should be as fast as possible, $\alpha=0.0$ means that the query should be as accurate as possible.
+
+\subsection{Search space}
+To reduce the size of the query plan search space, only plans that fit the following template are considered.
+First, all filters are pushed to the leaves of the query tree, immediately after the scans.
+Joins are performed after filtering, and only left-deep plans are considered.
+Any group-by/aggregate will be performed after the final join.
+Finally, projections are placed at the root of the query tree.
+The space of query plans is similar to that considered by Selinger \todo{Selinger cite}, with the addition of imputation operators.
+
+\subsection{Imputation operators}
+We introduce two new relational operators to perform imputation: impute ($\mu$) and drop ($\delta$).
+Each operator takes arguments $(C, R)$ where $C$ is a set of attributes and $R$ is a relation.
+Impute uses a machine learning method to replace all null values with non-null values for attributes in $C$ in the relation $R$.
+Drop simply removes all tuples which have a null value for some attribute in $C$ from $R$.
+Both operators guarantee that the resulting relation will contain no null values for attributes in $C$.
+
+\subsection{Imputation placement}
+Imputation operators must be placed so that no relational operator receives a tuple containing null in an attribute that the operator examines, regardless of the state of the data in the base tables.
+
+Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required.
+To track these cases, each query plan is associated with a set of dirty attributes $D$.
+An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null.
+We compute a dirty set for each base table using the table statistics, which track the number of null values in each column.
+If we apply an imputation operator to a relation $R$ with a dirty vector $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty vector $D' = D \ C$.
+Projections produce a new dirty vector by dropping each attribute that is not in the projection.
+A join R⨝_ψ▒R^′  ⁆ with dirty vectors D⁆ and D^′  produces a dirty vector D OR D^′⁆.
+Filters do not change the dirty vector.
+[MS It's possible that a filter φ_t   removes all of the rows that had null-values in some column t` != t, thus removing t` from the dirty vector.]
+
+The dirty set over-approximates the set of attributes that contain null.
+For example, a filter might remove all tuples which contain null without changing the dirty set, forcing an unnecessary imputation.
+We choose to over-approximate the dirty set to avoid the possibility of dropping a tuple that contains a null value without explicitly imputing the value or applying a drop operator.
+
+\subsection{Query planning}
+The input to our query planner is a tuple $(T, \Phi, \Psi, P, G, A)$: a set of tables $T$, a set of filter predicates $\phi_t, t \in T$, a set of join predicates $\psi_(t_1, t_2), t_1, t_2 \in T$, a set of attributes $P$, and an optional set of attributes $G$ and aggregator function $A \in \{\text{Max}, \text{Min}, \text{Sum}, \text{Avg}, \text{Count}\}$.
+
+To reduce the search space, we only consider the minimal imputation, the maximal imputation, and the minimal drop.
+The minimal imputation (resp.
+drop) only imputes (resp.
+drops) the columns required by the relational operator immediately following the imputation.
+The maximal imputation imputes all columns in the relation, regardless of which are required.
+
+# For a plan and a set of necessary attributes, add the allowed amount of imputation.
+AddImpute(q, attrs):
+	Dattr = Dirty(q) intersect attrs
+	if Dattr = empty:
+		return {q}
+	else:
+		return {impute_all(q), impute_min(q, Dattr), drop_min(q, Dattr)}
+
+# For a set of plans with different dirty sets, return the best plan for each dirty set.
+OptRel(Q):
+	D = {Dirty(q) | q in Q}
+	return { argmin_q Cost(q) s.t. Dirty(q) = d | d in D}
+
+# Given a table and a filter predicate, return the optimal plans.
+OptFilter(t⁆,φ⁆):
+	return OptRel(AddImpute(t, Attrs(φ)))
+
+# For a set of tables (with filter predicates) and join predicates, return the optimal plans.
+OptJoin(T⁆, join_preds):
+	Q = empty
+	for t, pred in T:
+		S = T \ {t, pred}
+		# Get the optimal plans for the left and right arguments to the join.
+		for r_l, r_r in OptJoin(S) x OptFilter(t, pred):
+			if r_l, r_r in join_preds:
+				# Add imputation to the left and right arguments.
+				Q = Q union OptRel(Join(r_l', r_r') for r_l' in AddImpute(r_l, Attrs(join_pred)), r_r' in AddImpute(r_r, Attrs(join_pred)))
+	return OptRel(Q)
+# Extended cost calculation to account for information loss/computation cost in imputations
+Cost(q):
+  case drop_min(q', d) =>  info_loss(drop, q', attrs(q') intersect d) + Cost(q')
+  case impute_all(q') => info_loss(impute, q', attr(q')) + impute_cost(q') + Cost(q')
+  case impute_min(q', d) => info_loss(impute, q', attrs(q') intersect d) + impute_cost(q'[d]) + Cost(q')
+  case _ => normal cost calculation
+
+info_loss(op, t, a):
+    # beta: some scaling factor
+    case info_loss(drop, t, a) => beta * sum(for ct_missing(t.a) in a)
+    case info_loss(impute, t, a) => beta *sum(for ct_missing(t.a) / ct_complete(t) in a)
+
+impute_cost(t):
+  # alpha: some scaling factor
+  alpha * (ntuples(t) + ct(attr(t)))
+
+\subsection{Histogram computation}
+Histogram adjustment algorithm:
+Invariant: the distribution of non-null values is preserved
+Drop(t, ix) => remove nulls from t table statistics for indices ix
+Impute(t, ix) => redistribute nulls from t table statistics for indices ix to other buckets based on existing distribution
+Filter(t, predicate) => scale all buckets of t table statistics and null counts by predicate selectivity
+Join(t1, t2) => estimate cardinality of join result (following selinger, if non-equality, then 30% of product,
+If equality and no primary/foreign key relationship then the maximum of the two cardinalities, otherwise
+If equality and primary/foreign key relationship then the size ofthe foreign key relationship table).
+Scale t1's statistics and null counts to total the new cardinality estimate.
+Do same for t2.
+Union histograms.
+
+\subsection{Complexity}
 

### 2016-12-01 simpledb/src/java/simpledb/DoubleField.java
fix: minor changes to get tests to run after merger

@@ -9,14 +9,6 @@ public class DoubleField implements Field {
 	private static final long serialVersionUID = -7652884565222581467L;
 	
 	private final Double value;
-
-    public double getValue() {
-        if (isMissing()) {
-            throw new UnsupportedOperationException("cannot get value on missing");
-        }
-        return value;
-    }
-
 	
 	public DoubleField(double value) {
 		if (value == Type.MISSING_DOUBLE) {

### 2016-12-01 simpledb/src/java/simpledb/Impute.java

@@ -2,6 +2,7 @@ package simpledb;
 
 import java.util.ArrayList;
 import java.util.Collection;
+import java.util.List;
 
 public abstract class Impute extends Operator {
 
@@ -17,8 +18,18 @@ public abstract class Impute extends Operator {
         this.td = child.getTupleDesc();
         
         // Validate drop fields and convert to indices.
-        this.dropFields        = dropFields;
-        this.dropFieldsIndices = extractDropFieldsIndices(dropFields, td);
+        if (dropFields != null) {
+            this.dropFields = dropFields;
+        } else {
+            // if no fields provided, then we assume all fields are to be used
+            List<String> allFields = new ArrayList<>();
+            for(int i = 0; i < td.numFields(); i++) {
+                allFields.add(td.getFieldName(i));
+            }
+
+            this.dropFields = allFields;
+        }
+        this.dropFieldsIndices = extractDropFieldsIndices(this.dropFields, td);
     }
     
     public static Collection<Integer> extractDropFieldsIndices(Collection<String> dropFields, TupleDesc td){
@@ -61,6 +72,10 @@ public abstract class Impute extends Operator {
     @Override
     public TupleDesc getTupleDesc() {
         return td;
+    }
+
+    public Collection<String> getDropFields() {
+        return dropFields;
     }
 
 }

### 2016-12-01 simpledb/src/java/simpledb/ImputeRandom.java

@@ -29,6 +29,10 @@ public class ImputeRandom extends Impute {
         buffer = new ArrayList<>();
         nextTupleIndex = 0;
     }
+
+    public ImputeRandom(DbIterator child) {
+        this(null, child);
+    }
     
     private void initRng(){
         random = new Random(GENERATOR_SEED);

### 2016-12-01 simpledb/src/java/simpledb/LogicalAccessNode.java

@@ -76,7 +76,7 @@ public class LogicalAccessNode extends ImputedPlan {
 			if (required.isEmpty()) {
 				throw new BadImputation();
 			}
-			pp = new Drop(pp, requiredAttrs);
+			pp = new Drop(requiredAttrs, pp);
 			tableDirtySet.removeAll(required);
 			dirtySet = tableDirtySet;
 			loss = subplanTableStats.estimateTotalNull(requiredIdx);
@@ -96,7 +96,7 @@ public class LogicalAccessNode extends ImputedPlan {
 			}
 			loss = subplanTableStats.estimateTotalNull(requiredIdx) * Math.pow(LOSS_FACTOR, -totalData);
 			time = subplanTableStats.estimateScanCost() + subplanTableStats.estimateImputeCost();
-			pp = new ImputeRandom(pp, requiredAttrs);
+			pp = new ImputeRandom(requiredAttrs, pp);
 			tableDirtySet.removeAll(required);
 			dirtySet = tableDirtySet;
 			adjustedTableStats = subplanTableStats.adjustForImpute(MINIMAL, requiredIdx);

### 2016-12-01 simpledb/src/java/simpledb/LogicalComposeImputation.java

@@ -71,21 +71,21 @@ public class LogicalComposeImputation extends ImputedPlan {
 
 		switch (imp) {
 		case DROP:
-			physicalPlan = new Drop(subplan.getPlan(), toNames(impute));
+			physicalPlan = new Drop(toNames(impute), subplan.getPlan());
 			dirtySet.removeAll(impute);
 			loss = estimateNumNulls(subplan, imputeIndices);
 			time = subplan.cardinality();
 			adjustedTableStats = subplanTableStats.adjustForImpute(DROP, imputeIndices);
 			return new LogicalComposeImputation(adjustedTableStats, physicalPlan, dirtySet, loss, time);
 		case MINIMAL:
-			physicalPlan = new ImputeRandom(subplan.getPlan(), toNames(impute));
+			physicalPlan = new ImputeRandom(toNames(impute), subplan.getPlan());
 			dirtySet.removeAll(impute);
 			loss = estimateNumNulls(subplan, imputeIndices) * Math.pow(LOSS_FACTOR, -totalData);
 			time = totalData;
 			adjustedTableStats = subplanTableStats.adjustForImpute(MINIMAL, imputeIndices);
 			return new LogicalComposeImputation(adjustedTableStats, physicalPlan, dirtySet, loss, time);
 		case MAXIMAL:
-			physicalPlan = new ImputeRandom(subplan.getPlan(), toNames(subplan.getDirtySet()));
+			physicalPlan = new ImputeRandom(toNames(subplan.getDirtySet()), subplan.getPlan());
 			dirtySet.clear();
 			loss = estimateNumNulls(subplan, imputeIndices) * Math.pow(LOSS_FACTOR, -totalData);
 			time = totalData;

### 2016-12-01 paper/algorithm.tex
continue formatting algorithm section

@@
 \subsection{Imputation placement}
+\label{sec:placement}
 Imputation operators must be placed so that no relational operator receives a tuple containing null in an attribute that the operator examines, regardless of the state of the data in the base tables.
@@
 We compute a dirty set for each base table using the table statistics, which track the number of null values in each column.
-If we apply an imputation operator to a relation $R$ with a dirty vector $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty vector $D' = D \ C$.
-Projections produce a new dirty vector by dropping each attribute that is not in the projection.
-A join R⨝_ψ▒R^′  ⁆ with dirty vectors D⁆ and D^′  produces a dirty vector D OR D^′⁆.
-Filters do not change the dirty vector.
-[MS It's possible that a filter φ_t   removes all of the rows that had null-values in some column t` != t, thus removing t` from the dirty vector.]
+If we apply an imputation operator to a relation $R$ with a dirty set $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty set $D' = D \ C$.
+Applying a projection $\pi_C(R)$ produces a dirty set $D' = D \cap C$.
+A join $R_1 \Join_\psi R_2$ with dirty sets $D_1$ and $D_2$  produces a dirty set $D' = D_1 \cup D_2$.
+Filters do not change the dirty set.
 
@@
 
+The query planner must select a join ordering in addition to placing imputation operators as described in Section~\ref{sec:placement}.
+
 To reduce the search space, we only consider the minimal imputation, the maximal imputation, and the minimal drop.
@@
 
-# For a plan and a set of necessary attributes, add the allowed amount of imputation.
-AddImpute(q, attrs):
-	Dattr = Dirty(q) intersect attrs
-	if Dattr = empty:
-		return {q}
-	else:
-		return {impute_all(q), impute_min(q, Dattr), drop_min(q, Dattr)}
+\begin{algorithm}
+  \begin{algorithmic}
+    \Require{$q$ is a query plan, $C_{req}$ is a set of attributes that must be imputed.}
+    \Ensure{Returns a set of query plans such that $\Call{Dirty}{q'} \cap C_{req} = \emptyset$.}
+    \Function{AddImpute}{$q, C_{req}$}
+    \State $C_{min} \gets \Call{Dirty}{q} \cap C_{req}$
+    \If{$C_{min} = \emptyset$}
+    \State \Return $\{q\}$
+    \Else
+    \State \Return $\{\mu_{\Call{Dirty}{q}}(q), \mu_{C_{min}}(q), \delta_{C_{min}}(q)\}$
+    \EndIf
+    \EndFunction
 
-# For a set of plans with different dirty sets, return the best plan for each dirty set.
-OptRel(Q):
-	D = {Dirty(q) | q in Q}
-	return { argmin_q Cost(q) s.t. Dirty(q) = d | d in D}
+    \State
 
-# Given a table and a filter predicate, return the optimal plans.
-OptFilter(t⁆,φ⁆):
-	return OptRel(AddImpute(t, Attrs(φ)))
+    \Require{$Q$ is a set of query plans.}
+    \Ensure{Returns an optimal query plan for each distinct dirty set in $Q$.}
+    \Function{OptRel}{$Q$}
+    \State $D \gets \{\Call{Dirty}{q} ~|~ q \in Q\}$
+    \State \Return $\{\argmin_{q \in Q \land \Call{Dirty}{q} = d} \Call{Cost}{q} ~|~ d \in D\}$
+    \EndFunction
 
+    \State
+
+    \Require{$t$ is a table and $\phi$ is a filter predicate.}
+    \Ensure{Returns a set of optimal query plans for scanning and filtering $t$, with distinct dirty sets.}
+    \Function{OptFilter}{$t, \phi$}
+    \State \Return $\Call{OptRel}{\Call{AddImpute}{t, \Call{Attrs}{\phi}}}$
+    \EndFunction
+
+    \State
+
+    \Require{$Q$ is a set of query plans and $\Psi$ relates query plans with join predicates.}
+    \Function{OptJoin}{$Q, \Phi$}
+    \State $Q \gets \emptyset$
+    \For{$q \in Q$}
+
+    \EndFor
+    \EndFunction
+  \end{algorithmic}
+  \caption{An algorithm for query planning with imputations.}
+\end{algorithm}
+
+\begin{verbatim}		
 # For a set of tables (with filter predicates) and join predicates, return the optimal plans.
@@
   alpha * (ntuples(t) + ct(attr(t)))
+\end{verbatim}
 

### 2016-12-02 paper/algorithm.tex
paper: added join algo sketched, added chained equations algo sketched, added histogram transformation algo

@@
 We compute a dirty set for each base table using the table statistics, which track the number of null values in each column.
-If we apply an imputation operator to a relation $R$ with a dirty set $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty set $D' = D \ C$.
+If we apply an imputation operator to a relation $R$ with a dirty set $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty set $D' = D \setminus C$.
 Applying a projection $\pi_C(R)$ produces a dirty set $D' = D \cap C$.
@@
 
-    \Require{$Q$ is a set of query plans and $\Psi$ relates query plans with join predicates.}
+    \Require{$Q$ is a map from tables and dirty sets to optimal query plans (possibly with imputations), and $\Psi$ relates query plans with join predicates.}
+    \Ensure{Returns a set of optimal query plans for all joins, with distinct dirty sets.}
     \Function{OptJoin}{$Q, \Phi$}
-    \State $Q \gets \emptyset$
-    \For{$q \in Q$}
-
+    \For{$size \in 1...|\Psi|$}
+    	\State $S \gets subset(\Psi, size)$
+	\For{$\phi \in S$}
+		\State $S' \gets S \setminus \phi$
+			\If{$joins(S', \phi)$}
+				\State $S'_{plans} \gets Q(\Call{Rels}{S'})$
+				\State $(t, \_) \gets \Call{Rels}{\phi}$
+				\State $t_{plans} \gets Q(t)$
+				\For {$l, r \in t_{plans} \times S'_{plans}$}
+					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\phi}}$
+					\State $P_r \gets \Call{AddImpute}{r, \Call{Attrs}{\phi}}$
+					\State \Call{Update}{Q, $\Call{OptRel}{ \{p_l \bowtie_{\phi} p_r | p_l \in P_l, p_r \in P_r \} }$}
 				\EndFor
+		\EndIf
+	\EndFor
+    \EndFor
+    \State \Return{Q(\Call{Rels}{$\Phi$})}
     \EndFunction
@@
 
-\begin{verbatim}		
-# For a set of tables (with filter predicates) and join predicates, return the optimal plans.
-OptJoin(T⁆, join_preds):
-	Q = empty
-	for t, pred in T:
-		S = T \ {t, pred}
-		# Get the optimal plans for the left and right arguments to the join.
-		for r_l, r_r in OptJoin(S) x OptFilter(t, pred):
-			if r_l, r_r in join_preds:
-				# Add imputation to the left and right arguments.
-				Q = Q union OptRel(Join(r_l', r_r') for r_l' in AddImpute(r_l, Attrs(join_pred)), r_r' in AddImpute(r_r, Attrs(join_pred)))
-	return OptRel(Q)
-# Extended cost calculation to account for information loss/computation cost in imputations
-Cost(q):
-  case drop_min(q', d) =>  info_loss(drop, q', attrs(q') intersect d) + Cost(q')
-  case impute_all(q') => info_loss(impute, q', attr(q')) + impute_cost(q') + Cost(q')
-  case impute_min(q', d) => info_loss(impute, q', attrs(q') intersect d) + impute_cost(q'[d]) + Cost(q')
-  case _ => normal cost calculation
-
-info_loss(op, t, a):
-    # beta: some scaling factor
-    case info_loss(drop, t, a) => beta * sum(for ct_missing(t.a) in a)
-    case info_loss(impute, t, a) => beta *sum(for ct_missing(t.a) / ct_complete(t) in a)
-
-impute_cost(t):
-  # alpha: some scaling factor
-  alpha * (ntuples(t) + ct(attr(t)))
-\end{verbatim}
+%\begin{verbatim}		
+%# For a set of tables (with filter predicates) and join predicates, return the optimal plans.
+%OptJoin(T⁆, join_preds):
+%	Q = empty
+%	for t, pred in T:
+%		S = T \ {t, pred}
+%		# Get the optimal plans for the left and right arguments to the join.
+%		for r_l, r_r in OptJoin(S) x OptFilter(t, pred):
+%			if r_l, r_r in join_preds:
+%				# Add imputation to the left and right arguments.
+%				Q = Q union OptRel(Join(r_l', r_r') for r_l' in AddImpute(r_l, Attrs(join_pred)), r_r' in AddImpute(r_r, Attrs(join_pred)))
+%	return OptRel(Q)
+%# Extended cost calculation to account for information loss/computation cost in imputations
+%Cost(q):
+%case drop_min(q', d) =>  info_loss(drop, q', attrs(q') intersect d) + Cost(q')
+%case impute_all(q') => info_loss(impute, q', attr(q')) + impute_cost(q') + Cost(q')
+%case impute_min(q', d) => info_loss(impute, q', attrs(q') intersect d) + impute_cost(q'[d]) + Cost(q')
+%case _ => normal cost calculation
+%
+%
+%info_loss(op, t, a):
+%# beta: some scaling factor
+%case info_loss(drop, t, a) => beta * sum(for ct_missing(t.a) in a)
+%case info_loss(impute, t, a) => beta *sum(for ct_missing(t.a) / ct_complete(t) in a)
+%
+%impute_cost(t):
+%# alpha: some scaling factor
+%alpha * (ntuples(t) + ct(attr(t)))
+%\end{verbatim}
 
 \subsection{Histogram computation}
-Histogram adjustment algorithm:
-Invariant: the distribution of non-null values is preserved
-Drop(t, ix) => remove nulls from t table statistics for indices ix
-Impute(t, ix) => redistribute nulls from t table statistics for indices ix to other buckets based on existing distribution
-Filter(t, predicate) => scale all buckets of t table statistics and null counts by predicate selectivity
-Join(t1, t2) => estimate cardinality of join result (following selinger, if non-equality, then 30% of product,
-If equality and no primary/foreign key relationship then the maximum of the two cardinalities, otherwise
-If equality and primary/foreign key relationship then the size ofthe foreign key relationship table).
-Scale t1's statistics and null counts to total the new cardinality estimate.
-Do same for t2.
-Union histograms.
+In order to correctly estimate the imputation cost and computation time parameters in our cost model,  the system must have updated cardinality estimates
+at each point in a query plan. These cardinality estimates can be impacted not just be filtering or joining, as in the traditional relational calculus, but also by
+the imputation operators. For example,  dropping versus imputing tuples has a clearly different effect on cardinality. For simplicity of design,
+each of the logical nodes in a query plan keeps a reference to a set of histograms. The optimizer creates a copy-on-write of the histograms and modifies them
+as necessary to account for new operations in the plan. In all cases, modifications maintain as an invariant that the distribution of data in the new histogram
+matches the distribution (but not necessarily the cardinality) of the original statistics. Algorithm~\ref{algo:histogram-transformation} provides the details
+of our implementation.
+
+\begin{algorithm}
+  \begin{algorithmic}
+    \Require{$H$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
+    \Ensure{Returns an updated $H$, such that the distribution of data matches the original $H$}
+
+    \Function{UpdateHistogram}{$H, op$}
+    	\If{$op \not\in \{\delta, \mu, \sigma, \bowtie \}$}
+		\Return{H}
+	\EndIf
+	\State $H' \gets \Call{Copy}{H}$
+    	\If{op = $\delta_C$}
+		\For{$c \in C$}
+			\State $H'[c][null] \gets 0$
+		\EndFor
+	\ElsIf{op = $\mu_C$}
+		\For{$c \in C$}
+			\State \Call{Add}{$H'[c], H'[c][null]$}
+		\EndFor
+	\ElsIf{op = $\sigma$}
+		\State \Call{ScaleBy}{H',  \Call{Selectivity}{$\sigma$}}
+	\ElsIf{op = $\bowtie_\phi$}
+		\State \Call{ScaleTo}{H', \Call{Cardinality}{$\bowtie_{\phi}$}}
+	\EndIf
+	\Return{H'}
+    \EndFunction
+  \end{algorithmic}
+  \caption{An algorithm for in-plan histogram updates}
+  \label{algo:histogram-transformation}
+\end{algorithm}
+
+\subsection{Imputation Strategies}
+The current version of the system uses an imputation strategy based on chained-equations regression trees (CERT) \todo{CITE}. Note that plugging in a new
+imputation strategy takes minimal effort, and doesn't require changes to the optimizer, so the system could in principle be tuned to a specific domain.
+The imputation algorithm is iterative in nature, and proceeds by repeatedly fitting regression trees to a subset of the data and the target imputation columns.
+In each iteration, the missing values of a column are replaced with newly imputed values. As the iterations proceed, the quality of the imputation
+should improve as values progressively reflect more accurate relationships amongst attributes. Algorithm~\ref{algo:imputation-strategy}
+provides details on the implementation.
+
+It is worth noting that given the nature of the imputation strategy used here, imputation operators become blocking in our system.
+
+\begin{algorithm}
+  \begin{algorithmic}
+    \Require{$T$ is a table. $D$ is a set of $T$ attributes that need to be imputed, and $C$ is a set of $T$ attributes that has complete data }
+    \Ensure{Returns an imputed $T$}
+
+    \Function{ImputeWithCERT}{$H, op$}
+    	\State $T' \gets \Call{ImputeRandom}{T, D}$
+	\For{$n\ EPOCHS$}
+		\For{$d \in D$}
+			\State $imp \gets \Call{TrainRT}{\pi_C(T'), \pi_d{T'}}$
+			\State $T' \gets (\pi_C(T'), \Call{PredictRT}{imp, T})$
+		\EndFor
+	\EndFor
+	\Return{T'}
+	\EndFunction
+  \end{algorithmic}
+  \caption{An algorithm for in-plan histogram updates}
+  \label{algo:histogram-transformation}
+\end{algorithm}
+
+\todo{Jose:Micah, take a look at the explanation above and see if it's missing anything (or is incorrect)}
 

### 2016-12-05 paper/relatedwork.tex
paper: added related work

--- 2016-11-28 paper/algorithm.tex
@@
-\section{Algorithm}
+\section{Related Work}
+\subsection{Missing Values and Statistics}
+Imputation of missing values is a widely studied field within the statistics and machine
+learning communities. As highlighted in \cite{gelman2006data}, missing data
+can appear for a variety of reasons, including both random and conditioned on
+existing values (observed and missing). Methods in the statistical community
+focus on correctly modeling relationships between the attributes to factor in
+such varied forms of missingness.
+
+Burgette and Reiter introduce imputation of missing data using sequences
+of regression trees in \cite{burgette2010multiple}. \ProjectName{} uses
+this technique to impute missing values as needed in the generated query plan.
+The flexibility of this non-parametric imputation approach aligns well with the
+spirit of query optimization, which aims to reconcile performance with a declarative interface to
+data manipulation. By removing concerns for imputation, users executing queries
+in \ProjectName{} don't need to consider the potential nature of the missing data.
+
+In \cite{akande2015empirical}, Akande et al analyze the performance of various
+imputation techniques on American Community Survey data, "an ongoing survey
+that provides vital information on a yearly basis"\cite{censusbureau}. The authors
+use a common technique: multiple imputation, which imputes missing values
+with various strategies on the base data, computes aggregates, and then
+combines estimates of these aggregates to account for additional uncertainty across
+strategies. The computational difficulties of imputing on large base tables
+are well known and can limit approaches. For example, Akande finds that
+one of their approaches (MI-GLM) crashes when attempting to impute on data that
+includes variables with potentially large domains (ten categories specifically).
+ In contrast, \ProjectName{} allows users to specify a tradeoff
+between information loss and computational complexity (in terms of time). 
+Furthermore, the query planner's imputation is guided by the requirements of
+each specific query's operators, rather than requiring broad assumptions
+about query workloads.
+
+\subsection{Missing Values and Databases}
+There is a long history in the database community surrounding the
+treatment of nulls. As early as 1973, \cite{codd1973understanding}
+provides a treatment of the semantics of null. Multiple
+papers have described various (at times conflicting) treatments
+of nulls \cite{grant1977null}. \ProjectName's main design invariant - that no relational operator
+see attributes with missing values for attributes it must operate on - eliminates
+the need to handle null value semantics, while guaranteeing soundness (modulo
+imputation strategy) of the query evaluation.
+
+BayesDB \cite{mansinghka2015bayesdb} provides users with a simple interface to 
+leverage statistical inference techniques on a database. Non-experts
+can use a simple declarative language, akin to SQL, to specify models
+which allow missing value imputation, amongst other broader functionality.
+Experts can further customize strategies and express domain knowledge to
+improve performance and accuracy.
+
+While BayesDB can be used for value imputation, this step is not framed
+within the context of query planning, but rather as an explicit step, using the
+\verb|INFER| operation.
+
+BayesDB provides a great alternative for bridging the gap between
+traditional databases and sophisticated modeling software. \ProjectName{}, in
+contrast, aims to remain squarely in the database realm, while allowing
+users to perform queries on a potentially larger subset of their data.
+
+\ProjectName's cost-based query planner 
+is partially based on the seminal work developed for System R's query planning\cite{blasgen1981system}.
+However, in contrast to System R, \ProjectName{} performs additional histogram transformations to account
+for the changing nature of missing values.
 

### 2016-12-06 paper/introduction.tex
paper: first pass at intro

--- 2016-11-28 paper/algorithm.tex
@@
-\section{Algorithm}
+\section{Introduction}
+
+Handling incorrect or dirty data is a complex and challenging problem for data scientists.
+One way in which a dataset can be dirty is for parts of it to be missing altogether.
+Missing data is one of the simpler variants of dirty data, but if handled naively, it can still cause analyses to be incorrect.
+To handle this problem, users may manually clean their dataset by performing some regression analysis to replace missing data elements with likely values.
+This process is called \emph{imputation}.
+Although manual imputation may solve the problem of missing data, an imputation algorithm may be too complex for most users to implement, and may also be computationally expensive to run on a large dataset.
+Additionally, users may have different requirements for their queries: some users may need queries to be fast, regardless of accuracy while others prefer accurate results and are willing to wait for them.
+
+In this paper, we present \ProjectName, a database system which is designed to interact with a dirty dataset as though it were clean.
+The guiding design principle of \ProjectName is that the user should never see missing data or have to modify their queries to account for it.
+To achieve this goal, we perform imputation on the fly, during query execution.
+We find that performing imputation during query execution allows our system to flexibly trade accuracy for computation time.
 

### 2016-12-07 paper/introduction.tex
paper: clean up in intro and relatedwork

@@
 
-In this paper, we present \ProjectName, a database system which is designed to interact with a dirty dataset as though it were clean.
-The guiding design principle of \ProjectName is that the user should never see missing data or have to modify their queries to account for it.
+In this paper, we present \ProjectName{}, a database system which is designed to interact with a dirty dataset as though it were clean.
+The guiding design principle of \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
 To achieve this goal, we perform imputation on the fly, during query execution.

### 2016-12-07 paper/relatedwork.tex
paper: clean up in intro and relatedwork

@@
 of nulls \cite{grant1977null}. \ProjectName's main design invariant - that no relational operator
-see attributes with missing values for attributes it must operate on - eliminates
+see attributes with missing values for attributes it must operate on and users should never see
+missing data - eliminates
 the need to handle null value semantics, while guaranteeing soundness (modulo

### 2016-12-07 paper/algorithm.tex
paper: fixed imputation algo label
paper: added top-level query planner algo

@@
 Finally, projections are placed at the root of the query tree.
-The space of query plans is similar to that considered by Selinger \todo{Selinger cite}, with the addition of imputation operators.
+The space of query plans is similar to that considered by Selinger \todo{Selinger cite}, with the addition of imputation operators appearing before/after traditional operators.
 
@@
 Impute uses a machine learning method to replace all null values with non-null values for attributes in $C$ in the relation $R$.
-Drop simply removes all tuples which have a null value for some attribute in $C$ from $R$.
+Drop simply removes from $R$ all tuples which have a null value for some attribute in $C$.
 Both operators guarantee that the resulting relation will contain no null values for attributes in $C$.
@@
 To track these cases, each query plan is associated with a set of dirty attributes $D$.
-An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null.
+An attribute $c$ is \emph{dirty} in some relation if the values for $c$ may contain null.
 We compute a dirty set for each base table using the table statistics, which track the number of null values in each column.
@@
 The maximal imputation imputes all columns in the relation, regardless of which are required.
+Algorithm~\ref{algo:plan-helpers} presents
+a series of helper functions used by the top level planner, shown in Algorithm~\ref{algo:top-level-planner}.
 
@@
     \Function{AddImpute}{$q, C_{req}$}
+    \todo{jose: I modified this, since in reality we return only the same plan if it has no dirty values, we may still impute when the intersection is empty, but only maximally}
     \State $C_{min} \gets \Call{Dirty}{q} \cap C_{req}$
-    \If{$C_{min} = \emptyset$}
-    \State \Return $\{q\}$
+    \If{$\Call{Dirty}{q} = \emptyset$}
+    	\State \Return $\{ q \}$
+   \ElsIf{$C_{min} = \emptyset$}
+    	\State \Return $\{ \mu_{\Call{Dirty}{q}}(q), q \}$
     \Else
@@
     \Function{OptFilter}{$t, \phi$}
-    \State \Return $\Call{OptRel}{\Call{AddImpute}{t, \Call{Attrs}{\phi}}}$
+    \todo{jose: I believe this was missing the sigma before}
+    \State \Return $\Call{OptRel}{\{  \sigma_{\phi}(q) | q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}} \} }$
     \EndFunction
@@
     \Ensure{Returns a set of optimal query plans for all joins, with distinct dirty sets.}
-    \Function{OptJoin}{$Q, \Phi$}
+    \Function{OptJoin}{$Q, \Psi$}
     \For{$size \in 1...|\Psi|$}
     	\State $S \gets subset(\Psi, size)$
-	\For{$\phi \in S$}
-		\State $S' \gets S \setminus \phi$
-			\If{$joins(S', \phi)$}
+	\For{$\psi \in S$}
+		\State $S' \gets S \setminus \psi$
+			\If{$joins(S', \psi)$} 
 				\State $S'_{plans} \gets Q(\Call{Rels}{S'})$
-				\State $(t, \_) \gets \Call{Rels}{\phi}$
+				\State $t \gets \Call{Rels}{\psi} \setminus \Call{Rels}{S'}$
 				\State $t_{plans} \gets Q(t)$
 				\For {$l, r \in t_{plans} \times S'_{plans}$}
-					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\phi}}$
-					\State $P_r \gets \Call{AddImpute}{r, \Call{Attrs}{\phi}}$
-					\State \Call{Update}{Q, $\Call{OptRel}{ \{p_l \bowtie_{\phi} p_r | p_l \in P_l, p_r \in P_r \} }$}
+					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\psi}}$
+					\State $P_r \gets \Call{AddImpute}{r, \Call{Attrs}{\psi}}$
+					\State \Call{Update}{Q, $\Call{OptRel}{ \{p_l \bowtie_{\psi} p_r | p_l \in P_l, p_r \in P_r \} }$}
 				\EndFor
@@
   \end{algorithmic}
-  \caption{An algorithm for query planning with imputations.}
+  \caption{Base algorithms for query planning with imputations.}
+  \label{algo:plan-helpers}
+\end{algorithm}
+
+\begin{algorithm}
+ \begin{algorithmic}
+   \Require{set of tables $T$, filter predicates $\Phi$, join predicates $\Psi$, a set of projection attributes $P$, optional set of grouping attributes $G$ and aggregator function $A$}
+   \Ensure{Returns optimized query plan.}
+
+  \Function{Plan}{$T, \Phi, \Psi, P, G, A$}
+ \State  $Q \gets \{ \}$ \Comment{map from table and dirty set to plans }
+   \For{$t \in T$}
+   	\State $\phi \gets \Call{GetFilters}{t, \Phi}$
+	\State $P_{t} \gets \Call{OptFilter}{t, \phi}$
+	\State $Q \gets Q \cup \{ (t, \Call{Dirty}{q}) \mapsto q | q \in P_t \}$
+   \EndFor
+
+   \State $M \gets \{ \}$ \Comment{map from dirty set to plans, contains all joins necessary}
+   \State $M \gets  \Call{OptJoin}{Q, \Psi}$	\Comment{optimize joins}
+   \If{$G \neq \emptyset \land A \neq NULL$}
+   	\State $attrs \gets \Call{Attrs}{G} \cup \Call{Attrs}{A}$
+   	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \Call{GroupBy}{\Call{AddImpute}{q, attrs}, G, A}}$
+   \Else
+  	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \pi_{P}(\Call{AddImpute}{q, P}) }$
+  \EndIf
+
+\Return $\argmin_{q \in M} Cost(q)$
+\EndFunction
+
+  \end{algorithmic}
+  \caption{Top-level query planner with imputations.}
+\label{algo:top-level-planner}
 \end{algorithm}
@@
   \end{algorithmic}
-  \caption{An algorithm for in-plan histogram updates}
-  \label{algo:histogram-transformation}
+  \caption{An algorithm for chained imputation using regression trees}
+  \label{algo:imputation-strategy}
 \end{algorithm}

### 2016-12-07 paper/algorithm.tex
paper: a fluffy complexity section

@@
 \subsection{Complexity}
+Our optimization algorithm builds off the approach taken by System R\cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
+note that if we remove our restriction on types of imputation($\delta_{min}, \mu_{min}, \mu_{max}$), and allow any arbitrary subset of attributes to be imputed,
+then every single operator in the query plan has a number of imputations exponential in the number of dirty columns. Our restriction, instead, increases the number
+of plans (in the worst case) at each operator by a factor of 3. Of course, this implies that in the worst case (where all dirty sets tracked are distinct throughout the query plan),
+we explore a number of plans that further scales the number of plans in the original algorithm by an exponential factor.
+
+However, we note that in all cases we have considered, this exponential blowup does not affect the practical performance of our optimizer.
 

### 2016-12-07 paper/gen_tables.py
paper: add code to generate tables

@@ -1,0 +1,84 @@
+#!/usr/bin/env python3
+
+import pandas as pd
+import sys
+
+def parse_time(t):
+    return t[2:][:-1]
+
+def err_to_str(err):
+    if err == 'null':
+        return r'$\bot$'
+    else:
+        return str(err)
+
+def get_first_row(df, l):
+    for (_, row) in df.iterrows():
+        if l(row):
+            return row
+
+def writel(f, l):
+    f.write(l + '\n')
+
+def main():
+    if len(sys.argv) != 2:
+        print('Usage: gen_tables.py QUERIES')
+        exit(-1)
+        
+    df = pd.read_csv(sys.argv[1])
+
+    with open('queries.tex', 'w') as f:
+        qidx = {}
+        idxq = {}
+        ctr = 0
+        writel(f, r'\begin{tabular}{ll}')
+        writel(f, r'\toprule')
+        writel(f, r'\# & Query \\')
+        writel(f, r'\midrule')
+        for query in df['query'].unique():
+            writel(f, r'%d & \verb|%s| \label{q%d} \\' % (ctr, query, ctr))
+            qidx[query] = ctr
+            idxq[ctr] = query
+            ctr += 1
+        writel(f, r'\bottomrule')
+        f.write(r'\end{tabular}' + '\n')
+
+    with open('runtimes.tex', 'w') as f:
+        f.write(r'\begin{tabular}{llllllll}' + '\n')
+        writel(f, r'\toprule')
+        writel(f, r'\multicolumn{2}{c}{} & \multicolumn{3}{c}{Imputed ($\alpha=0.0$)} & \multicolumn{3}{c}{Imputed ($\alpha=1.0$)} \\')
+        writel(f, r'\cmidrule(r){3-5}')
+        writel(f, r'\cmidrule(l){6-8}')
+        writel(f, r'Query \# & Base error & $\epsilon$ & $t_p$ & $t_r$ & $\epsilon$ & $t_p$ & $t_r$ \\')
+        writel(f, r'\midrule')
+        for query in df['query'].unique():
+            ctr = qidx[query]
+            
+            a0_row = get_first_row(df, lambda row: row['query'] == query and row['alpha'] == 0.0)
+            base_err = a0_row['base_err']
+            a0_err = a0_row['impute_err']
+            a0_ptime = parse_time(a0_row['plan_time'])
+            a0_rtime = parse_time(a0_row['run_time'])
+
[…]

### 2016-12-07 simpledb/test/acstest/AcsTestRule.java
feat: add runner to generate data for paper tables

@@ -62,9 +62,9 @@ public class AcsTestRule implements TestRule {
 		SCHEMA = new TupleDesc(types, fields);
 	}
 	
-	private class DbInitializer extends ExternalResource {
+	public class DbInitializer extends ExternalResource {
 		@Override 
-		protected void before() {
+		public void before() {
 			Database.getCatalog().clear();
 			ClassLoader loader = CleanTest.class.getClassLoader();
 			File acsData = new File(loader.getResource("testdata/acs.dat").getFile());
@@ -73,7 +73,7 @@ public class AcsTestRule implements TestRule {
 		}
 	}
 	
-	private final DbInitializer initializer = new DbInitializer();
+	public final DbInitializer initializer = new DbInitializer();
 
 	@Override
 	public Statement apply(Statement arg0, Description arg1) {

### 2016-12-07 simpledb/test/acstest/RunQueries.java

@@ -1,0 +1,134 @@
+package acstest;
+
+import java.io.ByteArrayInputStream;
+import java.io.File;
+import java.io.FileWriter;
+import java.io.IOException;
+import java.time.Duration;
+import java.time.Instant;
+import java.util.NoSuchElementException;
+import java.util.function.Function;
+
+import org.junit.Assert;
+
+import Zql.ParseException;
+import Zql.ZQuery;
+import Zql.ZStatement;
+import Zql.ZqlParser;
+import simpledb.Database;
+import simpledb.DbException;
+import simpledb.DbIterator;
+import simpledb.HeapFile;
+import simpledb.ImputedLogicalPlan;
+import simpledb.Insert;
+import simpledb.LogicalPlan;
+import simpledb.Parser;
+import simpledb.ParsingException;
+import simpledb.SeqScan;
+import simpledb.TableStats;
+import simpledb.TransactionAbortedException;
+import simpledb.TransactionId;
+import simpledb.TupleDesc;
+import simpledb.Type;
+
+public class RunQueries {
+	private static final String OUT_FN = "queries.csv";
+	
+	private static FileWriter outFile;
+	
+	private static DbIterator planQuery(String query, Function<Void, LogicalPlan> planFactory) throws ParseException, TransactionAbortedException, DbException, IOException, ParsingException {
+		ZqlParser p = new ZqlParser(new ByteArrayInputStream(query.getBytes("UTF-8")));
+		ZStatement s = p.readStatement();
+		Parser pp = new Parser(planFactory);
+		return pp.handleQueryStatement((ZQuery)s, new TransactionId()).getPhysicalPlan();
+	}
+	
+	private static void drain(DbIterator iter) throws DbException, TransactionAbortedException {
+		while (iter.hasNext()) {
+			try {
+				iter.next();
+			} catch(NoSuchElementException e) {
+				e.printStackTrace();
+			}
+		}
+	}
+	
+	private static void runQuery(String query) throws IOException, ParseException, TransactionAbortedException, DbException, ParsingException {
+		String dirtyQuery = query.replaceAll("acs", "acs_dirty");
+		
+		Instant startPlan, endPlan, startRun, endRun;
+		
+		System.err.println("Planning queries...");
+		DbIterator clean = planQuery(query, x -> new LogicalPlan());
[…]

### 2016-12-08 paper/conclusion.tex
add paper skeleton
paper: conclusion

--- 2016-11-28 paper/algorithm.tex
@@
-\section{Algorithm}
+\section{Conclusion}
+
+As shown in \ProjectName{}, missing values and their imputation can successfully be integrated into the relational calculus and
+existing plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
+learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
+information loss and time. In effect, user preferences for one or the other can be easily incorporated by adjusting a parameter.
+Simple histogram transformations
+provide incrementally updated cardinality estimates across operators in the plans, which allow us to provide more accurate cost estimates.
+By taking a dynamic programming approach, we can consider a variety of operator placements
+and input columns, while keeping planning tractable in real-world examples.
+
+In our experiments, we considered a series of analytic queries (using relevant aggregates) and set queries (with simple projections) on real-world
+American Community Survey data
+and showed that the chosen query plan improves accuracy of results relative to simply ignoring missing values. Furthermore,
+the variety of imputation operator placements across queries further emphasizes the lack of flexibility imposed by the coarse-grained
+pre-processing approach to imputation. By allowing different imputation strategies for different queries, we
+take a fine-grained approach to missing data, better addressing the often differing needs of database users.
+
+We highlight the long history of dealing with missing data both in the statistical learning and database communities.
+Similarly to existing work, we consider the impact of null values in databases and develop a simple set of invariants to 
+successfully plan around them. In contrast to the statistical learning work, our emphasis is not on the specific algorithm
+used to impute but rather on the timing of imputation in the execution of a query. In contrast to existing database work,
+we incorporate imputation into a cost-based optimizer and hide any details
+regarding missing values inside the system, allowing users to use traditional SQL and engage in normal workloads.
+
+\subsection{Future Work}
+\ProjectName{} opens up multiple avenues for further work. For instance, we could extend our minimal/maximal imputation
+operators to consider global information, such as the specific columns needed in operators higher up in the query plan.
+Our optimizer uses the same machine learning algorithm in all instances of imputation operators. Integrating multiple
+possible algorithms could allow for further fine-grained imputation. Finally, the underlying database for \ProjectName{}
+is far from a full-featured, production database, so a natural step is to integrate imputation planning in a widely-adopted
+production-quality database. This will likely expose further opportunities for improvement.
 

### 2016-12-10 paper/experiments.tex
paper: add code to generate tables
paper: add to experiments section

--- 2016-11-28 paper/algorithm.tex
@@
-\section{Algorithm}
+\section{Experiments}
+
+To evaluate the performance of our system, we generated plans for queries on two data sets: the American Community Survey and a synthetic data set containing business information.\todo{Synthetic queries?}
+
+\subsection{Data sets}
+
+The American Community Survey provides a number of public data sets collected by the U.S. Census Bureau.
+We used a cleaned version of the 2012 Public Use Microdata Sample (PUMS) data kindly provided by the authors of~\cite{akande2015empirical}.
+The cleaning procedure is described in detail in~\cite{akande2015empirical}, but to summarize, the following data were removed:
+\begin{itemize}
+\item Rows corresponding to vacant houses or single occupant households.
+\item Identification variables (e.g. area codes, serial numbers).
+\item Flag variables.
+\item Continuous variables.
+\end{itemize}
+The final data set consists of a single table with 671,153 rows and 35 columns, where all column entries are integers.
+
+\todo{Describe synthetic data}
+
+\subsection{Queries}
+
+We collected a set of queries that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
+
+The queries on the ACS data all consist only of selections, filters, and aggregations.
+The ACS data is contained in a single table, so there are relatively few join queries that are interesting on this data set.
+However, even with select, filter, and aggregate, there are interesting choices to make with imputation placement.
+
+\begin{figure}
+  \centerfloat
+  \input{queries}
+  \caption{The queries used in our experiments.}
+\end{figure}
+
+\begin{figure}
+  \centerfloat
+  \input{runtimes}
+  \caption{Running time and error for queries with different imputation levels.}
+\end{figure}
 

### 2016-12-10 paper/queries.tex
paper: add dummy experiments data

\begin{tabular}{ll}
\toprule
\# & Query \\
\midrule
0 & \verb|SELECT BLD FROM acs;| \label{q0} \\
1 & \verb|SELECT BLD as units_in_structure, COUNT(ST) as estimate FROM acs GROUP BY BLD;| \label{q1} \\
2 & \verb|SELECT BATH as has_bath, COUNT(ST) as ct FROM acs GROUP BY BATH;| \label{q2} \\
3 & \verb|SELECT ACR as lotsize, AVG(BDSP) as avg_num_bedrooms FROM acs GROUP BY ACR;| \label{q3} \\
4 & \verb|SELECT PSF as has_sub_families, SUM(NP) as num_people FROM acs GROUP BY PSF;| \label{q4} \\
5 & \verb|SELECT AVG(NP) as avg_num_people FROM acs;| \label{q5} \\
6 & \verb|SELECT BDSP as num_bedrooms, AVG(ACR) as avg_lot_size FROM acs WHERE VEH >= 2 GROUP BY BDSP;| \label{q6} \\
\bottomrule
\end{tabular}

### 2016-12-10 simpledb/src/java/simpledb/BadErrorException.java
fix: add uncommitted file

@@ -1,0 +1,5 @@
+package simpledb;
+
+public class BadErrorException extends Exception {
+	private static final long serialVersionUID = -2914912959397949282L;
+}

### 2016-12-10 simpledb/src/java/simpledb/ImputedLogicalPlan.java
feat: add logging to imputation cache

@@ -1,5 +1,7 @@
 package simpledb;
 
+import java.io.File;
+import java.io.FileNotFoundException;
 import java.util.*;
 import java.util.Map.Entry;
 
@@ -307,7 +309,13 @@ public class ImputedLogicalPlan extends LogicalPlan {
 	@Override
 	public DbIterator physicalPlan(TransactionId tid, Map<String, TableStats> baseTableStats, boolean explain)
 			throws ParsingException {
-		final ImputedPlanCache cache = new ImputedPlanCache();
+		LoggedImputedPlanCache cache = null;
+		try {
+			cache = new LoggedImputedPlanCache(new File("query_plans.log"));
+		} catch (FileNotFoundException e1) {
+			e1.printStackTrace();
+			throw new RuntimeException(e1);
+		}
 		optimizeFilters(tid, cache);
 		optimizeJoins(tid, cache);
 

### 2016-12-10 simpledb/src/java/simpledb/LoggedImputedPlanCache.java

@@ -1,0 +1,79 @@
+package simpledb;
+
+import java.io.File;
+import java.io.FileNotFoundException;
+import java.io.FileOutputStream;
+import java.io.PrintWriter;
+import java.util.Set;
+
+public class LoggedImputedPlanCache extends ImputedPlanCache {
+	private final PrintWriter writer;
+	private final QueryPlanVisualizer viz = new QueryPlanVisualizer();
+	
+	public LoggedImputedPlanCache(File log) throws FileNotFoundException {
+		super();
+		writer = new PrintWriter(new FileOutputStream(log, true));
+	}
+	
+	private <T> void writeSet(Set<T> s) {
+		int i = 0;
+		writer.print("{");
+		for (T e : s) {
+			writer.print(" ");
+			writer.print(e);
+			if (i < s.size()) {
+				writer.print(",");
+			} else {
+				writer.print(" ");
+			}
+			i++;
+		}
+		writer.print("}");
+	}
+	
+	@Override
+	void addPlan(Set<String> ts, Set<QuantifiedName> dirtySet, ImputedPlan newPlan, double lossWeight) {
+		super.addPlan(ts, dirtySet, newPlan, lossWeight);
+		
+		writer.print("Tables: ");
+		writeSet(ts);
+		writer.print("\n");
+		
+		writer.print("Dirty: ");
+		writeSet(dirtySet);
+		writer.print("\n");
+		
+		writer.format("Cost: %f\n", newPlan.cost(lossWeight));
+		
+		try {
+			writer.print(viz.getQueryPlanTree(newPlan.getPlan()));
+		} catch (ArrayIndexOutOfBoundsException e) {
+		}
+		writer.println();
+		
+		writer.flush();
+	}
+	
+	@Override
+	void addJoinPlan(Set<String> ts, Set<QuantifiedName> dirtySet, Set<LogicalJoinNode> joins, ImputedPlan newPlan, double lossWeight) {
+		super.addJoinPlan(ts, dirtySet, joins, newPlan, lossWeight);
+		
+		writer.print("Tables: ");
+		writeSet(ts);
+		writer.print("\n");
+		
+		writer.print("Dirty: ");
+		writeSet(dirtySet);
+		writer.print("\n");
+		
+		writer.format("Cost: %f\n", newPlan.cost(lossWeight));
+		
+		try {
+			writer.print(viz.getQueryPlanTree(newPlan.getPlan()));
[…]

### 2016-12-10 paper/algorithm.tex
paper: skeleton experiments

@@
 Finally, projections are placed at the root of the query tree.
-The space of query plans is similar to that considered by Selinger \todo{Selinger cite}, with the addition of imputation operators appearing before/after traditional operators.
+The space of query plans is similar to that considered by System R\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
 
@@
 \subsection{Imputation Strategies}
-The current version of the system uses an imputation strategy based on chained-equations regression trees (CERT) \todo{CITE}. Note that plugging in a new
+The current version of the system uses an imputation strategy based on chained-equations regression trees (CERT)\cite{burgette2010multiple}. Note that plugging in a new
 imputation strategy takes minimal effort, and doesn't require changes to the optimizer, so the system could in principle be tuned to a specific domain.

### 2016-12-10 paper/introduction.tex
paper: another pass over intro

@@
 This process is called \emph{imputation}.
-Although manual imputation may solve the problem of missing data, an imputation algorithm may be too complex for most users to implement, and may also be computationally expensive to run on a large dataset.
-Additionally, users may have different requirements for their queries: some users may need queries to be fast, regardless of accuracy while others prefer accurate results and are willing to wait for them.
+Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
+Although manual imputation solves the problem of missing data, in the age of big data, it may be very expensive to run an imputation algorithm on an entire dataset.
+Additionally, it may not even be necessary to completely clean the data to make it usable.
+Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
+Others may want to run queries on a subset of the data, so do not need to impute every field of every record.
 
@@
 To achieve this goal, we perform imputation on the fly, during query execution.
-We find that performing imputation during query execution allows our system to flexibly trade accuracy for computation time.
+Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows our system to flexibly trade accuracy for computation time.
 

### 2016-12-10 paper/algorithm.tex
paper: partial pass over algorithm section. add more cost model info

@@
 
-At a high level, our goal is to select a query plan that minimizes a metric which combines the cost of the query and the quality of the results.
-Precisely, we will find a query plan $Q$ that minimizes $\text{Cost}(Q) = \alpha \times \text{Time}(Q) + (1 - \alpha) \times \text{Loss}(Q)$.
+Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric which combines the cost of the query and the quality of the results.
+This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
+Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value.
+
+\subsection{Cost model}
+\label{sec:cost-model}
+We rank query plans $Q$ using a cost model $\text{Cost}(Q) = \alpha \times \text{Time}(Q) + (1 - \alpha) \times \text{Loss}(Q)$.
+$\text{Time}(Q)$ is an estimate of the runtime of the query which is derived from table statistics and selectivity estimation of the query predicates.
+$\text{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure.
 $\alpha$ is a parameter to the query optimizer that controls the emphasis on quality versus performance.
@@
 
+In order to correctly estimate $\text{Time}(Q)$ and $\text{Loss}(Q)$, the system must have cardinality estimates
+for each sub-query in each query plan.
+These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
+For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
+For simplicity, each of the logical nodes in a query plan refers to a set of histograms.
+When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
+Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
+
+The computation of $\text{Time}(Q)$ is standard and is comparable to the cost model used in any modern RDBMS.\@ Our runtime cost model is perhaps less sophisticated than most, as it is identical to the cost model in SimpleDB.
+
+$\text{Loss}(Q)$ is computed as follows.
+For each sub-query, in addition to a histogram, we keep track of the estimated number of null values in each column.
+\begin{align*}
+  \text{Loss}(Q) = \begin{cases}
+    \sum_{c \in C} \text{Null}(c, Q') & Q = \delta_C(Q') \\
+    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{Null}(c, Q') & Q = \mu_C(Q') \\
+    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
+    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
+  \end{cases}
+\end{align*}
+Dropping tuples which contain a null value incurs a loss penalty of 1 for each null field dropped.
+Imputing null fields incurs a loss penalty $0 < p \leq 1$ which decreases as the number of tuples available to train the imputation algorithm increases.
+The intuition behind this formula is that imputation performance should increase if there is more training data available, but there is a saturation point beyond which more data provides little additional benefit.
+
 \subsection{Search space}
+\label{sec:search-space}
 To reduce the size of the query plan search space, only plans that fit the following template are considered.
@@
 
-\subsection{Histogram computation}
-In order to correctly estimate the imputation cost and computation time parameters in our cost model,  the system must have updated cardinality estimates
-at each point in a query plan. These cardinality estimates can be impacted not just be filtering or joining, as in the traditional relational calculus, but also by
-the imputation operators. For example,  dropping versus imputing tuples has a clearly different effect on cardinality. For simplicity of design,
-each of the logical nodes in a query plan keeps a reference to a set of histograms. The optimizer creates a copy-on-write of the histograms and modifies them
-as necessary to account for new operations in the plan. In all cases, modifications maintain as an invariant that the distribution of data in the new histogram
-matches the distribution (but not necessarily the cardinality) of the original statistics. Algorithm~\ref{algo:histogram-transformation} provides the details
-of our implementation.
-
 \begin{algorithm}

### 2016-12-10 presentation/main.tex
presentation: added basic slides

\documentclass{beamer}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[style=ieee]{biblatex}
\usepackage{booktabs}
\usepackage{hyperref} % load after other packages
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}

\addbibresource{references.bib}

\makeatletter
\newcommand*{\centerfloat}{%
  \parindent \z@
  \leftskip \z@ \@plus 1fil \@minus \textwidth
  \rightskip\leftskip
  \parfillskip \z@skip}
\makeatother

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

\MakeRobust{\Call}

\renewcommand{\emptyset}{\varnothing}

\newcommand{\ProjectName}{ImputeDB}

\title{\ProjectName{}: A Database for Missing Data}
\author{Jose Cambronero \and Micah Smith \and John K. Feser}
\institute{MIT}
\date{\today{}}

\begin{document}
\frame{\titlepage}

\begin{frame}{Overview}
\begin{itemize}
	\item Handling incorrect or dirty data is a complex and challenging problem for data scientists
	\item $\text{missing data} \subset \text{dirty data}$
	\item Cost of running imputation on entire dataset
	\item Lack of flexibility
	\item \ProjectName{} allows users to interact with dataset as though it were clean.
	\item User should never see missing data or have to modify query.
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Motivating Example}
\todo{motivating example}
\end{frame}

\begin{frame}[fragile]{Query Planning: Overview}
\begin{itemize}
	\item $\mu_{min}, \mu_{max}, \delta$: new relational algebra operators
	\item Cost of computation vs. information loss: key tradeoff controlled by $\alpha$ parameter
	\item Placement of operators and dirty sets generated considered in planning.
	\item Histogram transformations used to provide up-to-date cardinality and missing value estimates during planning.
	\item Dynamic programming approach, extends System R optimization\cite{blasgen1981system}
	\item Exponential complexity, but tractable in practice with low planning times.
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Query Planning: Top-Level Algorithm}

\begin{algorithm}[H]
 \begin{algorithmic}
   \scriptsize
   \Require{set of tables $T$, filter predicates $\Phi$, join predicates $\Psi$, a set of projection attributes $P$, optional set of grouping attributes $G$ and aggregator function $A$}
   \Ensure{Returns optimized query plan.}

  \Function{Plan}{$T, \Phi, \Psi, P, G, A$}
 \State  $Q \gets \{ \}$ \Comment{map from table and dirty set to plans }
   \For{$t \in T$}
   	\State $\phi \gets \Call{GetFilters}{t, \Phi}$
	\State $P_{t} \gets \Call{OptFilter}{t, \phi}$
	\State $Q \gets Q \cup \{ (t, \Call{Dirty}{q}) \mapsto q | q \in P_t \}$
   \EndFor

   \State $M \gets \{ \}$ \Comment{map from dirty set to plans, contains all joins necessary}
   \State $M \gets  \Call{OptJoin}{Q, \Psi}$	\Comment{optimize joins}
   \If{$G \neq \emptyset \land A \neq NULL$}
   	\State $attrs \gets \Call{Attrs}{G} \cup \Call{Attrs}{A}$
   	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \Call{GroupBy}{\Call{AddImpute}{q, attrs}, G, A}}$
   \Else
  	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \pi_{P}(\Call{AddImpute}{q, P}) }$
  \EndIf

\Return $\argmin_{q \in M} Cost(q)$
\EndFunction

  \end{algorithmic}
  \caption{Top-level query planner with imputations.}
\label{algo:top-level-planner}
\end{algorithm}

\end{frame}

\begin{frame}[fragile]{Imputation Algorithm}
\begin{itemize}
	\item Iterative algorithm using regression trees: \textit{Chained Equations Regression Trees} (CERT)\cite{burgette2010multiple}
	\item Significant cost to perform on entire dataset
	\item \ProjectName{} allows flexible use
\end{itemize}

\begin{algorithm}[H]
  \scriptsize
  \begin{algorithmic}
    \Require{$T$ is a table. $D$ is a set of $T$ attributes that need to be imputed, and $C$ is a set of $T$ attributes that has complete data }
    \Ensure{Returns an imputed $T$}

    \Function{ImputeWithCERT}{$H, op$}
    	\State $T' \gets \Call{ImputeRandom}{T, D}$
	\For{$n\ EPOCHS$}
		\For{$d \in D$}
			\State $imp \gets \Call{TrainRT}{\pi_C(T'), \pi_d{T'}}$
			\State $T' \gets (\pi_C(T'), \Call{PredictRT}{imp, T})$
		\EndFor
	\EndFor
	\Return{T'}
	\EndFunction
  \end{algorithmic}
  \caption{An algorithm for chained imputation using regression trees}
  \label{algo:imputation-strategy}
\end{algorithm}
\end{frame}

\begin{frame}[fragile]{Experiments: Data}
\begin{itemize}
	\item Real dataset: American Community Survey (U.S. Census Bureau)
		\begin{itemize}
			\item Preprocessed version courtesy of authors of~\cite{akande2015empirical}.
			\item 671,153 rows and 37 columns, all integers.
			\item Randomly deleted 10\% of fields to create dirtied variant
			\item Can write typical data analyst queries involving filters and aggregates
		\end{itemize}
	\item Synthetic dataset: drawn from uniform distribution $[0,100)$
		\begin{itemize}
			\item10,000 rows and 10 columns
			\item Randomly deleted 30\% ti create dirtied variant
			\item Crafted ad-hoc queries using joins
		\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Experiments: Queries}
\begin{figure}
  \centerfloat
  \tiny
  \input{../paper/queries}
  \caption{The queries used in our experiments.}
  \label{fig:queries}
\end{figure}
\end{frame}

\begin{frame}[fragile]{Experiments: Results}
\begin{figure}
  \tiny
  \centerfloat
  \input{../paper/runtimes}
  \caption{Running time and error for queries with different imputation levels.}
  \label{fig:experiments}
\end{figure}
\end{frame}

\begin{frame}[fragile]{Related work}
\begin{itemize}
	\item Long history of imputation in statistical learning community.
		\begin{itemize}
			\item Akande et al (\cite{akande2015empirical}) explored ACS data to compare imputation strategies.
			\item Burgette and Reiter (\cite{burgette2010multiple}) introduce sequences of regression trees.
		\end{itemize}
	\item Long history of null value semantics in database community.
		\begin{itemize}
			\item Treatment of nulls described as early as  \cite{codd1973understanding}
			\item Statistical databases, such as BayesDB (\cite{mansinghka2015bayesdb}) combine statistical inference techniques and traditional querying.
		\end{itemize}
	\item \ProjectName{} first to integrate imputation into query planning with cost-based optimizer.
	\item \ProjectName{} treatment of nulls completely abstracted from user perspective: SQL logical independence.
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Conclusion and Future work}
\begin{itemize}
	\item Missing values and their imputation can successfully be integrated into the relational calculus
	\item By taking a dynamic programming approach, we can consider a variety of operator placements
and input columns, while keeping planning tractable in real-world examples.
	\item Considered series of real-world and fabricated queries, showing viability of approach.
	\item Future work:
		\begin{itemize}
			\item Extend imputation operators with global information (\textit{impute by need})
			\item Introduce multiple imputation algorithms into optimizer for consideration in plans \
			\item Implement in a production-quality database.
		\end{itemize}
\end{itemize}
\end{frame}

\end{document}

### 2016-12-10 paper/algorithm.tex

@@
 
-Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric which combines the cost of the query and the quality of the results.
+Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric (Sec.~\ref{sec:cost-model}) which combines the runtime cost of the query and the quality of the results.
 This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
 Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value.
+
+\subsection{Imputation operators}
+We introduce two new relational operators to perform imputation: impute ($\mu$) and drop ($\delta$).
+Each operator takes arguments $(C, R)$ where $C$ is a set of attributes and $R$ is a relation.
+Impute uses a machine learning method to replace all null values with non-null values for attributes in $C$ in the relation $R$.
+Drop simply removes from $R$ all tuples which have a null value for some attribute in $C$.
+Both operators guarantee that the resulting relation will contain no null values for attributes in $C$.
 
@@
 
+\todo{Check loss calculation w/ Micah}
 $\text{Loss}(Q)$ is computed as follows.
@@
 Finally, projections are placed at the root of the query tree.
-The space of query plans is similar to that considered by System R\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
-
-\subsection{Imputation operators}
-We introduce two new relational operators to perform imputation: impute ($\mu$) and drop ($\delta$).
-Each operator takes arguments $(C, R)$ where $C$ is a set of attributes and $R$ is a relation.
-Impute uses a machine learning method to replace all null values with non-null values for attributes in $C$ in the relation $R$.
-Drop simply removes from $R$ all tuples which have a null value for some attribute in $C$.
-Both operators guarantee that the resulting relation will contain no null values for attributes in $C$.
+The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
 
@@
 \subsection{Query planning}
-The input to our query planner is a tuple $(T, \Phi, \Psi, P, G, A)$: a set of tables $T$, a set of filter predicates $\phi_t, t \in T$, a set of join predicates $\psi_(t_1, t_2), t_1, t_2 \in T$, a set of attributes $P$, and an optional set of attributes $G$ and aggregator function $A \in \{\text{Max}, \text{Min}, \text{Sum}, \text{Avg}, \text{Count}\}$.
+The input to our query planner is a tuple $(T, \Phi, \Psi, P, G, A)$: a set of tables $T$, a set of filter predicates $\phi_t, t \in T$, a set of join predicates $\psi_(t_1, t_2), t_1, t_2 \in T$, a set of attributes $P$, and an optional set of attributes $G$ and aggregation function $A \in \{\text{Max}, \text{Min}, \text{Sum}, \text{Avg}, \text{Count}\}$.
 

### 2016-12-10 simpledb/src/java/simpledb/LogicalAccessNode.java
fix: switch to inverse sqrt loss model

@@ -89,7 +89,7 @@ public class LogicalAccessNode extends ImputedPlan {
 			break;
 		case MAXIMAL:
 			pp = new ImputeRegressionTree(DirtySet.toAttrs(tableDirtySet), pp);
-			loss = subplanTableStats.estimateTotalNull() * Math.pow(LOSS_FACTOR, -totalData);
+			loss = subplanTableStats.estimateTotalNull() * (1 / Math.sqrt(totalData));
 			time = subplanTableStats.estimateScanCost() + subplanTableStats.estimateImputeCost((Impute) pp, numDirtyMaximal, numComplete);
 			dirtySet = new HashSet<QuantifiedName>();
 			adjustedTableStats = subplanTableStats.adjustForImpute(MAXIMAL, requiredIdx);
@@ -99,7 +99,7 @@ public class LogicalAccessNode extends ImputedPlan {
 				throw new BadImputation();
 			}
 			pp = new ImputeRegressionTree(requiredAttrs, pp);
-			loss = subplanTableStats.estimateTotalNull(requiredIdx) * Math.pow(LOSS_FACTOR, -totalData);
+			loss = subplanTableStats.estimateTotalNull(requiredIdx) * (1 / Math.sqrt(totalData));
 			time = subplanTableStats.estimateScanCost() + subplanTableStats.estimateImputeCost((Impute) pp, numDirtyMinimal, numComplete);
 			tableDirtySet.removeAll(required);
 			dirtySet = tableDirtySet;

### 2016-12-10 simpledb/src/java/simpledb/LogicalComposeImputation.java

@@ -88,7 +88,7 @@ public class LogicalComposeImputation extends ImputedPlan {
 			Impute imputeOp = new ImputeRegressionTree(toNames(impute), subplan.getPlan());
 			physicalPlan = imputeOp;
 			dirtySet.removeAll(impute);
-			loss = estimateNumNulls(subplan, imputeIndices) * Math.pow(LOSS_FACTOR, -totalData);
+			loss = estimateNumNulls(subplan, imputeIndices) * (1 / Math.sqrt(totalData));
 			int numComplete = schema.numFields() - dirtySet.size();
 			time = imputeOp.getEstimatedCost(imputeIndices.size(), numComplete, (int) subplan.cardinality());
 			adjustedTableStats = subplanTableStats.adjustForImpute(MINIMAL, imputeIndices);
@@ -100,7 +100,7 @@ public class LogicalComposeImputation extends ImputedPlan {
 			Impute imputeOp = new ImputeRegressionTree(toNames(i), subplan.getPlan());
 			physicalPlan = imputeOp;
 			dirtySet.clear();
-			loss = estimateNumNulls(subplan, imputeIndices) * Math.pow(LOSS_FACTOR, -totalData);
+			loss = estimateNumNulls(subplan, imputeIndices) * (1 / Math.sqrt(totalData));
 			int numComplete = schema.numFields() - dirtySet.size();
 			time = imputeOp.getEstimatedCost(imputeIndices.size(), numComplete, (int) subplan.cardinality());
 			adjustedTableStats = subplanTableStats.adjustForImpute(MAXIMAL, imputeIndices);

### 2016-12-10 paper/algorithm.tex
paper: refactor algorithms into separate files

@@
 
-\begin{algorithm}
-  \begin{algorithmic}
-    \Require{$q$ is a query plan, $C_{req}$ is a set of attributes that must be imputed.}
-    \Ensure{Returns a set of query plans such that $\Call{Dirty}{q'} \cap C_{req} = \emptyset$.}
-    \Function{AddImpute}{$q, C_{req}$}
-    \todo{jose: I modified this, since in reality we return only the same plan if it has no dirty values, we may still impute when the intersection is empty, but only maximally}
-    \State $C_{min} \gets \Call{Dirty}{q} \cap C_{req}$
-    \If{$\Call{Dirty}{q} = \emptyset$}
-    	\State \Return $\{ q \}$
-   \ElsIf{$C_{min} = \emptyset$}
-    	\State \Return $\{ \mu_{\Call{Dirty}{q}}(q), q \}$
-    \Else
-    \State \Return $\{\mu_{\Call{Dirty}{q}}(q), \mu_{C_{min}}(q), \delta_{C_{min}}(q)\}$
-    \EndIf
-    \EndFunction
+\input{algorithms/plan-helpers.tex}
 
-    \State
+\input{algorithms/plan-toplevel.tex}
 
-    \Require{$Q$ is a set of query plans.}
-    \Ensure{Returns an optimal query plan for each distinct dirty set in $Q$.}
-    \Function{OptRel}{$Q$}
-    \State $D \gets \{\Call{Dirty}{q} ~|~ q \in Q\}$
-    \State \Return $\{\argmin_{q \in Q \land \Call{Dirty}{q} = d} \Call{Cost}{q} ~|~ d \in D\}$
-    \EndFunction
-
-    \State
-
-    \Require{$t$ is a table and $\phi$ is a filter predicate.}
-    \Ensure{Returns a set of optimal query plans for scanning and filtering $t$, with distinct dirty sets.}
-    \Function{OptFilter}{$t, \phi$}
-    \todo{jose: I believe this was missing the sigma before}
-    \State \Return $\Call{OptRel}{\{  \sigma_{\phi}(q) | q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}} \} }$
-    \EndFunction
-
-    \State
-
-    \Require{$Q$ is a map from tables and dirty sets to optimal query plans (possibly with imputations), and $\Psi$ relates query plans with join predicates.}
-    \Ensure{Returns a set of optimal query plans for all joins, with distinct dirty sets.}
-    \Function{OptJoin}{$Q, \Psi$}
-    \For{$size \in 1...|\Psi|$}
-    	\State $S \gets subset(\Psi, size)$
-	\For{$\psi \in S$}
-		\State $S' \gets S \setminus \psi$
-			\If{$joins(S', \psi)$} 
-				\State $S'_{plans} \gets Q(\Call{Rels}{S'})$
-				\State $t \gets \Call{Rels}{\psi} \setminus \Call{Rels}{S'}$
-				\State $t_{plans} \gets Q(t)$
-				\For {$l, r \in t_{plans} \times S'_{plans}$}
-					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\psi}}$
-					\State $P_r \gets \Call{AddImpute}{r, \Call{Attrs}{\psi}}$
-					\State \Call{Update}{Q, $\Call{OptRel}{ \{p_l \bowtie_{\psi} p_r | p_l \in P_l, p_r \in P_r \} }$}
-				\EndFor
-		\EndIf
-	\EndFor
-    \EndFor
-    \State \Return{Q(\Call{Rels}{$\Phi$})}
-    \EndFunction
-  \end{algorithmic}
-  \caption{Base algorithms for query planning with imputations.}
-  \label{algo:plan-helpers}
-\end{algorithm}
-
-\begin{algorithm}
- \begin{algorithmic}
-   \Require{set of tables $T$, filter predicates $\Phi$, join predicates $\Psi$, a set of projection attributes $P$, optional set of grouping attributes $G$ and aggregator function $A$}
-   \Ensure{Returns optimized query plan.}
-
-  \Function{Plan}{$T, \Phi, \Psi, P, G, A$}
- \State  $Q \gets \{ \}$ \Comment{map from table and dirty set to plans }
-   \For{$t \in T$}
-   	\State $\phi \gets \Call{GetFilters}{t, \Phi}$
-	\State $P_{t} \gets \Call{OptFilter}{t, \phi}$
-	\State $Q \gets Q \cup \{ (t, \Call{Dirty}{q}) \mapsto q | q \in P_t \}$
-   \EndFor
-
-   \State $M \gets \{ \}$ \Comment{map from dirty set to plans, contains all joins necessary}
-   \State $M \gets  \Call{OptJoin}{Q, \Psi}$	\Comment{optimize joins}
-   \If{$G \neq \emptyset \land A \neq NULL$}
-   	\State $attrs \gets \Call{Attrs}{G} \cup \Call{Attrs}{A}$
-   	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \Call{GroupBy}{\Call{AddImpute}{q, attrs}, G, A}}$
-   \Else
-  	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \pi_{P}(\Call{AddImpute}{q, P}) }$
-  \EndIf
-
-\Return $\argmin_{q \in M} Cost(q)$
-\EndFunction
-
-  \end{algorithmic}
-  \caption{Top-level query planner with imputations.}
-\label{algo:top-level-planner}
-\end{algorithm}
-
-%\begin{verbatim}		
-%# For a set of tables (with filter predicates) and join predicates, return the optimal plans.
-%OptJoin(T⁆, join_preds):
-%	Q = empty
-%	for t, pred in T:
-%		S = T \ {t, pred}
-%		# Get the optimal plans for the left and right arguments to the join.
-%		for r_l, r_r in OptJoin(S) x OptFilter(t, pred):
-%			if r_l, r_r in join_preds:
-%				# Add imputation to the left and right arguments.
-%				Q = Q union OptRel(Join(r_l', r_r') for r_l' in AddImpute(r_l, Attrs(join_pred)), r_r' in AddImpute(r_r, Attrs(join_pred)))
-%	return OptRel(Q)
-%# Extended cost calculation to account for information loss/computation cost in imputations
-%Cost(q):
-%case drop_min(q', d) =>  info_loss(drop, q', attrs(q') intersect d) + Cost(q')
-%case impute_all(q') => info_loss(impute, q', attr(q')) + impute_cost(q') + Cost(q')
-%case impute_min(q', d) => info_loss(impute, q', attrs(q') intersect d) + impute_cost(q'[d]) + Cost(q')
-%case _ => normal cost calculation
-%
-%
-%info_loss(op, t, a):
-%# beta: some scaling factor
-%case info_loss(drop, t, a) => beta * sum(for ct_missing(t.a) in a)
-%case info_loss(impute, t, a) => beta *sum(for ct_missing(t.a) / ct_complete(t) in a)
-%
-%impute_cost(t):
-%# alpha: some scaling factor
-%alpha * (ntuples(t) + ct(attr(t)))
-%\end{verbatim}
-
-\begin{algorithm}
-  \begin{algorithmic}
-    \Require{$H$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
-    \Ensure{Returns an updated $H$, such that the distribution of data matches the original $H$}
-
-    \Function{UpdateHistogram}{$H, op$}
-    	\If{$op \not\in \{\delta, \mu, \sigma, \bowtie \}$}
-		\Return{H}
-	\EndIf
-	\State $H' \gets \Call{Copy}{H}$
-    	\If{op = $\delta_C$}
-		\For{$c \in C$}
-			\State $H'[c][null] \gets 0$
-		\EndFor
-	\ElsIf{op = $\mu_C$}
-		\For{$c \in C$}
-			\State \Call{Add}{$H'[c], H'[c][null]$}
-		\EndFor
-	\ElsIf{op = $\sigma$}
-		\State \Call{ScaleBy}{H',  \Call{Selectivity}{$\sigma$}}
-	\ElsIf{op = $\bowtie_\phi$}
-		\State \Call{ScaleTo}{H', \Call{Cardinality}{$\bowtie_{\phi}$}}
-	\EndIf
-	\Return{H'}
-    \EndFunction
-  \end{algorithmic}
-  \caption{An algorithm for in-plan histogram updates}
-  \label{algo:histogram-transformation}
-\end{algorithm}
+\input{algorithms/histogram-transform.tex}
 
@@
 
-\begin{algorithm}
-  \begin{algorithmic}
-    \Require{$T$ is a table. $D$ is a set of $T$ attributes that need to be imputed, and $C$ is a set of $T$ attributes that has complete data }
-    \Ensure{Returns an imputed $T$}
-
-    \Function{ImputeWithCERT}{$H, op$}
-    	\State $T' \gets \Call{ImputeRandom}{T, D}$
-	\For{$n\ EPOCHS$}
-		\For{$d \in D$}
-			\State $imp \gets \Call{TrainRT}{\pi_C(T'), \pi_d{T'}}$
-			\State $T' \gets (\pi_C(T'), \Call{PredictRT}{imp, T})$
-		\EndFor
-	\EndFor
-	\Return{T'}
-	\EndFunction
-  \end{algorithmic}
-  \caption{An algorithm for chained imputation using regression trees}
-  \label{algo:imputation-strategy}
-\end{algorithm}
+\input{algorithms/cert.tex}
 

### 2016-12-10 paper/algorithms/cert.tex
paper: refactor algorithms into separate files

--- 2016-12-10 presentation/main.tex
@@
-\documentclass{beamer}
-
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage[english]{babel}
-\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
-\usepackage{amsfonts,amssymb,amsmath}
-\usepackage{xcolor}
-\usepackage{algorithm}
-\usepackage[noend]{algpseudocode}
-\usepackage[style=ieee]{biblatex}
-\usepackage{booktabs}
-\usepackage{hyperref} % load after other packages
-\usepackage{todonotes}
-\presetkeys{todonotes}{inline}{}
-
-\addbibresource{references.bib}
-
-\makeatletter
-\newcommand*{\centerfloat}{%
-  \parindent \z@
-  \leftskip \z@ \@plus 1fil \@minus \textwidth
-  \rightskip\leftskip
-  \parfillskip \z@skip}
-\makeatother
-
-\DeclareMathOperator*{\argmax}{arg\,max}
-\DeclareMathOperator*{\argmin}{arg\,min}
-
-\MakeRobust{\Call}
-
-\renewcommand{\emptyset}{\varnothing}
-
-\newcommand{\ProjectName}{ImputeDB}
-
-\title{\ProjectName{}: A Database for Missing Data}
-\author{Jose Cambronero \and Micah Smith \and John K. Feser}
-\institute{MIT}
-\date{\today{}}
-
-\begin{document}
-\frame{\titlepage}
-
-\begin{frame}{Overview}
-\begin{itemize}
-	\item Handling incorrect or dirty data is a complex and challenging problem for data scientists
-	\item $\text{missing data} \subset \text{dirty data}$
-	\item Cost of running imputation on entire dataset
-	\item Lack of flexibility
-	\item \ProjectName{} allows users to interact with dataset as though it were clean.
-	\item User should never see missing data or have to modify query.
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Motivating Example}
-\todo{motivating example}
-\end{frame}
-
-\begin{frame}[fragile]{Query Planning: Overview}
-\begin{itemize}
-	\item $\mu_{min}, \mu_{max}, \delta$: new relational algebra operators
-	\item Cost of computation vs. information loss: key tradeoff controlled by $\alpha$ parameter
-	\item Placement of operators and dirty sets generated considered in planning.
-	\item Histogram transformations used to provide up-to-date cardinality and missing value estimates during planning.
-	\item Dynamic programming approach, extends System R optimization\cite{blasgen1981system}
-	\item Exponential complexity, but tractable in practice with low planning times.
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Query Planning: Top-Level Algorithm}
-
-\begin{algorithm}[H]
- \begin{algorithmic}
-   \scriptsize
-   \Require{set of tables $T$, filter predicates $\Phi$, join predicates $\Psi$, a set of projection attributes $P$, optional set of grouping attributes $G$ and aggregator function $A$}
-   \Ensure{Returns optimized query plan.}
-
-  \Function{Plan}{$T, \Phi, \Psi, P, G, A$}
- \State  $Q \gets \{ \}$ \Comment{map from table and dirty set to plans }
-   \For{$t \in T$}
-   	\State $\phi \gets \Call{GetFilters}{t, \Phi}$
-	\State $P_{t} \gets \Call{OptFilter}{t, \phi}$
-	\State $Q \gets Q \cup \{ (t, \Call{Dirty}{q}) \mapsto q | q \in P_t \}$
-   \EndFor
-
-   \State $M \gets \{ \}$ \Comment{map from dirty set to plans, contains all joins necessary}
-   \State $M \gets  \Call{OptJoin}{Q, \Psi}$	\Comment{optimize joins}
-   \If{$G \neq \emptyset \land A \neq NULL$}
-   	\State $attrs \gets \Call{Attrs}{G} \cup \Call{Attrs}{A}$
-   	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \Call{GroupBy}{\Call{AddImpute}{q, attrs}, G, A}}$
-   \Else
-  	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \pi_{P}(\Call{AddImpute}{q, P}) }$
-  \EndIf
-
-\Return $\argmin_{q \in M} Cost(q)$
-\EndFunction
-
-  \end{algorithmic}
-  \caption{Top-level query planner with imputations.}
-\label{algo:top-level-planner}
-\end{algorithm}
-
-\end{frame}
-
-\begin{frame}[fragile]{Imputation Algorithm}
-\begin{itemize}
-	\item Iterative algorithm using regression trees: \textit{Chained Equations Regression Trees} (CERT)\cite{burgette2010multiple}
-	\item Significant cost to perform on entire dataset
-	\item \ProjectName{} allows flexible use
-\end{itemize}
-
-\begin{algorithm}[H]
-  \scriptsize
+\begin{algorithm}
   \begin{algorithmic}
@@
 \end{algorithm}
-\end{frame}
-
-\begin{frame}[fragile]{Experiments: Data}
-\begin{itemize}
-	\item Real dataset: American Community Survey (U.S. Census Bureau)
-		\begin{itemize}
-			\item Preprocessed version courtesy of authors of~\cite{akande2015empirical}.
-			\item 671,153 rows and 37 columns, all integers.
-			\item Randomly deleted 10\% of fields to create dirtied variant
-			\item Can write typical data analyst queries involving filters and aggregates
-		\end{itemize}
-	\item Synthetic dataset: drawn from uniform distribution $[0,100)$
-		\begin{itemize}
-			\item10,000 rows and 10 columns
-			\item Randomly deleted 30\% ti create dirtied variant
-			\item Crafted ad-hoc queries using joins
-		\end{itemize}
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Experiments: Queries}
-\begin{figure}
-  \centerfloat
-  \tiny
-  \input{../paper/queries}
-  \caption{The queries used in our experiments.}
-  \label{fig:queries}
-\end{figure}
-\end{frame}
-
-\begin{frame}[fragile]{Experiments: Results}
-\begin{figure}
-  \tiny
-  \centerfloat
-  \input{../paper/runtimes}
-  \caption{Running time and error for queries with different imputation levels.}
-  \label{fig:experiments}
-\end{figure}
-\end{frame}
-
-\begin{frame}[fragile]{Related work}
-\begin{itemize}
-	\item Long history of imputation in statistical learning community.
-		\begin{itemize}
-			\item Akande et al (\cite{akande2015empirical}) explored ACS data to compare imputation strategies.
-			\item Burgette and Reiter (\cite{burgette2010multiple}) introduce sequences of regression trees.
-		\end{itemize}
-	\item Long history of null value semantics in database community.
-		\begin{itemize}
-			\item Treatment of nulls described as early as  \cite{codd1973understanding}
-			\item Statistical databases, such as BayesDB (\cite{mansinghka2015bayesdb}) combine statistical inference techniques and traditional querying.
-		\end{itemize}
-	\item \ProjectName{} first to integrate imputation into query planning with cost-based optimizer.
-	\item \ProjectName{} treatment of nulls completely abstracted from user perspective: SQL logical independence.
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Conclusion and Future work}
-\begin{itemize}
-	\item Missing values and their imputation can successfully be integrated into the relational calculus
-	\item By taking a dynamic programming approach, we can consider a variety of operator placements
-and input columns, while keeping planning tractable in real-world examples.
-	\item Considered series of real-world and fabricated queries, showing viability of approach.
-	\item Future work:
-		\begin{itemize}
-			\item Extend imputation operators with global information (\textit{impute by need})
-			\item Introduce multiple imputation algorithms into optimizer for consideration in plans \
-			\item Implement in a production-quality database.
-		\end{itemize}
-\end{itemize}
-\end{frame}
-
-\end{document}

### 2016-12-10 paper/algorithms/histogram-transform.tex
paper: refactor algorithms into separate files

\begin{algorithm}
  \begin{algorithmic}
    \Require{$H$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
    \Ensure{Returns an updated $H$, such that the distribution of data matches the original $H$}

    \Function{UpdateHistogram}{$H, op$}
    	\If{$op \not\in \{\delta, \mu, \sigma, \bowtie \}$}
		\Return{H}
	\EndIf
	\State $H' \gets \Call{Copy}{H}$
    	\If{op = $\delta_C$}
		\For{$c \in C$}
			\State $H'[c][null] \gets 0$
		\EndFor
	\ElsIf{op = $\mu_C$}
		\For{$c \in C$}
			\State \Call{Add}{$H'[c], H'[c][null]$}
		\EndFor
	\ElsIf{op = $\sigma$}
		\State \Call{ScaleBy}{H',  \Call{Selectivity}{$\sigma$}}
	\ElsIf{op = $\bowtie_\phi$}
		\State \Call{ScaleTo}{H', \Call{Cardinality}{$\bowtie_{\phi}$}}
	\EndIf
	\Return{H'}
    \EndFunction
  \end{algorithmic}
  \caption{An algorithm for in-plan histogram updates}
  \label{algo:histogram-transformation}
\end{algorithm}

### 2016-12-10 paper/algorithms/plan-helpers.tex
paper: refactor algorithms into separate files

\begin{algorithm}
  \begin{algorithmic}
    \Require{$q$ is a query plan, $C_{req}$ is a set of attributes that must be imputed.}
    \Ensure{Returns a set of query plans such that $\Call{Dirty}{q'} \cap C_{req} = \emptyset$.}
    \Function{AddImpute}{$q, C_{req}$}
    \todo{jose: I modified this, since in reality we return only the same plan if it has no dirty values, we may still impute when the intersection is empty, but only maximally}
    \State $C_{min} \gets \Call{Dirty}{q} \cap C_{req}$
    \If{$\Call{Dirty}{q} = \emptyset$}
    	\State \Return $\{ q \}$
   \ElsIf{$C_{min} = \emptyset$}
    	\State \Return $\{ \mu_{\Call{Dirty}{q}}(q), q \}$
    \Else
    \State \Return $\{\mu_{\Call{Dirty}{q}}(q), \mu_{C_{min}}(q), \delta_{C_{min}}(q)\}$
    \EndIf
    \EndFunction

    \State

    \Require{$Q$ is a set of query plans.}
    \Ensure{Returns an optimal query plan for each distinct dirty set in $Q$.}
    \Function{OptRel}{$Q$}
    \State $D \gets \{\Call{Dirty}{q} ~|~ q \in Q\}$
    \State \Return $\{\argmin_{q \in Q \land \Call{Dirty}{q} = d} \Call{Cost}{q} ~|~ d \in D\}$
    \EndFunction

    \State

    \Require{$t$ is a table and $\phi$ is a filter predicate.}
    \Ensure{Returns a set of optimal query plans for scanning and filtering $t$, with distinct dirty sets.}
    \Function{OptFilter}{$t, \phi$}
    \todo{jose: I believe this was missing the sigma before}
    \State \Return $\Call{OptRel}{\{  \sigma_{\phi}(q) | q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}} \} }$
    \EndFunction

    \State

    \Require{$Q$ is a map from tables and dirty sets to optimal query plans (possibly with imputations), and $\Psi$ relates query plans with join predicates.}
    \Ensure{Returns a set of optimal query plans for all joins, with distinct dirty sets.}
    \Function{OptJoin}{$Q, \Psi$}
    \For{$size \in 1...|\Psi|$}
    	\State $S \gets subset(\Psi, size)$
	\For{$\psi \in S$}
		\State $S' \gets S \setminus \psi$
			\If{$joins(S', \psi)$} 
				\State $S'_{plans} \gets Q(\Call{Rels}{S'})$
				\State $t \gets \Call{Rels}{\psi} \setminus \Call{Rels}{S'}$
				\State $t_{plans} \gets Q(t)$
				\For {$l, r \in t_{plans} \times S'_{plans}$}
					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\psi}}$
					\State $P_r \gets \Call{AddImpute}{r, \Call{Attrs}{\psi}}$
					\State \Call{Update}{Q, $\Call{OptRel}{ \{p_l \bowtie_{\psi} p_r | p_l \in P_l, p_r \in P_r \} }$}
				\EndFor
		\EndIf
	\EndFor
    \EndFor
    \State \Return{Q(\Call{Rels}{$\Phi$})}
    \EndFunction
  \end{algorithmic}
  \caption{Base algorithms for query planning with imputations.}
  \label{algo:plan-helpers}
\end{algorithm}

### 2016-12-10 paper/algorithms/plan-toplevel.tex
paper: refactor algorithms into separate files

--- 2016-12-10 presentation/main.tex
@@
-\documentclass{beamer}
-
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage[english]{babel}
-\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
-\usepackage{amsfonts,amssymb,amsmath}
-\usepackage{xcolor}
-\usepackage{algorithm}
-\usepackage[noend]{algpseudocode}
-\usepackage[style=ieee]{biblatex}
-\usepackage{booktabs}
-\usepackage{hyperref} % load after other packages
-\usepackage{todonotes}
-\presetkeys{todonotes}{inline}{}
-
-\addbibresource{references.bib}
-
-\makeatletter
-\newcommand*{\centerfloat}{%
-  \parindent \z@
-  \leftskip \z@ \@plus 1fil \@minus \textwidth
-  \rightskip\leftskip
-  \parfillskip \z@skip}
-\makeatother
-
-\DeclareMathOperator*{\argmax}{arg\,max}
-\DeclareMathOperator*{\argmin}{arg\,min}
-
-\MakeRobust{\Call}
-
-\renewcommand{\emptyset}{\varnothing}
-
-\newcommand{\ProjectName}{ImputeDB}
-
-\title{\ProjectName{}: A Database for Missing Data}
-\author{Jose Cambronero \and Micah Smith \and John K. Feser}
-\institute{MIT}
-\date{\today{}}
-
-\begin{document}
-\frame{\titlepage}
-
-\begin{frame}{Overview}
-\begin{itemize}
-	\item Handling incorrect or dirty data is a complex and challenging problem for data scientists
-	\item $\text{missing data} \subset \text{dirty data}$
-	\item Cost of running imputation on entire dataset
-	\item Lack of flexibility
-	\item \ProjectName{} allows users to interact with dataset as though it were clean.
-	\item User should never see missing data or have to modify query.
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Motivating Example}
-\todo{motivating example}
-\end{frame}
-
-\begin{frame}[fragile]{Query Planning: Overview}
-\begin{itemize}
-	\item $\mu_{min}, \mu_{max}, \delta$: new relational algebra operators
-	\item Cost of computation vs. information loss: key tradeoff controlled by $\alpha$ parameter
-	\item Placement of operators and dirty sets generated considered in planning.
-	\item Histogram transformations used to provide up-to-date cardinality and missing value estimates during planning.
-	\item Dynamic programming approach, extends System R optimization\cite{blasgen1981system}
-	\item Exponential complexity, but tractable in practice with low planning times.
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Query Planning: Top-Level Algorithm}
-
-\begin{algorithm}[H]
+\begin{algorithm}
  \begin{algorithmic}
-   \scriptsize
    \Require{set of tables $T$, filter predicates $\Phi$, join predicates $\Psi$, a set of projection attributes $P$, optional set of grouping attributes $G$ and aggregator function $A$}
@@
 \end{algorithm}
-
-\end{frame}
-
-\begin{frame}[fragile]{Imputation Algorithm}
-\begin{itemize}
-	\item Iterative algorithm using regression trees: \textit{Chained Equations Regression Trees} (CERT)\cite{burgette2010multiple}
-	\item Significant cost to perform on entire dataset
-	\item \ProjectName{} allows flexible use
-\end{itemize}
-
-\begin{algorithm}[H]
-  \scriptsize
-  \begin{algorithmic}
-    \Require{$T$ is a table. $D$ is a set of $T$ attributes that need to be imputed, and $C$ is a set of $T$ attributes that has complete data }
-    \Ensure{Returns an imputed $T$}
-
-    \Function{ImputeWithCERT}{$H, op$}
-    	\State $T' \gets \Call{ImputeRandom}{T, D}$
-	\For{$n\ EPOCHS$}
-		\For{$d \in D$}
-			\State $imp \gets \Call{TrainRT}{\pi_C(T'), \pi_d{T'}}$
-			\State $T' \gets (\pi_C(T'), \Call{PredictRT}{imp, T})$
-		\EndFor
-	\EndFor
-	\Return{T'}
-	\EndFunction
-  \end{algorithmic}
-  \caption{An algorithm for chained imputation using regression trees}
-  \label{algo:imputation-strategy}
-\end{algorithm}
-\end{frame}
-
-\begin{frame}[fragile]{Experiments: Data}
-\begin{itemize}
-	\item Real dataset: American Community Survey (U.S. Census Bureau)
-		\begin{itemize}
-			\item Preprocessed version courtesy of authors of~\cite{akande2015empirical}.
-			\item 671,153 rows and 37 columns, all integers.
-			\item Randomly deleted 10\% of fields to create dirtied variant
-			\item Can write typical data analyst queries involving filters and aggregates
-		\end{itemize}
-	\item Synthetic dataset: drawn from uniform distribution $[0,100)$
-		\begin{itemize}
-			\item10,000 rows and 10 columns
-			\item Randomly deleted 30\% ti create dirtied variant
-			\item Crafted ad-hoc queries using joins
-		\end{itemize}
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Experiments: Queries}
-\begin{figure}
-  \centerfloat
-  \tiny
-  \input{../paper/queries}
-  \caption{The queries used in our experiments.}
-  \label{fig:queries}
-\end{figure}
-\end{frame}
-
-\begin{frame}[fragile]{Experiments: Results}
-\begin{figure}
-  \tiny
-  \centerfloat
-  \input{../paper/runtimes}
-  \caption{Running time and error for queries with different imputation levels.}
-  \label{fig:experiments}
-\end{figure}
-\end{frame}
-
-\begin{frame}[fragile]{Related work}
-\begin{itemize}
-	\item Long history of imputation in statistical learning community.
-		\begin{itemize}
-			\item Akande et al (\cite{akande2015empirical}) explored ACS data to compare imputation strategies.
-			\item Burgette and Reiter (\cite{burgette2010multiple}) introduce sequences of regression trees.
-		\end{itemize}
-	\item Long history of null value semantics in database community.
-		\begin{itemize}
-			\item Treatment of nulls described as early as  \cite{codd1973understanding}
-			\item Statistical databases, such as BayesDB (\cite{mansinghka2015bayesdb}) combine statistical inference techniques and traditional querying.
-		\end{itemize}
-	\item \ProjectName{} first to integrate imputation into query planning with cost-based optimizer.
-	\item \ProjectName{} treatment of nulls completely abstracted from user perspective: SQL logical independence.
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Conclusion and Future work}
-\begin{itemize}
-	\item Missing values and their imputation can successfully be integrated into the relational calculus
-	\item By taking a dynamic programming approach, we can consider a variety of operator placements
-and input columns, while keeping planning tractable in real-world examples.
-	\item Considered series of real-world and fabricated queries, showing viability of approach.
-	\item Future work:
-		\begin{itemize}
-			\item Extend imputation operators with global information (\textit{impute by need})
-			\item Introduce multiple imputation algorithms into optimizer for consideration in plans \
-			\item Implement in a production-quality database.
-		\end{itemize}
-\end{itemize}
-\end{frame}
-
-\end{document}

### 2016-12-11 paper/algorithm.tex
paper fix: extract algorithm begin/end markers to be able to use with [H] placement in slides
paper: update imputation algorithm section

--- 2016-12-10 paper/algorithm.tex
@@
 
-Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric which combines the cost of the query and the quality of the results.
-This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
-Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value.
-
-\subsection{Cost model}
-\label{sec:cost-model}
-We rank query plans $Q$ using a cost model $\text{Cost}(Q) = \alpha \times \text{Time}(Q) + (1 - \alpha) \times \text{Loss}(Q)$.
-$\text{Time}(Q)$ is an estimate of the runtime of the query which is derived from table statistics and selectivity estimation of the query predicates.
-$\text{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure.
+At a high level, our goal is to select a query plan that minimizes a metric which combines the cost of the query and the quality of the results.
+Precisely, we will find a query plan $Q$ that minimizes $\text{Cost}(Q) = \alpha \times \text{Time}(Q) + (1 - \alpha) \times \text{Loss}(Q)$.
 $\alpha$ is a parameter to the query optimizer that controls the emphasis on quality versus performance.
@@
 
-In order to correctly estimate $\text{Time}(Q)$ and $\text{Loss}(Q)$, the system must have cardinality estimates
-for each sub-query in each query plan.
-These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
-For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
-For simplicity, each of the logical nodes in a query plan refers to a set of histograms.
-When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
-Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
-
-The computation of $\text{Time}(Q)$ is standard and is comparable to the cost model used in any modern RDBMS.\@ Our runtime cost model is perhaps less sophisticated than most, as it is identical to the cost model in SimpleDB.
-
-$\text{Loss}(Q)$ is computed as follows.
-For each sub-query, in addition to a histogram, we keep track of the estimated number of null values in each column.
-\begin{align*}
-  \text{Loss}(Q) = \begin{cases}
-    \sum_{c \in C} \text{Null}(c, Q') & Q = \delta_C(Q') \\
-    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{Null}(c, Q') & Q = \mu_C(Q') \\
-    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
-    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
-  \end{cases}
-\end{align*}
-Dropping tuples which contain a null value incurs a loss penalty of 1 for each null field dropped.
-Imputing null fields incurs a loss penalty $0 < p \leq 1$ which decreases as the number of tuples available to train the imputation algorithm increases.
-The intuition behind this formula is that imputation performance should increase if there is more training data available, but there is a saturation point beyond which more data provides little additional benefit.
-
 \subsection{Search space}
-\label{sec:search-space}
 To reduce the size of the query plan search space, only plans that fit the following template are considered.
@@
 Finally, projections are placed at the root of the query tree.
-The space of query plans is similar to that considered by System R\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
+The space of query plans is similar to that considered by Selinger \todo{Selinger cite}, with the addition of imputation operators appearing before/after traditional operators.
 
@@
 
+\subsection{Histogram computation}
+In order to correctly estimate the imputation cost and computation time parameters in our cost model,  the system must have updated cardinality estimates
+at each point in a query plan. These cardinality estimates can be impacted not just be filtering or joining, as in the traditional relational calculus, but also by
+the imputation operators. For example,  dropping versus imputing tuples has a clearly different effect on cardinality. For simplicity of design,
+each of the logical nodes in a query plan keeps a reference to a set of histograms. The optimizer creates a copy-on-write of the histograms and modifies them
+as necessary to account for new operations in the plan. In all cases, modifications maintain as an invariant that the distribution of data in the new histogram
+matches the distribution (but not necessarily the cardinality) of the original statistics. Algorithm~\ref{algo:histogram-transformation} provides the details
+of our implementation.
+
 \begin{algorithm}
@@
 \subsection{Imputation Strategies}
-The current version of the system uses an imputation strategy based on chained-equations regression trees (CERT)\cite{burgette2010multiple}. Note that plugging in a new
-imputation strategy takes minimal effort, and doesn't require changes to the optimizer, so the system could in principle be tuned to a specific domain.
-The imputation algorithm is iterative in nature, and proceeds by repeatedly fitting regression trees to a subset of the data and the target imputation columns.
-In each iteration, the missing values of a column are replaced with newly imputed values. As the iterations proceed, the quality of the imputation
-should improve as values progressively reflect more accurate relationships amongst attributes. Algorithm~\ref{algo:imputation-strategy}
+
+We design our system such that any imputation strategy can be plugged in with minimal effort
+and without changes to the optimizer, so the strategy can in principle be targeted to a
+specific domain. The current version of the system uses a general-purpose imputation
+strategy based on chained-equation classification and regression trees (CE-CART). Chained
+equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
+regression} \cite{gelman2006data} and others) impute multiple missing attributes by
+iteratively estimating predictive models of one missing attribute conditional on all
+complete attributes and the other missing attributes. The individual predictive models can
+be customized by the researcher to the problem at hand. Decision tree algorithms, like CART,
+are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
+routines and are widely used in epidemiological domains \cite{burgette2011multiple}.
+
+These imputation algorithms are rarely used in isolation --- rather, the ultimate goal is to
+perform statistical analysis of the complete data. To this end, the technique of
+\textit{multiple imputation} can be used, in which multiple distinct copies of the complete
+data are generated and estimators are computed by averaging over the multiple datasets. In
+ImputeDB, we cannot assume the user desires multiple copies of the query result, though we
+note that in further work, multiple imputation could be used within ImputeDB to generate
+lower-variance estimates of some aggregates, like averages.
+
+Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
+the target imputation columns. In each iteration, the missing values of a column are
+replaced with newly imputed values. With each epoch, the quality of the imputation improves
+as values progressively reflect more accurate relationships amongst attributes. The
+algorithm terminates when convergence is achieved (the imputed values do not change across
+epochs) or a fixed number of epochs is reached.  Algorithm~\ref{algo:imputation-strategy}
 provides details on the implementation.
 
-It is worth noting that given the nature of the imputation strategy used here, imputation operators become blocking in our system.
+It is worth noting that given the nature of the imputation strategy used here, imputation
+operators become blocking in our system. An extension of ImputeDB could consider regression
+algorithms that learn in an online nature; this would reduce the cost of the imputation at
+the possible expense of quality.
 
@@
   \begin{algorithmic}
-    \Require{$T$ is a table. $D$ is a set of $T$ attributes that need to be imputed, and $C$ is a set of $T$ attributes that has complete data }
+    \Require{$T$ is a table. $D$ is a set of attributes of $T$ that need to be imputed, and
+      $C$ is a set of attributes of $T$ that has complete data }
     \Ensure{Returns an imputed $T$}
 
-    \Function{ImputeWithCERT}{$H, op$}
+    \Function{ImputeWithCE-CART}{$H, op$}
     	\State $T' \gets \Call{ImputeRandom}{T, D}$
@@
 		\For{$d \in D$}
-			\State $imp \gets \Call{TrainRT}{\pi_C(T'), \pi_d{T'}}$
-			\State $T' \gets (\pi_C(T'), \Call{PredictRT}{imp, T})$
+            \State $imp \gets \Call{TrainRT}{\pi_{C \cup D \backslash \{d\}}(T'), \pi_d{T'}}$
+            \State $T' \gets (\pi_{C \cup D \backslash \{d\}}(T'), \Call{PredictRT}{imp, T})$
 		\EndFor
@@
 
-\todo{Jose:Micah, take a look at the explanation above and see if it's missing anything (or is incorrect)}
+\todo{Micah: Explanation edited a bit. In the algorithm, when we impute attribute $d$, we
+condition on attributes $C \cup D \backslash \{d\}$. Hopefully the way I note it is clear;
+feel free to modify otherwise.}
 

### 2016-12-11 paper/conclusion.tex
paper: update imputation algorithm section

@@
 \subsection{Future Work}
-\ProjectName{} opens up multiple avenues for further work. For instance, we could extend our minimal/maximal imputation
-operators to consider global information, such as the specific columns needed in operators higher up in the query plan.
-Our optimizer uses the same machine learning algorithm in all instances of imputation operators. Integrating multiple
-possible algorithms could allow for further fine-grained imputation. Finally, the underlying database for \ProjectName{}
-is far from a full-featured, production database, so a natural step is to integrate imputation planning in a widely-adopted
-production-quality database. This will likely expose further opportunities for improvement.
+\ProjectName{} opens up multiple avenues for further work. For instance, we
+could extend our minimal/maximal imputation operators to consider global
+information, such as the specific columns needed in operators higher up in the
+query plan.  Our optimizer uses the same machine learning algorithm in all
+instances of imputation operators. Integrating multiple possible algorithms
+could allow for further fine-grained imputation. For example, algorithms that
+learn in an online manner could increase the efficiency of the system.
+Furthermore, a multiple imputation strategy could be followed for queries
+involving certain aggregates. Finally, the underlying database for
+\ProjectName{} is far from a full-featured, production database, so a natural
+step is to integrate imputation planning in a widely-adopted production-quality
+database. This will likely expose further opportunities for improvement.
 

### 2016-12-11 paper/algorithm.tex

--- 2016-12-10 paper/algorithm.tex
@@
 
+\begin{algorithm}
 \input{algorithms/plan-helpers.tex}
+\end{algorithm}
 
+\begin{algorithm}
 \input{algorithms/plan-toplevel.tex}
+\end{algorithm}
 
+\begin{algorithm}
 \input{algorithms/histogram-transform.tex}
+\end{algorithm}
 
 \subsection{Imputation Strategies}
-The current version of the system uses an imputation strategy based on chained-equations regression trees (CERT)\cite{burgette2010multiple}. Note that plugging in a new
-imputation strategy takes minimal effort, and doesn't require changes to the optimizer, so the system could in principle be tuned to a specific domain.
-The imputation algorithm is iterative in nature, and proceeds by repeatedly fitting regression trees to a subset of the data and the target imputation columns.
-In each iteration, the missing values of a column are replaced with newly imputed values. As the iterations proceed, the quality of the imputation
-should improve as values progressively reflect more accurate relationships amongst attributes. Algorithm~\ref{algo:imputation-strategy}
+
+We design our system such that any imputation strategy can be plugged in with minimal effort
+and without changes to the optimizer, so the strategy can in principle be targeted to a
+specific domain. The current version of the system uses a general-purpose imputation
+strategy based on chained-equation classification and regression trees (CE-CART). Chained
+equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
+regression} \cite{gelman2006data} and others) impute multiple missing attributes by
+iteratively estimating predictive models of one missing attribute conditional on all
+complete attributes and the other missing attributes. The individual predictive models can
+be customized by the researcher to the problem at hand. Decision tree algorithms, like CART,
+are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
+routines and are widely used in epidemiological domains \cite{burgette2011multiple}.
+
+These imputation algorithms are rarely used in isolation --- rather, the ultimate goal is to
+perform statistical analysis of the complete data. To this end, the technique of
+\textit{multiple imputation} can be used, in which multiple distinct copies of the complete
+data are generated and estimators are computed by averaging over the multiple datasets. In
+ImputeDB, we cannot assume the user desires multiple copies of the query result, though we
+note that in further work, multiple imputation could be used within ImputeDB to generate
+lower-variance estimates of some aggregates, like averages.
+
+Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
+the target imputation columns. In each iteration, the missing values of a column are
+replaced with newly imputed values. With each epoch, the quality of the imputation improves
+as values progressively reflect more accurate relationships amongst attributes. The
+algorithm terminates when convergence is achieved (the imputed values do not change across
+epochs) or a fixed number of epochs is reached.  Algorithm~\ref{algo:imputation-strategy}
 provides details on the implementation.
 
-It is worth noting that given the nature of the imputation strategy used here, imputation operators become blocking in our system.
+It is worth noting that given the nature of the imputation strategy used here, imputation
+operators become blocking in our system. An extension of ImputeDB could consider regression
+algorithms that learn in an online nature; this would reduce the cost of the imputation at
+the possible expense of quality.
 
-\input{algorithms/cert.tex}
+\begin{algorithm}
+  \begin{algorithmic}
+    \Require{$T$ is a table. $D$ is a set of attributes of $T$ that need to be imputed, and
+      $C$ is a set of attributes of $T$ that has complete data }
+    \Ensure{Returns an imputed $T$}
 
-\todo{Jose:Micah, take a look at the explanation above and see if it's missing anything (or is incorrect)}
+    \Function{ImputeWithCE-CART}{$H, op$}
+    	\State $T' \gets \Call{ImputeRandom}{T, D}$
+	\For{$n\ EPOCHS$}
+		\For{$d \in D$}
+            \State $imp \gets \Call{TrainRT}{\pi_{C \cup D \backslash \{d\}}(T'), \pi_d{T'}}$
+            \State $T' \gets (\pi_{C \cup D \backslash \{d\}}(T'), \Call{PredictRT}{imp, T})$
+		\EndFor
+	\EndFor
+	\Return{T'}
+	\EndFunction
+  \end{algorithmic}
+  \caption{An algorithm for chained imputation using regression trees}
+  \label{algo:imputation-strategy}
+%\input{algorithms/cert.tex}
+\end{algorithm}
+
+\todo{Micah: Explanation edited a bit. In the algorithm, when we impute attribute $d$, we
+condition on attributes $C \cup D \backslash \{d\}$. Hopefully the way I note it is clear;
+feel free to modify otherwise.}
 

### 2016-12-11 paper/experiments.tex

@@
 
-To evaluate the performance of our system, we generated plans for queries on two data sets: the American Community Survey and a synthetic data set containing business information.\todo{Synthetic queries?}
-
+To evaluate the performance of our system, we generated plans for queries on two data sets: the American Community Survey and a synthetic dataset.
 \subsection{Data sets}
 
-The American Community Survey provides a number of public data sets collected by the U.S. Census Bureau.
+The American Community Survey (ACS) provides a number of public data sets collected by the U.S. Census Bureau.
 We used a cleaned version of the 2012 Public Use Microdata Sample (PUMS) data kindly provided by the authors of~\cite{akande2015empirical}.
@@
 \end{itemize}
-The final data set consists of a single table with 671,153 rows and 35 columns, where all column entries are integers.
+The final data set consists of a single table with 671,153 rows and 37 columns, where all column entries are integers.
 
-\todo{Describe synthetic data}
+In order to evaluate our implementation, we created a dirtied variant of the ACS data by randomly eliminating
+10\% of the fields across the table.
+
+In order to evaluate additional queries, we constructed a synthetic table, consisting of 10 columns and 10,000 rows, with
+all values drawn from a uniform distribution between 0 and 100. To create a dirtied variant of the synthetic data, we randomly deleted 30\% of the fields
+across the table.
 
@@
 
-We collected a set of queries that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
+We collected a set of queries (Figure~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
 
 The queries on the ACS data all consist only of selections, filters, and aggregations.
-The ACS data is contained in a single table, so there are relatively few join queries that are interesting on this data set.
-However, even with select, filter, and aggregate, there are interesting choices to make with imputation placement.
+The ACS data is contained in a single table, so there are relatively few join queries that are interesting on this data set and remain relevant for analysts, rather
+than contrived solely for the purpose of experimentation.
+However, even with select, filter, and aggregate, there are interesting choices to make with imputation placement. In order to explore joins, we leverage the synthetic
+dataset and write various ad-hoc queries.
 
@@
   \caption{The queries used in our experiments.}
+  \label{fig:queries}
 \end{figure}
@@
   \caption{Running time and error for queries with different imputation levels.}
+  \label{fig:experiments}
 \end{figure}
+
+Figure~\ref{fig:experiments} provides a summary of our experimental results. For purposes of simplicity, we present only the results associated with
+queries that return a result size matching our base case. This addresses the fact that deriving a reasonable error metric for set-queries 
+is inherently task specific (e.g. consider tuning precision vs. recall). Meanwhile, for all queries that match, we apply a simple sum of errors squared
+(SSE). Figure~\ref{fig:experiments} indicates the query number and the error produced by a query engine with traditional SQL null semantics (base error).
+For imputations, we present two extremes of the $\alpha$ parameter: 0 and 1, which reflect a a focus on runtime and information loss, respectively.
+For each query, the table displays the new error ($\epsilon$), the planning time ($t_p$) and the running time ($t_r$) in seconds.
+
+The overall low planning times highlight the practicality of our algorithm, despite the exponential complexity. Runtimes clearly vary based on $\alpha$, allowing
+users to tune their query performance as desired.
+
+\todo{add result-specific analysis}
+
+Finally, we performed CERT imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
+We note that the latter took \todo{add time}, compared to our maximum of \todo{add time}. Given the varying needs of database users, \ProjectName{} clearly allows flexibility without
+a significant performance hit.
 

### 2016-12-11 paper/algorithm.tex
paper: break out cart algo to algorithms/cart
fix: remove extra end
fix: year in citation

@@
 are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
-routines and are widely used in epidemiological domains \cite{burgette2011multiple}.
+routines and are widely used in epidemiological domains \cite{burgette2010multiple}.
 
@@
 \begin{algorithm}
-  \begin{algorithmic}
-    \Require{$T$ is a table. $D$ is a set of attributes of $T$ that need to be imputed, and
-      $C$ is a set of attributes of $T$ that has complete data }
-    \Ensure{Returns an imputed $T$}
-
-    \Function{ImputeWithCE-CART}{$H, op$}
-    	\State $T' \gets \Call{ImputeRandom}{T, D}$
-	\For{$n\ EPOCHS$}
-		\For{$d \in D$}
-            \State $imp \gets \Call{TrainRT}{\pi_{C \cup D \backslash \{d\}}(T'), \pi_d{T'}}$
-            \State $T' \gets (\pi_{C \cup D \backslash \{d\}}(T'), \Call{PredictRT}{imp, T})$
-		\EndFor
-	\EndFor
-	\Return{T'}
-	\EndFunction
-  \end{algorithmic}
-  \caption{An algorithm for chained imputation using regression trees}
-  \label{algo:imputation-strategy}
-%\input{algorithms/cert.tex}
+    \input{algorithms/cart.tex}
 \end{algorithm}

### 2016-12-11 paper/algorithm.tex
paper: some cleanup, still missing motivating example and final experiments

@@
 This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
-Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value.
+Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value in any attribute that they directly operate on.
 
@@
 \label{sec:cost-model}
-We rank query plans $Q$ using a cost model $\text{Cost}(Q) = \alpha \times \text{Time}(Q) + (1 - \alpha) \times \text{Loss}(Q)$.
+We rank query plans $Q$ using a cost model $\text{Cost}(Q) = (1 - \alpha) \times \text{Time}(Q) + \alpha \times \text{Loss}(Q)$.
 $\text{Time}(Q)$ is an estimate of the runtime of the query which is derived from table statistics and selectivity estimation of the query predicates.
 $\text{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure.
-$\alpha$ is a parameter to the query optimizer that controls the emphasis on quality versus performance.
-$\alpha = 1.0$ means that the query should be as fast as possible, $\alpha=0.0$ means that the query should be as accurate as possible.
+$\alpha$ is a parameter to the query optimizer that controls the emphasis on quality over performance.
+$\alpha = 1.0$ means that the query should be as accurate as possible (at the expense of performance), $\alpha=0.0$ means that the query should be as fast
+as possible (at the expense of accuracy).
 
@@
 For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
-For simplicity, each of the logical nodes in a query plan refers to a set of histograms.
+For simplicity, each of the logical nodes in a query plan points to a set of histograms.
 When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
@@
 
-\todo{Micah: Explanation edited a bit. In the algorithm, when we impute attribute $d$, we
-condition on attributes $C \cup D \backslash \{d\}$. Hopefully the way I note it is clear;
-feel free to modify otherwise.}
-
 \subsection{Complexity}

### 2016-12-11 paper/experiments.tex
paper: some cleanup, still missing motivating example and final experiments

@@
 queries that return a result size matching our base case. This addresses the fact that deriving a reasonable error metric for set-queries 
-is inherently task specific (e.g. consider tuning precision vs. recall). Meanwhile, for all queries that match, we apply a simple sum of errors squared
+is inherently task specific (e.g. consider tuning precision vs. recall). Meanwhile, for all queries that match, we apply a simple sum of squared errors
 (SSE). Figure~\ref{fig:experiments} indicates the query number and the error produced by a query engine with traditional SQL null semantics (base error).
@@
 
-Finally, we performed CERT imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
+Finally, we performed $CE-CART$ imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
 We note that the latter took \todo{add time}, compared to our maximum of \todo{add time}. Given the varying needs of database users, \ProjectName{} clearly allows flexibility without

### 2016-12-12 paper/examples.tex
add paper skeleton
paper: skeleton experiments
paper: Add to examples section

--- 2016-11-28 paper/algorithm.tex
@@
-\section{Algorithm}
+\section{Examples}
+
+Consider the case of an analyst looking to explore the sources and causes of
+polling error in a presidential election. One hypothesis \cite{zukin2015s}
+is that the rapid switch from landlines to cellphones has increased the
+difficulty of contacting potential poll respondents. The analyst may
+begin her analysis by comparing the polling error in a state with the
+distribution of households with landlines, for the set of states for which
+polling error was most pronounced. In order to probe deeply, the analyst
+accesses a household-level survey of household characteristics that includes
+whether the household has a landline. One such survey --- which will be
+discussed further --- is the American Community Survey (ACS), conducted by the
+US Census Bureau.
+
+Surveys are rife with missing values due to non-response and other issues, and
+the ACS is no exception. The missing values could be imputed in an expensive
+operation\footnote{One would not necessarily want to impute the entire database once and
+store the imputed values, as in many settings the imputation strategy would be
+targeted for a specific question or domain.}, or the rows with missing values could be dropped entirely. With
+ImputeDB, the analyst could impute the relevant subset of data on-the-fly,
+achieving better performance without losing the information in the dropped
+examples.
+
+The analyst submits a query to ImputeDB which finds an optimized query plan.
+
+\begin{verbatim}
+SELECT polling.state, AVG(acs.TEL)
+FROM polling, acs
+WHERE polling.state = acs.ST
+  AND polling.error > 50
+GROUP BY polling.state
+\end{verbatim}
+
+\todo{explore a couple of plans generated, using dot visualizer}
 

### 2016-12-12 paper/algorithm.tex
paper: Add to examples section
paper: edit loss section and rearrange related work

@@
 
-The computation of $\text{Time}(Q)$ is standard and is comparable to the cost model used in any modern RDBMS.\@ Our runtime cost model is perhaps less sophisticated than most, as it is identical to the cost model in SimpleDB.
+The computation of $\text{Time}(Q)$ is standard and is comparable to the cost model used in
+any modern RDBMS.\@ Our runtime cost model is the cost model used in SimpleDB, though modern systems
+have more sophisticated approaches.
 
-\todo{Check loss calculation w/ Micah}
 $\text{Loss}(Q)$ is computed as follows.
-For each sub-query, in addition to a histogram, we keep track of the estimated number of null values in each column.
+For each sub-query, in addition to a histogram, we keep track of the estimated number of
+null values in each column, $\text{MissingCount}(c, Q)$.
 \begin{align*}
   \text{Loss}(Q) = \begin{cases}
-    \sum_{c \in C} \text{Null}(c, Q') & Q = \delta_C(Q') \\
-    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{Null}(c, Q') & Q = \mu_C(Q') \\
+    \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \delta_C(Q') \\
+    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \mu_C(Q') \\
     \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
@@
 \end{align*}
-Dropping tuples which contain a null value incurs a loss penalty of 1 for each null field dropped.
-Imputing null fields incurs a loss penalty $0 < p \leq 1$ which decreases as the number of tuples available to train the imputation algorithm increases.
-The intuition behind this formula is that imputation performance should increase if there is more training data available, but there is a saturation point beyond which more data provides little additional benefit.
+Dropping tuples which contain a null value incurs a loss of 1 for each null field dropped.
+Imputing null fields incurs a loss penalty $p \in (0, 1]$ which decreases as the number of
+tuples available to train the imputation algorithm increases.  Intuitively, imputation
+accuracy \textit{increases} when more complete attributes and complete tuples are available;
+correspondingly, the information loss \textit{decreases} with more complete values to work
+with. The inverse square loss can be seen as a first approximation to the learning guarantees of
+regression trees. \todo{Micah: Probably need to look into this more carefully since I don't
+know much about the learning guarantees of regression trees!} Since we conceive that
+alternate imputation strategies could be used, the analyst would provide a corresponding
+loss with a reasonable parameterization.
+
+\todo{Micah: Notation not super clear to me. I'm guessing $|Q'|$ is the number of tuples total, but
+$Q'$ is a query plan, and I would interpret it's cardinality as perhaps the height of the
+tree or number of elements in the tree, not in terms of underlying relation. Similarly, I'm
+not sure how to define $\text{MissingCount}(\cdot)$ above.}
 
@@
 
-The dirty set over-approximates the set of attributes that contain null.
+That is, the dirty set over-approximates the set of attributes that contain null.
 For example, a filter might remove all tuples which contain null without changing the dirty set, forcing an unnecessary imputation.
@@
 complete attributes and the other missing attributes. The individual predictive models can
-be customized by the researcher to the problem at hand. Decision tree algorithms, like CART,
+be customized by the analyst to the problem at hand. Decision tree algorithms, like CART,
 are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
@@
 provides details on the implementation.
+
+The flexibility of this non-parametric imputation approach aligns well with the spirit of
+query optimization, which aims to reconcile performance with a declarative interface to data
+manipulation. By removing concerns for imputation, users executing queries in \ProjectName{}
+don't need to consider the potential nature of the missing data.
 

### 2016-12-12 paper/algorithms/histogram-transform.tex

@@
-\begin{algorithm}
+%\begin{algorithm}
   \begin{algorithmic}
@@
 		\For{$c \in C$}
-			\State \Call{Add}{$H'[c], H'[c][null]$}
+			\State \Call{Add}{$H'[c], H'[c][null]$} 	\Comment{Adds to histogram based on distirbution}
 		\EndFor
 	\ElsIf{op = $\sigma$}
-		\State \Call{ScaleBy}{H',  \Call{Selectivity}{$\sigma$}}
+		\State \Call{ScaleBy}{$H',  \Call{Selectivity}{\sigma}$} 	\Comment{Scales all buckets by a constant factor}
 	\ElsIf{op = $\bowtie_\phi$}
-		\State \Call{ScaleTo}{H', \Call{Cardinality}{$\bowtie_{\phi}$}}
+		\State \Call{ScaleTo}{$H', \Call{Cardinality}{\bowtie_{\phi}}$} 	\Comment{Scales all buckets to sum to constant}
 	\EndIf
@@
   \label{algo:histogram-transformation}
-\end{algorithm}
+%\end{algorithm}

### 2016-12-12 paper/algorithms/plan-helpers.tex

@@
     \Function{AddImpute}{$q, C_{req}$}
-    \todo{jose: I modified this, since in reality we return only the same plan if it has no dirty values, we may still impute when the intersection is empty, but only maximally}
     \State $C_{min} \gets \Call{Dirty}{q} \cap C_{req}$
@@
     \Function{OptFilter}{$t, \phi$}
-    \todo{jose: I believe this was missing the sigma before}
     \State \Return $\Call{OptRel}{\{  \sigma_{\phi}(q) | q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}} \} }$
@@
     \Require{$Q$ is a map from tables and dirty sets to optimal query plans (possibly with imputations), and $\Psi$ relates query plans with join predicates.}
-    \Ensure{Returns a set of optimal query plans for all joins, with distinct dirty sets.}
+    \Ensure{Returns a map from dirty sets to optimal query plans involving all necessary joins}
     \Function{OptJoin}{$Q, \Psi$}
@@
 				\State $t_{plans} \gets Q(t)$
-				\For {$l, r \in t_{plans} \times S'_{plans}$}
+				\For {$l, r \in S'_{plans} \times t_{plans}$}
 					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\psi}}$
@@
   \label{algo:plan-helpers}
-\end{algorithm}
+%\end{algorithm}

### 2016-12-12 paper/examples.tex

@@
 whether the household has a landline. One such survey --- which will be
-discussed further --- is the American Community Survey (ACS), conducted by the
-US Census Bureau.
+discussed further (Sec~\ref{subsec:datasets}) --- is the American Community Survey (ACS),
+conducted by the US Census Bureau.
 
 Surveys are rife with missing values due to non-response and other issues, and
-the ACS is no exception. The missing values could be imputed in an expensive
-operation\footnote{One would not necessarily want to impute the entire database once and
-store the imputed values, as in many settings the imputation strategy would be
-targeted for a specific question or domain.}, or the rows with missing values could be dropped entirely. With
-ImputeDB, the analyst could impute the relevant subset of data on-the-fly,
-achieving better performance without losing the information in the dropped
-examples.
+the ACS is no exception. In this case, a non-negligible fraction of respondents omitted
+information on whether they have a landline. The analyst hypothesizes that older houses are
+more likely to have landlines and that households that are relatively new are less likely to
+have landlines.
 
-The analyst submits a query to ImputeDB which finds an optimized query plan.
+The analyst could take advantage of these presumed correlations to impute the missing values
+in an expensive operation\footnote{One would not necessarily want to impute the entire
+database once and store the imputed values, as in many settings the imputation strategy
+would be targeted for a specific question or domain.}. On the other hand, the query
+could be run directly and rows with missing values would be dropped entirely. With
+ImputeDB, the analyst could impute the relevant subset of data on-the-fly, achieving
+better performance without losing the information in the dropped examples.
+
+The analyst submits a query to ImputeDB which finds an optimized query plan. The resulting
+query imputes missing values for \verb!acs.TEL! after states with high polling error have
+been selected.
 
 \begin{verbatim}
-SELECT polling.state, AVG(acs.TEL)
+SELECT polling.ST, AVG(acs.TEL)
 FROM polling, acs
-WHERE polling.state = acs.ST
-  AND polling.error > 50
-GROUP BY polling.state
+WHERE polling.ST = acs.ST
+  AND polling.ERROR > 50         -- 5 percentage points
+GROUP BY polling.ST
 \end{verbatim}
 
-\todo{explore a couple of plans generated, using dot visualizer}
+\todo{Explore a couple of plans generated, using dot visualizer. Confirm that the actual
+plans match my guess in last sentence.}
 

### 2016-12-12 paper/introduction.tex

@@
 In this paper, we present \ProjectName{}, a database system which is designed to interact with a dirty dataset as though it were clean.
-The guiding design principle of \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
+The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
 To achieve this goal, we perform imputation on the fly, during query execution.
-Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows our system to flexibly trade accuracy for computation time.
+Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows users to flexibly trade accuracy for computation time.
 

### 2016-12-12 paper/relatedwork.tex

@@
 \section{Related Work}
+
+\todo{Micah: Should this section follow the introduction? A lot of this stuff
+becomes repetitive as we wouldn't be able to develop the body of the paper
+without having already made these citations.}
+
 \subsection{Missing Values and Statistics}
@@
 focus on correctly modeling relationships between the attributes to factor in
-such varied forms of missingness.
+varied forms of missingness.
 
-Burgette and Reiter introduce imputation of missing data using sequences
-of regression trees in \cite{burgette2010multiple}. \ProjectName{} uses
-this technique to impute missing values as needed in the generated query plan.
-The flexibility of this non-parametric imputation approach aligns well with the
-spirit of query optimization, which aims to reconcile performance with a declarative interface to
-data manipulation. By removing concerns for imputation, users executing queries
-in \ProjectName{} don't need to consider the potential nature of the missing data.
+Burgette and Reiter (\cite{burgette2010multiple}) discuss the usage of
+sequential regression trees for imputing missing data.
 
 In \cite{akande2015empirical}, Akande et al analyze the performance of various
-imputation techniques on American Community Survey data, "an ongoing survey
-that provides vital information on a yearly basis"\cite{censusbureau}. The authors
-use a common technique: multiple imputation, which imputes missing values
-with various strategies on the base data, computes aggregates, and then
-combines estimates of these aggregates to account for additional uncertainty across
-strategies. The computational difficulties of imputing on large base tables
-are well known and can limit approaches. For example, Akande finds that
-one of their approaches (MI-GLM) crashes when attempting to impute on data that
+multiple imputation techniques on the American Community Survey dataset
+(Sec~\ref{subsec:datasets}). The computational difficulties of imputing on large base
+tables are well known and can limit approaches. For example, Akande finds that
+one approaches (MI-GLM) crashes when attempting to impute on data that
 includes variables with potentially large domains (ten categories specifically).
- In contrast, \ProjectName{} allows users to specify a tradeoff
-between information loss and computational complexity (in terms of time). 
-Furthermore, the query planner's imputation is guided by the requirements of
-each specific query's operators, rather than requiring broad assumptions
-about query workloads.
+In contrast, \ProjectName{} allows users to specify a tradeoff between
+information loss and computational complexity (in terms of time). Furthermore,
+the query planner's imputation is guided by the requirements of each specific
+query's operators, rather than requiring broad assumptions about query
+workloads.
 

### 2016-12-12 presentation/main.tex

@@
 
-\begin{frame}{Overview}
+\begin{frame}{Introduction and Motivation}
 \begin{itemize}
 	\item Handling incorrect or dirty data is a complex and challenging problem for data scientists
-	\item $\text{missing data} \subset \text{dirty data}$
-	\item Cost of running imputation on entire dataset
-	\item Lack of flexibility
-	\item \ProjectName{} allows users to interact with dataset as though it were clean.
-	\item User should never see missing data or have to modify query.
+	\item $\text{missing data} \subset \text{dirty data}$ and requires imputation/dropping.
+	\item Cost of running imputation on entire dataset is a high barrier for analysis.
+	\item Lack of flexibility inherent in existing pre-processing approaches.
+	\item User should never see missing data or have to modify query. Want a DBMS that 
+		allow us to "pretend" data is perfect.
+	\item \ProjectName{} allows users to interact with dataset as though it were clean, planning
+	around necessary imputation/drops.
 \end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Motivating Example}
-\todo{motivating example}
 \end{frame}
@@
 	\item Exponential complexity, but tractable in practice with low planning times.
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Query Planning: Top-Level Algorithm}
-
-\begin{algorithm}[H]
- \begin{algorithmic}
-   \scriptsize
-   \Require{set of tables $T$, filter predicates $\Phi$, join predicates $\Psi$, a set of projection attributes $P$, optional set of grouping attributes $G$ and aggregator function $A$}
-   \Ensure{Returns optimized query plan.}
-
-  \Function{Plan}{$T, \Phi, \Psi, P, G, A$}
- \State  $Q \gets \{ \}$ \Comment{map from table and dirty set to plans }
-   \For{$t \in T$}
-   	\State $\phi \gets \Call{GetFilters}{t, \Phi}$
-	\State $P_{t} \gets \Call{OptFilter}{t, \phi}$
-	\State $Q \gets Q \cup \{ (t, \Call{Dirty}{q}) \mapsto q | q \in P_t \}$
-   \EndFor
-
-   \State $M \gets \{ \}$ \Comment{map from dirty set to plans, contains all joins necessary}
-   \State $M \gets  \Call{OptJoin}{Q, \Psi}$	\Comment{optimize joins}
-   \If{$G \neq \emptyset \land A \neq NULL$}
-   	\State $attrs \gets \Call{Attrs}{G} \cup \Call{Attrs}{A}$
-   	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \Call{GroupBy}{\Call{AddImpute}{q, attrs}, G, A}}$
-   \Else
-  	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \pi_{P}(\Call{AddImpute}{q, P}) }$
-  \EndIf
-
-\Return $\argmin_{q \in M} Cost(q)$
-\EndFunction
-
-  \end{algorithmic}
-  \caption{Top-level query planner with imputations.}
-\label{algo:top-level-planner}
-\end{algorithm}
-
-\end{frame}
-
-\begin{frame}[fragile]{Imputation Algorithm}
-\begin{itemize}
-	\item Iterative algorithm using regression trees: \textit{Chained Equations Regression Trees} (CERT)\cite{burgette2010multiple}
-	\item Significant cost to perform on entire dataset
-	\item \ProjectName{} allows flexible use
-\end{itemize}
-
-\begin{algorithm}[H]
-  \scriptsize
-  \begin{algorithmic}
-    \Require{$T$ is a table. $D$ is a set of $T$ attributes that need to be imputed, and $C$ is a set of $T$ attributes that has complete data }
-    \Ensure{Returns an imputed $T$}
-
-    \Function{ImputeWithCERT}{$H, op$}
-    	\State $T' \gets \Call{ImputeRandom}{T, D}$
-	\For{$n\ EPOCHS$}
-		\For{$d \in D$}
-			\State $imp \gets \Call{TrainRT}{\pi_C(T'), \pi_d{T'}}$
-			\State $T' \gets (\pi_C(T'), \Call{PredictRT}{imp, T})$
-		\EndFor
-	\EndFor
-	\Return{T'}
-	\EndFunction
-  \end{algorithmic}
-  \caption{An algorithm for chained imputation using regression trees}
-  \label{algo:imputation-strategy}
-\end{algorithm}
-\end{frame}
-
-\begin{frame}[fragile]{Experiments: Data}
-\begin{itemize}
-	\item Real dataset: American Community Survey (U.S. Census Bureau)
-		\begin{itemize}
-			\item Preprocessed version courtesy of authors of~\cite{akande2015empirical}.
-			\item 671,153 rows and 37 columns, all integers.
-			\item Randomly deleted 10\% of fields to create dirtied variant
-			\item Can write typical data analyst queries involving filters and aggregates
-		\end{itemize}
-	\item Synthetic dataset: drawn from uniform distribution $[0,100)$
-		\begin{itemize}
-			\item10,000 rows and 10 columns
-			\item Randomly deleted 30\% ti create dirtied variant
-			\item Crafted ad-hoc queries using joins
-		\end{itemize}
+	\item Imputation operator is a state-of-the-art regression-tree based method.
 \end{itemize}
@@
 \end{figure}
+\end{frame}
+
+\begin{frame}
+	\center
+	Extra slides with details for those interested
+\end{frame}
+
+\begin{frame}[fragile]{Motivating Example}
+\todo{motivating example}
+\end{frame}
+
+\begin{frame}[fragile]{Query Planning: Top-Level Algorithm}
+
+\begin{algorithm}[H]
+\scriptsize
+\input{../paper/algorithms/plan-toplevel.tex}
+\end{algorithm}
+
+\end{frame}
+
+\begin{frame}[fragile]{Imputation Algorithm}
+\begin{itemize}
+	\item Iterative algorithm using regression trees: \textit{Chained Equations Regression Trees} (CERT)\cite{burgette2010multiple}
+	\item Significant cost to perform on entire dataset
+	\item \ProjectName{} allows flexible use
+\end{itemize}
+\scriptsize
+
+\begin{algorithm}[H]
+\scriptsize
+\input{../paper/algorithms/cert.tex}
+\end{algorithm}
+\end{frame}
+
+\begin{frame}[fragile]{Experiments: Data}
+\begin{itemize}
+	\item Real dataset: American Community Survey (U.S. Census Bureau)
+		\begin{itemize}
+			\item Preprocessed version courtesy of authors of~\cite{akande2015empirical}.
+			\item 671,153 rows and 37 columns, all integers.
+			\item Randomly deleted 10\% of fields to create dirtied variant
+			\item Can write typical data analyst queries involving filters and aggregates
+		\end{itemize}
+	\item Synthetic dataset: drawn from uniform distribution $[0,100)$
+		\begin{itemize}
+			\item10,000 rows and 10 columns
+			\item Randomly deleted 30\% ti create dirtied variant
+			\item Crafted ad-hoc queries using joins
+		\end{itemize}
+\end{itemize}
 \end{frame}

### 2016-12-12 paper/relatedwork.tex
Add reference to data cleaning survey paper

@@
 \subsection{Missing Values and Statistics}
+
 Imputation of missing values is a widely studied field within the statistics and machine
@@
 focus on correctly modeling relationships between the attributes to factor in
-varied forms of missingness.
-
-Burgette and Reiter (\cite{burgette2010multiple}) discuss the usage of
-sequential regression trees for imputing missing data.
+varied forms of missingness. For example, Burgette and Reiter
+(\cite{burgette2010multiple}) discuss the usage of sequential regression trees
+for imputing missing data.
 
@@
 imputation strategy) of the query evaluation.
+
+Database system developers and others have worked on techniques to automatically
+detect dirty values, whether missing or otherwise, and rectify the errors is
+possible. A survey of methods and systems is provided in
+\cite{hellerstein2008quantitative}.
 

### 2016-12-12 paper/experiments.tex
paper: edit loss section and rearrange related work
paper: more todos
paper: cleaned up more todos

--- 2016-12-11 paper/experiments.tex
@@
 To evaluate the performance of our system, we generated plans for queries on two data sets: the American Community Survey and a synthetic dataset.
-\subsection{Data sets}
+\subsection{Data sets} \label{subsec:datasets}
 
@@
 
-\begin{figure}
+\todo{fix query references}
+\begin{table}
   \centerfloat
   \input{queries}
-  \caption{The queries used in our experiments.}
+  \caption{Queries used in our experiments.}
   \label{fig:queries}
-\end{figure}
+\end{table}
 
-\begin{figure}
+\begin{table}
   \centerfloat
   \input{runtimes}
-  \caption{Running time and error for queries with different imputation levels.}
+  \caption{Running time and errors (SSE) for queries with different imputation levels. base error corresponds to the
+error with traditional SQL null semantics. The errors for \ProjectName{} plans are labeled with $\epsilon$. The two
+$\alpha$ levels reflect a relative focus on information loss. $t_p$ and $t_r$ represent planning and run time, respectively, in seconds.}
   \label{fig:experiments}
-\end{figure}
+\end{table}
 
@@
 queries that return a result size matching our base case. This addresses the fact that deriving a reasonable error metric for set-queries 
-is inherently task specific (e.g. consider tuning precision vs. recall). Meanwhile, for all queries that match, we apply a simple sum of errors squared
-(SSE). Figure~\ref{fig:experiments} indicates the query number and the error produced by a query engine with traditional SQL null semantics (base error).
-For imputations, we present two extremes of the $\alpha$ parameter: 0 and 1, which reflect a a focus on runtime and information loss, respectively.
-For each query, the table displays the new error ($\epsilon$), the planning time ($t_p$) and the running time ($t_r$) in seconds.
-
+is inherently task specific (e.g. consider tuning precision vs. recall). For all other queries, we use sum of squared errors (SSE) as a metric.
 The overall low planning times highlight the practicality of our algorithm, despite the exponential complexity. Runtimes clearly vary based on $\alpha$, allowing
@@
 
-Finally, we performed CERT imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
+Finally, we performed CE-CART imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
 We note that the latter took \todo{add time}, compared to our maximum of \todo{add time}. Given the varying needs of database users, \ProjectName{} clearly allows flexibility without

### 2016-12-14 paper/experiments.tex
paper: keep working on experiment section

--- 2016-12-11 paper/experiments.tex
@@
 
-\begin{figure}
+\begin{table}
   \centerfloat
@@
   \label{fig:queries}
-\end{figure}
+\end{table}
 
-\begin{figure}
+\begin{table}
   \centerfloat
   \input{runtimes}
-  \caption{Running time and error for queries with different imputation levels.}
+  \caption{Running time and error for queries with different imputation levels. Base error is the RMSE between the query run on clean data and the query run on dirty data without imputation.}
   \label{fig:experiments}
-\end{figure}
+\end{table}
 
-Figure~\ref{fig:experiments} provides a summary of our experimental results. For purposes of simplicity, we present only the results associated with
-queries that return a result size matching our base case. This addresses the fact that deriving a reasonable error metric for set-queries 
-is inherently task specific (e.g. consider tuning precision vs. recall). Meanwhile, for all queries that match, we apply a simple sum of squared errors
-(SSE). Figure~\ref{fig:experiments} indicates the query number and the error produced by a query engine with traditional SQL null semantics (base error).
-For imputations, we present two extremes of the $\alpha$ parameter: 0 and 1, which reflect a a focus on runtime and information loss, respectively.
-For each query, the table displays the new error ($\epsilon$), the planning time ($t_p$) and the running time ($t_r$) in seconds.
+Figure~\ref{fig:experiments} provides a summary of our experimental results.
+We run each query in four configurations: without imputation on clean data, without imputation on dirty data, and with imputation on dirty data for $\alpha = 0$ and $\alpha = 1$.
+When a query returns a relation of the same ``shape'' (same number of tuples and same schema) on both clean and dirty data, we can compute an error value.
+For our experiments, we report root-mean-square error (RMSE).
+For imputations, we run queries using the two extreme values of $\alpha$: 0 and 1, representing a focus on runtime and on information loss, respectively.
+For each query, the table displays the error between the query run on clean data and the query run on dirty data with imputation and the running time in seconds.
 
-The overall low planning times highlight the practicality of our algorithm, despite the exponential complexity. Runtimes clearly vary based on $\alpha$, allowing
-users to tune their query performance as desired.
+Planning times are not shown, as they are fractions of a second for all of the queries that we tested.
+The overall low planning times highlight the practicality of our algorithm, despite the exponential complexity.
+Running times clearly vary based on $\alpha$, allowing users to tune their query performance as desired.
 
-\todo{add result-specific analysis}
+We can group our queries into five categories: sum/count queries, average queries, min/max queries, selection queries, and join queries on synthetic data.
+
+\begin{itemize}
+\item For sum and count queries (\ref{q1}, \ref{q2}, \ref{q4}), imputation does a good job of reducing the error.
+For each of these queries, we can reduce total error by several orders of magnitude.
+Imputation performs very well in these cases, even for data which is missing uniformly at random, because each missing value directly increases the error of the result.
+
+\item For average queries (\ref{q3}, \ref{q5}, \ref{q6}), imputation does not reduce the error, mainly because running these queries on dirty data incurs only a small penalty.
+This is because, in the case of average, removing data uniformly at random does not change the mean.
+If we were to remove data in a biased way, imputation would perform better because there would be more error for it to correct.
+
+\item For min and max queries (\ref{q7}, \ref{q8}, \ref{q12}), it is unlikely that imputation will ever help, because by definition these queries are looking for extreme values.
+These values are unlikely to be the result of imputation and are also unlikely to be randomly removed from the dataset.
+
+\item We do not compute error for selection queries (\ref{q9}, \ref{q10}, \ref{q11}), because the output of the query on clean data is not directly comparable to the output on dirty data.
+
+  \item Imputation performs poorly on the join queries on synthetic data (\ref{q13}, \ref{q14}) because the data is synthetic, so there are no correlations for imputation to find.
+We included these queries because we wanted to test the performance of our algorithm on queries that include joins, and we were pleased to find that simple joins cause no perceptible performance problems.
+\end{itemize}
 
 Finally, we performed $CE-CART$ imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
-We note that the latter took \todo{add time}, compared to our maximum of \todo{add time}. Given the varying needs of database users, \ProjectName{} clearly allows flexibility without
-a significant performance hit.
+We were unable to complete the full imputation due to resource constraints, but the process ran for more than one hour without terminating.
+This time compares favorably to our maximum of \todo{add time}.
+Given the varying needs of database users, \ProjectName{} clearly allows flexibility without a significant performance hit.
 

### 2016-12-14 paper/gen_tables.py

@@ -8,9 +8,15 @@ def parse_time(t):
 
 def err_to_str(err):
     if err == 'null':
-        return r'$\bot$'
+        return r'\multicolumn{1}{c}{--}'
     else:
-        return str(err)
+        return '%.2e' % float(err)
+
+def err_diff_to_str(err1, err2):
+    if err1 == 'null' or err2 == 'null':
+        return r'\multicolumn{1}{c}{--}'
+    else:
+        return '%.2e' % (float(err1) - float(err2))
 
 def get_first_row(df, l):
     for (_, row) in df.iterrows():
@@ -30,13 +36,14 @@ def main():
     with open('queries.tex', 'w') as f:
         qidx = {}
         idxq = {}
-        ctr = 0
-        writel(f, r'\begin{tabular}{ll}')
+        ctr = 1
+        writel(f, r'\newcounter{queryno}')
+        writel(f, r'\begin{tabular}{cl}')
         writel(f, r'\toprule')
-        writel(f, r'\# & Query \\')
+        writel(f, r'\# & \multicolumn{1}{c}{Query} \\')
         writel(f, r'\midrule')
         for query in df['query'].unique():
-            writel(f, r'%d & \verb|%s| \label{q%d} \\' % (ctr, query, ctr))
+            writel(f, r'%d & \verb|%s| \refstepcounter{queryno} \label{q%d} \\' % (ctr, query, ctr))
             qidx[query] = ctr
             idxq[ctr] = query
             ctr += 1
@@ -44,12 +51,12 @@ def main():
         f.write(r'\end{tabular}' + '\n')
 
     with open('runtimes.tex', 'w') as f:
-        f.write(r'\begin{tabular}{llllllll}' + '\n')
+        f.write(r'\begin{tabular}{cSSSSSSS}' + '\n')
         writel(f, r'\toprule')
-        writel(f, r'\multicolumn{2}{c}{} & \multicolumn{3}{c}{Imputed ($\alpha=0.0$)} & \multicolumn{3}{c}{Imputed ($\alpha=1.0$)} \\')
-        writel(f, r'\cmidrule(r){3-5}')
-        writel(f, r'\cmidrule(l){6-8}')
-        writel(f, r'Query \# & Base error & $\epsilon$ & $t_p$ & $t_r$ & $\epsilon$ & $t_p$ & $t_r$ \\')
[…]

### 2016-12-14 paper/algorithm.tex

@@
 $\text{Loss}(Q)$ is computed as follows.
-For each sub-query, in addition to a histogram, we keep track of the estimated number of
+For each sub-query, using the histogram information, we keep track of the estimated number of
 null values in each column, $\text{MissingCount}(c, Q)$.
@@
 correspondingly, the information loss \textit{decreases} with more complete values to work
-with. The inverse square loss can be seen as a first approximation to the learning guarantees of
-regression trees. \todo{Micah: Probably need to look into this more carefully since I don't
-know much about the learning guarantees of regression trees!} Since we conceive that
+with. The inverse square loss can be seen as a heuristic first approximation to the learning guarantees of
+regression trees. Since we conceive that
 alternate imputation strategies could be used, the analyst would provide a corresponding
 loss with a reasonable parameterization.
-
-\todo{Micah: Notation not super clear to me. I'm guessing $|Q'|$ is the number of tuples total, but
-$Q'$ is a query plan, and I would interpret it's cardinality as perhaps the height of the
-tree or number of elements in the tree, not in terms of underlying relation. Similarly, I'm
-not sure how to define $\text{MissingCount}(\cdot)$ above.}
 

### 2016-12-14 paper/experiments.tex
paper: fix unmerged file

@@
   \input{queries}
-  \caption{The queries used in our experiments.}
+  \caption{Queries used in our experiments.}
   \label{fig:queries}
@@
 
-Finally, we performed $CE-CART$ imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
+Finally, we performed CE-CART imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
 We were unable to complete the full imputation due to resource constraints, but the process ran for more than one hour without terminating.
-This time compares favorably to our maximum of \todo{add time}.
+This time compares favorably to our maximum of \todo{add time} for a query with imputation.
 Given the varying needs of database users, \ProjectName{} clearly allows flexibility without a significant performance hit.

### 2016-12-14 paper/algorithm.tex
paper: some cleanup in language/typos

@@
 
-The computation of $\text{Time}(Q)$ is standard and is comparable to the cost model used in
-any modern RDBMS.\@ Our runtime cost model is the cost model used in SimpleDB, though modern systems
+Our runtime cost model is the cost model used in SimpleDB, though modern systems
 have more sophisticated approaches.
 
-$\text{Loss}(Q)$ is computed as follows.
 For each sub-query, using the histogram information, we keep track of the estimated number of
-null values in each column, $\text{MissingCount}(c, Q)$.
+null values in a column  ($\text{MissingCount}(c, Q)$). We use this estimate to compute
+$\text{Loss}(Q)$ as follows.
+
 \begin{align*}
@@
 regression trees. Since we conceive that
-alternate imputation strategies could be used, the analyst would provide a corresponding
+alternate imputation strategies could be used, the analyst may provide a corresponding
 loss with a reasonable parameterization.
@@
 \subsection{Query planning}
-The input to our query planner is a tuple $(T, \Phi, \Psi, P, G, A)$: a set of tables $T$, a set of filter predicates $\phi_t, t \in T$, a set of join predicates $\psi_(t_1, t_2), t_1, t_2 \in T$, a set of attributes $P$, and an optional set of attributes $G$ and aggregation function $A \in \{\text{Max}, \text{Min}, \text{Sum}, \text{Avg}, \text{Count}\}$.
+The input to our query planner is a tuple $(T, \Phi, \Psi, P, G, A)$: a set of tables $T$, a set of filter predicates $\phi_t, t \in T$, a set of join predicates $\psi_{(t_1, t_2)}, t_1, t_2 \in T$, a set of attributes $P$, and an optional set of attributes $G$ and aggregation function $A \in \{\text{Max}, \text{Min}, \text{Sum}, \text{Avg}, \text{Count}\}$.
 
@@
 equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
-regression} \cite{gelman2006data} and others) impute multiple missing attributes by
+regression} \cite{gelman2006data}) impute multiple missing attributes by
 iteratively estimating predictive models of one missing attribute conditional on all

### 2016-12-14 paper/algorithms/histogram-transform.tex
paper: some cleanup in language/typos

@@
     \Require{$H$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
-    \Ensure{Returns an updated $H$, such that the distribution of data matches the original $H$}
+    \Ensure{Returns an updated histogram, such that the distribution of data matches the original $H$}
 
@@
 		\For{$c \in C$}
-			\State \Call{Add}{$H'[c], H'[c][null]$} 	\Comment{Adds to histogram based on distirbution}
+			\State \Call{Add}{$H'[c], H'[c][null]$} 	\Comment{Adds to histogram based on distribution}
 		\EndFor
 	\ElsIf{op = $\sigma$}
-		\State \Call{ScaleBy}{$H',  \Call{Selectivity}{\sigma}$} 	\Comment{Scales all buckets by a constant factor}
+		\State \Call{ScaleBy}{$H',  \Call{Selectivity}{\sigma}$} 	\Comment{Scales all buckets by constant factor}
 	\ElsIf{op = $\bowtie_\phi$}
@@
 	\EndIf
-	\Return{H'}
+	\Return{$H'$}
     \EndFunction

### 2016-12-14 paper/experiments.tex
paper: some cleanup in language/typos

@@
 
-In order to evaluate additional queries, we constructed a synthetic table, consisting of 10 columns and 10,000 rows, with
+In order to evaluate additional queries, we constructed a synthetic table, consisting of 10,000 rows and 10 columns, with
 all values drawn from a uniform distribution between 0 and 100. To create a dirtied variant of the synthetic data, we randomly deleted 30\% of the fields
@@
 This is because, in the case of average, removing data uniformly at random does not change the mean.
-If we were to remove data in a biased way, imputation would perform better because there would be more error for it to correct.
+If we were to remove data in a biased way, imputation would likely perform better because there would be more error for it to correct.
 

### 2016-12-14 paper/introduction.tex
paper: some cleanup in language/typos

@@
 One way in which a dataset can be dirty is for parts of it to be missing altogether.
-Missing data is one of the simpler variants of dirty data, but if handled naively, it can still cause analyses to be incorrect.
+Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
 To handle this problem, users may manually clean their dataset by performing some regression analysis to replace missing data elements with likely values.
@@
 Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
-Although manual imputation solves the problem of missing data, in the age of big data, it may be very expensive to run an imputation algorithm on an entire dataset.
+Although manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
 Additionally, it may not even be necessary to completely clean the data to make it usable.
 Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
-Others may want to run queries on a subset of the data, so do not need to impute every field of every record.
+Others may want to run queries on a subset of the data, so do not need to impute every field in every record.
 

### 2016-12-14 paper/main.tex
start filling out algorithm section from onenote
paper: added join algo sketched, added chained equations algo sketched, added histogram transformation algo
paper: macro for project name
paper: switch to biblatex + biber
paper: add code to generate tables
paper: add to experiments section
paper: some cleanup, still missing motivating example and final experiments
paper: move related work
paper: keep working on experiment section
paper: some cleanup in language/typos

@@
 \documentclass{article}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\usepackage[english]{babel}
+\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
+\usepackage{amsfonts,amssymb,amsmath}
+\usepackage{xcolor}
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
+\usepackage[style=ieee]{biblatex}
+\usepackage{booktabs}
+\usepackage{siunitx}
+\usepackage{hyperref} % load after other packages
+\addbibresource{references.bib}
 
-\title{Data Imputation as a Query Optimization}
+\makeatletter
+\newcommand*{\centerfloat}{%
+  \parindent \z@
+  \leftskip \z@ \@plus 1fil \@minus \textwidth
+  \rightskip\leftskip
+  \parfillskip \z@skip}
+\makeatother
+
+\DeclareMathOperator*{\argmax}{arg\,max}
+\DeclareMathOperator*{\argmin}{arg\,min}
+
+\MakeRobust{\Call}
+
+\renewcommand{\emptyset}{\varnothing}
+
+\newcommand{\ProjectName}{ImputeDB}
+
+\title{\ProjectName{}: Data Imputation as a Query Optimization}
 
@@
   \and
-  Micah Cohen
+  John K. Feser
   \and
-  John K. Feser}
+  Micah Smith}
 
@@
 \newcommand{\new}{\marginpar{NEW}}
+\newcommand{\todo}[1]{{\color{red}\textbf{TODO:} #1}}
 
@@
 In order to study the placement of an imputation step, we create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process in SimpleDB.
-We introduce measures of uncertainty and completeness of query results as well as the cost of imputation, which outline the main trade-offs in the imputation placement.
+We introduce measures for information loss and runtime for imputation operations, which outline the main trade-offs in the imputation placement.
 We add these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning.
@@
 \input{introduction}
+\input{relatedwork}
 \input{examples}
@@
 \input{experiments}
-\input{relatedwork}
 \input{conclusion}
 
-\bibliography{references}
+\printbibliography
 

### 2016-12-14 paper/relatedwork.tex
paper: some cleanup in language/typos

@@
 \section{Related Work}
-
-\todo{Micah: Should this section follow the introduction? A lot of this stuff
-becomes repetitive as we wouldn't be able to develop the body of the paper
-without having already made these citations.}
 
@@
 Database system developers and others have worked on techniques to automatically
-detect dirty values, whether missing or otherwise, and rectify the errors is
+detect dirty values, whether missing or otherwise, and rectify the errors if
 possible. A survey of methods and systems is provided in
@@
 BayesDB \cite{mansinghka2015bayesdb} provides users with a simple interface to 
-leverage statistical inference techniques on a database. Non-experts
+leverage statistical inference techniques in a database. Non-experts
 can use a simple declarative language, akin to SQL, to specify models

### 2016-12-14 simpledb/test/example/dirty2.jl
Generate example

@@ -1,0 +1,34 @@
+using DataFrames
+
+version = 2
+othp = 0.02
+telp = 0.11
+stp = 0.00
+infile = "acs.txt"
+outfile = "acs_dirty$version.txt"
+f = open(outfile, "w")
+
+df = readtable(infile, separator=',', header=false)
+
+iST = 1
+iTEL = 14
+d = ncol(df) # 37
+n = nrow(df) # 671153
+
+# fill in p
+p = othp * ones(d)
+p[iTEL] = telp
+p[iST] = stp
+
+for i in 1:n
+    row = Array{Any}(df[i,:]) # sketchy to allow us to print ""
+    for j in 1:d
+        if rand() < p[j]
+            row[j] = ""
+        end
+    end
+
+    println(f, join(row, ","))
+end
+
+close(f)

### 2016-12-14 simpledb/test/example/dirty3.jl

@@ -1,0 +1,34 @@
+using DataFrames
+
+version = 3
+othp = 0.00
+telp = 0.15
+stp = 0.00
+infile = "acs.txt"
+outfile = "acs_dirty$version.txt"
+f = open(outfile, "w")
+
+df = readtable(infile, separator=',', header=false)
+
+iST = 1
+iTEL = 14
+d = ncol(df) # 37
+n = nrow(df) # 671153
+
+# fill in p
+p = othp * ones(d)
+p[iTEL] = telp
+p[iST] = stp
+
+for i in 1:n
+    row = Array{Any}(df[i,:]) # sketchy to allow us to print ""
+    for j in 1:d
+        if rand() < p[j]
+            row[j] = ""
+        end
+    end
+
+    println(f, join(row, ","))
+end
+
+close(f)

### 2016-12-14 simpledb/test/example/run.sh

@@ -1,0 +1,16 @@
+#!/usr/bin/env bash
+
+# # convert dat to txt file so we can read it with other programs
+# java -jar dist/simpledb.jar print test/testdata/acs.dat 37 > acs.txt
+
+VERSION=3
+julia dirty$VERSION.jl
+java -jar dist/simpledb.jar convert acs_dirty$VERSION.txt 37 > acs_dirty$VERSION.dat
+echo "acs_dirty$VERSION (ST int, NP int, ACR int, AGS int, BATH int, BDSP int, BLD int, BUS int, REFR int, RMSP int, RWAT int, SINK int, STOV int, TEL int, TEN int, TOIL int, VEH int, YBL int, HHL int, HHT int, HUGCLNPP int, HUPAC int, HUPAOC int, HUPARC int, LNGI int, MULTG int, MV int, NR int, NRC int, PARTNER int, PSF int, R18 int, R65 int, SRNTVAL int, WIF int, WKEXREL int, WORKSTAT int)" \
+    | cat catalog1.txt - > catalog$VERSION.txt
+echo "SELECT polling.ST, AVG(acs_dirty$VERSION.TEL) FROM polling, acs_dirty$VERSION WHERE polling.ST = acs_dirty$VERSION.ST AND polling.ERROR > 50 GROUP BY polling.ST;" \
+    > query$VERSION.txt
+java -jar dist/simpledb.jar draw catalog$VERSION.txt query$VERSION.txt e9 0.1 > e9.txt
+java -jar dist/simpledb.jar draw catalog$VERSION.txt query$VERSION.txt e10 1.0 > e10.txt
+dot -Tpng e9_0.dot > e9_0.png
+dot -Tpng e10_0.dot > e10_0.png

### 2016-12-14 paper/examples.tex
Add example to paper

--- 2016-11-28 paper/algorithm.tex
@@
-\section{Algorithm}
+\section{Examples}
+
+Consider the case of an analyst looking to explore the sources and causes of
+polling error in a presidential election. One hypothesis \cite{zukin2015s}
+is that the rapid switch from landlines to cellphones increases the
+difficulty of contacting potential poll respondents. The analyst may
+begin her analysis by comparing the polling error in a state with the
+distribution of households with landlines, for the set of states for which
+polling error was most pronounced. In order to probe deeply, the analyst
+accesses a household-level survey that includes data on
+whether the household has a landline. One such survey --- which will be
+discussed further (Sec~\ref{subsec:datasets}) --- is the American Community Survey (ACS),
+conducted by the US Census Bureau.
+
+Surveys are rife with missing values due to non-response and other issues, and
+the ACS is no exception. The Census Bureau and other similar organization attempt to present
+data in as unadulterated a fashion as possible. The ACS dataset is not amenable to one-time
+imputation by either the Census Bureau, for this reason, or the analyst's organization,
+which likely uses the same database for a variety of unrelated uses (asking for different
+imputation strategies). Thus, the analyst must deal with missing values in one way or
+another.
+
+In this case, a non-negligible fraction of respondents omitted information on whether they
+have a landline. The analyst hypothesizes that older houses are more likely to have
+landlines and that households that are relatively new are less likely to have landlines. 
+The analyst could take advantage of these presumed correlations to impute the missing values
+on the base table in one expensive operation. On the other hand, the query could be run
+directly and rows with missing values may be dropped entirely. With ImputeDB, the analyst
+could impute the relevant subset of data on-the-fly, achieving better performance without
+losing the information in the dropped examples.
+
+The analyst submits a query (\todo{ref to listing}) to ImputeDB which finds an optimized query
+plan\footnote{The analyst specifies a relatively high value of $\alpha$, the quality
+emphasis parameter, in order to receive a plan that does an expensive imputation step. See
+Section~\ref{sec:cost-model} for more details.}. The resulting query plan
+(Figure~\ref{fig:query-plan})
+imputes missing values for \verb!acs.TEL! after states with high polling error have been
+selected. This decision reduces the set of tuples that must be input into the imputation
+algorithm to just those that are relevant to the query. Later, we show the algorithms that
+the ImputeDB query optimizer uses in order to detect these opportunities and control the
+tradeoffs between efficiency and imputation quality.
+
+\begin{verbatim}
+SELECT polling.ST, AVG(acs.TEL)
+FROM polling, acs
+WHERE polling.ST = acs.ST
+  AND polling.ERROR > 50         -- 5 percentage points
+GROUP BY polling.ST;
+\end{verbatim}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width=0.6\textwidth]{figures/example.png}
+    \caption{Query plan generated by ImputeDB prioritizing imputation quality.}
+    \label{fig:query-plan}
+\end{figure}
 

### 2016-12-14 paper/algorithm.tex
paper: Add some last edits

@@
 \subsection{Imputation operators}
-We introduce two new relational operators to perform imputation: impute ($\mu$) and drop ($\delta$).
-Each operator takes arguments $(C, R)$ where $C$ is a set of attributes and $R$ is a relation.
-Impute uses a machine learning method to replace all null values with non-null values for attributes in $C$ in the relation $R$.
-Drop simply removes from $R$ all tuples which have a null value for some attribute in $C$.
-Both operators guarantee that the resulting relation will contain no null values for attributes in $C$.
+We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
+\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
+attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
+replace all null values with non-null values for attributes in $C$ in the relation $R$.
+\textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
+Both operators guarantee that the resulting relation will contain no null values for
+attributes in $C$.
+
+\subsection{Search space}
+\label{sec:search-space}
+To reduce the size of the query plan search space, only plans that fit the following template are considered.
+First, all filters are pushed to the leaves of the query tree, immediately after the scans.
+Joins are performed after filtering, and only left-deep plans are considered.
+Any group-by/aggregate will be performed after the final join.
+Finally, projections are placed at the root of the query tree.
+The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
 
@@
 For each sub-query, using the histogram information, we keep track of the estimated number of
-null values in a column  ($\text{MissingCount}(c, Q)$). We use this estimate to compute
+null values in a column  ($\text{MissingCount}(c, Q)$), given the histograms associated with
+the sub-query. We use this estimate to compute
 $\text{Loss}(Q)$ as follows.
@@
 
-\subsection{Search space}
-\label{sec:search-space}
-To reduce the size of the query plan search space, only plans that fit the following template are considered.
-First, all filters are pushed to the leaves of the query tree, immediately after the scans.
-Joins are performed after filtering, and only left-deep plans are considered.
-Any group-by/aggregate will be performed after the final join.
-Finally, projections are placed at the root of the query tree.
-The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
+\todo{is it necessary to add a fudge factor to the loss calculation, for generality, and set
+it at 1.0 for our paper?}
 
@@
 are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
-routines and are widely used in epidemiological domains \cite{burgette2010multiple}.
+routines and are widely used in epidemiological domains \cite{burgette2010multiple}. Though
+we provide quantitative results for the CE-CART implementation, ImputeDB is designed to be
+agnostic to the choice of imputation algorithm, and additional algorithms --- with
+associated time and loss functions --- could be insert without any change in design.
 
@@
 ImputeDB, we cannot assume the user desires multiple copies of the query result, though we
-note that in further work, multiple imputation could be used within ImputeDB to generate
-lower-variance estimates of some aggregates, like averages.
+note that in further work, multiple imputation could be used within ImputeDB to reduce the
+variance of estimates of some aggregates, like averages.
 
@@
 algorithms that learn in an online nature; this would reduce the cost of the imputation at
-the possible expense of quality.
+the possible expense of quality. Again, given the imputation strategy-agnostic design of
+ImputeDB, this would not pose a problem.
 
@@
 \subsection{Complexity}
-Our optimization algorithm builds off the approach taken by System R\cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
+Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
 note that if we remove our restriction on types of imputation($\delta_{min}, \mu_{min}, \mu_{max}$), and allow any arbitrary subset of attributes to be imputed,

### 2016-12-14 paper/experiments.tex
paper: Add some last edits

@@
 
-To evaluate the performance of our system, we generated plans for queries on two data sets: the American Community Survey and a synthetic dataset.
+To evaluate the performance of our system, we generate plans for queries on two data sets: the American Community Survey and a synthetic dataset.
 \subsection{Data sets} \label{subsec:datasets}
@@
 
-We collected a set of queries (Figure~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
+We collected a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
 
@@
   \input{runtimes}
-  \caption{Running time and error for queries with different imputation levels. Base error is the RMSE between the query run on clean data and the query run on dirty data without imputation.}
+    \caption{Error ($\epsilon$, planning time ($t_p$), and running time ($t_r$)i or queries
+    with different imputation levels. Base error is the RMSE between the query run on clean
+    data and the query run on dirty data without imputation. \todo{Micah: Not clear to me
+    what the symbol means.}}
   \label{fig:experiments}
@@
 
-Figure~\ref{fig:experiments} provides a summary of our experimental results.
+Table~\ref{fig:experiments} provides a summary of our experimental results.
 We run each query in four configurations: without imputation on clean data, without imputation on dirty data, and with imputation on dirty data for $\alpha = 0$ and $\alpha = 1$.
@@
 
-Planning times are not shown, as they are fractions of a second for all of the queries that we tested.
+Planning times are not shown\todo{they are shown...}, as they are fractions of a second for all of the queries that we tested.
 The overall low planning times highlight the practicality of our algorithm, despite the exponential complexity.
@@
 
+\todo{consider giving up on getting the references to work?}
 \begin{itemize}

### 2016-12-14 paper/introduction.tex
paper: Add some last edits

@@
 Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
-To handle this problem, users may manually clean their dataset by performing some regression analysis to replace missing data elements with likely values.
+To handle this problem, users may manually clean their dataset by performing
+some statistical analysis to replace missing data elements with likely values.
 This process is called \emph{imputation}.
@@
 Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
-Others may want to run queries on a subset of the data, so do not need to impute every field in every record.
+Others may want to run queries on a subset of the data, so do not need to impute
+every field in every record. Yet others may want to customize the 
+imputation algorithm for the tradeoffs and demands of a particular domain.
 

### 2016-12-14 paper/relatedwork.tex
paper: Add some last edits

@@
 tables are well known and can limit approaches. For example, Akande finds that
-one approaches (MI-GLM) crashes when attempting to impute on data that
-includes variables with potentially large domains (ten categories specifically).
+one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
+includes variables with potentially large domains (ten categories in their case).
 In contrast, \ProjectName{} allows users to specify a tradeoff between
@@
 While BayesDB can be used for value imputation, this step is not framed
-within the context of query planning, but rather as an explicit step, using the
-\verb|INFER| operation.
+within the context of query planning, but rather as an explicit statistical
+inference step within the query language, using the \verb|INFER| operation.
 

### 2016-12-14 paper/examples.tex
paper: example query into lstlisting env, wrapped in figure and added ref

@@
 
-The analyst submits a query (\todo{ref to listing}) to ImputeDB which finds an optimized query
+The analyst submits a query (Figure~\ref{fig:example-query}) to ImputeDB which finds an optimized query
 plan\footnote{The analyst specifies a relatively high value of $\alpha$, the quality
@@
 the ImputeDB query optimizer uses in order to detect these opportunities and control the
-tradeoffs between efficiency and imputation quality.
+tradeoffs between efficiency and imputation quality. Finally, note that the optimizer doesn't
+place any imputation operator for the \verb|acs.ST| column, as optimizer uses the histogram to
+detect that there are no missing values there (as might be expected for a survey question that isn't
+impacted by non-responses).
 
-\begin{verbatim}
+\begin{figure}
+\begin{lstlisting}[language=SQL]
 SELECT polling.ST, AVG(acs.TEL)
@@
 GROUP BY polling.ST;
-\end{verbatim}
+\end{lstlisting}
+\caption{A typical analyst query on ACS data}
+\label{fig:example-query}
+\end{figure}
 

### 2016-12-14 paper/main.tex
Add example to paper
paper: example query into lstlisting env, wrapped in figure and added ref

@@
 \usepackage{hyperref} % load after other packages
+\usepackage{graphicx} % includegraphics
+\usepackage{listings} % code samples with syntax hl
 \addbibresource{references.bib}
+
+% listings settings
+\lstset{ %
+  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}; should come as last argument
+  basicstyle=\footnotesize,        % the size of the fonts that are used for the code
+  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
+  breaklines=true,                 % sets automatic line breaking
+  captionpos=b,                    % sets the caption-position to bottom
+  commentstyle=\color{red},    % comment style
+  frame=single,	                   % adds a frame around the code
+  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
+  keywordstyle=\color{blue},       % keyword style
+  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
+  numbersep=5pt,                   % how far the line-numbers are from the code
+  numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
+  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
+  stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
+  stringstyle=\color{mauve},     % string literal style
+  tabsize=2,	                   % sets default tabsize to 2 spaces
+}
 

### 2016-12-14 simpledb/test/example/run.sh
Add acs_dirty3 file used for example

@@ -5,12 +5,12 @@
 
 VERSION=3
 julia dirty$VERSION.jl
-java -jar dist/simpledb.jar convert acs_dirty$VERSION.txt 37 > acs_dirty$VERSION.dat
+java -jar ../../dist/simpledb.jar convert acs_dirty$VERSION.txt 37 > acs_dirty$VERSION.dat
 echo "acs_dirty$VERSION (ST int, NP int, ACR int, AGS int, BATH int, BDSP int, BLD int, BUS int, REFR int, RMSP int, RWAT int, SINK int, STOV int, TEL int, TEN int, TOIL int, VEH int, YBL int, HHL int, HHT int, HUGCLNPP int, HUPAC int, HUPAOC int, HUPARC int, LNGI int, MULTG int, MV int, NR int, NRC int, PARTNER int, PSF int, R18 int, R65 int, SRNTVAL int, WIF int, WKEXREL int, WORKSTAT int)" \
     | cat catalog1.txt - > catalog$VERSION.txt
 echo "SELECT polling.ST, AVG(acs_dirty$VERSION.TEL) FROM polling, acs_dirty$VERSION WHERE polling.ST = acs_dirty$VERSION.ST AND polling.ERROR > 50 GROUP BY polling.ST;" \
     > query$VERSION.txt
-java -jar dist/simpledb.jar draw catalog$VERSION.txt query$VERSION.txt e9 0.1 > e9.txt
-java -jar dist/simpledb.jar draw catalog$VERSION.txt query$VERSION.txt e10 1.0 > e10.txt
+java -jar ../../dist/simpledb.jar draw catalog$VERSION.txt query$VERSION.txt e9 0.1 > e9.txt
+java -jar ../../dist/simpledb.jar draw catalog$VERSION.txt query$VERSION.txt e10 1.0 > e10.txt
 dot -Tpng e9_0.dot > e9_0.png
 dot -Tpng e10_0.dot > e10_0.png

### 2016-12-14 presentation/main.tex
presentation: cleaned up
presentation: added example

@@
 \presetkeys{todonotes}{inline}{}
+\usepackage[style=ieee]{biblatex}
+
+\usepackage{listings} % code samples with syntax hl
+% listings settings
+\lstset{ %
+  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}; should come as last argument
+  basicstyle=\footnotesize,        % the size of the fonts that are used for the code
+  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
+  breaklines=true,                 % sets automatic line breaking
+  captionpos=b,                    % sets the caption-position to bottom
+  commentstyle=\color{red},    % comment style
+  frame=single,	                   % adds a frame around the code
+  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
+  keywordstyle=\color{blue},       % keyword style
+  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
+  numbersep=5pt,                   % how far the line-numbers are from the code
+  numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
+  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
+  stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
+  stringstyle=\color{mauve},     % string literal style
+  tabsize=2,	                   % sets default tabsize to 2 spaces
+}
 
@@
 \begin{frame}[fragile]{Motivating Example}
-\todo{motivating example}
+\begin{itemize}
+	\item An analyst wants to explore relationships between polling data and survey data on household landline ownership
+	\item There is an issue with non-responses in survey data
+	\item They issue an ImputeDB query and get an effective response
+\end{itemize}
+
+\begin{figure}
+\begin{lstlisting}[language=SQL]
+SELECT polling.ST, AVG(acs.TEL)
+FROM polling, acs
+WHERE polling.ST = acs.ST
+  AND polling.ERROR > 50         -- 5 percentage points
+GROUP BY polling.ST;
+\end{lstlisting}
+\caption{A typical analyst query on ACS data}
+\label{fig:example-query}
+\end{figure}
+\end{frame}
+
+\begin{frame}[fragile]{Motivating Example: Plan}
+\begin{itemize}
+	\item ImputeDB adds an impute as necessary for the aggregation
+	\item It does not need to impute before then as \verb|acs.ST| is not affected by non-responses and thus has no missing values
+\end{itemize}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width=0.4\textwidth]{../paper/figures/example.png}
+    \caption{Query plan generated by ImputeDB prioritizing imputation quality.}
+    \label{fig:query-plan}
+\end{figure}
 \end{frame}
@@
 \begin{itemize}
-	\item Iterative algorithm using regression trees: \textit{Chained Equations Regression Trees} (CERT)\cite{burgette2010multiple}
+	\item Iterative algorithm using regression trees: \textit{Chained Equations Regression Trees} (CE-CERT)\cite{burgette2010multiple}
 	\item Significant cost to perform on entire dataset
@@
 \scriptsize
-\input{../paper/algorithms/cert.tex}
+\input{../paper/algorithms/cart.tex}
 \end{algorithm}
@@
 
+\begin{frame}[t,allowframebreaks]{References}
+  \nocite{*}
+  \printbibliography
+ \end{frame}
+
 \end{document}

### 2016-12-14 paper/algorithm.tex
paper: Add discussion of time complexity of tree build

@@
 
-\todo{is it necessary to add a fudge factor to the loss calculation, for generality, and set
-it at 1.0 for our paper?}
+To compute $\text{Time}(Q)$, we retain the heuristics used by SimpleDB for the time
+complexity of the typical relational operators, and add new formulations for \textit{Drop}
+and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
+complexity can be expressed as a function of the number of heap pages read and the IO cost
+per page.
+
+Evaluation of the time complexity for \textit{Impute} is more thorny, as the properties of
+different decision tree building algorithms can vary significantly, the underlying IO cost
+is not easily extracted (especially in the case that the entire stream of tuples does not
+fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
+not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
+time complexity of the \textit{build tree} algorithm for one commonly-used class of
+decision trees is a function of the number of classes, several overhead constants, the
+parameterization of the partition and heuristic functions, and the \textit{arity} (the
+number of subsets considered for each split). Indeed, the \textit{build tree} phase often
+does not even dominate computation, as post-processing steps like pruning, which are vital
+in achieving good performance, can have cost exponential in the height of the tree. In
+practice, we find that a simple parameterization can be acceptable (yield intelligent query
+plans), though the range of $\alpha$ that does promote tradeoffs is condensed.
 

### 2016-12-14 paper/experiments.tex
paper: add new generated tables + experiments changes

@@
     \caption{Error ($\epsilon$, planning time ($t_p$), and running time ($t_r$)i or queries
-    with different imputation levels. Base error is the RMSE between the query run on clean
-    data and the query run on dirty data without imputation. \todo{Micah: Not clear to me
-    what the symbol means.}}
+    with different imputation levels. Base error is the root-mean-square error (RMSE) between the query run on clean
+    data and the query run on dirty data without imputation.}
   \label{fig:experiments}
@@
 
-Planning times are not shown\todo{they are shown...}, as they are fractions of a second for all of the queries that we tested.
+Planning times are not shown, as they are fractions of a second for all of the queries that we tested.
 The overall low planning times highlight the practicality of our algorithm, despite the exponential complexity.
@@
 
-\todo{consider giving up on getting the references to work?}
 \begin{itemize}

### 2016-12-14 paper/queries.tex
paper: add new generated tables + experiments changes

@@
-\begin{tabular}{ll}
+\newcounter{queryno}
+\begin{tabular}{cl}
 \toprule
-\# & Query \\
+\# & \multicolumn{1}{c}{Query} \\
 \midrule
-0 & \verb|SELECT BLD FROM acs;| \label{q0} \\
-1 & \verb|SELECT BLD as units_in_structure, COUNT(ST) as estimate FROM acs GROUP BY BLD;| \label{q1} \\
-2 & \verb|SELECT BATH as has_bath, COUNT(ST) as ct FROM acs GROUP BY BATH;| \label{q2} \\
-3 & \verb|SELECT ACR as lotsize, AVG(BDSP) as avg_num_bedrooms FROM acs GROUP BY ACR;| \label{q3} \\
-4 & \verb|SELECT PSF as has_sub_families, SUM(NP) as num_people FROM acs GROUP BY PSF;| \label{q4} \\
-5 & \verb|SELECT AVG(NP) as avg_num_people FROM acs;| \label{q5} \\
-6 & \verb|SELECT BDSP as num_bedrooms, AVG(ACR) as avg_lot_size FROM acs WHERE VEH >= 2 GROUP BY BDSP;| \label{q6} \\
+1 & \verb|SELECT BLD as units_in_structure, COUNT(ST) as estimate FROM acs GROUP BY BLD;| \refstepcounter{queryno} \label{q1} \\
+2 & \verb|SELECT BATH as has_bath, COUNT(ST) as ct FROM acs GROUP BY BATH;| \refstepcounter{queryno} \label{q2} \\
+3 & \verb|SELECT ACR as lotsize, AVG(BDSP) as avg_num_bedrooms FROM acs GROUP BY ACR;| \refstepcounter{queryno} \label{q3} \\
+4 & \verb|SELECT PSF as has_sub_families, SUM(NP) as num_people FROM acs GROUP BY PSF;| \refstepcounter{queryno} \label{q4} \\
+5 & \verb|SELECT AVG(NP) as avg_num_people FROM acs;| \refstepcounter{queryno} \label{q5} \\
+6 & \verb|SELECT BDSP as num_bedrooms, AVG(ACR) as avg_lot_size FROM acs WHERE VEH >= 2 GROUP BY BDSP;| \refstepcounter{queryno} \label{q6} \\
+7 & \verb|SELECT MIN(YBL) as earliest_built_bucket FROM acs WHERE ACR = 3;| \refstepcounter{queryno} \label{q7} \\
+8 & \verb|SELECT MIN(RMSP) as min_num_rooms FROM acs WHERE RWAT=2;| \refstepcounter{queryno} \label{q8} \\
+9 & \verb|SELECT * FROM acs WHERE REFR = 1 AND STOV = 1 AND TEL = 1 AND TOIL = 2;| \refstepcounter{queryno} \label{q9} \\
+10 & \verb|SELECT * FROM dirty m where m.f1 >= 2;| \refstepcounter{queryno} \label{q10} \\
+11 & \verb|SELECT * FROM dirty where dirty.f1 <= 2;| \refstepcounter{queryno} \label{q11} \\
+12 & \verb|SELECT MAX(f2) as max_f2 FROM dirty GROUP BY dirty.f1;| \refstepcounter{queryno} \label{q12} \\
+13 & \verb|SELECT m1.f1, m1.f2 FROM dirty m1, clean m2 WHERE m1.f1 = m2.f1 and m1.f1 = 1 and m2.f2 = 69;| \refstepcounter{queryno} \label{q13} \\
+14 & \verb|SELECT d1.f1 FROM dirty d1, clean d2 WHERE d1.f1 = d2.f1 and d2.f2 = 50;| \refstepcounter{queryno} \label{q14} \\
 \bottomrule

### 2016-12-14 paper/conclusion.tex
paper: add imputation cost/loss to further work

--- 2016-12-08 paper/conclusion.tex
@@
 \subsection{Future Work}
-\ProjectName{} opens up multiple avenues for further work. For instance, we could extend our minimal/maximal imputation
-operators to consider global information, such as the specific columns needed in operators higher up in the query plan.
-Our optimizer uses the same machine learning algorithm in all instances of imputation operators. Integrating multiple
-possible algorithms could allow for further fine-grained imputation. Finally, the underlying database for \ProjectName{}
-is far from a full-featured, production database, so a natural step is to integrate imputation planning in a widely-adopted
-production-quality database. This will likely expose further opportunities for improvement.
+\ProjectName{} opens up multiple avenues for further work. For instance, we
+could extend our minimal/maximal imputation operators to consider global
+information, such as the specific columns needed in operators higher up in the
+query plan.  Our optimizer uses the same machine learning algorithm in all
+instances of imputation operators. Integrating multiple possible algorithms
+could allow for further fine-grained imputation, and honing the time complexity and
+loss of these algorithms for an iterator model database could facilitate more
+intelligent query plans and consistent interpretations of $\alpha$ across
+strategies. For example, algorithms that learn in an online manner could
+increase the efficiency of the system.  Furthermore, a multiple imputation
+strategy could be followed for queries involving certain aggregates. Finally,
+the underlying database for \ProjectName{} is far from a full-featured,
+production database, so a natural step is to integrate imputation planning in a
+widely-adopted production-quality database. This will likely expose further
+opportunities for improvement.
 

### 2016-12-14 simpledb/src/java/simpledb/DbIterator.java
fix: changes to run benchmarks

@@ -81,7 +81,7 @@ public interface DbIterator extends Serializable {
 			}
 			count++;
 		}
-		return Math.sqrt(err) / count;
+		return Math.sqrt(err / count);
 	}
 	
 	public default void print(PrintStream strm) throws DbException, TransactionAbortedException {

### 2016-12-14 simpledb/test/acstest/RunQueries.java

@@ -71,7 +71,7 @@ public class RunQueries {
 		} catch (BadErrorException e) {}
 		dirty.close();
 		
-		for (double a = 0.9; a <= 1.0; a += 0.01) {
+		for (double a = 0.0; a <= 1.0; a += 1.0) {
 			final double aa = a;
 			startPlan = Instant.now();
 			DbIterator imputedDirty = planQuery(dirtyQuery, x -> new ImputedLogicalPlan(aa));
@@ -89,8 +89,8 @@ public class RunQueries {
 				imputeErr = clean.error(imputedDirty);
 			} catch (BadErrorException e) {}
 			
-			outFile.write(String.format("\"%s\",%f,%f,%s,%s,%f\n", query, baseErr, imputeErr, 
-					Duration.between(startPlan, endPlan), Duration.between(startRun, endRun), a));
+			outFile.write(String.format("\"%s\",%f,%f,%d,%d,%f\n", query, baseErr, imputeErr,
+					Duration.between(startPlan, endPlan).getSeconds(), Duration.between(startRun, endRun).getSeconds(), a));
 			
 			imputedDirty.close();
 		}

### 2016-12-14 paper/algorithm.tex
paper: minor word changes

@@
 
-For each sub-query, using the histogram information, we keep track of the estimated number of
-null values in a column  ($\text{MissingCount}(c, Q)$), given the histograms associated with
+For each sub-query, we keep track of the estimated number of
+null values in a column  ($\text{MissingCount}(c, Q)$) by using the histograms associated with
 the sub-query. We use this estimate to compute
@@
 
-Evaluation of the time complexity for \textit{Impute} is more thorny, as the properties of
+Evaluation of the time complexity for \textit{Impute} is thornier, as the properties of
 different decision tree building algorithms can vary significantly, the underlying IO cost
@@
 in achieving good performance, can have cost exponential in the height of the tree. In
-practice, we find that a simple parameterization can be acceptable (yield intelligent query
+practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
 plans), though the range of $\alpha$ that does promote tradeoffs is condensed.
@@
 agnostic to the choice of imputation algorithm, and additional algorithms --- with
-associated time and loss functions --- could be insert without any change in design.
+associated time and loss functions --- could be inserted without any change in design.
 
@@
 as values progressively reflect more accurate relationships amongst attributes. The
-algorithm terminates when convergence is achieved (the imputed values do not change across
+algorithm terminates when convergence is achieved (i.e. the imputed values do not change across
 epochs) or a fixed number of epochs is reached.  Algorithm~\ref{algo:imputation-strategy}
@@
 manipulation. By removing concerns for imputation, users executing queries in \ProjectName{}
-don't need to consider the potential nature of the missing data.
+don't need to consider the potential nature of missing data.
 

### 2016-12-14 paper/conclusion.tex
paper: minor word changes

@@
 In our experiments, we considered a series of analytic queries (using relevant aggregates) and set queries (with simple projections) on real-world
-American Community Survey data
+American Community Survey data and synthetic data
 and showed that the chosen query plan improves accuracy of results relative to simply ignoring missing values. Furthermore,

### 2016-12-14 paper/examples.tex
paper: Add discussion of time complexity of tree build
paper: minor word changes

@@
 tradeoffs between efficiency and imputation quality. Finally, note that the optimizer doesn't
-place any imputation operator for the \verb|acs.ST| column, as optimizer uses the histogram to
+place any imputation operator for the \verb|acs.ST| column, as it uses a histogram to
 detect that there are no missing values there (as might be expected for a survey question that isn't
@@
     \includegraphics[width=0.6\textwidth]{figures/example.png}
-    \caption{Query plan generated by ImputeDB prioritizing imputation quality.}
+    \caption{Query plan generated by ImputeDB prioritizing imputation quality}
     \label{fig:query-plan}

### 2016-12-14 paper/experiments.tex
paper: minor word changes
paper: reworded explanation of experiments table

@@
 
-The queries on the ACS data all consist only of selections, filters, and aggregations.
+The queries on the ACS data all consist only of projections, selections, and aggregations.
 The ACS data is contained in a single table, so there are relatively few join queries that are interesting on this data set and remain relevant for analysts, rather
@@
   \input{runtimes}
-    \caption{Error ($\epsilon$, planning time ($t_p$), and running time ($t_r$)i or queries
+    \caption{Base error, change in error and and running time for queries
     with different imputation levels. Base error is the root-mean-square error (RMSE) between the query run on clean
-    data and the query run on dirty data without imputation.}
+    data and the query run on dirty data without imputation. Change in error is relative to the base error.}
   \label{fig:experiments}
@@
 When a query returns a relation of the same ``shape'' (same number of tuples and same schema) on both clean and dirty data, we can compute an error value.
-For our experiments, we report root-mean-square error (RMSE).
+For our experiments, we report root-mean-square error (RMSE) for the base error (i.e. the error in the dirty data without imputation with respect to the clean data) and the change
+in RMSE for the imputed plans.
 For imputations, we run queries using the two extreme values of $\alpha$: 0 and 1, representing a focus on runtime and on information loss, respectively.
-For each query, the table displays the error between the query run on clean data and the query run on dirty data with imputation and the running time in seconds.
+For each query, the table displays the base error, the change in error in both imputation scenarios, and the running time in seconds.
 

### 2016-12-14 simpledb/test/acstest/DirtyTest.java
fix: remove some queries to experiments on time

@@ -31,11 +31,11 @@ public class DirtyTest {
 			{ "SELECT PSF as has_sub_families, SUM(NP) as num_people FROM acs GROUP BY PSF;" },
 			{ "SELECT AVG(NP) as avg_num_people FROM acs;" },
 			{ "SELECT BDSP as num_bedrooms, AVG(ACR) as avg_lot_size FROM acs WHERE VEH >= 2 GROUP BY BDSP;" },
-			{ "SELECT MIN(YBL) as earliest_built_bucket FROM acs WHERE ACR = 3;" },
+			//{ "SELECT MIN(YBL) as earliest_built_bucket FROM acs WHERE ACR = 3;" },
 			{ "SELECT MIN(RMSP) as min_num_rooms FROM acs WHERE RWAT=2;" },
-			{ "SELECT * FROM acs WHERE REFR = 1 AND STOV = 1 AND TEL = 1 AND TOIL = 2;" },
+//			{ "SELECT * FROM acs WHERE REFR = 1 AND STOV = 1 AND TEL = 1 AND TOIL = 2;" },
 			// simpledb cannot handle predicates vs other columns (only relative to constants)
-			{ "SELECT * FROM acs WHERE VEH >= 1 AND VEH <= 5 AND RMSP > 4;" },
+			//{ "SELECT * FROM acs WHERE VEH >= 1 AND VEH <= 5 AND RMSP > 4;" },
 			// queries on synthetic data
 			{ "SELECT * FROM dirty m where m.f1 >= 2;" },
 			{ "SELECT * FROM dirty where dirty.f1 <= 2;" },

### 2016-12-14 simpledb/test/acstest/RunQueries.java

@@ -91,7 +91,7 @@ public class RunQueries {
 			
 			outFile.write(String.format("\"%s\",%f,%f,%d,%d,%f\n", query, baseErr, imputeErr,
 					Duration.between(startPlan, endPlan).getSeconds(), Duration.between(startRun, endRun).getSeconds(), a));
-			
+			outFile.flush();
 			imputedDirty.close();
 		}
 		clean.close();
@@ -134,5 +134,6 @@ public class RunQueries {
 		for (Object[] query : DirtyTest.data()) {
 			runQuery((String)query[0]);
 		}
+		outFile.close();
 	}
 }

### 2016-12-14 paper/queries.tex
paper: fix table generator & generate new tables

--- 2016-12-10 paper/queries.tex
@@
-\begin{tabular}{ll}
+\newcounter{queryno}
+\begin{tabular}{cl}
 \toprule
-\# & Query \\
+\# & \multicolumn{1}{c}{Query} \\
 \midrule
-0 & \verb|SELECT BLD FROM acs;| \label{q0} \\
-1 & \verb|SELECT BLD as units_in_structure, COUNT(ST) as estimate FROM acs GROUP BY BLD;| \label{q1} \\
-2 & \verb|SELECT BATH as has_bath, COUNT(ST) as ct FROM acs GROUP BY BATH;| \label{q2} \\
-3 & \verb|SELECT ACR as lotsize, AVG(BDSP) as avg_num_bedrooms FROM acs GROUP BY ACR;| \label{q3} \\
-4 & \verb|SELECT PSF as has_sub_families, SUM(NP) as num_people FROM acs GROUP BY PSF;| \label{q4} \\
-5 & \verb|SELECT AVG(NP) as avg_num_people FROM acs;| \label{q5} \\
-6 & \verb|SELECT BDSP as num_bedrooms, AVG(ACR) as avg_lot_size FROM acs WHERE VEH >= 2 GROUP BY BDSP;| \label{q6} \\
+1 & \verb|SELECT BLD FROM acs;| \refstepcounter{queryno} \label{q1} \\
+2 & \verb|SELECT BLD as units_in_structure, COUNT(ST) as estimate FROM acs GROUP BY BLD;| \refstepcounter{queryno} \label{q2} \\
+3 & \verb|SELECT BATH as has_bath, COUNT(ST) as ct FROM acs GROUP BY BATH;| \refstepcounter{queryno} \label{q3} \\
+4 & \verb|SELECT ACR as lotsize, AVG(BDSP) as avg_num_bedrooms FROM acs GROUP BY ACR;| \refstepcounter{queryno} \label{q4} \\
+5 & \verb|SELECT PSF as has_sub_families, SUM(NP) as num_people FROM acs GROUP BY PSF;| \refstepcounter{queryno} \label{q5} \\
+6 & \verb|SELECT AVG(NP) as avg_num_people FROM acs;| \refstepcounter{queryno} \label{q6} \\
+7 & \verb|SELECT BDSP as num_bedrooms, AVG(ACR) as avg_lot_size FROM acs WHERE VEH >= 2 GROUP BY BDSP;| \refstepcounter{queryno} \label{q7} \\
+8 & \verb|SELECT MIN(RMSP) as min_num_rooms FROM acs WHERE RWAT=2;| \refstepcounter{queryno} \label{q8} \\
+9 & \verb|SELECT * FROM dirty m where m.f1 >= 2;| \refstepcounter{queryno} \label{q9} \\
+10 & \verb|SELECT * FROM dirty where dirty.f1 <= 2;| \refstepcounter{queryno} \label{q10} \\
+11 & \verb|SELECT MAX(f2) as max_f2 FROM dirty GROUP BY dirty.f1;| \refstepcounter{queryno} \label{q11} \\
+12 & \verb|SELECT m1.f1, m1.f2 FROM dirty m1, clean m2 WHERE m1.f1 = m2.f1 and m1.f1 = 1 and m2.f2 = 69;| \refstepcounter{queryno} \label{q12} \\
+13 & \verb|SELECT d1.f1 FROM dirty d1, clean d2 WHERE d1.f1 = d2.f1 and d2.f2 = 50;| \refstepcounter{queryno} \label{q13} \\
 \bottomrule

### 2016-12-14 paper/gen_tables.py

@@ -4,7 +4,7 @@ import pandas as pd
 import sys
 
 def parse_time(t):
-    return t[2:][:-1]
+    return t
 
 def err_to_str(err):
     if err == 'null':

### 2016-12-14 paper/experiments.tex
paper: reworded experiment references with new tables

@@
 \begin{itemize}
-\item For sum and count queries (\ref{q1}, \ref{q2}, \ref{q4}), imputation does a good job of reducing the error.
+\item For sum and count queries (\ref{q2}, \ref{q3}, \ref{q5}), imputation does a good job of reducing the error.
 For each of these queries, we can reduce total error by several orders of magnitude.
@@
 
-\item For average queries (\ref{q3}, \ref{q5}, \ref{q6}), imputation does not reduce the error, mainly because running these queries on dirty data incurs only a small penalty.
+\item For average queries (\ref{q4}, \ref{q6}, \ref{q7}), imputation does not reduce the error, mainly because running these queries on dirty data incurs only a small penalty.
 This is because, in the case of average, removing data uniformly at random does not change the mean.
@@
 
-\item For min and max queries (\ref{q7}, \ref{q8}, \ref{q12}), it is unlikely that imputation will ever help, because by definition these queries are looking for extreme values.
+\item For a min query (\ref{q8}), it is unlikely that imputation will ever help, because by definition these queries are looking for extreme values.
 These values are unlikely to be the result of imputation and are also unlikely to be randomly removed from the dataset.
 
-\item We do not compute error for selection queries (\ref{q9}, \ref{q10}, \ref{q11}), because the output of the query on clean data is not directly comparable to the output on dirty data.
+\item We do not compute error for selection queries (\ref{q1}, \ref{q9}, \ref{q10}, \ref{q12}), because the output of the query on clean data is not directly comparable to the output on dirty data.
 
-  \item Imputation performs poorly on the join queries on synthetic data (\ref{q13}, \ref{q14}) because the data is synthetic, so there are no correlations for imputation to find.
+  \item Imputation performs poorly on the join queries on synthetic data (\ref{q11}, \ref{q13}) because the data is synthetic, so there are no correlations for imputation to find.
 We included these queries because we wanted to test the performance of our algorithm on queries that include joins, and we were pleased to find that simple joins cause no perceptible performance problems.
@@
 We were unable to complete the full imputation due to resource constraints, but the process ran for more than one hour without terminating.
-This time compares favorably to our maximum of \todo{add time} for a query with imputation.
+This time compares favorably to our maximum of 55 seconds for a query with imputation on the ACS data.
 Given the varying needs of database users, \ProjectName{} clearly allows flexibility without a significant performance hit.

### 2016-12-15 paper/experiments.tex
paper: switch to percentage change in error

@@
   \input{runtimes}
-    \caption{Base error, change in error and and running time for queries
+    \caption{Base error, percent change in error and and running time for queries
     with different imputation levels. Base error is the root-mean-square error (RMSE) between the query run on clean
@@
 We were unable to complete the full imputation due to resource constraints, but the process ran for more than one hour without terminating.
-This time compares favorably to our maximum of 55 seconds for a query with imputation on the ACS data.
+This time compares favorably to our maximum of 72 seconds for a query with imputation on the ACS data.
 Given the varying needs of database users, \ProjectName{} clearly allows flexibility without a significant performance hit.

### 2016-12-15 paper/gen_tables.py

@@ -12,11 +12,13 @@ def err_to_str(err):
     else:
         return '%.2e' % float(err)
 
-def err_diff_to_str(err1, err2):
-    if err1 == 'null' or err2 == 'null':
+def err_diff_to_str(err, base_err):
+    if err == 'null' or base_err == 'null':
         return r'\multicolumn{1}{c}{--}'
+    if float(base_err) == 0.0:
+        return '0'
     else:
-        return '%.2e' % (float(err1) - float(err2))
+        return str(round(((float(err) - float(base_err)) / abs(float(base_err)) * 100)))
 
 def get_first_row(df, l):
     for (_, row) in df.iterrows():
@@ -56,7 +58,7 @@ def main():
         writel(f, r'\multicolumn{2}{c}{} & \multicolumn{2}{c}{Imputed ($\alpha=0.0$)} & \multicolumn{2}{c}{Imputed ($\alpha=1.0$)} \\')
         writel(f, r'\cmidrule(r){3-4}')
         writel(f, r'\cmidrule(l){5-6}')
-        writel(f, r'\# & \multicolumn{1}{c}{Base error} & \multicolumn{1}{c}{Error} & \multicolumn{1}{c}{Time (s)} & \multicolumn{1}{c}{Error} & \multicolumn{1}{c}{Time (s)} \\')
+        writel(f, r'\# & \multicolumn{1}{c}{Base error} & \multicolumn{1}{c}{Error \%$\Delta$} & \multicolumn{1}{c}{Time (s)} & \multicolumn{1}{c}{Error \%$\Delta$} & \multicolumn{1}{c}{Time (s)} \\')
         writel(f, r'\midrule')
         for query in df['query'].unique():
             ctr = qidx[query]

### 2017-02-07 paper/algorithm.tex
paper: rewrite algorithm section for clarity

@@
 To reduce the size of the query plan search space, only plans that fit the following template are considered.
-First, all filters are pushed to the leaves of the query tree, immediately after the scans.
-Joins are performed after filtering, and only left-deep plans are considered.
-Any group-by/aggregate will be performed after the final join.
-Finally, projections are placed at the root of the query tree.
+
+\begin{itemize}
+\item All filters are pushed to the leaves of the query tree, immediately after scanning a table.
+We use a pre-processing step which conjoins together filters which operate on the same table, so we can assume that each table has at most one relevant filter.
+
+\item Joins are performed after filtering, and only left-deep plans are considered.
+
+\item A grouping \& aggregation operator is optional, and if present will be performed after the final join.
+
+\item Projections are placed at the root of the query tree.
+\end{itemize}
+
 The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
@@
 
-In order to correctly estimate $\text{Time}(Q)$ and $\text{Loss}(Q)$, the system must have cardinality estimates
-for each sub-query in each query plan.
+In order to correctly estimate $\text{Time}(Q)$ and $\text{Loss}(Q)$, the system must have cardinality estimates for each sub-query in each query plan.
 These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
@@
 Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required.
-To track these cases, each query plan is associated with a set of dirty attributes $D$.
-An attribute $c$ is \emph{dirty} in some relation if the values for $c$ may contain null.
-We compute a dirty set for each base table using the table statistics, which track the number of null values in each column.
-If we apply an imputation operator to a relation $R$ with a dirty set $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty set $D' = D \setminus C$.
-Applying a projection $\pi_C(R)$ produces a dirty set $D' = D \cap C$.
-A join $R_1 \Join_\psi R_2$ with dirty sets $D_1$ and $D_2$  produces a dirty set $D' = D_1 \cup D_2$.
-Filters do not change the dirty set.
+To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$.
+An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null.
+We compute a dirty set for each table using the table statistics, which track the number of null values in each column.
+The dirty set for a query plan can be computed recursively as follows:
 
-That is, the dirty set over-approximates the set of attributes that contain null.
-For example, a filter might remove all tuples which contain null without changing the dirty set, forcing an unnecessary imputation.
-We choose to over-approximate the dirty set to avoid the possibility of dropping a tuple that contains a null value without explicitly imputing the value or applying a drop operator.
+\begin{align*}
+  \textsc{Dirty}(q) = \begin{cases}
+    \textsc{Dirty}(q') \setminus C & q = \delta_C(q')\ \text{or}\ q = \mu_C(q') \\
+    \textsc{Dirty}(q') \cap C & q = \pi_C(q') \\
+    \textsc{Dirty}(q_l) \cup \textsc{Dirty}(q_r) & q = q_l \bowtie_\psi q_r \\
+    \textsc{Dirty}(q') & q = \sigma_\phi(q') \\
+    \textsc{Dirty}(t) & q = \text{some table}\ t \\
+  \end{cases}
+\end{align*}
+
+Note that \textsc{Dirty} over-approximates the set of attributes that contain null.
+For example, a filter might remove all tuples which contain null, but the dirty set would be unchanged.
+We choose to over-approximate to ensure that all null values are explicitly imputed or dropped.
 

### 2017-02-07 paper/algorithms/plan-helpers.tex
paper: rewrite algorithm section for clarity

@@
   \begin{algorithmic}
-    \Require{$q$ is a query plan, $C_{req}$ is a set of attributes that must be imputed.}
-    \Ensure{Returns a set of query plans such that $\Call{Dirty}{q'} \cap C_{req} = \emptyset$.}
-    \Function{AddImpute}{$q, C_{req}$}
-    \State $C_{min} \gets \Call{Dirty}{q} \cap C_{req}$
-    \If{$\Call{Dirty}{q} = \emptyset$}
-    	\State \Return $\{ q \}$
-   \ElsIf{$C_{min} = \emptyset$}
-    	\State \Return $\{ \mu_{\Call{Dirty}{q}}(q), q \}$
+    \Require{$q$ is a query plan.
+$C_{req}$ is the set of attributes that must be imputed in the output of this query plan.
+$C_{vis}$ is the set of attributes which are visible in the output of the overall query.}
+    \Ensure{A set of query plans $Q$ with added imputation.}
+    \Function{AddImpute}{$q, C_{req}, C_{vis}$}
+    \State $D_{must} \gets \Call{Dirty}{q} \cap C_{req}$
+    \State $D_{may} \gets D_{must} \cup (\Call{Dirty}{q} \cap C_{vis})$
+    \State $Q \gets \emptyset$
+    \If{$D_{must} = \emptyset$}
+    \State $Q \gets Q \cup \{q\}$
     \Else
-    \State \Return $\{\mu_{\Call{Dirty}{q}}(q), \mu_{C_{min}}(q), \delta_{C_{min}}(q)\}$
+    \State $Q \gets Q \cup \{\mu_{D_{must}}(q),\ \delta_{D_{must}}(q)\}$
     \EndIf
-    \EndFunction
 
-    \State
+    \If{$D_{may} \neq \emptyset$}
+    \State $Q \gets Q \cup \{\mu_{D_{may}}(q)\}$
+    \EndIf
 
-    \Require{$Q$ is a set of query plans.}
-    \Ensure{Returns an optimal query plan for each distinct dirty set in $Q$.}
-    \Function{OptRel}{$Q$}
-    \State $D \gets \{\Call{Dirty}{q} ~|~ q \in Q\}$
-    \State \Return $\{\argmin_{q \in Q \land \Call{Dirty}{q} = d} \Call{Cost}{q} ~|~ d \in D\}$
+    \State \Return $Q$
     \EndFunction
@@
     \Require{$t$ is a table and $\phi$ is a filter predicate.}
-    \Ensure{Returns a set of optimal query plans for scanning and filtering $t$, with distinct dirty sets.}
-    \Function{OptFilter}{$t, \phi$}
-    \State \Return $\Call{OptRel}{\{  \sigma_{\phi}(q) | q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}} \} }$
+    \Ensure{A set of plans for filtering $t$ with $\phi$.}
+    \Function{OptFilter}{$t, \phi, C_{vis}$}
+    \State \Return $\{\sigma_{\phi}(q)\ |\ q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}, C_{vis}}\}$
     \EndFunction
@@
 
-    \Require{$Q$ is a map from tables and dirty sets to optimal query plans (possibly with imputations), and $\Psi$ relates query plans with join predicates.}
-    \Ensure{Returns a map from dirty sets to optimal query plans involving all necessary joins}
-    \Function{OptJoin}{$Q, \Psi$}
-    \For{$size \in 1...|\Psi|$}
-    	\State $S \gets subset(\Psi, size)$
-	\For{$\psi \in S$}
-		\State $S' \gets S \setminus \psi$
-			\If{$joins(S', \psi)$} 
-				\State $S'_{plans} \gets Q(\Call{Rels}{S'})$
-				\State $t \gets \Call{Rels}{\psi} \setminus \Call{Rels}{S'}$
-				\State $t_{plans} \gets Q(t)$
-				\For {$l, r \in S'_{plans} \times t_{plans}$}
-					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\psi}}$
-					\State $P_r \gets \Call{AddImpute}{r, \Call{Attrs}{\psi}}$
-					\State \Call{Update}{Q, $\Call{OptRel}{ \{p_l \bowtie_{\psi} p_r | p_l \in P_l, p_r \in P_r \} }$}
-				\EndFor
-		\EndIf
+    \Require{$q$ is a query plan, $G$ is a set of grouping attributes, $A$ is an aggregation function, and $C_{vis}$ is a set of visible attributes.}
+    \Ensure{A set of plans for grouping \& aggregating with $G$ and $A$.}
+    \Function{OptGroupBy}{$q, G, A, C_{vis}$}
+    \State $C \gets G \cup \Call{Attrs}{A}$
+    \State $Q' \gets \Call{AddImpute}{q, C, C_{vis}}$
+    \State \Return $\{\Call{GroupBy}{q', G, A} ~|~ q' \in Q'\}$
+    \EndFunction
+
+    \State
+
+    \Require{$q$ is a query plan, $P$ is a set of projection attributes.}
+    \Ensure{A set of plans with the attributes in $P$ projected.}
+    \Function{OptProject}{$q, P$}
+    \State \Return $\{\pi_P(q') ~|~ q' \in \Call{AddImpute}{q, P, P}\}$
+    \EndFunction
+
+    \State
+
+    \Require{A set of tables $T$, a plan cache $Q$ containing at least one query plan for each table in $T$, and a relation $J : T \times \Psi \times T$ between tables and join predicates.}
+    \Ensure{An updated plan cache $Q'$ which contains at least one plan that combines all of the tables in $T$.}
+    \Function{OptJoin}{$Q, T, J, C_{vis}$}
+    \For{$size \in 2...|T|$}
+    \For{$S \in \{\text{all length}\ size\ \text{subsets of}\ T\}$}
+    \For{$t \in S$}
+    \State $S' \gets S \setminus \{t\}$
+    \For{$(t, \psi, t') \in J$ where $t' \in S'$}
+    \State $C \gets \Call{Attrs}{\psi}$
+    \State $L \gets \{\Call{AddImpute}{q, C, C_{vis}} ~|~ q \in Q[t] \}$
+    \State $R \gets \{\Call{AddImpute}{q, C, C_{vis}} ~|~ q \in Q[S'] \}$
+    \State $Q[S] \lhd \{l \bowtie_\psi r ~|~ l \in L, r \in R\}$
     \EndFor
     \EndFor
-    \State \Return{Q(\Call{Rels}{$\Phi$})}
+    \EndFor
+    \EndFor
     \EndFunction
@@
 %\end{algorithm}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "../main"
+%%% End:

### 2017-02-07 paper/algorithms/plan-toplevel.tex
paper fix: extract algorithm begin/end markers to be able to use with [H] placement in slides
paper: rewrite algorithm section for clarity

@@
-\begin{algorithm}
+% \begin{algorithm}
 \begin{algorithmic}
-   \Require{set of tables $T$, filter predicates $\Phi$, join predicates $\Psi$, a set of projection attributes $P$, optional set of grouping attributes $G$ and aggregator function $A$}
-   \Ensure{Returns optimized query plan.}
+  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$, and an optional set of grouping attributes $G$ and aggregator function $A$.}
+  \Ensure{An optimized query plan.}
 
-  \Function{Plan}{$T, \Phi, \Psi, P, G, A$}
- \State  $Q \gets \{ \}$ \Comment{map from table and dirty set to plans }
-   \For{$t \in T$}
-   	\State $\phi \gets \Call{GetFilters}{t, \Phi}$
-	\State $P_{t} \gets \Call{OptFilter}{t, \phi}$
-	\State $Q \gets Q \cup \{ (t, \Call{Dirty}{q}) \mapsto q | q \in P_t \}$
+  \Function{Plan}{$T, F, J, P, G, A$}
+  \State Let $Q$ be an empty plan cache.
+  \For{$t \in T$} \Comment{Add selections to the plan cache.}
+  \If {$\exists \phi: (t, \phi) \in F$}
+  \State $Q[\{t\}] \lhd \Call{OptFilter}{Q, t, \phi, P}$
+  \Else
+  \State $Q[\{t\}] \lhd \{t\}$
+  \EndIf
   \EndFor
 
-   \State $M \gets \{ \}$ \Comment{map from dirty set to plans, contains all joins necessary}
-   \State $M \gets  \Call{OptJoin}{Q, \Psi}$	\Comment{optimize joins}
-   \If{$G \neq \emptyset \land A \neq NULL$}
-   	\State $attrs \gets \Call{Attrs}{G} \cup \Call{Attrs}{A}$
-   	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \Call{GroupBy}{\Call{AddImpute}{q, attrs}, G, A}}$
-   \Else
-  	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \pi_{P}(\Call{AddImpute}{q, P}) }$
+  \State $\Call{OptJoin}{Q, T, J, P}$	\Comment{Optimize joins.}
+  \State $B \gets Q[T]$ \Comment{Get the best plans for all tables.}
+  \If{$G \neq \emptyset$} \Comment{Add optional group \& aggregate.}
+  \State $B \gets \bigcup_{q \in B} \Call{OptGroupBy}{q, G, A, C_{vis}}$
   \EndIf
 
-\Return $\argmin_{q \in M} Cost(q)$
+  \State $B \gets \bigcup_{q \in B} \Call{OptProject}{q, P}$ \Comment{Add projections.}
+
+  \State \Return $\argmin_{q \in B} Cost(q)$ \Comment{Return cheapest plan.}
   \EndFunction
@@
 \label{algo:top-level-planner}
-\end{algorithm}
+% \end{algorithm}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "../main"
+%%% End:

### 2017-02-07 paper/algorithm.tex
paper: formatting

@@
 \label{sec:cost-model}
-We rank query plans $Q$ using a cost model $\text{Cost}(Q) = (1 - \alpha) \times \text{Time}(Q) + \alpha \times \text{Loss}(Q)$.
-$\text{Time}(Q)$ is an estimate of the runtime of the query which is derived from table statistics and selectivity estimation of the query predicates.
-$\text{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure.
-$\alpha$ is a parameter to the query optimizer that controls the emphasis on quality over performance.
-$\alpha = 1.0$ means that the query should be as accurate as possible (at the expense of performance), $\alpha=0.0$ means that the query should be as fast
+We rank query plans using a cost model \[\textsc{Cost}(Q) = (1 - \alpha) \times \textsc{Time}(Q) + \alpha \times \textsc{Loss}(Q).\] $\textsc{Time}(Q)$ is an estimate of the runtime of the query derived from table statistics and selectivity estimation of the query predicates.
+$\textsc{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure.
+$\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance.
+$\alpha = 1.0$ means that the optimal query should be as accurate as possible (at the expense of performance), $\alpha=0.0$ means that the optimal query should be as fast
 as possible (at the expense of accuracy).
 
-In order to correctly estimate $\text{Time}(Q)$ and $\text{Loss}(Q)$, the system must have cardinality estimates for each sub-query in each query plan.
+In order to correctly estimate $\textsc{Time}(Q)$ and $\textsc{Loss}(Q)$, the system must have cardinality estimates for each sub-query in each query plan.
 These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
@@
 For each sub-query, we keep track of the estimated number of
-null values in a column  ($\text{MissingCount}(c, Q)$) by using the histograms associated with
+null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
 the sub-query. We use this estimate to compute
@@
 \begin{align*}
-  \text{Loss}(Q) = \begin{cases}
-    \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \delta_C(Q') \\
-    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \mu_C(Q') \\
-    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
-    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
+  \textsc{Loss}(q) = \begin{cases}
+    \sum_{c \in C} \textsc{Missing}(c, q') & q = \delta_C(q') \\
+    \frac{1}{\sqrt{|q'|}} \sum_{c \in C} \textsc{Missing}(c, q') & q = \mu_C(q') \\
+    \textsc{Loss}(q_1') + \textsc{Loss}(q_2') & q = q_1' \bowtie_\psi q_2' \\
+    \textsc{Loss}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
   \end{cases}

### 2017-02-07 paper/algorithm.tex
removed references to SimpleDB, clarified some stuff, and added TODOs

@@
 The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
+\todo{add a schematic of a standard query plan}
 
@@
 When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
-Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
-
-Our runtime cost model is the cost model used in SimpleDB, though modern systems
-have more sophisticated approaches.
+Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans
 
@@
 
-To compute $\text{Time}(Q)$, we retain the heuristics used by SimpleDB for the time
-complexity of the typical relational operators, and add new formulations for \textit{Drop}
-and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
+To compute $\text{Time}(Q)$, we retain a simple set of heuristics used by a standard 
+database for the time. Scan costs are estimated based on the number of tuples, 
+page size, and an unit-less IO cost per page. Join costs are estimated as a function
+of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
+Filtering, projecting, and aggregating are all done in memory and have low computational overhead
+so we assume negligible time costs for those operations. We extend these heuristics to 
+account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
 complexity can be expressed as a function of the number of heap pages read and the IO cost
@@
 \subsection{Complexity}
+\todo{this section needs to be cleaned up/clarified}
 Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 

### 2017-02-07 paper/examples.tex
paper: formatting
removed references to SimpleDB, clarified some stuff, and added TODOs

--- 2016-12-14 paper/examples.tex
@@
 
-Consider the case of an analyst looking to explore the sources and causes of
-polling error in a presidential election. One hypothesis \cite{zukin2015s}
-is that the rapid switch from landlines to cellphones increases the
-difficulty of contacting potential poll respondents. The analyst may
-begin her analysis by comparing the polling error in a state with the
-distribution of households with landlines, for the set of states for which
-polling error was most pronounced. In order to probe deeply, the analyst
-accesses a household-level survey that includes data on
-whether the household has a landline. One such survey --- which will be
-discussed further (Sec~\ref{subsec:datasets}) --- is the American Community Survey (ACS),
-conducted by the US Census Bureau.
-
-Surveys are rife with missing values due to non-response and other issues, and
-the ACS is no exception. The Census Bureau and other similar organization attempt to present
-data in as unadulterated a fashion as possible. The ACS dataset is not amenable to one-time
-imputation by either the Census Bureau, for this reason, or the analyst's organization,
-which likely uses the same database for a variety of unrelated uses (asking for different
-imputation strategies). Thus, the analyst must deal with missing values in one way or
-another.
-
-In this case, a non-negligible fraction of respondents omitted information on whether they
-have a landline. The analyst hypothesizes that older houses are more likely to have
-landlines and that households that are relatively new are less likely to have landlines. 
-The analyst could take advantage of these presumed correlations to impute the missing values
-on the base table in one expensive operation. On the other hand, the query could be run
-directly and rows with missing values may be dropped entirely. With ImputeDB, the analyst
-could impute the relevant subset of data on-the-fly, achieving better performance without
-losing the information in the dropped examples.
-
-The analyst submits a query (\todo{ref to listing}) to ImputeDB which finds an optimized query
-plan\footnote{The analyst specifies a relatively high value of $\alpha$, the quality
-emphasis parameter, in order to receive a plan that does an expensive imputation step. See
-Section~\ref{sec:cost-model} for more details.}. The resulting query plan
-(Figure~\ref{fig:query-plan})
-imputes missing values for \verb!acs.TEL! after states with high polling error have been
-selected. This decision reduces the set of tuples that must be input into the imputation
-algorithm to just those that are relevant to the query. Later, we show the algorithms that
-the ImputeDB query optimizer uses in order to detect these opportunities and control the
-tradeoffs between efficiency and imputation quality.
-
-\begin{verbatim}
-SELECT polling.ST, AVG(acs.TEL)
-FROM polling, acs
-WHERE polling.ST = acs.ST
-  AND polling.ERROR > 50         -- 5 percentage points
-GROUP BY polling.ST;
-\end{verbatim}
-
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[width=0.6\textwidth]{figures/example.png}
-    \caption{Query plan generated by ImputeDB prioritizing imputation quality.}
-    \label{fig:query-plan}
-\end{figure}
+\todo{Replace example with a relevant example and plan to the new data used for experiments}
+%Consider the case of an analyst looking to explore the sources and causes of
+%polling error in a presidential election. One hypothesis \cite{zukin2015s}
+%is that the rapid switch from landlines to cellphones increases the
+%difficulty of contacting potential poll respondents. The analyst may
+%begin her analysis by comparing the polling error in a state with the
+%distribution of households with landlines, for the set of states for which
+%polling error was most pronounced. In order to probe deeply, the analyst
+%accesses a household-level survey that includes data on
+%whether the household has a landline. One such survey --- which will be
+%discussed further (Sec~\ref{subsec:datasets}) --- is the American Community Survey (ACS),
+%conducted by the US Census Bureau.
+%
+%Surveys are rife with missing values due to non-response and other issues, and
+%the ACS is no exception. The Census Bureau and other similar organization attempt to present
+%data in as unadulterated a fashion as possible. The ACS dataset is not amenable to one-time
+%imputation by either the Census Bureau, for this reason, or the analyst's organization,
+%which likely uses the same database for a variety of unrelated uses (asking for different
+%imputation strategies). Thus, the analyst must deal with missing values in one way or
+%another.
+%
+%In this case, a non-negligible fraction of respondents omitted information on whether they
+%have a landline. The analyst hypothesizes that older houses are more likely to have
+%landlines and that households that are relatively new are less likely to have landlines. 
+%The analyst could take advantage of these presumed correlations to impute the missing values
+%on the base table in one expensive operation. On the other hand, the query could be run
+%directly and rows with missing values may be dropped entirely. With ImputeDB, the analyst
+%could impute the relevant subset of data on-the-fly, achieving better performance without
+%losing the information in the dropped examples.
+%
+%The analyst submits a query (Figure~\ref{fig:example-query}) to ImputeDB which finds an optimized query
+%plan\footnote{The analyst specifies a relatively high value of $\alpha$, the quality
+%emphasis parameter, in order to receive a plan that does an expensive imputation step. See
+%Section~\ref{sec:cost-model} for more details.}. The resulting query plan
+%(Figure~\ref{fig:query-plan})
+%imputes missing values for \verb!acs.TEL! after states with high polling error have been
+%selected. This decision reduces the set of tuples that must be input into the imputation
+%algorithm to just those that are relevant to the query. Later, we show the algorithms that
+%the ImputeDB query optimizer uses in order to detect these opportunities and control the
+%tradeoffs between efficiency and imputation quality. Finally, note that the optimizer doesn't
+%place any imputation operator for the \verb|acs.ST| column, as it uses a histogram to
+%detect that there are no missing values there (as might be expected for a survey question that isn't
+%impacted by non-responses).
+%
+%\begin{figure}
+%\begin{lstlisting}[language=SQL]
+%SELECT polling.ST, AVG(acs.TEL)
+%FROM polling, acs
+%WHERE polling.ST = acs.ST
+%  AND polling.ERROR > 50         -- 5 percentage points
+%GROUP BY polling.ST;
+%\end{lstlisting}
+%\caption{A typical analyst query on ACS data}
+%\label{fig:example-query}
+%\end{figure}
+%
+%\begin{figure}[!ht]
+%    \centering
+%    \includegraphics[width=0.6\columnwidth]{figures/example.png}
+%    \caption{Query plan generated by ImputeDB prioritizing imputation quality}
+%    \label{fig:query-plan}
+%\end{figure}
 

### 2017-02-07 paper/main.tex
paper: make paper shorter
paper: conform to vldb style
paper: formatting
removed references to SimpleDB, clarified some stuff, and added TODOs

@@
-\documentclass{article}
+\documentclass{vldb}
+\usepackage{balance}
 \usepackage[utf8]{inputenc}
@@
 \usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
+\usepackage[subtle]{savetrees}
 \usepackage{amsfonts,amssymb,amsmath}
@@
 \usepackage[noend]{algpseudocode}
-\usepackage[style=ieee]{biblatex}
+\usepackage[style=numeric,firstinits=true]{biblatex}
 \usepackage{booktabs}
@@
 \addbibresource{references.bib}
+
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
 
@@
 \author{
-  Jose Cambronero \\
+  Jos\'e Cambronero \\
+  \texttt{jcamsan@mit.edu}
   \and
-  John K. Feser
+  John K. Feser \\
+  \texttt{feser@csail.mit.edu}
   \and
-  Micah Smith}
+  Micah Smith \\
+  \texttt{micahs@mit.edu}}
 
@@
 \begin{abstract}
-In order to study the placement of an imputation step, we create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process in SimpleDB.
+In order to study the placement of an imputation step, we create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process into a simple database prototype.
 We introduce measures for information loss and runtime for imputation operations, which outline the main trade-offs in the imputation placement.
@@
 
+\balance
 \printbibliography

### 2017-02-07 paper/relatedwork.tex
removed references to SimpleDB, clarified some stuff, and added TODOs
paper: added timeseries forecasting related work

@@
 
+\subsection{Forecasting and Databases}
+\cite{parisi2011embedding} introduces the idea of incorporating time-series forecast operators into
+databases, along with the necessary relational algebra extensions. Their work explores the theoretical
+properties of forecast operators and generalizes them into a family of operators, distinguished by
+the type of predictions returned. They highlight the use of forecasting for replacing missing values.
+In their follow on work\cite{parisi2013temporal}, they identify various equivalence and containment
+relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
+explore forecast-first and forecast-last plans, which perform forecasting operations before and after traditional
+relational operators, respectively.
+
+In contrast to this work, \ProjectName{} is targeted at generic value imputation, not necessarily tailored to 
+time-series. The optimizer is not based on equivalence transformations, nor are there guarantees of equal
+results under different conditions. Instead, the optimizer allows users to pick their tradeoff between
+runtime cost and information loss. The search space considered by our optimizer is broader, not just
+forecast-first/forecast-last plans, but rather imputation operators can be placed anywhere in the query plan
+(with some restrictions). The novelty of our contribution lies in the successful incorporation of
+imputation operations in non-trivial query plans with cost-based optimization.
+
+\cite{duan2007processing} describes the Fa system and its declarative language for time-series forecasting. Their
+system automatically searches the space of attribute combinations/transformations and statistical models
+to produce forecasts within a given accuracy threshold. Accuracy estimates are determined using
+standard techniques, such as k-fold cross-validation.
+
+\ProjectName{} provides a declarative language, as
+an extension to standard SQL, but in contrast to Fa, \ProjectName{} is not searching the space of possible
+imputation models, but rather the space of query plans that incorporate imputation operators. Another major point
+of distinction between \ProjectName{} and Fa is that the latter doesn't consider tradeoffs between accuracy and computation time, 
+but rather returns the most accurate forecast (with some stopping criterion).
+
 %%% Local Variables:

### 2017-02-09 simpledb/test/acstest/DirtyTest.java
chore: ignore acs tests

@@ -19,6 +19,7 @@ import Zql.ZStatement;
 import Zql.ZqlParser;
 import simpledb.*;
 
+@Ignore
 @RunWith(Parameterized.class)
 public class DirtyTest {
 	@Parameters(name = "{index}: {0}")

### 2017-02-09 paper/algorithm.tex
paper: added query schematic for basic plans
paper: keep refining algorithm section

--- 2017-02-07 paper/algorithm.tex
@@
 \subsection{Query planning}
-The input to our query planner is a tuple $(T, \Phi, \Psi, P, G, A)$: a set of tables $T$, a set of filter predicates $\phi_t, t \in T$, a set of join predicates $\psi_{(t_1, t_2)}, t_1, t_2 \in T$, a set of attributes $P$, and an optional set of attributes $G$ and aggregation function $A \in \{\text{Max}, \text{Min}, \text{Sum}, \text{Avg}, \text{Count}\}$.
+The input to our query planner is a tuple $(T, F, J, P, G, A)$:
+\begin{itemize}
+\item A set of tables $T$.
+\item A relation $F: T \times \Phi$ between tables and filter predicates.
+\item A relation $J : T \times \Psi \times T$ between tables and join predicates.
+\item A set of projection attributes $P$.
+\item An optional set of grouping attributes $G$ and aggregator function $A$.
+\end{itemize}
 
 The query planner must select a join ordering in addition to placing imputation operators as described in Section~\ref{sec:placement}.
+
+Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
+At a high level, a plan cache is a mapping from a set of tables to a set of optimal plans that perform the required joins and filters over these tables.
+We store multiple plans rather than a single best plan because each plan has a distinct set of dirty attributes.
+
+\begin{figure}
+  \begin{align*}
+    \llbracket Q[T] \rrbracket &= \begin{cases}
+      P & (T, P) \in Q \\
+      \emptyset & \text{otherwise} \\
+    \end{cases} \\
+    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned}
+                                           Q \cup \{(T,\ \argmin_{q \in P}\ \textsc{Cost}(q))\}\ \text{where} \\
+                                           P = \left\{q ~|~ \begin{aligned}
+                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
+                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p) 
+                                                                   \end{aligned}
+                                           \right\}
+                                         \end{aligned}
+    % \in 
+  \end{align*}
+  \caption{Semantics of the plan cache.}
+\end{figure}
+
+The query planner (Algorithm~\ref{algo:top-level-planner}), operates as follows.
+First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan.
+This set will be used to determine which attributes are imputed.
+Then, it constructs a plan cache
 
@@
 drops) the columns required by the relational operator immediately following the imputation.
-The maximal imputation imputes all columns in the relation, regardless of which are required.
-Algorithm~\ref{algo:plan-helpers} presents
-a series of helper functions used by the top level planner, shown in Algorithm~\ref{algo:top-level-planner}.
+The maximal imputation imputes all of the attributes in the relation which are used by other operators in the query plan or are visible in the query output.
+
+Algorithm~\ref{algo:plan-helpers} presents a series of helper functions used by the top level planner, shown in Algorithm~\ref{algo:top-level-planner}.
 

### 2017-02-09 paper/algorithms/plan-helpers.tex
paper: keep refining algorithm section

@@
   \begin{algorithmic}
-    \Require{$q$ is a query plan.
-$C_{req}$ is the set of attributes that must be imputed in the output of this query plan.
-$C_{vis}$ is the set of attributes which are visible in the output of the overall query.}
+    \Require{$q$ is a query plan. $C_l$ is a set of attributes that must be imputed in the output of this query plan. $C_g$ is the set of attributes which are used in the final plan.}
     \Ensure{A set of query plans $Q$ with added imputation.}
-    \Function{AddImpute}{$q, C_{req}, C_{vis}$}
-    \State $D_{must} \gets \Call{Dirty}{q} \cap C_{req}$
-    \State $D_{may} \gets D_{must} \cup (\Call{Dirty}{q} \cap C_{vis})$
+    \Function{AddImpute}{$q, C_l, C_g$}
+    \State $D_{must} \gets \Call{Dirty}{q} \cap C_l$
+    \State $D_{may} \gets D_{must} \cup (\Call{Dirty}{q} \cap C_g)$
     \State $Q \gets \emptyset$
@@
     \Ensure{A set of plans for filtering $t$ with $\phi$.}
-    \Function{OptFilter}{$t, \phi, C_{vis}$}
-    \State \Return $\{\sigma_{\phi}(q)\ |\ q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}, C_{vis}}\}$
+    \Function{OptFilter}{$t, \phi, C_g$}
+    \State \Return $\{\sigma_{\phi}(q)\ |\ q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}, C_g}\}$
     \EndFunction
@@
 
-    \Require{$q$ is a query plan, $G$ is a set of grouping attributes, $A$ is an aggregation function, and $C_{vis}$ is a set of visible attributes.}
+    \Require{$q$ is a query plan, $G$ is a set of grouping attributes, $A$ is an aggregation function, and $C_g$ is a set of visible attributes.}
     \Ensure{A set of plans for grouping \& aggregating with $G$ and $A$.}
-    \Function{OptGroupBy}{$q, G, A, C_{vis}$}
+    \Function{OptGroupBy}{$q, G, A, C_g$}
     \State $C \gets G \cup \Call{Attrs}{A}$
-    \State $Q' \gets \Call{AddImpute}{q, C, C_{vis}}$
+    \State $Q' \gets \Call{AddImpute}{q, C, C_g}$
     \State \Return $\{\Call{GroupBy}{q', G, A} ~|~ q' \in Q'\}$
@@
     \Ensure{An updated plan cache $Q'$ which contains at least one plan that combines all of the tables in $T$.}
-    \Function{OptJoin}{$Q, T, J, C_{vis}$}
+    \Function{OptJoin}{$Q, T, J, C_g$}
     \For{$size \in 2...|T|$}
@@
     \State $C \gets \Call{Attrs}{\psi}$
-    \State $L \gets \{\Call{AddImpute}{q, C, C_{vis}} ~|~ q \in Q[t] \}$
-    \State $R \gets \{\Call{AddImpute}{q, C, C_{vis}} ~|~ q \in Q[S'] \}$
+    \State $L \gets \{\Call{AddImpute}{q, C, C_g} ~|~ q \in Q[t] \}$
+    \State $R \gets \{\Call{AddImpute}{q, C, C_g} ~|~ q \in Q[S'] \}$
     \State $Q[S] \lhd \{l \bowtie_\psi r ~|~ l \in L, r \in R\}$

### 2017-02-09 paper/algorithm.tex

@@
 The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
+Figure \ref{fig:query-schematic} shows a simple schematic of what a query involving three tables might look like, absent any imputation operators.
+
+\begin{figure}
+\Tree [.$\pi$ [.$GroupBy$ [.$\bowtie$ [.$\bowtie$ [.$\sigma$ $t_1$ ] [.$\sigma$ $t_2$ ] ] [.$\sigma$ $t_3$ ] ] ] ]
+\caption{Query plan schematic for the type of traditional plans explored (absent imputation operators)}
+\label{fig:query-schematic}
+\end{figure}
 

### 2017-02-09 paper/experiments.tex
paper: formatting

@@
 \section{Experiments}
-
+\todo{we of course need to rewrite this whole section}
 To evaluate the performance of our system, we generate plans for queries on two data sets: the American Community Survey and a synthetic dataset.
@@
 
-\begin{table}
+\begin{table*}
   \centerfloat
@@
   \label{fig:queries}
-\end{table}
+\end{table*}
 

### 2017-02-09 paper/main.tex
paper: added query schematic for basic plans
paper: keep refining algorithm section
paper: added description of cdc data to paper, along with queries

--- 2016-12-14 paper/main.tex
@@
 \usepackage{hyperref} % load after other packages
+\usepackage{graphicx} % includegraphics
+\usepackage{listings} % code samples with syntax hl
+\usepackage{qtree} % for simple tree diagrams
+\usepackage{subcaption}
 \addbibresource{references.bib}
+
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
+
+% listings settings
+\lstset{ %
+  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}; should come as last argument
+  basicstyle=\ttfamily\footnotesize,        % the size of the fonts that are used for the code
+  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
+  breaklines=true,                 % sets automatic line breaking
+  captionpos=b,                    % sets the caption-position to bottom
+  commentstyle=\color{red},    % comment style
+  %frame=single,	                   % adds a frame around the code
+  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
+  keywordstyle=\color{blue},       % keyword style
+%  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
+  numbersep=5pt,                   % how far the line-numbers are from the code
+%  numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
+  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
+%  stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
+  stringstyle=\color{mauve},     % string literal style
+  tabsize=2,	                   % sets default tabsize to 2 spaces
+}
 

### 2017-02-09 paper/queries.tex
paper: added description of cdc data to paper, along with queries

--- 2016-12-10 paper/queries.tex
@@
-\begin{tabular}{ll}
+\newcounter{queryno}
+\begin{tabular}{cl}
 \toprule
-\# & Query \\
+\# & \multicolumn{1}{c}{Query} \\
 \midrule
-0 & \verb|SELECT BLD FROM acs;| \label{q0} \\
-1 & \verb|SELECT BLD as units_in_structure, COUNT(ST) as estimate FROM acs GROUP BY BLD;| \label{q1} \\
-2 & \verb|SELECT BATH as has_bath, COUNT(ST) as ct FROM acs GROUP BY BATH;| \label{q2} \\
-3 & \verb|SELECT ACR as lotsize, AVG(BDSP) as avg_num_bedrooms FROM acs GROUP BY ACR;| \label{q3} \\
-4 & \verb|SELECT PSF as has_sub_families, SUM(NP) as num_people FROM acs GROUP BY PSF;| \label{q4} \\
-5 & \verb|SELECT AVG(NP) as avg_num_people FROM acs;| \label{q5} \\
-6 & \verb|SELECT BDSP as num_bedrooms, AVG(ACR) as avg_lot_size FROM acs WHERE VEH >= 2 GROUP BY BDSP;| \label{q6} \\
+1 & \verb|SELECT income, AVG(weight) FROM demo, exams WHERE demo.id = exams.id id GROUP BY income;| \refstepcounter{queryno} \label{q1} \\
+2 & 
+\begin{minipage}{6in}
+\begin{lstlisting}[breaklines]
+SELECT income, COUNT(*), AVG(cholesterol) 
+FROM demo, exams, labs WHERE demos.id = exams.id AND exams.id = labs.id AND
+income = 15 OR income = 13 AND weight >= 63 GROUP BY income;
+\end{lstlisting}
+\end{minipage}
+\refstepcounter{queryno} \label{q2} \\
+3 & \verb|SELECT MAX(blood_lead) FROM exams, labs WHERE exams.id = labs.id AND waist_circumference >= 87;| \refstepcounter{queryno} \label{q3} \\
+% 1 & \verb|SELECT BLD FROM acs;| \refstepcounter{queryno} \label{q1} \\
+% 2 & \verb|SELECT BLD as units_in_structure, COUNT(ST) as estimate FROM acs GROUP BY BLD;| \refstepcounter{queryno} \label{q2} \\
+% 3 & \verb|SELECT BATH as has_bath, COUNT(ST) as ct FROM acs GROUP BY BATH;| \refstepcounter{queryno} \label{q3} \\
+% 4 & \verb|SELECT ACR as lotsize, AVG(BDSP) as avg_num_bedrooms FROM acs GROUP BY ACR;| \refstepcounter{queryno} \label{q4} \\
+% 5 & \verb|SELECT PSF as has_sub_families, SUM(NP) as num_people FROM acs GROUP BY PSF;| \refstepcounter{queryno} \label{q5} \\
+% 6 & \verb|SELECT AVG(NP) as avg_num_people FROM acs;| \refstepcounter{queryno} \label{q6} \\
+% 7 & \verb|SELECT BDSP as num_bedrooms, AVG(ACR) as avg_lot_size FROM acs WHERE VEH >= 2 GROUP BY BDSP;| \refstepcounter{queryno} \label{q7} \\
+% 8 & \verb|SELECT MIN(RMSP) as min_num_rooms FROM acs WHERE RWAT=2;| \refstepcounter{queryno} \label{q8} \\
+% 9 & \verb|SELECT * FROM dirty m where m.f1 >= 2;| \refstepcounter{queryno} \label{q9} \\
+% 10 & \verb|SELECT * FROM dirty where dirty.f1 <= 2;| \refstepcounter{queryno} \label{q10} \\
+% 11 & \verb|SELECT MAX(f2) as max_f2 FROM dirty GROUP BY dirty.f1;| \refstepcounter{queryno} \label{q11} \\
+% 12 & \verb|SELECT m1.f1, m1.f2 FROM dirty m1, clean m2 WHERE m1.f1 = m2.f1 and m1.f1 = 1 and m2.f2 = 69;| \refstepcounter{queryno} \label{q12} \\
+% 13 & \verb|SELECT d1.f1 FROM dirty d1, clean d2 WHERE d1.f1 = d2.f1 and d2.f2 = 50;| \refstepcounter{queryno} \label{q13} \\
 \bottomrule

### 2017-02-09 simpledb/src/java/simpledb/StringHistogram.java
fix: stupid string bug, hacky fix for now

@@ -69,7 +69,8 @@ public class StringHistogram {
         if (tup.getField(index).isMissing()) {
             hist.incrCtMissing();
         } else {
-            hist.addValue(tup, index);
+            int val = stringToInt(((StringField) tup.getField(index)).getValue());
+            hist.addValue(val);
         }
     }
 

### 2017-02-09 paper/queries.tex
paper-fix: modified queries so that they actually work with simpledb

@@
 \midrule
-1 & \verb|SELECT income, AVG(weight) FROM demo, exams WHERE demo.id = exams.id id GROUP BY income;| \refstepcounter{queryno} \label{q1} \\
+1 & \verb|SELECT income, AVG(weight) FROM demo, exams WHERE demo.id = exams.id GROUP BY income;| \refstepcounter{queryno} \label{q1} \\
 2 & 
@@
 \begin{lstlisting}[breaklines]
-SELECT income, COUNT(*), AVG(cholesterol) 
-FROM demo, exams, labs WHERE demos.id = exams.id AND exams.id = labs.id AND
-income = 15 OR income = 13 AND weight >= 63 GROUP BY income;
+SELECT income, AVG(cholesterol) 
+FROM demo, exams, labs WHERE demo.id = exams.id AND exams.id = labs.id AND
+income <= 15 AND income >= 13 AND weight >= 63 GROUP BY income;
 \end{lstlisting}

### 2017-02-09 simpledb/src/java/simpledb/ImputedLogicalPlan.java
fix: bug in adding imputation before group/aggregate

@@ -432,32 +432,17 @@ public class ImputedLogicalPlan extends LogicalPlan {
 		ImputedPlan bestPlan = null;
 		if (aggField != null) {
 			for (Entry<Set<QualifiedName>, ImputedPlan> entry : bestPlans.entrySet()) {
-				for (ImputationType imp : ImputationType.values()) {
-					try {
-						LogicalAggregateNode plan = new LogicalAggregateNode(entry.getValue(), imp, groupByField, aggOp, aggField, tableMap);
-						if (bestPlan == null || plan.cost(lossWeight) < bestPlan.cost(lossWeight)) {
-							bestPlan = plan;
-						}
-					} catch (BadImputation e) {}
+				for (ImputedPlan plan : addImputes(entry.getValue(), globalRequired, globalRequired)) {
+					if (bestPlan == null || plan.cost(lossWeight) < bestPlan.cost(lossWeight)) {
+						bestPlan = plan;
+					}
 				}
 			}
 		}
 		else {
 			// Otherwise impute other outfields and select lowest cost plan
-
-			// retrieve quantified names of out fields
-			Set<QualifiedName> required = new HashSet<>();
-			for(int fieldIx : outFields) {
-				// TODO: hackish!
-				String fieldNm = td.getFieldName(fieldIx);
-				String[] aliasAndAttr = fieldNm.split("\\.");
-				required.add(new QualifiedName(aliasAndAttr[0], aliasAndAttr[1]));
-			}
-
-			// search over possible plans
 			for (Entry<Set<QualifiedName>, ImputedPlan> entry : bestPlans.entrySet()) {
-				for (ImputationType imp : ImputationType.values()) {
-					ImputedPlan plan = LogicalComposeImputation.create(entry.getValue(), imp, required, null);
+				for (ImputedPlan plan : addImputes(entry.getValue(), globalRequired, globalRequired)) {
 					if (bestPlan == null || plan.cost(lossWeight) < bestPlan.cost(lossWeight)) {
 						bestPlan = plan;
 					}

### 2017-02-09 paper/experiments.tex

@@
 \subsection{Data sets} \label{subsec:datasets}
+\subsubsection{Center for Disease Control: Surveys}
+For our first set of experiments, we use survey data collected by the 
+Center for Disease Control (CDC) in the United States. In particular, we
+experiment on a set of tables collected as part of the National
+Health and Nutrition Examination Survey (NHANES), a series of studies
+conducted by the CDC on a national sample of several thousand individuals\cite{nhanes}.
+The data consists on survey responses, physical examinations, and laboratory
+results, amongst others. Our dataset corresponds to the 2013-2014 NHANES
+and is available for direct download on the CDC's organization site on 
+kaggle.com, a popular data science website.
 
-The American Community Survey (ACS) provides a number of public data sets collected by the U.S. Census Bureau.
-We used a cleaned version of the 2012 Public Use Microdata Sample (PUMS) data kindly provided by the authors of~\cite{akande2015empirical}.
-The cleaning procedure is described in detail in~\cite{akande2015empirical}, but to summarize, the following data were removed:
+There are a total of 6 tables associated with the NHANES data set. For purposes
+of our experiments, we use three tables:
+
 \begin{itemize}
-\item Rows corresponding to vacant houses or single occupant households.
-\item Identification variables (e.g. area codes, serial numbers).
-\item Flag variables.
-\item Continuous variables.
+	\item Demographics: outlines demographic information of subjects
+	\item Examinations: physical exam results
+	\item Laboratory: laboratory exam results
 \end{itemize}
-The final data set consists of a single table with 671,153 rows and 37 columns, where all column entries are integers.
 
-In order to evaluate our implementation, we created a dirtied variant of the ACS data by randomly eliminating
-10\% of the fields across the table.
+The original tables have a large number of attributes, in some cases providing more granular
+tests results or alternative metrics. In order to simplify the presentation and
+exploration of queries, we focused on a subset of the attributes for each table.
+Table~\ref{table:nhanes-description} shows the attributes selected, along with the
+percentage of nulls. It is worth noting that for readability, we have replaced the
+NHANES variable names with self-explanatory attribute names.
 
-In order to evaluate additional queries, we constructed a synthetic table, consisting of 10,000 rows and 10 columns, with
-all values drawn from a uniform distribution between 0 and 100. To create a dirtied variant of the synthetic data, we randomly deleted 30\% of the fields
-across the table.
+\begin{table}
+    \centering
+    \begin{subtable}{0.5\textwidth}
+        \centering
+       \input{tables/cdc_demo}
+        \caption{Demographics}
+    \end{subtable}
+    \begin{subtable}{0.5\textwidth}
+        \centering
+        \input{tables/cdc_labs}
+        \caption{Laboratory Results}
+    \end{subtable}
+       \begin{subtable}{0.5\textwidth}
+        \centering
+        \input{tables/cdc_exams}
+        \caption{Physical Results}
+    \end{subtable}
+    \caption{Missing values in CDC NHANES 2013-2014 data}
+\end{table}
+
+All tables used consists of 10 attributes. The laboratory and examinations data,
+referred to as \textit{labs} and \textit{exams} in our queries, consists of 9813 rows,
+while the demographics data (\textit{demo}) has 10175 rows.
+
+\subsubsection{Some Other TBD dataset}
+\todo{write data description for this}
 
 \subsection{Queries}
-
 We collected a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
 
-The queries on the ACS data all consist only of projections, selections, and aggregations.
-The ACS data is contained in a single table, so there are relatively few join queries that are interesting on this data set and remain relevant for analysts, rather
-than contrived solely for the purpose of experimentation.
-However, even with select, filter, and aggregate, there are interesting choices to make with imputation placement. In order to explore joins, we leverage the synthetic
-dataset and write various ad-hoc queries.
+The queries on the CDC NHANES data consists not only of projections and selections,
+but also interesting joins and aggregations. Our aim was to craft meaningful queries
+that would provide performance figures relevant to practitioners who may use
+similar datasets.
+
+Query~\ref{q1} calculates the average weight for users based on their income data.
+To do so, it joins demographics and physical examination data. Query~\ref{q2}
+compares cholesterol levels for individuals who are both ends of the income
+spectrum and who satisfy a minimum weight requirement. Query~\ref{q3}
+extracts the maximum blood lead levels for subjects who satisfy a particular
+waist circumference requirement.
+
+\todo{describe further queries on other data set}
 

### 2017-02-11 paper/algorithms/histogram-transform.tex
paper: cleaned up histograms in algos

--- 2016-12-10 paper/algorithms/histogram-transform.tex
@@
-\begin{algorithm}
+%\begin{algorithm}
   \begin{algorithmic}
-    \Require{$H$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
-    \Ensure{Returns an updated $H$, such that the distribution of data matches the original $H$}
+    \Require{$\mathbb{H}$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
+    \Ensure{Returns an updated map $\mathbb{H'}$}
 
-    \Function{UpdateHistogram}{$H, op$}
+    \Function{UpdateHistograms}{$\mathbb{H}, op$}
     	\If{$op \not\in \{\delta, \mu, \sigma, \bowtie \}$}
-		\Return{H}
+		\Return{$\mathbb{H}$}
 	\EndIf
-	\State $H' \gets \Call{Copy}{H}$
+	\State $\mathbb{H'} \gets \Call{Copy}{\mathbb{H}}$
     	\If{op = $\delta_C$}
 		\For{$c \in C$}
-			\State $H'[c][null] \gets 0$
+			\State $\mathbb{H'}[c][null] \gets 0$
 		\EndFor
@@
 		\For{$c \in C$}
-			\State \Call{Add}{$H'[c], H'[c][null]$}
+			\State \Call{Add}{$\mathbb{H'}[c], \mathbb{H'}[c][null]$}
 		\EndFor
 	\ElsIf{op = $\sigma$}
-		\State \Call{ScaleBy}{H',  \Call{Selectivity}{$\sigma$}}
+		\For{$c \in dom(\mathbb{H'})$}
+			\State \Call{ScaleBy}{$\mathbb{H'}[c],  \Call{Selectivity}{\sigma}$} 
+		\EndFor
 	\ElsIf{op = $\bowtie_\phi$}
-		\State \Call{ScaleTo}{H', \Call{Cardinality}{$\bowtie_{\phi}$}}
+		\For{$c \in dom(\mathbb{H'})$}
+			\State \Call{ScaleTo}{$\mathbb{H'}[c], \Call{Cardinality}{\bowtie_{\phi}}$}
+		\EndFor
 	\EndIf
-	\Return{H'}
+	\Return{$\mathbb{H'}$}
     \EndFunction

### 2017-02-11 paper/experiments.tex
paper: added data section for fcc dataset

@@
 \todo{we of course need to rewrite this whole section}
-To evaluate the performance of our system, we generate plans for queries on two data sets: the American Community Survey and a synthetic dataset.
+To evaluate the performance of our system, we generate plans for queries
+for two separate survey-based data sets, showing that our system
+is ideal for early dataset exploration.
+
 \subsection{Data sets} \label{subsec:datasets}
@@
 Health and Nutrition Examination Survey (NHANES), a series of studies
-conducted by the CDC on a national sample of several thousand individuals\cite{nhanes}.
+conducted by the CDC on a national sample of several thousand individuals\cite{cdc-data}.
 The data consists on survey responses, physical examinations, and laboratory
@@
     \caption{Missing values in CDC NHANES 2013-2014 data}
+    \label{table:nhanes-description} 
 \end{table}
@@
 
-\subsubsection{Some Other TBD dataset}
-\todo{write data description for this}
+\subsubsection{freeCodeCamp: 2016 New Coder Survey}
+For our second set of experiments, we chose to use data collected
+by freeCodeCamp as a part of a survey of new software developers
+(both professional and amateur). freeCodeCamp is an open-source
+community that intends to help users learn how to program. In addition,
+to providing training material for users, it also pairs users with
+nonprofits to develop software solutions. Their \textit{2016 New Coder Survey} consists of responses by over 15,000 people to 48 different
+questions. The survey targeted users who were in one way or another
+related to coding organizations.
+
+For our purposes, we use a version of the data that has been 
+pre-processed, but where plenty of missing values remain. It is
+worth noting that given the table design (where many attributes
+are related), many of these missing values are to be
+expected. The dataset we started with had 15,620 rows and 113
+columns. We chose 17 attributes, which are outlined in Table~\ref{table:fcc-description}, along with the percentage of missing values.
+
+\begin{table}
+   \centering
+    \input{tables/fcc}
+    \caption{Missing values in FCC Survey Data}
+   \label{table:fcc-description} 
+\end{table}
 

### 2017-02-11 paper/examples.tex
paper: rewrote example section to reference cdc data

--- 2016-12-14 paper/examples.tex
@@
-\section{Examples}
+\section{Motivating Example}
+Consider the case of an epidemiologist tasked with an initial
+exploratory analysis of data collected for individuals across
+a battery of exams. In particular, she is interested in exploring
+the relationship between income and the immune system,
+controlling for factors such as weight and gender.
 
-Consider the case of an analyst looking to explore the sources and causes of
-polling error in a presidential election. One hypothesis \cite{zukin2015s}
-is that the rapid switch from landlines to cellphones increases the
-difficulty of contacting potential poll respondents. The analyst may
-begin her analysis by comparing the polling error in a state with the
-distribution of households with landlines, for the set of states for which
-polling error was most pronounced. In order to probe deeply, the analyst
-accesses a household-level survey that includes data on
-whether the household has a landline. One such survey --- which will be
-discussed further (Sec~\ref{subsec:datasets}) --- is the American Community Survey (ACS),
-conducted by the US Census Bureau.
+While careful modeling techniques will be necessary before putting
+results into the final research paper, the epidemiologist is anxious
+to get a quick and relatively accurate view into the data. However,
+the data has been collected through surveys, namely CDC surveys (see Section~\ref{subsec:datasets} for details on such data),
+and there is a significant amount of missing data across all
+relevant attributes. The researcher must know develop a strategy
+for filling in such missing values. She could drop any records with
+missing values and continue her analysis with the remaining data,
+or she could take advantage of existing correlations between attributes
+to fit a more sophisticated model on the base tables, which may take
+a long time.
 
-Surveys are rife with missing values due to non-response and other issues, and
-the ACS is no exception. The Census Bureau and other similar organization attempt to present
-data in as unadulterated a fashion as possible. The ACS dataset is not amenable to one-time
-imputation by either the Census Bureau, for this reason, or the analyst's organization,
-which likely uses the same database for a variety of unrelated uses (asking for different
-imputation strategies). Thus, the analyst must deal with missing values in one way or
-another.
+To complicate matters further, the epidemiologist realizes that
+various of the steps she is interested in, including filtering and grouping,
+are impacted differently by missing data, and that these step may change
+repeatedly, as she wants to run various queries during the exploration phase.
 
-In this case, a non-negligible fraction of respondents omitted information on whether they
-have a landline. The analyst hypothesizes that older houses are more likely to have
-landlines and that households that are relatively new are less likely to have landlines. 
-The analyst could take advantage of these presumed correlations to impute the missing values
-on the base table in one expensive operation. On the other hand, the query could be run
-directly and rows with missing values may be dropped entirely. With ImputeDB, the analyst
-could impute the relevant subset of data on-the-fly, achieving better performance without
-losing the information in the dropped examples.
+Rather than work through
+the implications of each case, she submits a query, as shown in Figure~\ref{fig:example-query}, to ImputeDB. The system, parameterized by a value indicating
+a tradeoff between information loss and runtime cost, then finds
+the optimal query plan, as shown in Figure~\ref{xxx}.
 
-The analyst submits a query (Figure~\ref{fig:example-query}) to ImputeDB which finds an optimized query
-plan\footnote{The analyst specifies a relatively high value of $\alpha$, the quality
-emphasis parameter, in order to receive a plan that does an expensive imputation step. See
-Section~\ref{sec:cost-model} for more details.}. The resulting query plan
-(Figure~\ref{fig:query-plan})
-imputes missing values for \verb!acs.TEL! after states with high polling error have been
-selected. This decision reduces the set of tuples that must be input into the imputation
-algorithm to just those that are relevant to the query. Later, we show the algorithms that
-the ImputeDB query optimizer uses in order to detect these opportunities and control the
-tradeoffs between efficiency and imputation quality. Finally, note that the optimizer doesn't
-place any imputation operator for the \verb|acs.ST| column, as it uses a histogram to
-detect that there are no missing values there (as might be expected for a survey question that isn't
-impacted by non-responses).
+Tens of queries later, the epidemiologist has the holistic view of the
+data that they required before carefully crafting their own tailored
+imputation model. Knowing that there may be a need to explore
+this data set further, they can easily incorporate their imputation model
+into ImputeDB for future use.
 
@@
 \begin{lstlisting}[language=SQL]
-SELECT polling.ST, AVG(acs.TEL)
-FROM polling, acs
-WHERE polling.ST = acs.ST
-  AND polling.ERROR > 50         -- 5 percentage points
-GROUP BY polling.ST;
+SELECT
+income,
+AVG(white_blood_cell_ct)
+FROM demo, exams, labs
+WHERE 
+gender = 2 AND weight >= 120 AND
+demo.id = exams.id AND exams.id = labs.id
+GROUP BY demo.income
 \end{lstlisting}
-\caption{A typical analyst query on ACS data}
+\caption{A typical public health query on CDC's NHANES data}
 \label{fig:example-query}
@@
 
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[width=0.6\textwidth]{figures/example.png}
-    \caption{Query plan generated by ImputeDB prioritizing imputation quality}
-    \label{fig:query-plan}
-\end{figure}
-
+\todo{add a small comment about the number of plans in the plancache
+to highlight size of search space}
+\todo{add a figure with the resulting query plan, and what its doing}
 %%% Local Variables:

### 2017-02-12 simpledb/src/java/simpledb/ImputeRegressionTree.java
First steps on improving impute w/ regression tree

@@ -78,16 +78,17 @@ public class ImputeRegressionTree extends Impute {
                     completeFieldsIndices.add(i);
                 }
             }
-            for (int j=0; j<buffer.size(); j++){
-                Tuple t = buffer.get(j);
-                Iterator<Integer> it = completeFieldsIndices.iterator();
-                while (it.hasNext()){
-                    int i = it.next();
-                    if (t.getField(i).isMissing()){
-                        it.remove();
-                    }
-                }
-            }
+            // Can we populate the missing bit array above and then re-use that here?
+			Iterator<Integer> it = completeFieldsIndices.iterator();
+			while (it.hasNext()){
+				int i = it.next();
+				for (int j=0; j<buffer.size(); j++){
+					if (buffer.get(j).getField(i).isMissing()){
+						it.remove();
+						continue;
+					}
+				}
+			}
 
             // Populate Instances object
             List<Integer> allFieldsToInclude = new ArrayList<Integer>(completeFieldsIndices);

### 2017-02-12 simpledb/data.py
paper: added extra gdp reference table for join query with fcc data

@@ -5,6 +5,7 @@
 # Can also load into postgres instead
 
 import argparse
+from collections import defaultdict
 import numpy as np
 import os
 import pandas as pd
@@ -89,6 +90,15 @@ fcc_columns = [
  ]
 fcc_columns = dict(zip(fcc_columns, map(lambda x: x.lower(), fcc_columns)))
 
+### World Bank Data GDP
+# http://databank.worldbank.org/data/reports.aspx?Code=NY.GDP.PCAP.CD&id=af3ce82b&report_name=Popular_indicators&populartype=series&ispopular=y#
+
+gdp_columns = {
+  'Country Name'  : 'country',
+  '2015 [YR2015]' : 'gdp_per_capita'
+}
+
+
 
 def read_csv(path, col_map):
   # avoid deciding type before, these files are dirty and small
@@ -135,27 +145,47 @@ def get_schema(nm, df):
   schema_cols = ", ".join(["%s %s" % (col, typ) for col, typ in zip(df.columns, types)])
   return "%s(%s)" % (nm, schema_cols)
 
-def string_to_integer(df):
-  to_str = df.columns[df.dtypes == object].tolist()
-  rewrite = {}
-  for col in to_str:
-    vals = sorted(list(set(df[col])))
-    enum_map = {c : i for i, c in enumerate(vals)}
-    df[col] = df[col].map(lambda x: enum_map[x])
-    rewrite[col] = enum_map
-  return df, rewrite
-  
-def enum_to_string(enum):
-  msg = ""
-  for col, mapping in enum.iteritems():
-    msg += "Enumeration for %s\n" % col
-    for s, i in mapping.iteritems():
-      msg += "%s:%d\n" % (s, i)
-  return msg
-  
-  
+def get_str_cols(df):
+  return df.columns[df.dtypes == object].tolist()
+
+def collect_enum(df, enum):
+  ## collect enumeration values
+  str_cols = get_str_cols(df)
+  for col in str_cols:
+    keys = set(df[col])
+    missing = sorted([k for k in keys if not k in enum])
+    val = max(enum.values()) + 1 if len(enum) else 0
+    for k in missing:
+      enum[k] = val
+      val += 1
+
+def apply_enum(df, enum):
+  # applies enum and filters out
+  # any records that have a null in
+  # the cols once the enum has been applied
+  str_cols = get_str_cols(df)
+  if len(enum) == 0 or len(str_cols) == 0:
+    return df
[…]

### 2017-02-12 paper/experiments.tex
paper: added queries for fcc to experiments section

@@
 
-\todo{describe further queries on other data set}
+In Query~\ref{q4}, we calculate the average income for survey participants,
+based on their bootcamp attendance. Query~\ref{q5} estimates the average
+age of women from the Uniter States who participated. Query~\ref{q6} 
+calculates the average amount of money survey participants with student debt destined to
+learning based on their school degree. Finally, in Query~\ref{q7}, we join the FCC data with a reference table provided
+by the World Bank which summarizes GDP per Capita across various
+countries\cite{worldbank-data}. The query calculates the average GDP-per-capita for participants based on whether or not they
+attended a bootcamp.
 
-\begin{table*}
-  \centerfloat
-  \input{queries}
+Note that for all queries we have enumerated strings and encoded them with
+an appropriate integer value.
+
+\begin{table}
+\centering
+ \begin{subtable}{\linewidth}
+  \input{queries-cdc}
+  \caption{Queries on CDC data}
+  \label{fig:queries-cdc}
+ \end{subtable}
+ ~
+ \begin{subtable}{\linewidth}
+ \input{queries-fcc}
+ \caption{Queries on FCC data}
+ \label{fig:queries-fcc}
+ \end{subtable}
   \caption{Queries used in our experiments.}
   \label{fig:queries}
-\end{table*}
+\end{table}
 

### 2017-02-13 simpledb/src/java/simpledb/LogicalComposeImputation.java
feat: use new loss model

@@ -56,8 +56,8 @@ public class LogicalComposeImputation extends ImputedPlan {
 		switch (imp) {
 		case DROP: {
 			final DbIterator physicalPlan = new Drop(toNames(impute), subplan.getPlan());
-			final double loss = estimateNumNulls(subplan, imputeIndices);
 			final double time = subplan.cardinality();
+			final double loss = 1.0;
 			final TableStats adjustedTableStats = subplanTableStats.adjustForImpute(DROP, imputeIndices);
 			return new LogicalComposeImputation(adjustedTableStats, physicalPlan, subplan, dirtySet, loss, time);
 		}
@@ -65,7 +65,7 @@ public class LogicalComposeImputation extends ImputedPlan {
 		case MINIMAL: 
 			Impute imputeOp = new ImputeRegressionTree(toNames(impute), subplan.getPlan());
 			final DbIterator physicalPlan = imputeOp;
-			final double loss = estimateNumNulls(subplan, imputeIndices) * (1 / Math.sqrt(totalData));
+			final double loss = (1 / Math.sqrt(totalData));
 			final int numComplete = schema.numFields() - dirtySet.size();
 			final double time = imputeOp.getEstimatedCost(imputeIndices.size(), numComplete, (int) subplan.cardinality());
 			final TableStats adjustedTableStats = subplanTableStats.adjustForImpute(MAXIMAL, imputeIndices);

### 2017-02-13 simpledb/src/java/simpledb/LogicalComposeImputation.java
fix: drops should be free (like filters)

@@ -56,8 +56,8 @@ public class LogicalComposeImputation extends ImputedPlan {
 		switch (imp) {
 		case DROP: {
 			final DbIterator physicalPlan = new Drop(toNames(impute), subplan.getPlan());
-			final double time = subplan.cardinality();
 			final double loss = 1.0;
+			final double time = 0.0;
 			final TableStats adjustedTableStats = subplanTableStats.adjustForImpute(DROP, imputeIndices);
 			return new LogicalComposeImputation(adjustedTableStats, physicalPlan, subplan, dirtySet, loss, time);
 		}

### 2017-02-15 paper/algorithm.tex
paper: cleaned up histograms in algos
paper: minor revisions to algorithm section

@@
 
-Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric (Sec.~\ref{sec:cost-model}) which combines the runtime cost of the query and the quality of the results.
-This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
-Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value in any attribute that they directly operate on.
+Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans containing imputation operators (Sec.~\ref{sec:operators}) for a plan that minimizes a metric (Sec.~\ref{sec:cost-model}) which combines the runtime cost of the query and the estimated quality of the results.
+This plan is subject to two constraints (Sec.~\ref{sec:placement}).
+First, it must not emit any tuples which contain null values, regardless of the state of the base tables.
+Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
+Our planning algorithm is agnostic to the type of imputation used (Sec.~\ref{sec:imputation}), as it is treated as a black-box operation.
 
 \subsection{Imputation operators}
+\label{sec:operators}
 We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
@@
 replace all null values with non-null values for attributes in $C$ in the relation $R$.
+The imputation algorithm is discussed in detail in Sec.~\ref{sec:imputation}.
 \textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
@@
 \label{sec:search-space}
-To reduce the size of the query plan search space, only plans that fit the following template are considered.
+To keep the query plan search space tractable, only plans that fit the following template are considered.
 
@@
 
-The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
-Figure \ref{fig:query-schematic} shows a simple schematic of what a query involving three tables might look like, absent any imputation operators.
+The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
+Figure~\ref{fig:query-schematic} shows the layout of a query involving three tables, absent any imputation operators.
 
 \begin{figure}
-\Tree [.$\pi$ [.$GroupBy$ [.$\bowtie$ [.$\bowtie$ [.$\sigma$ $t_1$ ] [.$\sigma$ $t_2$ ] ] [.$\sigma$ $t_3$ ] ] ] ]
-\caption{Query plan schematic for the type of traditional plans explored (absent imputation operators)}
+\Tree [.$\pi$ [.$g$ [.$\bowtie$ [.$\bowtie$ [.$\sigma$ $t_1$ ] [.$\sigma$ $t_2$ ] ] [.$\sigma$ $t_3$ ] ] ] ]
+\caption{Query plan schematic for the type of traditional plans explored (absent imputation operators).}
 \label{fig:query-schematic}
@@
 \label{sec:cost-model}
-We rank query plans using a cost model \[\textsc{Cost}(Q) = (1 - \alpha) \times \textsc{Time}(Q) + \alpha \times \textsc{Loss}(Q).\] $\textsc{Time}(Q)$ is an estimate of the runtime of the query derived from table statistics and selectivity estimation of the query predicates.
-$\textsc{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure.
+\todo{Explain that loss is now averaged}
+\todo{Do we want to keep the behavior that imputation increases in quality with the number of cells, or switch to number of tuples.}
+We rank query plans using a cost model \[\textsc{Cost}(q) = (1 - \alpha) \times \textsc{Time}(q) + \alpha \times \textsc{Loss}(q).\] $\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
+$\textsc{Loss}(q)$ is an estimate of the amount of error introduced by the imputation procedure.
 $\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance.
-$\alpha = 1.0$ means that the optimal query should be as accurate as possible (at the expense of performance), $\alpha=0.0$ means that the optimal query should be as fast
-as possible (at the expense of accuracy).
+$\alpha = 1.0$ means that the optimal query should have the highest quality imputation (at the expense of performance), $\alpha=0.0$ means that the optimal query should be as fast as possible (at the expense of imputation quality).
 
-In order to correctly estimate $\textsc{Time}(Q)$ and $\textsc{Loss}(Q)$, the system must have cardinality estimates for each sub-query in each query plan.
+The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of $q$ and its sub-queries.
 These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
@@
 When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
-Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans
+Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
 
@@
 \begin{algorithm}
+\input{algorithms/histogram-helpers.tex}
+\end{algorithm}
+
+\begin{algorithm}
 \input{algorithms/histogram-transform.tex}
@@
 \subsection{Imputation Strategies}
+\label{sec:imputation}
 

### 2017-02-15 paper/examples.tex
paper: minor revisions to motivation section

@@
 \section{Motivating Example}
-Consider the case of an epidemiologist tasked with an initial
+Consider an epidemiologist tasked with an initial
 exploratory analysis of data collected for individuals across
@@
 results into the final research paper, the epidemiologist is anxious
-to get a quick and relatively accurate view into the data. However,
-the data has been collected through surveys, namely CDC surveys (see Section~\ref{subsec:datasets} for details on such data),
+to get a quick and accurate view into the data. However,
+the data has been collected through CDC surveys (see Section~\ref{subsec:datasets} for details),
 and there is a significant amount of missing data across all
-relevant attributes. The researcher must know develop a strategy
-for filling in such missing values. She could drop any records with
+relevant attributes. The researcher must develop a strategy
+for handling the missing values. She could drop any records with
 missing values and continue her analysis with the remaining data,
-or she could take advantage of existing correlations between attributes
-to fit a more sophisticated model on the base tables, which may take
-a long time.
+or she could fit a model on the base tables which can take advantage of existing correlations between attributes to fill in the missing values.
+If she drops the records with missing data, she will throw away the majority of her data set, as many records have missing fields.
+Fitting a model on the dataset will take more programming effort and may take a long time to run.
 
+\todo{Clarify. Can we get a little more detail on why they might change the imputation strategy?}
 To complicate matters further, the epidemiologist realizes that
-various of the steps she is interested in, including filtering and grouping,
+some of the steps she is interested in, including filtering and grouping,
 are impacted differently by missing data, and that these step may change
@@
 Rather than work through
-the implications of each case, she submits a query, as shown in Figure~\ref{fig:example-query}, to ImputeDB. The system, parameterized by a value indicating
+the implications of each case, she submits a query, as shown in Figure~\ref{fig:example-query}, to ImputeDB.\@ The system, parameterized by a value indicating
 a tradeoff between information loss and runtime cost, then finds
@@
 \begin{lstlisting}[language=SQL]
-SELECT
-income,
-AVG(white_blood_cell_ct)
+SELECT income, AVG(white_blood_cell_ct)
 FROM demo, exams, labs
-WHERE 
-gender = 2 AND weight >= 120 AND
-demo.id = exams.id AND exams.id = labs.id
+WHERE gender = 2 AND 
+      weight >= 120 AND
+      demo.id = exams.id AND 
+      exams.id = labs.id
 GROUP BY demo.income

### 2017-02-15 paper/examples.tex
paper: add query plan trees

@@
 
+\begin{figure}
+  \Tree
+  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
+    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
+      [.$\mu_{\text{demo.income}}$
+        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
+          [.$\mu_{\text{labs.white\_blood\_cell\_ct}}$
+            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
+              [.$\sigma_{\text{exams.weight} \geq 120}$ 
+                [.$\mu_{\text{exams.weight}}$ exams ] 
+              ] 
+              labs 
+            ]
+          ]
+        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
+      ] 
+    ] 
+  ] 
+]
+\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for quality.}
+\label{fig:quality-plan}
+\end{figure}
+
+\begin{figure}
+  \Tree
+  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
+    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
+      [.$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$
+        [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
+          [.$\bowtie_{\text{demo.id} = \text{exams.id}}$
+            [.$\sigma_{\text{demo.gender} = 2}$ demo ]
+            [.$\sigma_{\text{exams.weight} \geq 120}$ [.$\mu_{\text{exams.weight}}$ exams ] ] ] labs ] ] ] ]
+\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for performance.}
+\label{fig:fast-plan}
+\end{figure}
+
+First, consider the plan in Fig.~\ref{fig:fast-plan}.
+This plan will be generated for the query in Fig.~\ref{example-query} when the optimizer is requested to produce the fastest possible plan, without considering the accuracy penalty.
+There are several features of this query plan that highlight the subtleties of optimizing queries which contain imputation operators.
+First, this plan will drop tuples which contain null values, rather than
+
 \todo{add a small comment about the number of plans in the plancache

### 2017-02-16 paper/examples.tex
paper: moved motivating example before related work. Cleaned up and added one todo

@@
 
-\todo{Clarify. Can we get a little more detail on why they might change the imputation strategy?}
 To complicate matters further, the epidemiologist realizes that
@@
 repeatedly, as she wants to run various queries during the exploration phase.
+The imputation strategy itself may change through out time. For example, as she is getting an initial
+set of distribution statistics (such as mean and standard deviation), she may be willing to tolerate
+a simpler imputation approach, as long as she gets runtime performance in exchange. Later on,
+as more precise estimates are needed for comparing treatments or policy changes, the researcher
+may want to focus on the quality of the imputation.
 
@@
 a tradeoff between information loss and runtime cost, then finds
-the optimal query plan, as shown in Figure~\ref{xxx}.
+the optimal query plan. Figure~\ref{fig:quality-plan} and Figure~\ref{fig:fast-plan} show what this plan would be
+when optimizing for quality and performance, respectively. These two plans are found within a broader search space
+of XXXX plans. \todo{add a small comment about the number of plans in the plancache
+to highlight size of search space}.
 
-Tens of queries later, the epidemiologist has the holistic view of the
-data that they required before carefully crafting their own tailored
-imputation model. Knowing that there may be a need to explore
-this data set further, they can easily incorporate their imputation model
-into ImputeDB for future use.
+The differences between the two plans highlight the subtleties of optimizing queries which contain imputation operators.
+Note that imputation operators are placed in different points of the plan, operating over different columns, and indeed
+even influencing the order of joins. The quality-optimized plan (Figure~\ref{fig:quality-plan}) uses the impute operation, rather
+than dropping tuples with missing values. It waits to impute \verb|demo.income| until after the final join has taken place, but
+other imputations take place earlier on in the plan, some before filtering and join operations. 
+Meanwhile, the performance-optimized plan (Figure~\ref{fig:fast-plan}) performs a single impute operation (\verb|exams.weight|), and applies drops to tuples with missing values before aggregating.
 
@@
 
-First, consider the plan in Fig.~\ref{fig:fast-plan}.
-This plan will be generated for the query in Fig.~\ref{example-query} when the optimizer is requested to produce the fastest possible plan, without considering the accuracy penalty.
-There are several features of this query plan that highlight the subtleties of optimizing queries which contain imputation operators.
-First, this plan will drop tuples which contain null values, rather than
+Tens of queries later, the epidemiologist has the holistic view of the
+data that they required before carefully crafting their own tailored
+imputation model. Knowing that there may be a need to explore
+this data set further, they can easily incorporate their imputation model
+into ImputeDB for future use.
 
-\todo{add a small comment about the number of plans in the plancache
-to highlight size of search space}
-\todo{add a figure with the resulting query plan, and what its doing}
 %%% Local Variables:

### 2017-02-16 paper/main.tex
paper: \verb to \lstinline, add syntax highlighting
paper: moved motivating example before related work. Cleaned up and added one todo
paper: rewrote abstract from class project to vldb version

@@
 \usepackage{listings} % code samples with syntax hl
+\usepackage{stmaryrd}
 \usepackage{qtree} % for simple tree diagrams
@@
   tabsize=2,	                   % sets default tabsize to 2 spaces
+  language=SQL
 }
@@
 \begin{abstract}
-In order to study the placement of an imputation step, we create a logical imputation operator (along with respective physical instances) and incorporate its placement as part of the query plan optimization process into a simple database prototype.
-We introduce measures for information loss and runtime for imputation operations, which outline the main trade-offs in the imputation placement.
-We add these measures into our cost estimation, allowing us to intelligently place the data imputation step during query planning.
-We show the trade-offs between efficiency and accuracy for simple data imputation models.
+Missing values abound in data analysis and often present a serious challenge for users, who
+are forced to pick between removing tuples with missing values or creating a clean duplicate of their data
+by  applying a potentially expensive imputation strategy. We show that it is possible
+to incorporate imputation into a cost-based query optimizer, performing imputation on the fly
+for each query. This allows users to immediately explore
+their data, while the system picks the optimal placement of imputation operations based
+on the specific query issued and a parameter that the user selects for evaluating the tradeoff between data
+quality and runtime performance. We incorporate the imputation operators into the relational algebra, allowing
+for traditional optimizations to take place. Our experiments evaluate this approach on two real-world survey-based datasets
+from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}, showing that the 
+optimizer generates query plans that execute in a fraction of the time that it takes to impute values at the base
+tables, making exploratory data analysis using our system feasible. Furthermore, we show that the query results
+from on-the-fly imputation are comparable to those generated with the standard full-table imputation approach.
 \end{abstract}
@@
 \input{introduction}
+\input{examples}
 \input{relatedwork}
-\input{examples}
 \input{algorithm}

### 2017-02-16 paper/conclusion.tex
paper: added todos

@@
 \section{Conclusion}
-
+\todo{jose: rewrite this}
 As shown in \ProjectName{}, missing values and their imputation can successfully be integrated into the relational calculus and
@@
 \subsection{Future Work}
+\todo{Jose: rewrite this}
 \ProjectName{} opens up multiple avenues for further work. For instance, we

### 2017-02-16 paper/experiments.tex
paper: use full width figures for tables
paper: added todos

@@
 \section{Experiments}
-\todo{we of course need to rewrite this whole section}
-To evaluate the performance of our system, we generate plans for queries
+To evaluate the performance of our system, we generate and execute plans for queries
 for two separate survey-based data sets, showing that our system
@@
 \subsection{Queries}
+\todo{Add more queries: we want ~10 (balanced between the two datasets), explain them below}
 We collected a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
@@
 
-\begin{table}
+\begin{table*}
 \centering
@@
   \label{fig:queries}
-\end{table}
+\end{table*}
 
-\begin{table}
+\begin{table*}
   \centerfloat
@@
   \label{fig:experiments}
-\end{table}
+\end{table*}
 
+\todo{actually run experiments and rewrite this section}
 Table~\ref{fig:experiments} provides a summary of our experimental results.

### 2017-02-16 paper/conclusion.tex
paper: rewrote conclusion section

--- 2016-12-14 paper/conclusion.tex
@@
 \section{Conclusion}
-
-As shown in \ProjectName{}, missing values and their imputation can successfully be integrated into the relational calculus and
-existing plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
+As shown in \ProjectName{}, missing values and their imputation can be successfully integrated into the relational calculus and
+existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
 learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
-information loss and time. In effect, user preferences for one or the other can be easily incorporated by adjusting a parameter.
-Simple histogram transformations
-provide incrementally updated cardinality estimates across operators in the plans, which allow us to provide more accurate cost estimates.
-By taking a dynamic programming approach, we can consider a variety of operator placements
-and input columns, while keeping planning tractable in real-world examples.
-
-In our experiments, we considered a series of analytic queries (using relevant aggregates) and set queries (with simple projections) on real-world
-American Community Survey data
-and showed that the chosen query plan improves accuracy of results relative to simply ignoring missing values. Furthermore,
-the variety of imputation operator placements across queries further emphasizes the lack of flexibility imposed by the coarse-grained
-pre-processing approach to imputation. By allowing different imputation strategies for different queries, we
-take a fine-grained approach to missing data, better addressing the often differing needs of database users.
+information loss and runtime. We showed that different values for the tradeoff parameter can yield substantially
+different plans, allowing the user to express their preferences for performance on a per-query basis.
+Our experiments with \textit{CDC} and \textit{freeCodeCamp} survey-based datasets
+showed that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
+We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
+the data exploration phase. The plans selected for each query result in execution that
+takes a fraction of the time compared to the standard approach of imputing on the entire base tables and
+then formulating queries. Furthermore, the difference in query results between the two approaches is
+shown to be minimal. Data analysts need not commit to a specific imputation strategy and can instead
+vary this across queries.
 
@@
 regarding missing values inside the system, allowing users to use traditional SQL and engage in normal workloads.
+While prior work has incorporated operations such as prediction into their databases, this has been domain-specific
+in some cases, and in others, they have not integrated these operations into the planner. The novelty
+of our contribution lies in the formalization of missing value imputation for query planning, which results in performance
+gains over traditional approaches. This approach acknowledges that different users performing different queries on the same
+dataset will likely have varying imputation needs and that execution plans should appropriately reflect that variation.
 
 \subsection{Future Work}
+\todo{Jose: rewrite this}
 \ProjectName{} opens up multiple avenues for further work. For instance, we

### 2017-02-16 paper/relatedwork.tex
paper: check biblio, add nocite, add a bit to intro for importance of data cleaning, remove an old reference to acs

@@
 In \cite{akande2015empirical}, Akande et al analyze the performance of various
-multiple imputation techniques on the American Community Survey dataset
-(Sec~\ref{subsec:datasets}). The computational difficulties of imputing on large base
+multiple imputation techniques on the American Community Survey dataset. 
+The computational difficulties of imputing on large base
 tables are well known and can limit approaches. For example, Akande finds that

### 2017-02-17 paper/relatedwork.tex
paper: minor fix on bayesdb language

@@
 leverage statistical inference techniques in a database. Non-experts
-can use a simple declarative language, akin to SQL, to specify models
+can use a simple declarative language (an extension of SQL), to specify models
 which allow missing value imputation, amongst other broader functionality.

### 2017-02-18 simpledb/src/java/simpledb/AImputedPlanCache.java
feat: add plan cache which maintains pareto frontier
New plan cache keeps the pareto frontier of the plan space rather than
pruning using the cost value for the particular value of alpha.

@@ -1,0 +1,124 @@
+package simpledb;
+
+import java.util.Arrays;
+import java.util.HashMap;
+import java.util.HashSet;
+import java.util.Set;
+import java.util.TreeSet;
+
+public abstract class AImputedPlanCache implements ImputedPlanCache {
+	protected static class Key {
+		public final Set<String> tables;
+		public final Set<QualifiedName> dirtySet;
+
+		public Key(Set<String> tables, Set<QualifiedName> dirtySet) {
+			this.tables = tables;
+			this.dirtySet = dirtySet;
+		}
+
+		@Override
+		public boolean equals(Object o) {
+			if (o == null) {
+				return false;
+			}
+			if (!(o instanceof Key)) {
+				return false;
+			}
+
+			Key ok = (Key) o;
+			return tables.equals(ok.tables) && dirtySet.equals(ok.dirtySet);
+		}
+
+		@Override
+		public int hashCode() {
+			final int prime = 31;
+			int result = 1;
+			result = prime * result + tables.hashCode();
+			result = prime * result + dirtySet.hashCode();
+			return result;
+		}
+
+		@Override
+		public String toString() {
+			String tableStr = "Tables(" + Arrays.toString(this.tables.toArray()) + ")";
+			String dirtyStr = "Dirty(" + Arrays.toString(this.dirtySet.toArray()) + ")";
+			return tableStr + "," + dirtyStr;
+		}
+	}
+
+	public static class Value implements Comparable<Value> {
+		public final ImputedPlan plan;
+		public final Set<LogicalJoinNode> joins;
+
+		public Value(ImputedPlan plan, Set<LogicalJoinNode> joins) {
+			this.plan = plan;
+			this.joins = joins;
+		}
+
+		public Value(ImputedPlan plan) {
+			this(plan, new HashSet<LogicalJoinNode>());
+		}
+
+		public boolean dominates(Value o) {
+			return plan.loss().get() < o.plan.loss().get() && plan.time() < o.plan.time();
+		}
+
+		@Override
+		public int compareTo(Value o) {
+			int lcmp = Double.compare(plan.loss().get(), o.plan.loss().get());
+			if (lcmp != 0) { return lcmp; }
+			return Double.compare(plan.time(), o.plan.time());
+		}
+	}
+
[…]

### 2017-02-18 simpledb/src/java/simpledb/ImputedLogicalPlan.java

@@ -3,7 +3,6 @@ package simpledb;
 import java.io.File;
 import java.io.FileNotFoundException;
 import java.util.*;
-import java.util.Map.Entry;
 
 /**
  * The JoinOptimizer class for the extend imputation functionality. This not
@@ -113,7 +112,7 @@ public class ImputedLogicalPlan extends LogicalPlan {
 			
 			// If there's no filters, add the scan plan as is.
 			if (filters == null) { 
-				cache.addPlan(tablesInPlan, scanPlan.getDirtySet(), scanPlan, lossWeight);
+				cache.addPlan(tablesInPlan, scanPlan);
 			} 
 			
 			// Otherwise, impute as needed and add on the required filters.
@@ -128,7 +127,7 @@ public class ImputedLogicalPlan extends LogicalPlan {
 				
 				for (ImputedPlan imputedScan : imputedScans) {
 					ImputedPlan filterPlan = new LogicalImputedFilterNode(tid, imputedScan, filters);
-					cache.addPlan(tablesInPlan, filterPlan.getDirtySet(), filterPlan, lossWeight);
+					cache.addPlan(tablesInPlan, filterPlan);
 				}
 			}
 		}
@@ -198,8 +197,7 @@ public class ImputedLogicalPlan extends LogicalPlan {
 		news.clear(joinToRemove);
 
 		// possible plans to arrive at left and right sides of the join
-		Map<Set<QualifiedName>, ImputedPlan> leftPlans;
-		Map<Set<QualifiedName>, ImputedPlan> rightPlans;
+		Iterable<ImputedPlanCachePareto.Value> leftPlans, rightPlans;
 
 		// attributes that must be imputed for join
 		Set<QualifiedName> required = new HashSet<>();
@@ -207,9 +205,9 @@ public class ImputedLogicalPlan extends LogicalPlan {
 		required.add(new QualifiedName(j.t2Alias, j.f2PureName));
 
 		// names of table in already performed join
-		Set<String> leftTables = new HashSet<>();
+		final Set<String> leftTables = new HashSet<>();
 		// alias for table to join in
-		Set<String> rightTables = new HashSet<>();
+		final Set<String> rightTables = new HashSet<>();
 		Set<String> allTables = new HashSet<>();
 		// just joining existing base relations
 		boolean isSimpleJoin;
@@ -219,11 +217,11 @@ public class ImputedLogicalPlan extends LogicalPlan {
 			leftTables.add(j.t1Alias);
[…]

### 2017-02-18 simpledb/src/java/simpledb/ImputedPlanCache.java

@@ -1,149 +1,15 @@
 package simpledb;
-import java.util.*;
-import java.util.Map.Entry;
 
-/**
- * Keeps information on the best plan that includes a given set of tables and has a given set of dirty attributes
- * Insertions into the cache are handled directly to only overwrite if of lower cost.
- */
-public class ImputedPlanCache {	
-    private class ImputeKey {
-        public final Set<String> tables;
-        public final Set<QualifiedName> dirtySet;
-        
-        public ImputeKey(Set<String> tables, Set<QualifiedName> dirtySet) {
-            this.tables = tables;
-            this.dirtySet = dirtySet;
-        }
+import java.util.Set;
 
-        @Override
-        public boolean equals(Object o) {
-            if (o == null) {
-                return false;
-            }
-            if (!(o instanceof ImputeKey)) {
-                return false;
-            }
+import simpledb.AImputedPlanCache.Value;
 
-            ImputeKey ok = (ImputeKey) o;
-            return tables.equals(ok.tables) && dirtySet.equals(ok.dirtySet);
-        }
+public interface ImputedPlanCache {
+	Iterable<Value> bestPlans(Set<String> tables);
 
-        @Override
-        public int hashCode() {
-            final int prime = 31;
-            int result = 1;
-            result = prime * result + tables.hashCode();
-            result = prime * result + dirtySet.hashCode();
-            return result;
-        }
+	Iterable<Value> bestPlans(Set<String> tables, Set<QualifiedName> dirty);
 
-        @Override
-        public String toString() {
-            String tableStr = "Tables(" + Arrays.toString(this.tables.toArray()) + ")";
-            String dirtyStr = "Dirty(" + Arrays.toString(this.dirtySet.toArray()) + ")";
-            return tableStr + "," + dirtyStr;
-        }
-    }
+	void addPlan(Set<String> ts, ImputedPlan newPlanIP);
 
-    private final Map<ImputeKey, ImputedPlan> bestPlans;
-    // joins necessary to get to this state
-    private final Map<ImputeKey, Set<LogicalJoinNode>> necessaryJoins;
-    
-    public ImputedPlanCache() {
-        bestPlans = new HashMap<ImputeKey, ImputedPlan>();
-        necessaryJoins = new HashMap<ImputeKey, Set<LogicalJoinNode>>();
-    }
-
-    /**
-     * Add a plan if it is of lower cost for the appropriate (tables, dirty set) key
[…]

### 2017-02-18 simpledb/src/java/simpledb/ImputedPlanCacheDotted.java

@@ -4,13 +4,16 @@ import java.io.File;
 import java.io.FileNotFoundException;
 import java.io.IOException;
 import java.util.Set;
-import java.util.TreeSet;
 
-public class ImputedPlanCacheDotted extends ImputedPlanCache {
+import simpledb.AImputedPlanCache.Value;
+
+public class ImputedPlanCacheDotted implements ImputedPlanCache {
 	private final File outDir;
+	private final ImputedPlanCache cache;
 	
-	public ImputedPlanCacheDotted(File outDir) throws FileNotFoundException {
-		super();
+	public ImputedPlanCacheDotted(File outDir, ImputedPlanCache cache) throws FileNotFoundException {
+		this.cache = cache;
+		
 		if (!outDir.exists()) {
 			outDir.mkdir();
 		} else if (!outDir.isDirectory()) {
@@ -20,11 +23,17 @@ public class ImputedPlanCacheDotted extends ImputedPlanCache {
 		this.outDir = outDir;
 	}
 	
-	private void writePlan(Set<QualifiedName> dirtySet, ImputedPlan plan, double lossWeight) {
+	private void writePlan(Set<QualifiedName> dirtySet, ImputedPlan plan) {
 		StringBuilder prefix = new StringBuilder();
-		prefix.append(plan.cost(lossWeight));
+		prefix.append(plan.loss().get());
 		prefix.append(" ");
-		prefix.append(plan.cardinality());
+		prefix.append(plan.time());
+		prefix.append(" ");
+		try {
+			prefix.append(plan.cardinality());
+		} catch (UnsupportedOperationException e) {
+			prefix.append("??");
+		}
 		prefix.append(" ");
 		
 		try {
@@ -36,14 +45,24 @@ public class ImputedPlanCacheDotted extends ImputedPlanCache {
 		}
 	}
 	
-	void addJoinPlan(Set<String> ts, Set<QualifiedName> dirtySet, Set<LogicalJoinNode> joins, ImputedPlan newPlan, double lossWeight) {
-		super.addJoinPlan(ts, dirtySet, joins, newPlan, lossWeight);
-		writePlan(dirtySet, newPlan, lossWeight);
+	public void addJoinPlan(Set<String> ts, Set<LogicalJoinNode> joins, ImputedPlan newPlan) {
+		cache.addJoinPlan(ts, joins, newPlan);
[…]

### 2017-02-18 simpledb/src/java/simpledb/ImputedPlanCacheLogged.java

@@ -1,0 +1,76 @@
+package simpledb;
+
+import java.io.File;
+import java.io.FileNotFoundException;
+import java.io.FileOutputStream;
+import java.io.PrintWriter;
+import java.util.HashSet;
+import java.util.Set;
+
+import simpledb.AImputedPlanCache.Value;
+
+public class ImputedPlanCacheLogged implements ImputedPlanCache {
+	private final PrintWriter writer;
+	private final ImputedPlanCache cache;
+	private final QueryPlanVisualizer viz = new QueryPlanVisualizer();
+	
+	public ImputedPlanCacheLogged(File log, ImputedPlanCache cache) throws FileNotFoundException {
+		this.cache = cache;
+		writer = new PrintWriter(new FileOutputStream(log, true));
+	}
+	
+	private <T> void writeSet(Set<T> s) {
+		int i = 0;
+		writer.print("{");
+		for (T e : s) {
+			writer.print(" ");
+			writer.print(e);
+			if (i < s.size()) {
+				writer.print(",");
+			} else {
+				writer.print(" ");
+			}
+			i++;
+		}
+		writer.print("}");
+	}
+	
+	@Override
+	public void addPlan(Set<String> ts, ImputedPlan newPlan) {
+		addJoinPlan(ts, new HashSet<LogicalJoinNode>(), newPlan);
+	}
+	
+	@Override
+	public void addJoinPlan(Set<String> ts, Set<LogicalJoinNode> joins, ImputedPlan newPlan) {
+		cache.addJoinPlan(ts, joins, newPlan);
+		
+		writer.print("Tables: ");
+		writeSet(ts);
+		writer.print("\n");
+		
+		writer.print("Dirty: ");
+		writeSet(newPlan.getDirtySet());
+		writer.print("\n");
+		
+		writer.format("Loss: %f\n", newPlan.loss().get());
+		writer.format("Time: %f\n", newPlan.time());
+		
+		try {
+			writer.print(viz.getQueryPlanTree(newPlan.getPlan()));
+		} catch (ArrayIndexOutOfBoundsException e) {
+		}
+		writer.println();
+		
+		writer.flush();
+	}
+
+	@Override
+	public Iterable<Value> bestPlans(Set<String> tables) {
+		return cache.bestPlans(tables);
+	}
+
+	@Override
+	public Iterable<Value> bestPlans(Set<String> tables, Set<QualifiedName> dirty) {
+		return cache.bestPlans(tables, dirty);
+	}
+}

### 2017-02-18 simpledb/src/java/simpledb/ImputedPlanCachePareto.java

@@ -1,0 +1,51 @@
+package simpledb;
+import java.util.*;
+
+/**
+ * Keeps information on the best plan that includes a given set of tables and has a given set of dirty attributes
+ * Insertions into the cache are handled directly to only overwrite if of lower cost.
+ */
+public class ImputedPlanCachePareto extends AImputedPlanCache {	
+    public ImputedPlanCachePareto() {
+        super();
+    }
+
+    /**
+     * Add a plan if it is of lower cost for the appropriate (tables, dirty set) key
+     * @param ts tables in the underlying plan
+     * @param newPlan plan
+     */
+    public void addPlan(Set<String> ts, ImputedPlan newPlanIP) {
+        addJoinPlan(ts, new HashSet<LogicalJoinNode>(), newPlanIP);
+    }
+
+    /**
+     * Add a plan involving joins if it is of lower cost for the appropriate (tables, dirty set) key
+     * @param ts tables in the underlying plan
+     * @param joins join predicates involved in the plan
+     * @param newPlan plan
+     */
+    public void addJoinPlan(Set<String> ts, Set<LogicalJoinNode> joins, ImputedPlan newPlanIP) {
+    	Key key = new Key(ts, newPlanIP.getDirtySet());
+        Value newPlan = new Value(newPlanIP, joins);
+        if (bestPlans.containsKey(key)) {
+        	SortedSet<Value> plans = bestPlans.get(key);
+        	
+        	// Find dominating plans.
+        	for (Value plan : plans) {
+        		// Found a dominating plan.
+        		if (plan.dominates(newPlan)) { return; }
+        	}
+        	
+        	// Remove dominated plans.
+        	plans.removeIf(plan -> newPlan.dominates(plan));
+
+        	plans.add(newPlan);
+        } else {
+        	// always insert if we don't have any info on this key combo
+        	TreeSet<Value> plans = new TreeSet<>();
+        	plans.add(newPlan);
+            bestPlans.put(key, plans);
+        }
+    }
+}

### 2017-02-18 simpledb/src/java/simpledb/ImputedPlanCacheSingle.java

@@ -1,0 +1,39 @@
+package simpledb;
+
+import java.util.HashSet;
+import java.util.Set;
+import java.util.SortedSet;
+import java.util.TreeSet;
+
+public class ImputedPlanCacheSingle extends AImputedPlanCache {
+	private final double lossWeight;
+	
+	public ImputedPlanCacheSingle(double lossWeight) {
+		this.lossWeight = lossWeight;
+	}
+	
+	@Override
+	public void addPlan(Set<String> ts, ImputedPlan newPlanIP) {
+		addJoinPlan(ts, new HashSet<LogicalJoinNode>(), newPlanIP);
+	}
+
+	@Override
+	public void addJoinPlan(Set<String> ts, Set<LogicalJoinNode> joins, ImputedPlan newPlanIP) {
+		Key key = new Key(ts, newPlanIP.getDirtySet());
+        Value newPlan = new Value(newPlanIP, joins);
+        if (bestPlans.containsKey(key)) {
+        	SortedSet<Value> plans = bestPlans.get(key);
+        	assert plans.size() == 1;
+        	
+        	if (plans.first().plan.cost(lossWeight) > newPlan.plan.cost(lossWeight)) {
+        		plans.clear();
+        		plans.add(newPlan);
+        	}
+        } else {
+        	// always insert if we don't have any info on this key combo
+        	TreeSet<Value> plans = new TreeSet<>();
+        	plans.add(newPlan);
+            bestPlans.put(key, plans);
+        }
+	}
+}

### 2017-02-18 simpledb/src/java/simpledb/LoggedImputedPlanCache.java

@@ -1,79 +1,0 @@
-package simpledb;
-
-import java.io.File;
-import java.io.FileNotFoundException;
-import java.io.FileOutputStream;
-import java.io.PrintWriter;
-import java.util.Set;
-
-public class LoggedImputedPlanCache extends ImputedPlanCache {
-	private final PrintWriter writer;
-	private final QueryPlanVisualizer viz = new QueryPlanVisualizer();
-	
-	public LoggedImputedPlanCache(File log) throws FileNotFoundException {
-		super();
-		writer = new PrintWriter(new FileOutputStream(log, true));
-	}
-	
-	private <T> void writeSet(Set<T> s) {
-		int i = 0;
-		writer.print("{");
-		for (T e : s) {
-			writer.print(" ");
-			writer.print(e);
-			if (i < s.size()) {
-				writer.print(",");
-			} else {
-				writer.print(" ");
-			}
-			i++;
-		}
-		writer.print("}");
-	}
-	
-	@Override
-	void addPlan(Set<String> ts, Set<QualifiedName> dirtySet, ImputedPlan newPlan, double lossWeight) {
-		super.addPlan(ts, dirtySet, newPlan, lossWeight);
-		
-		writer.print("Tables: ");
-		writeSet(ts);
-		writer.print("\n");
-		
-		writer.print("Dirty: ");
-		writeSet(dirtySet);
-		writer.print("\n");
-		
-		writer.format("Cost: %f\n", newPlan.cost(lossWeight));
-		
-		try {
-			writer.print(viz.getQueryPlanTree(newPlan.getPlan()));
-		} catch (ArrayIndexOutOfBoundsException e) {
-		}
-		writer.println();
-		
-		writer.flush();
-	}
-	
-	@Override
-	void addJoinPlan(Set<String> ts, Set<QualifiedName> dirtySet, Set<LogicalJoinNode> joins, ImputedPlan newPlan, double lossWeight) {
-		super.addJoinPlan(ts, dirtySet, joins, newPlan, lossWeight);
-		
-		writer.print("Tables: ");
-		writeSet(ts);
-		writer.print("\n");
-		
-		writer.print("Dirty: ");
-		writeSet(dirtySet);
-		writer.print("\n");
-		
-		writer.format("Cost: %f\n", newPlan.cost(lossWeight));
-		
-		try {
-			writer.print(viz.getQueryPlanTree(newPlan.getPlan()));
[…]

### 2017-02-18 simpledb/src/java/simpledb/LogicalAggregateNode.java

@@ -27,7 +27,7 @@ public class LogicalAggregateNode extends ImputedPlan {
 
 	@Override
 	public Set<QualifiedName> getDirtySet() {
-		throw new RuntimeException("Not implemented.");
+		return Collections.emptySet();
 	}
 
 	@Override
@@ -47,6 +47,6 @@ public class LogicalAggregateNode extends ImputedPlan {
 
 	@Override
 	public double cardinality() {
-		throw new RuntimeException("Not implemented.");
+		throw new UnsupportedOperationException();
 	}
 }

### 2017-02-18 paper/conclusion.tex

@@
 \subsection{Future Work}
-\todo{Jose: rewrite this}
-\ProjectName{} opens up multiple avenues for further work. For instance, we
-could extend our minimal/maximal imputation operators to consider global
-information, such as the specific columns needed in operators higher up in the
-query plan.  Our optimizer uses the same machine learning algorithm in all
-instances of imputation operators. Integrating multiple possible algorithms
-could allow for further fine-grained imputation, and honing the time complexity and
-loss of these algorithms for an iterator model database could facilitate more
-intelligent query plans and consistent interpretations of $\alpha$ across
-strategies. For example, algorithms that learn in an online manner could
-increase the efficiency of the system.  Furthermore, a multiple imputation
-strategy could be followed for queries involving certain aggregates. Finally,
-the underlying database for \ProjectName{} is far from a full-featured,
-production database, so a natural step is to integrate imputation planning in a
-widely-adopted production-quality database. This will likely expose further
-opportunities for improvement.
+\ProjectName{} opens up multiple avenues for further work. We sketch out some of these
+directions.
+
+\textbf{Cost model intepretation:} While the current system
+provides a parameter ($\alpha$) to tradeoff between imputation quality and runtime,
+there is no direct interpretation of its unit-less value. One natural extension to this work
+is to reformulate the cost-model in an interpretable fashion.
+
+\textbf{Imputation confidence:} Obtaining a confidence measure for query results
+produced when missing values are imputed with a machine learning model is
+another possible improvement to the system. In most cases, the data imputed
+is unlikely to follow a standard distribution, which implies that any confidence
+estimates will result from iterative approaches such as cross-validation\cite{kohavi1995study}.
+In contrast to standard confidence calculations, there is the added complexity
+of imputation taking place at various points in the plan and the composition of results
+through standard query operations.
+
+\textbf{Imputation operators:} We explored a subset of possible imputation operators.
+In particular, we explored the two variants of the drop and impute operations. The system
+could be extended to consider a broader family of operators. As the search space grows,
+there will likely be additional steps needed in order to avoid sub-optimal plans.
+
+\textbf{Adaptive query planning:} Currently, the imputation is local to a given query, meaning
+no information is shared across queries. An intriguing direction would be to take advantage
+of imputation and execution data generated by repeated queries. Note that this goes beyond
+simply caching imputation results, and instead could entail operations such as extending
+intermediate results with prior imputation values to improve accuracy, or pruning out
+query plans which have shown to produce low-confidence imputations.
 

### 2017-02-18 paper/experiments.tex

@@
 by freeCodeCamp as a part of a survey of new software developers
-(both professional and amateur). freeCodeCamp is an open-source
+(both professional and amateur)\cite{fcc-data}. freeCodeCamp is an open-source
 community that intends to help users learn how to program. In addition,

### 2017-02-18 paper/introduction.tex
paper: minor word changes

@@
 
-Handling incorrect or dirty data is a complex and challenging problem for data scientists.
+Handling incorrect or dirty data is a complex and challenging problem for analysts.
+According to some estimates, data scientists spend between 50 and 80 percent of their
+time performing tasks such as collecting and cleaning datasets\cite{data-science-cleaning}.
 One way in which a dataset can be dirty is for parts of it to be missing altogether.
@@
 Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
-Others may want to run queries on a subset of the data, so do not need to impute
+Others may want to run queries on a subset of the data, and so do not need to impute
 every field in every record. Yet others may want to customize the 

### 2017-02-18 paper/algorithm.tex
paper: added todos
paper: removed histogram helpers and tried to rework into simpler form
paper: fixed general typos, added clearer todos

@@
 The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
-Figure~\ref{fig:query-schematic} shows the layout of a query involving three tables, absent any imputation operators.
+Figure~\ref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.
 
@@
 
+\todobox{this section must be rewritten to show new approach}{
 For each sub-query, we keep track of the estimated number of
@@
 loss with a reasonable parameterization.
+}
 
@@
 
+\todobox{This seems tailored to decision trees, but we should talk more generally since
+we claim the imputation step isn't the focus}{
 Evaluation of the time complexity for \textit{Impute} is thornier, as the properties of
@@
 plans), though the range of $\alpha$ that does promote tradeoffs is condensed.
+}
 
@@
 
+\todo{formatting of plan cache semantics needs to be fixed}
 \begin{figure}
@@
 
+\todobox{reorder these and explain algorithm, which right now is cutoff}{
 The query planner (Algorithm~\ref{algo:top-level-planner}), operates as follows.
@@
 Algorithm~\ref{algo:plan-helpers} presents a series of helper functions used by the top level planner, shown in Algorithm~\ref{algo:top-level-planner}.
-
+}
 \begin{algorithm}
+\todo{fix input descriptions and small error in optjoin}
 \input{algorithms/plan-helpers.tex}
@@
 
-\begin{algorithm}
-\input{algorithms/histogram-helpers.tex}
-\end{algorithm}
+%\begin{algorithm}
+%\input{algorithms/histogram-helpers.tex}
+%\end{algorithm}
 
@@
 Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
-note that if we remove our restriction on types of imputation($\delta_{min}, \mu_{min}, \mu_{max}$), and allow any arbitrary subset of attributes to be imputed,
+note that if we remove our restriction on types of imputation (i.e. minimal vs maximal), and allow any arbitrary subset of attributes to be imputed,
 then every single operator in the query plan has a number of imputations exponential in the number of dirty columns. Our restriction, instead, increases the number

### 2017-02-18 paper/conclusion.tex
paper: fixed general typos, added clearer todos

@@
 We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
-the data exploration phase. The plans selected for each query result in execution that
-takes a fraction of the time compared to the standard approach of imputing on the entire base tables and
+the data exploration phase. The plans selected for each query execute in a fraction of the time compared to the standard approach of imputing on the entire base tables and
 then formulating queries. Furthermore, the difference in query results between the two approaches is
@@
 
-\textbf{Cost model intepretation:} While the current system
+\textbf{Cost model interpretation:} While the current system
 provides a parameter ($\alpha$) to tradeoff between imputation quality and runtime,

### 2017-02-18 paper/examples.tex
paper: removed search space size todo note in examples
paper: fix mistake in figure
paper: fixed general typos, added clearer todos

@@
 to get a quick and accurate view into the data. However,
-the data has been collected through CDC surveys (see Section~\ref{subsec:datasets} for details),
+the data has been collected through CDC surveys (see Sec.~\ref{subsec:datasets} for details),
 and there is a significant amount of missing data across all
@@
 a tradeoff between information loss and runtime cost, then finds
-the optimal query plan. Figure~\ref{fig:quality-plan} and Figure~\ref{fig:fast-plan} show what this plan would be
-when optimizing for quality and performance, respectively. These two plans are found within a broader search space
-of XXXX plans. \todo{add a small comment about the number of plans in the plancache
-to highlight size of search space}.
+the optimal query plan. Figures~\ref{fig:quality-plan} and ~\ref{fig:fast-plan} show what this plan would be
+when optimizing for quality and performance, respectively.
 
 The differences between the two plans highlight the subtleties of optimizing queries which contain imputation operators.
-Note that imputation operators are placed in different points of the plan, operating over different columns, and indeed
+Note that imputation operators are placed in different points of the plan, operating over different columns, and 
 even influencing the order of joins. The quality-optimized plan (Figure~\ref{fig:quality-plan}) uses the impute operation, rather
@@
             [.$\sigma_{\text{demo.gender} = 2}$ demo ]
-            [.$\sigma_{\text{exams.weight} \geq 120}$ [.$\mu_{\text{exams.weight}}$ exams ] ] ] labs ] ] ] ]
+            [.$\sigma_{\text{exams.weight} \geq 120}$ [.$\delta_{\text{exams.weight}}$ exams ] ] ] labs ] ] ] ]
 \caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for performance.}
@@
 Tens of queries later, the epidemiologist has the holistic view of the
-data that they required before carefully crafting their own tailored
+data that she required before carefully crafting her own tailored
 imputation model. Knowing that there may be a need to explore
-this data set further, they can easily incorporate their imputation model
+this data set further, she can easily incorporate their imputation model
 into ImputeDB for future use.

### 2017-02-18 paper/experiments.tex
paper: fixed general typos, added clearer todos

@@
 \section{Experiments}
-To evaluate the performance of our system, we generate and execute plans for queries
+To evaluate the performance of our approach,  we implemented
+a prototype system in Java, following a traditional iterator model.
+For our experiments we plan and execute queries
 for two separate survey-based data sets, showing that our system
@@
 
-The queries on the CDC NHANES data consists not only of projections and selections,
+The queries on the CDC NHANES data consist not only of projections and selections,
 but also interesting joins and aggregations. Our aim was to craft meaningful queries
@@
 
-Query~\ref{q1} calculates the average weight for users based on their income data.
+Queries 1-3 are on the CDC data. Query~\ref{q1} calculates the average weight for users based on their income data.
 To do so, it joins demographics and physical examination data. Query~\ref{q2}
@@
 
-In Query~\ref{q4}, we calculate the average income for survey participants,
+Queries 4-7 are on the FCC data. In Query~\ref{q4}, we calculate the average income for survey participants,
 based on their bootcamp attendance. Query~\ref{q5} estimates the average
@@
 \begin{table*}
+\todo{clean up formatting here}
 \centering
@@
 
-\todo{actually run experiments and rewrite this section}
+\todobox{actually run experiments and rewrite this section}{
 Table~\ref{fig:experiments} provides a summary of our experimental results.
@@
 Given the varying needs of database users, \ProjectName{} clearly allows flexibility without a significant performance hit.
+}
+
+\todo{Add timing comparison to ACS table since that should take longer than base on the data we have}
 

### 2017-02-18 paper/introduction.tex
paper: fixed general typos, added clearer todos

@@
 Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
-Although manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
+Although this manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
 Additionally, it may not even be necessary to completely clean the data to make it usable.
-Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
+Some users may be willing to run queries on dirty data, simply ignoring any missing values, as long as they do not have to pay the cost of imputation.
 Others may want to run queries on a subset of the data, and so do not need to impute
@@
 To achieve this goal, we perform imputation on the fly, during query execution.
-Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows users to flexibly trade accuracy for computation time.
+Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows users to flexibly trade imputation quality for computation time.
 

### 2017-02-18 paper/relatedwork.tex
paper: fixed general typos, added clearer todos

@@
 papers have described various (at times conflicting) treatments
-of nulls \cite{grant1977null}. \ProjectName's main design invariant - that no relational operator
-see attributes with missing values for attributes it must operate on and users should never see
+of nulls \cite{grant1977null}. \ProjectName's main design invariant - no relational operator
+sees missing values for attributes it must operate on and users should never see
 missing data - eliminates

### 2017-02-18 paper/algorithm.tex
paper: algorithm helpers clean up, add text about planner

@@
 
+\begin{figure}
 \todo{formatting of plan cache semantics needs to be fixed}
-\begin{figure}
+\todo{I think the cache update step is not quite right, since p seems to be a single plan, but it is really a set}
   \begin{align*}
@@
 
-\todobox{reorder these and explain algorithm, which right now is cutoff}{
-The query planner (Algorithm~\ref{algo:top-level-planner}), operates as follows.
-First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan.
-This set will be used to determine which attributes are imputed.
-Then, it constructs a plan cache
-
 To reduce the search space, we only consider the minimal imputation, the maximal imputation, and the minimal drop.
@@
 
-Algorithm~\ref{algo:plan-helpers} presents a series of helper functions used by the top level planner, shown in Algorithm~\ref{algo:top-level-planner}.
-}
+Algorithm~\ref{algo:plan-helpers} presents a series of helper functions used by the top-level planner, shown in Algorithm~\ref{algo:top-level-planner}.
+
+The planner operates as follows.
+First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan.
+This set will be used to determine which attributes are imputed.
+Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
+a search over imputation operations produces various possible plans, which are added to the cache. If no selections are available, a 
+simple scan is added to the plan cache. Join order is jointly optimized with imputation placements and a set of
+plans encompassing all necessary tables is produced. This set is now extended for any necessary grouping and aggregation
+operations. After a similar step has been taken for projections, the planner returns the cheapest plan from the cache.
+
 \begin{algorithm}
-\todo{fix input descriptions and small error in optjoin}
 \input{algorithms/plan-helpers.tex}

### 2017-02-18 paper/algorithms/plan-helpers.tex
paper: algorithm helpers clean up, add text about planner

@@
   \begin{algorithmic}
-    \Require{$q$ is a query plan. $C_l$ is a set of attributes that must be imputed in the output of this query plan. $C_g$ is the set of attributes which are used in the final plan.}
+    \Require{$q$ is a query plan, $C_l$ is a set of attributes that must be imputed in the output of this query plan, $C_g$ is the set of attributes which are used in the final plan.}
     \Ensure{A set of query plans $Q$ with added imputation.}
@@
 
-    \Require{$t$ is a table and $\phi$ is a filter predicate.}
+    \Require{$t$ is a table, $\phi$ is a filter predicate, $C_g$ is a set of attributes which are used in the final plan.}
     \Ensure{A set of plans for filtering $t$ with $\phi$.}
@@
 
-    \Require{$q$ is a query plan, $G$ is a set of grouping attributes, $A$ is an aggregation function, and $C_g$ is a set of visible attributes.}
+    \Require{$q$ is a query plan, $G$ is a set of grouping attributes, $A$ is an aggregation function, $C_g$ is a set of attributes which are used in the final plan.}
     \Ensure{A set of plans for grouping \& aggregating with $G$ and $A$.}
@@
 
-    \Require{A set of tables $T$, a plan cache $Q$ containing at least one query plan for each table in $T$, and a relation $J : T \times \Psi \times T$ between tables and join predicates.}
+    \Require{A set of tables $T$, a plan cache $Q$ containing at least one query plan for each table in $T$, and a relation $J : T \times \Psi \times T$ between tables and join predicates, $C_g$ is the set of attributes which are used in the final plan.}
     \Ensure{An updated plan cache $Q'$ which contains at least one plan that combines all of the tables in $T$.}
@@
     \State $C \gets \Call{Attrs}{\psi}$
-    \State $L \gets \{\Call{AddImpute}{q, C, C_g} ~|~ q \in Q[t] \}$
-    \State $R \gets \{\Call{AddImpute}{q, C, C_g} ~|~ q \in Q[S'] \}$
+    \State $L \gets \{\Call{AddImpute}{q, C, C_g} ~|~ q \in Q[S'] \}$
+    \State $R \gets \{\Call{AddImpute}{q, C, C_g} ~|~ q \in Q[t] \}$
     \State $Q[S] \lhd \{l \bowtie_\psi r ~|~ l \in L, r \in R\}$

### 2017-02-18 paper/experiments.tex
paper: format queries
paper: added acs example to show base table imputation infeasible

@@
 
+For our final experiment, we introduce a simple aggregate query over data from the American Community Survey (ACS), which
+provides a number of public data sets collected by the U.S. Census Bureau.
+We used a cleaned version of the 2012 Public Use Microdata Sample (PUMS) data kindly provided by the authors of~\cite{akande2015empirical}.
+Given that the data had been cleaned, we uniformly introduced missing values across 40 percent of the values. The final dataset consists
+of 671,153 rows and 37 integer columns.
+
 \subsection{Queries}
@@
 \begin{table*}
-\todo{clean up formatting here}
 \centering
@@
 
-\todo{Add timing comparison to ACS table since that should take longer than base on the data we have}
+In many cases, applying the imputation step at the base table simply is not an option. To illustrate the increasing difficulty of such an approach
+as datasets scale, we used a simple aggregate query over the ACS dataset (Figure~\ref{query-acs}). Applying the imputation operation to the base table
+is extremely expensive, in our case the process did not terminate within 20 minutes. In contrast, \ProjectName{} executes a quality-optimized version
+in 9 seconds and a runtime-optimized version in 1 second. This highlights the potential benefit of using our system for early data exploration.
+
+\begin{figure}
+\begin{lstlisting}
+SELECT AVG(c0) FROM acs_dirty;
+\end{lstlisting}
+\caption{A simple aggregate query on ACS data. Imputing the base table is unfeasible (did not terminate within 20min). \ProjectName's
+quality-optimized and performance-optimized query execute in 9 seconds and 1 second, respectively.
+}
+\label{query-acs}
+\end{figure}
 

### 2017-02-18 simpledb/data.py

@@ -98,12 +98,28 @@ gdp_columns = {
   '2015 [YR2015]' : 'gdp_per_capita'
 }
 
-
-
 def read_csv(path, col_map):
   # avoid deciding type before, these files are dirty and small
   df = pd.read_csv(path, low_memory=False)
-  return df[col_map.keys()].rename(columns = col_map)
+  if not col_map is None:
+    df = df[col_map.keys()].rename(columns = col_map)
+  else:
+    # otherwise just name as c0 .... cn
+    cols = ["c%i" % c for c in range(df.shape[1])]
+    df.columns = cols
+  return df
+
+def add_nulls(df, pct):
+  print "Adding %f pct nulls" % pct
+  if pd.isnull(df).any().any():
+    return df
+  else:
+    vals = df.values.flatten().copy().astype(float)
+    n = len(vals)
+    ix = np.random.choice(n, size=int(pct * n), replace=False)
+    vals[ix] = np.nan
+    vals = vals.reshape(df.shape)
+    return pd.DataFrame(vals, columns=df.columns)
 
 def get_float_cols(df):
   return df.columns[df.dtypes == float].tolist()
@@ -266,6 +282,8 @@ def to_simpledb(dfs, prec, jar, path, suffix):
     df = float_fixed_precision(df, prec)
     # apply enumerations
     df = apply_enum(df, enums)
+    # if has no nulls naturally, then add some
+    df = add_nulls(df, 0.4)
     csv_path = os.path.join(path, name + suffix + '.csv')
     # no suffix for the dat files, since those need to match schema name
     dat_path = os.path.join(path, name + '.dat')
@@ -280,11 +298,12 @@ def to_simpledb(dfs, prec, jar, path, suffix):
   
 def read_csv_data(path):
   info  = {
-    'demo'  : ('demographic.csv', demo_columns),      # cdc data
-    'exams' : ('examination.csv', exams_columns),     # cdc data
-    'labs'  : ('labs.csv', labs_columns),             # cdc data
-    'fcc'   : ('fcc.csv', fcc_columns),               # fcc data
-    'gdp'   : ('gdp.csv', gdp_columns)                # gdp data
+    'demo'      : ('demographic.csv', demo_columns),      # cdc data
+    'exams'     : ('examination.csv', exams_columns),     # cdc data
+    'labs'      : ('labs.csv', labs_columns),             # cdc data
[…]

### 2017-02-18 paper/algorithm.tex
paper: clean up some TODOs for cost model

@@
 \label{sec:cost-model}
-\todo{Explain that loss is now averaged}
-\todo{Do we want to keep the behavior that imputation increases in quality with the number of cells, or switch to number of tuples.}
-We rank query plans using a cost model \[\textsc{Cost}(q) = (1 - \alpha) \times \textsc{Time}(q) + \alpha \times \textsc{Loss}(q).\] $\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
-$\textsc{Loss}(q)$ is an estimate of the amount of error introduced by the imputation procedure.
+We rank query plans using a cost model \[\textsc{Cost}(q) = (1 - \alpha) \times \textsc{Time}(q) + \alpha \times \frac{\Sigma_i \textsc{Loss}(q)_i}{|\textsc{Loss}(q)|}.\] $\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
+$\textsc{Loss}(q)$ is an estimate of the amount of error introduced by a sequence of imputation procedure.
+Our
+cost model averages over the losses from each such operator. 
 $\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance.
@@
 
-\todobox{this section must be rewritten to show new approach}{
 For each sub-query, we keep track of the estimated number of
 null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
-the sub-query. We use this estimate to compute
-$\text{Loss}(Q)$ as follows.
+the sub-query. We use this estimate to compute the set $\text{Loss}(Q)$ as follows.
 
@@
   \textsc{Loss}(q) = \begin{cases}
-    \sum_{c \in C} \textsc{Missing}(c, q') & q = \delta_C(q') \\
-    \frac{1}{\sqrt{|q'|}} \sum_{c \in C} \textsc{Missing}(c, q') & q = \mu_C(q') \\
-    \textsc{Loss}(q_1') + \textsc{Loss}(q_2') & q = q_1' \bowtie_\psi q_2' \\
+    \{1\} \cup \textsc{Loss}(q') & q = \delta_C(q') \\
+    \{\frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Cardinality}(q') }} \} \cup \textsc{Loss}(q')  & q = \mu_C(q') \\
+    \textsc{Loss}(q_1') \cup \textsc{Loss}(q_2') & q = q_1' \bowtie_\psi q_2' \\
     \textsc{Loss}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
@@
 \end{align*}
-Dropping tuples which contain a null value incurs a loss of 1 for each null field dropped.
-Imputing null fields incurs a loss penalty $p \in (0, 1]$ which decreases as the number of
-tuples available to train the imputation algorithm increases.  Intuitively, imputation
+
+Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
+the dropped tuples might have encoded). Imputing null fields incurs a loss penalty $p \in (0, 1]$,
+which is inversely proportional to the number of total cells available to train the imputation algorithm. 
+Total cells are defined as the product of the number of attributes and the number of tuples.
+Intuitively, imputation
 accuracy \textit{increases} when more complete attributes and complete tuples are available;
-correspondingly, the information loss \textit{decreases} with more complete values to work
-with. The inverse square loss can be seen as a heuristic first approximation to the learning guarantees of
-regression trees. Since we conceive that
+correspondingly, the information loss \textit{decreases} with more values to work
+with. Since we conceive that
 alternate imputation strategies could be used, the analyst may provide a corresponding
 loss with a reasonable parameterization.
-}
 
 To compute $\text{Time}(Q)$, we retain a simple set of heuristics used by a standard 
-database for the time. Scan costs are estimated based on the number of tuples, 
+database. Scan costs are estimated based on the number of tuples, 
 page size, and an unit-less IO cost per page. Join costs are estimated as a function
@@
 
-\todobox{This seems tailored to decision trees, but we should talk more generally since
-we claim the imputation step isn't the focus}{
-Evaluation of the time complexity for \textit{Impute} is thornier, as the properties of
-different decision tree building algorithms can vary significantly, the underlying IO cost
+The time computation for \textit{Impute} depends on the properties of the underlying
+algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
+and simply queries it for a time estimate. In the base implementation of our system,
+we use an iterative algorithm (see Sec.~\ref{sec:imputation}), where the runtime cost
+is a function of the number of attributes with no missing data, number of attributes with
+missing data being imputed, number of tuples, and the number of iterations for the algorithm.
+
+In order to avoid having the time component of our cost model dominate the information loss 
+component, we scale the former by an arbitrary constant.
+
+In general, evaluating the time complexity for different \textit{Impute} operators is
+complicated, as the properties of
+different algorithms can vary significantly, the underlying IO cost
 is not easily extracted (especially in the case that the entire stream of tuples does not
@@
 practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
-plans), though the range of $\alpha$ that does promote tradeoffs is condensed.
-}
+plans).
 

### 2017-02-21 simpledb/src/java/simpledb/ExperimentRunner.java
fix: output results on each iteration of experiments

@@ -127,11 +127,8 @@ public class ExperimentRunner {
         // write out times
         writeTime(id, alpha, iter, Duration.between(planStart, planEnd).toMillis(), Duration.between(runStart, runEnd).toMillis(), plan.hashCode());
 
-        // if its the first time, go ahead and write out results and plan
-        if (iter == 0) {
-            writeResult(id, alpha, results.toString());
-            writePlan(id, alpha, plan);
-        }
+        writeResult(id, alpha, results.toString());
+        writePlan(id, alpha, plan);
     }
 
     private String[] getQueries() throws IOException {

### 2017-02-21 simpledb/analyze_experiments.py
feat: added script to go through experiment results

@@ -1,0 +1,160 @@
+#!/usr/local/bin/python
+
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+import os.path
+import sys
+from StringIO import StringIO
+from collections import defaultdict
+
+if len(sys.argv) != 4:
+  print "Usage: python analyze_experiments.py <experiment-results-dir> <base-results-dir> <output-dir>"
+  sys.exit(1)
+
+experiments_dir = sys.argv[1]
+base_dir = sys.argv[2]
+output_dir = sys.argv[3]
+
+# basic utils
+def drop_warmup(df, by, keep=5):
+  return df.groupby(by).tail(keep)
+
+def summarize(df, by, a):
+  ops = {'mean': np.mean, 'std': np.std}
+  return df.groupby(by)[a].agg(ops).reset_index()
+  
+def plot(df,**kwargs):
+  _, ax = plt.subplots(1)
+  for a in set(df['alpha']):
+    data = df[df['alpha'] == a]
+    try:
+      data.plot(ax=ax, label=r'$\alpha=%.0f$' % a, **kwargs)
+    except TypeError:
+      data.plot(ax=ax, label=a, **kwargs)
+    # get around bug in legend drawing in pandas
+    plt.legend()
+  return ax
+  
+def get_query_results(path, headers):
+  lines = []
+  with open(path, 'r') as f:
+    lines = f.readlines()
+  lines = [l.strip() for l in lines]
+  lines = [l for l in lines if len(l) > 0]
+  results = defaultdict(lambda: [])
+  i = 0
+  # consume all results
+  while i < len(lines):
+    line = lines[i]
+    start_signal = 'alpha'
+    end_signal = 'query'
+    if line[:len(start_signal)].lower() == start_signal:
+      alpha = float(line.split(' ')[1])
+      query = int(lines[i-1].split(' ')[1])
+      data = []
+      # skip over Results:
+      i += 2
+      next_line = lines[i]
+      # consume current result as df
+      while len(next_line) > 0 and next_line[:len(end_signal)].lower() != end_signal:
+        # append raw record
+        data.append(next_line)
+        i += 1
+        if i < len(lines):
+          next_line = lines[i]
+        else:
+          next_line = []
+      if data:
+        str_data = '\n'.join(data)
+        buf = StringIO(str_data)
[…]

### 2017-02-21 paper/experiments.tex
paper: added some preliminary experiments

@@
 
-\begin{table*}
-  \centerfloat
-  \input{runtimes}
-    \caption{Base error, percent change in error and and running time for queries
-    with different imputation levels. Base error is the root-mean-square error (RMSE) between the query run on clean
-    data and the query run on dirty data without imputation. Change in error is relative to the base error.}
-  \label{fig:experiments}
-\end{table*}
+%\begin{table*}
+%  \centerfloat
+%  \input{runtimes}
+%    \caption{Base error, percent change in error and and running time for queries
+%    with different imputation levels. Base error is the root-mean-square error (RMSE) between the query run on clean
+%    data and the query run on dirty data without imputation. Change in error is relative to the base error.}
+%  \label{fig:experiments}
+%\end{table*}
 
-\todobox{actually run experiments and rewrite this section}{
-Table~\ref{fig:experiments} provides a summary of our experimental results.
-We run each query in four configurations: without imputation on clean data, without imputation on dirty data, and with imputation on dirty data for $\alpha = 0$ and $\alpha = 1$.
-When a query returns a relation of the same ``shape'' (same number of tuples and same schema) on both clean and dirty data, we can compute an error value.
-For our experiments, we report root-mean-square error (RMSE) for the base error (i.e. the error in the dirty data without imputation with respect to the clean data) and the change
-in RMSE for the imputed plans.
-For imputations, we run queries using the two extreme values of $\alpha$: 0 and 1, representing a focus on runtime and on information loss, respectively.
-For each query, the table displays the base error, the change in error in both imputation scenarios, and the running time in seconds.
+%\todobox{actually run experiments and rewrite this section}{
+All experiments were run on a Macbook Pro with 2 2.9 GHz Intel core i5 processors, 16 GB of main memory,  running OS X El Capitan.
 
-Planning times are not shown, as they are fractions of a second for all of the queries that we tested.
-The overall low planning times highlight the practicality of our algorithm, despite the exponential complexity.
-Running times clearly vary based on $\alpha$, allowing users to tune their query performance as desired.
+Figure~\ref{fig:runtimes} provides a summary of the performance results. Each query is run in two configurations: quality-optimized and runtime-optimized.
+To provide comparison, we also include a table with the average running time required to perform imputation at each of the base tables used in our analysis.
+Note that multiple queries perform joins between two or more base tables. In all cases, running a query through \ProjectName{} is cheaper than 
+imputing a single base table, with the added flexibility of imputation tailored to the query.
 
-We can group our queries into five categories: sum/count queries, average queries, min/max queries, selection queries, and join queries on synthetic data.
+\begin{figure}
+\begin{subfigure}{\linewidth}
+\includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
+\caption{Query runtimes with \ProjectName{}}
+\label{subfig:project-runtime-queries}
+\end{subfigure}
+~
+\begin{subfigure}{\linewidth}
+\includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
+\caption{Query runtimes with imputation on base tables}
+\label{subfig:base-runtime-queries}
+\end{subfigure}
+\caption{Comparing \ProjectName{} runtimes for each query (\ref{subfig:project-runtime-queries}) with the runtime for a query when base-table imputations are used (\ref{subfig:base-runtime-queries})}
+\end{figure}
 
-\begin{itemize}
-\item For sum and count queries (\ref{q2}, \ref{q3}, \ref{q5}), imputation does a good job of reducing the error.
-For each of these queries, we can reduce total error by several orders of magnitude.
-Imputation performs very well in these cases, even for data which is missing uniformly at random, because each missing value directly increases the error of the result.
+Figure~\ref{fig:plantimes} provides a summary of the planning times for each of the queries.  We exclude the planning time for queries that impute at base table, as that
+requires no planning. On average, planning constituted 38 percent of total runtime for plans generated by \ProjectName{}.
 
-\item For average queries (\ref{q4}, \ref{q6}, \ref{q7}), imputation does not reduce the error, mainly because running these queries on dirty data incurs only a small penalty.
-This is because, in the case of average, removing data uniformly at random does not change the mean.
-If we were to remove data in a biased way, imputation would likely perform better because there would be more error for it to correct.
+\begin{figure}
+\includegraphics[width=\columnwidth]{figures/planning_times_imputedb.png}
+\caption{Planning times for each query}
+\label{fig:plantimes}
+\end{figure}
 
-\item For a min query (\ref{q8}), it is unlikely that imputation will ever help, because by definition these queries are looking for extreme values.
-These values are unlikely to be the result of imputation and are also unlikely to be randomly removed from the dataset.
+Table~\ref{table:rmse} shows the Root-Mean-Squared-Error (RMSE) for \ProjectName{}'s query results when compared to running imputation on the base tables and executing the query on the
+cleaned data. Each query was run 10 iterations in both settings and then results from the two approaches were paired up and compared tuple-wise. We average the
+squared error across a single query before applying square root. We can see that optimizing for quality indeed reduces the RMSE of query results. In general, the RMSE relative to the
+base-imputation approach are acceptable.
 
-\item We do not compute error for selection queries (\ref{q1}, \ref{q9}, \ref{q10}, \ref{q12}), because the output of the query on clean data is not directly comparable to the output on dirty data.
-
-  \item Imputation performs poorly on the join queries on synthetic data (\ref{q11}, \ref{q13}) because the data is synthetic, so there are no correlations for imputation to find.
-We included these queries because we wanted to test the performance of our algorithm on queries that include joins, and we were pleased to find that simple joins cause no perceptible performance problems.
-\end{itemize}
-
-Finally, we performed CE-CART imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
-We were unable to complete the full imputation due to resource constraints, but the process ran for more than one hour without terminating.
-This time compares favorably to our maximum of 72 seconds for a query with imputation on the ACS data.
-Given the varying needs of database users, \ProjectName{} clearly allows flexibility without a significant performance hit.
-}
+\begin{table}
+\input{perf_summary.tex}
+\caption{Root-Mean-Squared-Error for queries run under different $\alpha$ parameterizations relative to results when imputing on base table}
+\label{table:rmse}
+\end{table}
 

### 2017-02-22 simpledb/src/java/simpledb/ExperimentRunner.java
fix: change cost model to use pareto frontier, pick final plan based on alpha-bound

@@ -48,7 +48,7 @@ public class ExperimentRunner {
 
     public ExperimentRunner(int iters, String catalog, String queries, String outputDir)
             throws IOException {
-        this(true, -1.0, -1.0, 1.0, iters, catalog, queries, outputDir);
+        this(true, 0.0, 0.0, 1.0, iters, catalog, queries, outputDir);
     }
 
     public ExperimentRunner(double minAlpha, double maxAlpha, double step, int iters, String catalog, String queries, String outputDir)

### 2017-02-22 simpledb/src/java/simpledb/ImputedPlanCache.java

@@ -12,4 +12,6 @@ public interface ImputedPlanCache {
 	void addPlan(Set<String> ts, ImputedPlan newPlanIP);
 
 	void addJoinPlan(Set<String> ts, Set<LogicalJoinNode> joins, ImputedPlan newPlanIP);
+
+	ImputedPlan getFinalPlan(double lossWeight, Set<String> tables);
 }

### 2017-02-22 paper/algorithm.tex
paper: rework algorithm and conclusion to account for new cost model

@@
 \label{sec:cost-model}
-We rank query plans using a cost model \[\textsc{Cost}(q) = (1 - \alpha) \times \textsc{Time}(q) + \alpha \times \frac{\Sigma_i \textsc{Loss}(q)_i}{|\textsc{Loss}(q)|}.\] $\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
+The cost for a plan $q$ is expressed as a tuple of $\langle \textsc{Loss}(q), \textsc{Time}(q) \rangle$.
+$\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
 $\textsc{Loss}(q)$ is an estimate of the amount of error introduced by a sequence of imputation procedure.
 Our
-cost model averages over the losses from each such operator. 
-$\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance.
-$\alpha = 1.0$ means that the optimal query should have the highest quality imputation (at the expense of performance), $\alpha=0.0$ means that the optimal query should be as fast as possible (at the expense of imputation quality).
+cost model averages over the losses from each such operator. We note that our loss calculation constrains $\textsc{Loss}(q)$ to $[0, 1]$, while $\textsc{Time}(q)$ can range within $(0, \infty)$.
+
+\begin{definition}
+We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$, if $l < l' \land t < t'$, and denote this as $\doms{p}{p'}$. 
+\end{definition}
+
+Given a query, \ProjectName{} produces a final set of possible plans
+
+\begin{gather*}
+ P= \{ p | \forall p'. \doms{p}{p'} \}
+ \end{gather*}
+
+This set outlines the best possible options for trading performance and cost for the current query, building a Pareto frontier\cite{pareto1964cours}.
+
+In order to pick a final plan from the frontier, our model introduces a parameter $\alpha$ which encodes an upper bound on the amount of information loss that a user is willing to tolerate
+relative to the minimum information loss possible given the imputation operations.
+
+\begin{definition}
+Given a final set of plans $P$, and $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
+We say $p$ is $\alpha$-bound optimal if $\forall p'^\alpha \in P: \textsc{Time}(p) \leq \textsc{Time}(p')$. 
+\end{definition}
+
+Given $\alpha$ and $P$, \ProjectName{} returns the $\alpha$-bound optimal plan in $P$. Note that the $\alpha$-bound optimal plan is unique, given that
+$P$ only maintains dominating plans.
+
+So in essence, $\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance, as plans that
+lose a lot of information through dropping are also the fastest.
+
+$\alpha = 1.0$ means that the optimal query should have the highest performance (at the expense of quality), while $\alpha=0.0$ means that the optimal query should lose
+as little information as possible (at the expense of performance).
 
@@
 Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
-At a high level, a plan cache is a mapping from a set of tables to a set of optimal plans that perform the required joins and filters over these tables.
-We store multiple plans rather than a single best plan because each plan has a distinct set of dirty attributes.
+At a high level, a plan cache is a mapping from a set of tables to a set of dominating plans that perform the required joins and filters over these tables.
+We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes and a choice over the plan frontier
+can only be made once the set is final. Details are shown in Figure~\ref{fig:semantics-plan-cache}.
 
 \begin{figure}
-\todo{formatting of plan cache semantics needs to be fixed}
-\todo{I think the cache update step is not quite right, since p seems to be a single plan, but it is really a set}
   \begin{align*}
@@
     \end{cases} \\
-    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned}
-                                           Q \cup \{(T,\ \argmin_{q \in P}\ \textsc{Cost}(q))\}\ \text{where} \\
-                                           P = \left\{q ~|~ \begin{aligned}
+    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
+                                            \text{ where } \\
+                                           P' &= \left\{q ~|~ \begin{aligned}
                                                                      q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
-                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p) 
+                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
+                                                                     \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
                                                                    \end{aligned}
                                            \right\}
-                                         \end{aligned}
-    % \in 
   \end{align*}
   \caption{Semantics of the plan cache.}
+  \label{fig:semantics-plan-cache}
 \end{figure}

### 2017-02-22 paper/algorithms/plan-toplevel.tex
paper: keep refining algorithm section
paper: rework algorithm and conclusion to account for new cost model

@@
 % \begin{algorithm}
-\begin{algorithmic}
-  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$, and an optional set of grouping attributes $G$ and aggregator function $A$.}
+\begin{algorithmic}[5]
+  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses trade-off performance and information loss}
   \Ensure{An optimized query plan.}
@@
   \Function{Plan}{$T, F, J, P, G, A$}
+  \State $C_g \gets \bigcup_{\phi \in F} \Call{Attr}{\phi}$ \Comment{Collect relevant attributes.} \label{lst:line:attr-start}
+  \State $C_g \gets \bigcup_{\psi \in J} \Call{Attr}{\psi} \cup P \cup G \cup \Call{Attr}{A} \cup C_g$ \label{lst:line:attr-end}
   \State Let $Q$ be an empty plan cache.
@@
   \If {$\exists \phi: (t, \phi) \in F$}
-  \State $Q[\{t\}] \lhd \Call{OptFilter}{Q, t, \phi, P}$
+  \State $Q[\{t\}] \lhd \Call{OptFilter}{Q, t, \phi, C_g}$
   \Else
@@
 
-  \State $\Call{OptJoin}{Q, T, J, P}$	\Comment{Optimize joins.}
+  \State $\Call{OptJoin}{Q, T, J, C_g}$	\Comment{Optimize joins.}
   \State $B \gets Q[T]$ \Comment{Get the best plans for all tables.}
   \If{$G \neq \emptyset$} \Comment{Add optional group \& aggregate.}
-  \State $B \gets \bigcup_{q \in B} \Call{OptGroupBy}{q, G, A, C_{vis}}$
+  \State $B \gets \bigcup_{q \in B} \Call{OptGroupBy}{q, G, A, C_g}$
   \EndIf
@@
   \State $B \gets \bigcup_{q \in B} \Call{OptProject}{q, P}$ \Comment{Add projections.}
-
-  \State \Return $\argmin_{q \in B} Cost(q)$ \Comment{Return cheapest plan.}
+  \State \Return $p \in B$ s.t. $p$ is $\alpha$-bound optimal  \Comment{Return fastest plan satisfying $\alpha$ bound.}
   \EndFunction

### 2017-02-22 paper/conclusion.tex
paper: rework algorithm and conclusion to account for new cost model

@@
 
-\textbf{Cost model interpretation:} While the current system
-provides a parameter ($\alpha$) to tradeoff between imputation quality and runtime,
-there is no direct interpretation of its unit-less value. One natural extension to this work
-is to reformulate the cost-model in an interpretable fashion.
-
 \textbf{Imputation confidence:} Obtaining a confidence measure for query results
@@
 intermediate results with prior imputation values to improve accuracy, or pruning out
-query plans which have shown to produce low-confidence imputations.
+query plans which have been shown to produce low-confidence imputations.
 

### 2017-02-22 paper/algorithm.tex
fix: small fix for dominating definition

@@
 
-Given a query, \ProjectName{} produces a final set of possible plans
+Given a query, \ProjectName{} produces a final set of possible plans $P$ by only maintaining plans that are not dominated by others in the search space. So if
+our search space is $S$
 
 \begin{gather*}
- P= \{ p | \forall p'. \doms{p}{p'} \}
+ P= \{ p | \not\exists p' \in S. \doms{p'}{p} \}
  \end{gather*}
@@
 
-Given $\alpha$ and $P$, \ProjectName{} returns the $\alpha$-bound optimal plan in $P$. Note that the $\alpha$-bound optimal plan is unique, given that
-$P$ only maintains dominating plans.
+Given $\alpha$ and $P$, \ProjectName{} returns the $\alpha$-bound optimal plan in $P$. Note that the $\alpha$-bound optimal plan is unique in $P$, given that
+$P$ does not keep any plans dominated by others in the set.
 

### 2017-02-22 paper/experiments.tex
paper: minor typos in experiments
fix: use smape instead of rmse

@@
 \caption{Comparing \ProjectName{} runtimes for each query (\ref{subfig:project-runtime-queries}) with the runtime for a query when base-table imputations are used (\ref{subfig:base-runtime-queries})}
+\label{fig:runtimes}
 \end{figure}
@@
 
-Table~\ref{table:rmse} shows the Root-Mean-Squared-Error (RMSE) for \ProjectName{}'s query results when compared to running imputation on the base tables and executing the query on the
-cleaned data. Each query was run 10 iterations in both settings and then results from the two approaches were paired up and compared tuple-wise. We average the
-squared error across a single query before applying square root. We can see that optimizing for quality indeed reduces the RMSE of query results. In general, the RMSE relative to the
-base-imputation approach are acceptable.
+Table~\ref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to 
+running imputation on the base tables and executing the query on the
+cleaned data. Each query was run 10 iterations in both settings and then results from the two approaches were paired up and compared tuple-wise. We average 
+absolute percentage deviations across each iteration of a query, and we report this value averaged over all iterations.
+We can see that optimizing for quality indeed reduces the SMAPE of query results. In general, the SMAPE relative to the
+base-imputation approach are low in all cases.
 
 \begin{table}
+\centering
 \input{perf_summary.tex}
-\caption{Root-Mean-Squared-Error for queries run under different $\alpha$ parameterizations relative to results when imputing on base table}
-\label{table:rmse}
+\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$ parameterizations relative to results when imputing on base table}
+\label{table:smape}
 \end{table}
@@
 \end{lstlisting}
-\caption{A simple aggregate query on ACS data. Imputing the base table is unfeasible (did not terminate within 20min). \ProjectName's
+\caption{A simple aggregate query on ACS data. Imputing the base table is undesirable (did not terminate within 20min). \ProjectName's
 quality-optimized and performance-optimized query execute in 9 seconds and 1 second, respectively.

### 2017-02-22 paper/algorithm.tex
paper: some typos, clarifications, fixes

@@
 
-Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans containing imputation operators (Sec.~\ref{sec:operators}) for a plan that minimizes a metric (Sec.~\ref{sec:cost-model}) which combines the runtime cost of the query and the estimated quality of the results.
+Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans containing imputation operators (Sec.~\ref{sec:operators}) for an optimal plan (Sec.~\ref{sec:cost-model}) given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.
+
 This plan is subject to two constraints (Sec.~\ref{sec:placement}).
@@
 The cost for a plan $q$ is expressed as a tuple of $\langle \textsc{Loss}(q), \textsc{Time}(q) \rangle$.
-$\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
-$\textsc{Loss}(q)$ is an estimate of the amount of error introduced by a sequence of imputation procedure.
-Our
-cost model averages over the losses from each such operator. We note that our loss calculation constrains $\textsc{Loss}(q)$ to $[0, 1]$, while $\textsc{Time}(q)$ can range within $(0, \infty)$.
+$\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of any query predicates.
+$\textsc{Loss}(q)$ is an estimate of the amount of information lost through the sequence of imputation procedures used.
+We note that our loss calculation constrains $\textsc{Loss}(q)$ to $[0, 1]$, while $\textsc{Time}(q)$ can range within $(0, \infty)$.
 
@@
 
-$\alpha = 1.0$ means that the optimal query should have the highest performance (at the expense of quality), while $\alpha=0.0$ means that the optimal query should lose
+$\alpha = 1.0$ means that the optimal query should have the fastest runtime (at the expense of quality), while $\alpha=0.0$ means that the optimal query should lose
 as little information as possible (at the expense of performance).
 
-The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of $q$ and its sub-queries.
+The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
 These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
@@
 
-For each sub-query, we keep track of the estimated number of
+For each sub-plan, we keep track of the estimated number of
 null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
-the sub-query. We use this estimate to compute the set $\text{Loss}(Q)$ as follows.
+that plan. We use this estimate to compute $\text{Loss}(Q)$ as follows.
 
 \begin{align*}
-  \textsc{Loss}(q) = \begin{cases}
+  \textsc{Losses}(q) &= \begin{cases}
     \{1\} \cup \textsc{Loss}(q') & q = \delta_C(q') \\
-    \{\frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Cardinality}(q') }} \} \cup \textsc{Loss}(q')  & q = \mu_C(q') \\
-    \textsc{Loss}(q_1') \cup \textsc{Loss}(q_2') & q = q_1' \bowtie_\psi q_2' \\
-    \textsc{Loss}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
-  \end{cases}
+    \{\frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Cardinality}(q') }} \} \cup \textsc{Losses}(q')  & q = \mu_C(q') \\
+    \textsc{Losses}(q_1') \cup \textsc{Losses}(q_2') & q = q_1' \bowtie_\psi q_2' \\
+    \textsc{Losses}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
+  \end{cases}\\
+  \textsc{Loss}(q) &= \frac{\Sigma_{l \in \textsc{Losses}(q)} l}{| \textsc{Losses}(q)|}
 \end{align*}
@@
 missing data being imputed, number of tuples, and the number of iterations for the algorithm.
-
-In order to avoid having the time component of our cost model dominate the information loss 
-component, we scale the former by an arbitrary constant.
 
@@
   \end{align*}
-  \caption{Semantics of the plan cache.}
+  \caption{Semantics of a plan cache $Q$}
   \label{fig:semantics-plan-cache}
@@
 plans encompassing all necessary tables is produced. This set is now extended for any necessary grouping and aggregation
-operations. After a similar step has been taken for projections, the planner returns the cheapest plan from the cache.
+operations. After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
+best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
+$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.
 
@@
 manipulation. By removing concerns for imputation, users executing queries in \ProjectName{}
-don't need to consider the potential nature of missing data.
+do not need to consider the potential nature of missing data.
 
@@
 Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
-note that if we remove our restriction on types of imputation (i.e. minimal vs maximal), and allow any arbitrary subset of attributes to be imputed,
+note that if we remove our restriction on imputation operators (i.e. minimal vs maximal), and allow any arbitrary subset of attributes to be imputed,
 then every single operator in the query plan has a number of imputations exponential in the number of dirty columns. Our restriction, instead, increases the number

### 2017-02-22 paper/conclusion.tex
paper: some typos, clarifications, fixes

@@
 then formulating queries. Furthermore, the difference in query results between the two approaches is
-shown to be minimal. Data analysts need not commit to a specific imputation strategy and can instead
+shown to be small in all queries considered. Data analysts need not commit to a specific imputation strategy and can instead
 vary this across queries.
@@
 Similarly to existing work, we consider the impact of null values in databases and develop a simple set of invariants to 
-successfully plan around them. In contrast to the statistical learning work, our emphasis is not on the specific algorithm
-used to impute but rather on the timing of imputation in the execution of a query. In contrast to existing database work,
+successfully plan around them. In contrast to existing statistical learning work, our emphasis is not on the specific algorithm
+used to impute but rather on the placement of imputation steps in the execution of a query. In contrast to existing database work,
 we incorporate imputation into a cost-based optimizer and hide any details
@@
 \textbf{Imputation operators:} We explored a subset of possible imputation operators.
-In particular, we explored the two variants of the drop and impute operations. The system
+In particular, we explored two variants of the \textit{Drop} and \textit{Impute} operations. The system
 could be extended to consider a broader family of operators. As the search space grows,

### 2017-02-22 paper/experiments.tex
paper: some typos, clarifications, fixes

@@
 For our first set of experiments, we use survey data collected by the 
-Center for Disease Control (CDC) in the United States. In particular, we
+Center for Disease Control and Prevention (CDC) in the United States. In particular, we
 experiment on a set of tables collected as part of the National
@@
 All tables used consists of 10 attributes. The laboratory and examinations data,
-referred to as \textit{labs} and \textit{exams} in our queries, consists of 9813 rows,
-while the demographics data (\textit{demo}) has 10175 rows.
+referred to as \textit{labs} and \textit{exams} in our queries, consist of 9813 rows,
+while the demographics data (\textit{demo}) have 10175 rows.
 
@@
 To do so, it joins demographics and physical examination data. Query~\ref{q2}
-compares cholesterol levels for individuals who are both ends of the income
+compares cholesterol levels for individuals who are at both ends of the income
 spectrum and who satisfy a minimum weight requirement. Query~\ref{q3}
@@
 based on their bootcamp attendance. Query~\ref{q5} estimates the average
-age of women from the Uniter States who participated. Query~\ref{q6} 
+age of women from the United States who participated. Query~\ref{q6} 
 calculates the average amount of money survey participants with student debt destined to
@@
 %\todobox{actually run experiments and rewrite this section}{
-All experiments were run on a Macbook Pro with 2 2.9 GHz Intel core i5 processors, 16 GB of main memory,  running OS X El Capitan.
+All experiments were run on a Macbook Pro with two 2.9 GHz Intel core i5 processors, 16 GB of main memory,  running OS X El Capitan.
 
-Figure~\ref{fig:runtimes} provides a summary of the performance results. Each query is run in two configurations: quality-optimized and runtime-optimized.
-To provide comparison, we also include a table with the average running time required to perform imputation at each of the base tables used in our analysis.
-Note that multiple queries perform joins between two or more base tables. In all cases, running a query through \ProjectName{} is cheaper than 
-imputing a single base table, with the added flexibility of imputation tailored to the query.
+Figure~\ref{fig:runtimes} provides a summary of the performance results. The first plot shows average running times for \ProjectName{}
+plans in two configurations: quality-optimized and runtime-optimized. The second plot shows, for comparison, the 
+average running times required for each query when performing imputation at the base tables.
+In all cases, running a query through \ProjectName{} is cheaper than 
+imputing a single base table, with the added advantage of imputation tailored to the query. This latter point adds
+significant flexibility to the user, who no longer needs to decide apriori what imputation strategy they have to stick to
+for all their queries.
 
@@
 running imputation on the base tables and executing the query on the
-cleaned data. Each query was run 10 iterations in both settings and then results from the two approaches were paired up and compared tuple-wise. We average 
-absolute percentage deviations across each iteration of a query, and we report this value averaged over all iterations.
+cleaned data. Each query was run 10 iterations in both settings and then results from the two approaches were paired up and compared tuple-wise. We average tuple-wise
+absolute percentage deviations within each iteration of a query, and we report this value averaged over all iterations.
 We can see that optimizing for quality indeed reduces the SMAPE of query results. In general, the SMAPE relative to the
-base-imputation approach are low in all cases.
+base-imputation approach are low in all cases, indicating that on-the-fly imputation produces similar results to
+imputation at the base tables.
 
@@
 In many cases, applying the imputation step at the base table simply is not an option. To illustrate the increasing difficulty of such an approach
-as datasets scale, we used a simple aggregate query over the ACS dataset (Figure~\ref{query-acs}). Applying the imputation operation to the base table
+as datasets scale, we used a simple aggregate query over the ACS dataset in our final experiment(Figure~\ref{query-acs}). Applying the imputation operation to the base table
 is extremely expensive, in our case the process did not terminate within 20 minutes. In contrast, \ProjectName{} executes a quality-optimized version

### 2017-02-22 paper/relatedwork.tex
paper: some typos, clarifications, fixes

@@
 In contrast, \ProjectName{} allows users to specify a tradeoff between
-information loss and computational complexity (in terms of time). Furthermore,
+information loss and runtime performance, allowing users to perform queries
+directly on the entire dataset. Furthermore,
 the query planner's imputation is guided by the requirements of each specific
@@
 contrast, aims to remain squarely in the database realm, while allowing
-users to perform queries on a potentially larger subset of their data.
+users to directly express queries on a potentially larger subset of their data.
 
@@
 \subsection{Forecasting and Databases}
-\cite{parisi2011embedding} introduces the idea of incorporating time-series forecast operators into
+Parisi et al\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
 databases, along with the necessary relational algebra extensions. Their work explores the theoretical
@@
 
-\ProjectName{} provides a declarative language, as
-an extension to standard SQL, but in contrast to Fa, \ProjectName{} is not searching the space of possible
+Similarly to Fa, \ProjectName{} provides a declarative language, as
+a subset of standard SQL. \ProjectName{}, however, is not searching the space of possible
 imputation models, but rather the space of query plans that incorporate imputation operators. Another major point
-of distinction between \ProjectName{} and Fa is that the latter doesn't consider tradeoffs between accuracy and computation time, 
-but rather returns the most accurate forecast (with some stopping criterion).
+of distinction between \ProjectName{} and Fa is that the latter doesn't consider tradeoffs between accuracy and computation time, but rather returns the most accurate forecast (with some stopping criterion).
 

### 2017-02-22 simpledb/src/java/simpledb/ExperimentRunner.java
Update directory structure of experiments output

@@ -31,8 +31,10 @@ public class ExperimentRunner {
     private FileWriter timesFileWriter;
     private FileWriter resultsFileWriter;
     private FileWriter plansFileWriter;
+	private String[] queries;
 
-    private ExperimentRunner(boolean imputeAtBase, double minAlpha, double maxAlpha, double step, int iters, String catalog, String queries, String outputBaseDir)
+    private ExperimentRunner(boolean imputeAtBase, double minAlpha, double maxAlpha, double step, int iters,
+		String catalog, String queries, String outputBaseDir)
             throws IOException {
         this.imputeAtBase = imputeAtBase;
         this.minAlpha = minAlpha;
@@ -49,7 +51,8 @@ public class ExperimentRunner {
         this(true, 0.0, 0.0, 1.0, iters, catalog, queries, outputDir);
     }
 
-    public ExperimentRunner(double minAlpha, double maxAlpha, double step, int iters, String catalog, String queries, String outputDir)
+    public ExperimentRunner(double minAlpha, double maxAlpha, double step, int iters, String catalog, String queries,
+		String outputDir)
             throws IOException {
         this(false, minAlpha, maxAlpha, step, iters, catalog, queries, outputDir);
     }
@@ -60,32 +63,6 @@ public class ExperimentRunner {
         TableStats.computeStatistics();
     }
 
-    /**
-     * Open FileWriters to three results files. Creates a subdirectory appropriate to the timing task for q and alpha.
-     * @param q
-     * @param alpha
-     * @throws IOException
-     */
-    private void openWriters(int q, double alpha) throws IOException{
-	Path outputDir = outputBaseDir.resolve(String.format("%02d/%03.0f", q, alpha*1000));
-	if (Files.notExists(outputDir)){
-		Files.createDirectories(outputDir);
-	}
-	this.timesFileWriter = new FileWriter(outputDir.resolve("timing.csv").toFile());
-	this.resultsFileWriter = new FileWriter(outputDir.resolve("results.txt").toFile());
-	this.plansFileWriter = new FileWriter(outputDir.resolve("plans.txt").toFile());
-
-        this.timesFileWriter.write("query,alpha,iter,plan_time,run_time,plan_hash\n");
-        this.resultsFileWriter.write("Query Results\n");
-        this.plansFileWriter.write("Query Plans\n");
-    }
-
-    public void closeWriters() throws IOException {
-        this.timesFileWriter.close();
-        this.resultsFileWriter.close();
-        this.plansFileWriter.close();
[…]

### 2017-02-22 simpledb/src/java/simpledb/ImputedPlanCacheDottedWriter.java
More imformative filenames from ImputedPlanCacheDottedWriter

@@ -29,7 +29,7 @@ public class ImputedPlanCacheDottedWriter implements ImputedPlanCacheWriter {
 		for (AImputedPlanCache.Value val : cache.bestPlans(tables)) {
 			ImputedPlan plan = val.plan;
 			StringBuilder prefix = new StringBuilder();
-			prefix.append("plandot_loss=");
+			prefix.append("queryplan_loss=");
 			prefix.append(plan.loss().get());
 			prefix.append("_time=");
 			prefix.append(plan.time());

### 2017-02-22 paper/algorithm.tex
paper: cleaned up complexity section, removed todos

@@
 \subsection{Complexity}
-\todo{this section needs to be cleaned up/clarified}
-Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
-note that if we remove our restriction on imputation operators (i.e. minimal vs maximal), and allow any arbitrary subset of attributes to be imputed,
-then every single operator in the query plan has a number of imputations exponential in the number of dirty columns. Our restriction, instead, increases the number
-of plans (in the worst case) at each operator by a factor of 3. Of course, this implies that in the worst case (where all dirty sets tracked are distinct throughout the query plan),
-we explore a number of plans that further scales the number of plans in the original algorithm by an exponential factor.
+Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our planning algorithm still operates in exponential time. 
+If we removed our restriction on imputation operators to consider not just minimal versus maximal imputations, and instead allowed any arbitrary subset
+of attributes to be imputed at a plan node (including those not operated on by that node), then every single operator in the query plan would have a
+number of plans exponential in the number of dirty columns.
 
-However, we note that in all cases we have considered, this exponential blowup does not affect the practical performance of our optimizer.
+By restricting imputations to two classes: those that impute only attributes used in the operator and those that impute all attributes needed downstream in the query,
+we limit the factor of plans at each operator to three: \textit{Drop}, \textit{Impute} over attributes used in operator, and \textit{Impute} over all attributes needed.
+Recall that the planner maintains different plans for different dirty sets, keeping only those that are not dominated by other plans with the same dirty set. So in many
+cases we can drop some of these intermediate plans generated at each operator. However, if the dirty sets tracked are distinct through the entire planning phase,
+then we could explore $\mathcal{O}(3^{\# of operators}) * plans$.
+
+However, we note that in all cases we have considered, this exponential blowup does not affect the practical performance of our optimizer and results in relatively short
+planning times.
 

### 2017-02-22 paper/experiments.tex
paper: cleaned up complexity section, removed todos

@@
 \subsection{Queries}
-\todo{Add more queries: we want ~10 (balanced between the two datasets), explain them below}
 We collected a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.

### 2017-02-22 simpledb/test/experiments/run_experiments.py
Entrypoint and queries for experiments

@@ -1,0 +1,30 @@
+#!/usr/bin/env python
+
+import os
+import subprocess
+
+executable = ["java","-jar","../../dist/simpledb.jar"]
+output_dir = os.path.join(os.path.dirname(os.path.realpath(__file__)), "output")
+catalog   = "../../catalog.txt"
+queries   = "queries.txt"
+
+def run_small_experiment():
+    this_output_dir = os.path.join(output_dir, "small")
+
+    iters     = 3
+    min_alpha = 0.01
+    max_alpha = 0.99
+    step      = 0.32
+
+    run_experiment(this_output_dir, iters, min_alpha, max_alpha, step)
+
+def run_experiment(this_output_dir, iters, min_alpha, max_alpha, step):
+    if not os.path.isdir(this_output_dir):
+        os.makedirs(this_output_dir)
+
+    subprocess.call(executable +
+        ["experiment", catalog, queries, this_output_dir,
+         str(iters), str(min_alpha), str(max_alpha), str(step)])
+
+if __name__ == "__main__":
+    run_small_experiment()

### 2017-02-23 paper/examples.tex
paper: pass over motivating examples section

@@
 \section{Motivating Example}
-Consider an epidemiologist tasked with an initial
-exploratory analysis of data collected for individuals across
+Consider an epidemiologist tasked with an 
+exploratory analysis of data collected from individuals across
 a battery of exams. In particular, she is interested in exploring
@@
 
-While careful modeling techniques will be necessary before putting
+While careful modeling will be necessary before putting
 results into the final research paper, the epidemiologist is anxious
@@
 for handling the missing values. She could drop any records with
-missing values and continue her analysis with the remaining data,
-or she could fit a model on the base tables which can take advantage of existing correlations between attributes to fill in the missing values.
-If she drops the records with missing data, she will throw away the majority of her data set, as many records have missing fields.
-Fitting a model on the dataset will take more programming effort and may take a long time to run.
+missing values and continue her analysis with the remaining data.
+She could also fit a model on the base tables, taking advantage of correlations between attributes to fill in the missing values.
+However, if she drops the records with missing data, she will throw away the majority of her data set, as many records have missing fields.
+Fitting a model on the dataset will take some up-front programming effort and may take a long time to run.
 
@@
 some of the steps she is interested in, including filtering and grouping,
-are impacted differently by missing data, and that these step may change
+are impacted differently by missing data, and that these steps may change
 repeatedly, as she wants to run various queries during the exploration phase.
-The imputation strategy itself may change through out time. For example, as she is getting an initial
-set of distribution statistics (such as mean and standard deviation), she may be willing to tolerate
-a simpler imputation approach, as long as she gets runtime performance in exchange. Later on,
+Finally, her preferences may change over time. For example, as she is getting an initial
+set of distribution statistics (such as mean and standard deviation), she may be willing to tolerate more error, as long as she gets runtime performance in exchange. Later on,
 as more precise estimates are needed for comparing treatments or policy changes, the researcher
-may want to focus on the quality of the imputation.
+may want to focus on the quality of the results.
 
-Rather than work through
-the implications of each case, she submits a query, as shown in Figure~\ref{fig:example-query}, to ImputeDB.\@ The system, parameterized by a value indicating
-a tradeoff between information loss and runtime cost, then finds
-the optimal query plan. Figures~\ref{fig:quality-plan} and ~\ref{fig:fast-plan} show what this plan would be
-when optimizing for quality and performance, respectively.
+Rather than work through the implications of each case, she writes her query in standard SQL and submits it to ImputeDB, which automatically performs the imputations necessary to fill in the missing data.
+An example query is shown in Fig.~\ref{fig:example-query}.
 
+She indicates her preference for performance or for accuracy by providing a configuration parameter $\alpha$ to ImputeDB in addition to her query.
+$\alpha$ is a dimensionless parameter in $[0,1]$, and it indicates the relative importance of the quality of query results and of query performance.
+So, a low $\alpha$ results in a fast query plan that may compromise accuracy, while a high $\alpha$ results in a high quality query plan that may take longer to run.
+In our intended workflow, the analyst picks a value of $\alpha$ and incrementally increases it to increase query result quality until the query takes too long to run.
+Fig.~\ref{fig:quality-plan} and Fig.~\ref{fig:fast-plan} show plans for the example query in Fig.~\ref{fig:example-query} when optimizing for quality and performance, respectively.
+
+In Fig.~\ref{fig:quality-plan} and Fig.~\ref{fig:fast-plan}, $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
+$\mu$ and $\delta$ are specialized imputation operators.
+$\mu$ fills in missing values using a regression model while $\delta$ simply drops tuples which contain a missing value in the specified attributes.
 The differences between the two plans highlight the subtleties of optimizing queries which contain imputation operators.
-Note that imputation operators are placed in different points of the plan, operating over different columns, and 
-even influencing the order of joins. The quality-optimized plan (Figure~\ref{fig:quality-plan}) uses the impute operation, rather
-than dropping tuples with missing values. It waits to impute \verb|demo.income| until after the final join has taken place, but
-other imputations take place earlier on in the plan, some before filtering and join operations. 
-Meanwhile, the performance-optimized plan (Figure~\ref{fig:fast-plan}) performs a single impute operation (\verb|exams.weight|), and applies drops to tuples with missing values before aggregating.
+Note that imputation operators are placed in different points of the plan, operating over different columns, and even influencing the order of joins.
+
+The quality-optimized plan (Fig.~\ref{fig:quality-plan}) uses the impute operation $\mu$, rather than dropping tuples with missing values.
+It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
+Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
+Meanwhile, the performance-optimized plan (Fig.~\ref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
+However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.
 
@@
 \begin{figure}
+  \begin{subfigure}[b]{\linewidth}
   \Tree
@@
     [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
-      [.$\mu_{\text{demo.income}}$
+      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
         [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
-          [.$\mu_{\text{labs.white\_blood\_cell\_ct}}$
+          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
             [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
               [.$\sigma_{\text{exams.weight} \geq 120}$ 
-                [.$\mu_{\text{exams.weight}}$ exams ] 
+                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
               ] 
@@
   ]
-\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for quality.}
+  \caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for quality ($\alpha=1$).}
   \label{fig:quality-plan}
-\end{figure}
-
-\begin{figure}
+\end{subfigure}
+\begin{subfigure}[b]{\linewidth}
   \Tree
@@
     [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
-      [.$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$
+      [.\colorbox{pink}{$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$}
         [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
@@
             [.$\sigma_{\text{demo.gender} = 2}$ demo ]
-            [.$\sigma_{\text{exams.weight} \geq 120}$ [.$\delta_{\text{exams.weight}}$ exams ] ] ] labs ] ] ] ]
-\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for performance.}
+            [.$\sigma_{\text{exams.weight} \geq 120}$ [.\colorbox{pink}{$\delta_{\text{exams.weight}}$} exams ] ] ] labs ] ] ] ]
+\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for performance ($\alpha=0$).}
 \label{fig:fast-plan}
+\end{subfigure}
+\vspace{0.5\baselineskip}
+\caption{Plans for the query in Fig.~\ref{fig:example-query}. Imputation operators are highlighted.}
 \end{figure}
 
-Tens of queries later, the epidemiologist has the holistic view of the
-data that she required before carefully crafting her own tailored
-imputation model. Knowing that there may be a need to explore
-this data set further, she can easily incorporate their imputation model
-into ImputeDB for future use.
+Returning to our epidemiologist, tens of queries later she has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.
+Knowing that there may be a need to explore this data set further, she can easily incorporate her imputation model into ImputeDB for future use.
 

### 2017-02-23 paper/algorithm.tex
paper: partial pass over algorithm section

@@
 
-Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans containing imputation operators (Sec.~\ref{sec:operators}) for an optimal plan (Sec.~\ref{sec:cost-model}) given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.
+Our optimizer searches (Sec.~\ref{sec:planning}) a restricted space (Sec.~\ref{sec:search-space}) of query plans containing imputation operators (Sec.~\ref{sec:operators}) for an optimal plan given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.
 
@@
 Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
-Our planning algorithm is agnostic to the type of imputation used (Sec.~\ref{sec:imputation}), as it is treated as a black-box operation.
+Our planning algorithm is agnostic to the type of imputation used (Sec.~\ref{sec:imputation}), as the properties of the imputation are summarized by the cost model (Sec.~\ref{sec:cost-model}).
 
@@
 attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
-replace all null values with non-null values for attributes in $C$ in the relation $R$.
-The imputation algorithm is discussed in detail in Sec.~\ref{sec:imputation}.
+replace all null values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in Sec.~\ref{sec:imputation}).
 \textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
@@
 \label{sec:cost-model}
-The cost for a plan $q$ is expressed as a tuple of $\langle \textsc{Loss}(q), \textsc{Time}(q) \rangle$.
-$\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of any query predicates.
-$\textsc{Loss}(q)$ is an estimate of the amount of information lost through the sequence of imputation procedures used.
-We note that our loss calculation constrains $\textsc{Loss}(q)$ to $[0, 1]$, while $\textsc{Time}(q)$ can range within $(0, \infty)$.
+The cost of a plan $q$ is expressed as a tuple $\langle \textsc{Loss}(q), \textsc{Time}(q) \rangle$.
+$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
+$\textsc{Loss}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
 
 \begin{definition}
-We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$, if $l < l' \land t < t'$, and denote this as $\doms{p}{p'}$. 
+We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$ if $l < l'$ and $t < t'$. We denote this as $\doms{p}{p'}$. 
 \end{definition}
 
-Given a query, \ProjectName{} produces a final set of possible plans $P$ by only maintaining plans that are not dominated by others in the search space. So if
-our search space is $S$
+Given a query, \ProjectName{} produces a final set of plans $P$ by only keeping plans that are not dominated by others in the search space.
+So for a search space $S$, \[P= \{ p ~|~ \not\exists p' \in S.\ p \neq p' \land \doms{p'}{p} \}.\]
 
-\begin{gather*}
- P= \{ p | \not\exists p' \in S. \doms{p'}{p} \}
- \end{gather*}
+This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of performance and cost for the current query.
 
-This set outlines the best possible options for trading performance and cost for the current query, building a Pareto frontier\cite{pareto1964cours}.
-
-In order to pick a final plan from the frontier, our model introduces a parameter $\alpha$ which encodes an upper bound on the amount of information loss that a user is willing to tolerate
-relative to the minimum information loss possible given the imputation operations.
+In order to pick a final plan from the frontier, our model introduces a parameter $\alpha$ which encodes an upper bound on the amount of information loss that a user is willing to tolerate relative to the minimum information loss possible given the imputation operations.\todo{Get rid of discussion of information loss.
+It's a technical term and we don't really tackle it.}
 
 \begin{definition}
-Given a final set of plans $P$, and $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
+Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
 We say $p$ is $\alpha$-bound optimal if $\forall p'^\alpha \in P: \textsc{Time}(p) \leq \textsc{Time}(p')$. 
@@
 
-Given $\alpha$ and $P$, \ProjectName{} returns the $\alpha$-bound optimal plan in $P$. Note that the $\alpha$-bound optimal plan is unique in $P$, given that
-$P$ does not keep any plans dominated by others in the set.
+Given $\alpha$ and $P$, \ProjectName{} returns the $\alpha$-bound optimal plan in $P$. Note that the $\alpha$-bound optimal plan is unique in $P$, since $P$ contains only non-dominated plans.
 
@@
   \textsc{Losses}(q) &= \begin{cases}
-    \{1\} \cup \textsc{Loss}(q') & q = \delta_C(q') \\
-    \{\frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Cardinality}(q') }} \} \cup \textsc{Losses}(q')  & q = \mu_C(q') \\
-    \textsc{Losses}(q_1') \cup \textsc{Losses}(q_2') & q = q_1' \bowtie_\psi q_2' \\
+     1 :: \textsc{Losses}(q') & q = \delta_C(q') \\
+    \frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q') }}  :: \textsc{Losses}(q')  & q = \mu_C(q') \\
+    \textsc{Losses}(q_1') ::: \textsc{Losses}(q_2') & q = q_1' \bowtie_\psi q_2' \\
     \textsc{Losses}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
+    Nil & q = t\\
   \end{cases}\\
@@
 \subsection{Query planning}
+\label{sec:planning}
 The input to our query planner is a tuple $(T, F, J, P, G, A)$:

### 2017-02-23 paper/algorithm.tex
paper: another partial pass over algorithm section

@@
 \begin{align*}
-  \textsc{Losses}(q) &= \begin{cases}
-     1 :: \textsc{Losses}(q') & q = \delta_C(q') \\
-    \frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q') }}  :: \textsc{Losses}(q')  & q = \mu_C(q') \\
-    \textsc{Losses}(q_1') ::: \textsc{Losses}(q_2') & q = q_1' \bowtie_\psi q_2' \\
-    \textsc{Losses}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
-    Nil & q = t\\
+  \textsc{L}(q) &= \begin{cases}
+     \{1\} \cup \textsc{L}(q') & q = \delta_C(q') \\
+    \left\{\frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}}\right\} \cup \textsc{L}(q')  & q = \mu_C(q') \\
+    \textsc{L}(q_1') \cup \textsc{L}(q_2') & q = q_1' \bowtie_\psi q_2' \\
+    \textsc{L}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
+    \emptyset & q = t\\
   \end{cases}\\
-  \textsc{Loss}(q) &= \frac{\Sigma_{l \in \textsc{Losses}(q)} l}{| \textsc{Losses}(q)|}
+  \textsc{Loss}(q) &= \frac{\Sigma \textsc{L}(q)}{|\textsc{L}(q)|}
 \end{align*}
+
+\todo{Similar algorithm for computing Time?}
 
@@
 The planner operates as follows.
-First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan.
+First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (Lines~\ref{lst:line:attr-start}-\ref{lst:line:attr-end}).
 This set will be used to determine which attributes are imputed.
 Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
-a search over imputation operations produces various possible plans, which are added to the cache. If no selections are available, a 
-simple scan is added to the plan cache. Join order is jointly optimized with imputation placements and a set of
-plans encompassing all necessary tables is produced. This set is now extended for any necessary grouping and aggregation
-operations. After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
+a search over imputation operations produces various possible plans, which are added to the cache (Line~\ref{lst:line:sel}). If no selections are available, a 
+simple scan is added to the plan cache (Line~\ref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
+plans encompassing all necessary tables is produced (Line~\ref{lst:line:join})\todo{Citation or explanation for join optimization algorithm}.
+This set is now extended for any necessary grouping and aggregation
+operations (Line~\ref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
 best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
@@
 cases we can drop some of these intermediate plans generated at each operator. However, if the dirty sets tracked are distinct through the entire planning phase,
-then we could explore $\mathcal{O}(3^{\# of operators}) * plans$.
+then we could explore $\mathcal{O}(3^{\# of operators}) * plans$\todo{this is unclear. what is plans?}.
 

### 2017-02-23 paper/algorithms/plan-toplevel.tex
paper: another partial pass over algorithm section

@@
 % \begin{algorithm}
-\begin{algorithmic}[5]
+\begin{algorithmic}[1]
   \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses trade-off performance and information loss}
@@
   \Function{Plan}{$T, F, J, P, G, A$}
-  \State $C_g \gets \bigcup_{\phi \in F} \Call{Attr}{\phi}$ \Comment{Collect relevant attributes.} \label{lst:line:attr-start}
+  \State $C_g \gets \bigcup_{\phi \in F} \Call{Attr}{\phi}$ \Comment{Collect relevant attributes.}~\label{lst:line:attr-start}
   \State $C_g \gets \bigcup_{\psi \in J} \Call{Attr}{\psi} \cup P \cup G \cup \Call{Attr}{A} \cup C_g$ \label{lst:line:attr-end}
@@
   \If {$\exists \phi: (t, \phi) \in F$}
-  \State $Q[\{t\}] \lhd \Call{OptFilter}{Q, t, \phi, C_g}$
+  \State $Q[\{t\}] \lhd \Call{OptFilter}{Q, t, \phi, C_g}$~\label{lst:line:sel}
   \Else
-  \State $Q[\{t\}] \lhd \{t\}$
+  \State $Q[\{t\}] \lhd \{t\}$~\label{lst:line:scan}
   \EndIf
@@
 
-  \State $\Call{OptJoin}{Q, T, J, C_g}$	\Comment{Optimize joins.}
+  \State $\Call{OptJoin}{Q, T, J, C_g}$	\Comment{Optimize joins.}~\label{lst:line:join}
   \State $B \gets Q[T]$ \Comment{Get the best plans for all tables.}
-  \If{$G \neq \emptyset$} \Comment{Add optional group \& aggregate.}
+  \If{$G \neq \emptyset$} \Comment{Add optional group \& aggregate.}~\label{lst:line:group}
   \State $B \gets \bigcup_{q \in B} \Call{OptGroupBy}{q, G, A, C_g}$
@@
   \State $B \gets \bigcup_{q \in B} \Call{OptProject}{q, P}$ \Comment{Add projections.}
-  \State \Return $p \in B$ s.t. $p$ is $\alpha$-bound optimal  \Comment{Return fastest plan satisfying $\alpha$ bound.}
+  \State \Return $p \in B$ s.t. $p$ is $\alpha$-bound optimal.
   \EndFunction

### 2017-02-23 paper/examples.tex
fix: meaning of alpha is swapped under new cost model

@@
 $\alpha$ is a dimensionless parameter in $[0,1]$, and it indicates the relative importance of the quality of query results and of query performance.
-So, a low $\alpha$ results in a fast query plan that may compromise accuracy, while a high $\alpha$ results in a high quality query plan that may take longer to run.
+So, a high $\alpha$ results in a fast query plan that may compromise accuracy, while a low $\alpha$ results in a high quality query plan that may take longer to run.
 In our intended workflow, the analyst picks a value of $\alpha$ and incrementally increases it to increase query result quality until the query takes too long to run.
@@
   ]
-  \caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for quality ($\alpha=1$).}
+  \caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for quality ($\alpha=0$).}
   \label{fig:quality-plan}
@@
             [.$\sigma_{\text{exams.weight} \geq 120}$ [.\colorbox{pink}{$\delta_{\text{exams.weight}}$} exams ] ] ] labs ] ] ] ]
-\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for performance ($\alpha=0$).}
+\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for performance ($\alpha=1$).}
 \label{fig:fast-plan}

### 2017-02-23 paper/algorithm.tex
fix: info loss => info lost, and adjusted small discussion

@@
 
-In order to pick a final plan from the frontier, our model introduces a parameter $\alpha$ which encodes an upper bound on the amount of information loss that a user is willing to tolerate relative to the minimum information loss possible given the imputation operations.\todo{Get rid of discussion of information loss.
-It's a technical term and we don't really tackle it.}
+In order to pick a final plan from the frontier, our model introduces a parameter $\alpha$ which encodes an upper bound on the $\textsc{Loss}$ value that the user is willing
+to tolerate relative to the minimum possible amongst frontier plans.
 
@@
 accuracy \textit{increases} when more complete attributes and complete tuples are available;
-correspondingly, the information loss \textit{decreases} with more values to work
+correspondingly, the information lost \textit{decreases} with more values to work
 with. Since we conceive that

### 2017-02-23 paper/algorithms/plan-toplevel.tex
fix: info loss => info lost, and adjusted small discussion

@@
 \begin{algorithmic}[1]
-  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses trade-off performance and information loss}
+  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses trade-off performance and information lost}
   \Ensure{An optimized query plan.}

### 2017-02-23 paper/conclusion.tex
fix: info loss => info lost, and adjusted small discussion

@@
 learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
-information loss and runtime. We showed that different values for the tradeoff parameter can yield substantially
+information lost and runtime. We showed that different values for the tradeoff parameter can yield substantially
 different plans, allowing the user to express their preferences for performance on a per-query basis.

### 2017-02-23 paper/relatedwork.tex
fix: info loss => info lost, and adjusted small discussion

@@
 In contrast, \ProjectName{} allows users to specify a tradeoff between
-information loss and runtime performance, allowing users to perform queries
+information lost and runtime performance, allowing users to perform queries
 directly on the entire dataset. Furthermore,
@@
 results under different conditions. Instead, the optimizer allows users to pick their tradeoff between
-runtime cost and information loss. The search space considered by our optimizer is broader, not just
+runtime cost and information lost. The search space considered by our optimizer is broader, not just
 forecast-first/forecast-last plans, but rather imputation operators can be placed anywhere in the query plan

### 2017-02-23 simpledb/src/java/simpledb/ExperimentRunner.java
Write to base/ subdirectory when using imputeAtBase

@@ -31,10 +31,10 @@ public class ExperimentRunner {
     private FileWriter timesFileWriter;
     private FileWriter resultsFileWriter;
     private FileWriter plansFileWriter;
-	private String[] queries;
+    private String[] queries;
 
     private ExperimentRunner(boolean imputeAtBase, double minAlpha, double maxAlpha, double step, int iters,
-		String catalog, String queries, String outputBaseDir)
+        String catalog, String queries, String outputBaseDir)
             throws IOException {
         this.imputeAtBase = imputeAtBase;
         this.minAlpha = minAlpha;
@@ -43,7 +43,11 @@ public class ExperimentRunner {
         this.iters = iters;
         this.catalogPath = Paths.get(catalog);
         this.queriesPath = Paths.get(queries);
-        this.outputBaseDir = Paths.get(outputBaseDir);
+        if (imputeAtBase){
+            this.outputBaseDir = Paths.get(outputBaseDir, "base");
+        } else {
+            this.outputBaseDir = Paths.get(outputBaseDir);
+        }
     }
 
     public ExperimentRunner(int iters, String catalog, String queries, String outputDir)
@@ -52,7 +56,7 @@ public class ExperimentRunner {
     }
 
     public ExperimentRunner(double minAlpha, double maxAlpha, double step, int iters, String catalog, String queries,
-		String outputDir)
+        String outputDir)
             throws IOException {
         this(false, minAlpha, maxAlpha, step, iters, catalog, queries, outputDir);
     }
@@ -72,11 +76,11 @@ public class ExperimentRunner {
     }
 
     private void openTimeWriter(int q, double alpha) throws IOException {
-	Path outputDir = outputBaseDir.resolve(String.format("q%02d/alpha%03.0f", q, alpha*1000));
-	if (Files.notExists(outputDir)){
-		Files.createDirectories(outputDir);
-	}
-	this.timesFileWriter = new FileWriter(outputDir.resolve("timing.csv").toFile());
+    Path outputDir = outputBaseDir.resolve(String.format("q%02d/alpha%03.0f", q, alpha*1000));
+    if (Files.notExists(outputDir)){
+        Files.createDirectories(outputDir);
+    }
+    this.timesFileWriter = new FileWriter(outputDir.resolve("timing.csv").toFile());
         this.timesFileWriter.write("query,alpha,iter,plan_time,run_time,plan_hash\n");
     }
 
@@ -85,12 +89,12 @@ public class ExperimentRunner {
     }
 
[…]

### 2017-02-23 simpledb/test/experiments/run_experiments.py
Add medium and large experiments

@@ -1,5 +1,6 @@
 #!/usr/bin/env python
 
+from __future__ import print_function
 import os
 import subprocess
 
@@ -18,13 +19,52 @@ def run_small_experiment():
 
     run_experiment(this_output_dir, iters, min_alpha, max_alpha, step)
 
+def run_medium_experiment():
+    this_output_dir = os.path.join(output_dir, "medium")
+
+    iters     = 100
+    min_alpha = 0.00
+    max_alpha = 1.00
+    step      = 0.20
+
+    run_experiment(this_output_dir, iters, min_alpha, max_alpha, step)
+
+def run_experiment_large():
+    this_output_dir = os.path.join(output_dir, "large")
+
+    iters     = 500
+    min_alpha = 0.00
+    max_alpha = 1.00
+    step      = 0.05
+
+    run_experiment(this_output_dir, iters, min_alpha, max_alpha, step)
+
 def run_experiment(this_output_dir, iters, min_alpha, max_alpha, step):
     if not os.path.isdir(this_output_dir):
         os.makedirs(this_output_dir)
 
+    # Timing using ImputeDB
     subprocess.call(executable +
         ["experiment", catalog, queries, this_output_dir,
          str(iters), str(min_alpha), str(max_alpha), str(step)])
 
+    # Timing using impute on base table
+    subprocess.call(executable +
+        ["experiment", catalog, queries, this_output_dir,
+         str(iters), "--base"])
+
 if __name__ == "__main__":
-    run_small_experiment()
+    import sys
+    if len(sys.argv) == 2:
+        experiment_size = sys.argv[1]
+    else:
+        experiment_size = "small"
+
+    if experiment_size == "small":
+        run_small_experiment()
+    elif experiment_size == "medium":
+        run_medium_experiment()
+    elif experiment_size == "large":
+        run_medium_experiment()
+    else:
+        raise Error

### 2017-02-23 simpledb/test/experiments/analyze_experiments.py
Update analyze_epxeriments for new dir structure and python3 compat

@@ -1,171 +1,204 @@
-#!/usr/local/bin/python
+#!/usr/bin/env python
+
+from __future__ import print_function
 
 import pandas as pd
 import numpy as np
+
+import matplotlib
+matplotlib.use('Agg')
 import matplotlib.pyplot as plt
-import os.path
+plt.ioff()
+
+import os
+import glob
 import sys
-from StringIO import StringIO
+
 from collections import defaultdict
 
-if len(sys.argv) != 4:
-  print "Usage: python analyze_experiments.py <experiment-results-dir> <base-results-dir> <output-dir>"
-  sys.exit(1)
-
-experiments_dir = sys.argv[1]
-base_dir = sys.argv[2]
-output_dir = sys.argv[3]
+try:
+    from StringIO import StringIO
+except ImportError:
+    from io import StringIO
 
 # basic utils
-def drop_warmup(df, by, keep=5):
-  return df.groupby(by).tail(keep)
+def drop_warmup(df, by, drop=20):
+    gb = df.groupby(by)
+    n  = max(gb.size())
+    # have the same number in each cat, so this is okay
+    return gb.tail(n-drop)
 
 def summarize(df, by, a):
-  ops = {'mean': np.mean, 'std': np.std}
-  return df.groupby(by)[a].agg(ops).reset_index()
+    ops = {'mean': np.mean, 'std': np.std}
+    return df.groupby(by)[a].agg(ops).reset_index()
   
 def plot(df,**kwargs):
-  _, ax = plt.subplots(1)
-  for a in set(df['alpha']):
-    data = df[df['alpha'] == a]
-    try:
-      data.plot(ax=ax, label=r'$\alpha=%.0f$' % a, **kwargs)
-    except TypeError:
-      data.plot(ax=ax, label=a, **kwargs)
-    # get around bug in legend drawing in pandas
-    plt.legend()
-  return ax
+    _, ax = plt.subplots(1)
+    for a in set(df['alpha']):
+        data = df[df['alpha'] == a]
+        try:
+            data.plot(ax=ax, label=r'$\alpha=%.0f$' % a, **kwargs)
+        except TypeError:
+            data.plot(ax=ax, label=a, **kwargs)
+        # get around bug in legend drawing in pandas
+        plt.legend()
+    return ax
   
-def get_query_results(path, headers):
-  lines = []
-  with open(path, 'r') as f:
-    lines = f.readlines()
-  lines = [l.strip() for l in lines]
[…]

### 2017-02-23 simpledb/test/experiments/run_experiments.py
Fix typo in large experiment

@@ -29,7 +29,7 @@ def run_medium_experiment():
 
     run_experiment(this_output_dir, iters, min_alpha, max_alpha, step)
 
-def run_experiment_large():
+def run_large_experiment():
     this_output_dir = os.path.join(output_dir, "large")
 
     iters     = 500
@@ -65,6 +65,6 @@ if __name__ == "__main__":
     elif experiment_size == "medium":
         run_medium_experiment()
     elif experiment_size == "large":
-        run_medium_experiment()
+        run_large_experiment()
     else:
         raise Error

### 2017-02-23 paper/experiments.tex
paper: Update queries and descriptions in experiments sec

@@
 \subsection{Queries}
-We collected a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
+We collect a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
 
 The queries on the CDC NHANES data consist not only of projections and selections,
-but also interesting joins and aggregations. Our aim was to craft meaningful queries
+but also interesting joins and aggregates. Our aim was to craft meaningful queries
 that would provide performance figures relevant to practitioners who may use
@@
 
-Queries 1-3 are on the CDC data. Query~\ref{q1} calculates the average weight for users based on their income data.
-To do so, it joins demographics and physical examination data. Query~\ref{q2}
-compares cholesterol levels for individuals who are at both ends of the income
-spectrum and who satisfy a minimum weight requirement. Query~\ref{q3}
-extracts the maximum blood lead levels for subjects who satisfy a particular
-waist circumference requirement.
+Queries 1-5 are on the CDC data. In Query~\ref{q1}, we calculate the average height for users based on their income data. In Query~\ref{q2},
+we compare cholesterol levels for individuals at either tail of the income
+spectrum and above a certain weight. In Query~\ref{q3},
+we extract the maximum blood lead levels for children under the age of 6 years.
+In Query ~\ref{q4}, we calculate the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity.
+In Query ~\ref{q5}, we extract the age (in months), creatine level, and head circumference, for subjects with high creatine levels.
 
-Queries 4-7 are on the FCC data. In Query~\ref{q4}, we calculate the average income for survey participants,
-based on their bootcamp attendance. Query~\ref{q5} estimates the average
-age of women from the United States who participated. Query~\ref{q6} 
-calculates the average amount of money survey participants with student debt destined to
-learning based on their school degree. Finally, in Query~\ref{q7}, we join the FCC data with a reference table provided
+% Perhaps more school makes your head bigger :)
+
+Queries 6-10 are on the FCC data. In Query~\ref{q6}, we calculate the average income for survey participants,
+based on their bootcamp attendance. In Query~\ref{q7} estimates the average
+age of women from the United States who participated. In Query~\ref{q8}, we 
+calculate the average amount of money survey participants with student debt destined to
+learning based on their school degree. In Query~\ref{q9}, we join the FCC data with a reference table provided
 by the World Bank which summarizes GDP per Capita across various
 countries\cite{worldbank-data}. The query calculates the average GDP-per-capita for participants based on whether or not they
-attended a bootcamp.
+attended a bootcamp. Finally, in Query ~\ref{q10}, we identify individuals from countries with GDP per-capita of \$5000 or
+less who earn \$100,000 or more post-bootcamp.
 
@@
 %\todobox{actually run experiments and rewrite this section}{
-All experiments were run on a Macbook Pro with two 2.9 GHz Intel core i5 processors, 16 GB of main memory,  running OS X El Capitan.
+All experiments were run on a single AWS EC2 c4.xlarge instance, with 4 2.9 GHz Intel Xeon E5-2666 v3 virtual CPUs and 7.5 GiB of main memory on Debian Linux.
 
@@
 
+%TODO MJS
 Figure~\ref{fig:plantimes} provides a summary of the planning times for each of the queries.  We exclude the planning time for queries that impute at base table, as that
@@
 running imputation on the base tables and executing the query on the
-cleaned data. Each query was run 10 iterations in both settings and then results from the two approaches were paired up and compared tuple-wise. We average tuple-wise
+cleaned data. Each query was run 10 %TODO MJS
+iterations in both settings and then results from the two approaches were paired up and compared tuple-wise. We average tuple-wise
 absolute percentage deviations within each iteration of a query, and we report this value averaged over all iterations.
@@
 
+% TODO MJS
 In many cases, applying the imputation step at the base table simply is not an option. To illustrate the increasing difficulty of such an approach
@@
 quality-optimized and performance-optimized query execute in 9 seconds and 1 second, respectively.
-}
+}%TODO MJS
 \label{query-acs}

### 2017-02-23 paper/experiments.tex
Update query in paper

@@
-\section{Experiments}
+\section{Experiments}\label{sec:experiments}
 To evaluate the performance of our approach,  we implemented
@@
 \subsection{Queries}
-We collect a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
+We collect a set of queries (Table~\ref{fig:queries}) that we think are both representative,
+in that they could reasonably be written by a user in the course of data analysis, and
+interesting to plan.
 
@@
 
-Queries 1-5 are on the CDC data. In Query~\ref{q1}, we calculate the average height for users based on their income data. In Query~\ref{q2},
-we compare cholesterol levels for individuals at either tail of the income
-spectrum and above a certain weight. In Query~\ref{q3},
-we extract the maximum blood lead levels for children under the age of 6 years.
-In Query ~\ref{q4}, we calculate the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity.
-In Query ~\ref{q5}, we extract the age (in months), creatine level, and head circumference, for subjects with high creatine levels.
+Queries 1-5 are on the CDC data. In Query~\ref{q1}, we calculate the average height for
+users based on their income data. In Query~\ref{q2}, we compare cholesterol levels for
+individuals with low, medium, and high incomes spectrum and above a certain weight. In
+Query~\ref{q3}, we extract the maximum blood lead levels for children under the age of 6
+years. In Query~\ref{q4}, we calculate the average systolic blood pressure, by gender, for
+subjects with a body mass index indicating obesity. In Query~\ref{q5}, we extract the age,
+gender, triglyceride levels, and waist circumference from individuals with high triglyceride
+levels.
 
-% Perhaps more school makes your head bigger :)
-
-Queries 6-10 are on the FCC data. In Query~\ref{q6}, we calculate the average income for survey participants,
-based on their bootcamp attendance. In Query~\ref{q7} estimates the average
-age of women from the United States who participated. In Query~\ref{q8}, we 
+Queries 6-10 are on the FCC data. In Query~\ref{q6}, we calculate the average income for
+survey participants, based on their bootcamp attendance. In Query~\ref{q7} estimates the
+average age of women from the United States who participated. In Query~\ref{q8}, we
 calculate the average amount of money survey participants with student debt destined to
-learning based on their school degree. In Query~\ref{q9}, we join the FCC data with a reference table provided
-by the World Bank which summarizes GDP per Capita across various
-countries\cite{worldbank-data}. The query calculates the average GDP-per-capita for participants based on whether or not they
-attended a bootcamp. Finally, in Query ~\ref{q10}, we identify individuals from countries with GDP per-capita of \$5000 or
-less who earn \$100,000 or more post-bootcamp.
+learning based on their school degree. In Query~\ref{q9}, we join the FCC data with a
+reference table provided by the World Bank which summarizes GDP per-capita across various
+countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
+participants based on whether or not they attended a bootcamp. Finally, in Query~\ref{q10},
+we identify individuals from countries with GDP per-capita of \$5000 or less who earn
+\$100,000 or more post-bootcamp.
 
@@
 %\todobox{actually run experiments and rewrite this section}{
-All experiments were run on a single AWS EC2 c4.xlarge instance, with 4 2.9 GHz Intel Xeon E5-2666 v3 virtual CPUs and 7.5 GiB of main memory on Debian Linux.
+All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
+four 2.9 GHz Intel Xeon E5-2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.
 
-Figure~\ref{fig:runtimes} provides a summary of the performance results. The first plot shows average running times for \ProjectName{}
-plans in two configurations: quality-optimized and runtime-optimized. The second plot shows, for comparison, the 
-average running times required for each query when performing imputation at the base tables.
-In all cases, running a query through \ProjectName{} is cheaper than 
-imputing a single base table, with the added advantage of imputation tailored to the query. This latter point adds
-significant flexibility to the user, who no longer needs to decide apriori what imputation strategy they have to stick to
-for all their queries.
+Figure~\ref{fig:runtimes} provides a summary of the performance results. The first plot
+shows average running times for \ProjectName{} plans in two configurations:
+quality-optimized and runtime-optimized. The second plot shows, for comparison, the average
+running times required for each query when performing imputation at the base tables.  In all
+cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
+with the added advantage of imputation tailored to the query. This latter point adds
+significant flexibility to the user, who no longer needs to decide apriori what imputation
+strategy they have to stick to for all their queries.
 
@@
 \end{subfigure}
-\caption{Comparing \ProjectName{} runtimes for each query (\ref{subfig:project-runtime-queries}) with the runtime for a query when base-table imputations are used (\ref{subfig:base-runtime-queries})}
+\caption{Comparing \ProjectName{} runtimes for each query
+    (\ref{subfig:project-runtime-queries}) with the runtime for a query when base-table
+    imputations are used (\ref{subfig:base-runtime-queries})}
 \label{fig:runtimes}
@@
 %TODO MJS
-Figure~\ref{fig:plantimes} provides a summary of the planning times for each of the queries.  We exclude the planning time for queries that impute at base table, as that
-requires no planning. On average, planning constituted 38 percent of total runtime for plans generated by \ProjectName{}.
+Figure~\ref{fig:plantimes} provides a summary of the planning times for each of the queries.
+We exclude the planning time for queries that impute at base table, as that requires no
+planning. On average, planning constituted 38 percent of total runtime for plans generated
+by \ProjectName{}.
 
@@
 
-Table~\ref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to 
-running imputation on the base tables and executing the query on the
+Table~\ref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
+\ProjectName{}'s query results when compared to running imputation on the base tables and
+executing the query on the
 cleaned data. Each query was run 10 %TODO MJS
-iterations in both settings and then results from the two approaches were paired up and compared tuple-wise. We average tuple-wise
-absolute percentage deviations within each iteration of a query, and we report this value averaged over all iterations.
-We can see that optimizing for quality indeed reduces the SMAPE of query results. In general, the SMAPE relative to the
-base-imputation approach are low in all cases, indicating that on-the-fly imputation produces similar results to
-imputation at the base tables.
+iterations in both settings and then results from the two approaches were paired up and
+compared tuple-wise. We average tuple-wise absolute percentage deviations within each
+iteration of a query, and we report this value averaged over all iterations.  We can see
+that optimizing for quality indeed reduces the SMAPE of query results. In general, the SMAPE
+relative to the base-imputation approach are low in all cases, indicating that on-the-fly
+imputation produces similar results to imputation at the base tables.
 
@@
 % TODO MJS
-In many cases, applying the imputation step at the base table simply is not an option. To illustrate the increasing difficulty of such an approach
-as datasets scale, we used a simple aggregate query over the ACS dataset in our final experiment(Figure~\ref{query-acs}). Applying the imputation operation to the base table
-is extremely expensive, in our case the process did not terminate within 20 minutes. In contrast, \ProjectName{} executes a quality-optimized version
-in 9 seconds and a runtime-optimized version in 1 second. This highlights the potential benefit of using our system for early data exploration.
+In many cases, applying the imputation step at the base table simply is not an option. To
+illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
+aggregate query over the ACS dataset in our final experiment(Figure~\ref{query-acs}).
+Applying the imputation operation to the base table is extremely expensive, in our case the
+process did not terminate within 20 minutes. In contrast, \ProjectName{} executes a
+quality-optimized version in 9 seconds and a runtime-optimized version in 1 second. This
+highlights the potential benefit of using our system for early data exploration.
 
@@
 \end{lstlisting}
-\caption{A simple aggregate query on ACS data. Imputing the base table is undesirable (did not terminate within 20min). \ProjectName's
-quality-optimized and performance-optimized query execute in 9 seconds and 1 second, respectively.
+\caption{A simple aggregate query on ACS data. Imputing the base table is prohibitive.
+    \ProjectName's quality-optimized and performance-optimized
+    query execute in 9 seconds and 1 second, respectively.
 }%TODO MJS

### 2017-02-23 paper/experiments.tex
Update query in paper

@@
-\section{Experiments}\label{sec:experiments}
+\section{Experiments}
 To evaluate the performance of our approach,  we implemented

### 2017-02-23 paper/introduction.tex
paper: some typos, clarifications, fixes
paper: work on intro

--- 2016-12-14 paper/introduction.tex
@@
 
-Handling incorrect or dirty data is a complex and challenging problem for data scientists.
+% Context
+Handling incorrect or dirty data is a complex and challenging problem for analysts.
+% Problem
 One way in which a dataset can be dirty is for parts of it to be missing altogether.
 Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
-To handle this problem, users may manually clean their dataset by performing
-some statistical analysis to replace missing data elements with likely values.
+According to some estimates, data scientists spend between 50 and 80 percent of their time performing tasks such as collecting and cleaning datasets~\cite{data-science-cleaning}.
+
+% Solution
+To handle this problem, users may manually clean their dataset by performing some statistical analysis to replace missing data elements with likely values.
 This process is called \emph{imputation}.
-Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
-Although manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
-Additionally, it may not even be necessary to completely clean the data to make it usable.
-Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
-Others may want to run queries on a subset of the data, so do not need to impute
-every field in every record. Yet others may want to customize the 
-imputation algorithm for the tradeoffs and demands of a particular domain.
+We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on-the-fly, during query execution.
+The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
 
-In this paper, we present \ProjectName{}, a database system which is designed to interact with a dirty dataset as though it were clean.
-The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
-To achieve this goal, we perform imputation on the fly, during query execution.
-Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows users to flexibly trade accuracy for computation time.
+% Prior work
+There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
+Prior work in statistics has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
+Prior work in databases has a established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
+Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
+In contrast, \ProjectName{} does not focus on the imputation algorithm itself but rather on the planning involved in its use in non-trivial query plans.
+
+% Implications
+Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set and get good results, even if that data has missing values.
+
+% Background
+Traditional imputation methods work over the entire data set and replace all missing values.
+Although this solves the problem of missing data, running a sophisticated imputation algorithm over a large data set can be very expensive.
+A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
+Finally, for many queries, it is not necessary to clean the entire data set, and a significant speed up can be obtained by cleaning only the necessary parts.
+
+% Scope
+\ProjectName{} is designed to enable data exploration, by allowing analysts to run their queries and get good results without dealing with imputation.
+However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.
+
+\subsection{Contributions}
+This paper makes the following contributions:
+\begin{itemize}
+\item \textbf{Relational algebra with imputation:}
+  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
+  Impute fills in missing data values using a regression model (CE-CART)~\cite{burgette2010multiple}.
+  Drop simply drops tuples which contain null values.
+\item \textbf{Model of imputation quality and cost:}
+  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
+  We use the cost model to abstract over the imputation algorithm used.
+  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
+\item \textbf{Query planning with imputation:}
+  We present a novel query planning algorithm which jointly optimizes for running time and the quality of the results.
+  It does so by maintaining the set of Pareto optimal plans according to the cost model.
+  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
+\end{itemize}
+
+% Summary of related work
+
+% Although this manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
+% Additionally, it may not be necessary to completely clean the data to make it usable.
+% Some users may be willing to run queries on dirty data, simply ignoring any missing values, as long as they do not have to pay the cost of imputation.
+% Others may want to run queries on a subset of the data, and so do not need to impute
+% every field in every record. Yet others may want to customize the 
+% imputation algorithm for the tradeoffs and demands of a particular domain.
+
+% In this paper, we present \ProjectName{}, a database system which is designed to interact with a dirty dataset as though it were clean.
+% To achieve this goal, we perform imputation on the fly, during query execution.
+% Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows users to flexibly trade imputation quality for computation time.
 

### 2017-02-23 paper/relatedwork.tex
paper: work on intro

@@
 Imputation of missing values is a widely studied field within the statistics and machine
-learning communities. As highlighted in \cite{gelman2006data}, missing data
+learning communities. As highlighted in~\cite{gelman2006data}, missing data
 can appear for a variety of reasons, including both random and conditioned on
@@
 focus on correctly modeling relationships between the attributes to factor in
-varied forms of missingness. For example, Burgette and Reiter
-(\cite{burgette2010multiple}) discuss the usage of sequential regression trees
+varied forms of missingness. For example, Burgette and Reiter~\cite{burgette2010multiple} discuss the usage of sequential regression trees
 for imputing missing data.
 
-In \cite{akande2015empirical}, Akande et al analyze the performance of various
+In~\cite{akande2015empirical}, Akande et al analyze the performance of various
 multiple imputation techniques on the American Community Survey dataset. 
@@
 There is a long history in the database community surrounding the
-treatment of nulls. As early as 1973, \cite{codd1973understanding}
+treatment of nulls. As early as 1973,~\cite{codd1973understanding}
 provides a treatment of the semantics of null. Multiple
 papers have described various (at times conflicting) treatments
-of nulls \cite{grant1977null}. \ProjectName's main design invariant - no relational operator
+of nulls~\cite{grant1977null}. \ProjectName's main design invariant - no relational operator
 sees missing values for attributes it must operate on and users should never see

### 2017-02-23 paper/relatedwork.tex
Add related work items

--- 2017-02-23 paper/relatedwork.tex
@@
 \cite{hellerstein2008quantitative}.
+
+In \cite{wolf2007query}, queries over databases with missing values are
+processed in a statistical approach, taking advantage of correlations between
+variables. Both tuples that match the original query as well as a ranking of
+tuples with incomplete data that may match are returned. Our work differs in
+that we allow any well-formed statistical technique to be used for imputation
+and focus on returning results to the analyst as if the database had been
+complete.
+
+Designers of data stream processing systems frequently confront missing values
+and consider them carefully in query processing. Often, if sensor error is the
+cause of missing values, values can be imputed with high confidence. In
+\cite{fernandez2009inter}, feedback punctuation is used to dynamically
+reconfigure the query plan as data continue to arrive for state-dependent
+optimizations. In this framework, expensive imputations can be avoided as
+real-time conditions change.
 

### 2017-02-23 paper/experiments.tex
Add results for acs_dirty query

@@
 
-% TODO MJS
-In many cases, applying the imputation step at the base table simply is not an option. To
-illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
+In many real-world cases, applying the imputation step at the base table is prohibitive.
+To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
 aggregate query over the ACS dataset in our final experiment(Figure~\ref{query-acs}).
-Applying the imputation operation to the base table is extremely expensive, in our case the
-process did not terminate within 20 minutes. In contrast, \ProjectName{} executes a
-quality-optimized version in 9 seconds and a runtime-optimized version in 1 second. This
-highlights the potential benefit of using our system for early data exploration.
+Applying the imputation operation to the base table is extremely expensive, in our case
+completing in 75 minutes. In contrast, \ProjectName{} executes a quality-optimized version
+in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
+benefit of using our system for early data exploration.
 
@@
 \caption{A simple aggregate query on ACS data. Imputing the base table is prohibitive.
-    \ProjectName's quality-optimized and performance-optimized
-    query execute in 9 seconds and 1 second, respectively.
-}%TODO MJS
+    \ProjectName's quality-optimized and performance-optimized queries execute in 7 seconds
+    and 1 second, respectively.
+}
 \label{query-acs}

### 2017-02-23 simpledb/test/experiments/run_experiments.py
Add results for acs_dirty query

@@ -3,6 +3,7 @@
 from __future__ import print_function
 import os
 import subprocess
+import tempfile
 
 executable = ["java","-jar","../../dist/simpledb.jar"]
 output_dir = os.path.join(os.path.dirname(os.path.realpath(__file__)), "output")
@@ -39,6 +40,35 @@ def run_large_experiment():
 
     run_experiment(this_output_dir, iters, min_alpha, max_alpha, step)
 
+def run_acs_experiment():
+    this_output_dir = os.path.join(output_dir, "acs")
+
+    # Impute on base table
+    (f, acs_query) = tempfile.mkstemp()
+    os.write(f, "SELECT AVG(c0) FROM acs_dirty;\n")
+
+    iters = 1
+
+    print("Running acs base...")
+    executable_max_heap = [executable[0]] + ["-Xmx3200m"] + executable[1:]
+    cmd = executable_max_heap + \
+        ["experiment", catalog, acs_query, this_output_dir, str(iters), "--base"]
+    print(cmd)
+    subprocess.call(cmd)
+    print("Running acs base...done.")
+
+    # Impute using ImputeDB
+    iters     = 220
+    min_alpha = 0.00
+    max_alpha = 1.00
+    step      = 1.00
+
+    subprocess.call(executable +
+        ["experiment", catalog, acs_query, this_output_dir,
+         str(iters), str(min_alpha), str(max_alpha), str(step)])
+
+    os.close(f)
+
 def run_experiment(this_output_dir, iters, min_alpha, max_alpha, step):
     if not os.path.isdir(this_output_dir):
         os.makedirs(this_output_dir)
@@ -66,5 +96,9 @@ if __name__ == "__main__":
         run_medium_experiment()
     elif experiment_size == "large":
         run_large_experiment()
+    elif experiment_size == "acs":
+        run_acs_experiment()
     else:
         raise Error
+
+    print("Done.")

### 2017-02-24 simpledb/test/experiments/analyze_experiments.py
Fix more bugs

@@ -82,11 +82,20 @@ def get_query_results(experiments_dir, headers, base=False):
         experiments_dir = os.path.join(experiments_dir, "base")
 
     # Iterate through queries, alpha, iters
-    for f in glob.glob(experiments_dir + os.path.sep + "q*/alpha*/it*/result.txt"):
+    print("get_query_results")
+    if restrict_alpha=True:
+        globs = glob.glob(experiments_dir + os.path.sep + "q*/alpha000/it*/result.txt") + \
+             glob.glob(experiments_dir + os.path.sep + "q*/alpha1000/it*/result.txt")
+    else:
+        globs = glob.glob(experiments_dir + os.path.sep + "q*/alpha*/it*/result.txt")
+
+    for f in globs:
+        print("Reading {}...".format(f))
         qi = f.rfind("/q")+len("/q")
         q = int(f[qi:qi+2])
         ai = f.rfind("/alpha") + len("/alpha")
-        alpha = int(f[ai:ai+3])/1000.0
+        aj = f.find("/", ai)
+        alpha = int(f[ai:aj])/1000.0
         df = pd.read_csv(f, header=None, names=headers[q])
         if len(df.columns) > 1:
             df = df.set_index(df.columns[0])
@@ -116,7 +125,7 @@ def get_smape(refs, results):
     metrics = np.array(metrics)
     return np.nanmean(metrics), np.nanstd(metrics)
 
-def get_timing_results(experiments_dir, base=False):
+def get_timing_results(experiments_dir, restrict_alpha=False, base=False):
     df = pd.DataFrame(columns=["query","alpha","iter","plan_time","run_time","plan_hash"])
 
     # Impute on base stored in base/ subdirectory, buth with same subdirectory
@@ -124,9 +133,14 @@ def get_timing_results(experiments_dir, base=False):
     if base:
         experiments_dir = os.path.join(experiments_dir, "base")
 
+    if restrict_alpha:
+        globs = glob.glob(experiments_dir + os.path.sep + "q*/alpha000/timing.csv") + \
+             glob.glob(experiments_dir + os.path.sep + "q*/alpha1000/timing.csv")
+     else:
+         globs = glob.glob(experiments_dir + os.path.sep + "q*/alpha*/timing.csv")
+
     # Iterate through queries, alpha, reading 'timing.csv' for each.
-    for f in glob.glob(experiments_dir + os.path.sep + "q*/alpha*/timing.csv"):
[…]

### 2017-02-24 paper/experiments.tex
Update perf summary
Update perf discussion

@@
 
-%\todobox{actually run experiments and rewrite this section}{
 All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
@@
 cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
-with the added advantage of imputation tailored to the query. This latter point adds
-significant flexibility to the user, who no longer needs to decide apriori what imputation
-strategy they have to stick to for all their queries.
+with performance improvement on a factor of ten.
 
@@
 
-%TODO MJS
 Figure~\ref{fig:plantimes} provides a summary of the planning times for each of the queries.
 We exclude the planning time for queries that impute at base table, as that requires no
-planning. On average, planning constituted 38 percent of total runtime for plans generated
-by \ProjectName{}.
+planning. In the median query across a variety of queries and choices of $\alpha$, planning
+constituted 8.5 percent of total runtime by \ProjectName{}.
 
@@
 \ProjectName{}'s query results when compared to running imputation on the base tables and
-executing the query on the
-cleaned data. Each query was run 10 %TODO MJS
-iterations in both settings and then results from the two approaches were paired up and
-compared tuple-wise. We average tuple-wise absolute percentage deviations within each
-iteration of a query, and we report this value averaged over all iterations.  We can see
-that optimizing for quality indeed reduces the SMAPE of query results. In general, the SMAPE
-relative to the base-imputation approach are low in all cases, indicating that on-the-fly
-imputation produces similar results to imputation at the base tables.
+executing the query on the cleaned data. Each query was run 200 iterations in both settings
+and then results from the two approaches were paired up and compared tuple-wise. We average
+tuple-wise absolute percentage deviations within each iteration of a query, and we report
+this value averaged over all iterations.  We can see that optimizing for quality indeed
+reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
+approach are low in all cases, indicating that on-the-fly imputation produces similar
+results to imputation at the base tables.
 
@@
 \centering
-\input{perf_summary.tex}
-\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$ parameterizations relative to results when imputing on base table}
+\input{tables/perf_summary.tex}
+\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
+    parameterizations relative to results when imputing on base table. Values of $0.0$,
+    $100.0$, or $NaN$ indicate uninformative values.} %TODO MJS
 \label{table:smape}

### 2017-02-24 paper/algorithm.tex
paper: added similar algo breakout for Time as for Loss

@@
 null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
-that plan. We use this estimate to compute $\text{Loss}(Q)$ as follows.
+that plan. We use this estimate to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.
 
@@
     \textsc{L}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
-    \emptyset & q = t\\
+    \emptyset\ multiset & q = t\\
   \end{cases}\\
@@
 \end{align*}
-
-\todo{Similar algorithm for computing Time?}
 
@@
 
-To compute $\text{Time}(Q)$, we retain a simple set of heuristics used by a standard 
-database. Scan costs are estimated based on the number of tuples, 
+To compute $\text{Time}(Q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard 
+database.
+
+%\begin{align*}
+%  \textsc{T}(q) &= \begin{cases}
+%     \{ f_1(pg_{fetch}, io_{cost}) \} \cup \textsc{T}(q') & q = \delta_C(q') \\
+%      \{ f_2(\textsc{Attr}(q'), C, \textsc{Card}(q'), iters) \} \cup \textsc{T}(q')  & q = \mu_C(q') \\  
+%     \{ f_3(\textsc{T}(q_1'), \textsc{T}(q_2')) \} & q = q_1' \bowtie_\psi q_2' \\
+%    \{ \hat{t} \} \cup \textsc{T}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
+%    \{ f_4(\textsc{Card}(t), pg_{size}, io_{cost}) \} & q = t\\
+%  \end{cases}
+%  \intertext{ \textit{for some set of functions  $f_{1,2,3,4}$, constant $\hat{t}$, IO Cost/Page $io_{cost}$, page fetches $pg_{fetch}$, page size $pg_{size}$, iterations $iters$} } 
+%  \textsc{Time}(q) &= \Sigma \textsc{T}(q)
+%\end{align*}
+
+Scan costs are estimated based on the number of tuples, 
 page size, and an unit-less IO cost per page. Join costs are estimated as a function
@@
 simple scan is added to the plan cache (Line~\ref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
-plans encompassing all necessary tables is produced (Line~\ref{lst:line:join})\todo{Citation or explanation for join optimization algorithm}.
-This set is now extended for any necessary grouping and aggregation
+plans encompassing all necessary tables is produced (Line~\ref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
 operations (Line~\ref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
@@
 $\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.
+
+The join optimization is based on the algorithm used in System R\cite{blasgen1981system},
+but rather than select the best plan at every point, we use our plan cache, which maintains the Pareto frontier of possible plans.
+The output of the join optimizer is the set of plans which satisfy all join predicates and lie along the frontier.
 
@@
 algorithm terminates when convergence is achieved (i.e. the imputed values do not change across
-epochs) or a fixed number of epochs is reached.  Algorithm~\ref{algo:imputation-strategy}
-provides details on the implementation.
+epochs) or a fixed number of epochs is reached.
 
@@
 
-\begin{algorithm}
-    \input{algorithms/cart.tex}
-\end{algorithm}
+%\begin{algorithm}
+%    \input{algorithms/cart.tex}
+%\end{algorithm}
 
@@
 cases we can drop some of these intermediate plans generated at each operator. However, if the dirty sets tracked are distinct through the entire planning phase,
-then we could explore $\mathcal{O}(3^{\# of operators}) * plans$\todo{this is unclear. what is plans?}.
+then we could explore exponentially more plans than System R.
 

### 2017-02-24 paper/algorithm.tex
paper: working on optimality proof

@@
 \begin{definition}
-We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$ if $l < l'$ and $t < t'$. We denote this as $\doms{p}{p'}$. 
+We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$ if $(l \leq l' \land t < t') \lor (l < l' \land t \leq t')$.
+We denote this as $\doms{p}{p'}$.
 \end{definition}
@@
 
+\subsection{Optimality}
+We prove the optimality of the final plan selected by showing that our cost-model and algorithm
+construct a complete Pareto frontier.
+
+%\begin{definition}
+%Given two plans $p$ and $p'$,  such that $|\textsc{Dirty}(p) \setminus \textsc{Dirty}(p')| = n$, and $\textsc{Time}(p) \leq \textsc{Time}(p')$,
+%we say $p$ yields $p'$ with dirty set size reduction $n$, if we can add any number of imputation and relational operators to $p$ to obtain
+%$p'$. We denote this $p \xrightarrow{n} p'$. 
+%\end{definition}
+
+\todo{define F and way its ok, why it depends solely on the dirty set... the algorithm keeps all other state relevant, so this is all thats needed for transformations}
+
+\begin{theorem}
+Our cost-model and plan-cache semantics results in a final set of plans that constitute a Pareto frontier.
+\end{theorem}
+
+\begin{proof}
+For contradiction, let $p_1$ and $p_2$ be plans such that $\doms{p_1}{p_2}$. Given a function $F: D \rightarrow O$ (d is dirty sets, o is operators),
+assume plans only $p_3 \in F(\textsc{Dirty}(p_2))(p_2)$, and $p_4 \in F(\textsc{Dirty}(p_1))(p_1)$, and that $\dom{p_3}{p_4}$.
+
+A contradiction arises from in one of two cases
+
+case 1: p1 doesn't dominate p2 (that con only happen if the dirty sets are different) (they are comparable), so not tossed
+case 2: they have the same dirty set, and since dirty set determines F exclusively, then p3 and p4 are reachable for both, so they are substituable
+\end{proof}
+
+%\begin{proof}
+%We prove this by induction on the reduction in the size of the dirty set for a plan.
+%
+%For the base case, given a plan $p$, we obtain $P_1 = \{ p' | p \xrightarrow{1} p'  \}$. We construct a set of plans for each
+%dirty set by keeping all non-dominated plans with the same dirty set. By definition, each of these sets are a Pareto frontier 
+%to reduce $\textsc{Dirty}(p)$ by 1.
+%
+%By the induction hypothesis, we can do the same on $P_n = \{ p' | p \xrightarrow{n} p' \}$ to produce a set of Pareto frontiers to
+%reduce $\textsc{Dirty}(p)$ by $n$.
+%
+%Given a Pareto optimal plan in our search space $p''$ s.t. $p \xrightarrow{n+1} p''$, it must be the case that there exists a plan $p''_{prev}  \xrightarrow{1} p''$. 
+%
+%If $p''_{prev}$ is Pareto optimal for its dirty set and a reduction of size $n$, then we can directly yield $p''$ from the appropriate
+%frontier.
+%
+%If $p''_{prev}$ is not Pareto optimal, there exists a Pareto optimal plan $\widehat{p''_{prev}}$ with the same dirty set
+%as $p''_{prev}$, which dominates it, so we can apply the same imputation operations to $\widehat{p''_{prev}}$ to yield $p''$.
+%\end{proof}
+%
+%\todo{We're working on a proof of optimality for the final set of plans created}
+
 \subsection{Complexity}

### 2017-02-24 paper/algorithm.tex
paper: cleaned up proof, first pass

--- 2017-02-24 paper/algorithm.tex
@@
 
+\subsection{Optimality}
+We prove the optimality of the final plan selected by showing that our cost-model and algorithm
+construct a complete Pareto frontier.
+
+\begin{definition}
+An abstract imputation operator $I: \mathcal{P}(String) \rightarrow \mathcal{P}(\{ \delta, \mu \})$ is a function that given a dirty set maps to
+a set of imputation operations. For convenience we denote applying the set of transformations to a plan $p$ as $I(p)$ and understand that this
+is shorthand for $\{ op(p)   | op \in I(\textsc{Dirty}(p)) \}$.
+\end{definition}
+
+Algorithm{~\ref{algo:top-level-planner} plans the query in stages. It first handles any pushed-down selections, it then optimizes joins, then groupby/aggregations,
+and finally projections. It maintains all necessary state information through a combination of dirty sets on intermediate plans and
+initializing empty plan caches at stages that require it. Given this, we can succinctly describe possible transformations to a plan at a given
+point in the algorithm solely through $I$.
+
+\begin{theorem}\label{theorem:no-bad-sit}
+By computing dirty-set specific Pareto frontiers, it is never the case that we discard a plan $p_1$ s.t.
+$\doms{p_2}{p_1}$, but $\exists p_1' \in I(p_1), p_2' \in I(p_2).
+\doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2')$
+\end{theorem}
+
+Theorem~\ref{theorem:no-bad-sit} states that there is no \textit{bad situation} such as that depicted in Figure~\ref{fig:bad-sit-diagram}
+
+\begin{figure}
+\begin{tikzpicture}
+% before I
+\node(p1) {$p_1$};
+\node[right=of p1](dom1){$\prec$};
+\node[right=of dom1](p2) {$p_2$};
+
+% after I
+\node[below=of p1](p1'){$p_1'$};
+\node[right=of p1'](dom2) {$\succ$};
+\node[right=of dom2](p2'){$p_2'$};
+
+% arrows
+\draw[->] (p1) -- node[right] {$\scriptstyle I$} (p1');
+\draw[->] (p2) -- node[right] {$\scriptstyle I$} (p2');
+\end{tikzpicture}
+\caption{A bad Pareto frontier selection would choose $p_2$ and never reach $p_1'$}
+\label{fig:bad-sit-diagram}
+\end{figure}
+
+\begin{proof}
+For contradiction, let $p_1$ and $p_2$ be plans such that $\doms{p_2}{p_1}$, $p_1$ is removed from the plan cache
+in favor of $p_2$, but there  $\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2')$
+and $p_1'$ is not reachable from $p_2$ by $I$.
+
+There are two possible cases :
+
+\begin{case}
+$\textsc{Dirty}(p_1) = \textsc{Dirty}(p_2)$, in which case $I(p_1) = I(p_2)$, as $I$ uniquely determines the possible transformations.
+This makes $p_2$ and $p_1$ substitutable, so $p_1'$ is reachable from $p_2$.
+This is a contradiction.
+\end{case}
+
+\begin{case}
+$\textsc{Dirty}(p_1) \not= \textsc{Dirty}(p_2)$, in which case $p_1$ is not removed from the plan cache as the $\doms{p_2}{p_1}$ is not within the same dirty set. This is a contradiction
+\end{case}
+
+So it must be the case that no such situation arises.
+\end{proof}
+
+Given that no desirable plans are discarded by computing Pareto frontiers for each dirty set, the final Pareto frontier (for dirty set $\emptyset$) is optimal.
+
 \subsection{Complexity}

### 2017-02-24 paper/experiments.tex
Update perf discussion
paper: intro feedback from martin

--- 2017-02-23 paper/experiments.tex
@@
 \subsection{Queries}
-We collect a set of queries (Table~\ref{fig:queries}) that we think are both representative,
-in that they could reasonably be written by a user in the course of data analysis, and
-interesting to plan.
+We collect a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.
 
@@
 
-Queries 1-5 are on the CDC data. In Query~\ref{q1}, we calculate the average height for
-users based on their income data. In Query~\ref{q2}, we compare cholesterol levels for
-individuals with low, medium, and high incomes spectrum and above a certain weight. In
-Query~\ref{q3}, we extract the maximum blood lead levels for children under the age of 6
-years. In Query~\ref{q4}, we calculate the average systolic blood pressure, by gender, for
-subjects with a body mass index indicating obesity. In Query~\ref{q5}, we extract the age,
-gender, triglyceride levels, and waist circumference from individuals with high triglyceride
-levels.
+Queries 1-5 are on the CDC data. In Query~\ref{q1}, we calculate the average height for users based on their income data. In Query~\ref{q2},
+we compare cholesterol levels for individuals at either tail of the income
+spectrum and above a certain weight. In Query~\ref{q3},
+we extract the maximum blood lead levels for children under the age of 6 years.
+In Query ~\ref{q4}, we calculate the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity.
+In Query ~\ref{q5}, we extract the age (in months), creatine level, and head circumference, for subjects with high creatine levels.
 
-Queries 6-10 are on the FCC data. In Query~\ref{q6}, we calculate the average income for
-survey participants, based on their bootcamp attendance. In Query~\ref{q7} estimates the
-average age of women from the United States who participated. In Query~\ref{q8}, we
+% Perhaps more school makes your head bigger :)
+
+Queries 6-10 are on the FCC data. In Query~\ref{q6}, we calculate the average income for survey participants,
+based on their bootcamp attendance. In Query~\ref{q7} estimates the average
+age of women from the United States who participated. In Query~\ref{q8}, we 
 calculate the average amount of money survey participants with student debt destined to
-learning based on their school degree. In Query~\ref{q9}, we join the FCC data with a
-reference table provided by the World Bank which summarizes GDP per-capita across various
-countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
-participants based on whether or not they attended a bootcamp. Finally, in Query~\ref{q10},
-we identify individuals from countries with GDP per-capita of \$5000 or less who earn
-\$100,000 or more post-bootcamp.
+learning based on their school degree. In Query~\ref{q9}, we join the FCC data with a reference table provided
+by the World Bank which summarizes GDP per Capita across various
+countries\cite{worldbank-data}. The query calculates the average GDP-per-capita for participants based on whether or not they
+attended a bootcamp. Finally, in Query ~\ref{q10}, we identify individuals from countries with GDP per-capita of \$5000 or
+less who earn \$100,000 or more post-bootcamp.
 
@@
 %\todobox{actually run experiments and rewrite this section}{
-All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
-four 2.9 GHz Intel Xeon E5-2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.
+All experiments were run on a single AWS EC2 c4.xlarge instance, with 4 2.9 GHz Intel Xeon E5-2666 v3 virtual CPUs and 7.5 GiB of main memory on Debian Linux.
 
-Figure~\ref{fig:runtimes} provides a summary of the performance results. The first plot
-shows average running times for \ProjectName{} plans in two configurations:
-quality-optimized and runtime-optimized. The second plot shows, for comparison, the average
-running times required for each query when performing imputation at the base tables.  In all
-cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
-with the added advantage of imputation tailored to the query. This latter point adds
-significant flexibility to the user, who no longer needs to decide apriori what imputation
-strategy they have to stick to for all their queries.
+Figure~\ref{fig:runtimes} provides a summary of the performance results. The first plot shows average running times for \ProjectName{}
+plans in two configurations: quality-optimized and runtime-optimized. The second plot shows, for comparison, the 
+average running times required for each query when performing imputation at the base tables.
+In all cases, running a query through \ProjectName{} is cheaper than 
+imputing a single base table, with the added advantage of imputation tailored to the query. This latter point adds
+significant flexibility to the user, who no longer needs to decide apriori what imputation strategy they have to stick to
+for all their queries.
 
@@
 \end{subfigure}
-\caption{Comparing \ProjectName{} runtimes for each query
-    (\ref{subfig:project-runtime-queries}) with the runtime for a query when base-table
-    imputations are used (\ref{subfig:base-runtime-queries})}
+\caption{Comparing \ProjectName{} runtimes for each query (\ref{subfig:project-runtime-queries}) with the runtime for a query when base-table imputations are used (\ref{subfig:base-runtime-queries})}
 \label{fig:runtimes}
@@
 %TODO MJS
-Figure~\ref{fig:plantimes} provides a summary of the planning times for each of the queries.
-We exclude the planning time for queries that impute at base table, as that requires no
-planning. On average, planning constituted 38 percent of total runtime for plans generated
-by \ProjectName{}.
+Figure~\ref{fig:plantimes} provides a summary of the planning times for each of the queries.  We exclude the planning time for queries that impute at base table, as that
+requires no planning. On average, planning constituted 38 percent of total runtime for plans generated by \ProjectName{}.
 
@@
 
-Table~\ref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
-\ProjectName{}'s query results when compared to running imputation on the base tables and
-executing the query on the
+Table~\ref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to 
+running imputation on the base tables and executing the query on the
 cleaned data. Each query was run 10 %TODO MJS
-iterations in both settings and then results from the two approaches were paired up and
-compared tuple-wise. We average tuple-wise absolute percentage deviations within each
-iteration of a query, and we report this value averaged over all iterations.  We can see
-that optimizing for quality indeed reduces the SMAPE of query results. In general, the SMAPE
-relative to the base-imputation approach are low in all cases, indicating that on-the-fly
-imputation produces similar results to imputation at the base tables.
+iterations in both settings and then results from the two approaches were paired up and compared tuple-wise. We average tuple-wise
+absolute percentage deviations within each iteration of a query, and we report this value averaged over all iterations.
+We can see that optimizing for quality indeed reduces the SMAPE of query results. In general, the SMAPE relative to the
+base-imputation approach are low in all cases, indicating that on-the-fly imputation produces similar results to
+imputation at the base tables.
 
@@
 
-In many real-world cases, applying the imputation step at the base table is prohibitive.
-To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
-aggregate query over the ACS dataset in our final experiment(Figure~\ref{query-acs}).
-Applying the imputation operation to the base table is extremely expensive, in our case
-completing in 75 minutes. In contrast, \ProjectName{} executes a quality-optimized version
-in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
-benefit of using our system for early data exploration.
+% TODO MJS
+In many cases, applying the imputation step at the base table simply is not an option. To illustrate the increasing difficulty of such an approach
+as datasets scale, we used a simple aggregate query over the ACS dataset in our final experiment(Figure~\ref{query-acs}). Applying the imputation operation to the base table
+is extremely expensive, in our case the process did not terminate within 20 minutes. In contrast, \ProjectName{} executes a quality-optimized version
+in 9 seconds and a runtime-optimized version in 1 second. This highlights the potential benefit of using our system for early data exploration.
 
@@
 \end{lstlisting}
-\caption{A simple aggregate query on ACS data. Imputing the base table is prohibitive.
-    \ProjectName's quality-optimized and performance-optimized queries execute in 7 seconds
-    and 1 second, respectively.
-}
+\caption{A simple aggregate query on ACS data. Imputing the base table is undesirable (did not terminate within 20min). \ProjectName's
+quality-optimized and performance-optimized query execute in 9 seconds and 1 second, respectively.
+}%TODO MJS
 \label{query-acs}

### 2017-02-24 paper/introduction.tex
paper: working on optimality proof
paper: intro feedback from martin

@@
 This process is called \emph{imputation}.
-We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on-the-fly, during query execution.
+We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on the fly, during query execution.
 The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
@@
 Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
-In contrast, \ProjectName{} does not focus on the imputation algorithm itself but rather on the planning involved in its use in non-trivial query plans.
+\ProjectName{} contributes a novel approach to planning data imputation on a per query basis, allowing users to use standard SQL on real-world data sets with missing values
+and express preferences in quality of data imputation versus running time.
+
+% Background
+While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive.
+A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
+Finally, for many queries, it is not necessary to clean the entire data set, and a significant speed up can be obtained by cleaning only the necessary parts.
+In contrast to existing systems, \ProjectName{} exploits this observation and avoids imputing over the entire data set, as done with traditional methods. \ProjectName{}
+carefully plans the placement of different imputation operations throughout the query plan, resulting in a 
+significant speedup in query execution. The focus on of our work is not on the imputation algorithm but rather on the planning involved in its use in non-trivial query plans.
 
 % Implications
-Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set and get good results, even if that data has missing values.
-
-% Background
-Traditional imputation methods work over the entire data set and replace all missing values.
-Although this solves the problem of missing data, running a sophisticated imputation algorithm over a large data set can be very expensive.
-A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
-Finally, for many queries, it is not necessary to clean the entire data set, and a significant speed up can be obtained by cleaning only the necessary parts.
+Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
+queries that execute in a fraction of the time it takes to impute base tables and then run queries (the traditional approach). Furthermore, the results obtained with this approach
+are similar in quality to those obtained with the base-table imputation approach, as indicated by low symmetric-Mean-Absolute-Percentage-Errors in our empirical results
+over real-world data sets (see Sec.~\ref{sec:experiments} for details).
 
 % Scope
-\ProjectName{} is designed to enable data exploration, by allowing analysts to run their queries and get good results without dealing with imputation.
+\ProjectName{} is designed to enable early data exploration, by allowing analysts to run their queries without explicitly dealing with imputation.
 However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.
@@
   We present a novel query planning algorithm which jointly optimizes for running time and the quality of the results.
-  It does so by maintaining the set of Pareto optimal plans according to the cost model.
+  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
   By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
+\item \textbf{Proof of optimality:}
+ We prove that our cost-model and planning algorithm produces a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
 \end{itemize}
+
+\ProjectName{} leverages the combined contributions to plan fast queries on data with missing values, while producing comparable results to traditional approaches.
+This provides the opportunity to increase productivity for users who frequently encounter missing values in their data.
 

### 2017-02-25 paper/examples.tex
paper: reworked motivation section

@@
 \section{Motivating Example}
-Consider an epidemiologist tasked with an 
+In this section, we present a common scenario where a scientists
+has to perform analysis on real-world data.
+
+\subsection{Analysis with missing data}
+An epidemiologist at the CDC is tasked with an 
 exploratory analysis of data collected from individuals across
@@
 
-While careful modeling will be necessary before putting
-results into the final research paper, the epidemiologist is anxious
+%While careful modeling will be necessary before putting
+%results into the final research paper,
+
+The epidemiologist is excited
 to get a quick and accurate view into the data. However,
-the data has been collected through CDC surveys (see Sec.~\ref{subsec:datasets} for details),
+the data has been collected through CDC surveys (see Sec.~\ref{subsec:datasets}),
 and there is a significant amount of missing data across all
-relevant attributes. The researcher must develop a strategy
-for handling the missing values. She could drop any records with
-missing values and continue her analysis with the remaining data.
-She could also fit a model on the base tables, taking advantage of correlations between attributes to fill in the missing values.
-However, if she drops the records with missing data, she will throw away the majority of her data set, as many records have missing fields.
-Fitting a model on the dataset will take some up-front programming effort and may take a long time to run.
+relevant attributes.
 
-To complicate matters further, the epidemiologist realizes that
-some of the steps she is interested in, including filtering and grouping,
-are impacted differently by missing data, and that these steps may change
-repeatedly, as she wants to run various queries during the exploration phase.
-Finally, her preferences may change over time. For example, as she is getting an initial
-set of distribution statistics (such as mean and standard deviation), she may be willing to tolerate more error, as long as she gets runtime performance in exchange. Later on,
-as more precise estimates are needed for comparing treatments or policy changes, the researcher
-may want to focus on the quality of the results.
+\todo{fill in the XX values}
+Before she can perform her queries, the researcher must develop a strategy
+for handling the missing values. She currently has two options:
+1) she could drop records with missing values in relevant fields,
+2) she could use a standard data imputation package on all of her data. Both of these
+approaches have significant drawbacks. For the query pictured in 
+Fig.~\ref{fig:example-query}, (1) drops XXXX percent of the relevant tuples,
+while her previous experiences with (2) have proven be expensive. In a previous project, she
+worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
+dataset took XXX hours. The analyst needs a more complete picture
+of the data, so (1) is insufficient, and she is under time pressure, so (2)
+is infeasible.
 
-Rather than work through the implications of each case, she writes her query in standard SQL and submits it to ImputeDB, which automatically performs the imputations necessary to fill in the missing data.
+She can run her queries immediately and finish her report if
+she uses \ProjectName{}, which allows her to write her
+query in standard SQL and automatically performs the imputations necessary to fill in the missing data.
 An example query is shown in Fig.~\ref{fig:example-query}.
-
-She indicates her preference for performance or for accuracy by providing a configuration parameter $\alpha$ to ImputeDB in addition to her query.
-$\alpha$ is a dimensionless parameter in $[0,1]$, and it indicates the relative importance of the quality of query results and of query performance.
-So, a high $\alpha$ results in a fast query plan that may compromise accuracy, while a low $\alpha$ results in a high quality query plan that may take longer to run.
-In our intended workflow, the analyst picks a value of $\alpha$ and incrementally increases it to increase query result quality until the query takes too long to run.
-Fig.~\ref{fig:quality-plan} and Fig.~\ref{fig:fast-plan} show plans for the example query in Fig.~\ref{fig:example-query} when optimizing for quality and performance, respectively.
-
-In Fig.~\ref{fig:quality-plan} and Fig.~\ref{fig:fast-plan}, $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
-$\mu$ and $\delta$ are specialized imputation operators.
-$\mu$ fills in missing values using a regression model while $\delta$ simply drops tuples which contain a missing value in the specified attributes.
-The differences between the two plans highlight the subtleties of optimizing queries which contain imputation operators.
-Note that imputation operators are placed in different points of the plan, operating over different columns, and even influencing the order of joins.
-
-The quality-optimized plan (Fig.~\ref{fig:quality-plan}) uses the impute operation $\mu$, rather than dropping tuples with missing values.
-It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
-Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
-Meanwhile, the performance-optimized plan (Fig.~\ref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
-However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.
 
@@
 \begin{figure}
-  \begin{subfigure}[b]{\linewidth}
   \Tree
@@
   ]
-  \caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for quality ($\alpha=0$).}
+\vspace{0.5\baselineskip}
+\caption{A quality-optimized plan for the query in Fig.~\ref{fig:example-query}.
+$\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
+$\mu$ and $\delta$ are specialized imputation operators (Sec.~\ref{sec:operators}).}
 \label{fig:quality-plan}
-\end{subfigure}
-\begin{subfigure}[b]{\linewidth}
+\end{figure}
+
+\subsection{Planning with \ProjectName{}}
+The multiple possible locations for performing imputation leads to a  search space contains plans of varying performance and quality of results.
+The user can influence the final plan selected by \ProjectName{}
+through a parameter $\alpha \in [0, 1]$, where low $\alpha$ 
+prioritizes the quality of the query results and high $\alpha$ prioritizes performance.
+
+For the query in Fig.~\ref{fig:example-query}, \ProjectName{} generates
+the plans in Fig.~\ref{fig:quality-plan} and~\ref{fig:fast-plan}, when 
+optimizing for quality and performance, respectively.
+
+The quality-optimized plan (Fig.~\ref{fig:quality-plan}) uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
+It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
+Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
+Meanwhile, the performance-optimized plan (Fig.~\ref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
+However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.
+In both cases, only performing
+imputation on the necessary data yields a query that is much faster
+than performing imputation on the whole dataset and then executing
+the query.
+
+\begin{figure}
   \Tree
@@
     [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
-      [.\colorbox{pink}{$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$}
+      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
+        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
+          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
             [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
-          [.$\bowtie_{\text{demo.id} = \text{exams.id}}$
+              [.$\sigma_{\text{exams.weight} \geq 120}$ 
+                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
+              ] 
+              labs 
+            ]
+          ]
         [.$\sigma_{\text{demo.gender} = 2}$ demo ]
-            [.$\sigma_{\text{exams.weight} \geq 120}$ [.\colorbox{pink}{$\delta_{\text{exams.weight}}$} exams ] ] ] labs ] ] ] ]
-\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for performance ($\alpha=1$).}
+      ] 
+    ] 
+  ] 
+  ]
+\caption{A performance-optimized plan for the query in Fig.~\ref{fig:example-query}.
+$\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
+$\mu$ and $\delta$ are specialized imputation operators (Sec.~\ref{sec:operators}).}
 \label{fig:fast-plan}
-\end{subfigure}
-\vspace{0.5\baselineskip}
-\caption{Plans for the query in Fig.~\ref{fig:example-query}. Imputation operators are highlighted.}
 \end{figure}
 
-Returning to our epidemiologist, tens of queries later she has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.
+\subsection{\ProjectName{} Workflow}
+Now that the analyst has \ProjectName{}, she can explore
+the dataset using SQL. For each query, she repeatedly chooses a value
+of $\alpha$ and runs the query until she is unsatisfied with the runtime.
+Rather than commit \emph{a priori} to a runtime bound for her query,
+which may be unsatisfiable, this iterative approach gives her 
+immediate feedback.
+
+Tens of queries later, our epidemiologist has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.
 Knowing that there may be a need to explore this data set further, she can easily incorporate her imputation model into ImputeDB for future use.
+
+%This approach allows her to maximize the query result quality, while
+%maintaining an acceptable runtime.
+%
+%%She indicates her preference for performance or for accuracy by providing a configuration parameter $\alpha$ to ImputeDB in addition to her query.
+%%$\alpha$ is a dimensionless parameter in $[0,1]$, and it indicates the relative importance of the quality of query results and of query performance.
+%So, a high $\alpha$ results in a fast query plan that may compromise accuracy, while a low $\alpha$ results in a high quality query plan that may take longer to run.
+%In our intended workflow, the analyst picks a value of $\alpha$ and incrementally increases it to increase query result quality until the query takes too long to run.
+%Fig.~\ref{fig:quality-plan} and Fig.~\ref{fig:fast-plan} show plans for the example query in Fig.~\ref{fig:example-query} when optimizing for quality and performance, respectively.
+
+%1 - How great life is with the project
+% - intended work flow
 

### 2017-02-25 paper/introduction.tex
paper: reworked motivation section

@@
 We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on the fly, during query execution.
-The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
+\emph{The key insight behind \ProjectName{} is that imputation
+only needs to be performed on the data relevant to a particular query and
+that this subset is generally much smaller than the entire database.}
+
+\todobox{see how this bit fits}{
+The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.}
 

### 2017-02-25 paper/algorithm.tex
paper: reworked motivation section
paper: some proof cleanup

@@
 \begin{definition}
-An abstract imputation operator $I: \mathcal{P}(String) \rightarrow \mathcal{P}(\{ \delta, \mu \})$ is a function that given a dirty set maps to
-a set of imputation operations. For convenience we denote applying the set of transformations to a plan $p$ as $I(p)$ and understand that this
+$I: \mathcal{P}(D) \rightarrow \mathcal{P}(\{ \delta, \mu \})$, where
+$D$ is the set of dirty attributes for a relation, is an abstract imputation operator mapping a set of dirty
+attributes to a set of imputation operations.
+
+For convenience we denote applying the set of imputation operations to a plan $p$ as $I(p)$ and understand that this
 is shorthand for $\{ op(p)   | op \in I(\textsc{Dirty}(p)) \}$.
@@
 
-Algorithm{~\ref{algo:top-level-planner} plans the query in stages. It first handles any pushed-down selections, it then optimizes joins, then groupby/aggregations,
-and finally projections. It maintains all necessary state information through a combination of dirty sets on intermediate plans and
-initializing empty plan caches at stages that require it. Given this, we can succinctly describe possible transformations to a plan at a given
-point in the algorithm solely through $I$.
+We can succinctly describe possible imputation transformations to a plan at a given point in the planning algorithm solely through $I$.
 
@@
 \begin{figure}
+\centering
 \begin{tikzpicture}

### 2017-02-25 paper/examples.tex
paper: minor fix

@@
 
-\todo{fill in the XX values}
 Before she can perform her queries, the researcher must develop a strategy
@@
 approaches have significant drawbacks. For the query pictured in 
-Fig.~\ref{fig:example-query}, (1) drops XXXX percent of the relevant tuples,
+Fig.~\ref{fig:example-query}, (1) drops 1492 potentially relevant tuples,
 while her previous experiences with (2) have proven be expensive. In a previous project, she
 worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
-dataset took XXX hours. The analyst needs a more complete picture
+dataset took \todo{XXX hours}. The analyst needs a more complete picture
 of the data, so (1) is insufficient, and she is under time pressure, so (2)

### 2017-02-26 paper/examples.tex
paper: Add full imputation timing to motivating example

@@
 worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
-dataset took \todo{XXX hours}. The analyst needs a more complete picture
+dataset took around 75 minutes\todo{Here, we can reference the number from our own experiment (done), or the number from Akande et al: ``performing imputation on just a 1.5 percent random sample took 45 minutes''}.
+The analyst needs a more complete picture
 of the data, so (1) is insufficient, and she is under time pressure, so (2)

### 2017-02-26 paper/algorithm.tex
paper: clarified caption for bad pareto situation (sam feedback)

@@
 \end{tikzpicture}
-\caption{A bad Pareto frontier selection would choose $p_2$ and never reach $p_1'$}
+\caption{A bad Pareto frontier selection would choose $p_2$ over $p_1$, as it is locally optimal, but with the available transformations ($I$) $p_2$  can only reach $p_2'$, which is dominated by the inaccessible $p_1'$.
+Theorem~\ref{theorem:no-bad-sit} says such a situation cannot arise when constructing Pareto frontiers on a dirty-set basis.}
 \label{fig:bad-sit-diagram}

### 2017-02-26 paper/experiments.tex
paper: removed experiments for queries 5 and 10

--- 2017-02-24 paper/experiments.tex
@@
 
-Queries 1-5 are on the CDC data. In Query~\ref{q1}, we calculate the average height for
+Queries 1-4 are on the CDC data. In Query~\ref{q1}, we calculate the average height for
 users based on their income data. In Query~\ref{q2}, we compare cholesterol levels for
@@
 years. In Query~\ref{q4}, we calculate the average systolic blood pressure, by gender, for
-subjects with a body mass index indicating obesity. In Query~\ref{q5}, we extract the age,
-gender, triglyceride levels, and waist circumference from individuals with high triglyceride
-levels.
+subjects with a body mass index indicating obesity.
 
-Queries 6-10 are on the FCC data. In Query~\ref{q6}, we calculate the average income for
+Queries 5-8 are on the FCC data. In Query~\ref{q6}, we calculate the average income for
 survey participants, based on their bootcamp attendance. In Query~\ref{q7} estimates the
@@
 countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
-participants based on whether or not they attended a bootcamp. Finally, in Query~\ref{q10},
-we identify individuals from countries with GDP per-capita of \$5000 or less who earn
-\$100,000 or more post-bootcamp.
+participants based on whether or not they attended a bootcamp.
 
@@
 cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
-with performance improvement on a factor of ten.
+with performance improvement on a factor of ten. This performance differential means it is feasible
+to explore multiple imputations operations (including more expensive operators) when using
+\ProjectName{}, in contrast to the traditional approach of base table imputation.
 
@@
 planning. In the median query across a variety of queries and choices of $\alpha$, planning
-constituted 8.5 percent of total runtime by \ProjectName{}.
+constituted 8.5 percent of total runtime by \ProjectName{}. The one-standard-deviation
+intervals around the mean planning time often overlap, suggesting the planning component is
+constant in $\alpha$.
 

### 2017-02-26 paper/algorithm.tex
paper: \ref to \Cref for standardized section names

--- 2017-02-25 paper/algorithm.tex
@@
 
-Our optimizer searches (Sec.~\ref{sec:planning}) a restricted space (Sec.~\ref{sec:search-space}) of query plans containing imputation operators (Sec.~\ref{sec:operators}) for an optimal plan given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.
+Our optimizer searches (\Cref{sec:planning}) a restricted space (\Cref{sec:search-space}) of query plans containing imputation operators (\Cref{sec:operators}) for an optimal plan given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.
 
-This plan is subject to two constraints (Sec.~\ref{sec:placement}).
+This plan is subject to two constraints (\Cref{sec:placement}).
 First, it must not emit any tuples which contain null values, regardless of the state of the base tables.
 Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
-Our planning algorithm is agnostic to the type of imputation used (Sec.~\ref{sec:imputation}), as the properties of the imputation are summarized by the cost model (Sec.~\ref{sec:cost-model}).
+Our planning algorithm is agnostic to the type of imputation used (\Cref{sec:imputation}), as the properties of the imputation are summarized by the cost model (\Cref{sec:cost-model}).
 
@@
 attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
-replace all null values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in Sec.~\ref{sec:imputation}).
+replace all null values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
 \textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
@@
 The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
-Figure~\ref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.
+\Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.
 
@@
 When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
-Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
+\Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
 
@@
 and simply queries it for a time estimate. In the base implementation of our system,
-we use an iterative algorithm (see Sec.~\ref{sec:imputation}), where the runtime cost
+we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
 is a function of the number of attributes with no missing data, number of attributes with
@@
 
-The query planner must select a join ordering in addition to placing imputation operators as described in Section~\ref{sec:placement}.
+The query planner must select a join ordering in addition to placing imputation operators as described in~\Cref{sec:placement}.
 
@@
 We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes and a choice over the plan frontier
-can only be made once the set is final. Details are shown in Figure~\ref{fig:semantics-plan-cache}.
+can only be made once the set is final. Details are shown in~\Cref{fig:semantics-plan-cache}.
 
@@
 
-Algorithm~\ref{algo:plan-helpers} presents a series of helper functions used by the top-level planner, shown in Algorithm~\ref{algo:top-level-planner}.
+\Cref{algo:plan-helpers} presents a series of helper functions used by the top-level planner, shown in~\Cref{algo:top-level-planner}.
 
 The planner operates as follows.
-First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (Lines~\ref{lst:line:attr-start}-\ref{lst:line:attr-end}).
+First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}-\Cref{lst:line:attr-end}).
 This set will be used to determine which attributes are imputed.
 Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
-a search over imputation operations produces various possible plans, which are added to the cache (Line~\ref{lst:line:sel}). If no selections are available, a 
-simple scan is added to the plan cache (Line~\ref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
-plans encompassing all necessary tables is produced (Line~\ref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
-operations (Line~\ref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
+a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}). If no selections are available, a 
+simple scan is added to the plan cache (\Cref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
+plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
+operations (\Cref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
 best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
@@
 
-Theorem~\ref{theorem:no-bad-sit} states that there is no \textit{bad situation} such as that depicted in Figure~\ref{fig:bad-sit-diagram}
+\Cref{theorem:no-bad-sit} states that there is no \textit{bad situation} such as that depicted in~\Cref{fig:bad-sit-diagram}
 
@@
 \end{tikzpicture}
-\caption{A bad Pareto frontier selection would choose $p_2$ and never reach $p_1'$}
+\caption{A bad Pareto frontier selection would choose $p_2$ over $p_1$, as it is locally optimal, but with the available transformations ($I$) $p_2$  can only reach $p_2'$, which is dominated by the inaccessible $p_1'$.~\Cref{theorem:no-bad-sit} says such a situation cannot arise when constructing Pareto frontiers on a dirty-set basis.}
 \label{fig:bad-sit-diagram}

### 2017-02-26 paper/experiments.tex
paper: \ref to \Cref for standardized section names

@@
 tests results or alternative metrics. In order to simplify the presentation and
-exploration of queries, we focused on a subset of the attributes for each table.
-Table~\ref{table:nhanes-description} shows the attributes selected, along with the
+exploration of queries, we focused on a subset of the attributes for each table.~\Cref{table:nhanes-description}
+shows the attributes selected, along with the
 percentage of nulls. It is worth noting that for readability, we have replaced the
@@
 expected. The dataset we started with had 15,620 rows and 113
-columns. We chose 17 attributes, which are outlined in Table~\ref{table:fcc-description}, along with the percentage of missing values.
+columns. We chose 17 attributes, which are outlined in~\Cref{table:fcc-description}, along with the percentage of missing values.
 
@@
 \subsection{Queries}
-We collect a set of queries (Table~\ref{fig:queries}) that we think are both representative,
+We collect a set of queries (\Cref{fig:queries}) that we think are both representative,
 in that they could reasonably be written by a user in the course of data analysis, and
@@
 
-Queries 1-4 are on the CDC data. In Query~\ref{q1}, we calculate the average height for
-users based on their income data. In Query~\ref{q2}, we compare cholesterol levels for
-individuals with low, medium, and high incomes spectrum and above a certain weight. In
-Query~\ref{q3}, we extract the maximum blood lead levels for children under the age of 6
-years. In Query~\ref{q4}, we calculate the average systolic blood pressure, by gender, for
+Queries 1-4 are on the CDC data. In~\Cref{q1}, we calculate the average height for
+users based on their income data. In~\Cref{q2}, we compare cholesterol levels for
+individuals with low, medium, and high incomes spectrum and above a certain weight. In~\Cref{q3}, 
+we extract the maximum blood lead levels for children under the age of 6
+years. In~\Cref{q4}, we calculate the average systolic blood pressure, by gender, for
 subjects with a body mass index indicating obesity.
 
-Queries 5-8 are on the FCC data. In Query~\ref{q6}, we calculate the average income for
-survey participants, based on their bootcamp attendance. In Query~\ref{q7} estimates the
-average age of women from the United States who participated. In Query~\ref{q8}, we
+Queries 5-8 are on the FCC data. In~\Cref{q5}, we calculate the average income for
+survey participants, based on their bootcamp attendance. In~\Cref{q6} estimates the
+average age of women from the United States who participated. In~\Cref{q7}, we
 calculate the average amount of money survey participants with student debt destined to
-learning based on their school degree. In Query~\ref{q9}, we join the FCC data with a
+learning based on their school degree. In~\Cref{q8}, we join the FCC data with a
 reference table provided by the World Bank which summarizes GDP per-capita across various
@@
 
-Figure~\ref{fig:runtimes} provides a summary of the performance results. The first plot
+\Cref{fig:runtimes} provides a summary of the performance results. The first plot
 shows average running times for \ProjectName{} plans in two configurations:
@@
 \caption{Comparing \ProjectName{} runtimes for each query
-    (\ref{subfig:project-runtime-queries}) with the runtime for a query when base-table
-    imputations are used (\ref{subfig:base-runtime-queries})}
+    (\Cref{subfig:project-runtime-queries}) with the runtime for a query when base-table
+    imputations are used (\Cref{subfig:base-runtime-queries})}
 \label{fig:runtimes}
@@
 
-Figure~\ref{fig:plantimes} provides a summary of the planning times for each of the queries.
+\Cref{fig:plantimes} provides a summary of the planning times for each of the queries.
 We exclude the planning time for queries that impute at base table, as that requires no
@@
 
-Table~\ref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
 \ProjectName{}'s query results when compared to running imputation on the base tables and
@@
 To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
-aggregate query over the ACS dataset in our final experiment(Figure~\ref{query-acs}).
+aggregate query over the ACS dataset in our final experiment(\Cref{query-acs}).
 Applying the imputation operation to the base table is extremely expensive, in our case

### 2017-02-26 paper/introduction.tex
paper: \ref to \Cref for standardized section names

@@
 are similar in quality to those obtained with the base-table imputation approach, as indicated by low symmetric-Mean-Absolute-Percentage-Errors in our empirical results
-over real-world data sets (see Sec.~\ref{sec:experiments} for details).
+over real-world data sets (see~\Cref{sec:experiments} for details).
 

### 2017-02-26 paper/experiments.tex
paper: removed planning times, just left in text, added todo for table of SMAPE

@@
 
-\Cref{fig:plantimes} provides a summary of the planning times for each of the queries.
-We exclude the planning time for queries that impute at base table, as that requires no
-planning. In the median query across a variety of queries and choices of $\alpha$, planning
-constituted 8.5 percent of total runtime by \ProjectName{}. The one-standard-deviation
-intervals around the mean planning time often overlap, suggesting the planning component is
-constant in $\alpha$.
+%\Cref{fig:plantimes} provides a summary of the planning times for each of the queries.
+%We exclude the planning time for queries that impute at base table, as that requires no
+%planning.
 
-\begin{figure}
-\includegraphics[width=\columnwidth]{figures/planning_times_imputedb.png}
-\caption{Planning times for each query}
-\label{fig:plantimes}
-\end{figure}
+In the median query across a variety of queries and choices of $\alpha$, planning
+constituted 8.5 percent of total runtime. In all cases, the optimizer
+returned a query plan within 40ms, with times roughly constant between levels of $\alpha$.
+
+% The one-standard-deviation
+%intervals around the mean planning time often overlap, suggesting the planning component is
+%constant in $\alpha$.
+
+%\begin{figure}
+%\includegraphics[width=\columnwidth]{figures/planning_times_imputedb.png}
+%\caption{Planning times for each query}
+%\label{fig:plantimes}
+%\end{figure}
 
@@
 \begin{table}
+\todo{Consider switching to graph, or changing table layout..not very intuitive right now.}
 \centering

### 2017-02-26 simpledb/test/experiments/run_experiments.py
Add alt experiment runner

@@ -40,6 +40,18 @@ def run_large_experiment():
 
     run_experiment(this_output_dir, iters, min_alpha, max_alpha, step)
 
+def run_alt_experiment():
+    this_output_dir = os.path.join(output_dir, "alt")
+
+    iters     = 220
+    min_alpha = 0.00
+    max_alpha = 1.00
+    step      = 0.05
+
+    queries = "queries1.txt"
+
+    run_experiment(this_output_dir, iters, min_alpha, max_alpha, step)
+
 def run_acs_experiment():
     this_output_dir = os.path.join(output_dir, "acs")
 
@@ -96,6 +108,8 @@ if __name__ == "__main__":
         run_medium_experiment()
     elif experiment_size == "large":
         run_large_experiment()
+    elif experiment_size == "alt":
+        run_alt_experiment()
     elif experiment_size == "acs":
         run_acs_experiment()
     else:

### 2017-02-26 paper/examples.tex

@@
 to get a quick and accurate view into the data. However,
-the data has been collected through CDC surveys (see Sec.~\ref{subsec:datasets}),
+the data has been collected through CDC surveys (see~\Cref{subsec:datasets}),
 and there is a significant amount of missing data across all
@@
 2) she could use a standard data imputation package on all of her data. Both of these
-approaches have significant drawbacks. For the query pictured in 
-Fig.~\ref{fig:example-query}, (1) drops 1492 potentially relevant tuples,
+approaches have significant drawbacks. For the query pictured in ~\Cref{fig:example-query}, 
+(1) drops 1492 potentially relevant tuples,
 while her previous experiences with (2) have proven be expensive. In a previous project, she
@@
 query in standard SQL and automatically performs the imputations necessary to fill in the missing data.
-An example query is shown in Fig.~\ref{fig:example-query}.
+An example query is shown in~\Cref{fig:example-query}.
 
@@
 \vspace{0.5\baselineskip}
-\caption{A quality-optimized plan for the query in Fig.~\ref{fig:example-query}.
+\caption{A quality-optimized plan for the query in~\Cref{fig:example-query}.
 $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
-$\mu$ and $\delta$ are specialized imputation operators (Sec.~\ref{sec:operators}).}
+$\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
 \label{fig:quality-plan}
@@
 
-For the query in Fig.~\ref{fig:example-query}, \ProjectName{} generates
-the plans in Fig.~\ref{fig:quality-plan} and~\ref{fig:fast-plan}, when 
+For the query in~\Cref{fig:example-query}, \ProjectName{} generates
+the plans in~\Cref{fig:quality-plan} and~\Cref{fig:fast-plan}, when 
 optimizing for quality and performance, respectively.
 
-The quality-optimized plan (Fig.~\ref{fig:quality-plan}) uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
+The quality-optimized plan (\Cref{fig:quality-plan}) uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
 It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
 Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
-Meanwhile, the performance-optimized plan (Fig.~\ref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
+Meanwhile, the performance-optimized plan (\Cref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
 However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.
@@
   ]
-\caption{A performance-optimized plan for the query in Fig.~\ref{fig:example-query}.
+\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}.
 $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
-$\mu$ and $\delta$ are specialized imputation operators (Sec.~\ref{sec:operators}).}
+$\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
 \label{fig:fast-plan}

### 2017-02-26 paper/introduction.tex

@@
 
-\todobox{see how this bit fits}{
-The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.}
-
 % Prior work
@@
 and express preferences in quality of data imputation versus running time.
+The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
 

### 2017-02-27 paper/main.tex
paper: added todos
paper: check biblio, add nocite, add a bit to intro for importance of data cleaning, remove an old reference to acs
paper fix: added option to hyperref package to avoid error with urls split across lines
paper: fixed general typos, added clearer todos
paper: added some preliminary experiments
paper: pass over motivating examples section
paper: partial pass over algorithm section
Fix whitespace in query
paper: working on optimality proof
paper: cleaned up proof, first pass
paper: intro feedback from martin
paper: abstract (phillip feedback)

@@
 \usepackage{siunitx}
-\usepackage{hyperref} % load after other packages
 \usepackage{graphicx} % includegraphics
@@
 \usepackage{qtree} % for simple tree diagrams
+\usepackage{caption}
 \usepackage{subcaption}
+\usepackage{xcolor}
+\usepackage{hyperref} % load after other packages
+\usepackage{tikz}
+\usetikzlibrary{positioning}
+\usepackage{cleveref}
+
 \addbibresource{references.bib}
@@
 \newcommand{\ProjectName}{ImputeDB}
+\newcommand{\doms}[2]{#1 \succ #2}
+
+% definitions/theorems etc
+\newtheorem{definition}{Definition}
+\newtheorem{theorem}{Theorem}
+\newtheorem{case}{Case}
+
+% new section names for cref
+\Crefname{query}{Query}{Query}
 
@@
   Jos\'e Cambronero \\
-  \texttt{jcamsan@mit.edu}
+  \texttt{jcamsan@csail.mit.edu}
   \and
@@
   \and
-  Micah Smith \\
+  Micah J. Smith \\
   \texttt{micahs@mit.edu}}
@@
 \newcommand{\todo}[1]{{\color{red}\textbf{TODO:} #1}}
-
+\newcommand{\todobox}[2]{{\color{red}\fbox{\parbox{\columnwidth}{\textbf{TODO:} #1\\#2}}}}
 \begin{document}
@@
 \begin{abstract}
-Missing values abound in data analysis and often present a serious challenge for users, who
+Missing values abound in data analysis and often present a serious challenge for usability of data. Users
 are forced to pick between removing tuples with missing values or creating a clean duplicate of their data
-by  applying a potentially expensive imputation strategy. We show that it is possible
+by  applying a potentially-expensive imputation strategy. We show that it is possible
 to incorporate imputation into a cost-based query optimizer, performing imputation on the fly
 for each query. This allows users to immediately explore
-their data, while the system picks the optimal placement of imputation operations based
+their data, while the system picks the optimal placement of imputation operations. This placement is based
 on the specific query issued and a parameter that the user selects for evaluating the tradeoff between data
@@
 for traditional optimizations to take place. Our experiments evaluate this approach on two real-world survey-based datasets
-from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}, showing that the 
-optimizer generates query plans that execute in a fraction of the time that it takes to impute values at the base
-tables, making exploratory data analysis using our system feasible. Furthermore, we show that the query results
-from on-the-fly imputation are comparable to those generated with the standard full-table imputation approach.
+from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}. Our results show that the optimizer
+generates query plans that execute between \todo{add [low x - high x] range} as fast as imputing at the base
+tables and then executing queries, making exploratory data analysis using our system feasible. Furthermore, we show that the query results
+from on-the-fly imputation have Symmetric-Mean-Absolute-Percentage-Error between \todo{add SMAPE range}  
+when compared to standard full-table imputation approach.
 \end{abstract}
@@
 \input{conclusion}
-
 \balance
+\nocite{*}
 \printbibliography

### 2017-02-27 paper/introduction.tex
paper: introduction (phillip feedback)

@@
 % Context
-Handling incorrect or dirty data is a complex and challenging problem for analysts.
+Handling incorrect or dirty data is a complex and challenging problem for analysts\todo{add significant citations here}.
 % Problem
 One way in which a dataset can be dirty is for parts of it to be missing altogether.
-Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
+Missing data, if handled naively, can cause analyses to be incorrect.
 According to some estimates, data scientists spend between 50 and 80 percent of their time performing tasks such as collecting and cleaning datasets~\cite{data-science-cleaning}.
@@
 This process is called \emph{imputation}.
-We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on the fly, during query execution.
+We have developed a system called \ProjectName{}, which is the first database to use a query optimization algorithm which allows imputation to occur on the fly, during query execution.
 \emph{The key insight behind \ProjectName{} is that imputation
@@
 that this subset is generally much smaller than the entire database.}
-
-% Prior work
-There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
-Prior work in statistics has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
-Prior work in databases has a established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
-Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
-\ProjectName{} contributes a novel approach to planning data imputation on a per query basis, allowing users to use standard SQL on real-world data sets with missing values
-and express preferences in quality of data imputation versus running time.
-The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
 
@@
 
+% Prior work
+There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
+Prior work in statistics\todo{add more specific subarea qualifier} has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
+Prior work in databases has a established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
+Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
+\ProjectName{} contributes the first optimizer to plan data imputation on a per query basis, allowing users to use standard SQL on real-world data sets with missing values
+and express preferences in quality of data imputation versus running time.
+The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
+
 % Implications
 Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
-queries that execute in a fraction of the time it takes to impute base tables and then run queries (the traditional approach). Furthermore, the results obtained with this approach
-are similar in quality to those obtained with the base-table imputation approach, as indicated by low symmetric-Mean-Absolute-Percentage-Errors in our empirical results
+queries that execute in a fraction of the time\todo{add actual performance figures here} it takes to impute base tables and then run queries (the traditional approach).
+Furthermore, the results obtained with this approach
+are similar in quality to those obtained with the base-table imputation approach, as indicated by low\todo{add actual value here} Symmetric-Mean-Absolute-Percentage-Errors in our empirical results
 over real-world data sets (see~\Cref{sec:experiments} for details).
@@
   We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
-  Impute fills in missing data values using a regression model (CE-CART)~\cite{burgette2010multiple}.
-  Drop simply drops tuples which contain null values.
+  Impute operation fills in missing data values using a regression model (CE-CART)~\cite{burgette2010multiple}.
+  Drop operation simply drops tuples which contain null values.
 \item \textbf{Model of imputation quality and cost:}
@@
 \item \textbf{Query planning with imputation:}
-  We present a novel query planning algorithm which jointly optimizes for running time and the quality of the results.
+  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
   It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.

### 2017-02-27 paper/examples.tex
paper: example (phillip feedback)

@@
 \section{Motivating Example}
-In this section, we present a common scenario where a scientists
-has to perform analysis on real-world data.
-
-\subsection{Analysis with missing data}
 An epidemiologist at the CDC is tasked with an 
@@
 controlling for factors such as weight and gender.
-
-%While careful modeling will be necessary before putting
-%results into the final research paper,
 
@@
 
-Before she can perform her queries, the researcher must develop a strategy
-for handling the missing values. She currently has two options:
+Before she can perform her queries, the epidemiologist must develop a strategy
+for handling missing values. She currently has two options:
 1) she could drop records with missing values in relevant fields,
@@
 (1) drops 1492 potentially relevant tuples,
-while her previous experiences with (2) have proven be expensive. In a previous project, she
-worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
-dataset took around 75 minutes\todo{Here, we can reference the number from our own experiment (done), or the number from Akande et al: ``performing imputation on just a 1.5 percent random sample took 45 minutes''}.
-The analyst needs a more complete picture
+while her previous experiences with (2) have proven be expensive. The epidemiologist needs a more complete picture
 of the data, so (1) is insufficient, and she is under time pressure, so (2)
@@
 \end{lstlisting}
-\caption{A typical public health query on CDC's NHANES data}
+\caption{A typical public health query on CDC's NHANES data.}
 \label{fig:example-query}
@@
 \caption{A quality-optimized plan for the query in~\Cref{fig:example-query}.
-$\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
+The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
 $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
@@
 The user can influence the final plan selected by \ProjectName{}
-through a parameter $\alpha \in [0, 1]$, where low $\alpha$ 
+through a tradeoff parameter $\alpha \in [0, 1]$, where low $\alpha$ 
 prioritizes the quality of the query results and high $\alpha$ prioritizes performance.
@@
 
-The quality-optimized plan (\Cref{fig:quality-plan}) uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
+\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
 It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
 Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
-Meanwhile, the performance-optimized plan (\Cref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
+Meanwhile, \Cref{fig:fast-plan}'s performance-optimized plan uses the drop operation $\delta$ instead of filling in missing values.
 However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.
@@
   ]
-\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}.
-$\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
+\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}.The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
 $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
@@
 \subsection{\ProjectName{} Workflow}
-Now that the analyst has \ProjectName{}, she can explore
+Now that the epidemiologist has \ProjectName{}, she can explore
 the dataset using SQL. For each query, she repeatedly chooses a value
-of $\alpha$ and runs the query until she is unsatisfied with the runtime.
+for $\alpha$ and runs the query until she is unsatisfied with the runtime.
 Rather than commit \emph{a priori} to a runtime bound for her query,

### 2017-02-27 paper/introduction.tex
paper: example (phillip feedback)

@@
 significant speedup in query execution. The focus on of our work is not on the imputation algorithm but rather on the planning involved in its use in non-trivial query plans.
+
+\todobox{find appropriate location for this}
+{In a previous project, she
+worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
+dataset took around 75 minutes\todo{Here, we can reference the number from our own experiment (done), or the number from Akande et al: ``performing imputation on just a 1.5 percent random sample took 45 minutes''}.}
 

### 2017-02-27 paper/conclusion.tex
paper: conclusion (phillip feedback)

@@
 \section{Conclusion}
-As shown in \ProjectName{}, missing values and their imputation can be successfully integrated into the relational calculus and
+Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational calculus and
 existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine

### 2017-02-27 paper/introduction.tex
paper: add references to intro

--- 2017-02-26 paper/introduction.tex
@@
 % Problem
-One way in which a dataset can be dirty is for parts of it to be missing altogether.
-Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
+One way in which a dataset can be dirty is for parts of it to be missing altogether~\cite{kim2003}.
+Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect~\cite{rubin1976}.
 According to some estimates, data scientists spend between 50 and 80 percent of their time performing tasks such as collecting and cleaning datasets~\cite{data-science-cleaning}.
@@
 This process is called \emph{imputation}.
-We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on the fly, during query execution.
+We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on-the-fly, during query execution.
 \emph{The key insight behind \ProjectName{} is that imputation

### 2017-02-27 paper/algorithm.tex

@@
 
+\todobox{this jumps around too much}{
 Our optimizer searches (\Cref{sec:planning}) a restricted space (\Cref{sec:search-space}) of query plans containing imputation operators (\Cref{sec:operators}) for an optimal plan given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.
@@
 Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
-Our planning algorithm is agnostic to the type of imputation used (\Cref{sec:imputation}), as the properties of the imputation are summarized by the cost model (\Cref{sec:cost-model}).
+Our planning algorithm is agnostic to the type of imputation used (\Cref{sec:imputation}), as the properties of the imputation are summarized by the cost model (\Cref{sec:cost-model}).}
 
@@
 replace all null values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
+In our experiments, we instantiate \textit{Impute} with the CE-CART\cite{burgette2010multiple} algorithm, but we highlight that
+our system is agnostic to this choice.
 \textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
@@
 
-\item A grouping \& aggregation operator is optional, and if present will be performed after the final join.
+\item A grouping and aggregation operator is optional and if present will be performed after the final join.
 
@@
 
-In order to pick a final plan from the frontier, our model introduces a parameter $\alpha$ which encodes an upper bound on the $\textsc{Loss}$ value that the user is willing
+In order to pick a final plan from the frontier, our model introduces a tradeoff parameter $\alpha$ which encodes an upper bound on the $\textsc{Loss}$ value that the user is willing
 to tolerate relative to the minimum possible amongst frontier plans.
@@
 Scan costs are estimated based on the number of tuples, 
-page size, and an unit-less IO cost per page. Join costs are estimated as a function
+page size, and an unit-less I/O cost per page. Join costs are estimated as a function
 of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
@@
 account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
-complexity can be expressed as a function of the number of heap pages read and the IO cost
+complexity can be expressed as a function of the number of heap pages read and the I/O cost
 per page.
@@
 complicated, as the properties of
-different algorithms can vary significantly, the underlying IO cost
+different algorithms can vary significantly, the underlying I/O cost
 is not easily extracted (especially in the case that the entire stream of tuples does not
@@
 \subsection{Complexity}
+\todo{Stengthen complexity argument, as per Sam's feedback}
 Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our planning algorithm still operates in exponential time. 

### 2017-02-27 paper/experiments.tex

@@
 for two separate survey-based data sets, showing that our system
-is ideal for early dataset exploration.
+is well suited for early dataset exploration.
 
@@
 \subsubsection{Center for Disease Control: Surveys}
-For our first set of experiments, we use survey data collected by the 
-Center for Disease Control and Prevention (CDC) in the United States. In particular, we
-experiment on a set of tables collected as part of the National
+We use survey data collected by the 
+Center for Disease Control and Prevention (CDC) in the United States. We
+experiment on a set of tables collected as part of the 2013-2014 National
 Health and Nutrition Examination Survey (NHANES), a series of studies
@@
 The data consists on survey responses, physical examinations, and laboratory
-results, amongst others. Our dataset corresponds to the 2013-2014 NHANES
-and is available for direct download on the CDC's organization site on 
+results, amongst others. Our dataset is available for direct download on the CDC's organization site on 
 kaggle.com, a popular data science website.
 
-There are a total of 6 tables associated with the NHANES data set. For purposes
-of our experiments, we use three tables:
+There are a total of 6 tables associated with the NHANES data set. We use three tables:
 
@@
     \end{subtable}
-    \caption{Missing values in CDC NHANES 2013-2014 data}
+    \caption{Missing value distribution within each table/attribute in CDC NHANES 2013-2014 data}
     \label{table:nhanes-description} 
@@
 For our purposes, we use a version of the data that has been 
-pre-processed, but where plenty of missing values remain. It is
+pre-processed, but where many missing values remain. It is
 worth noting that given the table design (where many attributes
@@
     \input{tables/fcc}
-    \caption{Missing values in FCC Survey Data}
+    \caption{Missing values in fCC Survey Data}
    \label{table:fcc-description} 
@@
 \subsection{Queries}
-We collect a set of queries (\Cref{fig:queries}) that we think are both representative,
-in that they could reasonably be written by a user in the course of data analysis, and
-interesting to plan.
+We collect a set of queries (\Cref{fig:queries}) that we think are both representative and interesting
+to plan. They could reasonably be written by a user in the course of data analysis.
 
@@
 
-Queries 5-8 are on the FCC data. In~\Cref{q5}, we calculate the average income for
+Queries 5-8 are on the fCC data. In~\Cref{q5}, we calculate the average income for
 survey participants, based on their bootcamp attendance. In~\Cref{q6} estimates the
@@
 calculate the average amount of money survey participants with student debt destined to
-learning based on their school degree. In~\Cref{q8}, we join the FCC data with a
+learning based on their school degree. In~\Cref{q8}, we join the fCC data with a
 reference table provided by the World Bank which summarizes GDP per-capita across various
@@
  \input{queries-fcc}
- \caption{Queries on FCC data}
+ \caption{Queries on fCC data}
  \label{fig:queries-fcc}
@@
 
-\Cref{fig:runtimes} provides a summary of the performance results. The first plot
+\Cref{fig:runtimes} shows a summary of the performance results. The first plot
 shows average running times for \ProjectName{} plans in two configurations:
@@
 cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
-with performance improvement on a factor of ten. This performance differential means it is feasible
+with performance improvement on a factor of ten\todo{give range of performance}. This performance differential means it is feasible
 to explore multiple imputations operations (including more expensive operators) when using
@@
 \begin{figure}
+\todo{change this to a single graph of X-times faster than base approach}
 \begin{subfigure}{\linewidth}
@@
 \ProjectName{}'s query results when compared to running imputation on the base tables and
-executing the query on the cleaned data. Each query was run 200 iterations in both settings
+executing the query on the cleaned data. Each query ran for 200 iterations in both settings
 and then results from the two approaches were paired up and compared tuple-wise. We average
@@
 reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
-approach are low in all cases, indicating that on-the-fly imputation produces similar
+approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces similar
 results to imputation at the base tables.

### 2017-02-27 paper/introduction.tex

--- 2017-02-27 paper/introduction.tex
@@
 % Problem
-One way in which a dataset can be dirty is for parts of it to be missing altogether.
-Missing data, if handled naively, can cause analyses to be incorrect.
+One way in which a dataset can be dirty is for parts of it to be missing altogether~\cite{kim2003}.
+Missing data, if handled naively, can cause analyses to be incorrect~\cite{rubin1976}.
 According to some estimates, data scientists spend between 50 and 80 percent of their time performing tasks such as collecting and cleaning datasets~\cite{data-science-cleaning}.
@@
 This process is called \emph{imputation}.
-We have developed a system called \ProjectName{}, which is the first database to use a query optimization algorithm which allows imputation to occur on the fly, during query execution.
+We have developed a system called \ProjectName{}, which is the first database to use a query optimization algorithm which allows imputation to occur on-the-fly, during query execution.
 \emph{The key insight behind \ProjectName{} is that imputation

### 2017-02-27 paper/experiments.tex
paper: pass over experiments section

@@
+\newcommand{\demorows}{10175}
+\newcommand{\labexrows}{9813}
+
 \section{Experiments}\label{sec:experiments}
@@
 \subsection{Data sets} \label{subsec:datasets}
-\subsubsection{Center for Disease Control: Surveys}
+\subsubsection{CDC NHANES}
 We use survey data collected by the 
-Center for Disease Control and Prevention (CDC) in the United States. We
-experiment on a set of tables collected as part of the 2013-2014 National
+Centers for Disease Control and Prevention (CDC) in the United States. We
+experiment on a set of tables collected as part of the 2013--2014 National
 Health and Nutrition Examination Survey (NHANES), a series of studies
-conducted by the CDC on a national sample of several thousand individuals\cite{cdc-data}.
+conducted by the CDC on a national sample of several thousand individuals~\cite{cdc-data}.
 The data consists on survey responses, physical examinations, and laboratory
-results, amongst others. Our dataset is available for direct download on the CDC's organization site on 
-kaggle.com, a popular data science website.
+results, amongst others.
 
-There are a total of 6 tables associated with the NHANES data set. We use three tables:
+There are 6 tables in the NHANES data set. We use three tables for our experiments:
 
 \begin{itemize}
-	\item Demographics: outlines demographic information of subjects
-	\item Examinations: physical exam results
-	\item Laboratory: laboratory exam results
+	\item \emph{Demographics}: demographic information of subjects
+	\item \emph{Examinations}: physical exam results
+	\item \emph{Laboratory}: laboratory exam results
 \end{itemize}
 
-The original tables have a large number of attributes, in some cases providing more granular
-tests results or alternative metrics. In order to simplify the presentation and
-exploration of queries, we focused on a subset of the attributes for each table.~\Cref{table:nhanes-description}
-shows the attributes selected, along with the
-percentage of nulls. It is worth noting that for readability, we have replaced the
-NHANES variable names with self-explanatory attribute names.
+The original tables have a large number of attributes, in some cases providing more granular tests results or alternative metrics.
+We focused on a subset of the attributes for each table to simplify the presentation and exploration of queries.
+\Cref{table:nhanes-description} shows the attributes selected, along with the percentage of null values for each attribute.
+For readability, we have replaced the NHANES variable names with self-explanatory attribute names.
 
@@
     \input{tables/cdc_demo}
-        \caption{Demographics}
+    \caption{Demographics. \demorows{} rows.}
   \end{subtable}
+  \par\medskip
   \begin{subtable}{0.5\textwidth}
@@
     \input{tables/cdc_labs}
-        \caption{Laboratory Results}
+    \caption{Laboratory Results. \labexrows{} rows.}
   \end{subtable}
+  \par\medskip  
   \begin{subtable}{0.5\textwidth}
@@
     \input{tables/cdc_exams}
-        \caption{Physical Results}
+    \caption{Physical Results. \labexrows{} rows.}
   \end{subtable}
-    \caption{Missing value distribution within each table/attribute in CDC NHANES 2013-2014 data}
-    \label{table:nhanes-description} 
+  \par\medskip  
+  \caption{Missing value distribution for each table/attribute in CDC NHANES 2013--2014 data.}\label{table:nhanes-description} 
 \end{table}
 
-All tables used consists of 10 attributes. The laboratory and examinations data,
-referred to as \textit{labs} and \textit{exams} in our queries, consist of 9813 rows,
-while the demographics data (\textit{demo}) have 10175 rows.
-
-\subsubsection{freeCodeCamp: 2016 New Coder Survey}
+\subsubsection{freeCodeCamp 2016 New Coder Survey}
 For our second set of experiments, we chose to use data collected
 by freeCodeCamp as a part of a survey of new software developers
-(both professional and amateur)\cite{fcc-data}. freeCodeCamp is an open-source
-community that intends to help users learn how to program. In addition,
-to providing training material for users, it also pairs users with
-nonprofits to develop software solutions. Their \textit{2016 New Coder Survey} consists of responses by over 15,000 people to 48 different
-questions. The survey targeted users who were in one way or another
-related to coding organizations.
+(both professional and amateur)~\cite{fcc-data}. freeCodeCamp is an open-source
+community that helps users learn how to program. Their \textit{2016 New Coder Survey} consists of responses by over 15,000 people to 48 different
+demographic and programming-related questions.
+The survey targeted users who were related to coding organizations.
 
-For our purposes, we use a version of the data that has been 
-pre-processed, but where many missing values remain. It is
-worth noting that given the table design (where many attributes
-are related), many of these missing values are to be
-expected. The dataset we started with had 15,620 rows and 113
-columns. We chose 17 attributes, which are outlined in~\Cref{table:fcc-description}, along with the percentage of missing values.
+We use a version of the data that has been pre-processed, but where missing values remain.
+\todobox{Give example of expected missing value}{However, many of the missing values are expected because the data has been de-normalized.}
+The original dataset has 15,620 rows and 113 attributes.
+We chose a subset of 17 attributes, shown in~\Cref{table:fcc-description}, along with the percentage of missing values.
 
@@
 
+\subsubsection{American Community Survey}
 For our final experiment, we introduce a simple aggregate query over data from the American Community Survey (ACS), which
@@
 We used a cleaned version of the 2012 Public Use Microdata Sample (PUMS) data kindly provided by the authors of~\cite{akande2015empirical}.
-Given that the data had been cleaned, we uniformly introduced missing values across 40 percent of the values. The final dataset consists
-of 671,153 rows and 37 integer columns.
+Given that the data had been cleaned, we artificially dirtied it by removing 40\% of the values uniformly at random.
+The final dataset consists of 671,153 rows and 37 integer columns.
 
 \subsection{Queries}
-We collect a set of queries (\Cref{fig:queries}) that we think are both representative and interesting
-to plan. They could reasonably be written by a user in the course of data analysis.
+We collected a set of queries (\Cref{fig:queries}) that we think are both representative and interesting
+to plan. We believe that they could reasonably be written by a user in the course of data analysis.
 
-The queries on the CDC NHANES data consist not only of projections and selections,
-but also interesting joins and aggregates. Our aim was to craft meaningful queries
-that would provide performance figures relevant to practitioners who may use
+The queries on the CDC NHANES data consist not only of projections and selections, but also interesting joins and aggregates.
+Our aim was to craft meaningful queries
+that would provide performance figures relevant to practitioners using
 similar datasets.

### 2017-02-27 paper/experiments.tex
paper: pass over experiments section

@@
 \subsection{Data sets} \label{subsec:datasets}
+We collected three data sets our experiments.
+For all data sets, we selected a subset of the original attributes.
+We also enumerated strings and encoded them with an appropriate integer value.
+\todo{describe all data transformations. jose?}
+
 \subsubsection{CDC NHANES}
-We use survey data collected by the 
+For our first set of experiments, we use survey data collected by the 
 Centers for Disease Control and Prevention (CDC) in the United States. We
@@
   \input{tables/fcc}
-    \caption{Missing values in fCC Survey Data}
-   \label{table:fcc-description} 
+  \caption{Missing values in freeCodeCamp Survey Data}\label{table:fcc-description} 
 \end{table}
@@
 \subsection{Queries}
-We collected a set of queries (\Cref{fig:queries}) that we think are both representative and interesting
-to plan. We believe that they could reasonably be written by a user in the course of data analysis.
+We collected a set of queries (\Cref{fig:queries}) that we think are interesting to plan.
+We believe that they could reasonably be written by a user in the course of data analysis.
 
@@
 
-Queries 5-8 are on the fCC data. In~\Cref{q5}, we calculate the average income for
-survey participants, based on their bootcamp attendance. In~\Cref{q6} estimates the
-average age of women from the United States who participated. In~\Cref{q7}, we
-calculate the average amount of money survey participants with student debt destined to
-learning based on their school degree. In~\Cref{q8}, we join the fCC data with a
-reference table provided by the World Bank which summarizes GDP per-capita across various
-countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
-participants based on whether or not they attended a bootcamp.
-
-Note that for all queries we have enumerated strings and encoded them with
-an appropriate integer value.
+Queries 5-8 are on the freeCodeCamp data.
+\Cref{q5} calculates the average income for survey participants, based on their bootcamp attendance.
+\Cref{q6} estimates the average age of women from the United States who participated.
+\Cref{q7} calculates the average amount of money survey participants with student debt spent on learning based on their school degree.
+\Cref{q8} joins the freeCodeCamp data with a reference table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
+The query calculates the average GDP per-capita of countries with and without bootcamp participants.
 
 \begin{table*}
+  \todo{do we really need headers on these tables?}
   \centering
@@
     \input{queries-cdc}
-  \caption{Queries on CDC data}
-  \label{fig:queries-cdc}
+    \caption{Queries on CDC data}\label{fig:queries-cdc}
   \end{subtable}
- ~
+  \par\medskip
   \begin{subtable}{\linewidth}
     \input{queries-fcc}
- \caption{Queries on fCC data}
- \label{fig:queries-fcc}
+    \caption{Queries on freeCodeCamp data}\label{fig:queries-fcc}
   \end{subtable}
-  \caption{Queries used in our experiments.}
-  \label{fig:queries}
+  \par\medskip  
+  \caption{Queries used in our experiments.}\label{fig:queries}
 \end{table*}
@@
 All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
-four 2.9 GHz Intel Xeon E5-2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.
+four 2.9 GHz Intel Xeon E5--2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.
 
@@
     \includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
-\caption{Query runtimes with \ProjectName{}}
-\label{subfig:project-runtime-queries}
+    \caption{Query runtimes with \ProjectName{}}\label{subfig:project-runtime-queries}
   \end{subfigure}
@@
     \includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
-\caption{Query runtimes with imputation on base tables}
-\label{subfig:base-runtime-queries}
+    \caption{Query runtimes with imputation on base tables}\label{subfig:base-runtime-queries}
   \end{subfigure}
+  \par\medskip
   \caption{Comparing \ProjectName{} runtimes for each query
     (\Cref{subfig:project-runtime-queries}) with the runtime for a query when base-table
-    imputations are used (\Cref{subfig:base-runtime-queries})}
-\label{fig:runtimes}
+    imputations are used (\Cref{subfig:base-runtime-queries})}\label{fig:runtimes}
 \end{figure}
@@
 
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
-\ProjectName{}'s query results when compared to running imputation on the base tables and
-executing the query on the cleaned data. Each query ran for 200 iterations in both settings
-and then results from the two approaches were paired up and compared tuple-wise. We average
-tuple-wise absolute percentage deviations within each iteration of a query, and we report
-this value averaged over all iterations.  We can see that optimizing for quality indeed
-reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
-approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces similar
-results to imputation at the base tables.
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running imputation on the base tables and executing the query on the cleaned data \todo{SMAPE citation}.
+Each query ran for 200 iterations \todo{iterations?
+is this for the imputation or is it the number of query re-runs?} in both settings and then results from the two approaches were paired up and compared tuple-wise.
+We average tuple-wise absolute percentage deviations within each iteration of a query, and we report this value averaged over all iterations.
+We can see that optimizing for quality indeed reduces the SMAPE of query results.
+In general, the SMAPE relative to the base-imputation approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces similar results to imputation at the base tables.
 
@@
     parameterizations relative to results when imputing on base table. Values of $0.0$,
-    $100.0$, or $NaN$ indicate uninformative values.} %TODO MJS
+    $100.0$, or $NaN$ indicate uninformative values\todo{use one flag value (-) or explain why value is uninformative}.} %TODO MJS
 \label{table:smape}
@@
 
-In many real-world cases, applying the imputation step at the base table is prohibitive.
-To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
-aggregate query over the ACS dataset in our final experiment(\Cref{query-acs}).
-Applying the imputation operation to the base table is extremely expensive, in our case
-completing in 75 minutes. In contrast, \ProjectName{} executes a quality-optimized version
-in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
-benefit of using our system for early data exploration.
-
-\begin{figure}
+In many real-world cases, applying the imputation step at the base table is prohibitively expensive.
+To illustrate the increasing difficulty of such an approach as datasets scale, we ran the following query over the ACS dataset:
 \begin{lstlisting}
@@
 \end{lstlisting}
-\caption{A simple aggregate query on ACS data. Imputing the base table is prohibitive.
-    \ProjectName's quality-optimized and performance-optimized queries execute in 7 seconds
-    and 1 second, respectively.
-}
-\label{query-acs}
+Imputing the base table and then running the query takes 75 minutes.
+In contrast, \ProjectName{} executes a quality-optimized query plan in 7 seconds and a runtime-optimized plan in 1 second.
+This highlights the benefit of using our system for early data exploration.
 \end{figure}

### 2017-02-27 paper/experiments.tex
Add second experiment results to paper

--- 2017-02-27 paper/experiments.tex
@@
 for two separate survey-based data sets, showing that our system
-is well suited for early dataset exploration.
+is ideal for early dataset exploration.
 
@@
 \subsubsection{Center for Disease Control: Surveys}
-We use survey data collected by the 
-Center for Disease Control and Prevention (CDC) in the United States. We
-experiment on a set of tables collected as part of the 2013-2014 National
+For our first set of experiments, we use survey data collected by the 
+Center for Disease Control and Prevention (CDC) in the United States. In particular, we
+experiment on a set of tables collected as part of the National
 Health and Nutrition Examination Survey (NHANES), a series of studies
@@
 The data consists on survey responses, physical examinations, and laboratory
-results, amongst others. Our dataset is available for direct download on the CDC's organization site on 
+results, amongst others. Our dataset corresponds to the 2013-2014 NHANES
+and is available for direct download on the CDC's organization site on 
 kaggle.com, a popular data science website.
 
-There are a total of 6 tables associated with the NHANES data set. We use three tables:
+There are a total of 6 tables associated with the NHANES data set. For purposes
+of our experiments, we use three tables:
 
@@
     \end{subtable}
-    \caption{Missing value distribution within each table/attribute in CDC NHANES 2013-2014 data}
+    \caption{Missing values in CDC NHANES 2013-2014 data}
     \label{table:nhanes-description} 
@@
 For our purposes, we use a version of the data that has been 
-pre-processed, but where many missing values remain. It is
+pre-processed, but where plenty of missing values remain. It is
 worth noting that given the table design (where many attributes
@@
     \input{tables/fcc}
-    \caption{Missing values in fCC Survey Data}
+    \caption{Missing values in FCC Survey Data}
    \label{table:fcc-description} 
@@
 \subsection{Queries}
-We collect a set of queries (\Cref{fig:queries}) that we think are both representative and interesting
-to plan. They could reasonably be written by a user in the course of data analysis.
+We collect a set of queries (\Cref{fig:queries}) that we think are both representative,
+in that they could reasonably be written by a user in the course of data analysis, and
+interesting to plan.
 
@@
 
-Queries 1-4 are on the CDC data. In~\Cref{q1}, we calculate the average height for
+Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}). In~\Cref{q1}, we calculate the average height for
 users based on their income data. In~\Cref{q2}, we compare cholesterol levels for
@@
 
-Queries 5-8 are on the fCC data. In~\Cref{q5}, we calculate the average income for
+Our second set of queries is on the FCC data (\Cref{fig:queries-fcc}). In~\Cref{q5}, we calculate the average income for
 survey participants, based on their bootcamp attendance. In~\Cref{q6} estimates the
@@
 calculate the average amount of money survey participants with student debt destined to
-learning based on their school degree. In~\Cref{q8}, we join the fCC data with a
+learning based on their school degree. In~\Cref{q8}, we join the FCC data with a
 reference table provided by the World Bank which summarizes GDP per-capita across various
@@
  \input{queries-fcc}
- \caption{Queries on fCC data}
+ \caption{Queries on FCC data}
  \label{fig:queries-fcc}
  \end{subtable}
+ ~
+ \begin{subtable}{\linewidth}
+ \input{query-acs}
+ \caption{Queries on ACS data}
+ \label{fig:query-acs}
+ \end{subtable}
+
   \caption{Queries used in our experiments.}
@@
 
-\Cref{fig:runtimes} shows a summary of the performance results. The first plot
+\Cref{fig:runtimes} provides a summary of the performance results. The first plot
 shows average running times for \ProjectName{} plans in two configurations:
@@
 cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
-with performance improvement on a factor of ten\todo{give range of performance}. This performance differential means it is feasible
+with performance improvement on a factor of ten. This performance differential means it is feasible
 to explore multiple imputations operations (including more expensive operators) when using
@@
 \begin{figure}
-\todo{change this to a single graph of X-times faster than base approach}
-\begin{subfigure}{\linewidth}
-\includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
-\caption{Query runtimes with \ProjectName{}}
-\label{subfig:project-runtime-queries}
-\end{subfigure}
-~
-\begin{subfigure}{\linewidth}
-\includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
-\caption{Query runtimes with imputation on base tables}
-\label{subfig:base-runtime-queries}
-\end{subfigure}
-\caption{Comparing \ProjectName{} runtimes for each query
-    (\Cref{subfig:project-runtime-queries}) with the runtime for a query when base-table
-    imputations are used (\Cref{subfig:base-runtime-queries})}
+\includegraphics[width=\columnwidth]{figures/running_times_combined.png}
+\caption{Query runtimes with imputation on base tables, \ProjectName{} imputation
+    optimizing for quality ($\alpha=0$), and \ProjectName{} imputation optimizing for
+    efficiency ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
+    magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
+    imputation can provide another order-of-magnitude speedup at the expense of quality. For
+    relatively large queries (Query 9), runtime of imputation can be reduced from 6.5
+    hours to 7 seconds. }
 \label{fig:runtimes}
@@
 \ProjectName{}'s query results when compared to running imputation on the base tables and
-executing the query on the cleaned data. Each query ran for 200 iterations in both settings
+executing the query on the cleaned data. After a warm-up period, each query was run 200 iterations in both settings
 and then results from the two approaches were paired up and compared tuple-wise. We average
@@
 reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
-approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces similar
+approach are low in all cases, indicating that on-the-fly imputation produces similar
 results to imputation at the base tables.
@@
 To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
-aggregate query over the ACS dataset in our final experiment(\Cref{query-acs}).
+aggregate query over the ACS dataset in our final experiment(\Cref{fig:query-acs}).
 Applying the imputation operation to the base table is extremely expensive, in our case
-completing in 75 minutes. In contrast, \ProjectName{} executes a quality-optimized version
+completing in 355 minutes(\ref{fig:runtimes}). In contrast, \ProjectName{} executes a quality-optimized version
 in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
 benefit of using our system for early data exploration.
-
-\begin{figure}
-\begin{lstlisting}
-SELECT AVG(c0) FROM acs_dirty;
-\end{lstlisting}
-\caption{A simple aggregate query on ACS data. Imputing the base table is prohibitive.
-    \ProjectName's quality-optimized and performance-optimized queries execute in 7 seconds
-    and 1 second, respectively.
-}
-\label{query-acs}
-\end{figure}
 

### 2017-02-27 paper/experiments.tex
paper: minor fixes

--- 2017-02-27 paper/experiments.tex
@@
 
-Queries 1-4 are on the CDC data. In~\Cref{q1}, we calculate the average height for
-users based on their income data. In~\Cref{q2}, we compare cholesterol levels for
-individuals with low, medium, and high incomes spectrum and above a certain weight. In~\Cref{q3}, 
-we extract the maximum blood lead levels for children under the age of 6
-years. In~\Cref{q4}, we calculate the average systolic blood pressure, by gender, for
-subjects with a body mass index indicating obesity.
+Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}). In~\Cref{q1}, we
+calculate the average height for users based on their income data. In~\Cref{q2}, we compare
+cholesterol levels for individuals with low, medium, and high incomes spectrum and above a
+certain weight. In~\Cref{q3}, we extract the maximum blood lead levels for children under
+the age of 6 years. In~\Cref{q4}, we calculate the average systolic blood pressure, by
+gender, for subjects with a body mass index indicating obesity.
 
-Queries 5-8 are on the freeCodeCamp data.
+Our second set of queries is on the FCC data (\Cref{fig:queries-fcc}).
 \Cref{q5} calculates the average income for survey participants, based on their bootcamp attendance.
@@
 The query calculates the average GDP per-capita of countries with and without bootcamp participants. 
-
 \begin{table*}
-  \todo{do we really need headers on these tables?}
 \centering
@@
   \input{queries-cdc}
-    \caption{Queries on CDC data}\label{fig:queries-cdc}
+  \caption{Queries on CDC data}
+  \label{fig:queries-cdc}
  \end{subtable}
-  \par\medskip
+ ~
  \begin{subtable}{\linewidth}
  \input{queries-fcc}
-    \caption{Queries on freeCodeCamp data}\label{fig:queries-fcc}
+ \caption{Queries on FCC data}
+ \label{fig:queries-fcc}
  \end{subtable}
-  \par\medskip  
-  \caption{Queries used in our experiments.}\label{fig:queries}
+ ~
+ \begin{subtable}{\linewidth}
+ \input{query-acs}
+ \caption{Queries on ACS data}
+ \label{fig:query-acs}
+ \end{subtable}
+
+  \caption{Queries used in our experiments.}
+  \label{fig:queries}
 \end{table*}
@@
 \begin{figure}
-  \todo{change this to a single graph of X-times faster than base approach}
-  \begin{subfigure}{\linewidth}
-    \includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
-    \caption{Query runtimes with \ProjectName{}}\label{subfig:project-runtime-queries}
-  \end{subfigure}
-  ~
-  \begin{subfigure}{\linewidth}
-    \includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
-    \caption{Query runtimes with imputation on base tables}\label{subfig:base-runtime-queries}
-  \end{subfigure}
-  \par\medskip
-  \caption{Comparing \ProjectName{} runtimes for each query
-    (\Cref{subfig:project-runtime-queries}) with the runtime for a query when base-table
-    imputations are used (\Cref{subfig:base-runtime-queries})}\label{fig:runtimes}
+\includegraphics[width=\columnwidth]{figures/running_times_combined.png}
+\caption{Query runtimes with imputation on base tables, \ProjectName{} imputation
+    optimizing for quality ($\alpha=0$), and \ProjectName{} imputation optimizing for
+    efficiency ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
+    magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
+    imputation can provide another order-of-magnitude speedup at the expense of quality. For
+    relatively large queries (Query 9), runtime of imputation can be reduced from 6.5
+    hours to 7 seconds. }
+\label{fig:runtimes}
 \end{figure}
@@
 \Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running imputation on the base tables and executing the query on the cleaned data \todo{SMAPE citation}.
-Each query ran for 200 iterations \todo{iterations?
+After a warm-up period, each query ran for 200 iterations \todo{iterations?
 is this for the imputation or is it the number of query re-runs?} in both settings and then results from the two approaches were paired up and compared tuple-wise.
@@
 
-In many real-world cases, applying the imputation step at the base table is prohibitively expensive.
-To illustrate the increasing difficulty of such an approach as datasets scale, we ran the following query over the ACS dataset:
-\begin{lstlisting}
-SELECT AVG(c0) FROM acs_dirty;
-\end{lstlisting}
-Imputing the base table and then running the query takes 75 minutes.
-In contrast, \ProjectName{} executes a quality-optimized query plan in 7 seconds and a runtime-optimized plan in 1 second.
-This highlights the benefit of using our system for early data exploration.
-\end{figure}
+In many real-world cases, applying the imputation step at the base table is prohibitively
+expensive.
+To illustrate the increasing difficulty of such an approach as datasets scale, we run the
+query in (\Cref{fig:query-acs}) over the ACS dataset.
+Applying the imputation operation to the base table is extremely expensive, in our case
+completing in 355 minutes(\ref{fig:runtimes}). In contrast, \ProjectName{} executes a quality-optimized version
+in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
+benefit of using our system for early data exploration.
 

### 2017-02-27 paper/experiments.tex
Add fcc acronym at first appearance

@@
 For our second set of experiments, we chose to use data collected
-by freeCodeCamp as a part of a survey of new software developers
+by freeCodeCamp (FCC) as a part of a survey of new software developers
 (both professional and amateur)~\cite{fcc-data}. freeCodeCamp is an open-source

### 2017-02-27 paper/experiments.tex
Fix todo describing number of query runs

@@
 \Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running imputation on the base tables and executing the query on the cleaned data \todo{SMAPE citation}.
-After a warm-up period, each query ran for 200 iterations \todo{iterations?
-is this for the imputation or is it the number of query re-runs?} in both settings and then results from the two approaches were paired up and compared tuple-wise.
+Each query was separately run 200 times in both settings and then results from the two approaches were paired up and compared tuple-wise.
 We average tuple-wise absolute percentage deviations within each iteration of a query, and we report this value averaged over all iterations.

### 2017-02-28 paper/introduction.tex
paper: another pass over intro and abstract

@@
 A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
-Finally, for many queries, it is not necessary to clean the entire data set, and a significant speed up can be obtained by cleaning only the necessary parts.
-In contrast to existing systems, \ProjectName{} exploits this observation and avoids imputing over the entire data set, as done with traditional methods. \ProjectName{}
+In contrast to existing systems, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. \ProjectName{}
 carefully plans the placement of different imputation operations throughout the query plan, resulting in a 
@@
 Prior work in statistics\todo{add more specific subarea qualifier} has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
-Prior work in databases has a established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
+Prior work in databases has established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
 Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
@@
   We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
-  Impute operation fills in missing data values using a regression model (CE-CART)~\cite{burgette2010multiple}.
+  Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
   Drop operation simply drops tuples which contain null values.
@@
 \item \textbf{Proof of optimality:}
- We prove that our cost-model and planning algorithm produces a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
+ We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
 \end{itemize}

### 2017-02-28 paper/main.tex
paper: minor fixes
paper: another pass over intro and abstract

@@
 their data, while the system picks the optimal placement of imputation operations. This placement is based
-on the specific query issued and a parameter that the user selects for evaluating the tradeoff between data
+on the specific query issued and a user-provided parameter that trades off data
 quality and runtime performance. We incorporate the imputation operators into the relational algebra, allowing
@@
 \balance
-\nocite{*}
 \printbibliography

### 2017-02-28 paper/examples.tex
paper: fixed examples, had wrong plan for performance optimized

@@
 2) she could use a standard data imputation package on all of her data. Both of these
-approaches have significant drawbacks. For the query pictured in ~\Cref{fig:example-query}, 
+approaches have significant drawbacks. For the query pictured in~\Cref{fig:example-query}, 
 (1) drops 1492 potentially relevant tuples,
-while her previous experiences with (2) have proven be expensive. The epidemiologist needs a more complete picture
+while her previous experiences with (2) have proven to be expensive. The epidemiologist needs a more complete picture
 of the data, so (1) is insufficient, and she is under time pressure, so (2)
@@
 \caption{A quality-optimized plan for the query in~\Cref{fig:example-query}.
-The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
-$\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
+The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate, $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
 \label{fig:quality-plan}
@@
 \subsection{Planning with \ProjectName{}}
-The multiple possible locations for performing imputation leads to a  search space contains plans of varying performance and quality of results.
+The search space contains plans of varying performance and quality of results, as a product of the multiple possible locations for imputation.
 The user can influence the final plan selected by \ProjectName{}
@@
 
-\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
+\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operation $\mu$, which employs an iterative regression model to fill in missing values rather than dropping tuples.
 It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
@@
     [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
-      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
-        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
-          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
+      [.\colorbox{pink}{$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$}
         [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
-              [.$\sigma_{\text{exams.weight} \geq 120}$ 
-                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
-              ] 
-              labs 
-            ]
-          ]
+          [.$\bowtie_{\text{demo.id} = \text{exams.id}}$
             [.$\sigma_{\text{demo.gender} = 2}$ demo ]
-      ] 
-    ] 
-  ] 
-  ]
-\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}.The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
-$\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
+            [.$\sigma_{\text{exams.weight} \geq 120}$ [.\colorbox{pink}{$\delta_{\text{exams.weight}}$} exams ] ] ] labs ] ] ] ]
+\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}.The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate, $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
 \label{fig:fast-plan}

### 2017-02-28 paper/relatedwork.tex
paper: pass over relatedwork

@@
 In contrast, \ProjectName{} allows users to specify a tradeoff between
-information lost and runtime performance, allowing users to perform queries
+imputation quality and runtime performance, allowing users to perform queries
 directly on the entire dataset. Furthermore,
@@
 \cite{fernandez2009inter}, feedback punctuation is used to dynamically
-reconfigure the query plan as data continue to arrive for state-dependent
-optimizations. In this framework, expensive imputations can be avoided as
-real-time conditions change.
+reconfigure the query plan for state-dependent
+optimizations as data continue to arrive. One of the applications of this framework is to avoid
+expensive imputations as real-time conditions change. 
+\ProjectName{} focuses on handling
+missing values, rather than providing a general framework for planning feedback. Furthermore,
+it provides the ability
+to trade imputation quality for runtime performance. We prove our
+algorithm and cost model produces an optimal set of plans to choose from,
+given the constraints of the search space.
 
@@
 relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
-explore forecast-first and forecast-last plans, which perform forecasting operations before and after traditional
+explore forecast-first and forecast-last plans, which perform forecasting operations before and after executing all traditional
 relational operators, respectively.
@@
 results under different conditions. Instead, the optimizer allows users to pick their tradeoff between
-runtime cost and information lost. The search space considered by our optimizer is broader, not just
+runtime cost and imputation quality. The search space considered by our optimizer is broader, not just
 forecast-first/forecast-last plans, but rather imputation operators can be placed anywhere in the query plan

### 2017-02-28 paper/algorithm.tex
paper: pass over algorithms

@@
 \section{Algorithm}
-
-\todobox{this jumps around too much}{
-Our optimizer searches (\Cref{sec:planning}) a restricted space (\Cref{sec:search-space}) of query plans containing imputation operators (\Cref{sec:operators}) for an optimal plan given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.
-
-This plan is subject to two constraints (\Cref{sec:placement}).
-First, it must not emit any tuples which contain null values, regardless of the state of the base tables.
-Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
-Our planning algorithm is agnostic to the type of imputation used (\Cref{sec:imputation}), as the properties of the imputation are summarized by the cost model (\Cref{sec:cost-model}).}
+In order to correctly plan around missing values, we first extend the relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
+with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
+aggregations. The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
+quality of the imputations performed and the runtime performance. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
+algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). Our imputation planning algorithm is
+agnostic to the type of imputation used (\Cref{sec:imputation}), and we prove that it is optimal, given our search space (\Cref{sec:optimal}). Finally, we show
+that while our algorithm is exponential in the number of joins(\Cref{sec:complexity}), in practice, planning times are not a concern.
 
@@
 The cost of a plan $q$ is expressed as a tuple $\langle \textsc{Loss}(q), \textsc{Time}(q) \rangle$.
+$\textsc{Loss}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
 $\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
-$\textsc{Loss}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
 
@@
 
-This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of performance and cost for the current query.
+This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Loss}$ for the current query.
 
@@
 For simplicity, each of the logical nodes in a query plan points to a set of histograms.
-When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
+When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan.
 \Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
 
-For each sub-plan, we keep track of the estimated number of
-null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
+For each sub-plan, we keep track of the estimated cardinality by using the histograms associated with
 that plan. We use this estimate to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.
@@
 
-To compute $\text{Time}(Q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard 
+To compute $\textsc{Time}(Q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard 
 database.
@@
 different algorithms can vary significantly, the underlying I/O cost
-is not easily extracted (especially in the case that the entire stream of tuples does not
+is not easily extracted (especially in the case that the entire set of tuples does not
 fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
@@
 An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null.
-We compute a dirty set for each table using the table statistics, which track the number of null values in each column.
+We compute a dirty set for each table using the table histograms, which track the number of null values in each column.
 The dirty set for a query plan can be computed recursively as follows:
@@
 Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
-At a high level, a plan cache is a mapping from a set of tables to a set of dominating plans that perform the required joins and filters over these tables.
-We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes and a choice over the plan frontier
-can only be made once the set is final. Details are shown in~\Cref{fig:semantics-plan-cache}.
+At a high level, a plan cache is a mapping from a set of tables to a set of dirty-set specific dominating plans that perform the required joins and filters over these tables.
+We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes, each of which maintains its own Pareto frontier. 
+Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.
 
@@
                                                                      \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
-                                                                     \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
+                                                                     \forall q' \in \llbracket Q[T] \rrbracket, \textsc{Dirty}(q')=\textsc{Dirty}(q).  \doms{q} {q'} 
                                                                    \end{aligned}
@@
 The join optimization is based on the algorithm used in System R\cite{blasgen1981system},
-but rather than select the best plan at every point, we use our plan cache, which maintains the Pareto frontier of possible plans.
-The output of the join optimizer is the set of plans which satisfy all join predicates and lie along the frontier.
+but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the dirty-set specific Pareto frontiers.
+The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.
 
@@
 query optimization, which aims to reconcile performance with a declarative interface to data
-manipulation. By removing concerns for imputation, users executing queries in \ProjectName{}
-do not need to consider the potential nature of missing data.
+manipulation.
 
@@
 
-\subsection{Optimality}
+\subsection{Optimality}\label{sec:optimal}
 We prove the optimality of the final plan selected by showing that our cost-model and algorithm
@@
 $I: \mathcal{P}(D) \rightarrow \mathcal{P}(\{ \delta, \mu \})$, where
-$D$ is the set of dirty attributes for a relation, is an abstract imputation operator mapping a set of dirty
-attributes to a set of imputation operations.
+$D$ is the dirty set for a relation, is an abstract imputation operator mapping a set of
+attributes with missing values to a set of imputation operations.
 
@@
 \begin{theorem}\label{theorem:no-bad-sit}
-By computing dirty-set specific Pareto frontiers, it is never the case that we discard a plan $p_1$ s.t.
-$\doms{p_2}{p_1}$, but $\exists p_1' \in I(p_1), p_2' \in I(p_2).
-\doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2')$
+By computing dirty-set specific Pareto frontiers, it is never the case that we discard a plan $p_1$ s.t. $\doms{p_2}{p_1}$, but 
+$\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2') \land p_1' \not \in I(p_2)$
 \end{theorem}
@@
 \begin{case}
-$\textsc{Dirty}(p_1) \not= \textsc{Dirty}(p_2)$, in which case $p_1$ is not removed from the plan cache as the $\doms{p_2}{p_1}$ is not within the same dirty set. This is a contradiction
+$\textsc{Dirty}(p_1) \not= \textsc{Dirty}(p_2)$, in which case $p_1$ is not removed from the plan cache as the $\doms{p_2}{p_1}$ relation does not occur within the same dirty set.
+This is a contradiction
 \end{case}
@@
 
-\subsection{Complexity}
+\subsection{Complexity}\label{sec:complexity}
 \todo{Stengthen complexity argument, as per Sam's feedback}
-Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our planning algorithm still operates in exponential time. 
+Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our planning algorithm is still exponential in the number of joins. 
 If we removed our restriction on imputation operators to consider not just minimal versus maximal imputations, and instead allowed any arbitrary subset

### 2017-02-28 paper/introduction.tex
pass on intro

@@
 
-% Context
-Handling incorrect or dirty data is a complex and challenging problem for analysts\todo{add significant citations here}.
-% Problem
-One way in which a dataset can be dirty is for parts of it to be missing altogether~\cite{kim2003}.
-Missing data, if handled naively, can cause analyses to be incorrect~\cite{rubin1976}.
-According to some estimates, data scientists spend between 50 and 80 percent of their time performing tasks such as collecting and cleaning datasets~\cite{data-science-cleaning}.
+Many databases have large numbers of missing or NULL values;  these can arise for a variety of reasons, including missing source data (e.g., unknown facts),
+missing columns during data integration, de-normalized databases, or as a result of outlier detection and cleaning where NULL values are substituted for
+bad values~\cite{kim2003}.  Such NULL values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
+before processing is often desirable.
 
 % Solution
-To handle this problem, users may manually clean their dataset by performing some statistical analysis to replace missing data elements with likely values.
+One common approach is to manually replace missing values using some sort of statistical or predictive model based on other values in the table and record.
 This process is called \emph{imputation}.
-We have developed a system called \ProjectName{}, which is the first database to use a query optimization algorithm which allows imputation to occur on-the-fly, during query execution.
+In this paper, we present a new system, called \ProjectName{}, which is the designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.
 \emph{The key insight behind \ProjectName{} is that imputation
@@
 that this subset is generally much smaller than the entire database.}  
+% Background
+While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive -- out experiments show that even a relatively simple regression algorithm algorithms tables about 75 minutes to train and run
+on a 600K row database;  prior work~\cite{AkandeEtAl} reports times of 45 minutes to impute missing values on a 1\% sample of an XXX row database.
+A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
+In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
+carefully plans the placement  imputation / row drop operations inside the query plan, resulting in a 
+significant speedup in query execution while minimizing the number of dropped rows.
+Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in placement of imputation operations
+in an optimal way into query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.
 
-% Background
-While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive.
-A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
-In contrast to existing systems, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. \ProjectName{}
-carefully plans the placement of different imputation operations throughout the query plan, resulting in a 
-significant speedup in query execution. The focus on of our work is not on the imputation algorithm but rather on the planning involved in its use in non-trivial query plans.
+% \todobox{find appropriate location for this}
+% {In a previous project, she
+% worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
+% dataset took around 75 minutes\todo{Here, we can reference the number from our own experiment (done), or the number from Akande et al: ``performing imputation on just a 1.5 percent random sample took 45 minutes''}.}
 
-\todobox{find appropriate location for this}
-{In a previous project, she
-worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
-dataset took around 75 minutes\todo{Here, we can reference the number from our own experiment (done), or the number from Akande et al: ``performing imputation on just a 1.5 percent random sample took 45 minutes''}.}
-
-% Prior work
-There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
-Prior work in statistics\todo{add more specific subarea qualifier} has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
-Prior work in databases has established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
-Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
-\ProjectName{} contributes the first optimizer to plan data imputation on a per query basis, allowing users to use standard SQL on real-world data sets with missing values
-and express preferences in quality of data imputation versus running time.
+% % Prior work
+% There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
+% Prior work in statistics\todo{add more specific subarea qualifier} has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
+% Prior work in databases has established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
+% Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
+% \ProjectName{} contributes the first optimizer to plan data imputation on a per query basis, allowing users to use standard SQL on real-world data sets with missing values
+% and express preferences in quality of data imputation versus running time.
 The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
@@
 Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
-queries that execute in a fraction of the time\todo{add actual performance figures here} it takes to impute base tables and then run queries (the traditional approach).
+queries that execute in a fraction of the time\todo{add actual performance figures here} it takes to impute all missing values and then run queries (the traditional approach).
 Furthermore, the results obtained with this approach
-are similar in quality to those obtained with the base-table imputation approach, as indicated by low\todo{add actual value here} Symmetric-Mean-Absolute-Percentage-Errors in our empirical results
+are similar in quality to those obtained with the complete imputation approach, as indicated by low\todo{add actual value here} Symmetric-Mean-Absolute-Percentage-Errors in our empirical results
 over real-world data sets (see~\Cref{sec:experiments} for details).
@@
 % Scope
-\ProjectName{} is designed to enable early data exploration, by allowing analysts to run their queries without explicitly dealing with imputation.
-However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.
+\ProjectName{} is designed to enable early data exploration, by allowing analysts to run their queries without explicitly decide with when to perform imputation.
+% SAM -- cut the following -- could consider moving later but feels out of place here
+%However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.
 
 \subsection{Contributions}
-This paper makes the following contributions:
+\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
+increasing productivity for users who frequently encounter missing values in their data. These contributions include:
 \begin{itemize}
-\item \textbf{Relational algebra with imputation:}
+\item \textbf{Relational algebra extended with imputation:}
   We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
@@
 \end{itemize}
-
-\ProjectName{} leverages the combined contributions to plan fast queries on data with missing values, while producing comparable results to traditional approaches.
-This provides the opportunity to increase productivity for users who frequently encounter missing values in their data.
 

### 2017-02-28 paper/experiments.tex
paper: pass over experiments

@@
 For our experiments we plan and execute queries
-for two separate survey-based data sets, showing that our system
+for three separate survey-based data sets, showing that our system
 is well suited for early dataset exploration.
@@
 For our second set of experiments, we chose to use data collected
-by freeCodeCamp (FCC) as a part of a survey of new software developers
+by freeCodeCamp (fCC) as a part of a survey of new software developers
 (both professional and amateur)~\cite{fcc-data}. freeCodeCamp is an open-source
@@
 We use a version of the data that has been pre-processed, but where missing values remain.
-\todobox{Give example of expected missing value}{However, many of the missing values are expected because the data has been de-normalized.}
-The original dataset has 15,620 rows and 113 attributes.
+For example, 46.6\% of entries for entry \textit{commutetime} are missing. However, it is worth noting that some of the missing values
+are also expected, given the way the data has been de-normalized. For example, \textit{bootcamploanyesno}, a binary attribute encoding
+whether a respondent had a loan for a bootcamp, is expected to be null for participants who did not attend a bootcamp.
+
 We chose a subset of 17 attributes, shown in~\Cref{table:fcc-description}, along with the percentage of missing values.
@@
   \input{tables/fcc}
-  \caption{Missing values in freeCodeCamp Survey Data}\label{table:fcc-description} 
+  \caption{Missing value distribution for each attribute in freeCodeCamp Survey Data}\label{table:fcc-description} 
 \end{table}
@@
 Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}). In~\Cref{q1}, we
-calculate the average height for users based on their income data. In~\Cref{q2}, we compare
-cholesterol levels for individuals with low, medium, and high incomes spectrum and above a
-certain weight. In~\Cref{q3}, we extract the maximum blood lead levels for children under
+calculate the average cuff size for users based on their income data. In~\Cref{q2}, we compare
+creatine levels for individuals with low, medium, and high incomes spectrum and above a
+certain weight. In~\Cref{q3}, we extract the average blood lead levels for children under
 the age of 6 years. In~\Cref{q4}, we calculate the average systolic blood pressure, by
@@
 
-Our second set of queries is on the FCC data (\Cref{fig:queries-fcc}).
+Our second set of queries is on the fCC data (\Cref{fig:queries-fcc}).
 \Cref{q5} calculates the average income for survey participants, based on their bootcamp attendance.
-\Cref{q6} estimates the average age of women from the United States who participated.
-\Cref{q7} calculates the average amount of money survey participants with student debt spent on learning based on their school degree.
+\Cref{q6} estimates the average commute time of women from the United States who participated.
+\Cref{q7} calculates the average amount of student debt based on school degree for survey participants who have student debt.
 \Cref{q8} joins the freeCodeCamp data with a reference table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
@@
 
-\Cref{fig:runtimes} shows a summary of the performance results. The first plot
-shows average running times for \ProjectName{} plans in two configurations:
-quality-optimized and runtime-optimized. The second plot shows, for comparison, the average
-running times required for each query when performing imputation at the base tables.  In all
-cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
-with performance improvement on a factor of ten\todo{give range of performance}. This performance differential means it is feasible
+\Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
+configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0$),
+\ProjectName{} optimizing for performance($\alpha=1$), and imputing at the base tables followed
+by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
+faster than imputing on the base tables. We can get another order-of-magnitude speedup when
+optimizing for performance. This performance differential means it is feasible
 to explore multiple imputations operations (including more expensive operators) when using
@@
 \includegraphics[width=\columnwidth]{figures/running_times_combined.png}
-\caption{Query runtimes with imputation on base tables, \ProjectName{} imputation
-    optimizing for quality ($\alpha=0$), and \ProjectName{} imputation optimizing for
-    efficiency ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
+\caption{Query runtimes with imputation on base tables, \ProjectName{}
+    optimizing for quality ($\alpha=0$), and \ProjectName{} optimizing for
+    performance ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
     magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
     imputation can provide another order-of-magnitude speedup at the expense of quality. For
-    relatively large queries (Query 9), runtime of imputation can be reduced from 6.5
+    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced from 6.5
     hours to 7 seconds. }
@@
 expensive.
-To illustrate the increasing difficulty of such an approach as datasets scale, we run the
-query in (\Cref{fig:query-acs}) over the ACS dataset.
+To illustrate the increasing difficulty of such an approach as datasets scale, we run~\Cref{q9} over the ACS dataset.
 Applying the imputation operation to the base table is extremely expensive, in our case
-completing in 355 minutes(\ref{fig:runtimes}). In contrast, \ProjectName{} executes a quality-optimized version
+completing in 355 minutes(\Cref{fig:runtimes})\todo{Micah: this doesn't match the 6.5 hours in the figure}. In contrast, \ProjectName{} executes a quality-optimized version
 in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential

### 2017-02-28 paper/conclusion.tex
paper: minor conclusion fix

@@
 learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
-information lost and runtime. We showed that different values for the tradeoff parameter can yield substantially
+imputation quality and runtime. We showed that different values for the tradeoff parameter can yield substantially
 different plans, allowing the user to express their preferences for performance on a per-query basis.
-Our experiments with \textit{CDC} and \textit{freeCodeCamp} survey-based datasets
+Our experiments with \textit{CDC}, \textit{freeCodeCamp} and \textit{ACS} survey-based datasets
 showed that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
 We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
-the data exploration phase. The plans selected for each query execute in a fraction of the time compared to the standard approach of imputing on the entire base tables and
+the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
+ the standard approach of imputing on the entire base tables and
 then formulating queries. Furthermore, the difference in query results between the two approaches is
-shown to be small in all queries considered. Data analysts need not commit to a specific imputation strategy and can instead
+shown to be small in all queries considered\todo{quantify this}. Data analysts need not commit to a specific imputation strategy and can instead
 vary this across queries.

### 2017-02-28 paper/experiments.tex
Pivot perf_summary table

@@
 \begin{table}
-\todo{Consider switching to graph, or changing table layout..not very intuitive right now.}
 \centering
@@
 \caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
-    parameterizations relative to results when imputing on base table. Values of $0.0$,
-    $100.0$, or $NaN$ indicate uninformative values\todo{use one flag value (-) or explain why value is uninformative}.} %TODO MJS
+    parameterizations relative to results when imputing on base table. Queries optimized for quality ($\alpha=0$) generally achieve lower error than queries optimized for efficiency ($\alpha=1$).}
 \label{table:smape}

### 2017-02-28 paper/experiments.tex
comments

@@
 
+\subsection{Results}
+
 \Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
@@
     relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced from 6.5
-    hours to 7 seconds. }
+    hours to 7 seconds. \srm{make this a bar graph}}
 \label{fig:runtimes}
@@
 
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running imputation on the base tables and executing the query on the cleaned data \todo{SMAPE citation}.
+\subsubsection{Accuracy vs. Base-table Imputation}
+
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running imputation on the base tables and executing the query on this cleaned base data \todo{SMAPE citation}.
 Each query was separately run 200 times in both settings and then results from the two approaches were paired up and compared tuple-wise.
@@
 
+\subsubsection{Runtime of Base-table Imputation}
 In many real-world cases, applying the imputation step at the base table is prohibitively
@@
 
+\srm{note somewhere that imputation requires both training and running model, and state why this is reasonable / the only choice}
+
 %%% Local Variables:

### 2017-02-28 paper/introduction.tex
comments

@@
 % Background
-While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive -- out experiments show that even a relatively simple regression algorithm algorithms tables about 75 minutes to train and run
+While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive -- out experiments show that even a relatively simple regression algorithm algorithms tables about XXX minutes to train and run
 on a 600K row database;  prior work~\cite{AkandeEtAl} reports times of 45 minutes to impute missing values on a 1\% sample of an XXX row database.
+\srm{Make numbers above consistent with experiments.}
 A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.

### 2017-02-28 paper/experiments.tex
Update acs timing

--- 2017-02-28 paper/experiments.tex
@@
 \newcommand{\labexrows}{9813}
+\newcommand{\acsbaseresultminutes}{355 minutes}
+\newcommand{\acsbaseresulthours}{just under 6 hours}
+\newcommand{\acsimputedbzeroresult}{4 seconds}
+\newcommand{\acsimputedboneresult}{1 second}
 
@@
     imputation can provide another order-of-magnitude speedup at the expense of quality. For
-    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced from 6.5
-    hours to 7 seconds. }
+    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced from \acsbaseresulthours~to \acsimputedbzeroresult. }
 \label{fig:runtimes}
@@
 Applying the imputation operation to the base table is extremely expensive, in our case
-completing in 355 minutes(\Cref{fig:runtimes})\todo{Micah: this doesn't match the 6.5 hours in the figure}. In contrast, \ProjectName{} executes a quality-optimized version
-in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
+completing in \acsbaseresultminutes (\Cref{fig:runtimes}. In contrast, \ProjectName{} executes a quality-optimized version
+in\acsimputedbzeroresult~and a runtime-optimized version in \acsimputedboneresult. This highlights the potential
 benefit of using our system for early data exploration.

### 2017-02-28 paper/experiments.tex

@@
     imputation can provide another order-of-magnitude speedup at the expense of quality. For
-    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced from \acsbaseresulthours~to \acsimputedbzeroresult. }
+    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced from \acsbaseresulthours~to \acsimputedbzeroresult. \srm{make this a bar graph}}
+
 \label{fig:runtimes}

### 2017-02-28 paper/experiments.tex
Switch to bar graph

@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{figures/running_times_combined.png}
+\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
 \caption{Query runtimes with imputation on base tables, \ProjectName{}
@@
     imputation can provide another order-of-magnitude speedup at the expense of quality. For
-    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced from \acsbaseresulthours~to \acsimputedbzeroresult. \srm{make this a bar graph}}
+    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced from \acsbaseresulthours~to \acsimputedbzeroresult.}
 

### 2017-02-28 paper/introduction.tex
paper: fix citation issues

@@
 % Background
-While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive -- out experiments show that even a relatively simple regression algorithm algorithms tables about XXX minutes to train and run
-on a 600K row database;  prior work~\cite{AkandeEtAl} reports times of 45 minutes to impute missing values on a 1\% sample of an XXX row database.
+While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression algorithm takes about XXX minutes to train and run
+on a 600K row database; prior work~\cite{akande2015empirical} reports times of 45 minutes to impute missing values on a 1\% sample of an XXX row database.
 \srm{Make numbers above consistent with experiments.}

### 2017-02-28 paper/algorithm.tex
comments
paper: remove formal proof

@@
 algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). Our imputation planning algorithm is
-agnostic to the type of imputation used (\Cref{sec:imputation}), and we prove that it is optimal, given our search space (\Cref{sec:optimal}). Finally, we show
+agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
 that while our algorithm is exponential in the number of joins(\Cref{sec:complexity}), in practice, planning times are not a concern.
@@
 
+\srm{I find this table hard to follow -- I think it'd be good to explain one or two of the rows in detail, and to explain what the set L contains.}
 Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
@@
 
-\subsection{Optimality}\label{sec:optimal}
-We prove the optimality of the final plan selected by showing that our cost-model and algorithm
-construct a complete Pareto frontier.
-
-\begin{definition}
-$I: \mathcal{P}(D) \rightarrow \mathcal{P}(\{ \delta, \mu \})$, where
-$D$ is the dirty set for a relation, is an abstract imputation operator mapping a set of
-attributes with missing values to a set of imputation operations.
-
-For convenience we denote applying the set of imputation operations to a plan $p$ as $I(p)$ and understand that this
-is shorthand for $\{ op(p)   | op \in I(\textsc{Dirty}(p)) \}$.
-\end{definition}
-
-We can succinctly describe possible imputation transformations to a plan at a given point in the planning algorithm solely through $I$.
-
-\begin{theorem}\label{theorem:no-bad-sit}
-By computing dirty-set specific Pareto frontiers, it is never the case that we discard a plan $p_1$ s.t. $\doms{p_2}{p_1}$, but 
-$\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2') \land p_1' \not \in I(p_2)$
-\end{theorem}
-
-\Cref{theorem:no-bad-sit} states that there is no \textit{bad situation} such as that depicted in~\Cref{fig:bad-sit-diagram}
-
-\begin{figure}
-\centering
-\begin{tikzpicture}
-% before I
-\node(p1) {$p_1$};
-\node[right=of p1](dom1){$\prec$};
-\node[right=of dom1](p2) {$p_2$};
-
-% after I
-\node[below=of p1](p1'){$p_1'$};
-\node[right=of p1'](dom2) {$\succ$};
-\node[right=of dom2](p2'){$p_2'$};
-
-% arrows
-\draw[->] (p1) -- node[right] {$\scriptstyle I$} (p1');
-\draw[->] (p2) -- node[right] {$\scriptstyle I$} (p2');
-\end{tikzpicture}
-\caption{A bad Pareto frontier selection would choose $p_2$ over $p_1$, as it is locally optimal, but with the available transformations ($I$) $p_2$  can only reach $p_2'$, which is dominated by the inaccessible $p_1'$.~\Cref{theorem:no-bad-sit} says such a situation cannot arise when constructing Pareto frontiers on a dirty-set basis.}
-\label{fig:bad-sit-diagram}
-\end{figure}
-
-\begin{proof}
-For contradiction, let $p_1$ and $p_2$ be plans such that $\doms{p_2}{p_1}$, $p_1$ is removed from the plan cache
-in favor of $p_2$, but there  $\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2')$
-and $p_1'$ is not reachable from $p_2$ by $I$.
-
-There are two possible cases :
-
-\begin{case}
-$\textsc{Dirty}(p_1) = \textsc{Dirty}(p_2)$, in which case $I(p_1) = I(p_2)$, as $I$ uniquely determines the possible transformations.
+%\subsection{Optimality}\label{sec:optimal}
+%We prove the optimality of the final plan selected by showing that our cost-model and algorithm
+%construct a complete Pareto frontier.
+%
+%\begin{definition}
+%$I: \mathcal{P}(D) \rightarrow \mathcal{P}(\{ \delta, \mu \})$, where
+%$D$ is the dirty set for a relation, is an abstract imputation operator mapping a set of
+%attributes with missing values to a set of imputation operations.
+%
+%For convenience we denote applying the set of imputation operations to a plan $p$ as $I(p)$ and understand that this
+%is shorthand for $\{ op(p)   | op \in I(\textsc{Dirty}(p)) \}$.
+%\end{definition}
+%
+%We can succinctly describe possible imputation transformations to a plan at a given point in the planning algorithm solely through $I$.
+%
+%\begin{theorem}\label{theorem:no-bad-sit}
+%By computing dirty-set specific Pareto frontiers, it is never the case that we discard a plan $p_1$ s.t. $\doms{p_2}{p_1}$, but 
+%$\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2') \land p_1' \not \in I(p_2)$
+%\end{theorem}
+%
+%\Cref{theorem:no-bad-sit} states that there is no \textit{bad situation} such as that depicted in~\Cref{fig:bad-sit-diagram}
+%
+%\begin{figure}
+%\centering
+%\begin{tikzpicture}
+%% before I
+%\node(p1) {$p_1$};
+%\node[right=of p1](dom1){$\prec$};
+%\node[right=of dom1](p2) {$p_2$};
+%
+%% after I
+%\node[below=of p1](p1'){$p_1'$};
+%\node[right=of p1'](dom2) {$\succ$};
+%\node[right=of dom2](p2'){$p_2'$};
+%
+%% arrows
+%\draw[->] (p1) -- node[right] {$\scriptstyle I$} (p1');
+%\draw[->] (p2) -- node[right] {$\scriptstyle I$} (p2');
+%\end{tikzpicture}
+%\caption{A bad Pareto frontier selection would choose $p_2$ over $p_1$, as it is locally optimal, but with the available transformations ($I$) $p_2$  can only reach $p_2'$, which is dominated by the inaccessible $p_1'$.~\Cref{theorem:no-bad-sit} says such a situation cannot arise when constructing Pareto frontiers on a dirty-set basis.}
+%\label{fig:bad-sit-diagram}
+%\end{figure}
+%
+%\begin{proof}
+%For contradiction, let $p_1$ and $p_2$ be plans such that $\doms{p_2}{p_1}$, $p_1$ is removed from the plan cache
+%in favor of $p_2$, but there  $\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2')$
+%and $p_1'$ is not reachable from $p_2$ by $I$.
+%
+%There are two possible cases :
+%
+%\begin{case}
+%$\textsc{Dirty}(p_1) = \textsc{Dirty}(p_2)$, in which case $I(p_1) = I(p_2)$, as $I$ uniquely determines the possible transformations.
 This makes $p_2$ and $p_1$ substitutable, so $p_1'$ is reachable from $p_2$.
 This is a contradiction.
-\end{case}
-
-\begin{case}
-$\textsc{Dirty}(p_1) \not= \textsc{Dirty}(p_2)$, in which case $p_1$ is not removed from the plan cache as the $\doms{p_2}{p_1}$ relation does not occur within the same dirty set.
+%\end{case}
+%
+%\begin{case}
+%$\textsc{Dirty}(p_1) \not= \textsc{Dirty}(p_2)$, in which case $p_1$ is not removed from the plan cache as the $\doms{p_2}{p_1}$ relation does not occur within the same dirty set.
 This is a contradiction
-\end{case}
-
-So it must be the case that no such situation arises.
-\end{proof}
-
-Given that no desirable plans are discarded by computing Pareto frontiers for each dirty set, the final Pareto frontier (for dirty set $\emptyset$) is optimal.
+%\end{case}
+%
+%So it must be the case that no such situation arises.
+%\end{proof}
+%
+%
+%Given that no desirable plans are discarded by computing Pareto frontiers for each dirty set, the final Pareto frontier (for dirty set $\emptyset$) is optimal.
 

### 2017-02-28 paper/introduction.tex
paper: remove formal proof

@@
 % Background
-While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression algorithm takes about XXX minutes to train and run
+While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive -- out experiments show that even a relatively simple regression algorithm algorithms tables about XXX minutes to train and run
 on a 600K row database;  prior work~\cite{akande2015empirical} reports times of 45 minutes to impute missing values on a 1\% sample of an XXX row database.
@@
   By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
-\item \textbf{Proof of optimality:}
- We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
+%\item \textbf{Proof of optimality:}
+% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
 \end{itemize}

### 2017-02-28 paper/experiments.tex
paper: inline ACS query

@@
 The query calculates the average GDP per-capita of countries with and without bootcamp participants.
+
 \begin{table*}
@@
     \input{queries-cdc}
-  \caption{Queries on CDC data}
-  \label{fig:queries-cdc}
+    \caption{Queries on CDC data}\label{fig:queries-cdc}
   \end{subtable}
- ~
+  \par\medskip
   \begin{subtable}{\linewidth}
     \input{queries-fcc}
- \caption{Queries on fCC data}
- \label{fig:queries-fcc}
+    \caption{Queries on fCC data}\label{fig:queries-fcc}
   \end{subtable}
- ~
- \begin{subtable}{\linewidth}
- \input{query-acs}
- \caption{Queries on ACS data}
- \label{fig:query-acs}
- \end{subtable}
-
-  \caption{Queries used in our experiments.}
-  \label{fig:queries}
+  \par\medskip
+  \caption{Queries used in our experiments.}\label{fig:queries}
 \end{table*}
@@
 %    \caption{Base error, percent change in error and and running time for queries
-%    with different imputation levels. Base error is the root-mean-square error (RMSE) between the query run on clean
+%    with different imputation levels. Base error is the root-mean-square error (RMSE) between the query on clean
 %    data and the query run on dirty data without imputation. Change in error is relative to the base error.}
@@
 expensive.
-To illustrate the increasing difficulty of such an approach as datasets scale, we run~\Cref{q9} over the ACS dataset.
+To illustrate the increasing difficulty of such an approach as datasets scale, we ran the following query over the ACS dataset:
+\begin{lstlisting}[breaklines]
+SELECT AVG(c0) FROM acs_dirty;
+\end{lstlisting}
 Applying the imputation operation to the base table is extremely expensive, in our case
-completing in \acsbaseresultminutes (\Cref{fig:runtimes}. In contrast, \ProjectName{} executes a quality-optimized version
-in\acsimputedbzeroresult~and a runtime-optimized version in \acsimputedboneresult. This highlights the potential
+completing in \acsbaseresultminutes{} (\Cref{fig:runtimes}). In contrast, \ProjectName{} executes a quality-optimized version
+in \acsimputedbzeroresult{} and a runtime-optimized version in \acsimputedboneresult{}. This highlights the potential
 benefit of using our system for early data exploration.

### 2017-02-28 paper/experiments.tex
Explain why we can't pre-train models

--- 2017-02-28 paper/experiments.tex
@@
     imputation can provide another order-of-magnitude speedup at the expense of quality. For
-    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced from \acsbaseresulthours~to \acsimputedbzeroresult.}
+    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced
+    from~\acsbaseresulthours~to~\acsimputedbzeroresult.}
 
@@
 
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running imputation on the base tables and executing the query on this cleaned base data \todo{SMAPE citation}.
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running
+imputation on the base tables and executing the query on this cleaned base data \todo{SMAPE citation}.
 Each query was separately run 200 times in both settings and then results from the two approaches were paired up and compared tuple-wise.
@@
 We can see that optimizing for quality indeed reduces the SMAPE of query results.
-In general, the SMAPE relative to the base-imputation approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces similar results to imputation at the base tables.
+In general, the SMAPE relative to the base-imputation approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces
+similar results to imputation at the base tables.
 
@@
 \caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
-    parameterizations relative to results when imputing on base table. Queries optimized for quality ($\alpha=0$) generally achieve lower error than queries optimized for efficiency ($\alpha=1$).}
+    parameterizations relative to results when imputing on base table. Queries optimized for quality ($\alpha=0$) generally achieve
+    lower error than queries optimized for efficiency ($\alpha=1$).}
 \label{table:smape}
@@
 Applying the imputation operation to the base table is extremely expensive, in our case
-completing in \acsbaseresultminutes (\Cref{fig:runtimes}. In contrast, \ProjectName{} executes a quality-optimized version
-in\acsimputedbzeroresult~and a runtime-optimized version in \acsimputedboneresult. This highlights the potential
+completing in \acsbaseresultminutes~(\Cref{fig:runtimes}). In contrast, \ProjectName{} executes a quality-optimized version
+in~\acsimputedbzeroresult~and a runtime-optimized version in~\acsimputedboneresult. This highlights the potential
 benefit of using our system for early data exploration.
 
-\srm{note somewhere that imputation requires both training and running model, and state why this is reasonable / the only choice}
+For every imputation inserted into the query plan, a new statistical model is instantiated and trained before being used for imputation.
+Although it is tempting to further optimize query execution in \ProjectName{} by pre-training imputation models while the analyst begins to make queries,
+this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are not known until runtime.
+Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.
 

### 2017-02-28 paper/experiments.tex

--- 2017-02-28 paper/experiments.tex
@@
 
-\srm{note somewhere that imputation requires both training and running model, and state why this is reasonable / the only choice}
+For every imputation inserted into the query plan, a new statistical model is instantiated and trained before being used for imputation.
+Although it is tempting to further optimize query execution in \ProjectName{} by pre-training imputation models while the analyst begins to make queries,
+this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are not known until runtime.
+Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.
 

### 2017-02-28 paper/algorithm.tex
paper: clean up complexity section

--- 2017-02-28 paper/algorithm.tex
@@
 \subsection{Complexity}\label{sec:complexity}
-\todo{Stengthen complexity argument, as per Sam's feedback}
-Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our planning algorithm is still exponential in the number of joins. 
-If we removed our restriction on imputation operators to consider not just minimal versus maximal imputations, and instead allowed any arbitrary subset
-of attributes to be imputed at a plan node (including those not operated on by that node), then every single operator in the query plan would have a
-number of plans exponential in the number of dirty columns.
+Our optimization algorithm builds off the approach taken by System R~\cite{blasgen1981system}.
+If imputation operators are ignored, we search the same plan space.
+Therefore, our algorithm has a complexity of at least $O(2^J)$, where $J$ is the number of joins.
 
-By restricting imputations to two classes: those that impute only attributes used in the operator and those that impute all attributes needed downstream in the query,
-we limit the factor of plans at each operator to three: \textit{Drop}, \textit{Impute} over attributes used in operator, and \textit{Impute} over all attributes needed.
-Recall that the planner maintains different plans for different dirty sets, keeping only those that are not dominated by other plans with the same dirty set. So in many
-cases we can drop some of these intermediate plans generated at each operator. However, if the dirty sets tracked are distinct through the entire planning phase,
-then we could explore exponentially more plans than System R.
+The addition of imputation operators increases the number of plans exponentially, as an imputation may be placed before any of the relational operators.
+We restrict imputations to two classes: those that impute only attributes used in the operator and those that impute all the attributes that are needed downstream in the query.
+By doing so we limit the number of imputations at any point to four: \textit{Drop} or \textit{Impute} over the attributes used in the operator or over all the downstream attributes.
+This modification increases the complexity of our planning algorithm to $O(16^J)$.\todo{I think this is right, but should check}
 
-However, we note that in all cases we have considered, this exponential blowup does not affect the practical performance of our optimizer and results in relatively short
-planning times.
+To motivate the restriction of imputation types, we consider the implications of allowing arbitrary imputations.
+If we allow any arbitrary subset of attributes to be imputed, then we would need to consider $O(2^{|D|+1})$ different imputation operators before each relational operator.
+This would increase the overall complexity of the algorithm drastically.
+
+Finally, we note that for the queries we have examined, this exponential blowup does not affect the practical performance of our optimizer.
+Recall that the planner maintains different plans for different dirty sets, keeping only those that are not dominated by other plans with the same dirty set.
+So in many cases we can drop some of these intermediate plans generated at each operator.
+The worst case complexity only occurs if the dirty sets tracked are distinct through the entire planning phase.
+Planning times are discussed in more detail in \Cref{sec:results}.
 

### 2017-02-28 paper/experiments.tex
Move commands so they can be used in any sec
paper: clean up complexity section
Add performance and smape figures to text

@@
-\newcommand{\demorows}{10175}
-\newcommand{\labexrows}{9813}
-\newcommand{\acsbaseresultminutes}{355 minutes}
-\newcommand{\acsbaseresulthours}{just under 6 hours}
-\newcommand{\acsimputedbzeroresult}{4 seconds}
-\newcommand{\acsimputedboneresult}{1 second}
-
 \section{Experiments}\label{sec:experiments}
@@
 We can see that optimizing for quality indeed reduces the SMAPE of query results.
-In general, the SMAPE relative to the base-imputation approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces
+In general, the SMAPE relative to the base-imputation approach are low in all cases --- between \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly imputation produces
 similar results to imputation at the base tables.

### 2017-02-28 paper/introduction.tex
Add performance and smape figures to text

--- 2017-02-28 paper/introduction.tex
@@
 % Background
-While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression algorithm takes about XXX minutes to train and run
-on a 600K row database; prior work~\cite{akande2015empirical} reports times of 45 minutes to impute missing values on a 1\% sample of an XXX row database.
-\srm{Make numbers above consistent with experiments.}
+While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression tree algorithm takes \acsbaseresulthours{} to train and run
+on a 600K row database; prior work~\cite{akande2015empirical} reports times of 45 minutes for multiple-imputation of missing values on a 1\% sample of the same database on a desktop computer.\todo{Explain the discrepancy between 45 * 100 and 355?
+They use L=10 datasets and unknown hardware.
+Or drop this comparison entirely?}
+
 A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
@@
 Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
-queries that execute in a fraction of the time\todo{add actual performance figures here} it takes to impute all missing values and then run queries (the traditional approach).
+queries that execute between \lowxalphazero{} and \highxalphaoneexacs{} times as fast as it takes to impute all missing values and then run queries (the traditional approach).
 Furthermore, the results obtained with this approach
-are similar in quality to those obtained with the complete imputation approach, as indicated by low\todo{add actual value here} Symmetric-Mean-Absolute-Percentage-Errors in our empirical results
+are similar in quality to those obtained with the complete imputation approach, as indicated by low Symmetric-Mean-Absolute-Percentage-Errors --- between \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent --- in our empirical results
 over real-world data sets (see~\Cref{sec:experiments} for details).

### 2017-02-28 paper/main.tex
pass on intro
paper: pass over experiments
move related work to end
comments
Move commands so they can be used in any sec
Add performance and smape figures to text

@@
 \usepackage{cleveref}
+\usepackage{xspace}
 
+\newcommand{\srm}[1]{\textcolor{red}{SAM:#1}}
 \addbibresource{references.bib}
@@
 %  stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
-  stringstyle=\color{mauve},     % string literal style
+  stringstyle=\color{red},     % string literal style
   tabsize=2,	                   % sets default tabsize to 2 spaces
@@
 
-\newcommand{\ProjectName}{ImputeDB}
+\newcommand{\ProjectName}{ImputeDB\xspace}
+\newcommand{\sys}{\ProjectName}
 \newcommand{\doms}[2]{#1 \succ #2}
@@
 \newcommand{\todobox}[2]{{\color{red}\fbox{\parbox{\columnwidth}{\textbf{TODO:} #1\\#2}}}}
+
+% results that can be used in any section
+\newcommand{\demorows}{10175}
+\newcommand{\labexrows}{9813}
+\newcommand{\acsbaseresultminutes}{355 minutes}
+\newcommand{\acsbaseresulthours}{just under 6 hours}
+\newcommand{\acsimputedbzeroresult}{4 seconds}
+\newcommand{\acsimputedboneresult}{1 second}
+\newcommand{\lowxalphazero}{10}
+\newcommand{\lowsmapealphazero}{0}
+\newcommand{\highxalphaoneexacs}{1000}
+\newcommand{\highsmapealphaoneexacs}{24}
+\newcommand{\highxalphaone}{16000}
+
 \begin{document}
@@
 from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}. Our results show that the optimizer
-generates query plans that execute between \todo{add [low x - high x] range} as fast as imputing at the base
+generates query plans that execute between \lowxalphazero{} and \highxalphaoneexacs{} times as fast as imputing at the base
 tables and then executing queries, making exploratory data analysis using our system feasible. Furthermore, we show that the query results
-from on-the-fly imputation have Symmetric-Mean-Absolute-Percentage-Error between \todo{add SMAPE range}  
+from on-the-fly imputation have Symmetric-Mean-Absolute-Percentage-Error between \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent
 when compared to standard full-table imputation approach.
@@
 \input{examples}
-\input{relatedwork}
 \input{algorithm}
 \input{experiments}
+\input{relatedwork}
 \input{conclusion}
 \balance
+\bibstyle{abbrv}
 \printbibliography

### 2017-02-28 paper/algorithm.tex

@@
 To motivate the restriction of imputation types, we consider the implications of allowing arbitrary imputations.
-If we allow any arbitrary subset of attributes to be imputed, then we would need to consider $O(2^{|D|+1})$ different imputation operators before each relational operator.
+If we allow any arbitrary subset of attributes to be imputed, then we would need to consider $O(2^{|D|+1})$ different imputation operators before each relational operator where $D$ is the set of dirty attributes in all tables.
 This would increase the overall complexity of the algorithm drastically.

### 2017-02-28 paper/experiments.tex
Add SMAPE citation

@@
 
-\subsection{Results}
+\subsection{Results}\label{sec:results}
 
@@
 
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running
-imputation on the base tables and executing the query on this cleaned base data \todo{SMAPE citation}.
-Each query was separately run 200 times in both settings and then results from the two approaches were paired up and compared tuple-wise.
-We average tuple-wise absolute percentage deviations within each iteration of a query, and we report this value averaged over all iterations.
-We can see that optimizing for quality indeed reduces the SMAPE of query results.
-In general, the SMAPE relative to the base-imputation approach are low in all cases --- between \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly imputation produces
-similar results to imputation at the base tables.
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
+\ProjectName{}'s query results when compared to running imputation on the base tables and
+executing the query on this cleaned base data \cite{tofallis2015better}.  Each query was
+separately run 200 times in both settings and then results from the two approaches were
+paired up and compared tuple-wise.  We average tuple-wise absolute percentage deviations
+within each iteration of a query, and we report this value averaged over all iterations.  We
+can see that optimizing for quality indeed reduces the SMAPE of query results.  In general,
+the SMAPE relative to the base-imputation approach are low in all cases --- between
+\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly
+imputation produces similar results to imputation at the base tables.
 

### 2017-02-28 paper/algorithm.tex
paper: removed proof and replaced with paragraph

--- 2017-02-28 paper/algorithm.tex
@@
 
+While $\textsc{Loss}$ is not monotonically increasing in our model, we maintain a collection of sound and complete Pareto frontiers by 
+computing these specific to each dirty set. That is to say, that the combination of $\textsc{Loss}(p) \times \textsc{Dirty}(p)$ for some
+plan $p$ behaves monotonically, as the dirty-set is strictly decreasing through the application of $\delta, \mu$ operations. 
+The final Pareto frontier corresponds to the collection of plans associated with the empty dirty set.
+
 To reduce the search space, we only consider the minimal imputation, the maximal imputation, and the minimal drop.

### 2017-02-28 paper/introduction.tex
paper: remove todo in intro

@@
 While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression tree algorithm takes \acsbaseresulthours{} to train and run
-on a 600K row database; prior work~\cite{akande2015empirical} reports times of 45 minutes for multiple-imputation of missing values on a 1\% sample of the same database on a desktop computer.\todo{Explain the discrepancy between 45 * 100 and 355?
-They use L=10 datasets and unknown hardware.
-Or drop this comparison entirely?}
+on a 600K row database.
 

### 2017-02-28 paper/experiments.tex
paper: fix experiment todo

@@
 
-\subsection{Data sets} \label{subsec:datasets}
+\subsection{Data sets}\label{subsec:datasets}
 We collected three data sets our experiments.
 For all data sets, we selected a subset of the original attributes.
-We also enumerated strings and encoded them with an appropriate integer value.
-\todo{describe all data transformations. jose?}
+We also transformed all data values to an integer representation by enumerating strings and transforming floating point values into an appropriate range.
 

### 2017-02-28 paper/algorithm.tex
paper: pass through imputation methods section

--- 2017-02-28 paper/algorithm.tex
@@
 To reduce the search space, we only consider the minimal imputation, the maximal imputation, and the minimal drop.
-The minimal imputation (resp.
-drop) only imputes (resp.
-drops) the columns required by the relational operator immediately following the imputation.
+The minimal imputation (resp.\ drop) only imputes (resp.\ drops) the columns required by the relational operator immediately following the imputation.
 The maximal imputation imputes all of the attributes in the relation which are used by other operators in the query plan or are visible in the query output.
@@
 The planner operates as follows.
-First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}-\Cref{lst:line:attr-end}).
+First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}--\Cref{lst:line:attr-end}).
 This set will be used to determine which attributes are imputed.
@@
 plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
-operations (\Cref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
+operations (\Cref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e.\ the
 best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
@@
 
-The join optimization is based on the algorithm used in System R\cite{blasgen1981system},
+The join optimization is based on the algorithm used in System R~\cite{blasgen1981system},
 but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the dirty-set specific Pareto frontiers.
@@
 
-We design our system such that any imputation strategy can be plugged in with minimal effort
-and without changes to the optimizer, so the strategy can in principle be targeted to a
-specific domain. The current version of the system uses a general-purpose imputation
-strategy based on chained-equation classification and regression trees (CE-CART). Chained
-equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
-regression} \cite{gelman2006data}) impute multiple missing attributes by
-iteratively estimating predictive models of one missing attribute conditional on all
-complete attributes and the other missing attributes. The individual predictive models can
-be customized by the analyst to the problem at hand. Decision tree algorithms, like CART,
-are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
-routines and are widely used in epidemiological domains \cite{burgette2010multiple}. Though
-we provide quantitative results for the CE-CART implementation, ImputeDB is designed to be
-agnostic to the choice of imputation algorithm, and additional algorithms --- with
-associated time and loss functions --- could be inserted without any change in design.
+\ProjectName{} uses a general-purpose imputation strategy based on chained-equation classification and regression trees (CE-CART).
+Chained equation imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting a predictive model of one missing attribute which is conditional on all complete attributes and the other missing attributes.
+Decision tree algorithms like CART are found to be effective~\cite{akande2015empirical} and are widely used in epidemiological domains~\cite{burgette2010multiple}.
+
+\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal effort and without changes to the optimizer, simply by adding time and loss functions (\Cref{sec:cost}) which describe the new algorithm.
+In principle, new imputation algorithms can be added which are targeted to a specific domain.
 
@@
 as values progressively reflect more accurate relationships amongst attributes. The
-algorithm terminates when convergence is achieved (i.e. the imputed values do not change across
+algorithm terminates when convergence is achieved (i.e.\ the imputed values do not change across
 epochs) or a fixed number of epochs is reached.
@@
 By doing so we limit the number of imputations at any point to four: \textit{Drop} or \textit{Impute} over the attributes used in the operator or over all the downstream attributes.
-This modification increases the complexity of our planning algorithm to $O(16^J)$.\todo{I think this is right, but should check}
+This modification increases the complexity of our planning algorithm to $O(2^{4J})$.
 
@@
 The worst case complexity only occurs if the dirty sets tracked are distinct through the entire planning phase.
-Planning times are discussed in more detail in \Cref{sec:results}.
+Planning times are discussed in \Cref{sec:results}.
 

### 2017-02-28 simpledb/test/experiments/run_experiments.py
Add slow impute option

@@ -83,7 +83,7 @@ def run_acs_experiment():
     max_alpha = 1.00
     step      = 1.00
 
-    subprocess.call(executable +
+    subprocess.call(executable_longimpute +
         ["experiment", catalog, acs_query, this_output_dir,
          str(iters), str(min_alpha), str(max_alpha), str(step)])
 

### 2017-02-28 simpledb/test/experiments/analyze_experiments.py
Update analyze_experiments

@@ -5,11 +5,6 @@ from __future__ import print_function
 import pandas as pd
 import numpy as np
 
-import matplotlib
-matplotlib.use('Agg')
-import matplotlib.pyplot as plt
-plt.ioff()
-
 import os
 import glob
 import sys
@@ -46,29 +41,6 @@ def drop_warmup(df, by, drop=20):
 def summarize(df, by, a):
     ops = {'mean': np.mean, 'std': np.std}
     return df.groupby(by)[a].agg(ops).reset_index()
-  
-def myplot(df,alphas=None,**kwargs):
-    _, ax = plt.subplots(1)
-
-    if alphas is not None:
-        aset = set(alphas).intersection(set(df['alpha']))
-    else:
-        aset = set(df['alpha'])
-
-    if set(aset).issubset(set([0.0,1.0])):
-        label_str = r'$\alpha=%.0f$'
-    else:
-        label_str = r'$\alpha=%.2f$'
-
-    for a in aset:
-        data = df[df['alpha'] == a]
-        try:
-            data.plot(ax=ax, label=(label_str % a), **kwargs)
-        except TypeError:
-            data.plot(ax=ax, label=a, **kwargs)
-        # get around bug in legend drawing in pandas
-        plt.legend()
-    return ax
   
 def get_query_results(experiments_dir, headers, restrict_alpha=False, base=False):
     # results has
@@ -168,17 +140,9 @@ def explore(experiments_dir, base=False):
     planning_times = summarize(data, by, 'plan_time')
     running_times = summarize(data, by, 'run_time')
 
-    print(planning_times)
-    print(running_times)
-
     return planning_times, running_times, data
 
-def main(experiments_dir, output_dir):
-    make_plots(experiments_dir, output_dir)
-    make_plots(experiments_dir, output_dir, base=True)
-    write_perf_summary(experiments_dir, output_dir)
-
-def make_plots(experiments_dir, output_dir, base=False):
+def collapse_results(experiments_dir, output_dir, base=False):
     if not os.path.isdir(experiments_dir):
         raise FileNotFoundError
 
@@ -200,10 +164,8 @@ def make_plots(experiments_dir, output_dir, base=False):
     by = ['query', 'alpha']
     if base:
         name = 'base_tables'
-        alphas = ['Impute at base tables']
     else:
[…]

### 2017-02-28 paper/algorithm.tex

@@
 
+\todobox{jose rewrite this clearly}{
+While $\textsc{Loss}$ is not monotonically increasing in our model, we maintain a collection of sound and complete Pareto frontiers by 
+computing these specific to each dirty set. That is to say, that the combination of $\textsc{Loss}(p) \times \textsc{Dirty}(p)$ for some
+plan $p$ behaves monotonically, as the dirty-set is strictly decreasing through the application of $\delta, \mu$ operations. 
+The final Pareto frontier corresponds to the collection of plans associated with the empty dirty set.}
+
 To reduce the search space, we only consider the minimal imputation, the maximal imputation, and the minimal drop.

### 2017-02-28 paper/algorithms/plan-helpers.tex

@@
     \Require{$q$ is a query plan, $G$ is a set of grouping attributes, $A$ is an aggregation function, $C_g$ is a set of attributes which are used in the final plan.}
-    \Ensure{A set of plans for grouping \& aggregating with $G$ and $A$.}
+    \Ensure{A set of plans for grouping and aggregating with $G$ and $A$.}
     \Function{OptGroupBy}{$q, G, A, C_g$}
@@
 
-    \Require{$q$ is a query plan, $P$ is a set of projection attributes.}
+    \Require{$q$ is a query plan, $P$ is a set of attributes to be projected at the top of the plan.}
     \Ensure{A set of plans with the attributes in $P$ projected.}

### 2017-02-28 paper/algorithms/plan-toplevel.tex

@@
 \begin{algorithmic}[1]
-  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses trade-off performance and information lost}
+  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses the trade-off between performance and imputation quality.}
   \Ensure{An optimized query plan.}

### 2017-02-28 simpledb/test/experiments/run_experiments.py
Options for new queries

@@ -57,6 +57,32 @@ def run_alt_experiment():
     run_experiment(this_output_dir, iters, min_alpha, max_alpha, step, queries =
             queries, executable = executable_longimpute)
 
+def run_large03_experiment():
+    this_output_dir = os.path.join(output_dir, "large03")
+
+    iters     = 220
+    min_alpha = 0.00
+    max_alpha = 1.00
+    step      = 1.00
+
+    queries = "queries03.txt"
+
+    run_experiment(this_output_dir, iters, min_alpha, max_alpha, step, queries =
+            queries, executable = executable_longimpute)
+
+def run_large04_experiment():
+    this_output_dir = os.path.join(output_dir, "large04")
+
+    iters     = 220
+    min_alpha = 0.00
+    max_alpha = 1.00
+    step      = 1.00
+
+    queries = "queries04.txt"
+
+    run_experiment(this_output_dir, iters, min_alpha, max_alpha, step, queries =
+            queries, executable = executable_longimpute)
+
 def run_acs_experiment():
     catalog = catalog_default
     executable = executable_default
@@ -128,6 +154,10 @@ if __name__ == "__main__":
         run_large_experiment()
     elif experiment_size == "alt":
         run_alt_experiment()
+    elif experiment_size == "large03":
+        run_large03_experiment()
+    elif experiment_size == "large04":
+        run_large04_experiment()
     elif experiment_size == "acs":
         run_acs_experiment()
     else:

### 2017-02-28 paper/algorithm.tex
paper: rewrote dominance definition, changed plan cache explanation to partial order

@@
 with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
-aggregations. The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
-quality of the imputations performed and the runtime performance. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
-algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). Our imputation planning algorithm is
+aggregations. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
+algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). 
+The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
+quality of the imputations performed and the runtime performance. Our imputation planning algorithm is
 agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
@@
 
+\subsection{Imputation placement}
+\label{sec:placement}
+Imputation operators must be placed so that no relational operator receives a tuple containing null in an attribute that the operator examines, regardless of the state of the data in the base tables.
+
+Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required.
+To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$.
+An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null.
+We compute a dirty set for each table using the table histograms, which track the number of null values in each column.
+The dirty set for a query plan can be computed recursively as follows:
+
+\begin{align*}
+  \textsc{Dirty}(q) = \begin{cases}
+    \textsc{Dirty}(q') \setminus C & q = \delta_C(q')\ \text{or}\ q = \mu_C(q') \\
+    \textsc{Dirty}(q') \cap C & q = \pi_C(q') \\
+    \textsc{Dirty}(q_l) \cup \textsc{Dirty}(q_r) & q = q_l \bowtie_\psi q_r \\
+    \textsc{Dirty}(q') & q = \sigma_\phi(q') \\
+    \textsc{Dirty}(t) & q = \text{some table}\ t \\
+  \end{cases}
+\end{align*}
+
+Note that \textsc{Dirty} over-approximates the set of attributes that contain null.
+For example, a filter might remove all tuples which contain null, but the dirty set would be unchanged.
+We choose to over-approximate to ensure that all null values are explicitly imputed or dropped.
+
 \subsection{Cost model}
@@
 \begin{definition}
-We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$ if $(l \leq l' \land t < t') \lor (l < l' \land t \leq t')$.
+We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(p) = \textsc{Dirty}(p') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$.
 We denote this as $\doms{p}{p'}$.
@@
 
-\subsection{Imputation placement}
-\label{sec:placement}
-Imputation operators must be placed so that no relational operator receives a tuple containing null in an attribute that the operator examines, regardless of the state of the data in the base tables.
-
-Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required.
-To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$.
-An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null.
-We compute a dirty set for each table using the table histograms, which track the number of null values in each column.
-The dirty set for a query plan can be computed recursively as follows:
-
-\begin{align*}
-  \textsc{Dirty}(q) = \begin{cases}
-    \textsc{Dirty}(q') \setminus C & q = \delta_C(q')\ \text{or}\ q = \mu_C(q') \\
-    \textsc{Dirty}(q') \cap C & q = \pi_C(q') \\
-    \textsc{Dirty}(q_l) \cup \textsc{Dirty}(q_r) & q = q_l \bowtie_\psi q_r \\
-    \textsc{Dirty}(q') & q = \sigma_\phi(q') \\
-    \textsc{Dirty}(t) & q = \text{some table}\ t \\
-  \end{cases}
-\end{align*}
-
-Note that \textsc{Dirty} over-approximates the set of attributes that contain null.
-For example, a filter might remove all tuples which contain null, but the dirty set would be unchanged.
-We choose to over-approximate to ensure that all null values are explicitly imputed or dropped.
-
 \subsection{Query planning}
@@
                                                                      \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
-                                                                     \forall q' \in \llbracket Q[T] \rrbracket, \textsc{Dirty}(q')=\textsc{Dirty}(q).  \doms{q} {q'} 
+                                                                     \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
                                                                    \end{aligned}
@@
 
-\todobox{jose rewrite this clearly}{
-While $\textsc{Loss}$ is not monotonically increasing in our model, we maintain a collection of sound and complete Pareto frontiers by 
-computing these specific to each dirty set. That is to say, that the combination of $\textsc{Loss}(p) \times \textsc{Dirty}(p)$ for some
-plan $p$ behaves monotonically, as the dirty-set is strictly decreasing through the application of $\delta, \mu$ operations. 
-The final Pareto frontier corresponds to the collection of plans associated with the empty dirty set.}
+The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Loss}, \textsc{Time} \rangle$ to
+collect sound and complete Pareto frontiers. For plans with the same dirty set, we can treat $\langle \textsc{Loss}, \textsc{Time} \rangle$
+as a total order, allowing us to appropriately select the frontier. Meanwhile, plans with different $\textsc{Dirty}$ sets cannot be compared. The final
+Pareto frontier corresponds to the collection of plans associate with the empty dirty set.
 

### 2017-02-28 paper/experiments.tex
paper: fcc -> freeCodeCamp

@@
 For our second set of experiments, we chose to use data collected
-by freeCodeCamp (fCC) as a part of a survey of new software developers
+by freeCodeCamp as a part of a survey of new software developers
 (both professional and amateur)~\cite{fcc-data}. freeCodeCamp is an open-source
@@
 
-Our second set of queries is on the fCC data (\Cref{fig:queries-fcc}).
+Our second set of queries is on the freeCodeCamp data (\Cref{fig:queries-fcc}).
 \Cref{q5} calculates the average income for survey participants, based on their bootcamp attendance.
@@
     \input{queries-fcc}
-    \caption{Queries on fCC data}\label{fig:queries-fcc}
+    \caption{Queries on freeCodeCamp data}\label{fig:queries-fcc}
   \end{subtable}

### 2017-02-28 paper/main.tex
paper: add sam as author
paper: title change
Update perf number in abstract

@@
 
-\title{\ProjectName{}: Data Imputation as a Query Optimization}
+\title{Query Optimization for Dynamic Imputation}
 
@@
   Micah J. Smith \\
-  \texttt{micahs@mit.edu}}
+  \texttt{micahs@mit.edu}
+  \and
+  Samuel Madden \\
+  \texttt{madden@csail.mit.edu}}
 
@@
 \newcommand{\lowsmapealphazero}{0}
+\newcommand{\highxalphazeroexacs}{200}
 \newcommand{\highxalphaoneexacs}{1000}
@@
 are forced to pick between removing tuples with missing values or creating a clean duplicate of their data
-by  applying a potentially-expensive imputation strategy. We show that it is possible
-to incorporate imputation into a cost-based query optimizer, performing imputation on the fly
+by  applying a potentially-expensive imputation strategy. Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, 
+performing imputation on the fly
 for each query. This allows users to immediately explore
@@
 for traditional optimizations to take place. Our experiments evaluate this approach on two real-world survey-based datasets
-from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}. Our results show that the optimizer
-generates query plans that execute between \lowxalphazero{} and \highxalphaoneexacs{} times as fast as imputing at the base
-tables and then executing queries, making exploratory data analysis using our system feasible. Furthermore, we show that the query results
-from on-the-fly imputation have Symmetric-Mean-Absolute-Percentage-Error between \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent
-when compared to standard full-table imputation approach.
+from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}. Our
+results show that the optimizer generates query plans that execute between \lowxalphazero{}
+and \highxalphazeroexacs{} times as fast as imputing at the base tables and then executing
+queries, making exploratory data analysis using our system feasible.  Furthermore, we show
+that the query results from on-the-fly imputation have
+Symmetric-Mean-Absolute-Percentage-Error between \lowsmapealphazero{} and
+\highsmapealphaoneexacs{} percent when compared to standard full-table imputation
+approach.
+
 \end{abstract}
@@
 \input{conclusion}
-\balance
 \bibstyle{abbrv}
 \printbibliography
+\balance
 

### 2017-02-28 paper/algorithm.tex
Update imputation strategy session
Clarify relation to multiple imputation

@@
 
-\ProjectName{} uses a general-purpose imputation strategy based on chained-equation classification and regression trees (CE-CART).
-Chained equation imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting a predictive model of one missing attribute which is conditional on all complete attributes and the other missing attributes.
-Decision tree algorithms like CART are found to be effective~\cite{akande2015empirical} and are widely used in epidemiological domains~\cite{burgette2010multiple}.
+\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal
+effort and without changes to the optimizer, simply by adding time and loss functions
+(\Cref{sec:cost}) which describe the new algorithm. In principle, new imputation algorithms
+can be added, possibly targeted to a specific domain. The flexibility of this approach
+aligns well with the spirit of query optimization, which aims to reconcile performance with
+a declarative interface to data manipulation.
 
-\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal effort and without changes to the optimizer, simply by adding time and loss functions (\Cref{sec:cost}) which describe the new algorithm.
-In principle, new imputation algorithms can be added which are targeted to a specific domain.
+As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
+based on chained-equation decision trees.  Chained equation
+imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
+regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
+a predictive model of one missing attribute, conditional on all complete attributes
+and all other missing attributes. Chained-equation decision tree algorithms are found to be
+effective~\cite{akande2015empirical} and are widely used in epidemiological
+domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
+implementation (\cite{witten2016data}).
 
-These imputation algorithms are rarely used in isolation --- rather, the ultimate goal is to
-perform statistical analysis of the complete data. To this end, the technique of
-\textit{multiple imputation} can be used, in which multiple distinct copies of the complete
-data are generated and estimators are computed by averaging over the multiple datasets. In
-ImputeDB, we cannot assume the user desires multiple copies of the query result, though we
-note that in further work, multiple imputation could be used within ImputeDB to reduce the
-variance of estimates of some aggregates, like averages.
+Our imputation algorithm proceeds by iteratively fitting regression trees to a subset of the
+data and the target imputation columns. In each iteration, the missing values of a column are
+replaced with newly imputed values. With each {\it epoch}, the quality of the imputation
+improves as values progressively reflect more accurate relationships amongst attributes. The
+algorithm terminates when convergence is achieved (i.e.\ the imputed values do not change
+across epochs) or a fixed number of epochs is reached.
 
-Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
-the target imputation columns. In each iteration, the missing values of a column are
-replaced with newly imputed values. With each epoch, the quality of the imputation improves
-as values progressively reflect more accurate relationships amongst attributes. The
-algorithm terminates when convergence is achieved (i.e.\ the imputed values do not change across
-epochs) or a fixed number of epochs is reached.
+These imputation algorithms are used as part of a larger \textit{multiple imputation}
+process, in which multiple distinct copies of the complete data are generated and estimators
+are computed by averaging over the multiple datasets to form a single completed dataset
+(\cite{gelman2006data}). In \ProjectName{}, we consider only single imputation, though in
+further work, multiple imputation could be considered as another avenue of assessing
+imputation confidence.
 
-The flexibility of this non-parametric imputation approach aligns well with the spirit of
-query optimization, which aims to reconcile performance with a declarative interface to data
-manipulation.
-
-It is worth noting that given the nature of the imputation strategy used here, imputation
-operators become blocking in our system. An extension of ImputeDB could consider regression
-algorithms that learn in an online nature; this would reduce the cost of the imputation at
-the possible expense of quality. Again, given the imputation strategy-agnostic design of
-ImputeDB, this would not pose a problem.
+Given the nature of the imputation strategy used here, {\it Impute} operators are blocking
+in our system. An extension of \ProjectName{} could consider regression algorithms that learn in
+an online nature; this would reduce the cost of the imputation at the possible expense of
+quality.
 

### 2017-02-28 paper/experiments.tex
Remove fcc references

@@
 \subsubsection{freeCodeCamp 2016 New Coder Survey}
-For our second set of experiments, we chose to use data collected
-by freeCodeCamp as a part of a survey of new software developers
-(both professional and amateur)~\cite{fcc-data}. freeCodeCamp is an open-source
-community that helps users learn how to program. Their \textit{2016 New Coder Survey} consists of responses by over 15,000 people to 48 different
-demographic and programming-related questions.
-The survey targeted users who were related to coding organizations.
+For our second set of experiments, we use data collected by freeCodeCamp, an open-source
+community for learning to code, as a part of a survey of new software
+developers~\cite{fcc-data}.  The \textit{2016 New Coder Survey} consists of responses by
+over 15,000 people to 48 different demographic and programming-related questions.  The
+survey targeted users who were related to coding organizations.
 
 We use a version of the data that has been pre-processed, but where missing values remain.
-For example, 46.6\% of entries for entry \textit{commutetime} are missing. However, it is worth noting that some of the missing values
-are also expected, given the way the data has been de-normalized. For example, \textit{bootcamploanyesno}, a binary attribute encoding
-whether a respondent had a loan for a bootcamp, is expected to be null for participants who did not attend a bootcamp.
+For example, 46.6\% of \textit{commutetime} responses are missing. However, it is worth
+noting that some of the missing values are also expected, given the way the data has been
+de-normalized. For example, \textit{bootcamploanyesno}, a binary attribute encoding whether
+a respondent had a loan for a bootcamp, is expected to be null for participants who did not
+attend a bootcamp.
 
-We chose a subset of 17 attributes, shown in~\Cref{table:fcc-description}, along with the percentage of missing values.
+We choose a subset of 17 attributes, which are shown in~\Cref{table:fcc-description} along
+with the percentage of missing values.
 
@@
 \subsubsection{American Community Survey}
-For our final experiment, we introduce a simple aggregate query over data from the American Community Survey (ACS), which
-provides a number of public data sets collected by the U.S. Census Bureau.
-We used a cleaned version of the 2012 Public Use Microdata Sample (PUMS) data kindly provided by the authors of~\cite{akande2015empirical}.
-Given that the data had been cleaned, we artificially dirtied it by removing 40\% of the values uniformly at random.
+For our final experiment, we run a simple aggregate query over data from the American
+Community Survey (ACS), a comprehensive survey conducted by the U.S.
+Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS)
+data kindly provided by the authors of~\cite{akande2015empirical}. Given that the data had
+been cleaned, we artificially dirtied it by removing 40\% of the values uniformly at random.
 The final dataset consists of 671,153 rows and 37 integer columns.
@@
 \subsection{Queries}
-We collected a set of queries (\Cref{fig:queries}) that we think are interesting to plan.
+We collect a set of queries (\Cref{fig:queries}) that we think are interesting to plan.
 We believe that they could reasonably be written by a user in the course of data analysis.
 
-The queries on the CDC NHANES data consist not only of projections and selections, but also interesting joins and aggregates.
-Our aim was to craft meaningful queries
-that would provide performance figures relevant to practitioners using
-similar datasets.
+The queries on the CDC NHANES data consist not only of projections and selections, but also
+interesting joins and aggregates.  Our aim was to craft meaningful queries that would
+provide performance figures relevant to practitioners using similar datasets.
 
-Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}). In~\Cref{q1}, we
-calculate the average cuff size for users based on their income data. In~\Cref{q2}, we compare
-creatine levels for individuals with low, medium, and high incomes spectrum and above a
-certain weight. In~\Cref{q3}, we extract the average blood lead levels for children under
-the age of 6 years. In~\Cref{q4}, we calculate the average systolic blood pressure, by
-gender, for subjects with a body mass index indicating obesity.
+Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}).
+\Cref{q1} calculates calculate the average cuff size for users based on their income data.
+\Cref{q2} compares creatine levels for individuals with low, medium, and high incomes spectrum and above a certain weight.
+\Cref{q3} extracts the average blood lead levels for children under the age of 6 years.
+\Cref{q4} calculates the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity.
 

### 2017-02-28 paper/experiments.tex
Move planing times results to commands

@@
 In the median query across a variety of queries and choices of $\alpha$, planning
-constituted 8.5 percent of total runtime. In all cases, the optimizer
-returned a query plan within 40ms, with times roughly constant between levels of $\alpha$.
+constituted \planningruntimepercent{} percent of total runtime. In all cases, the optimizer
+returned a query plan within \planningmaxtime{} ms, with times roughly constant between levels of $\alpha$.
 

### 2017-03-01 paper/algorithm.tex
paper: null made macro, since had null and NULL

@@
 attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
-replace all null values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
+replace all \nullv{} values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
 In our experiments, we instantiate \textit{Impute} with the CE-CART\cite{burgette2010multiple} algorithm, but we highlight that
 our system is agnostic to this choice.
-\textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
-Both operators guarantee that the resulting relation will contain no null values for
+\textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
+Both operators guarantee that the resulting relation will contain no \nullv{} values for
 attributes in $C$.
@@
 \label{sec:placement}
-Imputation operators must be placed so that no relational operator receives a tuple containing null in an attribute that the operator examines, regardless of the state of the data in the base tables.
+Imputation operators must be placed so that no relational operator receives a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.
 
-Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required.
+Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a \nullv{} value, there are cases where an imputation operator is required.
 To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$.
-An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null.
-We compute a dirty set for each table using the table histograms, which track the number of null values in each column.
+An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain \nullv{}.
+We compute a dirty set for each table using the table histograms, which track the number of \nullv{} values in each column.
 The dirty set for a query plan can be computed recursively as follows:
@@
 
-Note that \textsc{Dirty} over-approximates the set of attributes that contain null.
-For example, a filter might remove all tuples which contain null, but the dirty set would be unchanged.
-We choose to over-approximate to ensure that all null values are explicitly imputed or dropped.
+Note that \textsc{Dirty} over-approximates the set of attributes that contain \nullv{}.
+For example, a filter might remove all tuples which contain \nullv{}, but the dirty set would be unchanged.
+We choose to over-approximate to ensure that all \nullv{} values are explicitly imputed or dropped.
 
@@
 Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
-the dropped tuples might have encoded). Imputing null fields incurs a loss penalty $p \in (0, 1]$,
+the dropped tuples might have encoded). Imputing \nullv{} fields incurs a loss penalty $p \in (0, 1]$,
 which is inversely proportional to the number of total cells available to train the imputation algorithm. 

### 2017-03-01 paper/experiments.tex
paper: null made macro, since had null and NULL

--- 2017-02-28 paper/experiments.tex
@@
 We focused on a subset of the attributes for each table to simplify the presentation and exploration of queries.
-\Cref{table:nhanes-description} shows the attributes selected, along with the percentage of null values for each attribute.
+\Cref{table:nhanes-description} shows the attributes selected, along with the percentage of \nullv{} values for each attribute.
 For readability, we have replaced the NHANES variable names with self-explanatory attribute names.
@@
 de-normalized. For example, \textit{bootcamploanyesno}, a binary attribute encoding whether
-a respondent had a loan for a bootcamp, is expected to be null for participants who did not
+a respondent had a loan for a bootcamp, is expected to be \nullv{} for participants who did not
 attend a bootcamp.

### 2017-03-01 paper/introduction.tex
paper: null made macro, since had null and NULL

@@
 
-Many databases have large numbers of missing or NULL values;  these can arise for a variety of reasons, including missing source data (e.g., unknown facts),
-missing columns during data integration, de-normalized databases, or as a result of outlier detection and cleaning where NULL values are substituted for
-bad values~\cite{kim2003}.  Such NULL values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
+Many databases have large numbers of missing or \nullv{} values;  these can arise for a variety of reasons, including missing source data (e.g., unknown facts),
+missing columns during data integration, de-normalized databases, or as a result of outlier detection and cleaning where \nullv{} values are substituted for
+bad values~\cite{kim2003}.  Such \nullv{} values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
 before processing is often desirable.
@@
 This process is called \emph{imputation}.
-In this paper, we present a new system, called \ProjectName{}, which is the designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.
+In this paper, we present a new system, called \ProjectName{}, which is designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.
 \emph{The key insight behind \ProjectName{} is that imputation
@@
   Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
-  Drop operation simply drops tuples which contain null values.
+  Drop operation simply drops tuples which contain \nullv{} values.
 \item \textbf{Model of imputation quality and cost:}

### 2017-03-01 paper/relatedwork.tex
paper: remove formal proof
paper: null made macro, since had null and NULL

@@
 There is a long history in the database community surrounding the
-treatment of nulls. As early as 1973,~\cite{codd1973understanding}
-provides a treatment of the semantics of null. Multiple
+treatment of \nullv{}. As early as 1973,~\cite{codd1973understanding}
+provides a treatment of the semantics of \nullv{}. Multiple
 papers have described various (at times conflicting) treatments
@@
 missing data - eliminates
-the need to handle null value semantics, while guaranteeing soundness (modulo
+the need to handle \nullv{} value semantics, while guaranteeing soundness (modulo
 imputation strategy) of the query evaluation.
@@
 it provides the ability
-to trade imputation quality for runtime performance. We prove our
-algorithm and cost model produces an optimal set of plans to choose from,
-given the constraints of the search space.
+to trade imputation quality for runtime performance. 
+%We prove our
+%algorithm and cost model produces an optimal set of plans to choose from,
+%given the constraints of the search space.
 

### 2017-03-01 paper/algorithm.tex
paper: some minor fixes and todos added

@@
 \label{sec:placement}
-Imputation operators must be placed so that no relational operator receives a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.
+Imputation operators must be placed so that no relational operator encounters a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.
 
@@
 Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
-We say $p$ is $\alpha$-bound optimal if $\forall p'^\alpha \in P: \textsc{Time}(p) \leq \textsc{Time}(p')$. 
+We say $p$ is $\alpha$-bound optimal if $\forall p'^\alpha \in P: \textsc{Time}(p) \leq \textsc{Time}(p'^\alpha)$. 
 \end{definition}
@@
 
-For each sub-plan, we keep track of the estimated cardinality by using the histograms associated with
-that plan. We use this estimate to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.
+We use these histogram estimates to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.
 
@@
 Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
-the dropped tuples might have encoded). Imputing \nullv{} fields incurs a loss penalty $p \in (0, 1]$,
+the dropped tuples might have encoded). 
+\todobox{fix this explanation, not inverseley proprtional to what is stated here}{Imputing \nullv{} fields incurs a loss penalty $p \in (0, 1]$,
 which is inversely proportional to the number of total cells available to train the imputation algorithm. 
@@
 correspondingly, the information lost \textit{decreases} with more values to work
-with. Since we conceive that
+with.}
+Since we conceive that
 alternate imputation strategies could be used, the analyst may provide a corresponding
@@
 At a high level, a plan cache is a mapping from a set of tables to a set of dirty-set specific dominating plans that perform the required joins and filters over these tables.
-We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes, each of which maintains its own Pareto frontier. 
-Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.
+We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
+allows us to obtain the final set of plans that best tradeoff computation
+cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.
 
@@
 as a total order, allowing us to appropriately select the frontier. Meanwhile, plans with different $\textsc{Dirty}$ sets cannot be compared. The final
-Pareto frontier corresponds to the collection of plans associate with the empty dirty set.
-
-To reduce the search space, we only consider the minimal imputation, the maximal imputation, and the minimal drop.
-The minimal imputation (resp.\ drop) only imputes (resp.\ drops) the columns required by the relational operator immediately following the imputation.
-The maximal imputation imputes all of the attributes in the relation which are used by other operators in the query plan or are visible in the query output.
+Pareto frontier produced by the optimizer corresponds to the collection of plans associated with the empty dirty set.
 
@@
 The join optimization is based on the algorithm used in System R~\cite{blasgen1981system},
-but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the dirty-set specific Pareto frontiers.
+but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the Pareto frontiers.
 The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.
@@
 effort and without changes to the optimizer, simply by adding time and loss functions
-(\Cref{sec:cost}) which describe the new algorithm. In principle, new imputation algorithms
+(\Cref{sec:cost-model}) which describe the new algorithm. In principle, new imputation algorithms
 can be added, possibly targeted to a specific domain. The flexibility of this approach
@@
 Finally, we note that for the queries we have examined, this exponential blowup does not affect the practical performance of our optimizer.
-Recall that the planner maintains different plans for different dirty sets, keeping only those that are not dominated by other plans with the same dirty set.
-So in many cases we can drop some of these intermediate plans generated at each operator.
+Recall that the planner maintains different plans for different dirty sets, keeping only those plans that are not dominated by others.
+So in many cases we can drop some of the intermediate plans generated at each operator.
 The worst case complexity only occurs if the dirty sets tracked are distinct through the entire planning phase.

### 2017-03-01 paper/experiments.tex
paper: some minor fixes and todos added

@@
 \subsection{Data sets}\label{subsec:datasets}
-We collected three data sets our experiments.
+We collected three data sets for our experiments.
 For all data sets, we selected a subset of the original attributes.
@@
 data kindly provided by the authors of~\cite{akande2015empirical}. Given that the data had
-been cleaned, we artificially dirtied it by removing 40\% of the values uniformly at random.
+been cleaned, we artificially dirtied it by replacing 40\% of the values uniformly at random with \nullv{} values.
 The final dataset consists of 671,153 rows and 37 integer columns.
@@
 configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0$),
-\ProjectName{} optimizing for performance($\alpha=1$), and imputing at the base tables followed
+\ProjectName{} optimizing for performance ($\alpha=1$), and imputing at the base tables followed
 by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
@@
 
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)\todo{cite SMAPE} for
 \ProjectName{}'s query results when compared to running imputation on the base tables and
@@
 \begin{table}
+\todo{extend this table to have counts with new results}
 \centering
@@
 this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are not known until runtime.
-Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.
+
+\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}
 

### 2017-03-01 paper/relatedwork.tex
paper: some minor fixes and todos added

@@
 treatment of \nullv{}. As early as 1973,~\cite{codd1973understanding}
-provides a treatment of the semantics of \nullv{}. Multiple
+provided a treatment of the semantics of \nullv{}. Multiple
 papers have described various (at times conflicting) treatments
@@
 missing data - eliminates
-the need to handle \nullv{} value semantics, while guaranteeing soundness (modulo
-imputation strategy) of the query evaluation.
+the need to handle \nullv{} value semantics, while guaranteeing query evaluation soundness.
 
@@
 is partially based on the seminal work developed for System R's query planning\cite{blasgen1981system}.
-However, in contrast to System R, \ProjectName{} performs additional histogram transformations to account
-for the changing nature of missing values.
+However, in contrast to System R, \ProjectName{} performs additional optimizations for imputing missing data and
+uses histogram transformations to account for the changing nature of missing values.
 

### 2017-03-01 paper/algorithm.tex
paper: clarified definition of Loss in text, as per sam's feedback

@@
 
-\srm{I find this table hard to follow -- I think it'd be good to explain one or two of the rows in detail, and to explain what the set L contains.}
-Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
-the dropped tuples might have encoded). 
-\todobox{fix this explanation, not inverseley proprtional to what is stated here}{Imputing \nullv{} fields incurs a loss penalty $p \in (0, 1]$,
-which is inversely proportional to the number of total cells available to train the imputation algorithm. 
-Total cells are defined as the product of the number of attributes and the number of tuples.
-Intuitively, imputation
+Each imputation operator incurs a heuristic loss penalty. We define \textsc{Loss} as the average over the individual penalties incurred
+by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties in a multi-set, as the order of the penalties is irrelevant,
+and averages them at the root of the plan. For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
+might have encoded). Meanwhile, imputing \nullv{} fields with $\mu$ incurs a loss penalty $p \in (0, 1]$. In this case, the penalty is
+inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as
+as the product of the number of attributes and the number of tuples. Intuitively, imputation
 accuracy \textit{increases} when more complete attributes and complete tuples are available;
 correspondingly, the information lost \textit{decreases} with more values to work
-with.}
+with. The square root captures the intuition that the benefits provided by more training observations has diminishing returns.
+
 Since we conceive that

### 2017-03-01 paper/experiments.tex
paper: todos
Update figures for new query results
paper: temporary caption and paragraph explaining count stuff

@@
 \begin{table*}
+\todo{update this with final queries}
   \centering
@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
+\todo{Relabel 9 somehow, since it is no longer in table of queries but in lined later}
+\includegraphics[width=\columnwidth]{figures/running_times_combined_bar_13.png}
 \caption{Query runtimes with imputation on base tables, \ProjectName{}
@@
 within each iteration of a query, and we report this value averaged over all iterations.  We
-can see that optimizing for quality indeed reduces the SMAPE of query results.  In general,
-the SMAPE relative to the base-imputation approach are low in all cases --- between
+can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
+to the base-imputation approach are low in all cases --- between
 \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly
 imputation produces similar results to imputation at the base tables.
+
+We also calculate SMAPE over the number of tuples used to produce the results
+in each configuration with respect to the number of tuples satisfying the same query when
+the base tables have been imputed. In many cases where the SMAPE reduction between $\alpha = 1$ and
+$\alpha = 0$ is small, for example \Cref{q6}, we highlight that the count of tuples can be significantly different between
+the two scenarios. So while the aggregate used might not have been significantly affected by the different population,
+other aggregates may well be. This highlights a challenge for a user handling data imputation traditionally: it is unclear if
+the missing data will have a large or small negative impact on their analysis until they have paid the cost of
+running it. By using \ProjectName{} this cost can be lowered significantly.
 
@@
 \centering
-\input{tables/perf_summary.tex}
+\input{tables/perf_summary_13.tex}
 \caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
     parameterizations relative to results when imputing on base table. Queries optimized for quality ($\alpha=0$) generally achieve
-    lower error than queries optimized for efficiency ($\alpha=1$).}
+    lower error than queries optimized for efficiency ($\alpha=1$). Count values reflect the SMAPE over the number of tuples
+    used for each configuration with respect to the number of tuples used when executing the query after base table imputation. A higher
+    count difference reflects more potential for errors.}
 \label{table:smape}
@@
 
-\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}
+%\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}
 

### 2017-03-01 paper/experiments.tex
Update count interpretation

@@
 \begin{table}
-\todo{extend this table to have counts with new results}
 \centering
@@
 \caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
-    parameterizations relative to results when imputing on base table. Queries optimized for quality ($\alpha=0$) generally achieve
-    lower error than queries optimized for efficiency ($\alpha=1$). Count values reflect the SMAPE over the number of tuples
-    used for each configuration with respect to the number of tuples used when executing the query after base table imputation. A higher
-    count difference reflects more potential for errors.}
+    parameterizations relative to results when imputing on the base table. Queries optimized
+    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
+    efficiency ($\alpha=1$). Count shares show the number of results returned from each
+    query as a percentage of the number of results returned when imputing on the base table.
+    A lower count share reflects more potential for errors.}
 \label{table:smape}
@@
 %%% Local Variables:
+
 %%% mode: latex

### 2017-03-01 simpledb/test/experiments/analyze_experiments.py
Calculate count values and count shares

@@ -209,6 +209,73 @@ def write_perf_summary(experiments_dir, output_dir):
     perf_summary_latex.to_latex(os.path.join(output_dir, 'perf_summary.tex'),
             float_format='%.2f', index=False)
 
+    return perf
+
+def write_counts_summary(experiments_dir, output_dir):
+    if not os.path.isdir(experiments_dir):
+        raise FileNotFoundError
+
+    if not os.path.isdir(output_dir):
+        os.makedirs(output_dir)
+
+    experiment_results = get_query_results(experiments_dir, table_headers)
+    base_results = get_query_results(experiments_dir, table_headers, base=True)
+
+    def compute_average_count_items(d):
+        try:
+            myitems = d.iteritems()
+        except AttributeError:
+            myitems = d.items()
+
+        df = pd.DataFrame(columns=["query","alpha","mean","std"])
+        for (query, alpha), res in myitems:
+            sums = []
+            for df1 in res:
+                s = df1.sum()
+                if len(s) != 1:
+                    print(len(s))
+                    print(query, alpha)
+                    raise Exception
+                sums.append(s.values[0])
+            avg = np.mean(sums)
+            std = np.std(sums)
+            df = df.append(pd.DataFrame(
+                data={"query": [query],
+                      "alpha": [alpha],
+                      "mean": [avg],
+                      "std": [std],
+                }))
+
+        return df
+
+    df1 = compute_average_count_items(experiment_results)
+    df2 = compute_average_count_items(base_results)
+
+    df2["alpha"] = -1.0 # impute at base table
+
+    # Write to file
+    df = df1.append(df2).sort_values(["query"])
+
+    print("Writing means...")
+    dfmean = df.pivot(index="query",columns="alpha",values="mean") 
+    dfmean.to_latex(os.path.join(output_dir, 'counts_mean.tex'),
+            float_format='%.2f')
+
+    print("Writing stds...")
+    dfstd = df.pivot(index="query",columns="alpha",values="std")
+    dfstd.to_latex(os.path.join(output_dir, 'counts_std.tex'),
+            float_format='%.2f')
+
+    dfmean["0.0 pct"] = dfmean[0.0]/dfmean[-1.0]
+    dfmean["1.0 pct"] = dfmean[1.0]/dfmean[-1.0]
+
+    print("Writing means (pct)...")
+    dfmean[["0.0 pct","1.0 pct"]].to_latex(
[…]

### 2017-03-01 paper/main.tex
Move planing times results to commands
paper: null made macro, since had null and NULL
paper: some minor fixes and todos added
paper: add equal contribution mark

@@
 \author{
-  Jos\'e Cambronero \\
+  Jos\'e Cambronero\thanks{Author contributed equally to this paper.} \\
   \texttt{jcamsan@csail.mit.edu}
   \and
-  John K. Feser \\
+  John K. Feser\footnotemark[1] \\
   \texttt{feser@csail.mit.edu}
   \and
-  Micah J. Smith \\
+  Micah J. Smith\footnotemark[1] \\
   \texttt{micahs@mit.edu}
@@
 \newcommand{\todo}[1]{{\color{red}\textbf{TODO:} #1}}
-\newcommand{\todobox}[2]{{\color{red}\fbox{\parbox{\columnwidth}{\textbf{TODO:} #1\\#2}}}}
+\newcommand{\todobox}[2]{{\color{red}\fbox{\parbox{\columnwidth}{\textbf{TODO:} \textit{#1}\\#2}}}}
 
@@
 \newcommand{\highxalphaone}{16000}
+\newcommand{\nullv}{NULL}
+\newcommand{\planningruntimepercent}{8.5}
+\newcommand{\planningmaxtime}{40}
 
@@
 by applying a potentially-expensive imputation strategy. Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, 
-performing imputation on the fly
+performing any operations needed on the fly
 for each query. This allows users to immediately explore

### 2017-03-01 paper/relatedwork.tex
paper: one more relevant reference
paper: minor changes in related work

@@
 \section{Related Work}
+
+There is related work in three primary areas: statistics, database research, and forecasting.
 
@@
 
-Imputation of missing values is a widely studied field within the statistics and machine
+Imputation of missing values is widely studied in the statistics and machine
 learning communities. As highlighted in~\cite{gelman2006data}, missing data
-can appear for a variety of reasons, including both random and conditioned on
+can appear for a variety of reasons.
+It can be missing uniformly at random and or conditioned on
 existing values (observed and missing). Methods in the statistical community
-focus on correctly modeling relationships between the attributes to factor in
-varied forms of missingness. For example, Burgette and Reiter~\cite{burgette2010multiple} discuss the usage of sequential regression trees
+focus on correctly modeling relationships between attributes to account for different forms of missingness.
+For example, Burgette and Reiter~\cite{burgette2010multiple} discuss the usage of sequential regression trees
 for imputing missing data.
@@
 papers have described various (at times conflicting) treatments
-of nulls~\cite{grant1977null}. \ProjectName's main design invariant - no relational operator
+of nulls~\cite{grant1977null}. \ProjectName's main design invariant---no relational operator
 sees missing values for attributes it must operate on and users should never see
-missing data - eliminates
+missing data---eliminates
 the need to handle \nullv{} value semantics, while guaranteeing query evaluation soundness.
@@
 
+\cite{fischer2013towards} describe the architecture of a DBMS with integrated forecasting operations for time-series, detailing
+the abstractions necessary to do so.
+
 In contrast to this work, \ProjectName{} is targeted at generic value imputation, not necessarily tailored to 

### 2017-03-01 paper/experiments.tex
paper: remove references to query 9

--- 2017-03-01 paper/experiments.tex
@@
 \begin{figure}
-\todo{Relabel 9 somehow, since it is no longer in table of queries but in lined later}
 \includegraphics[width=\columnwidth]{figures/running_times_combined_bar_13.png}
@@
     magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
-    imputation can provide another order-of-magnitude speedup at the expense of quality. For
-    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced
-    from~\acsbaseresulthours~to~\acsimputedbzeroresult.}
+    imputation can provide another order-of-magnitude speedup at the expense of quality.}
 
@@
 Applying the imputation operation to the base table is extremely expensive, in our case
-completing in \acsbaseresultminutes{} (\Cref{fig:runtimes}). In contrast, \ProjectName{} executes a quality-optimized version
+completing in \acsbaseresultminutes{}. In contrast, \ProjectName{} executes a quality-optimized version
 in \acsimputedbzeroresult{} and a runtime-optimized version in \acsimputedboneresult{}. This highlights the potential

### 2017-03-01 paper/algorithm.tex
paper: break up definition of Loss

@@
   \textsc{L}(q) &= \begin{cases}
-     \{1\} \cup \textsc{L}(q') & q = \delta_C(q') \\
-    \left\{\frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}}\right\} \cup \textsc{L}(q')  & q = \mu_C(q') \\
-    \textsc{L}(q_1') \cup \textsc{L}(q_2') & q = q_1' \bowtie_\psi q_2' \\
-    \textsc{L}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
-    \emptyset\ multiset & q = t\\
+     1 + \textsc{L}(q') & q = \delta_C(q') \\
+    \frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}} + \textsc{L}(q')  & q = \mu_C(q') \\
+    \textsc{L}(q_1') + \textsc{L}(q_2') & q = q_1' \bowtie_\psi q_2' \\
+    \textsc{L}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
+    0 & q = t\\
   \end{cases}\\
-  \textsc{Loss}(q) &= \frac{\Sigma \textsc{L}(q)}{|\textsc{L}(q)|}
+    \textsc{N}(q) &= \begin{cases}
+     1 + \textsc{N}(q') & q = \delta_C(q') \\
+     1 + \textsc{L}(q')  & q = \mu_C(q') \\
+    \textsc{N}(q_1') + \textsc{N}(q_2') & q = q_1' \bowtie_\psi q_2' \\
+    \textsc{N}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
+    0 & q = t\\
+  \end{cases}\\
+  \textsc{Loss}(q) &= \frac{\textsc{L}(q)}{\textsc{N}(q)}
 \end{align*}
@@
 Each imputation operator incurs a heuristic loss penalty. We define \textsc{Loss} as the average over the individual penalties incurred
-by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties in a multi-set, as the order of the penalties is irrelevant,
-and averages them at the root of the plan. For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
+by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties, and $N$ counts the number of imputation operations.
+For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
 might have encoded). Meanwhile, imputing \nullv{} fields with $\mu$ incurs a loss penalty $p \in (0, 1]$. In this case, the penalty is
@@
 correspondingly, the information lost \textit{decreases} with more values to work
-with. The square root captures the intuition that the benefits provided by more training observations has diminishing returns.
+with. We wanted to encode the fact that the number of training observations has diminishing returns for imputation quality,
+and found that scaling the cell count by applying square root worked well in practice.
 

### 2017-03-01 paper/algorithm.tex
paper: polishing changes

--- 2017-03-01 paper/algorithm.tex
@@
 replace all \nullv{} values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
-In our experiments, we instantiate \textit{Impute} with the CE-CART\cite{burgette2010multiple} algorithm, but we highlight that
+In our experiments, we instantiate \textit{Impute} with the CE-CART~\cite{burgette2010multiple} algorithm, but we highlight that
 our system is agnostic to this choice.
@@
 inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as
-as the product of the number of attributes and the number of tuples. Intuitively, imputation
+the product of the number of attributes and the number of tuples. Intuitively, imputation
 accuracy \textit{increases} when more complete attributes and complete tuples are available;
@@
 Scan costs are estimated based on the number of tuples, 
-page size, and an unit-less I/O cost per page. Join costs are estimated as a function
+page size, and a unit-less I/O cost per page. Join costs are estimated as a function
 of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
@@
 fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
-not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
+not even consider CPU cost explicitly). For example,~\cite{martin1995time} find that the
 time complexity of the \textit{build tree} algorithm for one commonly-used class of

### 2017-03-01 paper/conclusion.tex
paper: polishing changes

@@
 is unlikely to follow a standard distribution, which implies that any confidence
-estimates will result from iterative approaches such as cross-validation\cite{kohavi1995study}.
+estimates will result from iterative approaches such as cross-validation~\cite{kohavi1995study}.
 In contrast to standard confidence calculations, there is the added complexity

### 2017-03-01 paper/experiments.tex
paper: add SMAPE citation
paper: polishing changes

@@
 For all data sets, we selected a subset of the original attributes.
-We also transformed all data values to an integer representation by enumerating strings and transforming floating point values into an appropriate range.
+We also transformed all data values to an integer representation by enumerating strings and transforming floating-point values into an appropriate range.
 
@@
 \Cref{q1} calculates calculate the average cuff size for users based on their income data.
-\Cref{q2} compares creatine levels for individuals with low, medium, and high incomes spectrum and above a certain weight.
+\Cref{q2} compares creatine levels for individuals with low, medium, and high incomes and above a certain weight.
 \Cref{q3} extracts the average blood lead levels for children under the age of 6 years.
@@
 
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)\todo{cite SMAPE} for
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for
 \ProjectName{}'s query results when compared to running imputation on the base tables and
-executing the query on this cleaned base data \cite{tofallis2015better}.  Each query was
+executing the query on this cleaned base data~\cite{tofallis2015better}.  Each query was
 separately run 200 times in both settings and then results from the two approaches were

### 2017-03-01 paper/introduction.tex
paper: polishing changes

@@
 
-A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
+A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all the data.
 In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
@@
 significant speedup in query execution while minimizing the number of dropped rows.
-Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in placement of imputation operations
-in an optimal way into query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.
+Unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
+optimally in query plans.   Specifically, our optimization algorithm generates the Pareto-optimal frontier of plans that trade imputation cost for result quality.
 
@@
   We use the cost model to abstract over the imputation algorithm used.
-  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
+  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
 \item \textbf{Query planning with imputation:}

### 2017-03-01 paper/relatedwork.tex
paper: polishing changes

@@
 The computational difficulties of imputing on large base
-tables are well known and can limit approaches. For example, Akande finds that
+tables are well-known and can limit approaches. For example, Akande finds that
 one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
@@
 detect dirty values, whether missing or otherwise, and rectify the errors if
-possible. A survey of methods and systems is provided in
-\cite{hellerstein2008quantitative}.
+possible. A survey of methods and systems is provided in~\cite{hellerstein2008quantitative}.
 
-In \cite{wolf2007query}, queries over databases with missing values are
-processed in a statistical approach, taking advantage of correlations between
-variables. Both tuples that match the original query as well as a ranking of
+In~\cite{wolf2007query}, queries over databases with missing values are
+processed using a statistical approach, taking advantage of correlations between
+variables. The tuples that match the original query as well as a ranking of
 tuples with incomplete data that may match are returned. Our work differs in
@@
 and consider them carefully in query processing. Often, if sensor error is the
-cause of missing values, values can be imputed with high confidence. In
-\cite{fernandez2009inter}, feedback punctuation is used to dynamically
+cause of missing values, values can be imputed with high confidence. In~\cite{fernandez2009inter}, feedback punctuation is used to dynamically
 reconfigure the query plan for state-dependent
@@
 
-BayesDB \cite{mansinghka2015bayesdb} provides users with a simple interface to 
+BayesDB~\cite{mansinghka2015bayesdb} provides users with a simple interface to 
 leverage statistical inference techniques in a database. Non-experts
@@
 \ProjectName's cost-based query planner 
-is partially based on the seminal work developed for System R's query planning\cite{blasgen1981system}.
+is partially based on the seminal work developed for System R's query planning~\cite{blasgen1981system}.
 However, in contrast to System R, \ProjectName{} performs additional optimizations for imputing missing data and
@@
 \subsection{Forecasting and Databases}
-Parisi et al\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
+Parisi et al.~\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
 databases, along with the necessary relational algebra extensions. Their work explores the theoretical
@@
 the type of predictions returned. They highlight the use of forecasting for replacing missing values.
-In their follow on work\cite{parisi2013temporal}, they identify various equivalence and containment
+In their follow on work~\cite{parisi2013temporal}, they identify various equivalence and containment
 relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
@@
 Similarly to Fa, \ProjectName{} provides a declarative language, as
-a subset of standard SQL. \ProjectName{}, however, is not searching the space of possible
+a subset of standard SQL.\@ \ProjectName{}, however, is not searching the space of possible
 imputation models, but rather the space of query plans that incorporate imputation operators. Another major point

### 2017-03-01 paper/algorithm.tex
paper: spelling

@@
 
-In order to pick a final plan from the frontier, our model introduces a tradeoff parameter $\alpha$ which encodes an upper bound on the $\textsc{Loss}$ value that the user is willing
+In order to pick a final plan from the frontier, our model introduces a trade-off parameter $\alpha$, which is an upper bound on the $\textsc{Loss}$ value that the user is willing
 to tolerate relative to the minimum possible amongst frontier plans.
@@
 We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
-allows us to obtain the final set of plans that best tradeoff computation
+allows us to obtain the final set of plans that best traded computation
 cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.

### 2017-03-01 paper/conclusion.tex
paper: spelling

@@
 existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
-learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
-imputation quality and runtime. We showed that different values for the tradeoff parameter can yield substantially
+learning technique, as operators in the algebra and use a simple, but effective, cost model to consider trade-offs in 
+imputation quality and runtime. We showed that different values for the trade-off parameter can yield substantially
 different plans, allowing the user to express their preferences for performance on a per-query basis.

### 2017-03-01 paper/examples.tex
paper: spelling

@@
 The user can influence the final plan selected by \ProjectName{}
-through a tradeoff parameter $\alpha \in [0, 1]$, where low $\alpha$ 
+through a trade-off parameter $\alpha \in [0, 1]$, where low $\alpha$ 
 prioritizes the quality of the query results and high $\alpha$ prioritizes performance.

### 2017-03-01 paper/relatedwork.tex
paper: spelling

@@
 includes variables with potentially large domains (ten categories in their case).
-In contrast, \ProjectName{} allows users to specify a tradeoff between
+In contrast, \ProjectName{} allows users to specify a trade-off between
 imputation quality and runtime performance, allowing users to perform queries
@@
 time-series. The optimizer is not based on equivalence transformations, nor are there guarantees of equal
-results under different conditions. Instead, the optimizer allows users to pick their tradeoff between
+results under different conditions. Instead, the optimizer allows users to pick their trade-off between
 runtime cost and imputation quality. The search space considered by our optimizer is broader, not just
@@
 imputation models, but rather the space of query plans that incorporate imputation operators. Another major point
-of distinction between \ProjectName{} and Fa is that the latter doesn't consider tradeoffs between accuracy and computation time, but rather returns the most accurate forecast (with some stopping criterion).
+of distinction between \ProjectName{} and Fa is that the latter doesn't consider trade-offs between accuracy and computation time, but rather returns the most accurate forecast (with some stopping criterion).
 

### 2017-03-01 paper/algorithm.tex
paper: feedback from ben

--- 2017-03-01 paper/algorithm.tex
@@
 \begin{itemize}
-\item All filters are pushed to the leaves of the query tree, immediately after scanning a table.
+\item All selections are pushed to the leaves of the query tree, immediately after scanning a table.
 We use a pre-processing step which conjoins together filters which operate on the same table, so we can assume that each table has at most one relevant filter.
@@
 \label{sec:placement}
-Imputation operators must be placed so that no relational operator encounters a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.
+We place imputation operators so that no relational operator encounters a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.
 
@@
 \begin{definition}
-We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(p) = \textsc{Dirty}(p') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$.
+We say that a plan $p$ with cost $\langle l, t \rangle$ \emph{dominates} a plan $p'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(p) = \textsc{Dirty}(p') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$.
 We denote this as $\doms{p}{p'}$.
@@
 Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
-We say $p$ is $\alpha$-bound optimal if $\forall p'^\alpha \in P: \textsc{Time}(p) \leq \textsc{Time}(p'^\alpha)$. 
+We define the set of $\alpha$-bound plans in $P$ as $P^\alpha$.
+We say a plan $p$ is $\alpha$-bound optimal if $\forall p' \in P^\alpha. \textsc{Time}(p) \leq \textsc{Time}(p'^\alpha)$. 
 \end{definition}
 
-Given $\alpha$ and $P$, \ProjectName{} returns the $\alpha$-bound optimal plan in $P$. Note that the $\alpha$-bound optimal plan is unique in $P$, since $P$ contains only non-dominated plans.
+Given $\alpha$ and $P$, \ProjectName{} returns an $\alpha$-bound optimal plan in $P$.
 
@@
 For simplicity, each of the logical nodes in a query plan points to a set of histograms.
-When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan.
+When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan, for example
+redistributing the count of tuples with \nullv{} in an imputed field.
 \Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
@@
 Filtering, projecting, and aggregating are all done in memory and have low computational overhead
-so we assume negligible time costs for those operations. We extend these heuristics to 
+so we assume negligible time costs for those operations.
+
+We extend these heuristics to 
 account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
 complexity can be expressed as a function of the number of heap pages read and the I/O cost
-per page.
-
-The time computation for \textit{Impute} depends on the properties of the underlying
+per page. The time computation for \textit{Impute} depends on the properties of the underlying
 algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
@@
 parameterization of the partition and heuristic functions, and the \textit{arity} (the
-number of subsets considered for each split). Indeed, the \textit{build tree} phase often
+number of subsets considered for each split). The \textit{build tree} phase often
 does not even dominate computation, as post-processing steps like pruning, which are vital
@@
   \end{align*}
-  \caption{Semantics of a plan cache $Q$}
+  \caption{Semantics of a plan cache $Q$, where $T$ is a set of tables and $p$ is a query plan}
   \label{fig:semantics-plan-cache}
@@
 
-Our imputation algorithm proceeds by iteratively fitting regression trees to a subset of the
+The algorithm used by the \textit{Impute} operator proceeds by iteratively fitting regression trees to a subset of the
 data and the target imputation columns. In each iteration, the missing values of a column are

### 2017-03-01 paper/conclusion.tex
paper: feedback from ben

--- 2017-02-28 paper/conclusion.tex
@@
 Our experiments with \textit{CDC}, \textit{freeCodeCamp} and \textit{ACS} survey-based datasets
-showed that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
+show that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
 We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
@@
 we incorporate imputation into a cost-based optimizer and hide any details
-regarding missing values inside the system, allowing users to use traditional SQL and engage in normal workloads.
+regarding missing values inside the system, allowing users to use traditional SQL and engage in normal workflows.
 While prior work has incorporated operations such as prediction into their databases, this has been domain-specific
@@
 another possible improvement to the system. In most cases, the data imputed
-is unlikely to follow a standard distribution, which implies that any confidence
-estimates will result from iterative approaches such as cross-validation\cite{kohavi1995study}.
+is unlikely to follow a standard distribution, but iterative approaches such as cross-validation\cite{kohavi1995study}
+can give confidence intervals.
 In contrast to standard confidence calculations, there is the added complexity
-of imputation taking place at various points in the plan and the composition of results
-through standard query operations.
+of imputation taking place at various points in the plan, so the composition of results
+through standard query operations needs to be taken into account when computing confidence intervals.
 

### 2017-03-01 paper/examples.tex
paper: feedback from ben

--- 2017-02-28 paper/examples.tex
@@
 Meanwhile, \Cref{fig:fast-plan}'s performance-optimized plan uses the drop operation $\delta$ instead of filling in missing values.
-However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.
+However, it is a significant improvement over dropping all tuples with missing data in any field, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.
 In both cases, only performing

### 2017-03-01 paper/experiments.tex
paper: feedback from ben

--- 2017-03-01 paper/experiments.tex
@@
 \Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
-configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0$),
-\ProjectName{} optimizing for performance ($\alpha=1$), and imputing at the base tables followed
+configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0.0$),
+\ProjectName{} optimizing for performance ($\alpha=1.0$), and imputing at the base tables followed
 by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
@@
 \Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)\todo{cite SMAPE} for
-\ProjectName{}'s query results when compared to running imputation on the base tables and
-executing the query on this cleaned base data \cite{tofallis2015better}.  Each query was
-separately run 200 times in both settings and then results from the two approaches were
+\ProjectName{}'s query results when compared to the results of running imputation on the base tables and
+executing the query on this cleaned base data \cite{tofallis2015better}. Given that the main datasets used
+have no ground truth, this measure indicates how much error is introduced by on-the-fly imputation compared
+to imputation on the base tables.
+
+Each query was separately run 200 times in both settings and then results from the two approaches were
 paired up and compared tuple-wise.  We average tuple-wise absolute percentage deviations
@@
 can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
-to the base-imputation approach are low in all cases --- between
+to the base-imputation approach is low in all cases --- between
 \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly
@@
 \caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
-    parameterizations relative to results when imputing on the base table. Queries optimized
+    parameterizations. To calculate SMAPE, \ProjectName{} results are compared to query results returned
+    from executions with base table imputation. Queries optimized
     for quality ($\alpha=0$) generally achieve lower error than queries optimized for
@@
 \end{lstlisting}
-Applying the imputation operation to the base table is extremely expensive, in our case
+Applying the imputation operation to the base table is extremely expensive, as it imputes over all attributes in all rows, in our case
 completing in \acsbaseresultminutes{}. In contrast, \ProjectName{} executes a quality-optimized version
 in \acsimputedbzeroresult{} and a runtime-optimized version in \acsimputedboneresult{}. This highlights the potential
-benefit of using our system for early data exploration.
+benefit of using our system for early data exploration. Note that the performance differential relative to base table imputation
+would likely widen further if selection predicates were added to the query.
 

### 2017-03-01 paper/introduction.tex
paper: feedback from ben

--- 2017-03-01 paper/introduction.tex
@@
 
-A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
+A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
 In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
-carefully plans the placement  imputation / row drop operations inside the query plan, resulting in a 
+carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
 significant speedup in query execution while minimizing the number of dropped rows.
-Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in placement of imputation operations
-in an optimal way into query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.
+Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in optimal placement of imputation operations
+in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.
 
@@
 % Scope
-\ProjectName{} is designed to enable early data exploration, by allowing analysts to run their queries without explicitly decide with when to perform imputation.
+\ProjectName{} is designed to enable early data exploration, by allowing analysts to run their queries without explicitly deciding when to perform imputation.
 % SAM -- cut the following -- could consider moving later but feels out of place here
@@
   We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
-  Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
-  Drop operation simply drops tuples which contain \nullv{} values.
+  The impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
+  The drop operation simply drops tuples which contain \nullv{} values.
 \item \textbf{Model of imputation quality and cost:}

### 2017-03-01 paper/relatedwork.tex
paper: feedback from ben

--- 2017-03-01 paper/relatedwork.tex
@@
 can appear for a variety of reasons.
-It can be missing uniformly at random and or conditioned on
+It can be missing uniformly at random or conditioned on
 existing values (observed and missing). Methods in the statistical community
@@
 
-In~\cite{akande2015empirical}, Akande et al analyze the performance of various
+~\cite{akande2015empirical} analyzes the performance of various
 multiple imputation techniques on the American Community Survey dataset. 
@@
 However, in contrast to System R, \ProjectName{} performs additional optimizations for imputing missing data and
-uses histogram transformations to account for the changing nature of missing values.
+uses histogram transformations to account for the way relational and imputation operators affect the underlying
+tuple distributions.
 

### 2017-03-01 paper/experiments.tex

@@
 
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)\todo{cite SMAPE} for
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for
 \ProjectName{}'s query results when compared to the results of running imputation on the base tables and

### 2017-03-01 paper/experiments.tex
Update query numbers

--- 2017-03-01 paper/experiments.tex
@@
 For all data sets, we selected a subset of the original attributes.
-We also transformed all data values to an integer representation by enumerating strings and transforming floating-point values into an appropriate range.
+We also transformed all data values to an integer representation by enumerating strings and transforming floating point values into an appropriate range.
 
@@
 \Cref{q2} compares creatine levels for individuals with low, medium, and high incomes and above a certain weight.
-\Cref{q3} extracts the average blood lead levels for children under the age of 6 years.
+\Cref{q3} extracts the average blood lead levels for children under 6 years of age.
 \Cref{q4} calculates the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity. 
+\Cref{q5} calculates the average waist circumference for subjects above a certain height and
+weight.
 
 Our second set of queries is on the freeCodeCamp data (\Cref{fig:queries-fcc}).
-\Cref{q5} calculates the average income for survey participants, based on their bootcamp attendance.
-\Cref{q6} estimates the average commute time of women from the United States who participated.
-\Cref{q7} calculates the average amount of student debt based on school degree for survey participants who have student debt.
-\Cref{q8} joins the freeCodeCamp data with a reference table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
-The query calculates the average GDP per-capita of countries with and without bootcamp participants.
+\Cref{q6} calculates the average income for higher-income survey participants, grouped by their bootcamp attendance.
+\Cref{q7} estimates the average commute time of women from the United States who participated.
+\Cref{q8} calculates the average amount of student debt based on school degree for survey participants who have student debt.
+\Cref{q9} joins the freeCodeCamp data with a reference table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
+The query calculates the average GDP per-capita of countries with and without bootcamp
+participants over 18 years of age.
 
 \begin{table*}
-\todo{update this with final queries}
   \centering
@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{figures/running_times_combined_bar_13.png}
+\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
 \caption{Query runtimes with imputation on base tables, \ProjectName{}
@@
 \ProjectName{}'s query results when compared to running imputation on the base tables and
-executing the query on this cleaned base data~\cite{tofallis2015better}.  Each query was
+executing the query on this cleaned base data \cite{tofallis2015better}.  Each query was
 separately run 200 times in both settings and then results from the two approaches were
@@
 the base tables have been imputed. In many cases where the SMAPE reduction between $\alpha = 1$ and
-$\alpha = 0$ is small, for example \Cref{q6}, we highlight that the count of tuples can be significantly different between
+$\alpha = 0$ is small, for example \Cref{q6}\todo{confirm}, we highlight that the count of tuples can be significantly different between
 the two scenarios. So while the aggregate used might not have been significantly affected by the different population,
@@
 \centering
-\input{tables/perf_summary_13.tex}
+\input{tables/perf_summary.tex}
 \caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$

### 2017-03-01 paper/conclusion.tex
paper: cleanup and remove todo in conclusion

--- 2017-03-01 paper/conclusion.tex
@@
 existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
-learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
-imputation quality and runtime. We showed that different values for the tradeoff parameter can yield substantially
+learning technique, as operators in the algebra and use a simple, but effective cost model to consider trade-offs in 
+imputation quality and runtime. We show that different values for the trade-off parameter can yield substantially
 different plans, allowing the user to express their preferences for performance on a per-query basis.
-Our experiments with \textit{CDC}, \textit{freeCodeCamp} and \textit{ACS} survey-based datasets
-show that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
+Our experiments with the \textit{CDC NHANES}, \textit{freeCodeCamp} and \textit{ACS} datasets
+show that \ProjectName{} can be used successfully with real-world data, to run standard SQL queries over data with missing values. 
 We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
@@
 then formulating queries. Furthermore, the difference in query results between the two approaches is
-shown to be small in all queries considered\todo{quantify this}. Data analysts need not commit to a specific imputation strategy and can instead
+shown to be small in all queries considered (0-\highsmapealphaoneexacs\% error, \Cref{sec:results}). Data analysts need not commit to a specific imputation strategy and can instead
 vary this across queries.
 
-We highlight the long history of dealing with missing data both in the statistical learning and database communities.
+We discuss the long history of dealing with missing data both in the statistical learning and database communities.
 Similarly to existing work, we consider the impact of null values in databases and develop a simple set of invariants to 
@@
 we incorporate imputation into a cost-based optimizer and hide any details
-regarding missing values inside the system, allowing users to use traditional SQL and engage in normal workflows.
+regarding missing values inside the system, allowing users to use traditional SQL as if the database were complete.
 While prior work has incorporated operations such as prediction into their databases, this has been domain-specific

### 2017-03-01 paper/introduction.tex

@@
 significant speedup in query execution while minimizing the number of dropped rows.
-Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in optimal placement of imputation operations
-in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.
+Unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
+optimally in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.
 

### 2017-03-01 paper/experiments.tex
paper: add implementation section

--- 2017-03-01 paper/experiments.tex
@@
 \section{Experiments}\label{sec:experiments}
-To evaluate the performance of our approach,  we implemented
-a prototype system in Java, following a traditional iterator model.
-For our experiments we plan and execute queries
-for three separate survey-based data sets, showing that our system
-is well suited for early dataset exploration.
+For our experiments we plan and execute queries for three separate survey-based data sets, showing that our system is well suited for early dataset exploration.
+We show that \ProjectName{} performs an order of magnitude better than imputation on the full dataset and produces results of comparable quality.
+
+\subsection{Implementation}
+To evaluate the performance of our approach, we implemented a prototype system in Java, following a traditional iterator model.
+Our database handles a subset of standard SQL queries, including filters, joins, projection, grouping and aggregation.
+For simplicity, \ProjectName{} only handles tables containing integer data, so we pre-process our experimental datasets before loading them.
 

### 2017-03-01 paper/experiments.tex
paper: minor formatting & reference fix

@@
 Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}).
-\Cref{q1} calculates calculate the average cuff size for users based on their income data.
+\Cref{q1} calculates calculate the average cuff size for individuals based on their income data, with a constraint on height.
 \Cref{q2} compares creatine levels for individuals with low, medium, and high incomes and above a certain weight.
@@
 \Cref{q4} calculates the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity. 
-\Cref{q5} calculates the average waist circumference for subjects above a certain height and
-weight.
+\Cref{q5} calculates the average waist circumference for subjects above a certain height and weight.
 
@@
 \Cref{q9} joins the freeCodeCamp data with a reference table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
-The query calculates the average GDP per-capita of countries with and without bootcamp
-participants over 18 years of age.
+The query calculates the average GDP per-capita of countries with and without bootcamp participants over 18 years of age.
 
 \begin{table*}
+  \todo{query parameter scaling}
   \centering
@@
 can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
-to the base-imputation approach is low in all cases --- between
-\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly
+to the base-imputation approach is low in all cases---between
+\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent---indicating that on-the-fly
 imputation produces similar results to imputation at the base tables.

### 2017-03-01 paper/experiments.tex
paper: fix query value ranges
paper: pass over section 4.4.1

@@
 \begin{table*}
-  \todo{query parameter scaling}
   \centering
@@
 
-We also calculate SMAPE over the number of tuples used to produce the results
-in each configuration with respect to the number of tuples satisfying the same query when
-the base tables have been imputed. In many cases where the SMAPE reduction between $\alpha = 1$ and
-$\alpha = 0$ is small, for example \Cref{q6}\todo{confirm}, we highlight that the count of tuples can be significantly different between
-the two scenarios. So while the aggregate used might not have been significantly affected by the different population,
-other aggregates may well be. This highlights a challenge for a user handling data imputation traditionally: it is unclear if
-the missing data will have a large or small negative impact on their analysis until they have paid the cost of
-running it. By using \ProjectName{} this cost can be lowered significantly.
+We also calculate the number of tuples used to produce each aggregate output.
+The Count columns in \Cref{table:smape} show the percentage difference between the number of tuples in the aggregate for $\alpha = 1$ and $\alpha = 0$ and running the query on the imputed base table.
+This shows that when optimizing for performance, not quality, many tuples are dropped.
+Even in cases where the SMAPE reduction between $\alpha = 1$ and $\alpha = 0$ is small (\Cref{q2}, \Cref{q6}) the tuple count is significantly different.
+In these cases, the aggregate value is not significantly impacted by the missing data.
+However, if the missing data is biased then the aggregate will have a significant error.
+This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.
+By using \ProjectName{} this cost can be lowered significantly.
 

### 2017-03-01 paper/introduction.tex
paper: fix intro

@@
 Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
-queries that execute between \lowxalphazero{} and \highxalphaoneexacs{} times as fast as it takes to impute all missing values and then run queries (the traditional approach).
+queries that execute between \lowxalphazero{} and \highxalphazeroexacs{} times as fast as it takes to impute all missing values and then run queries (the traditional approach).
 Furthermore, the results obtained with this approach

### 2017-03-01 paper/introduction.tex
Remove SMAPE before it is explained

@@
 Furthermore, the results obtained with this approach
-are similar in quality to those obtained with the complete imputation approach, as indicated by low Symmetric-Mean-Absolute-Percentage-Errors --- between \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent --- in our empirical results
-over real-world data sets (see~\Cref{sec:experiments} for details).
+are similar in quality to those obtained with the complete imputation approach, as indicated
+by one measure of error --- between \lowsmapealphazero{} and \highsmapealphaoneexacs{}
+percent --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
+for details).
 

### 2017-03-01 paper/algorithm.tex

@@
 replace all \nullv{} values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
-In our experiments, we instantiate \textit{Impute} with the CE-CART~\cite{burgette2010multiple} algorithm, but we highlight that
+In our experiments, we implement \textit{Impute} using a chained-equation decision trees
+algorithm, but we highlight that
 our system is agnostic to this choice.
@@
 We define the set of $\alpha$-bound plans in $P$ as $P^\alpha$.
-We say a plan $p$ is $\alpha$-bound optimal if $\forall p' \in P^\alpha. \textsc{Time}(p) \leq \textsc{Time}(p'^\alpha)$. 
+We say a plan $p^\alpha$ is $\alpha$-bound optimal if $\forall p' \in P^\alpha. \textsc{Time}(p) \leq \textsc{Time}(p')$. 
 \end{definition}
@@
 a predictive model of one missing attribute, conditional on all complete attributes
-and all other missing attributes. Chained-equation decision tree algorithms are found to be
+and all other missing attributes. Chained-equation decision trees algorithms are found to be
 effective~\cite{akande2015empirical} and are widely used in epidemiological

### 2017-03-01 paper/algorithm.tex
paper: phillip changes to algorithm section

--- 2017-03-01 paper/algorithm.tex
@@
 agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
-that while our algorithm is exponential in the number of joins(\Cref{sec:complexity}), in practice, planning times are not a concern.
+that while our algorithm is exponential in the number of joins (\Cref{sec:complexity}), in practice, planning times are not a concern.
 
@@
 attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
-replace all \nullv{} values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
+replace all \nullv{} values with non-\nullv{} values for attributes $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
 In our experiments, we instantiate \textit{Impute} with the CE-CART~\cite{burgette2010multiple} algorithm, but we highlight that
@@
 \item Joins are performed after filtering, and only left-deep plans are considered.
+  This significantly reduces the plan space while still allowing plans with interesting join patterns.
 
@@
 \label{sec:placement}
-We place imputation operators so that no relational operator encounters a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.
+We place imputation operators into the query plan so that no relational operator encounters a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.
 
-Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a \nullv{} value, there are cases where an imputation operator is required.
-To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$.
+Imputation operators can be placed at any point, but there are cases where an imputation operator is required to meet the guarantee that no non-imputation operator sees a \nullv{} value.
+To track these cases, we associate each query plan $q$ with a set of dirty attributes $\textsc{Dirty}(q)$.
 An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain \nullv{}.
@@
 
+\subsubsection{Pareto-optimal plans}
 \begin{definition}
@@
 
+\subsubsection{Cardinality estimation}
 The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
@@
 For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
-For simplicity, each of the logical nodes in a query plan points to a set of histograms.
-When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan, for example
-redistributing the count of tuples with \nullv{} in an imputed field.
+
+To compute cardinality, we maintain histograms for each column in the database.
+During query planning, each of the logical nodes in a query plan points to a set of histograms which describe the distribution of values in the output of the node.
+When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan.
+For example it will redistribute the count of tuples with \nullv{} in an imputed field.
 \Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
 
+\subsubsection{Imputation quality}
 We use these histogram estimates to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.
@@
 
-Since we conceive that
-alternate imputation strategies could be used, the analyst may provide a corresponding
-loss with a reasonable parameterization.
+To add an alternate imputation strategy to \ProjectName{} could be used, the analyst must provide a corresponding loss with a reasonable parameterization.
 
+\subsubsection{Query runtime}
 To compute $\textsc{Time}(Q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard 
@@
 effort and without changes to the optimizer, simply by adding time and loss functions
-(\Cref{sec:cost-model}) which describe the new algorithm. In principle, new imputation algorithms
+(\Cref{sec:cost-model}) which describe the new algorithm.
+An imputation strategy is a function $I(T, C) \rightarrow T'$ which fills in the missing attributes $C$ in a set of tuples $T$, returning a new set of tuples $T'$.
+In principle, new imputation algorithms
 can be added, possibly targeted to a specific domain. The flexibility of this approach

### 2017-03-01 paper/algorithm.tex
Small changes I had noted

--- 2017-03-01 paper/algorithm.tex
@@
 As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
-based on chained-equation decision trees.  Chained equation
+based on chained-equation decision trees.  Chained-equation
 imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
@@
 (\cite{gelman2006data}). In \ProjectName{}, we consider only single imputation, though in
-further work, multiple imputation could be considered as another avenue of assessing
+further work, multiple imputation could be another avenue of assessing
 imputation confidence.

### 2017-03-01 paper/experiments.tex
Small changes I had noted

@@
 
+\subsection{Results}\label{sec:results}
+
 All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
@@
 
-\subsection{Results}\label{sec:results}
+Each query was separately run 200 times in both settings and then
 
 \Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
-configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0.0$),
-\ProjectName{} optimizing for performance ($\alpha=1.0$), and imputing at the base tables followed
+configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0$),
+\ProjectName{} optimizing for performance ($\alpha=1$), and imputing at the base tables followed
 by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
@@
 
-Each query was separately run 200 times in both settings and then results from the two approaches were
-paired up and compared tuple-wise.  We average tuple-wise absolute percentage deviations
-within each iteration of a query, and we report this value averaged over all iterations.  We
+For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
+we compute tuple-wise absolute percentage deviations within each iteration of a query, and
+we report this value averaged over all iterations.  We
 can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
-to the base-imputation approach is low in all cases---between
+to the base-imputation approach is low---between
 \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent---indicating that on-the-fly
@@
 We also calculate the number of tuples used to produce each aggregate output.
-The Count columns in \Cref{table:smape} show the percentage difference between the number of tuples in the aggregate for $\alpha = 1$ and $\alpha = 0$ and running the query on the imputed base table.
+The Count columns in \Cref{table:smape} show the percentage difference between the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ and running the query on the imputed base table.
 This shows that when optimizing for performance, not quality, many tuples are dropped.
 Even in cases where the SMAPE reduction between $\alpha = 1$ and $\alpha = 0$ is small (\Cref{q2}, \Cref{q6}) the tuple count is significantly different.
-In these cases, the aggregate value is not significantly impacted by the missing data.
+In these cases, the aggregate value is not significantly impacted by the missing data. In particular, if values are missing uniformly at random,
+the aggregate should not be affected.
 However, if the missing data is biased then the aggregate will have a significant error.

### 2017-03-01 paper/algorithm.tex

--- 2017-03-01 paper/algorithm.tex
@@
 replace all \nullv{} values with non-\nullv{} values for attributes $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
-In our experiments, we instantiate \textit{Impute} with the CE-CART~\cite{burgette2010multiple} algorithm, but we highlight that
-our system is agnostic to this choice.
+In our experiments, we implement \textit{Impute} using a chained-equation decision trees algorithm, but our system is agnostic to this choice.
 \textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.

### 2017-03-01 paper/main.tex
paper: cleanup and remove todo in conclusion
paper: balance last page
paper: minor formatting & reference fix

@@
 \renewcommand{\algorithmicensure}{\textbf{Output:}}
+
+\hyphenation{Im-pute-DB}
 
@@
 \begin{abstract}
-Missing values abound in data analysis and often present a serious challenge for usability of data. Users
+Missing values abound in data analysis and often present a serious challenge for the usability of data. Users
 are forced to pick between removing tuples with missing values or creating a clean duplicate of their data
@@
 that the query results from on-the-fly imputation have
-Symmetric-Mean-Absolute-Percentage-Error between \lowsmapealphazero{} and
+error between \lowsmapealphazero{} and
 \highsmapealphaoneexacs{} percent when compared to standard full-table imputation
@@
 \bibstyle{abbrv}
+\balance
 \printbibliography
-\balance
 

### 2017-03-01 paper/algorithm.tex
paper: system R -> canonical system, in most mentions

--- 2017-03-01 paper/algorithm.tex
@@
 
-The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
+The space of query plans is similar to that considered in a canonical
+cost-based optimizer~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
 \Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.
@@
 As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
-based on chained-equation decision trees.  Chained-equation
+based on chained-equation decision trees.  Chained equation
 imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
@@
 (\cite{gelman2006data}). In \ProjectName{}, we consider only single imputation, though in
-further work, multiple imputation could be another avenue of assessing
+further work, multiple imputation could be considered as another avenue of assessing
 imputation confidence.
@@
 \subsection{Complexity}\label{sec:complexity}
-Our optimization algorithm builds off the approach taken by System R~\cite{blasgen1981system}.
+Our optimization algorithm builds off the approach taken by 
+a canonical cost-base optimizer~\cite{blasgen1981system}.
 If imputation operators are ignored, we search the same plan space.

### 2017-03-01 paper/algorithm.tex
paper: widows & orphans

@@
+\vspace{0.5in}
 \section{Algorithm}
@@
 
-The space of query plans is similar to that considered in a canonical
-cost-based optimizer~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
+The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
 \Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.
@@
 \subsection{Complexity}\label{sec:complexity}
-Our optimization algorithm builds off the approach taken by 
-a canonical cost-base optimizer~\cite{blasgen1981system}.
+Our optimization algorithm builds off the approach taken by System R~\cite{blasgen1981system}.
 If imputation operators are ignored, we search the same plan space.

### 2017-03-01 paper/experiments.tex
Update results subsec intro language
paper: widows & orphans
Fix remaining small items

@@
 
-All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
-four 2.9 GHz Intel Xeon E5--2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.
-
-Each query was separately run 200 times in both settings and then
-
 \Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
@@
 \ProjectName{}, in contrast to the traditional approach of base table imputation.
+
+All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
+four 2.9 GHz Intel Xeon E5--2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.
+Each query with each value of $\alpha$ was run 200 times, after discarding a warm-up period.
 
@@
 \ProjectName{}'s query results when compared to the results of running imputation on the base tables and
-executing the query on this cleaned base data \cite{tofallis2015better}. Given that the main datasets used
+executing the query on this cleaned base data. Given that the main datasets used
 have no ground truth, this measure indicates how much error is introduced by on-the-fly imputation compared

### 2017-03-01 paper/relatedwork.tex
paper: more widows & orphans
Fix remaining small items

@@
 \subsection{Forecasting and Databases}
-Parisi et al.~\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
+\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
 databases, along with the necessary relational algebra extensions. Their work explores the theoretical

### 2017-03-01 paper/algorithm.tex
paper: break planning algo section into subsections

@@
 
+\subsubsection{Plan cache}
 Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
@@
 
-\Cref{algo:plan-helpers} presents a series of helper functions used by the top-level planner, shown in~\Cref{algo:top-level-planner}.
-
-The planner operates as follows.
+\subsubsection{Planning algorithm}
+The planner (\Cref{algo:top-level-planner}, with helper functions in \Cref{algo:plan-helpers}) operates as follows.
 First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}--\Cref{lst:line:attr-end}).

### 2017-03-01 paper/conclusion.tex

@@
 
-\textbf{Imputation confidence:} Obtaining a confidence measure for query results
+\textbf{Imputation confidence:} Obtaining confidence measures for query results
 produced when missing values are imputed with a machine learning model is
-another possible improvement to the system. In most cases, the data imputed
-is unlikely to follow a standard distribution, but iterative approaches such as cross-validation\cite{kohavi1995study}
-can give confidence intervals.
-In contrast to standard confidence calculations, there is the added complexity
-of imputation taking place at various points in the plan, so the composition of results
+another possible improvement to the system. These measures could be obtained in general
+using resampling methods like cross-validation and bootstrapping\cite{kohavi1995study}.
+In contrast to standard confidence calculations, there is the added complexity that
+imputation takes place at various points in the plan, so the composition of results
 through standard query operations needs to be taken into account when computing confidence intervals.
+Finally, multiple imputation could be considered explicitly and confidence measures like variances could
+similarly propagate through the query plan.
 

### 2017-03-01 paper/algorithm.tex
Incorporate feedback on Imputation Strategies section

--- 2017-03-01 paper/algorithm.tex
@@
 
-\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal
-effort and without changes to the optimizer, simply by adding time and loss functions
-(\Cref{sec:cost-model}) which describe the new algorithm.
-An imputation strategy is a function $I(T, C) \rightarrow T'$ which fills in the missing attributes $C$ in a set of tuples $T$, returning a new set of tuples $T'$.
+\ProjectName{} is designed so that any imputation strategy, typically a predictive machine
+learning model,
+can be plugged in with minimal effort and without changes to the optimizer, simply by adding
+time and loss functions (\Cref{sec:cost-model}) which describe the new algorithm.
+An imputation strategy is a function $I(T, C) \rightarrow T'$ which fills in the missing
+attributes $C$ in a set of tuples $T$, returning a new set of tuples $T'$.
 In principle, new imputation algorithms
@@
 domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
-implementation (\cite{witten2016data}).
+implementation~\cite{witten2016data}.
 
-The algorithm used by the \textit{Impute} operator proceeds by iteratively fitting regression trees to a subset of the
-data and the target imputation columns. In each iteration, the missing values of a column are
-replaced with newly imputed values. With each {\it epoch}, the quality of the imputation
-improves as values progressively reflect more accurate relationships amongst attributes. The
-algorithm terminates when convergence is achieved (i.e.\ the imputed values do not change
-across epochs) or a fixed number of epochs is reached.
+The reference algorithm used by the \textit{Impute} operator proceeds by iteratively fitting
+regression trees to a subset of the data and the target imputation attribute. In each
+iteration, the missing values of a single attribute are replaced with newly imputed values.
+In one {\it epoch} of the algorithm, each of the target imputation
+attributes is filled in with imputed values in a single iteration.
+With each {\it epoch}, the quality of the imputation improves as values progressively
+reflect more accurate relationships between attributes. The algorithm terminates when
+convergence is achieved (i.e.\ the imputed values do not change across epochs) or a fixed
+number of epochs is reached.
 
@@
 process, in which multiple distinct copies of the complete data are generated and estimators
-are computed by averaging over the multiple datasets to form a single completed dataset
-(\cite{gelman2006data}). In \ProjectName{}, we consider only single imputation, though in
+are computed by averaging over the multiple datasets to form a single completed dataset~
+\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though in
 further work, multiple imputation could be another avenue of assessing
 imputation confidence.
-
-Given the nature of the imputation strategy used here, {\it Impute} operators are blocking
-in our system. An extension of \ProjectName{} could consider regression algorithms that learn in
-an online nature; this would reduce the cost of the imputation at the possible expense of
-quality.
 

### 2017-03-01 paper/experiments.tex
paper: removed table names from missing val tables
Incorporate feedback on Imputation Strategies section

@@
 \Cref{table:nhanes-description} shows the attributes selected, along with the percentage of \nullv{} values for each attribute.
-For readability, we have replaced the NHANES variable names with self-explanatory attribute names.
+For readability, we have replaced the NHANES attribute names with self-explanatory attribute names.
 
@@
     \input{tables/cdc_demo}
-    \caption{Demographics. \demorows{} rows.}
+    \caption{Demographics (\texttt{demo}). \demorows{} rows.}
   \end{subtable}
@@
     \input{tables/cdc_labs}
-    \caption{Laboratory Results. \labexrows{} rows.}
+    \caption{Laboratory Results (\texttt{labs}). \labexrows{} rows.}
   \end{subtable}
@@
     \input{tables/cdc_exams}
-    \caption{Physical Results. \labexrows{} rows.}
+    \caption{Physical Results (\texttt{exams}). \labexrows{} rows.}
   \end{subtable}
@@
   \input{tables/fcc}
-  \caption{Missing value distribution for each attribute in freeCodeCamp Survey Data}\label{table:fcc-description} 
+  \caption{Missing value distribution for each attribute in freeCodeCamp Survey Data (\texttt{fcc}}\label{table:fcc-description} 
 \end{table}
@@
 %\end{figure}
-
+\vspace{0.5in}
 \subsubsection{Accuracy vs. Base-table Imputation}

### 2017-03-01 paper/main.tex

@@
 approach. 
-
+We calculate the number of tuples used to produce each aggregate
+in our experiments, and find that when optimizing for quality between 0 to 10\% fewer tuples satisfy query constraints relative to the full-table imputation.
+When all tuples with missing values are dropped, the aggregate
+is computed with 3 to 88\% fewer tuples.
 \end{abstract}

### 2017-03-01 paper/main.tex
Update plan time results
paper: add affiliations
paper: revise abstract

--- 2017-03-01 paper/main.tex
@@
   Jos\'e Cambronero\thanks{Author contributed equally to this paper.} \\
+  MIT CSAIL \\
   \texttt{jcamsan@csail.mit.edu}
@@
   John K. Feser\footnotemark[1] \\
+  MIT CSAIL \\
   \texttt{feser@csail.mit.edu}
@@
   Micah J. Smith\footnotemark[1] \\
+  MIT LIDS \\
   \texttt{micahs@mit.edu}
@@
   Samuel Madden \\
+  MIT CSAIL \\
   \texttt{madden@csail.mit.edu}}
@@
 \begin{abstract}
-Missing values abound in data analysis and often present a serious challenge for the usability of data. Users
-are forced to pick between removing tuples with missing values or creating a clean duplicate of their data
-by applying a potentially-expensive imputation strategy. Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, 
-performing any operations needed on the fly
-for each query. This allows users to immediately explore
-their data, while the system picks the optimal placement of imputation operations. This placement is based
-on the specific query issued and a user-provided parameter that trades off data
-quality and runtime performance. We incorporate the imputation operators into the relational algebra, allowing
-for traditional optimizations to take place. Our experiments evaluate this approach on two real-world survey-based datasets
-from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}. Our
-results show that the optimizer generates query plans that execute between \lowxalphazero{}
-and \highxalphazeroexacs{} times as fast as imputing at the base tables and then executing
-queries, making exploratory data analysis using our system feasible.  Furthermore, we show
-that the query results from on-the-fly imputation have
-error between \lowsmapealphazero{} and
-\highsmapealphaoneexacs{} percent when compared to standard full-table imputation
-approach.
-
+  Missing values abound in data analysis and present a usability challenge.
+  Users are forced to pick between removing tuples with missing values or creating a clean duplicate of their data by applying a relatively expensive imputation strategy.
+  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on the fly for each query.
+  This allows users to immediately explore their data, while the system picks the optimal placement of imputation operations.
+  % This placement is based on the specific query issued and a user-provided parameter that trades off data quality and runtime performance.
+  % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
+  We  evaluate this approach on two real-world survey-based datasets and 
+  % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
+  our experiments show that our query plans are between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
+  Furthermore, we show that the query results from on-the-fly imputation differ in one error measure from the standard full-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%.
+  Finally, we show that while dropping tuples with missing values that fail query constraints discards 3--88\% of the data, on-the-fly imputation loses only 0--10\%.
 \end{abstract}

### 2017-03-01 paper/relatedwork.tex

@@
 
+\vspace{0.5in}
 \subsection{Missing Values and Statistics}
@@
 processed using a statistical approach, taking advantage of correlations between
-variables. The tuples that match the original query as well as a ranking of
+attributes. The tuples that match the original query as well as a ranking of
 tuples with incomplete data that may match are returned. Our work differs in

### 2017-03-01 paper/experiments.tex
Couple edits to results writeup

@@
 
-In the median query across a variety of queries and choices of $\alpha$, planning
-constituted \planningruntimepercent{} percent of total runtime. In all cases, the optimizer
-returned a query plan within \planningmaxtime{} ms, with times roughly constant between levels of $\alpha$.
+In the median query across a variety of queries and choices of $\alpha$, planning took
+\planningmediantime{} ms, 
+constituting \planningruntimepercent{} percent of total runtime. In all cases, the optimizer
+returned a query plan within \planningmaxtime{} ms, with times similar across values of $\alpha$.
 

### 2017-03-01 paper/algorithm.tex
Change regression trees -> decision trees

@@
 The reference algorithm used by the \textit{Impute} operator proceeds by iteratively fitting
-regression trees to a subset of the data and the target imputation attribute. In each
+decision trees to a subset of the data and the target imputation attribute. In each
 iteration, the missing values of a single attribute are replaced with newly imputed values.

### 2017-03-01 paper/introduction.tex
Remove remaining CART references
Change regression trees -> decision trees

@@
 % Background
-While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression tree algorithm takes \acsbaseresulthours{} to train and run
+While traditional imputation methods work over the entire data set and replace all missing
+values, running a sophisticated imputation algorithm over a large data set can be very
+expensive---our experiments show that even a relatively simple decision tree algorithm takes \acsbaseresulthours{} to train and run
 on a 600K row database.
@@
   We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
-  The impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
+  The impute operation fills in missing data values using any statistical imputation
+  technique, such as chained-equation decision trees ~\cite{burgette2010multiple}.
   The drop operation simply drops tuples which contain \nullv{} values.

### 2017-03-01 paper/algorithm.tex
paper: widows and orphans
paper: some edits

--- 2017-03-01 paper/algorithm.tex
@@
 \subsubsection{Cardinality estimation}
-The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
+The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ rely on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
 These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
@@
     \frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}} + \textsc{L}(q')  & q = \mu_C(q') \\
-    \textsc{L}(q_1') + \textsc{L}(q_2') & q = q_1' \bowtie_\psi q_2' \\
+    \textsc{L}(q_l') + \textsc{L}(q_r') & q = q_l' \bowtie_\psi q_r' \\
     \textsc{L}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
-    0 & q = t\\
+    0 & q = \text{some table } t\\
   \end{cases}\\
@@
      1 + \textsc{N}(q') & q = \delta_C(q') \\
-     1 + \textsc{L}(q')  & q = \mu_C(q') \\
-    \textsc{N}(q_1') + \textsc{N}(q_2') & q = q_1' \bowtie_\psi q_2' \\
+     1 + \textsc{N}(q')  & q = \mu_C(q') \\
+    \textsc{N}(q_l') + \textsc{N}(q_r') & q = q_l' \bowtie_\psi q_r' \\
     \textsc{N}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
-    0 & q = t\\
+    0 & q = \text{some table } t\\
   \end{cases}\\
@@
 
-To add an alternate imputation strategy to \ProjectName{} could be used, the analyst must provide a corresponding loss with a reasonable parameterization.
+To add an alternate imputation strategy to \ProjectName{}, the analyst must provide a corresponding loss with a reasonable parameterization.
 
@@
 We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
-allows us to obtain the final set of plans that best traded computation
+allows us to obtain the final set of plans that best trade computation
 cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.
@@
 
-The join optimization is based on the algorithm used in System R~\cite{blasgen1981system},
+The join order and imputation optimization is based on the algorithm used in System R~\cite{blasgen1981system},
 but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the Pareto frontiers.
@@
 can be added, possibly targeted to a specific domain. The flexibility of this approach
-aligns well with the spirit of query optimization, which aims to reconcile performance with
+aligns with the spirit of query optimization, which aims to reconcile performance with
 a declarative interface to data manipulation.
@@
 
-These imputation algorithms are used as part of a larger \textit{multiple imputation}
+These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
 process, in which multiple distinct copies of the complete data are generated and estimators
@@
 Our optimization algorithm builds off the approach taken by 
-a canonical cost-base optimizer~\cite{blasgen1981system}.
+a canonical cost-based optimizer~\cite{blasgen1981system}.
 If imputation operators are ignored, we search the same plan space.

### 2017-03-01 paper/introduction.tex
paper: some edits

--- 2017-03-01 paper/introduction.tex
@@
 carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
-significant speedup in query execution while minimizing the number of dropped rows.
-Unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
+significant speedup in query execution while mitigating the number of dropped rows.
+Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
 optimally in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

### 2017-03-01 paper/algorithm.tex

@@
-\vspace{0.5in}
 \section{Algorithm}

### 2017-03-01 paper/relatedwork.tex
Change regression trees -> decision trees

@@
 existing values (observed and missing). Methods in the statistical community
-focus on correctly modeling relationships between attributes to account for different forms of missingness.
-For example, Burgette and Reiter~\cite{burgette2010multiple} discuss the usage of sequential regression trees
-for imputing missing data.
+focus on correctly modeling relationships between attributes to account for different forms
+of missingness. For example, \cite{burgette2010multiple} discuss the
+usage of iterative decision trees for imputing missing data.
 
-~\cite{akande2015empirical} analyzes the performance of various
+\cite{akande2015empirical} analyzes the performance of various
 multiple imputation techniques on the American Community Survey dataset. 
 The computational difficulties of imputing on large base
-tables are well-known and can limit approaches. For example, Akande finds that
+tables are well-known and can limit approaches. For example, \cite{akande2015empirical} finds that
 one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
@@
 papers have described various (at times conflicting) treatments
-of nulls~\cite{grant1977null}. \ProjectName's main design invariant---no relational operator
-sees missing values for attributes it must operate on and users should never see
+of \nullv{}s~\cite{grant1977null}. \ProjectName's main design invariant---no relational operator
+sees missing values for attributes it must operate on nor do users see
 missing data---eliminates
@@
 In~\cite{wolf2007query}, queries over databases with missing values are
-processed using a statistical approach, taking advantage of correlations between
+processed directly using a statistical approach, taking advantage of correlations between
 attributes. The tuples that match the original query as well as a ranking of
@@
 reconfigure the query plan for state-dependent
-optimizations as data continue to arrive. One of the applications of this framework is to avoid
+optimizations as data continues to arrive. One of the applications of this framework is to avoid
 expensive imputations as real-time conditions change. 
-\ProjectName{} focuses on handling
-missing values, rather than providing a general framework for planning feedback. Furthermore,
-it provides the ability
-to trade imputation quality for runtime performance. 
 %We prove our
@@
 to produce forecasts within a given accuracy threshold. Accuracy estimates are determined using
-standard techniques, such as k-fold cross-validation.
+standard techniques, such as cross-validation.
 

### 2017-03-01 paper/examples.tex
paper: minor fixes

@@
 for handling missing values. She currently has two options:
-1) she could drop records with missing values in relevant fields,
-2) she could use a standard data imputation package on all of her data. Both of these
+(1) she could drop records with missing values in relevant fields,
+(2) she could use a standard data imputation package on all of her data. Both of these
 approaches have significant drawbacks. For the query pictured in~\Cref{fig:example-query}, 

### 2017-03-01 paper/introduction.tex
paper: minor fixes

--- 2017-03-01 paper/introduction.tex
@@
 % Background
-While traditional imputation methods work over the entire data set and replace all missing
-values, running a sophisticated imputation algorithm over a large data set can be very
-expensive---our experiments show that even a relatively simple decision tree algorithm takes \acsbaseresulthours{} to train and run
+While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression tree algorithm takes \acsbaseresulthours{} to train and run
 on a 600K row database.
@@
 \subsection{Contributions}
-\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
-increasing productivity for users who frequently encounter missing values in their data. These contributions include:
+\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches.
+These contributions include:
 \begin{itemize}

### 2017-03-01 paper/conclusion.tex

@@
 existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
-learning technique, as operators in the algebra and use a simple, but effective cost model to consider trade-offs in 
+learning technique, as operators in the algebra and use a simple yet effective cost model to consider trade-offs in 
 imputation quality and runtime. We show that different values for the trade-off parameter can yield substantially
 different plans, allowing the user to express their preferences for performance on a per-query basis.
-Our experiments with the \textit{CDC NHANES}, \textit{freeCodeCamp} and \textit{ACS} datasets
-show that \ProjectName{} can be used successfully with real-world data, to run standard SQL queries over data with missing values. 
+Our experiments with the CDC NHANES, freeCodeCamp and ACS datasets
+show that \ProjectName{} can be used successfully with real-world data to run standard SQL queries over data with missing values. 
 We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
@@
 We discuss the long history of dealing with missing data both in the statistical learning and database communities.
-Similarly to existing work, we consider the impact of null values in databases and develop a simple set of invariants to 
+Similar to existing work, we consider the impact of \nullv{} values in databases and develop a simple set of invariants to 
 successfully plan around them. In contrast to existing statistical learning work, our emphasis is not on the specific algorithm
@@
 \subsection{Future Work}
-\ProjectName{} opens up multiple avenues for further work. We sketch out some of these
-directions.
+\ProjectName{} opens up multiple avenues for further work.
 
 \textbf{Imputation confidence:} Obtaining confidence measures for query results
-produced when missing values are imputed with a machine learning model is
+produced when missing values are imputed with a statistical model is
 another possible improvement to the system. These measures could be obtained in general
 using resampling methods like cross-validation and bootstrapping\cite{kohavi1995study}.
-In contrast to standard confidence calculations, there is the added complexity that
+In contrast to standard confidence calculations, one added complexity is that
 imputation takes place at various points in the plan, so the composition of results
-through standard query operations needs to be taken into account when computing confidence intervals.
-Finally, multiple imputation could be considered explicitly and confidence measures like variances could
+through standard query operations needs to be taken into account when computing confidence
+measures.
+In addition, multiple imputation could be used within query execution and confidence measures like variances could
 similarly propagate through the query plan.

### 2017-03-01 paper/introduction.tex

--- 2017-03-01 paper/introduction.tex
@@
 carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
-significant speedup in query execution while mitigating the number of dropped rows.
+significant speedup in query execution while reducing the number of dropped rows.
 Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations

### 2017-03-01 paper/experiments.tex
paper: minor fixes

@@
   \input{tables/fcc}
-  \caption{Missing value distribution for each attribute in freeCodeCamp Survey Data (\texttt{fcc}}\label{table:fcc-description} 
+  \caption{Missing value distribution for each attribute in freeCodeCamp Survey Data (\texttt{fcc})}\label{table:fcc-description} 
 \end{table}
@@
 \Cref{q9} joins the freeCodeCamp data with a reference table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
-The query calculates the average GDP per-capita of countries with and without bootcamp participants over 18 years of age.
+The query calculates the average GDP per-capita by grouping 18+ year old participants who attended bootcamp versus those who did not.
 

### 2017-03-01 paper/experiments.tex
minor fix

@@
 Although it is tempting to further optimize query execution in \ProjectName{} by pre-training imputation models while the analyst begins to make queries,
-this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are not known until runtime.
+this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are unknown until runtime.
 

### 2017-03-01 paper/experiments.tex
Minor edits to experiments

--- 2017-03-01 paper/experiments.tex
@@
 For our experiments we plan and execute queries for three separate survey-based data sets, showing that our system is well suited for early dataset exploration.
-We show that \ProjectName{} performs an order of magnitude better than imputation on the full dataset and produces results of comparable quality.
+We show that \ProjectName{} performs an order of magnitude better than traditional
+base-table imputation and produces results of comparable quality.
 
 \subsection{Implementation}
-To evaluate the performance of our approach, we implemented a prototype system in Java, following a traditional iterator model.
+To evaluate the performance of our approach, we implement a prototype system in Java, following a traditional iterator model.
 Our database handles a subset of standard SQL queries, including filters, joins, projection, grouping and aggregation.
-For simplicity, \ProjectName{} only handles tables containing integer data, so we pre-process our experimental datasets before loading them.
+For simplicity, \ProjectName{} only handles tables containing integer data.
 
 \subsection{Data sets}\label{subsec:datasets}
-We collected three data sets for our experiments.
+We collect three data sets for our experiments.
 For all data sets, we selected a subset of the original attributes.
-We also transformed all data values to an integer representation by enumerating strings and transforming floating-point values into an appropriate range.
+We also transform all data values to an integer representation by enumerating strings and transforming floating-point values into an appropriate range.
 
@@
 For our first set of experiments, we use survey data collected by the 
-Centers for Disease Control and Prevention (CDC) in the United States. We
+U.S. Centers for Disease Control and Prevention (CDC). We
 experiment on a set of tables collected as part of the 2013--2014 National
@@
 \begin{itemize}
-	\item \emph{Demographics}: demographic information of subjects
-	\item \emph{Examinations}: physical exam results
-	\item \emph{Laboratory}: laboratory exam results
+	\item \texttt{demo}: demographic information of subjects
+	\item \texttt{exams}: physical exam results
+	\item \texttt{labs}: laboratory exam results
 \end{itemize}
@@
 The original tables have a large number of attributes, in some cases providing more granular tests results or alternative metrics.
-We focused on a subset of the attributes for each table to simplify the presentation and exploration of queries.
+We focus on a subset of the attributes for each table to simplify the presentation and exploration of queries.
 \Cref{table:nhanes-description} shows the attributes selected, along with the percentage of \nullv{} values for each attribute.
@@
 
-The queries on the CDC NHANES data consist not only of projections and selections, but also
+The queries consist not only of projections and selections, but also
 interesting joins and aggregates. Our aim was to craft meaningful queries that would
@@
 Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}).
-\Cref{q1} calculates calculate the average cuff size for individuals based on their income data, with a constraint on height.
+\Cref{q1} calculates the average cuff size for individuals based on their income data, with a constraint on height.
 \Cref{q2} compares creatine levels for individuals with low, medium, and high incomes and above a certain weight.
@@
 \Cref{q8} calculates the average amount of student debt based on school degree for survey participants who have student debt.
-\Cref{q9} joins the freeCodeCamp data with a reference table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
+\Cref{q9} joins the freeCodeCamp data with data provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
 The query calculates the average GDP per-capita of countries with and without bootcamp participants over 18 years of age.
@@
 
-\Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
-configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0$),
-\ProjectName{} optimizing for performance ($\alpha=1$), and imputing at the base tables followed
-by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
-faster than imputing on the base tables. We can get another order-of-magnitude speedup when
-optimizing for performance. This performance differential means it is feasible
+For each query, we evaluate three different configurations: \ProjectName{} optimizing for
+quality ($\alpha=0$), \ProjectName{} optimizing for performance ($\alpha=1$), and imputing
+at the base tables followed by executing the query traditionally.
+
+\Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
+queries are an order-of-magnitude faster than imputing on the base tables. We can get
+another order-of-magnitude speedup when optimizing for performance. This performance
+differential means it is feasible
 to explore multiple imputations operations (including more expensive operators) when using
 \ProjectName{}, in contrast to the traditional approach of base table imputation.
-
-All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
-four 2.9 GHz Intel Xeon E5--2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.
-Each query with each value of $\alpha$ was run 200 times, after discarding a warm-up period.
 
@@
 
+All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
+four 2.9 GHz Intel Xeon E5--2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.
+Each query with each value of $\alpha$ was run 200 times, after discarding a warm-up period.
+
 % The one-standard-deviation
@@
 
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for
-\ProjectName{}'s query results when compared to the results of running imputation on the base tables and
-executing the query on this cleaned base data. Given that the main datasets used
-have no ground truth, this measure indicates how much error is introduced by on-the-fly imputation compared
-to imputation on the base tables.
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error
+(SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results when compared to
+imputing on the base tables and executing the query on this cleaned base data.
+This measure indicates how much error is introduced by on-the-fly imputation compared to
+imputation on the base tables. In a real use-case, the data do not have ground-truth
+anyways, so the appropriate comparison is to this full-fledged impute on the base table.
 
@@
 can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
-to the base-imputation approach is low---between
-\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent---indicating that on-the-fly
-imputation produces similar results to imputation at the base tables.
+to the base-imputation approach is low---between \lowsmapealphazero{} and
+\highsmapealphaoneexacs{} percent---indicating that on-the-fly imputation produces similar
+results to imputation at the base tables. Notably, the results with largest error correspond
+most closely to one existing approach of dropping all \nullv{}s.
 
@@
 expensive.
-To illustrate the increasing difficulty of such an approach as datasets scale, we ran the following query over the ACS dataset:
+To illustrate the increasing difficulty of such an approach as datasets scale, we run the following query over the ACS dataset:
 \begin{lstlisting}[breaklines]

### 2017-03-01 paper/experiments.tex

@@
 \Cref{q9} joins the freeCodeCamp data with data provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
-The query calculates the average GDP per-capita of countries with and without bootcamp participants over 18 years of age.
+The query calculates the average GDP per-capita by grouping 18+ year old participants who attended bootcamp versus those who did not.
 

### 2017-03-01 paper/main.tex
Couple edits to results writeup

@@
 \newcommand{\nullv}{NULL}
-\newcommand{\planningruntimepercent}{8.5}
-\newcommand{\planningmaxtime}{40}
+\newcommand{\planningruntimepercent}{1.8}
+\newcommand{\planningmediantime}{1.5}
+\newcommand{\planningmaxtime}{34}
 
@@
 \begin{abstract}
-  Missing values abound in data analysis and present a usability challenge.
-  Users are forced to pick between removing tuples with missing values or creating a clean duplicate of their data by applying a relatively expensive imputation strategy.
-  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on the fly for each query.
+  Missing values are common in data analysis and present a usability challenge.
+  Users are forced to pick between removing tuples with missing values or creating a cleaned version of their data by applying a relatively expensive imputation strategy.
+  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on-the-fly for each query.
   This allows users to immediately explore their data, while the system picks the optimal placement of imputation operations.
@@
   % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
-  We  evaluate this approach on two real-world survey-based datasets and 
+  We evaluate this approach on three real-world survey-based datasets and 
   % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
   our experiments show that our query plans are between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
-  Furthermore, we show that the query results from on-the-fly imputation differ in one error measure from the standard full-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%.
+  Furthermore, we show that the query results from on-the-fly imputation differ from the standard full-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
   Finally, we show that while dropping tuples with missing values that fail query constraints discards 3--88\% of the data, on-the-fly imputation loses only 0--10\%.

### 2017-03-01 paper/experiments.tex
paper:fix count fractions

@@
 We also calculate the number of tuples used to produce each aggregate output.
-The Count columns in \Cref{table:smape} show the percentage difference between the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ and running the query on the imputed base table.
+The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples
+used when running the query on the imputed base table.
 This shows that when optimizing for performance, not quality, many tuples are dropped.
@@
     for quality ($\alpha=0$) generally achieve lower error than queries optimized for
-    efficiency ($\alpha=1$). Count shares show the number of results returned from each
-    query as a percentage of the number of results returned when imputing on the base table.
+    efficiency ($\alpha=1$). The count fraction column shows the number of tuples used in calculating each aggregate
+     as a fraction of the number of tuples used when running the same query after imputing on the base table.
     A lower count share reflects more potential for errors.}

### 2017-03-01 paper/algorithm.tex
Minor edits to algorithm

@@
 
-$\alpha = 1.0$ means that the optimal query should have the fastest runtime (at the expense of quality), while $\alpha=0.0$ means that the optimal query should lose
-as little information as possible (at the expense of performance).
+If $\alpha=0.0$, then the optimal query should lose as little information as possible (at
+the expense of performance). If $\alpha = 1.0$, then the optimal query should have the
+fastest runtime (at the expense of quality).
 
@@
 correspondingly, the information lost \textit{decreases} with more values to work
-with. We wanted to encode the fact that the number of training observations has diminishing returns for imputation quality,
-and found that scaling the cell count by applying square root worked well in practice.
-
-To add an alternate imputation strategy to \ProjectName{}, the analyst must provide a corresponding loss with a reasonable parameterization.
+with. We encode the fact that the number of training observations has diminishing returns for imputation quality,
+and find that scaling the cell count by applying square root works well in practice.
 
@@
 algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
-and simply queries it for a time estimate. In the base implementation of our system,
+and simply queries it for a time estimate. In our reference implementation,
 we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
@@
 is not easily extracted (especially in the case that the entire set of tuples does not
-fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
-not even consider CPU cost explicitly). For example,~\cite{martin1995time} find that the
+fit in memory), and the CPU cost dominates (in contrast to the other operators, which do
+not consider CPU cost explicitly). For example,~\cite{martin1995time} finds that the
 time complexity of the \textit{build tree} algorithm for one commonly-used class of
@@
 does not even dominate computation, as post-processing steps like pruning, which are vital
-in achieving good performance, can have cost exponential in the height of the tree. In
-practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
-plans).
+in achieving good performance, can have cost exponential in the height of the tree.
+Nonetheless, in practice, we find that a simple parameterization can be acceptable (i.e.
+yield intelligent query plans).
 
@@
 Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
-At a high level, a plan cache is a mapping from a set of tables to a set of dirty-set specific dominating plans that perform the required joins and filters over these tables.
+At a high level, a plan cache is a mapping from a set of tables to a set of dirty set-specific dominating plans that perform the required joins and filters over these tables.
 We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
-allows us to obtain the final set of plans that best trade computation
+allows us to obtain the final set of plans that best trade-off computation
 cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.
@@
 
-\ProjectName{} is designed so that any imputation strategy, typically a predictive machine
-learning model,
+\ProjectName{} is designed so that any imputation strategy
 can be plugged in with minimal effort and without changes to the optimizer, simply by adding
 time and loss functions (\Cref{sec:cost-model}) which describe the new algorithm.
-An imputation strategy is a function $I(T, C) \rightarrow T'$ which fills in the missing
-attributes $C$ in a set of tuples $T$, returning a new set of tuples $T'$.
-In principle, new imputation algorithms
-can be added, possibly targeted to a specific domain. The flexibility of this approach
-aligns with the spirit of query optimization, which aims to reconcile performance with
-a declarative interface to data manipulation.
+For our purposes, an imputation strategy is a function $I(R, C) \rightarrow R'$ which fills in the missing
+attributes $C$ in a relation $R$, returning a new set of tuples $R'$.
+The flexibility of this approach aligns with the spirit of query optimization, which aims to
+reconcile performance with a declarative interface to data manipulation.
+
+In principle, new imputation algorithms can be added, possibly targeted to a specific
+domain. However, \ProjectName{}'s planner will consider an arbitrary subset of the
+attributes in any imputation, thus it is feasible to customize the \textit{type} of
+algorithm but not necessarily a specific modelling relationships between attributes.
 
@@
 imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
-regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
+regression}~\cite{gelman2006data}) impute multiple missing attributes being imputed by iteratively fitting
 a predictive model of one missing attribute, conditional on all complete attributes
@@
 The reference algorithm used by the \textit{Impute} operator proceeds by iteratively fitting
-decision trees to a subset of the data and the target imputation attribute. In each
-iteration, the missing values of a single attribute are replaced with newly imputed values.
+decision trees for the target imputation attribute on a subset of the data. In each
+iteration, the missing values of a separate attribute are replaced with newly imputed values.
 In one {\it epoch} of the algorithm, each of the target imputation
 attributes is filled in with imputed values in a single iteration.
-With each {\it epoch}, the quality of the imputation improves as values progressively
+With each epoch, the quality of the imputation improves as values progressively
 reflect more accurate relationships between attributes. The algorithm terminates when
@@
 are computed by averaging over the multiple datasets to form a single completed dataset~
-\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though in
-further work, multiple imputation could be another avenue of assessing
-imputation confidence.
+\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
+multiple imputation could incorporated as another avenue of assessing imputation confidence.
 

### 2017-03-01 paper/algorithm.tex
paper: tweaks

--- 2017-03-01 paper/algorithm.tex
@@
 \section{Algorithm}
-In order to correctly plan around missing values, we first extend the relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
+In order to correctly plan around missing values, we first extend the set of relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
 with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
@@
 \subsubsection{Pareto-optimal plans}
+Given a query, \ProjectName{} produces a final set of plans $P$ by only keeping plans that are not dominated by others in the search space.
+So for a search space $S$, \[P= \{ p ~|~ \not\exists p' \in S.\ p \neq p' \land \doms{p'}{p} \}.\]
+
 \begin{definition}
@@
 \end{definition}
-
-Given a query, \ProjectName{} produces a final set of plans $P$ by only keeping plans that are not dominated by others in the search space.
-So for a search space $S$, \[P= \{ p ~|~ \not\exists p' \in S.\ p \neq p' \land \doms{p'}{p} \}.\]
 
@@
 algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
-and simply queries it for a time estimate. In the base implementation of our system,
+and simply queries it for a time estimate. In the implementation of our system,
 we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost

### 2017-03-01 paper/algorithm.tex

--- 2017-03-01 paper/algorithm.tex
@@
 algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
-and simply queries it for a time estimate. In our reference implementation,
+and simply queries it for a time estimate. In the implementation of our system,
 we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost

### 2017-03-01 paper/experiments.tex
paper: more tweaks

@@
 \subsection{Implementation}
-To evaluate the performance of our approach, we implement a prototype system in Java, following a traditional iterator model.
+To evaluate the performance of our approach, we implemented a prototype system in Java, following a traditional iterator model.
 Our database handles a subset of standard SQL queries, including filters, joins, projection, grouping and aggregation.
@@
 %\end{figure}
-\vspace{0.5in}
 \subsubsection{Accuracy vs. Base-table Imputation}

### 2017-03-02 paper/introduction.tex
Minor changes to intro and motivating example

@@
 % Solution
-One common approach is to manually replace missing values using some sort of statistical or predictive model based on other values in the table and record.
+One common approach is to manually replace missing values using a statistical or predictive model based on other values in the table and record.
 This process is called \emph{imputation}.
-In this paper, we present a new system, called \ProjectName{}, which is designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.
+In this paper, we introduce \ProjectName{}, a new system designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.
 \emph{The key insight behind \ProjectName{} is that imputation
@@
 A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
-In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
-carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
+In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set. Specifically, \ProjectName{}
+carefully plans the placement of imputation/row-drop operations inside the query plan, resulting in a 
 significant speedup in query execution while reducing the number of dropped rows.
 Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
-optimally in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.
+optimally in query plans.   Specifically, our optimization algorithms generate the
+Pareto-optimal frontier of plans that trade imputation cost for result quality,
+and allow the analyst to specify their desired plan from this frontier.
 
@@
 % Implications
-Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
-queries that execute between \lowxalphazero{} and \highxalphazeroexacs{} times as fast as it takes to impute all missing values and then run queries (the traditional approach).
+Our approach enables an exploratory analysis workflow in which the analyst can issue
+standard SQL queries over a data set, even if that data has missing values.
+These queries execute between \lowxalphazero{} and \highxalphazeroexacs{} times as fast as it takes to impute all missing values and then run queries (the traditional approach).
 Furthermore, the results obtained with this approach
-are similar in quality to those obtained with the complete imputation approach, as indicated
-by one measure of error --- between \lowsmapealphazero{} and \highsmapealphaoneexacs{}
-percent --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
+are similar in quality to those obtained with the traditional imputation approach
+--- between \lowsmapealphazero{} and \highsmapealphaoneexacs{}
+percent, in one error measure --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
 for details).
@@
 % Scope
-\ProjectName{} is designed to enable early data exploration, by allowing analysts to run their queries without explicitly deciding when to perform imputation.
+\ProjectName{} is designed to enable early data exploration, by allowing analysts to run
+their queries without an explicit base-table imputation step.
 % SAM -- cut the following -- could consider moving later but feels out of place here
@@
   We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
-  The impute operation fills in missing data values using any statistical imputation
+  The \textit{Impute} operation fills in missing data values using any statistical imputation
   technique, such as chained-equation decision trees ~\cite{burgette2010multiple}.
-  The drop operation simply drops tuples which contain \nullv{} values.
+  The \textit{Drop} operation simply drops tuples which contain \nullv{} values.
 \item \textbf{Model of imputation quality and cost:}
-  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
+  We extend the traditional cost model for query plans to incorporate a measure of the
+  \textit{quality} of the imputations performed.
   We use the cost model to abstract over the imputation algorithm used.

### 2017-03-02 paper/main.tex
paper: more tweaks
Minor changes to intro and motivating example

@@
   our experiments show that our query plans are between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
-  Furthermore, we show that the query results from on-the-fly imputation differ from the standard full-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
+  Furthermore, we show that the query results from on-the-fly imputation differ from the
+  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
   Finally, we show that while dropping tuples with missing values that fail query constraints discards 3--88\% of the data, on-the-fly imputation loses only 0--10\%.

### 2017-03-02 paper/algorithm.tex

@@
 Nonetheless, in practice, we find that a simple parameterization can be acceptable (i.e.
-yield intelligent query plans).
+yields intelligent query plans).
 
@@
 collect sound and complete Pareto frontiers. For plans with the same dirty set, we can treat $\langle \textsc{Loss}, \textsc{Time} \rangle$
-as a total order, allowing us to appropriately select the frontier. Meanwhile, plans with different $\textsc{Dirty}$ sets cannot be compared. The final
+as a total order, allowing us to appropriately select the frontier. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
 Pareto frontier produced by the optimizer corresponds to the collection of plans associated with the empty dirty set.
@@
 For our purposes, an imputation strategy is a function $I(R, C) \rightarrow R'$ which fills in the missing
-attributes $C$ in a relation $R$, returning a new set of tuples $R'$.
+attributes $C$ in a relation $R$, returning a new relation $R'$.
 The flexibility of this approach aligns with the spirit of query optimization, which aims to

### 2017-03-02 paper/relatedwork.tex
paper: more tweaks
paper: add paragraph intro

@@
 
-\vspace{0.5in}
 \subsection{Missing Values and Statistics}
@@
 
-BayesDB~\cite{mansinghka2015bayesdb} provides users with a simple interface to 
-leverage statistical inference techniques in a database. Non-experts
+Other work has looked at integrating statistical models into databases.
+For example, BayesDB~\cite{mansinghka2015bayesdb} provides users with a simple interface to leverage statistical inference techniques in a database. Non-experts
 can use a simple declarative language (an extension of SQL), to specify models

### 2017-03-02 paper/examples.tex

--- 2017-03-01 paper/examples.tex
@@
 for handling missing values. She currently has two options:
-1) she could drop records with missing values in relevant fields,
-2) she could use a standard data imputation package on all of her data. Both of these
+(1) she could drop records that have missing values in relevant fields,
+(2) she could use a traditional imputation package on her entire dataset. Both of these
 approaches have significant drawbacks. For the query pictured in~\Cref{fig:example-query}, 
 (1) drops 1492 potentially relevant tuples,
-while her previous experiences with (2) have proven to be expensive. The epidemiologist needs a more complete picture
-of the data, so (1) is insufficient, and she is under time pressure, so (2)
+while from her experience, (2) has proven to take too long. The epidemiologist needs a more complete picture
+of the data, so (1) is insufficient, and for this quick exploratory analysis, (2)
 is infeasible.
@@
 
-\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operation $\mu$, which employs an iterative regression model to fill in missing values rather than dropping tuples.
-It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
+\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operation $\mu$,
+which employs a reference imputation strategy to fill in missing values rather than dropping tuples.
+It waits to impute \verb|demo.income| until after the final join has taken place, though other imputations take place earlier on in the plan, some before filtering and join operations.
 Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
@@
 Now that the epidemiologist has \ProjectName{}, she can explore
-the dataset using SQL. For each query, she repeatedly chooses a value
-for $\alpha$ and runs the query until she is unsatisfied with the runtime.
-Rather than commit \emph{a priori} to a runtime bound for her query,
-which may be unsatisfiable, this iterative approach gives her 
+the dataset using SQL. She begins by choosing a value for $\alpha$ and can adjust it up or
+down until she is satisfied with her query runtime. This iterative approach gives her 
 immediate feedback.

### 2017-03-02 paper/conclusion.tex
paper: add itemize to conclusion

@@
 
-\textbf{Imputation confidence:} Obtaining confidence measures for query results
+\begin{itemize}
+\item \textbf{Imputation confidence:} Obtaining confidence measures for query results
 produced when missing values are imputed with a statistical model is
@@
 
-\textbf{Imputation operators:} We explored a subset of possible imputation operators.
+\item \textbf{Imputation operators:} We explored a subset of possible imputation operators.
 In particular, we explored two variants of the \textit{Drop} and \textit{Impute} operations. The system
@@
 
-\textbf{Adaptive query planning:} Currently, the imputation is local to a given query, meaning
+\item \textbf{Adaptive query planning:} Currently, the imputation is local to a given query, meaning
 no information is shared across queries. An intriguing direction would be to take advantage
@@
 query plans which have been shown to produce low-confidence imputations.
+\end{itemize}
 

### 2017-03-02 paper/main.tex

@@
 \usepackage{xspace}
+\usepackage[all]{nowidow}
 
@@
   % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
-  We evaluate this approach on three real-world survey-based datasets and 
+  We evaluate this approach on three real-world survey-based datasets.
   % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
-  our experiments show that our query plans are between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
+  Our experiments show that our query plans execute between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
   Furthermore, we show that the query results from on-the-fly imputation differ from the

### 2017-04-25 rebuttal.tex
rebut: add rebuttal outline

\documentclass[draft,12pt]{article}
\usepackage[margin=1in]{geometry}

\begin{document}
\section{Reviewer 1}
\begin{itemize}
\item W1. A number of the algorithms/figures are a bit redundant (e.g., Histogram estimation for classical join operators in Alg. 3) Overall, the paper in its current form feels padded -- it could be trimmed down by 1-2 pages. See below for a few ideas of things to use this space for. I will note that I'd be perfectly happy accepting this as a shorter, unpadded paper.

\item W2. The paper completely obscures the ML technique being used: (1) It would be nice to have some discussion of the complete API for implementing an ML technique, as well as how one might implement a cost-estimator for one or two examples. (2) It would be helpful to see the impact of a few different approaches (e.g., Tree-based, Interpolation, Regression, FD-based) on performance and accuracy. (3) Given that the ML time estimator is being treated as a black box, it would be useful to see how sensitive the plan selector is to inaccurate estimates.

\item W3. Reducing runtime to ~10s is excellent, but comes at a very serious accuracy cost (3-5x worse SMAPE). Worse still, there's no way for the user to differentiate between the cases of Q4 and Q8 at runtime. 24\% error relative to direct imputation is huge.

\item - Although it doesn't consider cost or the drop operator, a similar technique appears in Yang et. al. "Lenses: An On-Demand Approach to ETL." A comparison would be nice.

\item - S1: "Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach — between 0 and 24 percent" 24 percent error is pretty bad.

\item - S2: The drop operator appears to be a weaker form of SQL's null-value / 3-valued Boolean semantics. The latter might actually be more useful. For example, consider the case of a predicate (A $>$ 5 OR B $<$ 10). If A $>$ 5, then the condition is independent of B, and the tuple can still be salvaged. If A $\leq$ 5, then the expression evaluates to UNKNOWN and the tuple is dropped anyway. For aggregates there is a greater need, as one might have different aggregates computed over different subsets of the data, but even here it's not clear that this is any less informative than just throwing data away.

\item - S2: Why no unions?

\item - S3.4.1, Definition 2: On my first read-through, I was puzzled about why there was no plan with zero-loss. It would help to clarify that Loss is not about tuple loss, but rather a heuristic penalty function.

\item - S3.4.3: An extremely nice emergent property from the L-value function for the imputation operator seems to be that the operator will be moved to a position in the plan where it has access to the maximum training data. I am slightly concerned that by focusing entirely on size, this might lead to skew in the training set. For example, consider a one-to-many join \verb/µ(R) |X| S/. Your optimization algorithm would likely wish to move imputation up: \verb/µ(R |X| S)/. Yes, there is more training data, but if the fanout for each tuple is skewed, some tuples in R will become over-represented in the join result. A similar observation holds for selection. For example take your Q8, where the conditions "studentdebtowe $>$ 0 AND schooldegree >= 0" presumably indicate invalid data that should not be used for training. (is this perhaps why you have so much error for full imputation on Q8?)

\item - S3.4.4: It appears that Time() assumes that plans are IO-bound. This is a perfectly reasonable assumption, but should be stated explicitly, as datasets that fit in memory will require a different cost model.

\item - S3.4.4: "Drop is a special case of a sequential scan, and its time complexity can be expressed as a function of the number of heap pages read and the I/O cost per page." seems to suggest that drop is implemented as a blocking operator. Why not just take the same approach as selection and treat it as having negligible cost?

\item - S4: All of the workload queries are (data-size-independent) AVG queries. What about other types of aggregates (e.g., SUM, COUNT)?

\item - S4.4.1: Imputation over the raw data is being used as ground truth. Is there any reason to suspect that this is the correct thing to do? Perhaps joining multiple tables together exposes new attributes that makes imputation *more* reliable? At the very least, it would be useful to understand why there is so much error in $\alpha=0$ on Q2 and Q8.

\item - There are a number of opportunities for trimming in the figures, tables, and algorithms: Algorithms 1-3 could be compacted a bunch, Table 1 is full of whitespace (e.g., make it a 3-column {figure*}), Tables 1-2 could also be summarized as CDFs, Figure 6 is full of whitespace, Table 3 could be put queries on a single row (e.g., Q1 would easily fit on 1 line), and the references have a lot of redundant information (URLs, for example).

\item - Additional space recovered from these sections could be used to discuss the choice of techniques for imputation, and the confidence estimation idea suggested in the future work section.

\item - Another interesting use of the space, or at least an idea for future work might be the role of aggregation on inference accuracy. In my experience aggregate queries tend to be *much* more forgiving of approximations (e.g., Online Aggregation) and data errors.
\end{itemize}

\section{Reviewer 2}
\begin{itemize}
\item 1) No new imputation method is presented. The model relies on external imputation techniques, and given it's done on-the-fly, presumably imputed values depend on qualifying query records. 
2) No comparative evaluation against existing imputation/data cleaning techniques.

\item D1) The model relies on external imputation techniques, and given it's done on-the-fly, the imputed values will differ based on the qualifying query records?

\item D2) The paper does not propose new imputation techniques but rely on existing statistical imputation algorithms. There is no comparative evaluation, nor samples of imputed values to show the goodness of the imputed values. It would be nice to see a comparative evaluation of different statistical imputation methods, and suggestions for extensions that can fit within the proposed imputation model.

\item D3) Table 3 uses a lot of space, and contains simple queries that are already described in the text. The table can be removed to include additional experiments as suggested above. 
\end{itemize}

\section{Reviewer 3}
\begin{itemize}
\item W1. The paper feels like an early submission that is not ready for publication at its current state.

\item W2. More work is required in the presentation: important parts are not explained in depth, while there are many details in other sections that can be omitted.

\item W3. More experiments needed.

\item D1. Section 3, which is the meat of the paper, is very overloaded, without getting in depth in various parts of the system. It definitely should be split in more sections, and more details should be given. In particular:
-- Give some more information about cardinality estimation. Also, Algorithm 3 is too far from the text, and needs some explanation too.
-- Some more details are also required in the Imputation quality section. Why are you choosing these specific formulas for Loss, especially for the impute operator. Some examples would help.
-- Algorithms 1 and 2 could probably be presented in an easier to grasp way.

\item D2. It would be nice to show how exactly your approach can be implemented in an existing optimizer. This comes up in various places in the paper:
-- When defining the search space, why not first define the transformation operators that inject the imputation operators. Then any optimizer can pick these rules up and add them in its plan enumeration. Of course, the used approach that follows System-R's plan enumeration is useful, but why lose generality when presenting the search space. Moreover, showing the transformation rules can help you show the size of the search space that will better justify the restrictions you apply.
-- Why use simplified cost formulas for the traditional relational operators and not existing systems' formulas?
-- It would be great (but not necessary of course) to add your techniques in an existing DBMS, such as Postgres to see the impact on a real system in your experiments.

\item D3. Is there any way for the user to know what error to expect based on the $\alpha$ value? Moreover, is its value somehow proportional to the error (e.g., $\alpha=0.5$ would give roughly double the error of $\alpha=0.25$)? Or should the user simply try multiple execution with different values of $\alpha$ until the desired results are given?

\item D4. How would a two-step approach that first finds the optimal plan (w/o imputation) and then adds the imputation operators perform?

\item D5. Can you have plans with a mix of drop and impute operators?

\item D6. During plan enumeration, can you safely prune plans at some point, similar to what a traditional optimizer would do?

\item D7. When defining the cost of impute operator, how do you know a priori the number of iterations for the algorithm (given it has to converge)?

\item D8. Are there impute implementation that are non-blocking? Or do they always have to perform multiple passes over the data?

\item D9. Additional experiments I suggest to include:
-- Consider a scenario where the user submits multiple similar queries. Show after how many such queries your approach has no longer a competitive advantage compared to performing imputation on the data.
-- Try different values for $\alpha$ (not just 0 and 1).
-- Show runtime of enumeration algorithm as queries get more complex.

\item D10. The scenario you are considering, with multiple queries during exploration, begs for reuse of previous query results, as you also mention in the future work section.

\item D11. Less important points:
-- Sections 4.2 and 4.3 in the experiments are too verbose.
-- In Section 3.2, you mention that joins are performed after filtering. When filters are on results of joins, this is not possible.
-- Given you optimize for both performance and quality, Tummer's and Koch's work on multi-objective QO is related.
-- Could be interesting for the optimizer to pick imputation strategy based on the type of the query.
\end{itemize}
\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2017-04-25 paper/algorithm.tex

@@
 Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
-At a high level, a plan cache is a mapping from a set of tables to a set of dirty set-specific dominating plans that perform the required joins and filters over these tables.
+At a high level, a plan cache is a mapping from a set of tables to a set of dirty set-specific dominating plans over these tables.
 We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
@@
 
-In principle, new imputation algorithms can be added, possibly targeted to a specific
-domain. However, \ProjectName{}'s planner will consider an arbitrary subset of the
-attributes in any imputation, thus it is feasible to customize the \textit{type} of
-algorithm but not necessarily a specific modelling relationships between attributes.
-
-As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
+New imputation algorithms can also be added. As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
 based on chained-equation decision trees.  Chained-equation
 imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
-regression}~\cite{gelman2006data}) impute multiple missing attributes being imputed by iteratively fitting
+regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
 a predictive model of one missing attribute, conditional on all complete attributes

### 2017-04-25 paper/examples.tex

@@
 She can run her queries immediately and finish her report if
-she uses \ProjectName{}, which allows her to write her
-query in standard SQL and automatically performs the imputations necessary to fill in the missing data.
+she uses \ProjectName{}, which takes standard SQL and automatically performs the imputations
+necessary to fill in the missing data.
 An example query is shown in~\Cref{fig:example-query}.

### 2017-04-25 paper/experiments.tex

@@
 To evaluate the performance of our approach, we implemented a prototype system in Java, following a traditional iterator model.
-Our database handles a subset of standard SQL queries, including filters, joins, projection, grouping and aggregation.
+Our database handles a subset of standard SQL, including filters, joins, projection, grouping and aggregation.
 For simplicity, \ProjectName{} only handles tables containing integer data.
@@
 \Cref{q8} calculates the average amount of student debt based on school degree for survey participants who have student debt.
-\Cref{q9} joins the freeCodeCamp data with data provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
+\Cref{q9} joins the freeCodeCamp data with a table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
 The query calculates the average GDP per-capita by grouping 18+ year old participants who attended bootcamp versus those who did not.
@@
 
-In the median query across a variety of queries and choices of $\alpha$, planning took
+In the median across a variety of queries and choices of $\alpha$, planning took
 \planningmediantime{} ms, 
@@
 four 2.9 GHz Intel Xeon E5--2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.
-Each query with each value of $\alpha$ was run 200 times, after discarding a warm-up period.
+Each query and each value of $\alpha$ was run 200 times, after discarding a warm-up period.
 
@@
 This measure indicates how much error is introduced by on-the-fly imputation compared to
-imputation on the base tables. In a real use-case, the data do not have ground-truth
-anyways, so the appropriate comparison is to this full-fledged impute on the base table.
+imputation on the base tables. In our case, the relevant comparison is against traditional
+imputation on the base table, given the unavailability of ground truth.
 

### 2017-04-25 paper/introduction.tex

@@
 values, running a sophisticated imputation algorithm over a large data set can be very
-expensive---our experiments show that even a relatively simple decision tree algorithm takes \acsbaseresulthours{} to train and run
+expensive: our experiments show that even a relatively simple decision tree algorithm takes \acsbaseresulthours{} to train and run
 on a 600K row database.
 
-A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
+A simpler approach might drop all rows with missing values, which can not only introduce
+bias into the results, but also result in discarding much of the data.
 In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set. Specifically, \ProjectName{}

### 2017-04-25 rebuttal.tex
rebut: start adding comments

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{xcolor}
+\newcommand{\resp}[1]{{\color{blue}{#1}}}
 
@@
 
+\resp{We skipped unions for brevity. There is nothing fundamental about our algorithm which makes unions difficult to implement.}
+
 \item - S3.4.1, Definition 2: On my first read-through, I was puzzled about why there was no plan with zero-loss.
@@
 A similar observation holds for selection.
-For example take your Q8, where the conditions "studentdebtowe $>$ 0 AND schooldegree >= 0" presumably indicate invalid data that should not be used for training.
+For example take your Q8, where the conditions "studentdebtowe $>$ 0 AND schooldegree $\geq$ 0" presumably indicate invalid data that should not be used for training.
 (is this perhaps why you have so much error for full imputation on Q8?)
+
+  % \resp{This is a reasonable concern.
+Optimizing for the maximum training set is flawed when the training data is skewed.
+We think that this problem is avoidable with a minor tweak to the loss function.
+Instead of optimizing for a large training set, we can optimize for a \emph{diverse} training set by tracking the
 
@@
 This is a perfectly reasonable assumption, but should be stated explicitly, as datasets that fit in memory will require a different cost model.
+
+  \resp{Agreed. Our database stores tables in files, so our plans are I/O bound, but a database which uses an in-memory representation will need a different Time() function.}
 
@@
 \item 1) No new imputation method is presented. The model relies on external imputation techniques, and given it's done on-the-fly, presumably imputed values depend on qualifying query records.
-2) No comparative evaluation against existing imputation/data cleaning techniques.
+
+  \resp{Yes, the imputed results depend on the records which are relevant to particular query.
+We do not present a new imputation method, but our technique is agnostic to the particular imputation method used.
+We reduce the overhead of imputation by performing it only on the tuples which are relevant to a particular query.}
+
+\item 2) No comparative evaluation against existing imputation/data cleaning techniques.
+
+  \resp{Our evaluation compares our technique to an imputation of the full dataset.
+We use the same imputation algorithm for our on-the-fly imputation as for the full dataset imputation.
+Cleaning the full dataset is the standard method of handling missing values.}
 
 \item D1) The model relies on external imputation techniques, and given it's done on-the-fly, the imputed values will differ based on the qualifying query records?
+
+  \resp{Yes, as discussed above.}
 
@@
 It would be nice to see a comparative evaluation of different statistical imputation methods, and suggestions for extensions that can fit within the proposed imputation model.
+
+  \resp{The error measures that we provide compare our on-the-fly application of imputation to imputation over the full dataset.}
 
@@
 
+\resp{Existing systems have very complex formulas for estimating query costs. We felt that the overhead of reimplementing an existing cost model would not be worth the benefit.}
+
 \item D3.
@@
 
+  \resp{In our intended usage, the user starts by running with $\alpha = 0$ and increases $\alpha$ if the query takes too long to complete.}
+
 \item D4. How would a two-step approach that first finds the optimal plan (w/o imputation) and then adds the imputation operators perform?
+
+  \resp{This approach would not be able to improve on the plans produced by our approach, as it would produce a subset of the plans that we consider.
+It's unclear how much worse the plans would be, but a major issue is that the Drop and Impute operators have an impact on the cardinality of their outputs.
+Our optimizer can use this information to improve the plan, but adding imputation after the fact would not be able to rearrange the plan to capitalize on changes in cardinality.}
 
@@
 
+  \resp{Yes.}
+
 \item D6. During plan enumeration, can you safely prune plans at some point, similar to what a traditional optimizer would do?
+
+  \resp{We maintain the set of pareto optimal plans at each point in the optimization.
+This set is significantly smaller than the set of all plans, so it represents some pruning.
+We found that the size of these pareto sets is small enough that we did not need to prune further.
+
+    If necessary, it should be possible to sample from the pareto optimal plans during the plan enumeration.
+The final selected plan may not be pareto optimal, but this is a potential way to control the number of plans.}
 
@@
 -- In Section 3.2, you mention that joins are performed after filtering. When filters are on results of joins, this is not possible.
--- Given you optimize for both performance and quality, Tummer's and Koch's work on multi-objective QO is related.
--- Could be interesting for the optimizer to pick imputation strategy based on the type of the query.
+
+\item -- Given you optimize for both performance and quality, Tummer's and Koch's work on multi-objective QO is related.
+
+  \resp{Yes, Tummer \& Koch is closely related.
+However, much of the complexity of their paper is in dealing with cost functions which are not well behaved.
+Our cost functions are both monotonic, which allows us to use a simpler optimizer than what Tummer \& Koch propose.}
+
+\item -- Could be interesting for the optimizer to pick imputation strategy based on the type of the query.
 \end{itemize}

### 2017-04-25 paper/main.tex
paper: rework queries table to make it smaller

@@
   basicstyle=\ttfamily\footnotesize,        % the size of the fonts that are used for the code
-  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
+  breakatwhitespace=true,         % sets if automatic breaks should only happen at whitespace
   breaklines=true,                 % sets automatic line breaking

### 2017-04-25 paper/queries.tex
paper: rework queries table to make it smaller

--- 2016-12-14 paper/queries.tex
@@
 \toprule
-\# & \multicolumn{1}{c}{Query} \\
+\# & \multicolumn{1}{c}{Queries on CDC data} \\
 \midrule
-1 & \verb|SELECT BLD FROM acs;| \refstepcounter{queryno} \label{q1} \\
-2 & \verb|SELECT BLD as units_in_structure, COUNT(ST) as estimate FROM acs GROUP BY BLD;| \refstepcounter{queryno} \label{q2} \\
-3 & \verb|SELECT BATH as has_bath, COUNT(ST) as ct FROM acs GROUP BY BATH;| \refstepcounter{queryno} \label{q3} \\
-4 & \verb|SELECT ACR as lotsize, AVG(BDSP) as avg_num_bedrooms FROM acs GROUP BY ACR;| \refstepcounter{queryno} \label{q4} \\
-5 & \verb|SELECT PSF as has_sub_families, SUM(NP) as num_people FROM acs GROUP BY PSF;| \refstepcounter{queryno} \label{q5} \\
-6 & \verb|SELECT AVG(NP) as avg_num_people FROM acs;| \refstepcounter{queryno} \label{q6} \\
-7 & \verb|SELECT BDSP as num_bedrooms, AVG(ACR) as avg_lot_size FROM acs WHERE VEH >= 2 GROUP BY BDSP;| \refstepcounter{queryno} \label{q7} \\
-8 & \verb|SELECT MIN(RMSP) as min_num_rooms FROM acs WHERE RWAT=2;| \refstepcounter{queryno} \label{q8} \\
-9 & \verb|SELECT * FROM dirty m where m.f1 >= 2;| \refstepcounter{queryno} \label{q9} \\
-10 & \verb|SELECT * FROM dirty where dirty.f1 <= 2;| \refstepcounter{queryno} \label{q10} \\
-11 & \verb|SELECT MAX(f2) as max_f2 FROM dirty GROUP BY dirty.f1;| \refstepcounter{queryno} \label{q11} \\
-12 & \verb|SELECT m1.f1, m1.f2 FROM dirty m1, clean m2 WHERE m1.f1 = m2.f1 and m1.f1 = 1 and m2.f2 = 69;| \refstepcounter{queryno} \label{q12} \\
-13 & \verb|SELECT d1.f1 FROM dirty d1, clean d2 WHERE d1.f1 = d2.f1 and d2.f2 = 50;| \refstepcounter{queryno} \label{q13} \\
+1 & 
+\begin{minipage}{6in}
+\begin{lstlisting}[breaklines]
+SELECT income, AVG(cuff_size) FROM demo, exams 
+WHERE demo.id = exams.id AND height >= 150 GROUP BY income;
+\end{lstlisting}
+\end{minipage}\refstepcounter{queryno}\label[query]{q1} \\
+2 & 
+\begin{minipage}{6in}
+\begin{lstlisting}[breaklines]
+SELECT income, AVG(creatine) FROM demo, exams, labs 
+WHERE demo.id = exams.id AND exams.id = labs.id AND income >= 13 AND income <= 15 AND weight >= 63 GROUP BY income;
+\end{lstlisting}
+\end{minipage}
+\refstepcounter{queryno}\label[query]{q2} \\
+3 & 
+\begin{minipage}{6in}
+\begin{lstlisting}[breaklines]
+SELECT AVG(blood_lead) FROM demo, exams, labs 
+WHERE demo.id = labs.id AND labs.id = exams.id AND age_yrs <= 6;
+\end{lstlisting}
+\end{minipage}\refstepcounter{queryno}\label[query]{q3}\\
+4 & 
+\begin{minipage}{6in}
+\begin{lstlisting}[breaklines]
+SELECT gender, AVG(blood_pressure_systolic) FROM demo, labs, exams 
+WHERE demo.id = labs.id AND labs.id = exams.id AND body_mass_index >= 30 GROUP BY gender;
+\end{lstlisting}
+\end{minipage}\refstepcounter{queryno}\label[query]{q4}\\
+5 & 
+\begin{minipage}{6in}
+\begin{lstlisting}[breaklines]
+SELECT AVG(waist_circumference) FROM demo, exams 
+WHERE demo.id = exams.id AND height >= 150 AND weight >= 100;
+\end{lstlisting}
+\end{minipage}\refstepcounter{queryno}\label[query]{q5}\\
+\# & \multicolumn{1}{c}{Queries on freeCodeCamp data} \\
+\midrule
+6 & 
+\begin{minipage}{6in}
+\begin{lstlisting}[breaklines]
+SELECT attendedbootcamp, AVG(income) FROM fcc
+WHERE income >= 50000 GROUP BY attendedbootcamp;
+\end{lstlisting}
+\end{minipage}\refstepcounter{queryno} \label[query]{q6} \\
+7 & 
+\begin{minipage}{6in}
+\begin{lstlisting}[breaklines]
+SELECT AVG(commutetime) FROM fcc
+WHERE gender = "female" AND countrycitizen = "United States";
+\end{lstlisting}
+\end{minipage}\refstepcounter{queryno} \label[query]{q7} \\
+8 & 
+\begin{minipage}{6in}
+\begin{lstlisting}[breaklines]
+SELECT schooldegree, AVG(studentdebtowe) FROM fcc
+WHERE studentdebtowe > 0 AND schooldegree >= 0 GROUP BY schooldegree;
+\end{lstlisting}
+\end{minipage}\refstepcounter{queryno} \label[query]{q8}\\
+9 & 
+\begin{minipage}{6in}
+\begin{lstlisting}[breaklines]
+SELECT attendedbootcamp, AVG(gdp_per_capita) FROM fcc, gdp
+WHERE fcc.countrycitizen = gdp.country AND age >= 18
+GROUP BY attendedbootcamp;
+\end{lstlisting}
+\end{minipage}\refstepcounter{queryno} \label[query]{q9}\\
 \bottomrule
 \end{tabular}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2017-04-27 rebuttal.tex
addressed Q8 24% error in rebuttal

@@
 \newcommand{\resp}[1]{{\color{blue}{#1}}}
+\newcommand{\todo}[1]{\color{red}{\textit{#1}}}
 
@@
 24\% error relative to direct imputation is huge.
+\resp{
+The outsized error in Q8 is a result of a drop operator placed very early on in the plan (a result of optimizing for performance). This produced significantly smaller sample sizes for various
+groups, which skewed the average calculation for each group significantly. For example, the average student debt for subjects with no high school degree was almost 83\% off from the estimate
+using the base table imputation, but if we look at the underlying population the former was calculated with 6 people, while the latter was with 258 people.
+\todo{We have extended ImputeDB to provide users with a estimate of the number of tuples dropped in a calculation due to nulls, which may be indicative of the underlying error or skew
+in the results presented. This is a simple heuristic but we find that it would have provided appropriate warning in cases such as Q8.}
+}
 

### 2017-04-27 rebuttal.tex
added a section of implementation components

@@
 \newcommand{\todo}[1]{\color{red}{\textit{#1}}}
+\newcommand{\ProjectName}[0]{ImputeDB}
 
@@
 
-\resp{Existing systems have very complex formulas for estimating query costs. We felt that the overhead of reimplementing an existing cost model would not be worth the benefit.}
+\resp{
+Existing systems have very complex formulas for estimating query costs. We felt that the overhead of reimplementing an existing cost model would not be worth the benefit.
+However, we added a new section that discusses in detail the main components for \ProjectName{}, so that developers of existing systems can use this as a roadmap
+for implementation.
+}
 

### 2017-04-27 paper/implementation.tex
added a section of implementation components

\section{Implementation}
\todo{this is a preliminary version of this... clean up and trim language}
We discuss the main components required to implement an optimizer for dynamic query imputation. 
We use as reference a standard tuple-iterator prototype database.

\textbf{Extended histograms}: To plan dynamic imputation, we require estimates of the distribution of missing values in
the underlying tables. We found this information easy to store by extending our existing histograms to include the count
of missing values in each bucket across table attributes. On the base tables, this provides an exact count of the
missing fields. In intermediate planning stages, it allows us to use simple heuristics to estimate the number of missing 
values remaining after standard relational operations, such as selections. While statistics about missing value counts
could be kept separate from standard cardinality estimation histograms, cardinality estimation in the optimizer would
frequently need to adjust one based on the other, so its simplest to merge them.

\textbf{Dirty sets}: We keep track of the attributes that missing values but have not yet been imputed (using $\mu$ or $\delta$). We extend
the logical nodes in our planning data structures to carry a set of fully qualified attribute names representing the dirty set. This required 
small modifications in all standard relation operator nodes in our system.

\textbf{Additional Logical Operators}: While the $\mu$ and $\delta$ operations are not exposed directly to the user, they are a core extension
the the standard set of logical operator nodes used by a query planner. Both operators operate on a subset of attributes, determined by
a combination of the dirty set and planner-provided attributes, used to enumerate local imputation alternatives. Both operators, just like
selection or join nodes in the standard planner, can provide estimates of the new cardinality and do so by adjusting the appropriate histograms
handed off to the next operator in the plan.

\textbf{Core optimizer}: Imputations early in the plan (e.g. before a selection) could conceivably be added to existing optimizers. However, the full
planning used in \ProjectName{} is tightly integrated with core parts of the optimizer, such as join enumeration. This tight integration allows \ProjectName{}
to effectively prune out plans that are not in the relevant Pareto set. It also allows imputation costs to influence choose of join orders. For these reasons
we believe substituting the standard optimizer for a custom version tailored to dynamic imputation optimization is critical.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-04-28 rebuttal.tex
Add a couple points to rebuttal

@@
 
+\resp{We emphasize that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense by renaming it to \textit{Penalty}.}
+
 \item - S3.4.3: An extremely nice emergent property from the L-value function for the imputation operator seems to be that the operator will be moved to a position in the plan where it has access to the maximum training data.
@@
   % \resp{This is a reasonable concern.
-Optimizing for the maximum training set is flawed when the training data is skewed.
+Though often an appropriate proxy, optimizing for the maximum training set is flawed when the training data is skewed.
 We think that this problem is avoidable with a minor tweak to the loss function.
 Instead of optimizing for a large training set, we can optimize for a \emph{diverse} training set by tracking the
+% TODO also add an explanation of the coding of schooldegree -- and why it is filter >= 0,
+    % which seems unreasonable to an observer.
 
@@
 Why not just take the same approach as selection and treat it as having negligible cost?
+    \resp{We mistakenly suggest in S3 that we treat \textit{Drop} as a blocking operator. In
+        our implementation, we do endow it with $0$ cost, and we have updated the text to
+        reflect that.}
 
 \item - S4: All of the workload queries are (data-size-independent) AVG queries. What about other types of aggregates (e.g., SUM, COUNT)?
+
+    \resp{Though we do not show this explicitly, we repeat all of the queries using COUNT
+        aggregates in order to compute the ``Count Fraction''. The timing results are
+        clearly identical in this case and the number of tuples returned can be seen
+        directly in Table 4.}
+    \todo{For SUM queries, that may depend more directly on the number of rows returned than the
+        quality of imputations for any individual row, the effect of different COUNT results
+        dominates the computation AVG*COUNT. How do we handle this?}
 
@@
 At the very least, it would be useful to understand why there is so much error in $\alpha=0$ on Q2 and Q8.
+
+    \resp{For our purposes, imputation over the raw data is a reasonable ground truth. This
+        is because the analyst, in practice, should be considering the tradeoffs on quality
+        and performance of using ImputeDB versus a production imputation system on the raw
+        data. Unfortunately, they would never have access to a real ground truth in such a
+        scenario, and moreover, finding ground truth for use even in an empirical evaluation
+        of imputation methods can be challenging because artificially dirtying the data
+        often does not capture the complex reasons for missingness in real world datasets.
+        We update the discussion in S4.4.1 to emphasize this point.}
 

### 2017-04-28 paper/experiments.tex
paper: rework queries table to make it smaller
Add a couple points to rebuttal

@@
   \centering
-  \begin{subtable}{\linewidth}
-    \input{queries-cdc}
-    \caption{Queries on CDC data}\label{fig:queries-cdc}
-  \end{subtable}
-  \par\medskip
-  \begin{subtable}{\linewidth}
-    \input{queries-fcc}
-    \caption{Queries on freeCodeCamp data}\label{fig:queries-fcc}
-  \end{subtable}
-  \par\medskip
-  \caption{Queries used in our experiments.}\label{fig:queries}
+  \input{queries}
 \end{table*}
@@
 This measure indicates how much error is introduced by on-the-fly imputation compared to
-imputation on the base tables. In our case, the relevant comparison is against traditional
+imputation on the base tables. This comparison is the relevant one in practice --- an
+analyst would only be considering the trade-offs in imputation quality and time in the first
+place for dirty input data, for which ground truth is unavailable.
+
+In our case, the relevant comparison is against traditional
 imputation on the base table, given the unavailability of ground truth.
@@
 %%% Local Variables:
-
 %%% mode: latex

### 2017-05-02 paper/experiments.tex
prettify missing values tables

@@
 
-\begin{table}
+\begin{table*}
   \centering
-  \begin{subtable}{0.5\textwidth}
+  \begin{subtable}{0.3\textwidth}
     \centering
@@
     \caption{Demographics (\texttt{demo}). \demorows{} rows.}
-  \end{subtable}
-  \par\medskip
-  \begin{subtable}{0.5\textwidth}
+  \end{subtable}\quad\begin{subtable}{0.3\textwidth}
     \centering
@@
     \caption{Laboratory Results (\texttt{labs}). \labexrows{} rows.}
-  \end{subtable}
-  \par\medskip  
-  \begin{subtable}{0.5\textwidth}
+  \end{subtable}\quad\begin{subtable}{0.3\textwidth}
     \centering
@@
   \end{subtable}
-  \par\medskip  
-  \caption{Missing value distribution for each table/attribute in CDC NHANES 2013--2014 data.}\label{table:nhanes-description} 
-\end{table}
+  \caption{Percentage of values missing in the CDC NHANES 2013--2014 data.}\label{table:nhanes-description} 
+\end{table*}
 
@@
   \input{tables/fcc}
-  \caption{Missing value distribution for each attribute in freeCodeCamp Survey Data (\texttt{fcc})}\label{table:fcc-description} 
+  \caption{Percentage of values missing in the freeCodeCamp Survey data (\texttt{fcc}).}\label{table:fcc-description} 
 \end{table}
@@
 This measure indicates how much error is introduced by on-the-fly imputation compared to
-imputation on the base tables. This comparison is the relevant one in practice --- an
-analyst would only be considering the trade-offs in imputation quality and time in the first
-place for dirty input data, for which ground truth is unavailable.
-
-In our case, the relevant comparison is against traditional
+imputation on the base tables. In our case, the relevant comparison is against traditional
 imputation on the base table, given the unavailability of ground truth.

### 2017-05-02 rebuttal.tex
missed bracket in command
clean up algorithms

--- 2017-04-27 rebuttal.tex
@@
-\documentclass[draft,12pt]{article}
-\usepackage[margin=1in]{geometry}
+\documentclass{vldb}
 \usepackage{xcolor}
 \newcommand{\resp}[1]{{\color{blue}{#1}}}
-\newcommand{\todo}[1]{\color{red}{\textit{#1}}}
 
@@
 24\% error relative to direct imputation is huge.
-\resp{
-The outsized error in Q8 is a result of a drop operator placed very early on in the plan (a result of optimizing for performance). This produced significantly smaller sample sizes for various
-groups, which skewed the average calculation for each group significantly. For example, the average student debt for subjects with no high school degree was almost 83\% off from the estimate
-using the base table imputation, but if we look at the underlying population the former was calculated with 6 people, while the latter was with 258 people.
-\todo{We have extended ImputeDB to provide users with a estimate of the number of tuples dropped in a calculation due to nulls, which may be indicative of the underlying error or skew
-in the results presented. This is a simple heuristic but we find that it would have provided appropriate warning in cases such as Q8.}
-}
 
@@
 
+  \resp{We have clarified this in the paper.}
+
 \item - S3.4.3: An extremely nice emergent property from the L-value function for the imputation operator seems to be that the operator will be moved to a position in the plan where it has access to the maximum training data.
@@
 
-  \resp{Agreed. Our database stores tables in files, so our plans are I/O bound, but a database which uses an in-memory representation will need a different Time() function.}
+  \resp{Agreed.
+Our database stores tables in files, so our plans are I/O bound, but a database which uses an in-memory representation will need a different Time() function.
+We have clarified this in the paper.}
 
@@
 Why not just take the same approach as selection and treat it as having negligible cost?
+
+  \resp{This was an error in the paper. In the implementation, we do treat Drop as taking negligible time.}
 

### 2017-05-02 paper/algorithm.tex
paper: make figure smaller
Add a couple points to rebuttal
clean up algorithms

@@
 \begin{figure}
+  \begin{minipage}[c]{0.5\columnwidth}
     \Tree [.$\pi$ [.$g$ [.$\bowtie$ [.$\bowtie$ [.$\sigma$ $t_1$ ] [.$\sigma$ $t_2$ ] ] [.$\sigma$ $t_3$ ] ] ] ]
-\caption{Query plan schematic for the type of traditional plans explored (absent imputation operators).}
-\label{fig:query-schematic}
+  \end{minipage}\begin{minipage}[c]{0.5\columnwidth}
+    \caption{
+       Query plan schematic for the type of traditional plans explored (absent imputation operators).
+    } \label{fig:query-schematic}
+  \end{minipage}
 \end{figure}
@@
 \label{sec:cost-model}
-The cost of a plan $q$ is expressed as a tuple $\langle \textsc{Loss}(q), \textsc{Time}(q) \rangle$.
-$\textsc{Loss}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
+The cost of a plan $q$ is expressed as a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
+$\textsc{Penalty}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
 $\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
@@
 
-This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Loss}$ for the current query.
+This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.
 
-In order to pick a final plan from the frontier, our model introduces a trade-off parameter $\alpha$, which is an upper bound on the $\textsc{Loss}$ value that the user is willing
+In order to pick a final plan from the frontier, our model introduces a trade-off parameter $\alpha$, which is an upper bound on the $\textsc{Penalty}$ value that the user is willing
 to tolerate relative to the minimum possible amongst frontier plans.
@@
 \begin{definition}
-Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
+Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Penalty}(p) - \min_{p' \in P}(\textsc{Penalty}(p')) \leq \alpha$. We denote this as $p^\alpha$.
 We define the set of $\alpha$-bound plans in $P$ as $P^\alpha$.
@@
 \subsubsection{Cardinality estimation}
-The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ rely on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
+The computation of $\textsc{Time}(q)$ and $\textsc{Penalty}(q)$ rely on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
 These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
@@
 \subsubsection{Imputation quality}
-We use these histogram estimates to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.
+We estimate the quality of the imputations in a plan $Q$ using a heuristic measure $\text{Penalty}(Q)$, which is computed as follows.
 
@@
   \end{cases}\\
-  \textsc{Loss}(q) &= \frac{\textsc{L}(q)}{\textsc{N}(q)}
+  \textsc{Penalty}(q) &= \frac{\textsc{L}(q)}{\textsc{N}(q)}
 \end{align*}
 
-Each imputation operator incurs a heuristic loss penalty. We define \textsc{Loss} as the average over the individual penalties incurred
-by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties, and $N$ counts the number of imputation operations.
-For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
-might have encoded). Meanwhile, imputing \nullv{} fields with $\mu$ incurs a loss penalty $p \in (0, 1]$. In this case, the penalty is
-inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as
-the product of the number of attributes and the number of tuples. Intuitively, imputation
-accuracy \textit{increases} when more complete attributes and complete tuples are available;
-correspondingly, the information lost \textit{decreases} with more values to work
-with. We encode the fact that the number of training observations has diminishing returns for imputation quality,
-and find that scaling the cell count by applying square root works well in practice.
+Each imputation operator incurs an individual penalty.
+We define \textsc{Penalty} as the average over the penalties incurred by the imputation operators.
+Our recursive definition of $L$ accumulates these penalties and $N$ counts the number of imputation operations.
+For example, $\delta$ contributes a penalty of 1 (i.e. we lose all information  that dropped tuples might have encoded).
+Imputing \nullv{} fields with $\mu$ incurs a penalty $p \in (0, 1]$.
+In this case, the penalty is inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as the product of the number of attributes and the number of tuples.
+Intuitively, imputation accuracy \textit{increases} when more complete attributes and complete tuples are available; correspondingly, the penalty \textit{decreases} with more values to work with.
+Scaling the cell count by applying square root incorporates the intuition that the number of training observations has diminishing returns for imputation quality and we find that it works well in practice.
 
@@
 Scan costs are estimated based on the number of tuples, 
-page size, and a unit-less I/O cost per page. Join costs are estimated as a function
+page size, and a unit-less I/O cost per page. 
+We assume that I/O costs dominate CPU costs for relational operators. 
+A database implementation for which this assumption is not true, i.e. an in memory database, will need a different cost model.
+Join costs are estimated as a function
 of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
@@
 We extend these heuristics to 
-account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
-complexity can be expressed as a function of the number of heap pages read and the I/O cost
-per page. The time computation for \textit{Impute} depends on the properties of the underlying
+account for the new operators: \textit{Drop} and \textit{Impute}.
+\textit{Drop} is a special case of a filter, so its time complexity is correspondingly negligible.
+The time computation for \textit{Impute} depends on the properties of the underlying
 algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
@@
 
-The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Loss}, \textsc{Time} \rangle$ to
-collect sound and complete Pareto frontiers. For plans with the same dirty set, we can treat $\langle \textsc{Loss}, \textsc{Time} \rangle$
+The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Penalty}, \textsc{Time} \rangle$ to
+collect sound and complete Pareto frontiers. For plans with the same dirty set, we can treat $\langle \textsc{Penalty}, \textsc{Time} \rangle$
 as a total order, allowing us to appropriately select the frontier. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
@@
 \begin{algorithm}
-\input{algorithms/plan-helpers.tex}
-\end{algorithm}
-
-\begin{algorithm}
 \input{algorithms/plan-toplevel.tex}
@@
 can be plugged in with minimal effort and without changes to the optimizer, simply by adding
-time and loss functions (\Cref{sec:cost-model}) which describe the new algorithm.
+time and penalty functions (\Cref{sec:cost-model}) which describe the new algorithm.
 For our purposes, an imputation strategy is a function $I(R, C) \rightarrow R'$ which fills in the missing

### 2017-05-02 paper/algorithms/plan-toplevel.tex
clean up algorithms

--- 2017-02-23 paper/algorithms/plan-toplevel.tex
@@
-% \begin{algorithm}
+\newcommand{\InlineIte}[3]{#2\ \textbf{if}\ #1\ \textbf{else}\ #3}
+\renewcommand{\algorithmicindent}{1em}
+
 \begin{algorithmic}[1]
-  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses trade-off performance and information lost}
-  \Ensure{An optimized query plan.}
+    \Require{
+    \Statex
+    \begin{itemize}[leftmargin=*,noitemsep]
+    \item $q$: A query plan.
+    \item $C_l$: A set of attributes that must be imputed in the output of this query plan.
+    \item $C_g$: The set of attributes which are used in the final plan.
+    \end{itemize}}
+  \Function{Impute}{$q,\ C_l,\ C_g$}
+  \State $D_{must} \gets \Call{Dirty}{q} \cap C_l$
+  \State $D_{may} \gets D_{must} \cup (\Call{Dirty}{q} \cap C_g)$
+  \State $Q \gets (\InlineIte{D_{must} = \emptyset}{\{q\}}{\{\mu_{D_{must}}(q),\ \delta_{D_{must}}(q)\}})$
+  \State \Return $(\InlineIte{D_{may} = \emptyset}{Q}{Q \cup \{\mu_{D_{may}}(q), \delta_{D_{may}}(q) \}})$
+  \EndFunction
 
-  \Function{Plan}{$T, F, J, P, G, A$}
-  \State $C_g \gets \bigcup_{\phi \in F} \Call{Attr}{\phi}$ \Comment{Collect relevant attributes.}~\label{lst:line:attr-start}
-  \State $C_g \gets \bigcup_{\psi \in J} \Call{Attr}{\psi} \cup P \cup G \cup \Call{Attr}{A} \cup C_g$ \label{lst:line:attr-end}
+  \Statex
+
+  \Require{
+    \Statex
+    \begin{itemize}[leftmargin=*,noitemsep]
+    \item $T$: A set of tables.
+    \item $F$: A $T \times \Phi$ relation between tables and filter predicates.
+    \item $J$: A $T \times \Psi \times T$ relation between tables and join predicates.
+    \item $P$: A set of projection attributes.
+    \item $G$: A set of grouping attributes.
+    \item $A$: An aggregation function.
+    \item $\alpha$: A parameter in $[0, 1]$ that expresses the trade-off between performance and imputation quality.
+    \end{itemize}}
+  \Function{Plan}{$T,\ F,\ J,\ P,\ G,\ A,\ \alpha$}
+  \State $C_g \gets \bigcup_{\psi \in J} \Call{Attr}{\psi}$ \Comment{Collect relevant attributes.}~\label{lst:line:attr-start} 
+  \State $C_g \gets C_g\ \cup\ \bigcup_{\phi \in F} \Call{Attr}{\phi}\ \cup\ P \cup\ G\ \cup\ \Call{Attr}{A}$~\label{lst:line:attr-end}
+  \Statex
   \State Let $Q$ be an empty plan cache.
@@
   \If {$\exists \phi: (t, \phi) \in F$}
-  \State $Q[\{t\}] \lhd \Call{OptFilter}{Q, t, \phi, C_g}$~\label{lst:line:sel}
-  \Else
-  \State $Q[\{t\}] \lhd \{t\}$~\label{lst:line:scan}
+  \State $Q[\{t\}] \lhd \{\sigma_{\phi}(q)\ |\ q \in \Call{Impute}{t, \Call{Attrs}{\phi}, C_g}\}$~\label{lst:line:sel}
+  \Else\ $Q[\{t\}] \lhd \{t\}$~\label{lst:line:scan}
   \EndIf
   \EndFor
-
-  \State $\Call{OptJoin}{Q, T, J, C_g}$	\Comment{Optimize joins.}~\label{lst:line:join}
+  \Statex
+  \For{$size \in 2\dots|T|$} \Comment{Optimize joins.}~\label{lst:line:join}
+  \For{$S \in \{\text{all length}\ size\ \text{subsets of}\ T\}$}
+  \For{$t \in S$}
+  \For{$(t,\ \psi,\ t') \in J$ where $t' \in S \setminus t$}
+  \State $L \gets \{\Call{Impute}{q,\ \Call{Attrs}{\psi},\ C_g} ~|~ q \in Q[S \setminus t] \}$
+  \State $R \gets \{\Call{Impute}{q,\ \Call{Attrs}{\psi},\ C_g} ~|~ q \in Q[\{t\}] \}$
+  \State $Q[S] \lhd \{l \bowtie_\psi r ~|~ l \in L,\ r \in R\}$
+  \EndFor
+  \EndFor
+  \EndFor
+  \EndFor
+  \Statex
   \State $B \gets Q[T]$ \Comment{Get the best plans for all tables.}
   \If{$G \neq \emptyset$} \Comment{Add optional group \& aggregate.}~\label{lst:line:group}
-  \State $B \gets \bigcup_{q \in B} \Call{OptGroupBy}{q, G, A, C_g}$
+  \State $C_l \gets G \cup \Call{Attrs}{A}$
+  \State $B \gets \bigcup_{q \in B} \{g(q', G, A) ~|~ q' \in \Call{Impute}{q, C_l,\ C_g}\}$
   \EndIf
 
-  \State $B \gets \bigcup_{q \in B} \Call{OptProject}{q, P}$ \Comment{Add projections.}
+  \State $B \gets \bigcup_{q \in B} \{\pi_P(q') ~|~ q' \in \Call{Impute}{q,\ P,\ P}\}$ \Comment{Add projections.}
   \State \Return $p \in B$ s.t. $p$ is $\alpha$-bound optimal.
@@
 \end{algorithmic}
-\caption{Top-level query planner with imputations.}
+\caption{A query planner with imputations.}
 \label{algo:top-level-planner}
-% \end{algorithm}
 

### 2017-05-02 rebuttal.tex

--- 2017-04-28 rebuttal.tex
@@
-\documentclass[draft,12pt]{article}
-\usepackage[margin=1in]{geometry}
+\documentclass{vldb}
 \usepackage{xcolor}
 \newcommand{\resp}[1]{{\color{blue}{#1}}}
-\newcommand{\todo}[1]{\color{red}{\textit{#1}}}
+\newcommand{\todo}[1]{{\color{red}{\textit{#1}}}}
 \newcommand{\ProjectName}[0]{ImputeDB}
@@
 
+  \resp{We have clarified this in the paper.}
+
 \resp{We emphasize that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense by renaming it to \textit{Penalty}.}
@@
         reflect that.}
+
+  \resp{This was an error in the paper. In the implementation, we do treat Drop as taking negligible time.}
 

### 2017-05-05 simpledb/test/experiments/analyze_experiments.py
njoins -> ntables misnomer in generating workloads

@@ -154,8 +154,8 @@ def get_timing_results(experiments_dir, restrict_alpha=False, base=False, joins=
         df0 = pd.read_csv(f)
         if joins:
           # get number of joins
-          njoins = int(f.split(os.path.sep)[-4])
-          df0['njoins'] = njoins
+          ntables = int(f.split(os.path.sep)[-4])
+          df0['njoins'] = ntables - 1
         df = df.append(df0)
 
     return df

### 2017-05-05 simpledb/test/experiments/generate_big_joins.py

@@ -20,12 +20,12 @@ def join_query(tables):
 def self_join(tbl, n):
   return join_query([tbl] * n)
 
-def create_join_workload(n_joins, n_queries):
+def create_join_workload(n_tables, n_queries):
   random.seed(1)
   workload = []
   # self joins of the given size
   for table in tables:
-    workload.append(self_join(table, n_joins))
+    workload.append(self_join(table, n_tables))
 
   # random joins of given size
   n_queries -= len(workload)

### 2017-05-05 rebuttal.tex
rebut: add detail on reformatting
added discussion of planning times to paper and rebuttal

@@
 
+      \resp{We have compacted the algorithms and tables.
+Note that Tables 1-2 show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs.
+We think that the URLs add value to the references, particularly for data sets.}
+
 \item - Additional space recovered from these sections could be used to discuss the choice of techniques for imputation, and the confidence estimation idea suggested in the future work section.
@@
 
+\resp{We explored the planning times for queries with increasing number of joins. The standard algorithm presented in the original version submitted handles
+up to 5 joins, with columns requiring imputation in all tables, with planning time of approximately one second. We added a small optimization to help reduce
+planning times for a larger number of joins. We maintain an approximation of the Pareto sets, rather than the exact set, by only adding non-dominated plans that
+are sufficiently different. This helps reduce some of the increases in planning time for a larger number of joins.}
+
 \item D10. The scenario you are considering, with multiple queries during exploration, begs for reuse of previous query results, as you also mention in the future work section.

### 2017-05-05 paper/algorithm.tex
paper: remove reference to missing algorithm
added discussion of planning times to paper and rebuttal

@@
 The worst case complexity only occurs if the dirty sets tracked are distinct through the entire planning phase.
-Planning times are discussed in \Cref{sec:results}.
+
+In order to support a large number of joins, with reasonable planning times, \ProjectName{} can maintain approximations of the Pareto sets,
+rather than exact sets. To do so, dominated plans are still removed based on new non-dominated plans, but new plans are only added
+if they are significantly different from any existing plan in the set. The distance between plans is determined based on the cost penalty and time
+values associated with a plan.  Planning times are discussed further in \Cref{sec:results}.
 

### 2017-05-05 paper/experiments.tex
added discussion of planning times to paper and rebuttal

@@
 
+\subsubsection{Planning times}
+\begin{table}
+\centering
+\input{tables/planning_summary_exact_pareto.tex}
+\caption{Planning times are sub-second for up to 4 joins. For 5+  joins planning times rise. Many practical queries
+can be formulated with a smaller number of joins, but to support more complex queries \ProjectName{} provides an
+optimization that maintains approximate Pareto sets for queries with many joins. This reduces planning time
+for queries with 7 joins by approximately 45\%, on average.}
+\label{table:planning-times}
+\end{table}
+
+We explored the practical implications of our planning algorithm by constructing a series of queries with increasing number of joins.
+We used the CDC tables and evaluated a combination of self-joins and joins across reference tables. Planning times for queries
+involving 1 to 5 joins, which although limited is a practical value for real world exploratory queries, take less than a second.
+For queries with 7 joins, planning time increases substantially to 29 seconds, on average. We introduced an optimization that reduces
+the plan space by maintaining only approximations of the Pareto set for queries involving a large number of joins.
+With this optimization, the same set of queries have an average planning time of 16 seconds (2 seconds SE).
+
 %\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

### 2017-05-05 paper/examples.tex

@@
 \begin{figure}
+  \begin{subfigure}{\columnwidth}
     \Tree
@@
   ]
-\vspace{0.5\baselineskip}
-\caption{A quality-optimized plan for the query in~\Cref{fig:example-query}.
-The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate, $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
+  \caption{A quality-optimized plan for the query in~\Cref{fig:example-query}.}
 \label{fig:quality-plan}
+\end{subfigure}
+  \begin{subfigure}{\columnwidth}
+\vspace{1em}
+\Tree
+  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
+    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
+      [.\colorbox{pink}{$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$}
+        [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
+          [.$\bowtie_{\text{demo.id} = \text{exams.id}}$
+            [.$\sigma_{\text{demo.gender} = 2}$ demo ]
+            [.$\sigma_{\text{exams.weight} \geq 120}$ [.\colorbox{pink}{$\delta_{\text{exams.weight}}$} exams ] ] ] labs ] ] ] ]
+\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}.}
+\label{fig:fast-plan}
+\end{subfigure}
+\vspace{1em}
+
+\caption{The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate, $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).
+The imputation operators are highlighted.}
 \end{figure}
@@
 the query.
-
-\begin{figure}
-\Tree
-  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
-    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
-      [.\colorbox{pink}{$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$}
-        [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
-          [.$\bowtie_{\text{demo.id} = \text{exams.id}}$
-            [.$\sigma_{\text{demo.gender} = 2}$ demo ]
-            [.$\sigma_{\text{exams.weight} \geq 120}$ [.\colorbox{pink}{$\delta_{\text{exams.weight}}$} exams ] ] ] labs ] ] ] ]
-\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}.The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate, $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
-\label{fig:fast-plan}
-\end{figure}
 

### 2017-05-06 rebuttal.tex
added plots of plan frontiers for queries 1 and 6 to paper and rebuttal

@@
 
-\resp{We explored the planning times for queries with increasing number of joins. The standard algorithm presented in the original version submitted handles
+\resp{
+The sensitivity of different values of $\alpha$ depends on the query itself. Performing a scan of values of $\alpha$ with small step sizes can return
+a lot of the same plans for execution. In lieu of this, we extended the results section to include a subsection that presents the final frontiers for
+two queries, one from each dataset used during evaluation. This frontier can be exposed to the user to help in their workflow.
+\\
+We explored the planning times for queries with increasing number of joins. The standard algorithm presented in the original version submitted handles
 up to 5 joins, with columns requiring imputation in all tables, with planning time of approximately one second. We added a small optimization to help reduce

### 2017-05-06 paper/experiments.tex
added plots of plan frontiers for queries 1 and 6 to paper and rebuttal

@@
 
+\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm.
+The diversity of plans highlights the possible tradeoffs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
+can be exposed to the user to help in guiding their workflow for exploring values of $\alpha$.
+
 \begin{table}
@@
 \end{table}
+
+\begin{figure}
+\includegraphics[width=\columnwidth]{figures/pareto_frontiers_plot.png}
+\caption{\ProjectName{}'s final Pareto frontier for \Cref{q1} and \Cref{q6}. The user can tradeoff between the plans by adjusting values of $\alpha$. Depending on the query
+there may be small clusters of plans centered around a similar penalty or time value. The frontier is produced before executing a plan and can be exposed
+to the user.}
+\label{fig:pareto-frontiers}
+\end{figure}
 

### 2017-05-06 simpledb/src/java/simpledb/ImputedPlanCachePareto.java
be explicit to lookup empty dirty set when choosing final plan; same result but makes intent clear

@@ -72,7 +72,8 @@ public class ImputedPlanCachePareto extends AImputedPlanCache {
         double minLoss = getMinLoss(tables);
         double minTime = Double.MAX_VALUE;
         ImputedPlan chosen = null;
-        for (ImputedPlanCachePareto.Value val : bestPlans(tables)) {
+        // be explicit about fetching plan for all tables with empty dirty set
+        for (ImputedPlanCachePareto.Value val : bestPlans(tables, new HashSet<>())) {
             double time = val.plan.getTime();
             double loss = val.plan.getLoss();
             if ((loss - minLoss) <= lossBound && time < minTime) {

### 2017-05-08 paper/examples.tex
small paragraph on time-bound approach

@@
 
+An alternative strategy involves starting with $\alpha=1$ and iterating to lower values, as long as execution
+of the query completes within some user-defined time bound. \ProjectName{}'s planning
+algorithm can expose the final set of plans that tradeoff execution time and imputation quality, which allows
+the user to adjust their choice of $\alpha$ to uniquely select amongst these plans.
+
 Tens of queries later, our epidemiologist has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.

### 2017-05-08 paper/implementation.tex
tightened up language in implementation section

@@
 \section{Implementation}
-\todo{this is a preliminary version of this... clean up and trim language}
-We discuss the main components required to implement an optimizer for dynamic query imputation. 
-We use as reference a standard tuple-iterator prototype database.
+We discuss the main components required to implement dynamic query imputation.
 
-\textbf{Extended histograms}: To plan dynamic imputation, we require estimates of the distribution of missing values in
-the underlying tables. We found this information easy to store by extending our existing histograms to include the count
-of missing values in each bucket across table attributes. On the base tables, this provides an exact count of the
-missing fields. In intermediate planning stages, it allows us to use simple heuristics to estimate the number of missing 
-values remaining after standard relational operations, such as selections. While statistics about missing value counts
-could be kept separate from standard cardinality estimation histograms, cardinality estimation in the optimizer would
-frequently need to adjust one based on the other, so its simplest to merge them.
+\textbf{Extended histograms}: Planning imputation requires estimates of the distribution of missing values. 
+We extended standard histograms to include the count of missing values across table attributes.
+On the base tables, this provides an exact count of the missing fields. 
+In sub-plans, it allows us to use simple heuristics to estimate the number of missing values after applying
+both standard and imputation operators. These cardinality estimates are critical to the optimizer.
 
-\textbf{Dirty sets}: We keep track of the attributes that missing values but have not yet been imputed (using $\mu$ or $\delta$). We extend
-the logical nodes in our planning data structures to carry a set of fully qualified attribute names representing the dirty set. This required 
-small modifications in all standard relation operator nodes in our system.
+\textbf{Dirty sets}: We over-approximate the attributes in a plan that may have missing values and need to be
+imputed (using $\mu$ or $\delta$). We extended
+the logical nodes in our planning data structures to carry a set of fully qualified attribute names as the dirty set.
 
 \textbf{Additional Logical Operators}: While the $\mu$ and $\delta$ operations are not exposed directly to the user, they are a core extension
-the the standard set of logical operator nodes used by a query planner. Both operators operate on a subset of attributes, determined by
-a combination of the dirty set and planner-provided attributes, used to enumerate local imputation alternatives. Both operators, just like
-selection or join nodes in the standard planner, can provide estimates of the new cardinality and do so by adjusting the appropriate histograms
-handed off to the next operator in the plan.
+the the standard logical operator nodes in the planner.
+Both operate on a subset of attributes, determined by
+a combination of the dirty set and planner-provided attributes. These are used to enumerate local imputation alternatives. Both operators
+can compute cardinality and adjust histogram estimates. In contrast to standard operators, they also provide $\textsc{Penalty}$ and $\textsc{Time}$ estimates (\Cref{sec:cost-model}),
+based on the underlying imputation algorithm and the data input.
 
-\textbf{Core optimizer}: Imputations early in the plan (e.g. before a selection) could conceivably be added to existing optimizers. However, the full
-planning used in \ProjectName{} is tightly integrated with core parts of the optimizer, such as join enumeration. This tight integration allows \ProjectName{}
-to effectively prune out plans that are not in the relevant Pareto set. It also allows imputation costs to influence choose of join orders. For these reasons
-we believe substituting the standard optimizer for a custom version tailored to dynamic imputation optimization is critical.
+\textbf{Core optimizer}: Imputations early in the plan (e.g. before a selection) could be added to existing optimizers. However, the full
+planning algorithm used in \ProjectName{} is tightly integrated with core parts of the optimizer, such as join enumeration. This tight integration allows \ProjectName{}
+to influence join and imputation order, through the imputation cost model, and effectively prune out plans that are not in the relevant Pareto set.
+Substituting the standard optimizer for a custom version tailored to dynamic imputation optimization is critical to obtain the
+performance provided by \ProjectName{}.
 

### 2017-05-09 paper/algorithm.tex
paper: remove input section in "query planning"

@@
 \label{sec:planning}
-The input to our query planner is a tuple $(T, F, J, P, G, A)$:
-\begin{itemize}
-\item A set of tables $T$.
-\item A relation $F: T \times \Phi$ between tables and filter predicates.
-\item A relation $J : T \times \Psi \times T$ between tables and join predicates.
-\item A set of projection attributes $P$.
-\item An optional set of grouping attributes $G$ and aggregator function $A$.
-\end{itemize}
-
 The query planner must select a join ordering in addition to placing imputation operators as described in~\Cref{sec:placement}.
@@
 \subsubsection{Planning algorithm}
-The planner (\Cref{algo:top-level-planner}, with helper functions in \Cref{algo:plan-helpers}) operates as follows.
+The planner (\Cref{algo:top-level-planner}) operates as follows.
 First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}--\Cref{lst:line:attr-end}).

### 2017-05-09 paper/experiments.tex
reworded caption to make clear that planning times shown don't gave optimization, just referenced in text
paper: remove implementation subsection of experiments

@@
 base-table imputation and produces results of comparable quality.
-
-\subsection{Implementation}
-To evaluate the performance of our approach, we implemented a prototype system in Java, following a traditional iterator model.
-Our database handles a subset of standard SQL, including filters, joins, projection, grouping and aggregation.
-For simplicity, \ProjectName{} only handles tables containing integer data.
 
@@
 can be formulated with a smaller number of joins, but to support more complex queries \ProjectName{} provides an
-optimization that maintains approximate Pareto sets for queries with many joins. This reduces planning time
-for queries with 7 joins by approximately 45\%, on average.}
+optimization that maintains approximate Pareto sets for queries with many joins. This reduces the 29 second
+planning time shown for 7 joins by approximately 45\%, on average.}
 \label{table:planning-times}

### 2017-05-10 simpledb/src/java/simpledb/Drop.java
impute constructor back to DbIterator and take ImputedPlan as argument to method

@@ -10,10 +10,10 @@ public class Drop extends Impute {
      * Deal with missing data on a set of column by simply dropping the rows
      * with missing data.
      * @param dropFields set of columns to consider
-     * @param subplan
+     * @param child
      */
-    public Drop(Collection<String> dropFields, ImputedPlan subplan) {
-        super(dropFields, subplan);
+    public Drop(Collection<String> dropFields, DbIterator child) {
+        super(dropFields, child);
     }
 
     @Override
@@ -40,12 +40,12 @@ public class Drop extends Impute {
     }
 
 	@Override
-	public double getEstimatedTime() {
+	public double getEstimatedTime(ImputedPlan subplan) {
 		return 0.01 * subplan.cardinality();
 	}
 
 	@Override
-    public double getEstimatedPenalty() {
+    public double getEstimatedPenalty(ImputedPlan subplan) {
         return 1.0;
     }
 }

### 2017-05-10 simpledb/src/java/simpledb/Impute.java

@@ -8,15 +8,13 @@ public abstract class Impute extends Operator {
 
     private static final long serialVersionUID = 1L;
 
-    protected ImputedPlan subplan;
     protected DbIterator child;
     protected TupleDesc td;
     protected final Collection<String> dropFields;
     protected final Collection<Integer> dropFieldsIndices;
 
-    public Impute(Collection<String> dropFields, ImputedPlan subplan){
-        this.subplan = subplan;
-        this.child = subplan.getPlan();
+    public Impute(Collection<String> dropFields, DbIterator child){
+        this.child = child;
         this.td = child.getTupleDesc();
         
         // Validate drop fields and convert to indices.
@@ -90,9 +88,9 @@ public abstract class Impute extends Operator {
      * imputation. This is a unitless quantity
      * @return estimated time
      */
-	public abstract double getEstimatedTime();
+	public abstract double getEstimatedTime(ImputedPlan subplan);
 
-    public abstract double getEstimatedPenalty();
+    public abstract double getEstimatedPenalty(ImputedPlan subplan);
 
 
 }

### 2017-05-10 simpledb/src/java/simpledb/ImputeRandom.java

@@ -20,18 +20,18 @@ public class ImputeRandom extends Impute {
      * non-missing elements of the same column. Thus, this method requires that
      * we store all of the child tuples in a buffer until the child is
      * exhausted. Child is the physical plan obtained from subplan.
-     * @param subplan
+     * @param child
      */
-    public ImputeRandom(Collection<String> dropFields, ImputedPlan subplan){
-        super(dropFields, subplan);
+    public ImputeRandom(Collection<String> dropFields, DbIterator child){
+        super(dropFields, child);
         initRng();
         
         buffer = new ArrayList<>();
         nextTupleIndex = 0;
     }
 
-    public ImputeRandom(ImputedPlan subplan) {
-        this(null, subplan);
+    public ImputeRandom(DbIterator child) {
+        this(null, child);
     }
     
     private void initRng(){
@@ -109,7 +109,7 @@ public class ImputeRandom extends Impute {
      * @see simpledb.Impute#getEstimatedTime()
      */
 	@Override
-	public double getEstimatedTime() {
+	public double getEstimatedTime(ImputedPlan subplan) {
         int numTuples = (int) subplan.cardinality();
         int numDirty= subplan.getDirtySet().size(); // number of dirty attributes
 		double T = numTuples * numDirty;
@@ -118,7 +118,7 @@ public class ImputeRandom extends Impute {
 	}
 
 	@Override
-    public double getEstimatedPenalty() {
+    public double getEstimatedPenalty(ImputedPlan subplan) {
         return 0.5;
     }
 

### 2017-05-10 simpledb/src/java/simpledb/ImputeRegressionTree.java

@@ -51,8 +51,8 @@ public class ImputeRegressionTree extends Impute {
 	// Map from indices in the Instances object to indices in the Tuple buffer.
 	private HashMap<Integer, Integer> dropFieldsIndicesMap;
 
-    public ImputeRegressionTree(Collection<String> dropFields, ImputedPlan subplan) {
-        super(dropFields, subplan);
+    public ImputeRegressionTree(Collection<String> dropFields, DbIterator child) {
+        super(dropFields, child);
         setNumImputationEpochs();
         initRng();
         buffer = new ArrayList<>();
@@ -277,11 +277,11 @@ public class ImputeRegressionTree extends Impute {
      * @see simpledb.Impute#getEstimatedTime()
      */
 	@Override
-	public double getEstimatedTime() {
+	public double getEstimatedTime(ImputedPlan subplan) {
 		TupleDesc schema = subplan.getPlan().getTupleDesc();
 		Set<QualifiedName> dirytSet = subplan.getDirtySet();
 		int m_c = schema.numFields() - dirytSet.size(); // number of clean attributes
-		int m_i = dropFieldsIndices.size(); // number of dirty attributes
+		int m_i = dropFieldsIndices.size(); // number of attributes to be imputed
 		int n = (int) subplan.cardinality(); // number of tuples
 		int k = numImputationEpochs;
 		double T = k * m_i * n * (m_c + m_i - 1) * Math.log(n);
@@ -290,7 +290,7 @@ public class ImputeRegressionTree extends Impute {
 	}
 
 	@Override
-	public double getEstimatedPenalty() {
+	public double getEstimatedPenalty(ImputedPlan subplan) {
 		TupleDesc schema = subplan.getPlan().getTupleDesc();
 		double totalData = subplan.cardinality() * schema.numFields();
 		return (1 / Math.sqrt(totalData));

### 2017-05-10 simpledb/src/java/simpledb/ImputeTotallyRandom.java

@@ -16,10 +16,10 @@ public class ImputeTotallyRandom extends Impute {
      * are drawn from a uniform distribution on [Integer.MIN_VALUE,
      * Integer.MAX_VALUE] and String fields are drawn from a uniform
      * distribution on ["a"*maxSize, "z"*maxSize].
-     * @param subplan
+     * @param child
      */
-    public ImputeTotallyRandom(Collection<String> dropFields, ImputedPlan subplan){
-        super(dropFields, subplan);
+    public ImputeTotallyRandom(Collection<String> dropFields, DbIterator child){
+        super(dropFields, child);
         initRng();
     }
     
@@ -99,12 +99,12 @@ public class ImputeTotallyRandom extends Impute {
     }
 
 	@Override
-	public double getEstimatedTime() {
+	public double getEstimatedTime(ImputedPlan subplan) {
 		return 0;
 	}
 
 	@Override
-    public double getEstimatedPenalty() {
+    public double getEstimatedPenalty(ImputedPlan subplan) {
         return 0.5;
     }
 

### 2017-05-10 simpledb/src/java/simpledb/LogicalComposeImputation.java

@@ -56,17 +56,17 @@ public class LogicalComposeImputation extends ImputedPlan {
 
 		switch (imp) {
 		case DROP: {
-			Impute dropOp = new Drop(toNames(impute), subplan);
-			final double loss = dropOp.getEstimatedPenalty();
-			final double time = dropOp.getEstimatedTime();
+			Impute dropOp = new Drop(toNames(impute), subplan.getPlan());
+			final double loss = dropOp.getEstimatedPenalty(subplan);
+			final double time = dropOp.getEstimatedTime(subplan);
 			final TableStats adjustedTableStats = subplanTableStats.adjustForImpute(DROP, imputeIndices);
 			return new LogicalComposeImputation(adjustedTableStats, dropOp, subplan, dirtySet, loss, time);
 		}
 		case MAXIMAL:
 		case MINIMAL: 
-			Impute imputeOp = new ImputeRegressionTree(toNames(impute), subplan);
-			final double loss = imputeOp.getEstimatedPenalty();
-			final double time = imputeOp.getEstimatedTime();
+			Impute imputeOp = new ImputeRegressionTree(toNames(impute), subplan.getPlan());
+			final double loss = imputeOp.getEstimatedPenalty(subplan);
+			final double time = imputeOp.getEstimatedTime(subplan);
 			final TableStats adjustedTableStats = subplanTableStats.adjustForImpute(MAXIMAL, imputeIndices);
 			return new LogicalComposeImputation(adjustedTableStats, imputeOp, subplan, dirtySet, loss, time);
 		case NONE:

### 2017-05-10 simpledb/src/java/simpledb/AImputedPlanCache.java
most loss renamed to penalty

@@ -61,18 +61,18 @@ public abstract class AImputedPlanCache implements ImputedPlanCache {
 		}
 
 		public boolean dominates(Value o) {
-			return plan.loss().get() <= o.plan.loss().get() && plan.time() <= o.plan.time();
+			return plan.penalty().get() <= o.plan.penalty().get() && plan.time() <= o.plan.time();
 		}
 
 		public boolean isApproximate(Value o) {
-			double lossRatio = o.plan.loss().get() / plan.loss().get();
+			double penaltyRatio = o.plan.penalty().get() / plan.penalty().get();
 			double timeRatio = o.plan.time() / plan.time();
-			return lossRatio >= 0.95 && lossRatio <= 1.05 && timeRatio >= 0.5 && timeRatio <= 1.5;
+			return penaltyRatio >= 0.95 && penaltyRatio <= 1.05 && timeRatio >= 0.5 && timeRatio <= 1.5;
 		}
 
 		@Override
 		public int compareTo(Value o) {
-			int lcmp = Double.compare(plan.loss().get(), o.plan.loss().get());
+			int lcmp = Double.compare(plan.penalty().get(), o.plan.penalty().get());
 			if (lcmp != 0) { return lcmp; }
 			return Double.compare(plan.time(), o.plan.time());
 		}

### 2017-05-10 simpledb/src/java/simpledb/ImputedPlan.java

@@ -15,7 +15,7 @@ import simpledb.Utility.AvgAgg;
  * Each ImputedPlan carries a copy of its table statistics. These are modified according to imputation/selection/joins.
  */
 public abstract class ImputedPlan {
-	protected abstract AvgAgg loss();
+	protected abstract AvgAgg penalty();
 	protected abstract double time();
 	
 	public abstract Set<QualifiedName> getDirtySet();
@@ -25,11 +25,11 @@ public abstract class ImputedPlan {
 
 	public double cost(double lossWeight) {
 		assert 0.0 <= lossWeight && lossWeight <= 1.0;
-		return lossWeight * loss().get() + (1 - lossWeight) * time();
+		return lossWeight * penalty().get() + (1 - lossWeight) * time();
 	}
 
-	public double getLoss() {
-		return loss().get();
+	public double getPenalty() {
+		return penalty().get();
 	}
 
 	public double getTime() {

### 2017-05-10 simpledb/src/java/simpledb/ImputedPlanCacheDottedWriter.java

@@ -30,7 +30,7 @@ public class ImputedPlanCacheDottedWriter implements ImputedPlanCacheWriter {
 			ImputedPlan plan = val.plan;
 			StringBuilder prefix = new StringBuilder();
 			prefix.append("queryplan_loss=");
-			prefix.append(plan.loss().get());
+			prefix.append(plan.penalty().get());
 			prefix.append("_time=");
 			prefix.append(plan.time());
 			prefix.append("_card=");

### 2017-05-10 simpledb/src/java/simpledb/LogicalComposeImputation.java

@@ -16,7 +16,7 @@ public class LogicalComposeImputation extends ImputedPlan {
 	private final DbIterator physicalPlan;
 	private final ImputedPlan subplan;
 	private final Set<QualifiedName> dirtySet;
-	private final double loss;
+	private final double penalty;
 	private final double time;
 	private TableStats tableStats;
 
@@ -25,15 +25,15 @@ public class LogicalComposeImputation extends ImputedPlan {
 	 * @param tableStats
 	 * @param physicalPlan
 	 * @param dirtySet
-	 * @param loss
+	 * @param penalty
 	 * @param time
 	 */
 	private LogicalComposeImputation(TableStats tableStats, DbIterator physicalPlan, ImputedPlan subplan, Set<QualifiedName> dirtySet,
-			double loss, double time) {
+			double penalty, double time) {
 		super();
 		this.physicalPlan = physicalPlan;
 		this.dirtySet = dirtySet;
-		this.loss = loss;
+		this.penalty = penalty;
 		this.time = time;
 		this.tableStats = tableStats;
 		this.subplan = subplan;
@@ -57,18 +57,18 @@ public class LogicalComposeImputation extends ImputedPlan {
 		switch (imp) {
 		case DROP: {
 			Impute dropOp = new Drop(toNames(impute), subplan.getPlan());
-			final double loss = dropOp.getEstimatedPenalty(subplan);
+			final double penalty = dropOp.getEstimatedPenalty(subplan);
 			final double time = dropOp.getEstimatedTime(subplan);
 			final TableStats adjustedTableStats = subplanTableStats.adjustForImpute(DROP, imputeIndices);
-			return new LogicalComposeImputation(adjustedTableStats, dropOp, subplan, dirtySet, loss, time);
+			return new LogicalComposeImputation(adjustedTableStats, dropOp, subplan, dirtySet, penalty, time);
 		}
 		case MAXIMAL:
 		case MINIMAL: 
 			Impute imputeOp = new ImputeRegressionTree(toNames(impute), subplan.getPlan());
-			final double loss = imputeOp.getEstimatedPenalty(subplan);
+			final double penalty = imputeOp.getEstimatedPenalty(subplan);
 			final double time = imputeOp.getEstimatedTime(subplan);
 			final TableStats adjustedTableStats = subplanTableStats.adjustForImpute(MAXIMAL, imputeIndices);
[…]

### 2017-05-10 simpledb/src/java/simpledb/LogicalImputedScanNode.java

@@ -38,7 +38,7 @@ public class LogicalImputedScanNode extends ImputedPlan {
 	}
 	
 	@Override
-	protected AvgAgg loss() {
+	protected AvgAgg penalty() {
 		return AvgAgg.EMPTY;
 	}
 

### 2017-05-10 paper/experiments.tex
pareto plot with representative plans, and caption explaining precision point

@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{figures/pareto_frontiers_plot.png}
+\centering
+\includegraphics[scale=0.5]{figures/pareto_frontiers_plot.png}
 \caption{\ProjectName{}'s final Pareto frontier for \Cref{q1} and \Cref{q6}. The user can tradeoff between the plans by adjusting values of $\alpha$. Depending on the query
-there may be small clusters of plans centered around a similar penalty or time value. The frontier is produced before executing a plan and can be exposed
-to the user.}
+there may be small clusters of plans centered around a similar penalty or time value. This may occur as no plan actually dominates another at a given level of precision.
+For clarity, we have removed from the plot plans that provide similar tradeoffs in either dimension. }
 \label{fig:pareto-frontiers}

### 2017-05-10 paper/pareto_plot.py

@@ -1,0 +1,26 @@
+import pandas as pd
+import matplotlib.pyplot as plt
+ANNOTATE = False
+
+df = pd.read_csv('tables/pareto_frontiers.csv')  
+df['norm_time'] = df.groupby('query')['time'].apply(lambda x: x / x.max())
+df['id'] = range(0, df.shape[0])
+interesting_queries = [1, 6] # 2 is also nice, but want one from each set of experiments
+# plans that are different mainly due to floating point precision, collapse
+collapse = [0, 1, 2, 4, 7] + [31, 33]
+fig, axes = plt.subplots(2, sharex=True, figsize=(4.5, 6))
+plt.subplots_adjust(hspace=0.2)
+for i, q in enumerate(interesting_queries):
+  data  = df[(df['query'] == q) & ~df['id'].isin(collapse)]
+  ax = axes[i]
+  data.plot(kind='scatter', x='norm_time', y = 'penalty', ax=ax, s=20)
+  if ANNOTATE:
+    for row in data.itertuples():
+      ax.annotate('%d' % row.id, (row.norm_time, row.penalty), (row.norm_time * 1.01, row.penalty * 1.01))
+  ax.set_title('Pareto Frontier for Query %d' % q)
+  ax.set_xlabel('Time estimate/Max(Time estimate)')
+  ax.set_ylabel('Penalty Estimate')
+plt.subplots_adjust(left=0.2)
+  
+fig.savefig('figures/pareto_frontiers_plot.png')
+

### 2017-05-10 paper/algorithm.tex
paper: move histogram algorithm closer to cardinality estimation
paper: start reworking quality + cost sections

@@
 \label{sec:cost-model}
-The cost of a plan $q$ is expressed as a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
+The cost of a plan $q$ is a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
 $\textsc{Penalty}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
@@
 the expense of performance). If $\alpha = 1.0$, then the optimal query should have the
-fastest runtime (at the expense of quality).
+lowest runtime (at the expense of quality).
 
 \subsubsection{Cardinality estimation}
-The computation of $\textsc{Time}(q)$ and $\textsc{Penalty}(q)$ rely on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
-These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
+The computation of $\textsc{Time}(q)$ and $\textsc{Penalty}(q)$ rely accurate cardinality estimates for the plan $q$ and its sub-plans.
+These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational algebra, but also by the imputation operators.
 For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
@@
 
+\begin{algorithm}
+\input{algorithms/histogram-transform.tex}
+\end{algorithm}
+
 \subsubsection{Imputation quality}
-We estimate the quality of the imputations in a plan $Q$ using a heuristic measure $\text{Penalty}(Q)$, which is computed as follows.
+We estimate the quality of the imputations in a plan $q$ using a heuristic measure $\text{Penalty}(q, P)$, which is computed as follows.
 
 \begin{align*}
-  \textsc{L}(q) &= \begin{cases}
+  \textsc{L}(q, f) &= \begin{cases}
      1 + \textsc{L}(q') & q = \delta_C(q') \\
-    \frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}} + \textsc{L}(q')  & q = \mu_C(q') \\
+     P(q') + \textsc{L}(q') & q = \mu_C(q') \\
      \textsc{L}(q_l') + \textsc{L}(q_r') & q = q_l' \bowtie_\psi q_r' \\
-    \textsc{L}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
+     \textsc{L}(q') & q = \sigma_\phi(q'),\ q = \pi_C(q') \\
      0 & q = \text{some table } t\\
   \end{cases}\\
-    \textsc{N}(q) &= \begin{cases}
-     1 + \textsc{N}(q') & q = \delta_C(q') \\
-     1 + \textsc{N}(q')  & q = \mu_C(q') \\
-    \textsc{N}(q_l') + \textsc{N}(q_r') & q = q_l' \bowtie_\psi q_r' \\
-    \textsc{N}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
-    0 & q = \text{some table } t\\
-  \end{cases}\\
-  \textsc{Penalty}(q) &= \frac{\textsc{L}(q)}{\textsc{N}(q)}
+  %   \textsc{N}(q) &= \begin{cases}
+  %    1 + \textsc{N}(q') & q = \delta_C(q') \\
+  %    1 + \textsc{N}(q')  & q = \mu_C(q') \\
+  %   \textsc{N}(q_l') + \textsc{N}(q_r') & q = q_l' \bowtie_\psi q_r' \\
+  %   \textsc{N}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
+  %   0 & q = \text{some table } t\\
+  % \end{cases}\\
+  \textsc{Penalty}(q, P) &= \frac{\textsc{L}(q, P)}{\textsc{L}(q, 1)}
 \end{align*}
 
-Each imputation operator incurs an individual penalty.
-We define \textsc{Penalty} as the average over the penalties incurred by the imputation operators.
-Our recursive definition of $L$ accumulates these penalties and $N$ counts the number of imputation operations.
-For example, $\delta$ contributes a penalty of 1 (i.e. we lose all information  that dropped tuples might have encoded).
-Imputing \nullv{} fields with $\mu$ incurs a penalty $p \in (0, 1]$.
-In this case, the penalty is inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as the product of the number of attributes and the number of tuples.
-Intuitively, imputation accuracy \textit{increases} when more complete attributes and complete tuples are available; correspondingly, the penalty \textit{decreases} with more values to work with.
-Scaling the cell count by applying square root incorporates the intuition that the number of training observations has diminishing returns for imputation quality and we find that it works well in practice.
+$\textsc{Penalty}(q, P)$ assigns a cost in $(0,1]$ to each imputation operator in the plan $q$ according to its estimated quality, then averages over the individual costs.
+For example, each occurrence of $\delta$ incurs a cost of 1 (i.e.\ we lose all information that dropped tuples might have encoded).
+Imputing \nullv{} fields with $\mu$ incurs a cost in $(0, 1]$ according to a cost function $P$, which is specific to the imputation method in use.
+
+The choice of $P$ depends on the properties of the imputation method being used.
+A well chosen $P$ should provide a reasonable estimate of the quality of imputation from the histogram describing the input data.
+From this histogram, $P$ can obtain the number of attributes in its input, its cardinality, and an estimate of its variance.
+Intuitively, imputation accuracy \textit{increases} when more complete attributes and complete tuples are available; correspondingly, the $P$ should \textit{decrease} as the number of values increases.
+$P$ should never decrease to zero, as this would mean that it can perform perfect imputation.
+It should reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
+In Section~\ref{sec:imputation}, we discuss choices for $P$ for several common imputation methods.
+
+% In this case, the penalty is inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as the product of the number of attributes and the number of tuples.
+% Intuitively, imputation accuracy \textit{increases} when more complete attributes and complete tuples are available; correspondingly, the penalty \textit{decreases} with more values to work with.
+% Scaling the cell count by applying square root incorporates the intuition that the number of training observations has diminishing returns for imputation quality and we find that it works well in practice.
 
 \subsubsection{Query runtime}
-To compute $\textsc{Time}(Q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard 
-database.
+To compute $\textsc{Time}(q, T)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard database.
 
@@
 
-Scan costs are estimated based on the number of tuples, 
-page size, and a unit-less I/O cost per page. 
+Scan costs are estimated based on the number of tuples, page size, and a unit-less I/O cost per page. 
 We assume that I/O costs dominate CPU costs for relational operators. 
-A database implementation for which this assumption is not true, i.e. an in memory database, will need a different cost model.
-Join costs are estimated as a function
-of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
-Filtering, projecting, and aggregating are all done in memory and have low computational overhead
-so we assume negligible time costs for those operations.
+A database implementation for which this assumption is not true, e.g. an in memory database, will need a different cost model.
+Join costs are estimated as a function of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
+Filtering, projecting, and aggregating are all done in memory and have low computational overhead so we assume negligible time costs for those operations.
 
-We extend these heuristics to 
-account for the new operators: \textit{Drop} and \textit{Impute}.
+We extend these heuristics to account for the new operators: \textit{Drop} and \textit{Impute}.
 \textit{Drop} is a special case of a filter, so its time complexity is correspondingly negligible. 
-The time computation for \textit{Impute} depends on the properties of the underlying
-algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
-and simply queries it for a time estimate. In the implementation of our system,
-we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
-is a function of the number of attributes with no missing data, number of attributes with
-missing data being imputed, number of tuples, and the number of iterations for the algorithm.
+As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on the properties of the underlying algorithm via a function $T$.
 
-In general, evaluating the time complexity for different \textit{Impute} operators is
-complicated, as the properties of
-different algorithms can vary significantly, the underlying I/O cost
-is not easily extracted (especially in the case that the entire set of tuples does not
-fit in memory), and the CPU cost dominates (in contrast to the other operators, which do
-not consider CPU cost explicitly). For example,~\cite{martin1995time} finds that the
-time complexity of the \textit{build tree} algorithm for one commonly-used class of
-decision trees is a function of the number of classes, several overhead constants, the
-parameterization of the partition and heuristic functions, and the \textit{arity} (the
-number of subsets considered for each split). The \textit{build tree} phase often
-does not even dominate computation, as post-processing steps like pruning, which are vital
-in achieving good performance, can have cost exponential in the height of the tree.
-Nonetheless, in practice, we find that a simple parameterization can be acceptable (i.e.
-yields intelligent query plans).
+The choice of $T$ depends on the imputation method in use, and as with $P$, there are several intuitions which guide the choice of $T$.
+First, $\textsc{Time}$ is computed in terms of the I/O cost per page so $T$ must return a value in these units.
+It can be unnatural to compute the cost of imputation in terms of the I/O cost, so it may be simpler to compute the per-tuple cost and then convert.
+Second, $T$ should \emph{increase} as the number of tuples passed to the imputation function increases and it should be proportional to the complexity of the imputation algorithm.
+As with $P$, the ordering between queries generated by $T$ is more important than the precision of its estimates.
+\todo{Micah, how did you deal with the IO/CPU cost problem?}
+
+% In the implementation of our system,
+% we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
+% is a function of the number of attributes with no missing data, number of attributes with
+% missing data being imputed, number of tuples, and the number of iterations for the algorithm.
+
+% In general, evaluating the time complexity for different \textit{Impute} operators is complicated, as the properties of different algorithms can vary significantly, the underlying I/O cost is not easily extracted (especially in the case that the entire set of tuples does not fit in memory), and the CPU cost dominates (in contrast to the other operators, which do not consider CPU cost explicitly).
+% For example,~\cite{martin1995time} finds that the time complexity of the \textit{build tree} algorithm for one commonly-used class of decision trees is a function of the number of classes, several overhead constants, the parameterization of the partition and heuristic functions, and the \textit{arity} (the number of subsets considered for each split).
+% The \textit{build tree} phase often does not even dominate computation, as post-processing steps like pruning, which are vital in achieving good performance, can have cost exponential in the height of the tree.
+% Nonetheless, in practice, we find that a simple parameterization can be acceptable (i.e.\ yields intelligent query plans).
+
+\subsection{Imputation strategies}
+\label{sec:imputation}
+
+\ProjectName{} is designed so that any imputation strategy
+can be plugged in with minimal effort and without changes to the optimizer, simply by adding
+time and penalty functions (\Cref{sec:cost-model}) which describe the new algorithm.
+For our purposes, an imputation strategy is a function $I(R, C) \rightarrow R'$ which fills in the missing
+attributes $C$ in a relation $R$, returning a new relation $R'$.
+The flexibility of this approach aligns with the spirit of query optimization, which aims to
+reconcile performance with a declarative interface to data manipulation.
+
+\subsubsection{Regression trees}
+
+New imputation algorithms can also be added. As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
+based on chained-equation decision trees.  Chained-equation
+imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
+regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
+a predictive model of one missing attribute, conditional on all complete attributes
+and all other missing attributes. Chained-equation decision trees algorithms are found to be
+effective~\cite{akande2015empirical} and are widely used in epidemiological
+domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
+implementation~\cite{witten2016data}.
+
+\[P(q) = \frac{1}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}\]
+
+The reference algorithm used by the \textit{Impute} operator proceeds by iteratively fitting
+decision trees for the target imputation attribute on a subset of the data. In each
+iteration, the missing values of a separate attribute are replaced with newly imputed values.
+In one {\it epoch} of the algorithm, each of the target imputation
+attributes is filled in with imputed values in a single iteration.
+With each epoch, the quality of the imputation improves as values progressively
+reflect more accurate relationships between attributes. The algorithm terminates when
+convergence is achieved (i.e.\ the imputed values do not change across epochs) or a fixed
+number of epochs is reached.
+
+These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
+process, in which multiple distinct copies of the complete data are generated and estimators
+are computed by averaging over the multiple datasets to form a single completed dataset~
+\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
+multiple imputation could incorporated as another avenue of assessing imputation confidence.
+
+\subsubsection{Mean}
+
+\todo{Fill this in.}
+
+\subsubsection{Hot-deck}
+
+\todo{Fill this in.}
+
+%\begin{algorithm}
+%    \input{algorithms/cart.tex}
+%\end{algorithm}
 
@@
 %\input{algorithms/histogram-helpers.tex}
-%\end{algorithm}
-
-\begin{algorithm}
-\input{algorithms/histogram-transform.tex}
-\end{algorithm}
-
-\subsection{Imputation Strategies}
-\label{sec:imputation}
-
-\ProjectName{} is designed so that any imputation strategy
-can be plugged in with minimal effort and without changes to the optimizer, simply by adding
-time and penalty functions (\Cref{sec:cost-model}) which describe the new algorithm.
-For our purposes, an imputation strategy is a function $I(R, C) \rightarrow R'$ which fills in the missing
-attributes $C$ in a relation $R$, returning a new relation $R'$.
-The flexibility of this approach aligns with the spirit of query optimization, which aims to
-reconcile performance with a declarative interface to data manipulation.
-
-New imputation algorithms can also be added. As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
-based on chained-equation decision trees.  Chained-equation
-imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
-regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
-a predictive model of one missing attribute, conditional on all complete attributes
-and all other missing attributes. Chained-equation decision trees algorithms are found to be
-effective~\cite{akande2015empirical} and are widely used in epidemiological
-domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
-implementation~\cite{witten2016data}.
-
-The reference algorithm used by the \textit{Impute} operator proceeds by iteratively fitting
-decision trees for the target imputation attribute on a subset of the data. In each
-iteration, the missing values of a separate attribute are replaced with newly imputed values.
-In one {\it epoch} of the algorithm, each of the target imputation
-attributes is filled in with imputed values in a single iteration.
-With each epoch, the quality of the imputation improves as values progressively
-reflect more accurate relationships between attributes. The algorithm terminates when
-convergence is achieved (i.e.\ the imputed values do not change across epochs) or a fixed
-number of epochs is reached.
-
-These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
-process, in which multiple distinct copies of the complete data are generated and estimators
-are computed by averaging over the multiple datasets to form a single completed dataset~
-\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
-multiple imputation could incorporated as another avenue of assessing imputation confidence.
-
-%\begin{algorithm}
-%    \input{algorithms/cart.tex}
 %\end{algorithm}

### 2017-05-10 paper/implementation.tex
paper: pass over implementation section

@@
 \section{Implementation}
-We discuss the main components required to implement dynamic query imputation.
+Adding imputation and our optimization algorithm to a standard SQL database requires modifications to several key database components.
+In this section we discuss the changes which must be made to implement dynamic imputation.
 
-\textbf{Extended histograms}: Planning imputation requires estimates of the distribution of missing values. 
-We extended standard histograms to include the count of missing values across table attributes.
-On the base tables, this provides an exact count of the missing fields. 
-In sub-plans, it allows us to use simple heuristics to estimate the number of missing values after applying
-both standard and imputation operators. These cardinality estimates are critical to the optimizer.
+\begin{itemize}
+\item \textbf{Extended histograms:}
+  Ranking query plans with imputation requires estimates of the distribution of missing values in the base tables and in the outputs of sub-plans.
+  We extended standard histograms to include the count of missing values for each attribute.
+  On the base tables, this provides an exact count of the missing values.
+  In sub-plans we use simple heuristics---as discussed in Section~\ref{sec:cardinal}---to estimate the number of missing values after applying both standard and imputation operators.
+  These cardinality estimates are critical to the optimizer's ability to compare the performance and imputation quality of different plans.
 
-\textbf{Dirty sets}: We over-approximate the attributes in a plan that may have missing values and need to be
-imputed (using $\mu$ or $\delta$). We extended
-the logical nodes in our planning data structures to carry a set of fully qualified attribute names as the dirty set.
+\item \textbf{Dirty sets:}
+  The planner needs to know which attributes in the output of a plan might contain missing values so it can insert the correct imputation operators.
+  To provide the planner with this information, we over-approximate the set of attributes in a plan that may have missing values.
+  Each plan has an associated \emph{dirty set}, described in Section~\ref{sec:placement}, which tracks these attributes.
 
-\textbf{Additional Logical Operators}: While the $\mu$ and $\delta$ operations are not exposed directly to the user, they are a core extension
-the the standard logical operator nodes in the planner.
-Both operate on a subset of attributes, determined by
-a combination of the dirty set and planner-provided attributes. These are used to enumerate local imputation alternatives. Both operators
-can compute cardinality and adjust histogram estimates. In contrast to standard operators, they also provide $\textsc{Penalty}$ and $\textsc{Time}$ estimates (\Cref{sec:cost-model}),
-based on the underlying imputation algorithm and the data input.
+\item \textbf{Imputation operators:}
+  We extend the set of logical operations available to the planner with the \emph{Impute} and \emph{Drop} operators.
+  The database must have implementations for both operators and be able to correctly place them while planning.
+  In addition to the normal heuristics for estimating query runtime, these operators have a cost function $\textsc{Penalty}$ (Section~\ref{sec:quality}) which estimates the quality of their output.
+  The planner must be able to optimize both cost functions and select an appropriate plan from the set of Pareto-optimal plans.
 
-\textbf{Core optimizer}: Imputations early in the plan (e.g. before a selection) could be added to existing optimizers. However, the full
-planning algorithm used in \ProjectName{} is tightly integrated with core parts of the optimizer, such as join enumeration. This tight integration allows \ProjectName{}
-to influence join and imputation order, through the imputation cost model, and effectively prune out plans that are not in the relevant Pareto set.
-Substituting the standard optimizer for a custom version tailored to dynamic imputation optimization is critical to obtain the
-performance provided by \ProjectName{}.
+  % While the \emph{Impute} and \emph{Drop} operations are not exposed directly to the user, they are a core extension the the standard logical operator nodes in the planner.
+  % Both operate on a subset of attributes, determined by a combination of the dirty set and planner-provided attributes. These are used to enumerate local imputation alternatives.
+  % Both operators can compute cardinality and adjust histogram estimates.
+  % In contrast to standard operators, they also provide $\textsc{Penalty}$ and $\textsc{Time}$ estimates (\Cref{sec:cost-model}), based on the underlying imputation algorithm and the data input.
+
+\item \textbf{Integrated optimizer:}
+  We believe that it would be relatively simple to extend an existing query optimizer to insert imputations immediately after scanning the base tables.
+  However, integrating the imputation placement rules into the optimizer allows us to explore a large space of query plans which contain imputations.
+  In particular, this tight integration allows us to jointly choose the most effective join order and imputation placement.
+  Our goal of integrating imputation placement into the optimizer requires the other modifications that we have discussed.
+\end{itemize}
 

### 2017-05-10 paper/algorithm.tex
paper: add section labels

@@
 \subsubsection{Cardinality estimation}
+\label{sec:cardinal}
 The computation of $\textsc{Time}(q)$ and $\textsc{Penalty}(q)$ rely accurate cardinality estimates for the plan $q$ and its sub-plans.
@@
 \subsubsection{Imputation quality}
+\label{sec:quality}
 We estimate the quality of the imputations in a plan $q$ using a heuristic measure $\text{Penalty}(q, P)$, which is computed as follows.

### 2017-05-10 paper/relatedwork.tex
paper: endless orphans
paper: add some discussion of trummer & koch paper

@@
 
+The planning algorithm that we present has similarities to the multi-objective query optimizer in~\cite{trummer2014}.
+Both algorithms handle plans which have multiple cost functions by maintaining sets of Pareto optimal plans and pruning plans which become dominated.
+In \ProjectName{}, costs are expected to grow monotonically as the size of the plan increases, which simplifies the optimization problem significantly.
+\cite{trummer2014} handles the more complex case of cost functions which are piecewise linear, but not necessarily monotonic.
+
 \subsection{Forecasting and Databases}

### 2017-05-10 paper/algorithm.tex

@@
 replace all \nullv{} values with non-\nullv{} values for attributes $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
-In our experiments, we implement \textit{Impute} using a chained-equation decision trees algorithm, but our system is agnostic to this choice.
+In our experiments, we implement \textit{Impute} using a general-purpose chained-equation decision trees algorithm, but our system is agnostic to this choice.
 \textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
@@
 The cost of a plan $q$ is a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
-$\textsc{Penalty}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
+$\textsc{Penalty}(q) \in [0,1]$ is a heuristic that attempts to quantify the amount of information lost by the imputation procedures used.
 $\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
@@
 \subsubsection{Pareto-optimal plans}
-Given a query, \ProjectName{} produces a final set of plans $P$ by only keeping plans that are not dominated by others in the search space.
-So for a search space $S$, \[P= \{ p ~|~ \not\exists p' \in S.\ p \neq p' \land \doms{p'}{p} \}.\]
+Given a query, \ProjectName{} produces a final set of query plans $Q$ by only keeping plans that are not dominated by others in the search space.
+So for a search space $S$, \[Q= \{ q ~|~ \not\exists q' \in S.\ q \neq q' \land \doms{q'}{q} \}.\]
 
 \begin{definition}
-We say that a plan $p$ with cost $\langle l, t \rangle$ \emph{dominates} a plan $p'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(p) = \textsc{Dirty}(p') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$.
-We denote this as $\doms{p}{p'}$.
+We say that a plan $q$ with cost $\langle l, t \rangle$ \emph{dominates} a plan $q'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(q) = \textsc{Dirty}(q') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$.
+We denote this as $\doms{q}{q'}$.
 \end{definition}
 
-This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.
+This set $Q$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.
 
@@
 \begin{definition}
-Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Penalty}(p) - \min_{p' \in P}(\textsc{Penalty}(p')) \leq \alpha$. We denote this as $p^\alpha$.
-We define the set of $\alpha$-bound plans in $P$ as $P^\alpha$.
-We say a plan $p^\alpha$ is $\alpha$-bound optimal if $\forall p' \in P^\alpha. \textsc{Time}(p) \leq \textsc{Time}(p')$. 
+Let $Q$ be a set of plans. For $q \in Q$, we say $q$ is $\alpha$-bound, if $(\textsc{Penalty}(q) - \min_{q' \in Q}(\textsc{Penalty}(q')) \leq \alpha$. We denote this as $q^\alpha$.
+We define the set of $\alpha$-bound plans in $Q$ as $Q^\alpha$.
+We say a plan $q^\alpha$ is $\alpha$-bound optimal if $\forall q' \in Q^\alpha. \textsc{Time}(q) \leq \textsc{Time}(q')$. 
 \end{definition}
 
-Given $\alpha$ and $P$, \ProjectName{} returns an $\alpha$-bound optimal plan in $P$.
+Given $\alpha$ and $Q$, \ProjectName{} returns an $\alpha$-bound optimal plan in $Q$.
 
@@
 \label{sec:cardinal}
-The computation of $\textsc{Time}(q)$ and $\textsc{Penalty}(q)$ rely accurate cardinality estimates for the plan $q$ and its sub-plans.
+The computation of $\textsc{Penalty}(q)$ and$\textsc{Time}(q)$ relies on accurate cardinality estimates for the plan $q$ and its sub-plans.
 These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational algebra, but also by the imputation operators.
@@
 
-$\textsc{Penalty}(q, P)$ assigns a cost in $(0,1]$ to each imputation operator in the plan $q$ according to its estimated quality, then averages over the individual costs.
-For example, each occurrence of $\delta$ incurs a cost of 1 (i.e.\ we lose all information that dropped tuples might have encoded).
-Imputing \nullv{} fields with $\mu$ incurs a cost in $(0, 1]$ according to a cost function $P$, which is specific to the imputation method in use.
+$\textsc{Penalty}(q, P)$ assigns a penalty in $(0,1]$ to each imputation operator in the
+plan $q$ according to its estimated quality, then averages over the individual penalties.
+For example, each occurrence of $\delta$ incurs a penalty of 1 (i.e.\ we lose all information that dropped tuples might have encoded).
+Imputing \nullv{} fields with $\mu$ incurs a penalty in $(0, 1]$ according to a cost function $P$, which is specific to the imputation method in use.
 
 The choice of $P$ depends on the properties of the imputation method being used.
-A well chosen $P$ should provide a reasonable estimate of the quality of imputation from the histogram describing the input data.
+A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the histogram describing the input data.
 From this histogram, $P$ can obtain the number of attributes in its input, its cardinality, and an estimate of its variance.

### 2017-05-10 paper/algorithm.tex
Clean up imputation quality notation and explanation

@@
 \label{sec:quality}
-We estimate the quality of the imputations in a plan $q$ using a heuristic measure $\text{Penalty}(q, P)$, which is computed as follows.
+We estimate the quality of the imputations in a plan $q$ using a heuristic measure $\text{Penalty}_{P}(q)$, which is computed as follows.
 
 \begin{align*}
-  \textsc{L}(q, f) &= \begin{cases}
+    &\textsc{L}_{P}(q) = \begin{cases}
      1 + \textsc{L}(q') & q = \delta_C(q') \\
@@
   \end{cases}\\
-  %   \textsc{N}(q) &= \begin{cases}
-  %    1 + \textsc{N}(q') & q = \delta_C(q') \\
-  %    1 + \textsc{N}(q')  & q = \mu_C(q') \\
-  %   \textsc{N}(q_l') + \textsc{N}(q_r') & q = q_l' \bowtie_\psi q_r' \\
-  %   \textsc{N}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
-  %   0 & q = \text{some table } t\\
-  % \end{cases}\\
-  \textsc{Penalty}(q, P) &= \frac{\textsc{L}(q, P)}{\textsc{L}(q, 1)}
+    &\textsc{Penalty}_{P}(q) = \frac{\textsc{L}_{P}(q)}{\textsc{L}_{P_1}(q)}
 \end{align*}
 
-$\textsc{Penalty}(q, P)$ assigns a penalty in $(0,1]$ to each imputation operator in the
-plan $q$ according to its estimated quality, then averages over the individual penalties.
-For example, each occurrence of $\delta$ incurs a penalty of 1 (i.e.\ we lose all information that dropped tuples might have encoded).
-Imputing \nullv{} fields with $\mu$ incurs a penalty in $(0, 1]$ according to a cost function $P$, which is specific to the imputation method in use.
+$\textsc{Penalty}_{P}(q)$ aggregates penalties from individual imputations to produce a measure of the quality of imputation for the entire plan.
+For imputation on a subplan $q'$, a penalty $P(q')$ in $(0,1]$ is assigned to $\mu(q')$
+and a penalty of $1$ is assigned to $\delta(q')$. Here, $P$ is an imputation-specific penalty function that estimates the heuristic penalty for imputing \nullv{} fields.
+The function $\textsc{L}_{P}(q')$ is then defined recursively to aggregate these penalties for the entire plan.
+The result is normalized by $\textsc{L}_{P_\delta}(q')$, which is the total penalty from choosing everywhere to drop tuples with \nullv{} values and equates to the count of imputations considered.
 
-The choice of $P$ depends on the properties of the imputation method being used.
-A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the histogram describing the input data.
-From this histogram, $P$ can obtain the number of attributes in its input, its cardinality, and an estimate of its variance.
-Intuitively, imputation accuracy \textit{increases} when more complete attributes and complete tuples are available; correspondingly, the $P$ should \textit{decrease} as the number of values increases.
-$P$ should never decrease to zero, as this would mean that it can perform perfect imputation.
-It should reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
+The form of $P$ is chosen to reflect the properties of the imputation method being used.
+A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of histograms describing the input data.
+From these histogram, $P$ can obtain the number of clean and dirty attributes in its input, their cardinality, and estimates of means, variances, and other statistics.
+Intuitively, imputation quality \textit{increases} when more complete attributes and complete tuples are available; correspondingly, $P(q')$ \textit{decreases} as the number of values increases.
+The heuristic penalty should also reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
+Finally, the heuristic penalty may increase as amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.
+
+Imputation quality $P(q')=0$ is unachievable, as this suggests the values of missing elements are known exactly.
+On the other hand, $P(q')=1$ occurs when all tuples with missing values are dropped, as this represents the worst case in terms of recovering information from these tuples.
 In Section~\ref{sec:imputation}, we discuss choices for $P$ for several common imputation methods.
-
-% In this case, the penalty is inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as the product of the number of attributes and the number of tuples.
-% Intuitively, imputation accuracy \textit{increases} when more complete attributes and complete tuples are available; correspondingly, the penalty \textit{decreases} with more values to work with.
-% Scaling the cell count by applying square root incorporates the intuition that the number of training observations has diminishing returns for imputation quality and we find that it works well in practice.
 

### 2017-05-11 rebuttal.tex
first stab at D2/3 reviewer 3 rebuttal

@@
 for implementation.
+
+\todo{More sophisticated cost formulas from exiting systems could further improve our time estimates for traditional operators. In this case, we focused on developing the heuristics
+relevant for the new operators, at the expense of simplifying implementation of common formulas, as we felt that was core to our contribution.}
 }
@@
 
-  \resp{In our intended usage, the user starts by running with $\alpha = 0$ and increases $\alpha$ if the query takes too long to complete.}
+  \resp{In our intended usage, the user starts by running with $\alpha = 0$ and increases $\alpha$ if the query takes too long to complete.
+  \todo{
+  We agree that having a way of providing an estimate on error would be interesting. Since error in this case relates to the imputation values
+  generated by the whole-table imputation (as there is no ground truth), this would likely need to be on a case-by-case basis for distinct
+  instantiations of the impute operator. 
+  }
+  }
 

### 2017-05-11 paper/algorithm.tex
algorithms pass, nitpicks

@@
 agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
-that while our algorithm is exponential in the number of joins (\Cref{sec:complexity}), in practice, planning times are not a concern.
+that while our algorithm is exponential in the number of joins (\Cref{sec:complexity}), in practice, planning times are not a concern for most queries.
 
@@
 The cost of a plan $q$ is a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
-$\textsc{Penalty}(q) \in [0,1]$ is a heuristic that attempts to quantify the amount of information lost by the imputation procedures used.
-$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.
+$\textsc{Penalty}(q) \in (0,1]$ is a heuristic that attempts to quantify the amount of information lost by the imputation procedure used.
+$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics, selectivity estimation of the query predicates, and the complexity of the
+impute operation.
 
@@
 The computation of $\textsc{Penalty}(q)$ and$\textsc{Time}(q)$ relies on accurate cardinality estimates for the plan $q$ and its sub-plans.
-These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational algebra, but also by the imputation operators.
+These estimates are impacted not just by filtering or joining, as in the traditional relational algebra, but also by the imputation operators.
 For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
@@
 
+\todo{Add a verbal explanation of the scaling steps}
 \begin{algorithm}

### 2017-05-11 paper/algorithm.tex
verbal explanation of histogram estimation

@@
 These estimates are impacted not just by filtering or joining, as in the traditional relational algebra, but also by the imputation operators.
-For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
 
@@
 When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan.
-For example it will redistribute the count of tuples with \nullv{} in an imputed field.
+
 \Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
+Drop operators always reduce the cardinality of the relation. Given $\delta_C$, each column $c \in C$ clears its count of tuples with missing values,
+which can yield distinct cardinality estimates across attributes. To provide consistent estimates, we take the lowest such estimate $e$ and
+scale all attribute histograms based on their distribution such that each provides an estimate of $e$. In the case of $\mu_C$, the cardinality estimate
+remains constant. The count of missing values across attributes is cleared and redistributed along the appropriate histograms based on their distribution.
+In the case of $\sigma$ and $\bowtie$ operations, we estimate the cardinality of the output relation and adjust histogram counts based on their distribution to sum to that estimate.
 
-\todo{Add a verbal explanation of the scaling steps}
 \begin{algorithm}

### 2017-05-11 paper/algorithm.tex
quick pass over api section

@@
 \label{sec:quality}
-We estimate the quality of the imputations in a plan $q$ using a heuristic measure $\text{Penalty}_{P}(q)$, which is computed as follows.
+Every instantiation of the impute-operator defines a custom penalty function $P$ for imputing \nullv{} fields.
+\ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\text{Penalty}_{P}(q)$, computed as follows.
 
@@
   \end{cases}\\
-    &\textsc{Penalty}_{P}(q) = \frac{\textsc{L}_{P}(q)}{\textsc{L}_{P_1}(q)}
+    &\textsc{Penalty}_{P}(q) = \frac{\textsc{L}_{P}(q)}{\textsc{L}_{P_\delta}(q)}
 \end{align*}
@@
 For imputation on a subplan $q'$, a penalty $P(q')$ in $(0,1]$ is assigned to $\mu(q')$
-and a penalty of $1$ is assigned to $\delta(q')$. Here, $P$ is an imputation-specific penalty function that estimates the heuristic penalty for imputing \nullv{} fields.
+and a penalty of $1$ is assigned to $\delta(q')$.
 The function $\textsc{L}_{P}(q')$ is then defined recursively to aggregate these penalties for the entire plan. 
-The result is normalized by $\textsc{L}_{P_\delta}(q')$, which is the total penalty from choosing everywhere to drop tuples with \nullv{} values and equates to the count of imputations considered.
+The result is normalized by $\textsc{L}_{P_\delta}(q')$, which is the total penalty for always dropping tuples with \nullv{} values and equates to the count of imputation operators considered.
 
-The form of $P$ is chosen to reflect the properties of the imputation method being used.
-A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of histograms describing the input data.
-From these histogram, $P$ can obtain the number of clean and dirty attributes in its input, their cardinality, and estimates of means, variances, and other statistics.
+\noindent{\textbf{Constructing $P$}}: The form of $P$ is chosen to reflect the properties of the imputation method being used.
+A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of plan histograms.
+From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations to summary statistics such as mean and variance
 Intuitively, imputation quality \textit{increases} when more complete attributes and complete tuples are available; correspondingly, $P(q')$ \textit{decreases} as the number of values increases.
@@
 \subsubsection{Query runtime}
-To compute $\textsc{Time}(q, T)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard database.
+To compute $\textsc{Time}_{T}(q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard database.
 
@@
 
-The choice of $T$ depends on the imputation method in use, and as with $P$, there are several intuitions which guide the choice of $T$.
+\noindent{\textbf{Constructing $T$}}: The choice of $T$ depends on the imputation method in use, and as with $P$, there are several intuitions which guide the choice of $T$.
 First, $\textsc{Time}$ is computed in terms of the I/O cost per page so $T$ must return a value in these units.
@@
 \todo{Fill this in.}
+\[P(q) = \frac{1}{n} (1 + \overline{Variance(q)}) \]
 

### 2017-05-11 paper/algorithm.tex
fixing some wrong comments in algo section

@@
 The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Penalty}, \textsc{Time} \rangle$ to
-collect sound and complete Pareto frontiers. For plans with the same dirty set, we can treat $\langle \textsc{Penalty}, \textsc{Time} \rangle$
-as a total order, allowing us to appropriately select the frontier. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
-Pareto frontier produced by the optimizer corresponds to the collection of plans associated with the empty dirty set.
+collect sound and complete Pareto frontiers. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
+Pareto frontier produced by the optimizer corresponds to the collection of plans that best tradeoff imputation quality and
+execution time.
 
@@
 
-In order to support a large number of joins, with reasonable planning times, \ProjectName{} can maintain approximations of the Pareto sets,
-rather than exact sets. To do so, dominated plans are still removed based on new non-dominated plans, but new plans are only added
-if they are significantly different from any existing plan in the set. The distance between plans is determined based on the cost penalty and time
-values associated with a plan.  Planning times are discussed further in \Cref{sec:results}.
+In order to support a large number of joins, with reasonable planning times, \ProjectName{} can maintain approximations of the Pareto sets. 
+A non-dominated plan can be approximated by an existing plan in the set, if the distance between them is smaller than some predefined bound.
+Approximated plans are not added to the frontier, pruning the search space further. 
+Planning times across both exact and approximate Pareto frontiers are discussed further in \Cref{sec:results}.
 

### 2017-05-11 paper/experiments.tex
removed redundant filter in a query and addressed in rebuttal
some more cleanup in implementation and experiments

--- 2017-05-09 paper/experiments.tex
@@
 
-Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}).
-\Cref{q1} calculates the average cuff size for individuals based on their income data, with a constraint on height.
-\Cref{q2} compares creatine levels for individuals with low, medium, and high incomes and above a certain weight.
-\Cref{q3} extracts the average blood lead levels for children under 6 years of age.
-\Cref{q4} calculates the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity. 
-\Cref{q5} calculates the average waist circumference for subjects above a certain height and weight.
-
-Our second set of queries is on the freeCodeCamp data (\Cref{fig:queries-fcc}).
-\Cref{q6} calculates the average income for higher-income survey participants, grouped by their bootcamp attendance.
-\Cref{q7} estimates the average commute time of women from the United States who participated.
-\Cref{q8} calculates the average amount of student debt based on school degree for survey participants who have student debt.
-\Cref{q9} joins the freeCodeCamp data with a table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
-The query calculates the average GDP per-capita by grouping 18+ year old participants who attended bootcamp versus those who did not.
+%Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}).
+%\Cref{q1} calculates the average cuff size for individuals based on their income data, with a constraint on height.
+%\Cref{q2} compares creatine levels for individuals with low, medium, and high incomes and above a certain weight.
+%\Cref{q3} extracts the average blood lead levels for children under 6 years of age.
+%\Cref{q4} calculates the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity. 
+%\Cref{q5} calculates the average waist circumference for subjects above a certain height and weight.
+%
+%Our second set of queries is on the freeCodeCamp data (\Cref{fig:queries-fcc}).
+%\Cref{q6} calculates the average income for higher-income survey participants, grouped by their bootcamp attendance.
+%\Cref{q7} estimates the average commute time of women from the United States who participated.
+%\Cref{q8} calculates the average amount of student debt based on school degree for survey participants who have student debt.
+%\Cref{q9} joins the freeCodeCamp data with a table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
+%The query calculates the average GDP per-capita by grouping 18+ year old participants who attended bootcamp versus those who did not.
 
@@
 \Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm.
-The diversity of plans highlights the possible tradeoffs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
-can be exposed to the user to help in guiding their workflow for exploring values of $\alpha$.
+and encapsulate the possible tradeoffs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
+can be exposed to the user to help in guiding their workflow.
 
@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{figures/pareto_frontiers_plot.png}
-\caption{\ProjectName{}'s final Pareto frontier for \Cref{q1} and \Cref{q6}. The user can tradeoff between the plans by adjusting values of $\alpha$. Depending on the query
-there may be small clusters of plans centered around a similar penalty or time value. The frontier is produced before executing a plan and can be exposed
-to the user.}
+\centering
+\includegraphics[scale=0.4]{figures/pareto_frontiers_plot.png}
+\caption{\ProjectName{}'s final Pareto frontier for \Cref{q1} and \Cref{q6}. 
+For clarity, we have removed from the plot plans that provide nearly indistinguishable tradeoffs in either dimension. }
 \label{fig:pareto-frontiers}

### 2017-05-11 paper/experiments.tex
sprinkled remaining todos, to make sure we don't miss

@@
 
-For each query, we evaluate three different configurations: \ProjectName{} optimizing for
+For each query, we evaluate \todo{update: three different configurations}: \ProjectName{} optimizing for
 quality ($\alpha=0$), \ProjectName{} optimizing for performance ($\alpha=1$), and imputing
@@
 \Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
-queries are an order-of-magnitude faster than imputing on the base tables. We can get
-another order-of-magnitude speedup when optimizing for performance. This performance
+queries are an \todo{update/validate: order-of-magnitude faster than imputing on the base tables. We can get
+another order-of-magnitude speedup when optimizing for performance}. This performance
 differential means it is feasible
@@
     imputation can provide another order-of-magnitude speedup at the expense of quality.}
-
+\todo{update with new figure}   
 \label{fig:runtimes}
@@
 
-In the median across a variety of queries and choices of $\alpha$, planning took
+\todo{update: In the median across a variety of queries and choices of $\alpha$, planning took
 \planningmediantime{} ms, 
 constituting \planningruntimepercent{} percent of total runtime. In all cases, the optimizer
-returned a query plan within \planningmaxtime{} ms, with times similar across values of $\alpha$.
+returned a query plan within \planningmaxtime{} ms, with times similar across values of $\alpha$.}
 
@@
 
+\todo{Distinguish between underlying imputation methods}
 For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
@@
 can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
-to the base-imputation approach is low---between \lowsmapealphazero{} and
-\highsmapealphaoneexacs{} percent---indicating that on-the-fly imputation produces similar
+to the base-imputation approach is low---\todo{update: between \lowsmapealphazero{} and
+\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces similar
 results to imputation at the base tables. Notably, the results with largest error correspond
@@
     A lower count share reflects more potential for errors.}
+\todo{update with new figure}
 \label{table:smape}
@@
 optimization that maintains approximate Pareto sets for queries with many joins. This reduces the 29 second
-planning time shown for 7 joins by approximately 45\%, on average.}
+planning time shown for 7 joins by approximately 45\%, on average.
+\todo{update with new table and values}
+}
 \label{table:planning-times}
@@
 involving 1 to 5 joins, which although limited is a practical value for real world exploratory queries, take less than a second.
-For queries with 7 joins, planning time increases substantially to 29 seconds, on average. We introduced an optimization that reduces
+\todo{update: For queries with 7 joins, planning time increases substantially to 29 seconds, on average. We introduced an optimization that reduces
 the plan space by maintaining only approximations of the Pareto set for queries involving a large number of joins.
-With this optimization, the same set of queries have an average planning time of 16 seconds (2 seconds SE).
+With this optimization, the same set of queries have an average planning time of 16 seconds (2 seconds SE).}
 

### 2017-05-11 paper/algorithm.tex
Update Imputation Strategies section

@@
 We assume that I/O costs dominate CPU costs for relational operators. 
-A database implementation for which this assumption is not true, e.g. an in memory database, will need a different cost model.
+A database implementation for which this assumption is not true, e.g. an in-memory database, will need a different cost model.
 Join costs are estimated as a function of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
@@
 \textit{Drop} is a special case of a filter, so its time complexity is correspondingly negligible.
-As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on the properties of the underlying algorithm via a function $T$.
 
-\noindent{\textbf{Constructing $T$}}: The choice of $T$ depends on the imputation method in use, and as with $P$, there are several intuitions which guide the choice of $T$.
-First, $\textsc{Time}$ is computed in terms of the I/O cost per page so $T$ must return a value in these units.
-It can be unnatural to compute the cost of imputation in terms of the I/O cost, so it may be simpler to compute the per-tuple cost and then convert.
-Second, $T$ should \emph{increase} as the number of tuples passed to the imputation function increases and it should be proportional to the complexity of the imputation algorithm.
-As with $P$, the ordering between queries generated by $T$ is more important than the precision of its estimates.
-\todo{Micah, how did you deal with the IO/CPU cost problem?}
+\noindent{\textbf{Constructing $T$}}: As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on
+the properties of the underlying algorithm via a function $T$. If we assume
+that the \textit{Impute} routine is CPU-bound, then we will adjust
+\textsc{Time} by some constant factor $f$, thereby scaling the result to be
+comparable with IO-bound routines.
+
+The form of $T$ depends on the specific imputation method, and as with $P$,
+there are some principles to capture the intuition.  $T$ should \emph{increase}
+as the number of tuples and complete and dirty attributes passed to the
+imputation function increases and it should be take into account the complexity
+of the imputation algorithm. Analysis of the time complexity of different
+imputation algorithms, especially with in-database execution, is beyond the
+scope of this paper. As with $P$, it is the ordering between queries
+generated by $T$ that is more important than the precision of estimates.
 
@@
 \ProjectName{} is designed so that any imputation strategy
-can be plugged in with minimal effort and without changes to the optimizer, simply by adding
-time and penalty functions (\Cref{sec:cost-model}) which describe the new algorithm.
-For our purposes, an imputation strategy is a function $I(R, C) \rightarrow R'$ which fills in the missing
-attributes $C$ in a relation $R$, returning a new relation $R'$.
-The flexibility of this approach aligns with the spirit of query optimization, which aims to
+can be plugged in with minimal effort and without changes to the optimizer, simply by supplying
+penalty and time functions (\Cref{sec:cost-model}) pertinent to the algorithm.
+These can then be queried by the optimizer during planning.  The flexibility of
+this approach aligns with the spirit of query optimization, which aims to
 reconcile performance with a declarative interface to data manipulation.
 
-\subsubsection{Regression trees}
+\subsubsection{Decision trees}
 
-New imputation algorithms can also be added. As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
-based on chained-equation decision trees.  Chained-equation
-imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
-regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
-a predictive model of one missing attribute, conditional on all complete attributes
-and all other missing attributes. Chained-equation decision trees algorithms are found to be
-effective~\cite{akande2015empirical} and are widely used in epidemiological
-domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
-implementation~\cite{witten2016data}.
-
-\[P(q) = \frac{1}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}\]
-
-The reference algorithm used by the \textit{Impute} operator proceeds by iteratively fitting
-decision trees for the target imputation attribute on a subset of the data. In each
-iteration, the missing values of a separate attribute are replaced with newly imputed values.
-In one {\it epoch} of the algorithm, each of the target imputation
-attributes is filled in with imputed values in a single iteration.
-With each epoch, the quality of the imputation improves as values progressively
-reflect more accurate relationships between attributes. The algorithm terminates when
-convergence is achieved (i.e.\ the imputed values do not change across epochs) or a fixed
-number of epochs is reached.
+New imputation algorithms can also be added. As a reference implementation,
+\ProjectName{} uses a general-purpose imputation strategy based on
+chained-equation decision trees.  Chained-equation imputation
+methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
+regression}~\cite{gelman2006data}) impute multiple missing attributes by
+iteratively fitting a predictive model of one missing attribute, conditional on
+all complete attributes and all other missing attributes. Chained-equation
+decision trees algorithms are found to be effective~\cite{akande2015empirical}
+and are widely used in epidemiological domains~\cite{burgette2010multiple}.
 
@@
 \cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
-multiple imputation could incorporated as another avenue of assessing imputation confidence.
+multiple imputation could be incorporated as another avenue of assessing imputation confidence.
+
+In the reference implementation, the chained-equation algorithm proceeds by
+iteratively fitting decision trees for the target imputation attributes. In
+each iteration, the missing values of a separate attribute are replaced with
+newly imputed values.  Values in a single column can be updated over multiple
+iterations of the algorithm. With more iterations, the quality of the
+imputation improves as values progressively reflect more accurate relationships
+between attributes. The algorithm terminates after a fixed number of
+iterations, or earlier if convergence is achieved. (That is, when the imputed
+values of one column are not updated by subsequent iterations that update that
+column.) For our experiments, we use Weka's \texttt{REPTree}
+implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
+algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
+decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
+and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
+$k * |\textsc{Dirty}(q)|$ trees, we have
+
+\begin{equation*}
+    \begin{aligned}
+    T(q) = k * &|\textsc{Dirty}(q)| * \textsc{Card}(q) * \\
+               &(|\textsc{Clean}(q)| + |\textsc{Dirty}(q)| - 1) * log(\textsc{Card}(q))
+    \end{aligned}
+\end{equation*}
+
+Separately, we specify a heuristic penalty that takes into account the
+improvement in imputation quality from more tuples and attributes as well as
+the diminishing returns to more data.
+
+\[
+    P(q) = \frac{1}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}
+\]
 

### 2017-05-11 paper/algorithm.tex
Add mean and hot-deck descriptions

@@
 \subsubsection{Query runtime}
+\label{sec:query-runtime}
+
 To compute $\textsc{Time}_{T}(q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard database.
@@
 this approach aligns with the spirit of query optimization, which aims to
-reconcile performance with a declarative interface to data manipulation.
+reconcile performance with a declarative interface to data manipulation. To
+motivate this API, we describe how an analyst would specify the form of these
+functions for a general-purpose imputation algorithm, as well as two simpler
+methods.
 
@@
 \[
-    P(q) = \frac{1}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}
+    P(q) = \frac{|\textsc{Dirty}(q)|}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}
 \]
@@
 
-\todo{Fill this in.}
-\[P(q) = \frac{1}{n} (1 + \overline{Variance(q)}) \]
+At the other end of the spectrum of complexity, an analyst could use a simpler
+method that replaces a missing value with the mean of the complete values of
+that attribute. Mean value imputation is a common choice for use in exploratory
+data analysis. One benefit of mean imputation in the context of \ProjectName{}
+is that taking advantage of the histograms associated with a plan $q$, we can
+estimate the mean and impute missing values in a non-blocking fashion. Thus,
+applying the heuristics from \ref{sec:quality} and \ref{sec:query-runtime}, we
+have
+
+\begin{equation*}
+\begin{aligned}
+    T(q) &\approx 0 \\
+    P(q) &= \frac{|\textsc{Dirty}(q)|}{\sqrt{\textsc{Card}(q)}} \\
+    P(q) &= \frac{1}{n} (1 + \overline{Variance(q)})
+\end{aligned}
+\end{equation*}
+>>>>>>> Add mean and hot-deck descriptions
 
@@
 
-\todo{Fill this in.}
+Hot-deck imputation is a nonparametric method in which missing values are
+replaced with randomly selected complete values. This method preserves the
+distribution of the complete values. The heuristic $T$ and $P$ functions are
+similar except for the fact that hot-deck is blocking in our implementation.
+
+\begin{equation*}
+\begin{aligned}
+    T(q) &= \textsc{Card}(q) \\
+    P(q) &= \frac{|\textsc{Dirty}(q)|}{\sqrt{\textsc{Card}(q)}}
+\end{aligned}
+\end{equation*}
 

### 2017-05-11 paper/experiments.tex
paper: pass over experiments

--- 2017-05-10 paper/experiments.tex
@@
 \subsection{Queries}
-We collect a set of queries (\Cref{fig:queries}) that we think are interesting to plan.
+We collect a set of queries (\Cref{tbl:queries}) that we think are interesting to plan.
 We believe that they could reasonably be written by a user in the course of data analysis.
@@
 
-Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}).
+The first set of queries is on the CDC data.
 \Cref{q1} calculates the average cuff size for individuals based on their income data, with a constraint on height.
@@
 
-Our second set of queries is on the freeCodeCamp data (\Cref{fig:queries-fcc}).
+The second set of queries is on the freeCodeCamp data.
 \Cref{q6} calculates the average income for higher-income survey participants, grouped by their bootcamp attendance.
@@
   \input{queries}
+  \vspace{1ex}
+  \caption{Queries used in our experiments.}\label{tbl:queries}
 \end{table*}
@@
 
-\Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
-queries are an order-of-magnitude faster than imputing on the base tables. We can get
+\Cref{fig:runtimes} shows the performance results. The quality-optimized
+queries are an order-of-magnitude faster than imputing on the base tables. We get
 another order-of-magnitude speedup when optimizing for performance. This performance
 differential means it is feasible
-to explore multiple imputations operations (including more expensive operators) when using
+to explore multiple imputation operations (including more expensive operators) when using
 \ProjectName{}, in contrast to the traditional approach of base table imputation.
@@
 %planning.
-
-In the median across a variety of queries and choices of $\alpha$, planning took
-\planningmediantime{} ms, 
-constituting \planningruntimepercent{} percent of total runtime. In all cases, the optimizer
-returned a query plan within \planningmaxtime{} ms, with times similar across values of $\alpha$.
 
@@
 
-We explored the practical implications of our planning algorithm by constructing a series of queries with increasing number of joins.
-We used the CDC tables and evaluated a combination of self-joins and joins across reference tables. Planning times for queries
-involving 1 to 5 joins, which although limited is a practical value for real world exploratory queries, take less than a second.
-For queries with 7 joins, planning time increases substantially to 29 seconds, on average. We introduced an optimization that reduces
-the plan space by maintaining only approximations of the Pareto set for queries involving a large number of joins.
+Planning times are reasonable, with a median planning time of \planningmediantime{} ms. over the benchmark queries, constituting \planningruntimepercent{}\% of total runtime.
+In all cases, the optimizer returned a query plan within \planningmaxtime{} ms.
+The choice of $\alpha$ has no impact on planning times, because query selection is done after collecting the Pareto frontier.
+
+However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
+We explored the limits our planning algorithm by constructing a series of queries with increasing number of joins.
+We used the CDC tables and evaluated a combination of self-joins and joins across reference tables.
+Planning times for queries involving 1--5 joins, which is a practical value for real world exploratory queries, take less than a second.
+For queries with 7 joins, planning time increases substantially to 29 seconds, on average.
+We introduced an optimization that reduces the plan space by maintaining an approximation of the Pareto set for queries involving a large number of joins.
 With this optimization, the same set of queries have an average planning time of 16 seconds (2 seconds SE).

### 2017-05-11 rebuttal.tex
removed redundant filter in a query and addressed in rebuttal
rebut: answer more questions

--- 2017-05-06 rebuttal.tex
@@
 
+  \resp{We agree that it could be interesting to allow drops to be omitted before filters.
+This would increase the plan space somewhat and might lead to better plans in some cases.
+For the conjunctive predicates in our experiments, our implementation splits the predicate into multiple filter nodes, so imputations or drops and filters can interleave.
+This should have the same effect, as the filter and the drop operator behave the same way on a tuple containing null.
+However, this optimization would also apply to disjunctive predicates, which cannot be split, and is worth definitely investigating.}
+
 \item - S2: Why no unions?
@@
 
-  % \resp{This is a reasonable concern.
-Though often an appropriate proxy, optimizing for the maximum training set is flawed when the training data is skewed.
-We think that this problem is avoidable with a minor tweak to the loss function.
-Instead of optimizing for a large training set, we can optimize for a \emph{diverse} training set by tracking the
-% TODO also add an explanation of the coding of schooldegree -- and why it is filter >= 0,
-    % which seems unreasonable to an observer.
+  \resp{This is a reasonable concern.
+Though often an appropriate proxy, optimizing for the maximum training set is flawed when the query requires the construction a large, skewed data set.
+The problem seems to be particular to queries which generate large intermediate tables.
+In some cases, as when joining two datasets, the intermediate table is a good target for training because it exposes correlations which would be hidden in the separate tables.
+In other cases, this intermediate table is large and skewed, which causes problems for imputation.
+Avoiding this case requires us to estimate the skew in these intermediate tables.}
+  \todo{Need a better explanation here.}
+\todo{Add an explanation of the coding of schooldegree -- and why it is filter $>=$ 0, which seems unreasonable to an observer.}
 
@@
 
+  \resp{Yes.
+For example, in time series data, it is reasonable to fill the most recent non-null value forward.
+This kind of imputation isn't useful for survey data, so we didn't consider it for our experiments.
+We have also extended the experiments section to consider a simple imputation method which fills missing values with the mean of the column, which only needs one pass over the data.}
+
 \item D9. Additional experiments I suggest to include:

### 2017-05-11 rebuttal.tex

@@
 In other cases, this intermediate table is large and skewed, which causes problems for imputation.
-Avoiding this case requires us to estimate the skew in these intermediate tables.}
+Avoiding this case requires us to estimate the skew in these intermediate tables.
+
+To clarify the point about Q8, we've removed the predicate condition $schooldegree \geq 0$, this was in fact a trivial satisfied predicate, given the way our dataset was set up.}
   \todo{Need a better explanation here.}
-\todo{Add an explanation of the coding of schooldegree -- and why it is filter $>=$ 0, which seems unreasonable to an observer.}
 

### 2017-05-11 paper/main.tex
added a section of implementation components
clean up algorithms
paper: move implementation section

@@
 \usepackage{xcolor}
-\usepackage{hyperref} % load after other packages
 \usepackage{tikz}
 \usetikzlibrary{positioning}
-\usepackage{cleveref}
 \usepackage{xspace}
 \usepackage[all]{nowidow}
+\usepackage{enumitem}
+\usepackage{hyperref} % load after other packages
+\usepackage{cleveref}
 
@@
   % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
-  Our experiments show that our query plans execute between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
+  Our experiments show that our query plans execute between \todo{update: \lowxalphazero{} and \highxalphazeroexacs{}} times faster than first imputing the base tables.
   Furthermore, we show that the query results from on-the-fly imputation differ from the
   traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
-  Finally, we show that while dropping tuples with missing values that fail query constraints discards 3--88\% of the data, on-the-fly imputation loses only 0--10\%.
+  Finally, we show that while dropping tuples with missing values that fail query constraints \todo{update: discards 3--88\% of the data, on-the-fly imputation loses only 0--10\%}.
 \end{abstract}
@@
 \input{algorithm}
+\input{implementation}
 \input{experiments}

### 2017-05-11 rebuttal.tex
rebut: pass over rebuttal

@@
 \begin{document}
+\section{Multiple reviewers}
+\begin{itemize}
+\item Redundant figures, excessive whitespace, formatting.
+
+  \resp{We've compressed many of the figures and removed whitespace to make room for additional material.}
+\end{itemize}
+
   \section{Reviewer 1}
   \begin{itemize}
-\item W1.
-A number of the algorithms/figures are a bit redundant (e.g., Histogram estimation for classical join operators in Alg.
-3) Overall, the paper in its current form feels padded -- it could be trimmed down by 1-2 pages.
-See below for a few ideas of things to use this space for.
-I will note that I'd be perfectly happy accepting this as a shorter, unpadded paper.
-
   \item W2.
@@
 (3) Given that the ML time estimator is being treated as a black box, it would be useful to see how sensitive the plan selector is to inaccurate estimates.
+    \todo{}
 
@@
 24\% error relative to direct imputation is huge.
+
     \resp{
@@
 
-\item - Although it doesn't consider cost or the drop operator, a similar technique appears in Yang et. al. "Lenses: An On-Demand Approach to ETL." A comparison would be nice.
+  \item Although it doesn't consider cost or the drop operator, a similar technique appears in Yang et. al. "Lenses: An On-Demand Approach to ETL." A comparison would be nice.
+    \todo{}
 
-\item - S1: "Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach — between 0 and 24 percent" 24 percent error is pretty bad.
+  \item S1: "Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach — between 0 and 24 percent" 24 percent error is pretty bad.
+    \todo{}
 
-\item - S2: The drop operator appears to be a weaker form of SQL's null-value / 3-valued Boolean semantics.
+  \item S2: The drop operator appears to be a weaker form of SQL's null-value / 3-valued Boolean semantics.
 The latter might actually be more useful.
@@
 
-\item - S2: Why no unions?
+  \item S2: Why no unions?
 
@@
 
-\item - S3.4.1, Definition 2: On my first read-through, I was puzzled about why there was no plan with zero-loss.
+  \item S3.4.1, Definition 2: On my first read-through, I was puzzled about why there was no plan with zero-loss.
 It would help to clarify that Loss is not about tuple loss, but rather a heuristic penalty function.
 
-  \resp{We have clarified this in the paper.}
+    \resp{We have clarified this in the paper by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and by renaming it to \textit{Penalty}.}
 
-\resp{We emphasize that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense by renaming it to \textit{Penalty}.}
-
-\item - S3.4.3: An extremely nice emergent property from the L-value function for the imputation operator seems to be that the operator will be moved to a position in the plan where it has access to the maximum training data.
+  \item S3.4.3: An extremely nice emergent property from the L-value function for the imputation operator seems to be that the operator will be moved to a position in the plan where it has access to the maximum training data.
 I am slightly concerned that by focusing entirely on size, this might lead to skew in the training set.
@@
 
-To clarify the point about Q8, we've removed the predicate condition $schooldegree \geq 0$, this was in fact a trivial satisfied predicate, given the way our dataset was set up.}
+      To clarify the point about Q8, we've removed the predicate condition $schooldegree \geq 0$. This was in fact a trivially satisfied predicate given the way our dataset was set up.}
     \todo{Need a better explanation here.}
 
-\item - S3.4.4: It appears that Time() assumes that plans are IO-bound.
+  \item S3.4.4: It appears that Time() assumes that plans are IO-bound.
 This is a perfectly reasonable assumption, but should be stated explicitly, as datasets that fit in memory will require a different cost model.
@@
 
-\item - S3.4.4: "Drop is a special case of a sequential scan, and its time complexity can be expressed as a function of the number of heap pages read and the I/O cost per page." seems to suggest that drop is implemented as a blocking operator.
+  \item S3.4.4: "Drop is a special case of a sequential scan, and its time complexity can be expressed as a function of the number of heap pages read and the I/O cost per page." seems to suggest that drop is implemented as a blocking operator.
 Why not just take the same approach as selection and treat it as having negligible cost?
     \resp{We mistakenly suggest in S3 that we treat \textit{Drop} as a blocking operator. In
-        our implementation, we do endow it with $0$ cost, and we have updated the text to
-        reflect that.}
+      our implementation, it has $0$ cost, and we have updated the text to reflect that.}
 
-  \resp{This was an error in the paper. In the implementation, we do treat Drop as taking negligible time.}
-
-\item - S4: All of the workload queries are (data-size-independent) AVG queries. What about other types of aggregates (e.g., SUM, COUNT)?
+  \item S4: All of the workload queries are (data-size-independent) AVG queries. What about other types of aggregates (e.g., SUM, COUNT)?
 
@@
       directly in Table 4.}
+
     \todo{For SUM queries, that may depend more directly on the number of rows returned than the
@@
 
-\item - S4.4.1: Imputation over the raw data is being used as ground truth.
+  \item S4.4.1: Imputation over the raw data is being used as ground truth.
 Is there any reason to suspect that this is the correct thing to do?
@@
 
-    \item - There are a number of opportunities for trimming in the figures, tables, and algorithms: Algorithms 1-3 could be compacted a bunch, Table 1 is full of whitespace (e.g., make it a 3-column {figure*}), Tables 1-2 could also be summarized as CDFs, Figure 6 is full of whitespace, Table 3 could be put queries on a single row (e.g., Q1 would easily fit on 1 line), and the references have a lot of redundant information (URLs, for example).
+  \item There are a number of opportunities for trimming in the figures, tables, and algorithms: Algorithms 1-3 could be compacted a bunch, Table 1 is full of whitespace (e.g., make it a 3-column {figure*}), Tables 1-2 could also be summarized as CDFs, Figure 6 is full of whitespace, Table 3 could be put queries on a single row (e.g., Q1 would easily fit on 1 line), and the references have a lot of redundant information (URLs, for example).
 
@@
 
-\item - Additional space recovered from these sections could be used to discuss the choice of techniques for imputation, and the confidence estimation idea suggested in the future work section.
+  \item Additional space recovered from these sections could be used to discuss the choice of techniques for imputation, and the confidence estimation idea suggested in the future work section.
 
-\item - Another interesting use of the space, or at least an idea for future work might be the role of aggregation on inference accuracy.
+  \item Another interesting use of the space, or at least an idea for future work might be the role of aggregation on inference accuracy.
 In my experience aggregate queries tend to be *much* more forgiving of approximations (e.g., Online Aggregation) and data errors.
@@
 
-  \resp{Yes, the imputed results depend on the records which are relevant to particular query.
+    \resp{Our contribution is the planning algorithm which allows imputation to be interleaved into the query plan.
       We do not present a new imputation method, but our technique is agnostic to the particular imputation method used.
-We reduce the overhead of imputation by performing it only on the tuples which are relevant to a particular query.}
+      We reduce the overhead of imputation by performing it only on the tuples which are relevant to a particular query, so the imputed results do depend on the records which are relevant to particular query.}
 
@@
 
-  \resp{Our evaluation compares our technique to an imputation of the full dataset.
+    \resp{
+      We compare our planning technique for on-the-fly imputation to imputation performed on the full dataset, using the same imputation algorithm in both cases.
+      We are not aware of another database technology which offers on-the-fly imputation, so there is no direct comparison for the planning algorithm.
+      % Our evaluation compares our technique to an imputation of the full dataset.
 We use the same imputation algorithm for our on-the-fly imputation as for the full dataset imputation.
-Cleaning the full dataset is the standard method of handling missing values.}
+Cleaning the full dataset is the standard method of handling missing values.
+    }
 
@@
 
-  \resp{Yes, as discussed above.}
+    \resp{Yes, as discussed in the response to 1.}
 
@@
 
-  \resp{The error measures that we provide compare our on-the-fly application of imputation to imputation over the full dataset.}
-
-\item D3) Table 3 uses a lot of space, and contains simple queries that are already described in the text. The table can be removed to include additional experiments as suggested above.
+    \resp{
+      We have added two new imputation methods to the experiments, to show that the planning algorithm can be used successfully with different imputation methods.
+      We have also added a subsection to the algorithm section explaining how we chose the cost functions for each imputation method that we test and giving rules for choosing cost functions for a new imputation method.
+    }
   \end{itemize}
@@
   \begin{itemize}
-\item W1. The paper feels like an early submission that is not ready for publication at its current state.
-
   \item W2. More work is required in the presentation: important parts are not explained in depth, while there are many details in other sections that can be omitted.
@@
 
-    If necessary, it should be possible to sample from the pareto optimal plans during the plan enumeration.
+      If necessary, it is possible to sample from the pareto optimal plans during the plan enumeration.
 The final selected plan may not be pareto optimal, but this is a potential way to control the number of plans.}
+    \todo{Talk about pareto set approximation here.}
 
   \item D7. When defining the cost of impute operator, how do you know a priori the number of iterations for the algorithm (given it has to converge)?
+    \todo{Micah: what do we actually do here?}
 
@@
 
+    \resp{We agree.
+      Reusing previous query results is a tricky problem because it requires the database to speculatively store some of the intermediate results of its query plans.
+      A simple heuristic which should work well is to store the results of full table imputation whenever a query forces this to occur. These results are guaranteed to be useful later.
+      Storing other intermediate results is problematic, particularly those which are computed after filtering or joining, as we would need to have a mapping back to the tuples in the base tables.}
+
   \item D11. Less important points:
@@
 
+    \resp{The subset of SQL that we consider does not allow filtering on the results of joins (see Section 3.2).
+Extending the planner to cover this case is not complex.
+Before adding filters, it must add imputations on the required attributes.}
+
   \item -- Given you optimize for both performance and quality, Tummer's and Koch's work on multi-objective QO is related.
 
-  \resp{Yes, Tummer \& Koch is closely related.
-However, much of the complexity of their paper is in dealing with cost functions which are not well behaved.
-Our cost functions are both monotonic, which allows us to use a simpler optimizer than what Tummer \& Koch propose.}
+    \resp{Yes, Tummer \& Koch is definitely related.
+      However, much of the complexity of their method is in dealing with cost functions which are not well behaved.
+      Our cost functions are monotonic, which allows us to use a simpler optimizer than what Tummer \& Koch propose.
+    We have added a discussion of their work to the related works section.}
 

### 2017-05-11 paper/algorithm.tex
Update formulae
paper: continue to reduce figures
a pass over cost model and impute api

--- 2017-05-11 paper/algorithm.tex
@@
 The cost of a plan $q$ is a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
-$\textsc{Penalty}(q) \in (0,1]$ is a heuristic that attempts to quantify the amount of information lost by the imputation procedure used.
+$\textsc{Penalty}(q) \in (0,1]$ is a heuristic that quantifies the amount of information lost by the imputation procedure used.
 $\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics, selectivity estimation of the query predicates, and the complexity of the
@@
 
-This set $Q$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.
+This set $Q$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all feasible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.
 
@@
 
-Given $\alpha$ and $Q$, \ProjectName{} returns an $\alpha$-bound optimal plan in $Q$.
-
-So in essence, $\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance, as plans that
+Given $\alpha$ and $Q$, \ProjectName{} returns an $\alpha$-bound optimal plan in $Q$. In essence, $\alpha$ is a tunable parameter 
+that determines whether the optimizer focuses on quality or on performance, as plans that
 lose a lot of information through dropping are also the fastest.
@@
 A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of plan histograms.
-From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations to summary statistics such as mean and variance
+From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations to distribution statistics such as mean and variance.
 Intuitively, imputation quality \textit{increases} when more complete attributes and complete tuples are available; correspondingly, $P(q')$ \textit{decreases} as the number of values increases.
@@
 the properties of the underlying algorithm via a function $T$. If we assume
-that the \textit{Impute} routine is CPU-bound, then we will adjust
+that the \textit{Impute} routine is CPU-bound, then we adjust
 \textsc{Time} by some constant factor $f$, thereby scaling the result to be
@@
 
-The form of $T$ depends on the specific imputation method, and as with $P$,
-there are some principles to capture the intuition.  $T$ should \emph{increase}
+The form of $T$ depends on the specific imputation method, but there are various
+principles useful in capturing the intuition of a practical choice.
+$T$ should \emph{increase}
 as the number of tuples and complete and dirty attributes passed to the
@@
 
-\subsection{Imputation strategies}
+\subsection{Imputation strategies and their API}
 \label{sec:imputation}
@@
 can be plugged in with minimal effort and without changes to the optimizer, simply by supplying
-penalty and time functions (\Cref{sec:cost-model}) pertinent to the algorithm.
+penalty ($P$) and time ($T$) functions (\Cref{sec:cost-model}) pertinent to the algorithm.
 These can then be queried by the optimizer during planning.  The flexibility of
 this approach aligns with the spirit of query optimization, which aims to
-reconcile performance with a declarative interface to data manipulation.
+reconcile performance with a declarative interface to data manipulation. To
+motivate this API, we describe how an analyst would specify the form of these
+functions for a general-purpose imputation algorithm, as well as two simpler
+methods.
 
@@
 
-New imputation algorithms can also be added. As a reference implementation,
+As a reference implementation,
 \ProjectName{} uses a general-purpose imputation strategy based on
@@
 
-These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
-process, in which multiple distinct copies of the complete data are generated and estimators
-are computed by averaging over the multiple datasets to form a single completed dataset~
-\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
-multiple imputation could be incorporated as another avenue of assessing imputation confidence.
+% Jose: I think this is a bit out of place. So trimmed for space
+%These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
+%process, in which multiple distinct copies of the complete data are generated and estimators
+%are computed by averaging over the multiple datasets to form a single completed dataset~
+%\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
+%multiple imputation could be incorporated as another avenue of assessing imputation confidence.
 
-In the reference implementation, the chained-equation algorithm proceeds by
+The chained-equation algorithm proceeds by
 iteratively fitting decision trees for the target imputation attributes. In
@@
 between attributes. The algorithm terminates after a fixed number of
-iterations, or earlier if convergence is achieved. (That is, when the imputed
-values of one column are not updated by subsequent iterations that update that
-column.) For our experiments, we use Weka's \texttt{REPTree}
+iterations, or earlier if convergence is achieved, indicated by unchanged values in a 
+column in subsequent iterations. For our experiments, we use Weka's \texttt{REPTree}
 implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
@@
 
-\todo{Fill this in.}
-\[P(q) = \frac{1}{n} (1 + \overline{Variance(q)}) \]
+At the other end of the spectrum of complexity, an analyst could use a simple
+method that replaces a missing value with the mean over available values
+in that attribute, a common choice in exploratory data analysis known
+as mean value imputation.
+
+We can take advantage of the histograms maintained by \ProjectName{}
+to estimate the mean and construct a non-blocking impute operator.
+Applying the heuristics from \ref{sec:quality} and \ref{sec:query-runtime}, we
+have
+
+\begin{equation*}
+\begin{aligned}
+    T(q) &\approx 0 \\
+    P(q) &= \sum_{A \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(A)}{\textsc{Card}(q)}
+\end{aligned}
+\end{equation*}
 
@@
 
-\todo{Fill this in.}
+Hot-deck imputation is a nonparametric method in which missing values are
+replaced with randomly selected complete values. This method preserves the
+distribution of the complete values. The heuristic $P$ remains the same, 
+but we modify $T$ to account for the fact that hot-deck is blocking in our
+implementation.
+
+\begin{equation*}
+\begin{aligned}
+    T(q) &= \textsc{Card}(q) \\
+    P(q) &= \sum_{A \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(A)}{\textsc{Card}(q)}
+\end{aligned}
+\end{equation*}
 

### 2017-05-11 simpledb/src/java/simpledb/ImputeMean.java
Add ImputeMean

@@ -1,0 +1,116 @@
+package simpledb;
+
+import java.util.ArrayList;
+import java.util.Collection;
+import java.util.Iterator;
+import java.util.Random;
+import java.util.Set;
+
+
+public class ImputeMean extends Impute {
+
+    private static final long serialVersionUID = 1L;
+
+	private TableStats tableStats = null;
+
+    /**
+     * Impute missing data for a column by using the mean of the
+     * non-missing elements of the same column. Thus, this method requires that
+     * we store all of the child tuples in a buffer until the child is
+     * exhausted. Alternately, we could try to make use of the histograms for the table.
+     * @param child
+     */
+    public ImputeMean(Collection<String> dropFields, DbIterator child){
+        super(dropFields, child);
+
+        // get tablestats?
+        int thisTableId = -1;
+        Catalog catalog = Database.getCatalog();
+        Iterator<Integer> it = catalog.tableIdIterator();
+        while (it.hasNext()){
+		int tableId = it.next();
+		if (catalog.getTupleDesc(tableId).equals(td)){
+			thisTableId = tableId;
+			break;
+		}
+        }
+        if (thisTableId != -1){
+		String tablename = catalog.getTableName(thisTableId);
+		this.tableStats = TableStats.getTableStats(tablename);
+        }
+    }
+
+    public ImputeMean(DbIterator child) {
+        this(null, child);
+    }
+
+    @Override
+    public void rewind() throws DbException, TransactionAbortedException {
+        child.rewind();
+    }
+
+    @Override
+    protected Tuple fetchNext() throws DbException, TransactionAbortedException {
+	if (this.tableStats == null){
+		throw new DbException("TableStats not loaded.");
+	}
+
+	if (child.hasNext()){
+		Tuple t = child.next();
+		if (t.hasMissingFields()){
+			return impute(child.next());
+		} else {
+			return t;
+		}
+	} else {
+		return null;
+	}
+    }
+
+    private Tuple impute(Tuple t) throws DbException {
+        Tuple tc = new Tuple(t);
+
+        for (int j : dropFieldsIndices){
+            // Don't impute if not missing.
+            if (!tc.getField(j).isMissing())
+                continue;
+
+            // set to mean
+            tc.setField(j, new IntField((int) tableStats.estimateMean(j)));
+        }
+
[…]

### 2017-05-11 simpledb/src/java/simpledb/TableStats.java

@@ -294,6 +294,14 @@ public class TableStats {
     	return ret;
     }
 
+    public double estimateMean(int field){
+	return this.intStats[field].mean();
+    }
+
+    public double estimateVariance(int field){
+	return this.intStats[field].variance();
+    }
+
     /**
      * return the total number of tuples in this table
      * */

### 2017-05-11 simpledb/src/java/simpledb/TableStats.java
Add ImputeMean

@@ -293,6 +293,14 @@ public class TableStats {
     	}
     	return ret;
     }
+    
+    public double estimateMean(int field){
+    	return this.intStats[field].mean();
+    }
+    
+    public double estimateVariance(int field){
+    	return this.intStats[field].variance();
+    }
 
     /**
      * return the total number of tuples in this table

### 2017-05-11 paper/experiments.tex

--- 2017-05-11 paper/experiments.tex
@@
 \Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
-queries are an \todo{update/validate: order-of-magnitude faster than imputing on the base tables. We can get
-another order-of-magnitude speedup when optimizing for performance}. This performance
+queries are an \todo{update/validate: order-of-magnitude faster than imputing on the base tables. We get another order-of-magnitude speedup when optimizing for performance}. This performance
 differential means it is feasible
-to explore multiple imputations operations (including more expensive operators) when using
+to explore multiple imputation operations (including more expensive operators) when using
 \ProjectName{}, in contrast to the traditional approach of base table imputation.
@@
 %planning.
-
-\todo{update: In the median across a variety of queries and choices of $\alpha$, planning took
-\planningmediantime{} ms, 
-constituting \planningruntimepercent{} percent of total runtime. In all cases, the optimizer
-returned a query plan within \planningmaxtime{} ms, with times similar across values of $\alpha$.}
 
@@
 
-We explored the practical implications of our planning algorithm by constructing a series of queries with increasing number of joins.
-We used the CDC tables and evaluated a combination of self-joins and joins across reference tables. Planning times for queries
-involving 1 to 5 joins, which although limited is a practical value for real world exploratory queries, take less than a second.
+\todo{Update: Planning times are reasonable, with a median planning time of \planningmediantime{} ms. over the benchmark queries, constituting \planningruntimepercent{}\% of total runtime.
+In all cases, the optimizer returned a query plan within \planningmaxtime{} ms.}
+The choice of $\alpha$ has no impact on planning times, because query selection is done after collecting the Pareto frontier.
+
+However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
+We explored the limits our planning algorithm by constructing a series of queries with increasing number of joins.
+We used the CDC tables and evaluated a combination of self-joins and joins across reference tables.
+Planning times for queries involving 1--5 joins, which is a practical value for real world exploratory queries, take less than a second.
 \todo{update: For queries with 7 joins, planning time increases substantially to 29 seconds, on average. We introduced an optimization that reduces

### 2017-05-11 paper/queries.tex
removed redundant filter in a query and addressed in rebuttal

@@
 \newcounter{queryno}
+{
+  \lstset{breaklines=true, basicstyle=\scriptsize\ttfamily}
 \begin{tabular}{cl}
@@
       \begin{minipage}{6in}
-\begin{lstlisting}[breaklines]
+\begin{lstlisting}[breaklines, basicstyle=\scriptsize\ttfamily]
 SELECT income, AVG(cuff_size) FROM demo, exams 
@@
 \end{lstlisting}
-\end{minipage}\refstepcounter{queryno}\label[query]{q1} \\
+      \end{minipage}\refstepcounter{queryno}\label[query]{q1} \\[-1.5ex]
   2 & 
       \begin{minipage}{6in}
-\begin{lstlisting}[breaklines]
-SELECT income, AVG(creatine) FROM demo, exams, labs 
-WHERE demo.id = exams.id AND exams.id = labs.id AND income >= 13 AND income <= 15 AND weight >= 63 GROUP BY income;
+\begin{lstlisting}[breaklines, basicstyle=\scriptsize\ttfamily]
+SELECT income, AVG(creatine) FROM demo, exams, labs WHERE demo.id = exams.id AND exams.id = labs.id AND income >= 13 AND income <= 15 AND weight >= 63 GROUP BY income;
 \end{lstlisting}
       \end{minipage}
-\refstepcounter{queryno}\label[query]{q2} \\
+      \refstepcounter{queryno}\label[query]{q2} \\[-1.5ex]
   3 & 
@@
 \end{lstlisting}
-\end{minipage}\refstepcounter{queryno}\label[query]{q3}\\
+      \end{minipage}\refstepcounter{queryno}\label[query]{q3}\\[-1.5ex]
   4 & 
@@
 \end{lstlisting}
-\end{minipage}\refstepcounter{queryno}\label[query]{q4}\\
+      \end{minipage}\refstepcounter{queryno}\label[query]{q4}\\[-1.5ex]
   5 & 
@@
       \end{minipage}\refstepcounter{queryno}\label[query]{q5}\\
+  \midrule
   \# & \multicolumn{1}{c}{Queries on freeCodeCamp data} \\
@@
 \begin{lstlisting}[breaklines]
-SELECT attendedbootcamp, AVG(income) FROM fcc
-WHERE income >= 50000 GROUP BY attendedbootcamp;
+SELECT attendedbootcamp, AVG(income) FROM fcc WHERE income >= 50000 GROUP BY attendedbootcamp;
 \end{lstlisting}
-\end{minipage}\refstepcounter{queryno} \label[query]{q6} \\
+      \end{minipage}\refstepcounter{queryno} \label[query]{q6} \\[-1.5ex]
   7 & 
@@
 \begin{lstlisting}[breaklines]
-SELECT AVG(commutetime) FROM fcc
-WHERE gender = "female" AND countrycitizen = "United States";
+SELECT AVG(commutetime) FROM fcc WHERE gender = "female" AND countrycitizen = "United States";
 \end{lstlisting}
-\end{minipage}\refstepcounter{queryno} \label[query]{q7} \\
+      \end{minipage}\refstepcounter{queryno} \label[query]{q7} \\[-1.5ex]
   8 & 
@@
 \begin{lstlisting}[breaklines]
-SELECT schooldegree, AVG(studentdebtowe) FROM fcc
-WHERE studentdebtowe > 0 AND schooldegree >= 0 GROUP BY schooldegree;
+SELECT schooldegree, AVG(studentdebtowe) FROM fcc WHERE studentdebtowe > 0 GROUP BY schooldegree;
 \end{lstlisting}
-\end{minipage}\refstepcounter{queryno} \label[query]{q8}\\
+      \end{minipage}\refstepcounter{queryno} \label[query]{q8}\\[-1.5ex]
   9 & 
@@
 SELECT attendedbootcamp, AVG(gdp_per_capita) FROM fcc, gdp 
-WHERE fcc.countrycitizen = gdp.country AND age >= 18
-GROUP BY attendedbootcamp;
+WHERE fcc.countrycitizen = gdp.country AND age >= 18 GROUP BY attendedbootcamp;
 \end{lstlisting}
-\end{minipage}\refstepcounter{queryno} \label[query]{q9}\\
+      \end{minipage}\refstepcounter{queryno} \label[query]{q9}\\[-1.5ex]
   \bottomrule
 \end{tabular}
-
+}
 %%% Local Variables:

### 2017-05-11 simpledb/data.py
Data.py script is python2

@@ -1,4 +1,4 @@
-#!/usr/local/bin/python
+#!/usr/bin/env python2
 
 # Convert CDC and other data to simpledb format
 # Make floating point columns integer

### 2017-05-11 paper/conclusion.tex

@@
 \section{Conclusion}
-Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational calculus and
+Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational algebra and
 existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine

### 2017-05-11 paper/implementation.tex

@@
 Adding imputation and our optimization algorithm to a standard SQL database requires modifications to several key database components.
-In this section we discuss the changes which must be made to implement dynamic imputation.
+In this section we discuss the changes required to implement dynamic imputation.
 
@@
 \item \textbf{Extended histograms:}
-  Ranking query plans with imputation requires estimates of the distribution of missing values in the base tables and in the outputs of sub-plans.
+  Ranking query plans with imputation needs estimates of the distribution of missing values in the base tables and in the outputs of sub-plans.
   We extended standard histograms to include the count of missing values for each attribute.

### 2017-05-11 simpledb/src/java/simpledb/ImputeFactory.java
Add ImputeFactory

@@ -1,0 +1,17 @@
+package simpledb;
+
+import java.util.Collection;
+
+public class ImputeFactory {
+	public static Impute newImpute(String imputationMethod, Collection<String> arg1, DbIterator arg2){
+		if (imputationMethod.equals("REGRESSION_TREE")){
+			return new ImputeRegressionTree(arg1, arg2);
+		} else if (imputationMethod.equals("HOTDECK")){
+			return new ImputeHotDeck(arg1, arg2);
+		} else if (imputationMethod.equals("MEAN")){
+			return new ImputeMean(arg1, arg2);
+		} else {
+			throw new UnsupportedOperationException();
+		}
+	}
+}

### 2017-05-11 rebuttal.tex
rebut: revise response on drops before filters

@@
 
-    \resp{We agree that it could be interesting to allow drops to be omitted before filters.
+    \resp{We agree that it would be interesting to allow drops to be omitted before filters.
       This would increase the plan space somewhat and might lead to better plans in some cases.
-For the conjunctive predicates in our experiments, our implementation splits the predicate into multiple filter nodes, so imputations or drops and filters can interleave.
-This should have the same effect, as the filter and the drop operator behave the same way on a tuple containing null.
-However, this optimization would also apply to disjunctive predicates, which cannot be split, and is worth definitely investigating.}
+      This optimization would only change the output of the filter when the predicate is disjunctive, so would not have an impact on the queries we used in our experiments.
+      Filters with conjunctive predicates behave identically with and without the drop, although removing the drop will give a minor performance improvement.}
 

### 2017-05-11 rebuttal.tex

@@
 (3) Given that the ML time estimator is being treated as a black box, it would be useful to see how sensitive the plan selector is to inaccurate estimates.
-    \todo{}
+
+  \resp{
+  We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation. We also added
+ details on how to pick the heuristic functions that define the API for impute operators. We provide details on how these functions are defined
+ for all the methods used in our evaluation. 
+  }
 
@@
     \resp{
-      The outsized error in Q8 is a result of a drop operator placed very early on in the plan (a result of optimizing for performance). This produced significantly smaller sample sizes for various
+      The outsized error in Q8 is due to a drop operator placed very early on in the plan (a result of optimizing for performance). This produced significantly smaller sample sizes for various
       groups, which skewed the average calculation for each group significantly. For example, the average student debt for subjects with no high school degree was almost 83\% off from the estimate
       using the base table imputation, but if we look at the underlying population the former was calculated with 6 people, while the latter was with 258 people.
-      \todo{We have extended ImputeDB to provide users with a estimate of the number of tuples dropped in a calculation due to nulls, which may be indicative of the underlying error or skew
-        in the results presented. This is a simple heuristic but we find that it would have provided appropriate warning in cases such as Q8.}
     }
@@
   \item Although it doesn't consider cost or the drop operator, a similar technique appears in Yang et. al. "Lenses: An On-Demand Approach to ETL." A comparison would be nice.
-    \todo{}
+   \resp{
+   Yang's work is definitely related, and we've added it to our related work section. However, it is worth noting that their decision process for application of lenses is at the tuple-level,
+   rather than with respect to the query plan.
+   }
 
-  \item S1: "Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach — between 0 and 24 percent" 24 percent error is pretty bad.
-    \todo{}
+  \item S1: "Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach between 0 and 24 percent" 24 percent error is pretty bad.
 
@@
 
-      To clarify the point about Q8, we've removed the predicate condition $schooldegree \geq 0$. This was in fact a trivially satisfied predicate given the way our dataset was set up.}
-    \todo{Need a better explanation here.}
+     To clarify the point about Q8, we've removed the predicate condition $schooldegree \geq 0$. 
+      We originally enumerated strings with positive integers and replaced missing values strings with 0, rather than perform imputation on them, so every enumerated value
+      satisfied this predicate trivially. We have changed our evaluation to impute string enumerations as well and dropped this predicate.
+      }
+    %\todo{Need a better explanation here.}
 
@@
 
+  \resp{
+  We've extended the evaluation section to include more impute operators and an evaluation of planning times.
+  }
+
   \item D1.
@@
     -- Algorithms 1 and 2 could probably be presented in an easier to grasp way.
+
+    \resp{
+    We've broken up this section and consolidated algoirthm notation. The cardinality estimation section provides more detail about the algorithm. We also
+    added intuition for how the heuristic Penalty (a renaming of Loss, for clarity) and Time should be designed. We walk through these definitions for three different
+    impute operators.
+    }
 

### 2017-05-11 paper/algorithm.tex

@@
 \label{sec:cardinal}
-The computation of $\textsc{Penalty}(q)$ and$\textsc{Time}(q)$ relies on accurate cardinality estimates for the plan $q$ and its sub-plans.
+The computation of $\textsc{Penalty}(q)$ and $\textsc{Time}(q)$ relies on accurate cardinality estimates for the plan $q$ and its sub-plans.
 These estimates are impacted not just by filtering or joining, as in the traditional relational algebra, but also by the imputation operators.
@@
 To compute cardinality, we maintain histograms for each column in the database.
+These histograms track the distribution of values in the column as well as the number of missing values.
 During query planning, each of the logical nodes in a query plan points to a set of histograms which describe the distribution of values in the output of the node.
-When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan.
+When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as described in~\Cref{algo:histogram-transformation} to account for the new operation in the plan.
 
-\Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
-Drop operators always reduce the cardinality of the relation. Given $\delta_C$, each column $c \in C$ clears its count of tuples with missing values,
-which can yield distinct cardinality estimates across attributes. To provide consistent estimates, we take the lowest such estimate $e$ and
-scale all attribute histograms based on their distribution such that each provides an estimate of $e$. In the case of $\mu_C$, the cardinality estimate
-remains constant. The count of missing values across attributes is cleared and redistributed along the appropriate histograms based on their distribution.
-In the case of $\sigma$ and $\bowtie$ operations, we estimate the cardinality of the output relation and adjust histogram counts based on their distribution to sum to that estimate.
+Drop operators (\Crefrange{lst:drop-start}{lst:drop-end}) reduce the cardinality of their input relation by removing tuples which contain missing values.
+For a drop $\delta_C$, for each column $c \in C$ we set its missing value count to zero, because all of these tuples will be dropped.
+This can result in a set of histograms which have different cardinalities if some columns have more missing values than others.
+To account for the discrepancy, we rescale all the histograms without changing the shape of their distributions to match the lowest cardinality column.
 
+Impute operators (\Crefrange{lst:impute-start}{lst:impute-end}) maintain the cardinality of their inputs, but they also fill in missing values.
+For an impute $\mu_C$, we set the missing value count to zero, and rescale each histogram so that its new cardinality, with no missing values, is the same as its old cardinality.
+We assume that the imputed values will not change the shape of the distribution of values in each column.
+
+Cardinality estimation for standard relational operators (\Crefrange{lst:std-start}{lst:std-end}) is the same as in a normal database.
 \begin{algorithm}
@@
     \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
-                                            \text{ where } \\
-                                           P' &= \left\{q ~|~ \begin{aligned}
+    \text{ where }\ P' &= \left\{q ~\Bigg|~ \begin{aligned}
         q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\

### 2017-05-11 paper/relatedwork.tex

@@
 
+\todo{confirm that this reading of Yang is correct}
+\cite{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
+of probabilistic compositional repair operators, called Lenses. These can be used
+to repair missing values by imposing data constraints such as \verb|NOT NULL|. 
+Similarly to \ProjectName{}, they provide a way to balance between the quality of the cleaning 
+performed (expressed by the confidence of their output) and the cost of those operations. However, their 
+repairs take place on a per-tuple basis, with the non-deterministic layer of their system replacing place-holder
+variables with inferred values. Meanwhile, \ProjectName{} optimizes imputation at the plan-level and
+provides the same impute technique for all tuples streaming through a given operator.
+
 \ProjectName's cost-based query planner 

### 2017-05-11 simpledb/test/experiments/analyze_experiments.py
Update

@@ -17,21 +17,17 @@ except ImportError:
     from io import StringIO
 
 # configure this
-nqueries = 10
+nqueries = 9
 table_headers = [
     ['income', 'cuff_size'],                # query 0
     ['income', 'creatine'],                 # query 1
     ['blood_lead'],                         # query 2
     ['gender', 'blood_pressure_systolic'],  # query 3
-    ['attendedbootcamp', 'income'],         # query 4
-    ['commutetime'],                        # query 5
-    ['schooldegree', 'studentdebtowe'],     # query 6
-    ['attendedbootcamp', 'gdp_per_capita'], # query 7
-    ['waist_circumference'],                # query 8
-    ['income','cuff_size'],                 # query 9
-    ['attendedbootcamp','income'],          # query 10
-    ['commutetime'],                        # query 11
-    ['gender','age'],                       # query 12
+    ['waist_circumference'],                # query 4
+    ['attendedbootcamp', 'income'],         # query 5
+    ['commutetime'],                        # query 6
+    ['schooldegree', 'studentdebtowe'],     # query 7
+    ['attendedbootcamp', 'gdp_per_capita'], # query 8
 ]
 
 # basic utils
@@ -161,7 +157,7 @@ def collapse_results(experiments_dir, output_dir, base=False):
         data['is_experiment'] = False
         data['alpha'] = 'Impute at base tables'
 
-    data = drop_warmup(data, ['query', 'alpha'], drop=20)
+    data = drop_warmup(data, ['query', 'alpha'], drop=0)
 
     # time measures
     by = ['query', 'alpha']

### 2017-05-12 paper/algorithm.tex
paper: tweak time & penalty sections

@@
 The function $\textsc{L}_{P}(q')$ is then defined recursively to aggregate these penalties for the entire plan. 
-The result is normalized by $\textsc{L}_{P_\delta}(q')$, which is the total penalty for always dropping tuples with \nullv{} values and equates to the count of imputation operators considered.
+The result is normalized by $\textsc{L}_{P_\delta}(q')$, which is the total penalty for always dropping tuples with \nullv{} values and is equal to the number of imputation operators in the plan.
 
@@
 A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of plan histograms.
-From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations to distribution statistics such as mean and variance.
+From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations of distribution statistics such as mean and variance.
 Intuitively, imputation quality \textit{increases} when more complete attributes and complete tuples are available; correspondingly, $P(q')$ \textit{decreases} as the number of values increases.
-The heuristic penalty should also reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
-Finally, the heuristic penalty may increase as amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.
+The penalty should also reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
+Finally, the penalty may increase as amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.
 
-Imputation quality $P(q')=0$ is unachievable, as this suggests the values of missing elements are known exactly.
+Imputation quality $P(q')=0$ is unattainable, as this suggests the values of missing elements are known exactly.
 On the other hand, $P(q')=1$ occurs when all tuples with missing values are dropped, as this represents the worst case in terms of recovering information from these tuples.
@@
 Join costs are estimated as a function of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
-Filtering, projecting, and aggregating are all done in memory and have low computational overhead so we assume negligible time costs for those operations.
+Filtering, projecting, and aggregating never require repeated tuple fetches and have low computational overhead so we assume that they take negligible time.
 
@@
 
-\noindent{\textbf{Constructing $T$}}: As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on
-the properties of the underlying algorithm via a function $T$. If we assume
-that the \textit{Impute} routine is CPU-bound, then we adjust
-\textsc{Time} by some constant factor $f$, thereby scaling the result to be
-comparable with IO-bound routines.
+\noindent{\textbf{Constructing $T$}}:
+As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on the properties of the underlying algorithm via a function $T$.
+If we assume that the \textit{Impute} routine is CPU-bound, then we adjust $T$ by some constant factor $f$, thereby scaling the result to be comparable with IO-bound routines.
 
-The form of $T$ depends on the specific imputation method, but there are various
-principles useful in capturing the intuition of a practical choice.
-$T$ should \emph{increase}
-as the number of tuples and complete and dirty attributes passed to the
-imputation function increases and it should be take into account the complexity
-of the imputation algorithm. Analysis of the time complexity of different
-imputation algorithms, especially with in-database execution, is beyond the
-scope of this paper. As with $P$, it is the ordering between queries
-generated by $T$ that is more important than the precision of estimates.
+The form of $T$ depends on the specific imputation method, but there are several considerations in choosing $T$.
+In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases and it should be take into account the complexity of the imputation algorithm.
+Analysis of the time complexity of different imputation algorithms, especially with in-database execution, is beyond the scope of this paper.
+As with $P$, it is the ordering between queries generated by $T$ that is more important than the precision of estimates.
+
+We discuss several values of $T$ for different imputation methods in \Cref{sec:imputation}.
 

### 2017-05-12 paper/algorithm.tex
paper: pass over imputation methods section

@@
 
-\ProjectName{} is designed so that any imputation strategy
-can be plugged in with minimal effort and without changes to the optimizer, simply by supplying
-penalty ($P$) and time ($T$) functions (\Cref{sec:cost-model}) pertinent to the algorithm.
-These can then be queried by the optimizer during planning.  The flexibility of
-this approach aligns with the spirit of query optimization, which aims to
-reconcile performance with a declarative interface to data manipulation. To
-motivate this API, we describe how an analyst would specify the form of these
-functions for a general-purpose imputation algorithm, as well as two simpler
-methods.
+\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal effort and without changes to the optimizer, by supplying penalty ($P$) and time ($T$) functions (\Cref{sec:cost-model}) pertinent to the algorithm.
+These cost functions are then used by the optimizer during planning.
+The flexibility of this approach is in the spirit of query optimization, which aims to reconcile performance with a declarative interface to data manipulation.
+To motivate this API, we describe how a developer would specify these functions for a general-purpose imputation algorithm, as well as for two simpler imputation methods.
 
@@
 all complete attributes and all other missing attributes. Chained-equation
-decision trees algorithms are found to be effective~\cite{akande2015empirical}
+decision trees algorithms are effective~\cite{akande2015empirical}
 and are widely used in epidemiological domains~\cite{burgette2010multiple}.
@@
 The chained-equation algorithm proceeds by
-iteratively fitting decision trees for the target imputation attributes. In
-each iteration, the missing values of a separate attribute are replaced with
-newly imputed values.  Values in a single column can be updated over multiple
+iteratively fitting decision trees for the target attributes. 
+In each iteration, the missing values of a single attribute are replaced with newly imputed values.  Values in a single column can be updated over multiple
 iterations of the algorithm. With more iterations, the quality of the
@@
 improvement in imputation quality from more tuples and attributes as well as
-the diminishing returns to more data.
+the diminishing returns from more data.
 
@@
 
-At the other end of the spectrum of complexity, an analyst could use a simple
-method that replaces a missing value with the mean over available values
-in that attribute, a common choice in exploratory data analysis known
-as mean value imputation.
+At the other end of the complexity spectrum, mean value imputation is a common choice for exploratory data analysis.
+Mean value imputation replaces missing values in a column with the mean of the available values in the column.
 
-We can take advantage of the histograms maintained by \ProjectName{}
-to estimate the mean and construct a non-blocking impute operator.
-Applying the heuristics from \ref{sec:quality} and \ref{sec:query-runtime}, we
+We can take advantage of the histograms maintained by \ProjectName{} to estimate the mean and construct a non-blocking impute operator.
+Applying the heuristics from~\Cref{sec:quality} and~\Cref{sec:query-runtime}, we
 have
@@
 
+Note that $P$ increases with the variance of the columns and decreases with their cardinality, reflecting the intuition that the quality of the mean is lower with high variance and higher with more data.
+
 \subsubsection{Hot-deck}
 
-Hot-deck imputation is a nonparametric method in which missing values are
-replaced with randomly selected complete values. This method preserves the
-distribution of the complete values. The heuristic $P$ remains the same, 
-but we modify $T$ to account for the fact that hot-deck is blocking in our
-implementation.
+Hot-deck imputation is a nonparametric method in which missing values are replaced with randomly selected complete values.
+This method preserves the distribution of the complete values.
+We use the same $P$ as for mean value imputation, but we modify $T$ to account for the fact that hot-deck is blocking in our implementation.
 
@@
 \end{equation*}
-
-%\begin{algorithm}
-%    \input{algorithms/cart.tex}
-%\end{algorithm}
 

### 2017-05-12 paper/experiments.tex
paper: fix up title capitalization

@@
 
-\subsection{Data sets}\label{subsec:datasets}
+\subsection{Data Sets}\label{subsec:datasets}
 We collect three data sets for our experiments.
@@
 
-\subsubsection{Planning times}
+\subsubsection{Planning Times}
 \begin{table}
@@
 We explored the limits our planning algorithm by constructing a series of queries with increasing number of joins.
-We used the CDC tables and evaluated a combination of self-joins and joins across reference tables.
+We used the CDC tables and evaluated increasing number of joins across tables.
 Planning times for queries involving 1--5 joins, which is a practical value for real world exploratory queries, take less than a second.

### 2017-05-12 paper/relatedwork.tex
paper: mark sections for sam to review

@@
 
-\todo{confirm that this reading of Yang is correct}
+\review{
 \cite{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
 of probabilistic compositional repair operators, called Lenses. These can be used
-to repair missing values by imposing data constraints such as \verb|NOT NULL|. 
+to repair missing values by imposing data constraints such as \texttt{NOT NULL}. 
 Similarly to \ProjectName{}, they provide a way to balance between the quality of the cleaning 
@@
 provides the same impute technique for all tuples streaming through a given operator.
+}
 
@@
 
+\review{
 The planning algorithm that we present has similarities to the multi-objective query optimizer in~\cite{trummer2014}.
@@
 \cite{trummer2014} handles the more complex case of cost functions which are piecewise linear, but not necessarily monotonic.
+}
 

### 2017-05-12 rebuttal.tex

@@
       aggregates in order to compute the ``Count Fraction''. The timing results are
-      clearly identical in this case and the number of tuples returned can be seen
+      identical in this case and the number of tuples returned can be seen
       directly in Table 4.}
 
-    \todo{For SUM queries, that may depend more directly on the number of rows returned than the
+    \todo{Do we need to address this?:
+    For SUM queries, that may depend more directly on the number of rows returned than the
       quality of imputations for any individual row, the effect of different COUNT results
@@
       We compare our planning technique for on-the-fly imputation to imputation performed on the full dataset, using the same imputation algorithm in both cases.
-      We are not aware of another database technology which offers on-the-fly imputation, so there is no direct comparison for the planning algorithm.
+      We are not aware of another database technology which offers on-the-fly imputation optimization, so there is no direct comparison for the planning algorithm.
       % Our evaluation compares our technique to an imputation of the full dataset.
@@
       We have added two new imputation methods to the experiments, to show that the planning algorithm can be used successfully with different imputation methods.
-      We have also added a subsection to the algorithm section explaining how we chose the cost functions for each imputation method that we test and giving rules for choosing cost functions for a new imputation method.
+      We have also added a subsection to the algorithm section explaining how we chose the cost functions for each imputation method that we test and giving guidelines for choosing cost functions for a new imputation method.
     }
@@
     \resp{
-    We've broken up this section and consolidated algoirthm notation. The cardinality estimation section provides more detail about the algorithm. We also
+    We've broken up this section and consolidated algorithm notation. The cardinality estimation section provides more detail about the algorithm. We also
     added intuition for how the heuristic Penalty (a renaming of Loss, for clarity) and Time should be designed. We walk through these definitions for three different
@@
     \resp{
-      Existing systems have very complex formulas for estimating query costs. We felt that the overhead of reimplementing an existing cost model would not be worth the benefit.
+    \todo{this is a bad response...}
+      We felt that the overhead of reimplementing an existing cost model would not be worth the benefit.
       However, we added a new section that discusses in detail the main components for \ProjectName{}, so that developers of existing systems can use this as a roadmap
@@
 
-    \resp{We maintain the set of pareto optimal plans at each point in the optimization.
+    \resp{We maintain the set of Pareto optimal plans at each point in the optimization.
 This set is significantly smaller than the set of all plans, so it represents some pruning.
-We found that the size of these pareto sets is small enough that we did not need to prune further.
+We found that the size of these Pareto sets is small enough that we did not need to prune further.
 
-      If necessary, it is possible to sample from the pareto optimal plans during the plan enumeration.
-The final selected plan may not be pareto optimal, but this is a potential way to control the number of plans.}
-    \todo{Talk about pareto set approximation here.}
+      If necessary, it is possible to sample from the Pareto optimal plans during the plan enumeration.
+The final selected plan may not be Pareto optimal, but this is a potential way to control the number of plans.
+      To perform this approximation, we only add plans to the frontier that are non-dominated and are sufficiently different from existing plans in the frontier.
+The plan already in the frontier should
+      provide similar tradeoffs and reduces the search space explored. \ProjectName{} only employs this optimization when a large number of tables requiring imputation are involved in the query.
+      }
 
   \item D7. When defining the cost of impute operator, how do you know a priori the number of iterations for the algorithm (given it has to converge)?
-    \todo{Micah: what do we actually do here?}
+  For the iterative regression impute operatore, we set a predefined number of iterations as an upper bound on its execution. If the imputation converges,
+  we can return immediately, otherwise we return after this upper bound is reached. This is a common approach used in iterative optimization techniques, such
+  as gradient descent.
 
@@
 This kind of imputation isn't useful for survey data, so we didn't consider it for our experiments.
-We have also extended the experiments section to consider a simple imputation method which fills missing values with the mean of the column, which only needs one pass over the data.}
+We have extended the experiments section to consider a simple non-block imputation method relevant to our dataset: mean value imputation.
+This is a common technique, which fills in  missing values
+    with the average observed value. In our case, we implement a non-blocking variant that estimates the mean from the plan's histograms, avoiding the need to perform a full pass over
+    the data before imputing.}
 

### 2017-05-12 paper/experiments.tex
paper: pass over experiments section

@@
 
-For each query, we evaluate \todo{update: three different configurations}: \ProjectName{} optimizing for
-quality ($\alpha=0$), \ProjectName{} optimizing for performance ($\alpha=1$), and imputing
-at the base tables followed by executing the query traditionally.
+For each query, we evaluate \todo{update: three different configurations}: \ProjectName{} optimizing for quality ($\alpha=0$) and \ProjectName{} optimizing for performance ($\alpha=1$) and as a baseline, we fully impute the tables used by each query then run the query on the imputed tables, simulating the process that an analyst would follow to apply existing imputation techniques.
 
 \Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
-queries are an \todo{update/validate: order-of-magnitude faster than imputing on the base tables. We get another order-of-magnitude speedup when optimizing for performance}. This performance
+queries are an \todo{update/validate: order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance}. This performance
 differential means it is feasible
 to explore multiple imputation operations (including more expensive operators) when using
-\ProjectName{}, in contrast to the traditional approach of base table imputation.
+\ProjectName{}, in contrast to the traditional approach of fully imputing the dataset.
 
@@
 \includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
-\caption{Query runtimes with imputation on base tables, \ProjectName{}
-    optimizing for quality ($\alpha=0$), and \ProjectName{} optimizing for
-    performance ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
-    magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
-    imputation can provide another order-of-magnitude speedup at the expense of quality.}
+\caption{Runtimes for the queries in~\Cref{tbl:queries}.
+$\alpha=0$ (resp.
+$\alpha=1$) is \ProjectName{} run at the highest (resp.
+lowest) quality setting.
+Baseline is full table imputation followed by running the query.}
 \todo{update with new figure}   
@@
 
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error
-(SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results when compared to
-imputing on the base tables and executing the query on this cleaned base data.
-This measure indicates how much error is introduced by on-the-fly imputation compared to
-imputation on the base tables. This comparison is the relevant one in practice --- an
-analyst would only be considering the trade-offs in imputation quality and time in the first
-place for dirty input data, for which ground truth is unavailable.
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline.
+This measures the error introduced by on-the-fly imputation as compared to full imputation on the base tables.
+This comparison is the relevant one in practice --- an analyst would only be considering the trade-offs in imputation quality and time in the first place for dirty input data, for which ground truth is unavailable.
 
-In our case, the relevant comparison is against traditional
-imputation on the base table, given the unavailability of ground truth.
+In our case, the relevant comparison is against traditional imputation on the base table, given the unavailability of ground truth.
 
@@
 can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
-to the base-imputation approach is low---\todo{update: between \lowsmapealphazero{} and
-\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces similar
-results to imputation at the base tables. Notably, the results with largest error correspond
+to the baseline is low---\todo{update: between \lowsmapealphazero{} and
+\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with largest error correspond
 most closely to one existing approach of dropping all \nullv{}s.
@@
 We also calculate the number of tuples used to produce each aggregate output.
-The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples
-used when running the query on the imputed base table.
+The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples used when running the query on the imputed base table.
 This shows that when optimizing for performance, not quality, many tuples are dropped.
@@
 
+Queries optimized
+    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
+    efficiency ($\alpha=1$). The count fraction column shows the number of tuples used in calculating each aggregate
+     as a fraction of the number of tuples used when running the same query after imputing on the base table.
+    A lower count share reflects more potential for errors.
+
 \begin{table}
@@
 \input{tables/perf_summary.tex}
-\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
-    parameterizations. To calculate SMAPE, \ProjectName{} results are compared to query results returned
-    from executions with base table imputation. Queries optimized
-    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
-    efficiency ($\alpha=1$). The count fraction column shows the number of tuples used in calculating each aggregate
-     as a fraction of the number of tuples used when running the same query after imputing on the base table.
-    A lower count share reflects more potential for errors.}
+\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$ parameterizations, as compared to the baseline.}
 \todo{update with new figure}
@@
 \caption{\ProjectName{}'s final Pareto frontier for \Cref{q1} and \Cref{q6}. 
-For clarity, we have removed from the plot plans that provide nearly indistinguishable tradeoffs in either dimension. }
+For clarity, we have removed from the plot plans that provide nearly indistinguishable tradeoffs in either dimension.}
 \label{fig:pareto-frontiers}
@@
 
-\subsubsection{Runtime of Base-table Imputation}
-In many real-world cases, applying the imputation step at the base table is prohibitively
-expensive.
-To illustrate the increasing difficulty of such an approach as datasets scale, we run the following query over the ACS dataset:
+\subsubsection{Baseline Runtime}
+In many real-world cases, applying imputation to an entire dataset is prohibitively expensive.
+For example, if we run CART imputation on the full ACS dataset, it completes in \acsbaseresultminutes{}.
+
+In contrast, \ProjectName{} executes the following query:
 \begin{lstlisting}[breaklines]
-SELECT AVG(c0) FROM acs_dirty;
+SELECT AVG(c0) FROM acs;
 \end{lstlisting}
-Applying the imputation operation to the base table is extremely expensive, as it imputes over all attributes in all rows, in our case
-completing in \acsbaseresultminutes{}. In contrast, \ProjectName{} executes a quality-optimized version
-in \acsimputedbzeroresult{} and a runtime-optimized version in \acsimputedboneresult{}. This highlights the potential
-benefit of using our system for early data exploration. Note that the performance differential relative to base table imputation
-would likely widen further if selection predicates were added to the query.
+over the ACS dataset (also using CART imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
+The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.
+Adding selection predicates would further reduce the query runtime by reducing the amount of required imputation.
+An analyst could do the same, but tracking the required imputations would get increasingly complicated as the queries became more complex.
+This highlights the benefit of using our system for early data exploration.
 
-For every imputation inserted into the query plan, a new statistical model is instantiated and trained before being used for imputation.
+\todo{Should this be here?}
+For every imputation operator inserted into the query plan, a new statistical model is instantiated and trained before being used for imputation.
 Although it is tempting to further optimize query execution in \ProjectName{} by pre-training imputation models while the analyst begins to make queries,
@@
 \input{tables/planning_summary_exact_pareto.tex}
-\caption{Planning times are sub-second for up to 4 joins. For 5+  joins planning times rise. Many practical queries
-can be formulated with a smaller number of joins, but to support more complex queries \ProjectName{} provides an
-optimization that maintains approximate Pareto sets for queries with many joins. This reduces the 29 second
-planning time shown for 7 joins by approximately 45\%, on average.
-\todo{update with new table and values}
-}
+\caption{Planning times for varying numbers of joins. \todo{update with new table and values}}
+  % Planning times are sub-second for up to 4 joins. For 5+  joins planning times rise. Many practical queries
+% can be formulated with a smaller number of joins, but to support more complex queries \ProjectName{} provides an
+% optimization that maintains approximate Pareto sets for queries with many joins. This reduces the 29 second
+% planning time shown for 7 joins by approximately 45\%, on average.
 \label{table:planning-times}
@@
 However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
-We explored the limits our planning algorithm by constructing a series of queries with increasing number of joins.
-We used the CDC tables and evaluated increasing number of joins across tables.
-Planning times for queries involving 1--5 joins, which is a practical value for real world exploratory queries, take less than a second.
-\todo{update: For queries with 7 joins, planning time increases substantially to 29 seconds, on average. We introduced an optimization that reduces
-the plan space by maintaining only approximations of the Pareto set for queries involving a large number of joins.
-With this optimization, the same set of queries have an average planning time of 16 seconds (2 seconds SE).}
+We explored the limits our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
+Planning times for queries involving 1--5 joins---a practical value for real world exploratory queries---are less than 1 second.
+\todo{update: For queries with 7 joins, planning time increases substantially to 29 seconds, on average.
+  It is possible to reduce planning time by approximating the Pareto set, particularly for queries with a large number of joins.
+  With this optimization, the queries with 7 joins have an average planning time of 16 seconds (2 seconds SE).}
 

### 2017-05-12 paper/pareto_plot.py
plot for planning

@@ -1,30 +1,76 @@
 import pandas as pd
 import matplotlib.pyplot as plt
+import numpy as np
+import sys
 ANNOTATE = False
 
-df = pd.read_csv('tables/pareto_frontiers.csv')  
-df['norm_time'] = df.groupby('query')['time'].apply(lambda x: x / x.max())
-df['id'] = range(0, df.shape[0])
-interesting_queries = [1, 6] # 2 is also nice, but want one from each set of experiments
-# plans that are different mainly due to floating point precision, collapse
-collapse = [0, 1, 2, 4, 7] + [31, 33]
-fig, axes = plt.subplots(nrows=1, ncols=2, sharey=True, sharex=True, figsize=(8,4))
-plt.subplots_adjust(hspace=0.2)
-for i, q in enumerate(interesting_queries):
-  data  = df[(df['query'] == q) & ~df['id'].isin(collapse)]
-  ax = axes[i]
-  data.plot(kind='scatter', x='norm_time', y = 'penalty', ax=ax, s=20)
-  if ANNOTATE:
-    for row in data.itertuples():
-      ax.annotate('%d' % row.id, (row.norm_time, row.penalty), (row.norm_time * 1.01, row.penalty * 1.01))
-  ax.set_title('Query %d' % q)
-  # damn you matplotlib
-  ax.set_xlabel('')
-  ax.set_ylabel('')
+def plot_frontier():
+  df = pd.read_csv('tables/pareto_frontiers.csv')
+  df['norm_time'] = df.groupby('query')['time'].apply(lambda x: x / x.max())
+  df['id'] = range(0, df.shape[0])
+  interesting_queries = [1, 6] # 2 is also nice, but want one from each set of experiments
+  # plans that are different mainly due to floating point precision, collapse
+  collapse = [0, 1, 2, 4, 7] + [31, 33]
+  fig, axes = plt.subplots(nrows=1, ncols=2, sharey=True, sharex=True, figsize=(8,4))
+  plt.subplots_adjust(hspace=0.2)
+  for i, q in enumerate(interesting_queries):
+    data  = df[(df['query'] == q) & ~df['id'].isin(collapse)]
+    ax = axes[i]
+    data.plot(kind='scatter', x='norm_time', y = 'penalty', ax=ax, s=20)
+    if ANNOTATE:
+      for row in data.itertuples():
[…]

### 2017-05-12 rebuttal.tex
added planning time plot etc to paper and discussion in rebuttal

@@
       We explored the planning times for queries with increasing number of joins. The standard algorithm presented in the original version submitted handles
-      up to 5 joins, with columns requiring imputation in all tables, with planning time of approximately one second. We added a small optimization to help reduce
+      up to 5 joins, with columns requiring imputation in all tables, with planning time of approximately one second or less. We added a small optimization to help reduce
       planning times for a larger number of joins. We maintain an approximation of the Pareto sets, rather than the exact set, by only adding non-dominated plans that
-      are sufficiently different. This helps reduce some of the increases in planning time for a larger number of joins.}
+      are sufficiently different. This optimization reduces planning time linearly with respect to the reduction in the size of the final frontier. Users can increase
+      the level of approximation to reduce planning times further.}
 

### 2017-05-12 paper/experiments.tex
added planning time plot etc to paper and discussion in rebuttal

--- 2017-05-12 paper/experiments.tex
@@
 \subsubsection{Planning Times}
-\begin{table}
+\begin{figure}
 \centering
-\input{tables/planning_summary_exact_pareto.tex}
-\caption{Planning times are sub-second for up to 4 joins. For 5+  joins planning times rise. Many practical queries
-can be formulated with a smaller number of joins, but to support more complex queries \ProjectName{} provides an
-optimization that maintains approximate Pareto sets for queries with many joins. This reduces the 29 second
-planning time shown for 7 joins by approximately 45\%, on average.
-\todo{update with new table and values}
+\includegraphics[scale=0.4]{figures/planning_times_plot.png}
+\caption{Planning times are sub-second for up to 5 joins. For more joins, we implement an approximation to Pareto sets.
+This reduces planning time linearly with respect to the reduction in the final frontier size. Users can increase the level of
+approximation for more aggressive planning reductions.
 }
-\label{table:planning-times}
-\end{table}
+\label{fig:planning-times}
+\end{figure}
 
-\todo{Update: Planning times are reasonable, with a median planning time of \planningmediantime{} ms. over the benchmark queries, constituting \planningruntimepercent{}\% of total runtime.
+\todo{Update: Planning times for our benchmark queries are reasonable, with a median planning time of \planningmediantime{} ms.
+over the benchmark queries, constituting \planningruntimepercent{}\% of total runtime.
 In all cases, the optimizer returned a query plan within \planningmaxtime{} ms.}
@@
 However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
-We explored the limits our planning algorithm by constructing a series of queries with increasing number of joins.
-We used the CDC tables and evaluated increasing number of joins across tables.
-Planning times for queries involving 1--5 joins, which is a practical value for real world exploratory queries, take less than a second.
-\todo{update: For queries with 7 joins, planning time increases substantially to 29 seconds, on average. We introduced an optimization that reduces
-the plan space by maintaining only approximations of the Pareto set for queries involving a large number of joins.
-With this optimization, the same set of queries have an average planning time of 16 seconds (2 seconds SE).}
-
+\Cref{fig:planning-times} shows the average planning time on a log scale for queries of the CDC data with an increasing number of joins. As the number of tables requiring imputation grows,
+so does the search space and corresponding planning time. Planning time for queries involving 1--5 joins are sub-second. While a useful subset of exploratory queries can be expressed
+with a smaller number of joins, we extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
+displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. Users can increase the level of approximation to reduce planning times further.
 %\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

### 2017-05-12 paper/experiments.tex

--- 2017-05-12 paper/experiments.tex
@@
 \subsubsection{Planning Times}
-\begin{table}
+\begin{figure}
 \centering
-\input{tables/planning_summary_exact_pareto.tex}
-\caption{Planning times for varying numbers of joins. \todo{update with new table and values}}
-  % Planning times are sub-second for up to 4 joins. For 5+  joins planning times rise. Many practical queries
-% can be formulated with a smaller number of joins, but to support more complex queries \ProjectName{} provides an
-% optimization that maintains approximate Pareto sets for queries with many joins. This reduces the 29 second
-% planning time shown for 7 joins by approximately 45\%, on average.
-\label{table:planning-times}
-\end{table}
+\includegraphics[scale=0.4]{figures/planning_times_plot.png}
+\caption{Planning times are sub-second for up to 5 joins. For more joins, we implement an approximation to Pareto sets. Users can increase the level of
+approximation for more aggressive planning reductions.
+}
+\label{fig:planning-times}
+\end{figure}
 
-\todo{Update: Planning times are reasonable, with a median planning time of \planningmediantime{} ms. over the benchmark queries, constituting \planningruntimepercent{}\% of total runtime.
+\todo{Update: Planning times for our benchmark queries are reasonable, with a median planning time of \planningmediantime{} ms.
+over the benchmark queries, constituting \planningruntimepercent{}\% of total runtime.
 In all cases, the optimizer returned a query plan within \planningmaxtime{} ms.}
@@
 However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
-We explored the limits our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
-Planning times for queries involving 1--5 joins---a practical value for real world exploratory queries---are less than 1 second.
-\todo{update: For queries with 7 joins, planning time increases substantially to 29 seconds, on average.
-  It is possible to reduce planning time by approximating the Pareto set, particularly for queries with a large number of joins.
-  With this optimization, the queries with 7 joins have an average planning time of 16 seconds (2 seconds SE).}
-
+We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
+Planning times (~\Cref{fig:planning-times}) for queries involving 1--5 joins---a practical value for a practical subset of real world exploratory queries---are less than 1 second.
+We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
+displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. Users can increase the level of approximation to reduce planning times further.
 %\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

### 2017-05-12 paper/experiments.tex
paper: remove extra paragraph
removed panning time plot, mentioned reductions in text

@@
 
-\subsubsection{Planning Times}
-\begin{figure}
-\centering
-\includegraphics[scale=0.4]{figures/planning_times_plot.png}
-\caption{Planning times are sub-second for up to 5 joins. For more joins, we implement an approximation to Pareto sets. Users can increase the level of
-approximation for more aggressive planning reductions.
-}
-\label{fig:planning-times}
-\end{figure}
+%\subsubsection{Planning Times}
+%\begin{figure}
+%\centering
+%\includegraphics[scale=0.4]{figures/planning_times_plot.png}
+%\caption{Planning times are sub-second for up to 5 joins. For more joins, we implement an approximation to Pareto sets. Users can increase the level of
+%approximation for more aggressive planning reductions.
+%}
+%\label{fig:planning-times}
+%\end{figure}
 
@@
 We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
-Planning times (~\Cref{fig:planning-times}) for queries involving 1--5 joins---a practical value for a practical subset of real world exploratory queries---are less than 1 second.
+Planning times (~\Cref{fig:planning-times}) for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
 We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
-displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. Users can increase the level of approximation to reduce planning times further.
+displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
+in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
 %\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

### 2017-05-12 rebuttal.tex
added comment about bayesdb running times for single model fitting

@@
       We are not aware of another database technology which offers on-the-fly imputation optimization, so there is no direct comparison for the planning algorithm.
+      However, we ran a simple experiment using BayesDB to compare possible imputation times. In order to make the comparison as fair as possible in terms of running
+      time, we reduced the number of models that BayesDB instantiates to a single model per base table. We fit each model with 100 iterations, in an optimized setting.
+      The fitting times on the CDC tables ranged between 119 and 320 seconds (per table). The FCC model fitting took 237 seconds. The user would then still need to use these models to
+      perform imputation.  The underlying modeling technique for BayesDB is quite sophisticated, so the high running time is not a surprise, but should provide some context
+      for the challenges faced by users trying to explore new datasets. In a real world scenario, the user would instantiate more than a single model, increasing these running
+      times further.
       % Our evaluation compares our technique to an imputation of the full dataset.

### 2017-05-14 simpledb/test/experiments/analyze_experiments.py
Use fire with analyze_experiments

@@ -9,6 +9,8 @@ import os
 import glob
 import sys
 
+import fire
+
 from collections import defaultdict
 
 try:
@@ -220,25 +222,31 @@ def write_perf_summary(experiments_dir, output_dir):
 
     return perf
 
-def analyze_join_planning(experiment_dir, output_dir):
-  if not os.path.isdir(output_dir):
-      os.makedirs(output_dir)
-  # time data
-  data = get_timing_results(experiment_dir, joins=True)
-  # drop warmup iterations
-  data = drop_warmup(data, ['njoins', 'query', 'alpha'], drop=10)
-  # average planning time and se by number of joins
-  # alpha doesn't have a real impact on the planning time (as we would expect), so don't aggregate with it
-  planning_summary = data.groupby('njoins')['plan_time'].agg({'mean': np.mean, 'std': np.std}).reset_index()
-  planning_summary_latex = planning_summary.copy()
-  planning_summary_latex = planning_summary_latex[['njoins', 'mean', 'std']]
-  planning_summary_latex = planning_summary_latex.rename(columns={'njoins': '# of Joins', 'mean': 'Avg (ms)', 'std': 'SE'})
-  planning_summary_latex.to_latex(os.path.join(output_dir, 'plan_summary.tex'), index=False, float_format='%.2f')
+def analyze_join_planning(experiment_dir, output_dir=""):
+    if not output_dir:
+        output_dir = os.path.join(experiment_dir, "analysis")
+
+    if not os.path.isdir(output_dir):
+        os.makedirs(output_dir)
+    # time data
+    data = get_timing_results(experiment_dir, joins=True)
+    # drop warmup iterations
+    data = drop_warmup(data, ['njoins', 'query', 'alpha'], drop=10)
+    # average planning time and se by number of joins
+    # alpha doesn't have a real impact on the planning time (as we would expect), so don't aggregate with it
+    planning_summary = data.groupby('njoins')['plan_time'].agg({'mean': np.mean, 'std': np.std}).reset_index()
+    planning_summary_latex = planning_summary.copy()
+    planning_summary_latex = planning_summary_latex[['njoins', 'mean', 'std']]
+    planning_summary_latex = planning_summary_latex.rename(columns={'njoins': '# of Joins', 'mean': 'Avg (ms)', 'std': 'SE'})
+    planning_summary_latex.to_latex(os.path.join(output_dir, 'plan_summary.tex'), index=False, float_format='%.2f')
 
 
[…]

### 2017-05-14 paper/experiments.tex
Move some text inside captions

@@
 \includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
-\caption{Runtimes for the queries in~\Cref{tbl:queries}.
-$\alpha=0$ (resp.
-$\alpha=1$) is \ProjectName{} run at the highest (resp.
-lowest) quality setting.
-Baseline is full table imputation followed by running the query.}
+\caption{Runtimes for the queries in~\Cref{tbl:queries} using reference
+    chained-equation decision trees algorithm. $\alpha=0$ (resp. $\alpha=1$) is
+    \ProjectName{} run at the highest (resp. lowest) quality setting. Baseline
+    is full table imputation followed by running the query. Each query and each
+    value of $\alpha$ was run 200 times.}
 \todo{update with new figure}   
@@
 %planning.
-
-All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
-four 2.9 GHz Intel Xeon E5--2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.
-Each query and each value of $\alpha$ was run 200 times, after discarding a warm-up period.
 
@@
 This comparison is the relevant one in practice --- an analyst would only be considering the trade-offs in imputation quality and time in the first place for dirty input data, for which ground truth is unavailable.
-
-In our case, the relevant comparison is against traditional imputation on the base table, given the unavailability of ground truth.
 
@@
 
+\begin{table}
+\centering
+\input{tables/perf_summary.tex}
+\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$ parameterizations, as compared to the baseline.
 Queries optimized
@@
      as a fraction of the number of tuples used when running the same query after imputing on the base table.
-    A lower count share reflects more potential for errors.
-
-\begin{table}
-\centering
-\input{tables/perf_summary.tex}
-\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$ parameterizations, as compared to the baseline.}
+    A lower count share reflects more potential for errors.}
 \todo{update with new figure}

### 2017-05-14 paper/experiments.tex
Update count fraction, planning times

@@
     value of $\alpha$ was run 200 times.}
-\todo{update with new figure}   
 \label{fig:runtimes}
@@
     A lower count share reflects more potential for errors.}
-\todo{update with new figure}
 \label{table:smape}
@@
 In many real-world cases, applying imputation to an entire dataset is prohibitively expensive.
-For example, if we run CART imputation on the full ACS dataset, it completes in \acsbaseresultminutes{}.
+For example, if we run chained-equation decision trees imputation on the full ACS dataset, it completes in \acsbaseresultminutes{}.
 
@@
 \end{lstlisting}
-over the ACS dataset (also using CART imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
+over the ACS dataset (also using chained-equation decision trees imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
 The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.
@@
 
-\todo{Update: Planning times for our benchmark queries are reasonable, with a median planning time of \planningmediantime{} ms.
-over the benchmark queries, constituting \planningruntimepercent{}\% of total runtime.
-In all cases, the optimizer returned a query plan within \planningmaxtime{} ms.}
-The choice of $\alpha$ has no impact on planning times, because query selection is done after collecting the Pareto frontier.
+\todo{Update: Planning times for our benchmark queries are reasonable, with mean planning
+times across the benchmark queries ranging from \planninglowtime{} ms to
+\planninghightime{} ms, constituting \planninglowtimepct{} to \planninghightimepct{} of
+total runtime, respectively.
+
+In all cases, the optimizer returned a query
+plan within \planningmaxtime{} ms.} The choice of $\alpha$ has no impact on planning times,
+because query selection is done after collecting the Pareto frontier.
 

### 2017-05-14 paper/main.tex
paper: mark sections for sam to review
Update count fraction, planning times

--- 2017-04-25 paper/main.tex
@@
-\documentclass{vldb}
+\documentclass[preprint]{vldb}
 \usepackage{balance}
@@
 \newcommand{\todobox}[2]{{\color{red}\fbox{\parbox{\columnwidth}{\textbf{TODO:} \textit{#1}\\#2}}}}
+\newcommand{\review}[1]{{\color{blue}#1}}
 
@@
 \newcommand{\acsimputedboneresult}{1 second}
-\newcommand{\lowxalphazero}{10}
-\newcommand{\lowsmapealphazero}{0}
-\newcommand{\highxalphazeroexacs}{200}
-\newcommand{\highxalphaoneexacs}{1000}
-\newcommand{\highsmapealphaoneexacs}{24}
-\newcommand{\highxalphaone}{16000}
+\newcommand{\lowxalphazero}{10} % the smallest speed up, comparing a=0 to base table imputation
+\newcommand{\highxalphazeroexacs}{140} % the largest speed up, comparing a=0 to base table imputation 
+\newcommand{\highxalphaoneexacs}{1400} % the largest speed up, comparing a=0 OR a=1 to base table imputation 
+\newcommand{\lowsmapealphazero}{0} % the smallest SMAPE difference, comparing a=0 to base table imputation
+\newcommand{\highsmapealphazero}{6} % the largest SMAPE difference, comparing a=0 to base table imputation
+\newcommand{\highsmapealphaoneexacs}{20} % the largest SMAPE difference, comparing a=1 to base table imputation
+%\newcommand{\highxalphaone}{16000}
 \newcommand{\nullv}{NULL}
-\newcommand{\planningruntimepercent}{1.8}
-\newcommand{\planningmediantime}{1.5}
-\newcommand{\planningmaxtime}{34}
+
+\newcommand{\planninglowtime}{0.7}
+\newcommand{\planninglowtimepct}{0}
+\newcommand{\planninghightime}{2.8}
+\newcommand{\planninghightimepct}{14}
+
+\newcommand{\planningmaxtime}{34} % TODO
 
@@
   traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
-  Finally, we show that while dropping tuples with missing values that fail query constraints discards 3--88\% of the data, on-the-fly imputation loses only 0--10\%.
+  Finally, we show that while dropping tuples with missing values that fail query constraints discards 6--78\% of the data, on-the-fly imputation loses only 0--21\%.
 \end{abstract}
@@
 \input{conclusion}
+
+\section{Acknowledgments}
+The authors would like to acknowledge the authors of~\cite{akande2015empirical} for giving us a cleaned copy of the 2012 ACS PUMS.
+
 \bibstyle{abbrv}

### 2017-05-14 rebuttal.tex
rebut: add intro

@@
 \begin{document}
+\section{Introduction}
+The authors would like to thank the reviewers for their helpful and insightful comments.
+We have taken steps to address each comment and we believe that the paper is significantly improved.
+In this document, we give a point-by-point response to the comments from each reviewer.
+
+The reviewer's original comments are in black and our responses are in \resp{blue}.
+
+In our draft, new text is highlighted in {\color{blue}blue} and changed text is highlighted in {\color{red}red}.
+
 \section{Multiple reviewers}

### 2017-05-14 rebuttal.tex
rebut: update response to w2

@@
   \resp{
-  We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation. We also added
- details on how to pick the heuristic functions that define the API for impute operators. We provide details on how these functions are defined
- for all the methods used in our evaluation. 
+    We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation.
+    These methods are in addition to the original imputation method (CART), which is based on regression trees.
+    The regression tree model incorporates correlations between attributes, so should be able to take advantage of any functional dependencies to make better predictions.
+    Interpolation based imputation is useful for data where the ordering of tuples is significant, like time series, but is not well suited for survey data where the ordering of tuples is arbitrary.
+
+    We also added details on how to pick the cost functions that define the API for impute operators.
+    We provide details on how these functions are defined for all the methods used in our evaluation.
+
+    Our optimizer (like any query optimizer) is sensitive to the quality of its estimates.
+    We believe that the cost functions we have chosen lead to good results overall, but there may be cases where our estimates lead to bad outcomes.
+    It is an interesting and difficult problem to accurately estimate the quality of a regression model before running the model.
+    Any work in this area could be used immediately to improve our algorithm.
   }

### 2017-05-14 rebuttal.tex
rebut: add overall changes

@@
 
-The reviewer's original comments are in black and our responses are in \resp{blue}.
+In the rebuttal, the reviewer's original comments are in black and our responses are in \resp{blue}.
 
@@
 
-\section{Multiple reviewers}
+\section{Overall changes}
 \begin{itemize}
@@
 
-  \resp{We've compressed many of the figures and removed whitespace to make room for additional material.}
+  \resp{Several reviewers commented on the amount of whitespace, so we've compressed many of the figures and removed whitespace to make room for additional material.}
+
+\item \resp{
+    We have improved our system since the paper was submitted by adding imputation for categorical variables.
+    This has a minor impact on the experimental results.
+    It improves the ability of the regression tree model to perform imputation by making these variables available to the model.
+  }
 \end{itemize}

### 2017-05-14 paper/algorithms/histogram-transform.tex
paper: removed histogram helpers and tried to rework into simpler form
paper: fixed general typos, added clearer todos
paper: revise histogram algorithm
paper: pass over cardinality estimation section
paper: highlight major changes & additions

@@
-%\begin{algorithm}
-  \begin{algorithmic}
-    \Require{$\mathbb{H}$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
-    \Ensure{Returns an updated map $\mathbb{H'}$}
+\newcommand{\InlineIte}[3]{#2\ \textbf{if}\ #1\ \textbf{else}\ #3}
+\renewcommand{\algorithmicindent}{1em}
 
-    \Function{UpdateHistograms}{$\mathbb{H}, op$}
-    	\If{$op \not\in \{\delta, \mu, \sigma, \bowtie \}$}
-		\Return{$\mathbb{H}$}
-	\EndIf
-	\State $\mathbb{H'} \gets \Call{Copy}{\mathbb{H}}$
-    	\If{op = $\delta_C$}
-		\For{$c \in C$}
-			\State $\mathbb{H'}[c][null] \gets 0$
-		\EndFor
-	\ElsIf{op = $\mu_C$}
-		\For{$c \in C$}
-			\State \Call{Add}{$\mathbb{H'}[c], \mathbb{H'}[c][null]$}
-		\EndFor
-	\ElsIf{op = $\sigma$}
-		\For{$c \in dom(\mathbb{H'})$}
-			\State \Call{ScaleBy}{$\mathbb{H'}[c],  \Call{Selectivity}{\sigma}$} 
-		\EndFor
+\review{
+\begin{algorithmic}[1]
+    \Require{
+    \Statex
+    \begin{itemize}[leftmargin=*,noitemsep]
+    \item $H$: A map from attribute names to histograms.
+    \item $op$: A relational operator.
+    \end{itemize}}
+
+  \Function{UpdateHistograms}{$H,\ op$}
+  \If{op = $\delta_C$}\label{lst:drop-start}
+  \State Set null count to zero for $H[c]$ where $c \in C$.
+  \State Scale $H[c]$ s.t. $\Call{Card}{H[c]} = \min_d \Call{Card}{H[d]}$, $\forall c$.\label{lst:drop-end}
+  \ElsIf{op = $\mu_C$}\label{lst:impute-start}
+  \State Scale $H[c]$ s.t. $\Call{Card}{H[c]}$ is unchanged and null
+  \State count is zero, $\forall c$.\label{lst:impute-end}
+  \ElsIf{op = $\sigma_\phi$}\label{lst:std-start}
+  \State Scale $H[c]$ by $\Call{Selectivity}{\phi, H}$, $\forall c$.
   \ElsIf{op = $\bowtie_\phi$}
-		\For{$c \in dom(\mathbb{H'})$}
-			\State \Call{ScaleTo}{$\mathbb{H'}[c], \Call{Cardinality}{\bowtie_{\phi}}$}
-		\EndFor
-	\EndIf
-	\Return{$\mathbb{H'}$}
+  \State Scale $H[c]$ s.t. $\Call{Card}{H[c]} = \Call{Card}{\bowtie_\phi}$, $\forall c$.
+  \EndIf\label{lst:std-end}
+  \State \Return $H$
   \EndFunction
 \end{algorithmic}
-  \caption{An algorithm for in-plan histogram updates}
+}
+\caption{\review{An algorithm for in-plan histogram updates}}
 \label{algo:histogram-transformation}
-%\end{algorithm}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "../main"
+%%% End:

### 2017-05-14 paper/examples.tex
paper: highlight major changes & additions

@@
 
+\review{
 An alternative strategy involves starting with $\alpha=1$ and iterating to lower values, as long as execution
@@
 the user to adjust their choice of $\alpha$ to uniquely select amongst these plans.
+}
 

### 2017-05-14 paper/experiments.tex
paper: highlight major changes & additions

@@
 
+\review{
 \Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm.
@@
 can be exposed to the user to help in guiding their workflow.
+}
 
@@
 \includegraphics[scale=0.4]{figures/pareto_frontiers_plot.png}
-\caption{\ProjectName{}'s final Pareto frontier for \Cref{q1} and \Cref{q6}. 
-For clarity, we have removed from the plot plans that provide nearly indistinguishable tradeoffs in either dimension.}
+\caption{\review{
+    \ProjectName{}'s final Pareto frontier for \Cref{q1} and \Cref{q6}. 
+    For clarity, we have removed from the plot plans that provide nearly indistinguishable tradeoffs in either dimension.
+}}
 \label{fig:pareto-frontiers}
@@
 
-In contrast, \ProjectName{} executes the following query:
-\begin{lstlisting}[breaklines]
-SELECT AVG(c0) FROM acs;
-\end{lstlisting}
-over the ACS dataset (also using chained-equation decision trees imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
+In contrast, \ProjectName{} executes the query \lstinline{SELECT AVG(c0) FROM acs} over the ACS dataset (also using chained-equation decision trees imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
 The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.
@@
 
+\review{
 However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
@@
 in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
+}
 %\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

### 2017-05-14 rebuttal.tex
rebut: minor response
paper: highlight major changes & additions
rebut: pass over rebuttal

@@
 
-In our draft, new text is highlighted in {\color{blue}blue} and changed text is highlighted in {\color{red}red}.
+In our draft, new or significantly changed text is highlighted in {\color{blue}blue}.
 
@@
   \item Although it doesn't consider cost or the drop operator, a similar technique appears in Yang et. al. "Lenses: An On-Demand Approach to ETL." A comparison would be nice.
+
    \resp{
-   Yang's work is definitely related, and we've added it to our related work section. However, it is worth noting that their decision process for application of lenses is at the tuple-level,
-   rather than with respect to the query plan.
+     Yang's work is definitely related, and we've added it to our related work section.
+   Their system helps an analyst prioritize cleaning tasks for a particular query.
+   It uses pretrained regression models as one of a number of repair strategies for missing data rather than creating a query plan which incorporates the training and running of such models.
    }
@@
  \item S1: "Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach between 0 and 24 percent" 24 percent error is pretty bad.
+
+   \resp{See the response to W3 for discussion of Q8.}
 
@@
 Why not just take the same approach as selection and treat it as having negligible cost?
+
     \resp{We mistakenly suggest in S3 that we treat \textit{Drop} as a blocking operator. In
@@
     \item D7. When defining the cost of impute operator, how do you know a priori the number of iterations for the algorithm (given it has to converge)?
-  For the iterative regression impute operatore, we set a predefined number of iterations as an upper bound on its execution. If the imputation converges,
+
+  \resp{For the iterative regression impute operator, we set a predefined number of iterations as an upper bound on its execution. If the imputation converges,
   we can return immediately, otherwise we return after this upper bound is reached. This is a common approach used in iterative optimization techniques, such
   as gradient descent.
+}
 
@@
 
-  \item -- Given you optimize for both performance and quality, Tummer's and Koch's work on multi-objective QO is related.
+  \item Given you optimize for both performance and quality, Tummer's and Koch's work on multi-objective QO is related.
 
@@
 
-\item -- Could be interesting for the optimizer to pick imputation strategy based on the type of the query.
+\item Could be interesting for the optimizer to pick imputation strategy based on the type of the query.
   \end{itemize}

### 2017-05-14 rebuttal.tex
rebut: shrink
pass over rebuttal

@@
 \documentclass{vldb}
+\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
+\usepackage[subtle]{savetrees}
 \usepackage{xcolor}
@@
 The authors would like to thank the reviewers for their helpful and insightful comments.
-We have taken steps to address each comment and we believe that the paper is significantly improved.
-In this document, we give a point-by-point response to the comments from each reviewer.
+We have taken steps to address each comment and we believe that the paper is significantly improved
+as a result. In this document, we give a point-by-point response to the comments from each reviewer.
 
 In the rebuttal, the reviewer's original comments are in black and our responses are in \resp{blue}.
-
 In our draft, new or significantly changed text is highlighted in {\color{blue}blue}.
@@
       using the base table imputation, but if we look at the underlying population the former was calculated with 6 people, while the latter was with 258 people.
+
+      We can trivially extend \ProjectName{} to warn users when the query chosen for execution has a high $\textsc{Penalty}$ estimate, along with the number of tuples that have
+      been dropped (after execution), so that situations such as those observed in Q8 can be identified by the user.
     }
@@
 
-    \resp{We have clarified this in the paper by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and by renaming it to \textit{Penalty}.}
+    \resp{We have clarified this in the paper by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and by renaming it to $\textsc{Penalty}$.}
 
@@
 In other cases, this intermediate table is large and skewed, which causes problems for imputation.
-Avoiding this case requires us to estimate the skew in these intermediate tables.
 
-     To clarify the point about Q8, we've removed the predicate condition $schooldegree \geq 0$. 
-      We originally enumerated strings with positive integers and replaced missing values strings with 0, rather than perform imputation on them, so every enumerated value
-      satisfied this predicate trivially. We have changed our evaluation to impute string enumerations as well and dropped this predicate.
+    Avoiding this case requires us to estimate the skew in these intermediate tables, which would be in itself an interesting and challenging research problem. If we had improved intermediate
+    estimates, we could incorporate those to improve our system.
       }
@@
       identical in this case and the number of tuples returned can be seen
-      directly in Table 4.}
-
-    \todo{Do we need to address this?:
-    For SUM queries, that may depend more directly on the number of rows returned than the
-      quality of imputations for any individual row, the effect of different COUNT results
-      dominates the computation AVG*COUNT. How do we handle this?}
+      directly in Table 4. We would expect benchmark queries that use aggregates such as SUM 
+      to work well in our system, with errors proportional to the fraction of relevant tuples retrieved
+      and the AVG value estimated.}
 
@@
       often does not capture the complex reasons for missingness in real world datasets.
-      We update the discussion in S4.4.1 to emphasize this point.}
+      We updated the discussion in S4.4.1 to emphasize this point.}
 
@@
       We are not aware of another database technology which offers on-the-fly imputation optimization, so there is no direct comparison for the planning algorithm.
-      However, we ran a simple experiment using BayesDB to compare possible imputation times. In order to make the comparison as fair as possible in terms of running
-      time, we reduced the number of models that BayesDB instantiates to a single model per base table. We fit each model with 100 iterations, in an optimized setting.
-      The fitting times on the CDC tables ranged between 119 and 320 seconds (per table). The FCC model fitting took 237 seconds. The user would then still need to use these models to
-      perform imputation.  The underlying modeling technique for BayesDB is quite sophisticated, so the high running time is not a surprise, but should provide some context
-      for the challenges faced by users trying to explore new datasets. In a real world scenario, the user would instantiate more than a single model, increasing these running
-      times further.
+%      However, we ran a simple experiment using BayesDB to compare possible imputation times. In order to make the comparison as fair as possible in terms of running
+%      time, we reduced the number of models that BayesDB instantiates to a single model per base table. We fit each model with 100 iterations, in an optimized setting.
+%      The fitting times on the CDC tables ranged between 119 and 320 seconds (per table). The FCC model fitting took 237 seconds. The user would then still need to use these models to
+%      perform imputation.  The underlying modeling technique for BayesDB is quite sophisticated, so the high running time is not a surprise, but should provide some context
+%      for the challenges faced by users trying to explore new datasets. In a real world scenario, the user would instantiate more than a single model, increasing these running
+%      times further.
       % Our evaluation compares our technique to an imputation of the full dataset.
@@
    \resp{
-    \todo{this is a bad response...}
-      We felt that the overhead of reimplementing an existing cost model would not be worth the benefit.
-      However, we added a new section that discusses in detail the main components for \ProjectName{}, so that developers of existing systems can use this as a roadmap
-      for implementation.
+   We've added a section that describes in detail the components that need to be added to an existing database system to implement our technique. 
+   The details come directly from our implementation of \ProjectName{} on top of SimpleDB, a popular teaching database used at MIT, University of Washington, and Northwestern University, 
+   amongst others. These extensions could further exploit any existing sophistication in the underlying system for tasks such as execution cost estimates.
 
-      \todo{More sophisticated cost formulas from exiting systems could further improve our time estimates for traditional operators. In this case, we focused on developing the heuristics
-        relevant for the new operators, at the expense of simplifying implementation of common formulas, as we felt that was core to our contribution.}
+   Extending a rule-based optimizer to perform dynamic imputation would certainly be an interesting idea. However, the planning algorithm presented benefits from the tight
+   interaction between traditional query planning and imputation operator placement. The latter can influence aspects such as join ordering and cardinality estimates for subplans.
+   Incorporating these into a rule-based system would be possible, but we believe the plan selected through this approach will be suboptimal, compared to that chosen through
+   our integrated planning approach.
    }
@@
     \resp{In our intended usage, the user starts by running with $\alpha = 0$ and increases $\alpha$ if the query takes too long to complete.
-      \todo{
-        We agree that having a way of providing an estimate on error would be interesting. Since error in this case relates to the imputation values
-        generated by the whole-table imputation (as there is no ground truth), this would likely need to be on a case-by-case basis for distinct
-        instantiations of the impute operator. 
-      }
+
+     We agree that having a way of providing an estimate on error, based on $\alpha$, would be both useful and interesting to develop.
+     Providing error estimates further requires a reasonable notion of ground-truth. For evaluation, we've used the values imputed on the 
+     tables when applying whole-table imputation. Based on our evaluation, higher values of $\alpha$ correctly identify increased error
+     within a query, however, extending that to be predictive is not something we implemented in the current system.
      }
@@
 
-    \resp{Yes.}
+    \resp{Yes, we have now highlighted this for clarity in the paper as well.}
 
@@
 This kind of imputation isn't useful for survey data, so we didn't consider it for our experiments.
-We have extended the experiments section to consider a simple non-block imputation method relevant to our dataset: mean value imputation.
+We have extended the experiments section to consider a simple non-blocking imputation method relevant to our dataset: mean value imputation.
 This is a common technique, which fills in  missing values
@@
       \\
-      We explored the planning times for queries with increasing number of joins. The standard algorithm presented in the original version submitted handles
+      We explored the planning times for queries with increasing number of joins by running additional experiments with up to 8 joins. 
+      The standard algorithm presented in the original version submitted handles
       up to 5 joins, with columns requiring imputation in all tables, with planning time of approximately one second or less. We added a small optimization to help reduce

### 2017-05-14 paper/pareto_plot.py
frontier plots with stylized axis labels to match cost function font

@@ -4,7 +4,15 @@ import numpy as np
 import sys
 ANNOTATE = False
 
+
 def plot_frontier():
+  # setup latex rendering to use \textsc below
+  from matplotlib import rc
+  rc('font',**{'family':'sans-serif','sans-serif':['Helvetica']})
+  ## for Palatino and other serif fonts use:
+  #rc('font',**{'family':'serif','serif':['Palatino']})
+  rc('text', usetex=True)
+  
   df = pd.read_csv('tables/pareto_frontiers.csv')
   df['norm_time'] = df.groupby('query')['time'].apply(lambda x: x / x.max())
   df['id'] = range(0, df.shape[0])
@@ -26,9 +34,9 @@ def plot_frontier():
     ax.set_ylabel('')
 
     plt.subplots_adjust(bottom=0.2)
-    fig.suptitle('Pareto Frontiers', size=16)
-    fig.text(0.5, 0.04, 'Time estimate/Max(Time estimate)', ha='center')
-    fig.text(0.04, 0.5, 'Penalty Estimate', va='center', rotation='vertical')
+    fig.suptitle('Planner Pareto Frontiers', size=14)
+    fig.text(0.5, 0.04, r'$\textsc{Time}/\textsc{Max}(\textsc{Time})$', ha='center')
+    fig.text(0.04, 0.5, r'$\textsc{Penalty}$', va='center', rotation='vertical')
     fig.savefig('figures/pareto_frontiers_plot.png')
 
 

### 2017-05-15 paper/algorithm.tex
paper: mark sections for sam to review
paper: highlight major changes & additions
paper: stop using citations as nouns

@@
 
-\subsection{Imputation operators}
+\subsection{Imputation Operators}
 \label{sec:operators}
@@
 
-\subsection{Search space}
+\subsection{Search Space}
 \label{sec:search-space}
@@
 
-\subsection{Imputation placement}
+\subsection{Imputation Placement}
 \label{sec:placement}
@@
 
-\subsection{Cost model}
+\subsection{Cost Model}
 \label{sec:cost-model}
@@
 
-This set $Q$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all feasible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.
+This set $Q$ will be the Pareto frontier~\cite{pareto1964cours} of $S$, and it contains the best option for all feasible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.
 
@@
 
+\review{
 \subsubsection{Cardinality estimation}
@@
 
-\subsection{Imputation strategies and their API}
+\subsection{Imputation Strategies and their API}
 \label{sec:imputation}
@@
 \end{equation*}
+}
 
-\subsection{Query planning}
+\subsection{Query Planning}
 \label{sec:planning}
@@
 
-In order to support a large number of joins, with reasonable planning times, \ProjectName{} can maintain approximations of the Pareto sets. 
+\review{
+In order to support a large number of joins, with lower planning times, \ProjectName{} can maintain approximations of the Pareto sets.
 A non-dominated plan can be approximated by an existing plan in the set, if the distance between them is smaller than some predefined bound.
@@
 Planning times across both exact and approximate Pareto frontiers are discussed further in \Cref{sec:results}.
+}
 

### 2017-05-15 paper/experiments.tex
frontier plots with stylized axis labels to match cost function font
paper: stop using citations as nouns

@@
 Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS)
-data kindly provided by the authors of~\cite{akande2015empirical}. Given that the data had
+data kindly provided by \citeauthor*{akande2015empirical}. Given that the data had
 been cleaned, we artificially dirtied it by replacing 40\% of the values uniformly at random with \nullv{} values.
@@
 \review{
-\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm.
+\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}.
+These are the resulting set of plans after pruning dominated plans throughout our algorithm.
 and encapsulate the possible tradeoffs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
@@
 \caption{\review{
-    \ProjectName{}'s final Pareto frontier for \Cref{q1} and \Cref{q6}. 
+    \ProjectName{}'s final planner Pareto frontier for \Cref{q1} and \Cref{q6}. 
     For clarity, we have removed from the plot plans that provide nearly indistinguishable tradeoffs in either dimension.

### 2017-05-15 paper/main.tex
paper: stop using citations as nouns

@@
 \usepackage[noend]{algpseudocode}
-\usepackage[style=numeric,firstinits=true]{biblatex}
+\usepackage[style=numeric,firstinits=true,maxcitenames=2]{biblatex}
 \usepackage{booktabs}
@@
 \section{Acknowledgments}
-The authors would like to acknowledge the authors of~\cite{akande2015empirical} for giving us a cleaned copy of the 2012 ACS PUMS.
+The authors would like to acknowledge \AtNextCite{\defcounter{maxnames}{3}}\citeauthor{akande2015empirical} for giving us a cleaned copy of the 2012 ACS PUMS.
 

### 2017-05-15 paper/relatedwork.tex
paper: stop using citations as nouns

@@
 Imputation of missing values is widely studied in the statistics and machine
-learning communities. As highlighted in~\cite{gelman2006data}, missing data
-can appear for a variety of reasons.
+learning communities. Missing data
+can appear for a variety of reasons~\cite{gelman2006data}.
 It can be missing uniformly at random or conditioned on
@@
 focus on correctly modeling relationships between attributes to account for different forms
-of missingness. For example, \cite{burgette2010multiple} discuss the
-usage of iterative decision trees for imputing missing data.
+of missingness.
+For example, \textcite*{burgette2010multiple} use iterative decision trees for imputing missing data.
 
-\cite{akande2015empirical} analyzes the performance of various
-multiple imputation techniques on the American Community Survey dataset. 
 The computational difficulties of imputing on large base
-tables are well-known and can limit approaches. For example, \cite{akande2015empirical} finds that
-one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
-includes variables with potentially large domains (ten categories in their case).
+tables are well-known and can limit approaches. % \citeauthor{akande2015empirical} analyze the performance of various
+% multiple imputation techniques on the American Community Survey dataset~\cite{akande2015empirical}.
+For example, \textcite*{akande2015empirical} find that one approach (MI-GLM) is prohibitively expensive when attempting to impute on the American Community Survey dataset.
+In this case, the complexity comes from the large domains of the variables in the ACS data (up to ten categories in their case).
 In contrast, \ProjectName{} allows users to specify a trade-off between
 imputation quality and runtime performance, allowing users to perform queries
-directly on the entire dataset. Furthermore,
+directly on datasets which are too large to fully impute quickly. Furthermore,
 the query planner's imputation is guided by the requirements of each specific
@@
 There is a long history in the database community surrounding the
-treatment of \nullv{}. As early as 1973,~\cite{codd1973understanding}
-provided a treatment of the semantics of \nullv{}. Multiple
-papers have described various (at times conflicting) treatments
-of \nullv{}s~\cite{grant1977null}. \ProjectName's main design invariant---no relational operator
+treatment of \nullv{}.
+Multiple papers have described various (at times conflicting) treatments of \nullv{}s~\cite{codd1973understanding,grant1977null}.
+\ProjectName's main design invariant---no relational operator
 sees missing values for attributes it must operate on nor do users see
@@
 detect dirty values, whether missing or otherwise, and rectify the errors if
-possible. A survey of methods and systems is provided in~\cite{hellerstein2008quantitative}.
+possible. \textcite*{hellerstein2008quantitative} surveys the methods and systems related to missing value detection and correction.
 
-In~\cite{wolf2007query}, queries over databases with missing values are
-processed directly using a statistical approach, taking advantage of correlations between
+\textcite*{wolf2007query} treat queries over databases with missing values directly using a statistical approach, taking advantage of correlations between
 attributes. The tuples that match the original query as well as a ranking of
@@
 and consider them carefully in query processing. Often, if sensor error is the
-cause of missing values, values can be imputed with high confidence. In~\cite{fernandez2009inter}, feedback punctuation is used to dynamically
+cause of missing values, values can be imputed with high confidence. \todo{What does this mean?} \textcite*{fernandez2009inter} use feedback punctuation to dynamically
 reconfigure the query plan for state-dependent
@@
 \review{
-\cite{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
+\textcite*{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
 of probabilistic compositional repair operators, called Lenses. These can be used
@@
 repairs take place on a per-tuple basis, with the non-deterministic layer of their system replacing place-holder
-variables with inferred values. Meanwhile, \ProjectName{} optimizes imputation at the plan-level and
+variables with inferred values. In contrast, \ProjectName{} optimizes imputation at the plan-level and
 provides the same impute technique for all tuples streaming through a given operator.
@@
 \review{
-The planning algorithm that we present has similarities to the multi-objective query optimizer in~\cite{trummer2014}.
+The planning algorithm that we present has similarities to work on multi-objective query optimization~\cite{trummer2014}.
 Both algorithms handle plans which have multiple cost functions by maintaining sets of Pareto optimal plans and pruning plans which become dominated.
 In \ProjectName{}, costs are expected to grow monotonically as the size of the plan increases, which simplifies the optimization problem significantly.
-\cite{trummer2014} handles the more complex case of cost functions which are piecewise linear, but not necessarily monotonic.
+\textcite*{trummer2014} handle the more complex case of cost functions which are piecewise linear, but not necessarily monotonic.
 }
@@
 \subsection{Forecasting and Databases}
-\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
+\textcite*{parisi2011embedding} introduces the idea of incorporating time-series forecast operators into
 databases, along with the necessary relational algebra extensions. Their work explores the theoretical
@@
 the type of predictions returned. They highlight the use of forecasting for replacing missing values.
-In their follow on work~\cite{parisi2013temporal}, they identify various equivalence and containment
+In subsequent work~\cite{parisi2013temporal}, they identify various equivalence and containment
 relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
@@
 
-\cite{fischer2013towards} describe the architecture of a DBMS with integrated forecasting operations for time-series, detailing
+\textcite*{fischer2013towards} describe the architecture of a DBMS with integrated forecasting operations for time-series, detailing
 the abstractions necessary to do so.
 
-In contrast to this work, \ProjectName{} is targeted at generic value imputation, not necessarily tailored to 
-time-series. The optimizer is not based on equivalence transformations, nor are there guarantees of equal
+In contrast to this work, \ProjectName{} is targeted at generic value imputation, and is not tailored to time-series.
+The optimizer is not based on equivalence transformations, nor are there guarantees of equal
 results under different conditions. Instead, the optimizer allows users to pick their trade-off between
@@
 
-\cite{duan2007processing} describes the Fa system and its declarative language for time-series forecasting. Their
+\textcite*{duan2007processing} describe the Fa system and its declarative language for time-series forecasting. Their
 system automatically searches the space of attribute combinations/transformations and statistical models

### 2017-05-15 rebuttal.tex
rebut: respond to point about # of queries

@@
 \usepackage{xcolor}
+\usepackage{enumitem}
 \newcommand{\resp}[1]{{\color{blue}{#1}}}
@@
 \section{Overall changes}
-\begin{itemize}
+\begin{itemize}[leftmargin=*]
 \item Redundant figures, excessive whitespace, formatting.
@@
   \section{Reviewer 1}
-  \begin{itemize}
+  \begin{itemize}[leftmargin=*]
   \item W2.
@@
   \item W3.
-Reducing runtime to ~10s is excellent, but comes at a very serious accuracy cost (3-5x worse SMAPE).
+Reducing runtime to 10s is excellent, but comes at a very serious accuracy cost (3-5x worse SMAPE).
 Worse still, there's no way for the user to differentiate between the cases of Q4 and Q8 at runtime.
@@
 
-    Avoiding this case requires us to estimate the skew in these intermediate tables, which would be in itself an interesting and challenging research problem. If we had improved intermediate
-    estimates, we could incorporate those to improve our system.
+      Avoiding this case requires us to estimate the skew in these intermediate tables, which is an area of active research.
+      However, if we had improved intermediate estimates, we could incorporate them to improve our system.
       }
@@
   \section{Reviewer 2}
-  \begin{itemize}
+  \begin{itemize}[leftmargin=*]
   \item 1) No new imputation method is presented. The model relies on external imputation techniques, and given it's done on-the-fly, presumably imputed values depend on qualifying query records.
@@
   \section{Reviewer 3}
-  \begin{itemize}
+  \begin{itemize}[leftmargin=*]
   \item W2. More work is required in the presentation: important parts are not explained in depth, while there are many details in other sections that can be omitted.
@@
     \resp{
+      It is straightforward to compute the number of queries before we lose our advantage.
+      For example, performing full imputation on the ACS dataset takes 355 minutes (Sec.
+5.3.2) and we can run a simple query in 4 seconds, so we can run approximately $5,000$ queries before taking more time than full imputation.
+
       The sensitivity of different values of $\alpha$ depends on the query itself. Performing a scan of values of $\alpha$ with small step sizes can return
@@
       two queries, one from each dataset used during evaluation. This frontier can be exposed to the user to help in their workflow.
-      \\
+
       We explored the planning times for queries with increasing number of joins by running additional experiments with up to 8 joins. 

### 2017-05-15 paper/algorithm.tex
paper: grammar & spelling

@@
 In order to pick a final plan from the frontier, our model introduces a trade-off parameter $\alpha$, which is an upper bound on the $\textsc{Penalty}$ value that the user is willing
-to tolerate relative to the minimum possible amongst frontier plans.
+to tolerate relative to the minimum possible among the frontier plans.
 
@@
 The form of $T$ depends on the specific imputation method, but there are several considerations in choosing $T$.
-In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases and it should be take into account the complexity of the imputation algorithm.
+In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases and it should take into account the complexity of the imputation algorithm.
 Analysis of the time complexity of different imputation algorithms, especially with in-database execution, is beyond the scope of this paper.
@@
 allows us to obtain the final set of plans that best trade-off computation
-cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.
+cost and imputation quality. The plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.
 
@@
 collect sound and complete Pareto frontiers. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
-Pareto frontier produced by the optimizer corresponds to the collection of plans that best tradeoff imputation quality and
-execution time.
+Pareto frontier produced by the optimizer is the set of plans with the best imputation quality-execution time trade-offs.
 

### 2017-05-15 paper/examples.tex
paper: grammar & spelling

--- 2017-05-05 paper/examples.tex
@@
 
+\review{
+As an alternative strategy, the analyst can start by executing their query with $\alpha=1$ and incrementally reducing $\alpha$, as long as execution of the query completes within some user-defined time bound.
+\ProjectName{}'s planning algorithm can expose the final set of plans with optimal time-quality trade-offs, which allows the user to adjust their choice of $\alpha$ to select from these plans.
+}
+
 Tens of queries later, our epidemiologist has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.

### 2017-05-15 paper/experiments.tex
old ref
paper: grammar & spelling

@@
 to the baseline is low---\todo{update: between \lowsmapealphazero{} and
-\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with largest error correspond
+\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with the largest error correspond
 most closely to one existing approach of dropping all \nullv{}s.
@@
 These are the resulting set of plans after pruning dominated plans throughout our algorithm.
-and encapsulate the possible tradeoffs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
+and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
 can be exposed to the user to help in guiding their workflow.
@@
     \ProjectName{}'s final planner Pareto frontier for \Cref{q1} and \Cref{q6}. 
-    For clarity, we have removed from the plot plans that provide nearly indistinguishable tradeoffs in either dimension.
+    For clarity, we have removed from the plot plans that provide nearly indistinguishable trade-offs in either dimension.
 }}
@@
 We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
-Planning times (~\Cref{fig:planning-times}) for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
+Planning times for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
 We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets

### 2017-05-15 paper/relatedwork.tex
paper: grammar & spelling

@@
 to repair missing values by imposing data constraints such as \texttt{NOT NULL}. 
-Similarly to \ProjectName{}, they provide a way to balance between the quality of the cleaning 
+As with \ProjectName{}, they provide a way to balance between the quality of the cleaning 
 performed (expressed by the confidence of their output) and the cost of those operations. However, their 
@@
 \subsection{Forecasting and Databases}
-\textcite*{parisi2011embedding} introduces the idea of incorporating time-series forecast operators into
+\textcite*{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
 databases, along with the necessary relational algebra extensions. Their work explores the theoretical
@@
 
-Similarly to Fa, \ProjectName{} provides a declarative language, as
+As with Fa, \ProjectName{} provides a declarative language, as
 a subset of standard SQL.\@ \ProjectName{}, however, is not searching the space of possible

### 2017-05-15 paper/algorithm.tex
Add qualitative results from other imputation methods

--- 2017-05-15 paper/algorithm.tex
@@
 
+We note here that for every imputation operator inserted into the query plan, a new
+statistical model is instantiated and trained. Though it is tempting to try to pre-train
+imputation models while the analyst begins to make queries, the exact rows of data, set
+of complete attributes, and set of attributes to impute are unknown until runtime, making
+this strategy infeasible. Thus, if applicable, $T$ should take into account both
+imputation model training and application.
+
 We discuss several values of $T$ for different imputation methods in \Cref{sec:imputation}.
@@
 
-\subsubsection{Hot-deck}
+\subsubsection{Hot deck}
 
-Hot-deck imputation is a nonparametric method in which missing values are replaced with randomly selected complete values.
-This method preserves the distribution of the complete values.
-We use the same $P$ as for mean value imputation, but we modify $T$ to account for the fact that hot-deck is blocking in our implementation.
+Random hot deck imputation is a nonparametric method in which missing values in a column are
+replaced with randomly selected complete values from the same column, preserving the
+distribution of the complete values. We use the same $P$ as for mean value imputation, but
+we modify $T$ to account for the fact that hot-deck is blocking in our implementation.
 

### 2017-05-15 paper/experiments.tex
Add qualitative results from other imputation methods

--- 2017-05-15 paper/experiments.tex
@@
 
-For each query, we evaluate \todo{update: three different configurations}: \ProjectName{} optimizing for quality ($\alpha=0$) and \ProjectName{} optimizing for performance ($\alpha=1$) and as a baseline, we fully impute the tables used by each query then run the query on the imputed tables, simulating the process that an analyst would follow to apply existing imputation techniques.
+We evaluate \ProjectName{} optimizing for quality ($\alpha=0$), optimizing for performance
+($\alpha=1$), and targeting a balance of each. As a baseline, we fully impute the tables
+used by each query and then run the query on the imputed tables, simulating the process that
+an analyst would follow to apply existing imputation techniques.
 
 \Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
-queries are an \todo{update/validate: order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance}. This performance
+queries are an order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance. This performance
 differential means it is feasible
@@
 \caption{Runtimes for the queries in~\Cref{tbl:queries} using reference
-    chained-equation decision trees algorithm. $\alpha=0$ (resp. $\alpha=1$) is
-    \ProjectName{} run at the highest (resp. lowest) quality setting. Baseline
-    is full table imputation followed by running the query. Each query and each
-    value of $\alpha$ was run 200 times.}
+    chained-equation decision trees algorithm. Using $\alpha=0$, \ProjectName{}
+    is run at the highest quality setting, while with $\alpha=1$, performance
+    is prioritized. Each query and each value of $\alpha$ was run 200 times.}
 \label{fig:runtimes}
@@
 
-\todo{Distinguish between underlying imputation methods}
-For $\alpha=0$ and $\alpha=1$, we pair up tuples and compute SMAPE for all queries. That is,
-we compute tuple-wise absolute percentage deviations within each iteration of a query, and
-we report this value averaged over all iterations.  We
+To calculate SMAPE for $\alpha=0$, $\alpha=0.5$, and $\alpha=1$, 
+we compute tuple-wise absolute percentage deviations within each iteration of each query, and
+average this value over all iterations. We
 can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
-to the baseline is low---\todo{update: between \lowsmapealphazero{} and
-\highsmapealphaoneexacs{} percent}---indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with largest error correspond
+to the baseline is low --- between \lowsmapealphazero{} and
+\highsmapealphaoneexacs{} percent --- indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with largest error correspond
 most closely to one existing approach of dropping all \nullv{}s.
+
+In \Cref{table:smape}, we report results using the reference chained-equation
+decision trees model. We also experiment with mean-value imputation and hot
+deck imputation, as outlined in \Cref{sec:imputation}. One approach to
+mean-value imputation, to facilitate non-blocking operation within the iterator
+model, is to estimate the column mean from base table histograms. In this case,
+imputations are identical no matter where the operator is placed in the query
+plan. Moreover, runtimes are essentially identical regardless of the value of
+$\alpha$ because the time cost of the imputation is relatively low no matter
+how many tuples need to be imputed. In hot deck imputation, runtimes are also
+similar across $\alpha$, because the cost of buffering
+tuples and sampling at random is negligible compared to the rest of query
+execution. In terms of accuracy, relative errors using hot deck are close to
+zero; since the baseline, in this case, is hot deck imputation on the base
+table, only skew in the distributions arising from filters and joins would lead
+to larger errors. In some cases, skew does arise, and errors are on the same
+magnitude as in the decision trees case. Indeed, \ProjectName{} yields larger
+benefit when using relatively sophisticated, higher quality imputation
+algorithms as opposed to simpler strategies that would even be quick to
+implement on the base table.
 
@@
 
-\todo{Should this be here?}
-For every imputation operator inserted into the query plan, a new statistical model is instantiated and trained before being used for imputation.
-Although it is tempting to further optimize query execution in \ProjectName{} by pre-training imputation models while the analyst begins to make queries,
-this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are unknown until runtime.
-
 %\subsubsection{Planning Times}
@@
 
-\todo{Update: Planning times for our benchmark queries are reasonable, with mean planning
+Planning times for our benchmark queries are reasonable, with mean planning
 times across the benchmark queries ranging from \planninglowtime{} ms to
-\planninghightime{} ms, constituting \planninglowtimepct{} to \planninghightimepct{} of
-total runtime, respectively.
-
-In all cases, the optimizer returned a query
-plan within \planningmaxtime{} ms.} The choice of $\alpha$ has no impact on planning times,
-because query selection is done after collecting the Pareto frontier.
+\planninghightime{} ms, and ranging from \planninglowtimepct{} to
+\planninghightimepct{} as a share of total runtime.  In all cases, the
+optimizer returned a query plan within \planningmaxtime{} ms. The choice of
+$\alpha$ has no impact on planning times, because query selection is done after
+collecting the Pareto frontier.
 

### 2017-05-15 paper/relatedwork.tex
Add qualitative results from other imputation methods

--- 2017-05-15 paper/relatedwork.tex
@@
 Designers of data stream processing systems frequently confront missing values
-and consider them carefully in query processing. Often, if sensor error is the
-cause of missing values, values can be imputed with high confidence. \todo{What does this mean?} \textcite*{fernandez2009inter} use feedback punctuation to dynamically
+and consider them carefully in query processing. Often, if a physical process
+like sensor error is the cause of missing values, values can be imputed with
+high confidence. \textcite*{fernandez2009inter} use feedback punctuation to
+dynamically
 reconfigure the query plan for state-dependent

### 2017-05-15 rebuttal.tex
changes based on latest sam comments

@@
   }
+
+\item \resp{We extended the original paper two additional popular imputation techniques: mean value imputation and hot-deck imputation.}
+
+\item \resp{We extended experiments to include the new operators and provide more details on results for $\alpha$. We also added the Pareto frontiers for various plans and
+explored planning times for more complex queries.}
+
+\item \resp{We added a section discussing how to construct the imputation-related functions used by our cost model, which represent the API for a new imputation operator.}
+We provide details on the definition of these functions for all imputation methods used.
+
+\item \resp{We included a new section that provides a roadmap for implementing \ProjectName{} on top of an existing database.}
 \end{itemize}
@@
     \resp{
-      The outsized error in Q8 is due to a drop operator placed very early on in the plan (a result of optimizing for performance). This produced significantly smaller sample sizes for various
+      The outsized error in Q8 is due to a drop operator placed very early on in the plan-- this is a result of optimizing for performance
+      and the fact that this data set has many missing values for relevant attributes (77.50\% for \texttt{studentdebtowe}). This produced significantly smaller sample sizes for various
       groups, which skewed the average calculation for each group significantly. For example, the average student debt for subjects with no high school degree was almost 83\% off from the estimate
@@
       We can trivially extend \ProjectName{} to warn users when the query chosen for execution has a high $\textsc{Penalty}$ estimate, along with the number of tuples that have
-      been dropped (after execution), so that situations such as those observed in Q8 can be identified by the user.
+      been dropped (after execution), so that situations such as those observed in Q8 can be identified by the user. We've added this detail to the paper.
     }
@@
 
-      Avoiding this case requires us to estimate the skew in these intermediate tables, which is an area of active research.
+      Avoiding this case requires us to estimate the skew in these intermediate tables, which is an area that traditional query optimizers struggle with as well.
       However, if we had improved intermediate estimates, we could incorporate them to improve our system.
@@
       to work well in our system, with errors proportional to the fraction of relevant tuples retrieved
-      and the AVG value estimated.}
+      and the AVG value estimated. We have noted this in the experiments section now.}
 
@@
       and performance of using ImputeDB versus a production imputation system on the raw
-      data. Unfortunately, they would never have access to a real ground truth in such a
+      data. Unfortunately, they would never have access to real ground truth in such a
       scenario, and moreover, finding ground truth for use even in an empirical evaluation
@@
       We are not aware of another database technology which offers on-the-fly imputation optimization, so there is no direct comparison for the planning algorithm.
+      Again, the point of this paper is not to evaluate different imputation techniques. However, we have added a new section describing how new imputation techniques can be integrated
+      into the paper, and also evaluated the effectiveness of these new algorithms in our experiments.
 %      However, we ran a simple experiment using BayesDB to compare possible imputation times. In order to make the comparison as fair as possible in terms of running
@@
 
-    \resp{Yes, we have now highlighted this for clarity in the paper as well.}
+    \resp{Yes, we have now highlighted this for clarity in the paper as well.\todo{add specific section}}
 
@@
 
-    \resp{We maintain the set of Pareto optimal plans at each point in the optimization.
-This set is significantly smaller than the set of all plans, so it represents some pruning.
-We found that the size of these Pareto sets is small enough that we did not need to prune further.
+    \resp{We maintain the set of Pareto optimal plans at each point in the optimization. This set is significantly smaller than the set of all plans, so it represents some pruning (a point we have
+    now clarified in the paper). We found that the size of these Pareto sets is small enough that we did not need to prune further.
 
@@
   we can return immediately, otherwise we return after this upper bound is reached. This is a common approach used in iterative optimization techniques, such
-  as gradient descent.
+  as gradient descent. We've clarified this technique in the section discussing this imputation operation.
 }
@@
 5.3.2) and we can run a simple query in 4 seconds, so we can run approximately $5,000$ queries before taking more time than full imputation.
+We've made this point explicit in the ACS experiment section now.
 
-      The sensitivity of different values of $\alpha$ depends on the query itself. Performing a scan of values of $\alpha$ with small step sizes can return
-      a lot of the same plans for execution. In lieu of this, we extended the results section to include a subsection that presents the final frontiers for
+      The sensitivity to different values of $\alpha$ depends on the query itself. Performing a scan of values of $\alpha$ with small step sizes can return
+      a lot of the same plans for execution. We did run experiments with a range of $\alpha$ values, but instead of presenting a figure with each value of $\alpha$,
+      which can result in many duplicate plans (since different $\alpha$ values may result in the same plan), we extended the results section to include a subsection that presents the final frontiers for
       two queries, one from each dataset used during evaluation. This frontier can be exposed to the user to help in their workflow.

### 2017-05-15 paper/experiments.tex

@@
 
-In \Cref{table:smape}, we report results using the reference chained-equation
+\review{In \Cref{table:smape}, we report results using the reference chained-equation
 decision trees model. We also experiment with mean-value imputation and hot
@@
 tuples and sampling at random is negligible compared to the rest of query
-execution. In terms of accuracy, relative errors using hot deck are close to
+execution. \todo{be quantitative: In terms of accuracy, relative errors using hot deck are close to
 zero; since the baseline, in this case, is hot deck imputation on the base
@@
 algorithms as opposed to simpler strategies that would even be quick to
-implement on the base table.
+implement on the base table.}
 
@@
 This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.
-By using \ProjectName{} this cost can be lowered significantly.
+By using \ProjectName{} this cost can be lowered significantly. We can also trivially extend \ProjectName{} to warn users when the query chosen for execution has a high
+$\textsc{Penalty}$ estimate, along with the number of tuples that have been dropped (after execution), so that situations with high potential for skewed results can be identified by the user.
 
-\review{
+Our benchmark queries performed well for both cases of \texttt{AVG} and \texttt{COUNT} aggregates, so we expect \ProjectName{} would provide similar performance for \texttt{SUM}, with 
+errors proportional to the fraction of relevant tuples retrieved and the \texttt{AVG} value estimated.
+
 \Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}.
@@
 In contrast, \ProjectName{} executes the query \lstinline{SELECT AVG(c0) FROM acs} over the ACS dataset (also using chained-equation decision trees imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
+Given the runtime difference, we can run approximately 5,000 queries before taking more time than full imputation.
 The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.

### 2017-05-15 paper/conclusion.tex
Update discussion of regression tree error results

@@
 then formulating queries. Furthermore, the difference in query results between the two approaches is
-shown to be small in all queries considered (0-\highsmapealphaoneexacs\% error, \Cref{sec:results}). Data analysts need not commit to a specific imputation strategy and can instead
+shown to be small in all queries considered (\lowsmapealphazero{}-\highsmapealphaone{}\% error, \Cref{sec:results}). Data analysts need not commit to a specific imputation strategy and can instead
 vary this across queries.

### 2017-05-15 paper/experiments.tex
Update discussion of regression tree error results

@@
 
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline.
+\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline, using the reference chained-equation decision trees model.
 This measures the error introduced by on-the-fly imputation as compared to full imputation on the base tables.
@@
 
-To calculate SMAPE for $\alpha=0$, $\alpha=0.5$, and $\alpha=1$, 
+To calculate SMAPE for each $\alpha$,
 we compute tuple-wise absolute percentage deviations within each iteration of each query, and
-average this value over all iterations. We
-can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
-to the baseline is low --- between \lowsmapealphazero{} and
-\highsmapealphaoneexacs{} percent --- indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with largest error correspond
-most closely to one existing approach of dropping all \nullv{}s.
+average this value over all iterations.
+Relative errors of query results, when optimizing for quality
+($\alpha=0$), are low --- between \lowsmapealphazero{} and
+\highsmapealphazero{} --- indicating that on-the-fly imputation produces
+results similar to the baseline. When optimizing for performance ($\alpha=1$),
+relative errors can be modestly higher --- up to \highsmapealphaone{} in the
+worst case, as this approach most closely corresponds to dropping all
+\nullv{}s. Thus, it is important to recognize the tradeoffs between quality and
+performance for the specific use case.
 
-\review{In \Cref{table:smape}, we report results using the reference chained-equation
-decision trees model. We also experiment with mean-value imputation and hot
+\review{
+We also experiment with mean-value imputation and hot
 deck imputation, as outlined in \Cref{sec:imputation}. One approach to
@@
 algorithms as opposed to simpler strategies that would even be quick to
-implement on the base table.}
+implement on the base table.
+}
+}
 
@@
 
+\review{
 \Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}.
@@
     For clarity, we have removed from the plot plans that provide nearly indistinguishable trade-offs in either dimension.
-}}
+}
 \label{fig:pareto-frontiers}
+}
 \end{figure}

### 2017-05-15 paper/introduction.tex
Update discussion of regression tree error results

@@
 standard SQL queries over a data set, even if that data has missing values.
-These queries execute between \lowxalphazero{} and \highxalphazeroexacs{} times as fast as it takes to impute all missing values and then run queries (the traditional approach).
+These queries execute between \lowxalphazero{} and \highxalphazero{} times as fast as it takes to impute all missing values and then run queries (the traditional approach).
 Furthermore, the results obtained with this approach
 are similar in quality to those obtained with the traditional imputation approach
---- between \lowsmapealphazero{} and \highsmapealphaoneexacs{}
+--- between \lowsmapealphazero{} and \highsmapealphaone{}
 percent, in one error measure --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
@@
 
-% Scope
+\subsection{Contributions}
 \ProjectName{} is designed to enable early data exploration, by allowing analysts to run
-their queries without an explicit base-table imputation step.
-% SAM -- cut the following -- could consider moving later but feels out of place here
-%However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.
+their queries without an explicit base-table imputation step. To do so, it leverages a number of contributions
+to minimize imputation time while producing comparable results to traditional approaches.
 
-\subsection{Contributions}
-\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches.
 These contributions include:

### 2017-05-15 paper/main.tex
Add qualitative results from other imputation methods
Update discussion of regression tree error results

@@
 \newcommand{\lowxalphazero}{10} % the smallest speed up, comparing a=0 to base table imputation
-\newcommand{\highxalphazeroexacs}{140} % the largest speed up, comparing a=0 to base table imputation 
-\newcommand{\highxalphaoneexacs}{1400} % the largest speed up, comparing a=0 OR a=1 to base table imputation 
+\newcommand{\highxalphazero}{140} % the largest speed up, comparing a=0 to base table imputation 
+\newcommand{\highxalphaone}{1400} % the largest speed up, comparing a=0 OR a=1 to base table imputation 
 \newcommand{\lowsmapealphazero}{0} % the smallest SMAPE difference, comparing a=0 to base table imputation
 \newcommand{\highsmapealphazero}{6} % the largest SMAPE difference, comparing a=0 to base table imputation
-\newcommand{\highsmapealphaoneexacs}{20} % the largest SMAPE difference, comparing a=1 to base table imputation
+\newcommand{\highsmapealphaone}{20} % the largest SMAPE difference, comparing a=1 to base table imputation
 %\newcommand{\highxalphaone}{16000}
@@
 
-\newcommand{\planningmaxtime}{34} % TODO
+\newcommand{\planningmaxtime}{\todo{XXX}} % {34} % TODO
 
@@
   % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
-  Our experiments show that our query plans execute between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
+  Our experiments show that our query plans execute between \lowxalphazero{} and \highxalphazero{} times faster than first imputing the base tables.
   Furthermore, we show that the query results from on-the-fly imputation differ from the
-  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
+  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaone{}\%, in one error measure.
   Finally, we show that while dropping tuples with missing values that fail query constraints discards 6--78\% of the data, on-the-fly imputation loses only 0--21\%.

### 2017-05-15 paper/experiments.tex
Add max planning time result

@@
 ($\alpha=0$), are low --- between \lowsmapealphazero{} and
-\highsmapealphazero{} --- indicating that on-the-fly imputation produces
+\highsmapealphazero{}\% --- indicating that on-the-fly imputation produces
 results similar to the baseline. When optimizing for performance ($\alpha=1$),
-relative errors can be modestly higher --- up to \highsmapealphaone{} in the
+relative errors can be modestly higher --- up to \highsmapealphaone{}\% in the
 worst case, as this approach most closely corresponds to dropping all
 \nullv{}s. Thus, it is important to recognize the tradeoffs between quality and
-performance for the specific use case.
+performance for a specific use case.
 
@@
 
-Planning times for our benchmark queries are reasonable, with mean planning
-times across the benchmark queries ranging from \planninglowtime{} ms to
-\planninghightime{} ms, and ranging from \planninglowtimepct{} to
-\planninghightimepct{} as a share of total runtime.  In all cases, the
-optimizer returned a query plan within \planningmaxtime{} ms. The choice of
-$\alpha$ has no impact on planning times, because query selection is done after
-collecting the Pareto frontier.
+Planning times for our benchmark queries are reasonable, with mean planning times across the
+benchmark queries ranging from \planningtimelow{} to \planningtimehigh{}, and ranging from
+\planningtimelowpct{} to \planningtimehighpct{}\% of total runtime. In 99\% of individual
+cases, the optimizer returned a query plan with \planningtimepninetynine{}. The choice of
+$\alpha$ has no impact on planning times, because query selection is done after collecting
+the Pareto frontier.
 

### 2017-05-15 paper/experiments.tex
some moves and highlight new text

@@
 provide performance figures relevant to practitioners using similar datasets.
+\review{
+Our benchmark queries performed well for both cases of \texttt{AVG} and \texttt{COUNT} aggregates, 
+so we expect \ProjectName{} would provide similar performance for \texttt{SUM}, with 
+errors proportional to the fraction of relevant tuples retrieved and the \texttt{AVG} value estimated.}
 
@@
 This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.
+\review{
 By using \ProjectName{} this cost can be lowered significantly. We can also trivially extend \ProjectName{} to warn users when the query chosen for execution has a high
@@
 
-Our benchmark queries performed well for both cases of \texttt{AVG} and \texttt{COUNT} aggregates, so we expect \ProjectName{} would provide similar performance for \texttt{SUM}, with 
-errors proportional to the fraction of relevant tuples retrieved and the \texttt{AVG} value estimated.
-
-\review{
 \Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}.
@@
 In contrast, \ProjectName{} executes the query \lstinline{SELECT AVG(c0) FROM acs} over the ACS dataset (also using chained-equation decision trees imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
-Given the runtime difference, we can run approximately 5,000 queries before taking more time than full imputation.
+\review{Given the runtime difference, we can run approximately 5,000 queries before taking more time than full imputation.
 The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.
@@
 An analyst could do the same, but tracking the required imputations would get increasingly complicated as the queries became more complex.
-This highlights the benefit of using our system for early data exploration.
+This highlights the benefit of using our system for early data exploration.}
 

### 2017-05-15 paper/experiments.tex
paper: minor changes

--- 2017-05-15 paper/experiments.tex
@@
 We evaluate \ProjectName{} optimizing for quality ($\alpha=0$), optimizing for performance
-($\alpha=1$), and targeting a balance of each. As a baseline, we fully impute the tables
+($\alpha=1$), and targeting a balance of each ($\alpha = 0.5$). As a baseline, we fully impute the tables
 used by each query and then run the query on the imputed tables, simulating the process that
@@
 
-\Cref{fig:runtimes} shows a summary of the performance results. The quality-optimized
+\Cref{fig:runtimes} summarizes the performance results. The quality-optimized
 queries are an order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance. This performance
@@
 We also experiment with mean-value imputation and hot
-deck imputation, as outlined in \Cref{sec:imputation}. One approach to
+deck imputation (\Cref{sec:imputation}). One approach to
 mean-value imputation, to facilitate non-blocking operation within the iterator
@@
 \planningtimelowpct{} to \planningtimehighpct{}\% of total runtime. In 99\% of individual
-cases, the optimizer returned a query plan with \planningtimepninetynine{}. The choice of
+cases, the optimizer returned a query plan within \planningtimepninetynine{}. The choice of
 $\alpha$ has no impact on planning times, because query selection is done after collecting

### 2017-05-15 paper/experiments.tex
paper: squeeze

@@
 
-There are 6 tables in the NHANES data set. We use three tables for our experiments:
-
-\begin{itemize}
-	\item \texttt{demo}: demographic information of subjects
-	\item \texttt{exams}: physical exam results
-	\item \texttt{labs}: laboratory exam results
-\end{itemize}
+There are 6 tables in the NHANES data set.
+We use three tables for our experiments: demographic information of subjects (\texttt{demo}), physical exam results (\texttt{exams}), and laboratory exam results (\texttt{labs}).
 
@@
 \Cref{fig:runtimes} summarizes the performance results. The quality-optimized
-queries are an order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance. This performance
-differential means it is feasible
-to explore multiple imputation operations (including more expensive operators) when using
-\ProjectName{}, in contrast to the traditional approach of fully imputing the dataset.
+queries are an order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance.
 

### 2017-05-15 paper/main.tex
Add max planning time result
paper: minor changes
paper: squeeze

@@
 \usepackage[noend]{algpseudocode}
-\usepackage[style=numeric,firstinits=true,maxcitenames=2]{biblatex}
+\usepackage[style=savetrees,firstinits=true,maxcitenames=2]{biblatex}
 \usepackage{booktabs}
@@
 
-\newcommand{\planninglowtime}{0.7}
-\newcommand{\planninglowtimepct}{0}
-\newcommand{\planninghightime}{2.8}
-\newcommand{\planninghightimepct}{14}
+\newcommand{\planningtimelow}{0.7 ms}
+\newcommand{\planningtimelowpct}{0}
+\newcommand{\planningtimehigh}{2.8 ms}
+\newcommand{\planningtimehighpct}{14}
 
-\newcommand{\planningmaxtime}{\todo{XXX}} % {34} % TODO
+\newcommand{\planningtimemax}{113 ms} % max planning time of any single query
+\newcommand{\planningtimepninetynine}{4 ms} % 99 percentile planning time of any single query
+\newcommand{\planningtimepninetyninepointnine}{14 ms} % 99.9 percentile planning time of any single query
 
@@
 \bibstyle{abbrv}
+\printbibliography
 \balance
-\printbibliography
 

### 2017-05-15 paper/algorithm.tex
small changes

@@
 This set $Q$ will be the Pareto frontier~\cite{pareto1964cours} of $S$, and it contains the best option for all feasible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.
+\review{This
+set of plans is significantly smaller than the set of all plans, which our algorithm prunes away.}
 
@@
 Every instantiation of the impute-operator defines a custom penalty function $P$ for imputing \nullv{} fields.
-\ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\text{Penalty}_{P}(q)$, computed as follows.
+\ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\textsc{Penalty}_{P}(q)$, computed as follows.
 
@@
 As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on the properties of the underlying algorithm via a function $T$.
-If we assume that the \textit{Impute} routine is CPU-bound, then we adjust $T$ by some constant factor $f$, thereby scaling the result to be comparable with IO-bound routines.
+If we assume that the \textit{Impute} routine is CPU-bound, then we adjust $T$ by some constant factor $f$, thereby scaling the result to be comparable with I/O-bound routines.
 
@@
 Separately, we specify a heuristic penalty that takes into account the
-improvement in imputation quality from more tuples and attributes as well as
+improvement in imputation quality from more tuples and attributes, as well as
 the diminishing returns from more data.
@@
 At the other end of the complexity spectrum, mean value imputation is a common choice for exploratory data analysis.
-Mean value imputation replaces missing values in a column with the mean of the available values in the column.
+Mean value imputation replaces missing values in a column with the mean of the available values in that column.
 

### 2017-05-15 paper/implementation.tex
paper: mark sections for sam to review
added comment about simpledb in implementation
paper: minor change
paper: add simpledb ref
small changes

@@
+\review{
 \section{Implementation}
 Adding imputation and our optimization algorithm to a standard SQL database requires modifications to several key database components.
-In this section we discuss the changes required to implement dynamic imputation.
+In this section we discuss the changes required to implement dynamic imputation. We based this directly on our experience implementing
+\ProjectName{} on top of SimpleDB~\cite{simpledb}, a teaching database used at MIT, University of Washington, and Northwestern University, among others.
 
@@
   In particular, this tight integration allows us to jointly choose the most effective join order and imputation placement.
-  Our goal of integrating imputation placement into the optimizer requires the other modifications that we have discussed.
 \end{itemize}
+}
 

### 2017-05-15 paper/algorithm.tex
Pass through entire paper

--- 2017-05-15 paper/algorithm.tex
@@
 replace all \nullv{} values with non-\nullv{} values for attributes $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
-In our experiments, we implement \textit{Impute} using a general-purpose chained-equation decision trees algorithm, but our system is agnostic to this choice.
+% In our experiments, we implement \textit{Impute} using a general-purpose chained-equation decision trees algorithm, but our system is agnostic to this choice.
 \textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
@@
 This set $Q$ will be the Pareto frontier~\cite{pareto1964cours} of $S$, and it contains the best option for all feasible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.
+This
+set of plans is significantly smaller than the set of all plans, which our algorithm prunes away.
 
@@
 lose a lot of information through dropping are also the fastest. 
-
 If $\alpha=0.0$, then the optimal query should lose as little information as possible (at
@@
 
-\noindent{\textbf{Constructing $P$}}: The form of $P$ is chosen to reflect the properties of the imputation method being used.
+\textbf{Constructing $P$}:
+The form of $P$ is chosen to reflect the properties of the imputation method being used.
 A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of plan histograms.
@@
 
-\noindent{\textbf{Constructing $T$}}:
+\textbf{Constructing $T$}:
 As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on the properties of the underlying algorithm via a function $T$.
@@
 this strategy infeasible. Thus, if applicable, $T$ should take into account both
-imputation model training and application.
-
-We discuss several values of $T$ for different imputation methods in \Cref{sec:imputation}.
+training and application of the imputation model.
 
@@
 The flexibility of this approach is in the spirit of query optimization, which aims to reconcile performance with a declarative interface to data manipulation.
-To motivate this API, we describe how a developer would specify these functions for a general-purpose imputation algorithm, as well as for two simpler imputation methods.
+To motivate this API, we describe how a developer might specify these functions for a general-purpose imputation algorithm, as well as for two simpler imputation methods.
 
@@
 column in subsequent iterations. For our experiments, we use Weka's \texttt{REPTree}
-implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
+decision tree implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
 algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
@@
 \begin{aligned}
-    T(q) &\approx 0 \\
-    P(q) &= \sum_{A \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(A)}{\textsc{Card}(q)}
+    P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
+    T(q) &\approx 0
 \end{aligned}
@@
 \begin{aligned}
-    T(q) &= \textsc{Card}(q) \\
-    P(q) &= \sum_{A \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(A)}{\textsc{Card}(q)}
+    P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
+    T(q) &= \textsc{Card}(q)
 \end{aligned}
@@
 collect sound and complete Pareto frontiers. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
-Pareto frontier produced by the optimizer is the set of plans with the best imputation quality-execution time trade-offs.
+Pareto frontier produced by the optimizer is the set of plans with the best imputation
+quality and runtime trade-off.
 
@@
 This set will be used to determine which attributes are imputed.
-Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
+Then, it constructs a plan cache. For each table used in the query, any available selections
+are pushed to the leaves, and
 a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}). If no selections are available, a 
@@
 If we allow any arbitrary subset of attributes to be imputed, then we would need to consider $O(2^{|D|+1})$ different imputation operators before each relational operator where $D$ is the set of dirty attributes in all tables.
-This would increase the overall complexity of the algorithm drastically.
+This would increase the overall complexity of the algorithm dramatically.
 

### 2017-05-15 paper/examples.tex
Pass through entire paper

--- 2017-05-08 paper/examples.tex
@@
 \subsection{Planning with \ProjectName{}}
-The search space contains plans of varying performance and quality of results, as a product of the multiple possible locations for imputation.
+The search space contains plans with varying performance and imputation quality
+characteristics, as a product of the multiple possible locations for imputation.
 The user can influence the final plan selected by \ProjectName{}
@@
 
-\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operation $\mu$,
+\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operator $\mu$,
 which employs a reference imputation strategy to fill in missing values rather than dropping tuples.
@@
 
-An alternative strategy involves starting with $\alpha=1$ and iterating to lower values, as long as execution
-of the query completes within some user-defined time bound. \ProjectName{}'s planning
-algorithm can expose the final set of plans that tradeoff execution time and imputation quality, which allows
-the user to adjust their choice of $\alpha$ to uniquely select amongst these plans.
+\review{
+As an alternative strategy, the analyst can start by executing her query with $\alpha=1$ and
+incrementally reducing $\alpha$, as long as execution of the query completes within some
+acceptable time bound.
+\ProjectName{}'s planning algorithm can expose the final set of plans with optimal
+time-quality trade-offs, which allows the user to adjust her choice of $\alpha$ to select from these plans.
+}
 
 Tens of queries later, our epidemiologist has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.
-Knowing that there may be a need to explore this data set further, she can easily incorporate her imputation model into ImputeDB for future use.
+Knowing that there may be a need to explore this data set further, she can also incorporate her imputation model into ImputeDB for future use.
 

### 2017-05-15 paper/experiments.tex
small changes
Pass through entire paper

@@
 \section{Experiments}\label{sec:experiments}
-For our experiments we plan and execute queries for three separate survey-based data sets, showing that our system is well suited for early dataset exploration.
+For our experiments we plan and execute queries for three separate survey-based data sets.
 We show that \ProjectName{} performs an order of magnitude better than traditional
-base-table imputation and produces results of comparable quality.
+base-table imputation and produces results of comparable quality, showing that our system is well suited for early dataset exploration.
 
@@
 conducted by the CDC on a national sample of several thousand individuals~\cite{cdc-data}.
-The data consists on survey responses, physical examinations, and laboratory
-results, amongst others.
+The data consists of survey responses, physical examinations, and laboratory
+results, among others.
 
@@
 
-The original tables have a large number of attributes, in some cases providing more granular tests results or alternative metrics.
+The original tables have a large number of attributes, in some cases providing more granular test results or alternative metrics.
 We focus on a subset of the attributes for each table to simplify the presentation and exploration of queries.
@@
 For our second set of experiments, we use data collected by freeCodeCamp, an open-source
-community for learning to code, as a part of a survey of new software
+community for learning to code, as part of a survey of new software
 developers~\cite{fcc-data}.  The \textit{2016 New Coder Survey} consists of responses by
@@
 Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS)
-data kindly provided by \citeauthor*{akande2015empirical}. Given that the data had
-been cleaned, we artificially dirtied it by replacing 40\% of the values uniformly at random with \nullv{} values.
+data, which we then artificially dirty by replacing 40\% of the values uniformly at random with \nullv{} values.
 The final dataset consists of 671,153 rows and 37 integer columns.
@@
 worst case, as this approach most closely corresponds to dropping all
-\nullv{}s. Thus, it is important to recognize the tradeoffs between quality and
+\nullv{}s. Thus, it is important to recognize the trade-offs between quality and
 performance for a specific use case.
@@
 We also experiment with mean-value imputation and hot
-deck imputation (\Cref{sec:imputation}). One approach to
-mean-value imputation, to facilitate non-blocking operation within the iterator
-model, is to estimate the column mean from base table histograms. In this case,
+deck imputation (\Cref{sec:imputation}). In our implementation of mean-value imputation,
+to facilitate non-blocking operation within the iterator
+model, we estimate the column mean from base table histograms. Therefore,
 imputations are identical no matter where the operator is placed in the query
@@
 The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples used when running the query on the imputed base table.
-This shows that when optimizing for performance, not quality, many tuples are dropped.
-Even in cases where the SMAPE reduction between $\alpha = 1$ and $\alpha = 0$ is small (\Cref{q2}, \Cref{q6}) the tuple count is significantly different.
-In these cases, the aggregate value is not significantly impacted by the missing data. In particular, if values are missing uniformly at random,
+This shows that when optimizing for performance, not quality, many tuples are dropped due to
+insertion of $\lambda$ operators.
+Even in cases where the SMAPE reduction from $\alpha = 1$ and $\alpha = 0$ is small
+(\Cref{q2} and \Cref{q6}) the tuple count is significantly different.
+In these cases, the aggregate value is not significantly impacted by the missing data. In
+particular, if values are missing completely at random,
 the aggregate should not be affected.
@@
 \Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}.
-These are the resulting set of plans after pruning dominated plans throughout our algorithm.
+These are the resulting set of plans after pruning dominated plans throughout our algorithm
 and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
@@
 
+Planning times for our benchmark queries are reasonable, with mean planning times across the
+benchmark queries ranging from \planningtimelow{} to \planningtimehigh{}, and ranging from
+\planningtimelowpct{} to \planningtimehighpct{}\% of total runtime. In 99\% of individual
+cases, the optimizer returned a query plan within \planningtimepninetynine{}. The choice of
+$\alpha$ has no impact on planning times, because query selection is done after collecting
+the Pareto frontier.
+
+\review{
+However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
+We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
+Planning times for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
+We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
+displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
+in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
+}
+
 \subsubsection{Baseline Runtime}
@@
 %\end{figure}
-
-Planning times for our benchmark queries are reasonable, with mean planning times across the
-benchmark queries ranging from \planningtimelow{} to \planningtimehigh{}, and ranging from
-\planningtimelowpct{} to \planningtimehighpct{}\% of total runtime. In 99\% of individual
-cases, the optimizer returned a query plan within \planningtimepninetynine{}. The choice of
-$\alpha$ has no impact on planning times, because query selection is done after collecting
-the Pareto frontier.
-
-\review{
-However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
-We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
-Planning times for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
-We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
-displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
-in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
-}
 %\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

### 2017-05-15 paper/implementation.tex
Pass through entire paper

@@
 \item \textbf{Extended histograms:}
-  Ranking query plans with imputation needs estimates of the distribution of missing values in the base tables and in the outputs of sub-plans.
-  We extended standard histograms to include the count of missing values for each attribute.
+  Ranking query plans that include imputation requires estimates of the distribution of missing values in the base tables and in the outputs of sub-plans.
+  We extend standard histograms to include the count of missing values for each attribute.
   On the base tables, this provides an exact count of the missing values.
@@
   In particular, this tight integration allows us to jointly choose the most effective join order and imputation placement.
+  Our goal of integrating imputation placement into the optimizer requires the other modifications that we have discussed.
 \end{itemize}

### 2017-05-15 paper/introduction.tex
Pass through entire paper

@@
 
-Many databases have large numbers of missing or \nullv{} values;  these can arise for a variety of reasons, including missing source data (e.g., unknown facts),
+Many databases have large numbers of missing or \nullv{} values;  these can arise for a
+variety of reasons, including missing source data,
 missing columns during data integration, de-normalized databases, or as a result of outlier detection and cleaning where \nullv{} values are substituted for
@@
 
-A simpler approach might drop all rows with missing values, which can not only introduce
+A simpler approach might drop all rows with any missing values, which can not only introduce
 bias into the results, but also result in discarding much of the data.
 In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set. Specifically, \ProjectName{}
-carefully plans the placement of imputation/row-drop operations inside the query plan, resulting in a 
+carefully plans the placement of imputation or row-drop operations inside the query plan, resulting in a 
 significant speedup in query execution while reducing the number of dropped rows.
@@
 standard SQL queries over a data set, even if that data has missing values.
-These queries execute between \lowxalphazero{} and \highxalphazero{} times as fast as it takes to impute all missing values and then run queries (the traditional approach).
-Furthermore, the results obtained with this approach
+These queries execute between \lowxalphazero{} and \highxalphazero{} times faster than
+imputing all missing values and then running queries (the traditional approach). Furthermore, the results obtained with this approach
 are similar in quality to those obtained with the traditional imputation approach

### 2017-05-15 paper/relatedwork.tex
Pass through entire paper

@@
 can appear for a variety of reasons~\cite{gelman2006data}.
-It can be missing uniformly at random or conditioned on
+It can be missing completely at random or conditioned on
 existing values (observed and missing). Methods in the statistical community
@@
 treatment of \nullv{}.
-Multiple papers have described various (at times conflicting) treatments of \nullv{}s~\cite{codd1973understanding,grant1977null}.
-\ProjectName's main design invariant---no relational operator
-sees missing values for attributes it must operate on nor do users see
-missing data---eliminates
+Multiple papers have described various (and at times conflicting) treatments of \nullv{}s~\cite{codd1973understanding,grant1977null}.
+\ProjectName's design invariants eliminate
 the need to handle \nullv{} value semantics, while guaranteeing query evaluation soundness.
@@
 
-\textcite*{wolf2007query} treat queries over databases with missing values directly using a statistical approach, taking advantage of correlations between
+\textcite*{wolf2007query} directly treat queries over databases with missing values using a statistical approach, taking advantage of correlations between
 attributes. The tuples that match the original query as well as a ranking of
@@
 can use a simple declarative language (an extension of SQL), to specify models
-which allow missing value imputation, amongst other broader functionality.
-Experts can further customize strategies and express domain knowledge to
+which allow missing value imputation, among other broader functionality.
+Experts can further customize strategies and incorporate domain knowledge to
 improve performance and accuracy.
-
 While BayesDB can be used for value imputation, this step is not framed
@@
 inference step within the query language, using the \verb|INFER| operation. 
-
 BayesDB provides a great alternative for bridging the gap between
@@
 \review{
-\textcite*{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
-of probabilistic compositional repair operators, called Lenses. These can be used
+\textcite*{yang2015lenses} develop an incremental approach to ETL data cleaning, through the use
+of probabilistic compositional repair operators called \textit{Lenses}. These can be used
 to repair missing values by imposing data constraints such as \texttt{NOT NULL}. 
-As with \ProjectName{}, they provide a way to balance between the quality of the cleaning 
+As with \ProjectName{}, they provide a way to balance the quality of the cleaning 
 performed (expressed by the confidence of their output) and the cost of those operations. However, their 
@@
 the abstractions necessary to do so.
-
 In contrast to this work, \ProjectName{} is targeted at generic value imputation, and is not tailored to time-series.
@@
 standard techniques, such as cross-validation. 
-
 As with Fa, \ProjectName{} provides a declarative language, as

### 2017-05-15 paper/relatedwork.tex
reword of yang

--- 2017-05-15 paper/relatedwork.tex
@@
 \textcite*{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
-of probabilistic compositional repair operators, called Lenses. These can be used
-to repair missing values by imposing data constraints such as \texttt{NOT NULL}. 
-As with \ProjectName{}, they provide a way to balance between the quality of the cleaning 
-performed (expressed by the confidence of their output) and the cost of those operations. However, their 
-repairs take place on a per-tuple basis, with the non-deterministic layer of their system replacing place-holder
-variables with inferred values. In contrast, \ProjectName{} optimizes imputation at the plan-level and
-provides the same impute technique for all tuples streaming through a given operator.
+of probabilistic compositional repair operators, called \emph{Lenses}, which balance quality of cleaning
+and cost of those operations. These operators can be used
+to repair missing values by imposing data constraints such as \texttt{NOT NULL}. However, the model
+used to impute missing values must be pre-trained and all repairs are done on a per-tuple basis
+rather than at the plan-level.
 }

### 2017-05-15 paper/algorithm.tex

@@
 This set $Q$ will be the Pareto frontier~\cite{pareto1964cours} of $S$, and it contains the best option for all feasible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.
-This
-set of plans is significantly smaller than the set of all plans, which our algorithm prunes away.
+\review{This
+set of plans is significantly smaller than the set of all plans, which our algorithm prunes away.}
 

### 2017-05-15 paper/experiments.tex

@@
 Community Survey (ACS), a comprehensive survey conducted by the U.S.
-Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS)
-data, which we then artificially dirty by replacing 40\% of the values uniformly at random with \nullv{} values.
+Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS) data,
+which we then artificially dirty by replacing 40\% of the values uniformly at random with
+\nullv{} values.
 The final dataset consists of 671,153 rows and 37 integer columns.

### 2017-05-15 paper/relatedwork.tex

--- 2017-05-15 paper/relatedwork.tex
@@
 \textcite*{yang2015lenses} develop an incremental approach to ETL data cleaning, through the use
-of probabilistic compositional repair operators called \textit{Lenses}. These can be used
-to repair missing values by imposing data constraints such as \texttt{NOT NULL}. 
-As with \ProjectName{}, they provide a way to balance the quality of the cleaning 
-performed (expressed by the confidence of their output) and the cost of those operations. However, their 
-repairs take place on a per-tuple basis, with the non-deterministic layer of their system replacing place-holder
-variables with inferred values. In contrast, \ProjectName{} optimizes imputation at the plan-level and
-provides the same impute technique for all tuples streaming through a given operator.
+of probabilistic compositional repair operators, called \emph{Lenses}, which balance quality of cleaning
+and cost of those operations. These operators can be used
+to repair missing values by imposing data constraints such as \texttt{NOT NULL}. However, the model
+used to impute missing values must be pre-trained and all repairs are done on a per-tuple basis
+rather than at the plan-level.
 }

### 2017-05-15 paper/experiments.tex
cleaned up figure and some review paragraphs

--- 2017-05-15 paper/experiments.tex
@@
 Community Survey (ACS), a comprehensive survey conducted by the U.S.
-Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS)
-data kindly provided by \citeauthor*{akande2015empirical}. Given that the data had
+Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS). Given that the data had
 been cleaned, we artificially dirtied it by replacing 40\% of the values uniformly at random with \nullv{} values.
@@
 
-We evaluate \ProjectName{} optimizing for quality ($\alpha=0$), optimizing for performance
+\review{We evaluate \ProjectName{} optimizing for quality ($\alpha=0$), optimizing for performance
 ($\alpha=1$), and targeting a balance of each ($\alpha = 0.5$). As a baseline, we fully impute the tables
 used by each query and then run the query on the imputed tables, simulating the process that
-an analyst would follow to apply existing imputation techniques.
+an analyst would follow to apply existing imputation techniques.}
 
@@
 model, is to estimate the column mean from base table histograms. In this case,
-imputations are identical no matter where the operator is placed in the query
+imputations are identical regardless of where the operator is placed in the query
 plan. Moreover, runtimes are essentially identical regardless of the value of
-$\alpha$ because the time cost of the imputation is relatively low no matter
+$\alpha$ because the time cost of the imputation is relatively low independent of
 how many tuples need to be imputed. In hot deck imputation, runtimes are also
@@
 and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
-can be exposed to the user to help in guiding their workflow.
+can be exposed to the user to guide their workflow.
 }

### 2017-05-15 paper/pareto_plot.py

@@ -32,11 +32,20 @@ def plot_frontier():
     # damn you matplotlib
     ax.set_xlabel('')
     ax.set_ylabel('')
+    xtick_labels = ax.get_xticks().tolist()
+    xtick_labels[0] = ''
+    ax.set_xticklabels(xtick_labels)
+    
+    ytick_labels = ax.get_yticks().tolist()
+    ytick_labels[0] = ''
+    ytick_labels[-1] = ''
+    ytick_labels[-2] = ''
+    ax.set_yticklabels(ytick_labels)
 
     plt.subplots_adjust(bottom=0.2)
     fig.suptitle('Planner Pareto Frontiers', size=14)
-    fig.text(0.5, 0.04, r'$\textsc{Time}/\textsc{Max}(\textsc{Time})$', ha='center')
-    fig.text(0.04, 0.5, r'$\textsc{Penalty}$', va='center', rotation='vertical')
+    fig.text(0.5, 0.04, r'$\textsc{Time}/\textsc{Max}(\textsc{Time})$', ha='center', size=14)
+    fig.text(0.04, 0.5, r'$\textsc{Penalty}$', va='center', rotation='vertical', size=14)
     fig.savefig('figures/pareto_frontiers_plot.png')
 
 

### 2017-05-15 paper/conclusion.tex

@@
 \section{Conclusion}
-Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational algebra and
+Our evaluation of \ProjectName{} shows that imputation of missing values can be successfully integrated into the relational algebra and
 existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
@@
 While prior work has incorporated operations such as prediction into their databases, this has been domain-specific
-in some cases, and in others, they have not integrated these operations into the planner. The novelty
+in some cases, and in others, operations have not been integrated into the planner. The novelty
 of our contribution lies in the formalization of missing value imputation for query planning, which results in performance

### 2017-05-15 paper/experiments.tex

--- 2017-05-15 paper/experiments.tex
@@
 plan. Moreover, runtimes are essentially identical regardless of the value of
-$\alpha$ because the time cost of the imputation is relatively low no matter
+$\alpha$ because the time cost of the imputation is relatively low independent of
 how many tuples need to be imputed. In hot deck imputation, runtimes are also

### 2017-05-15 paper/experiments.tex
paper: break up results section

--- 2017-05-15 paper/experiments.tex
@@
 
+\review{
 \subsection{Results}\label{sec:results}
@@
 
-\Cref{fig:runtimes} summarizes the performance results. The quality-optimized
+\textbf{Runtime vs. Base-table Imputation}\quad \Cref{fig:runtimes} summarizes the performance results. The quality-optimized
 queries are an order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance.
@@
 %\end{figure}
-\subsubsection{Accuracy vs. Base-table Imputation}
-
-\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline, using the reference chained-equation decision trees model.
+\textbf{Accuracy vs.
+Base-table Imputation (AVG)}\quad \Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline, using the reference chained-equation decision trees model.
 This measures the error introduced by on-the-fly imputation as compared to full imputation on the base tables.
@@
 
-\review{
+\textbf{Accuracy vs. Base-table Imputation (COUNT)}\quad
+We also calculate the number of tuples used to produce each aggregate output.
+The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples used when running the query on the imputed base table.
+This shows that when optimizing for performance, not quality, many tuples are dropped due to
+insertion of $\delta$ operators.
+Even in cases where the SMAPE reduction from $\alpha = 1$ and $\alpha = 0$ is small
+(\Cref{q2} and \Cref{q6}) the tuple count is significantly different.
+In these cases, the aggregate value is not significantly impacted by the missing data. In
+particular, if values are missing completely at random,
+the aggregate should not be affected.
+However, if the missing data is biased then the aggregate will have a significant error.
+This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.
+
+By using \ProjectName{} this cost can be lowered significantly. We can also trivially extend \ProjectName{} to warn users when the query chosen for execution has a high
+$\textsc{Penalty}$ estimate, along with the number of tuples that have been dropped (after execution), so that situations with high potential for skewed results can be identified by the user.
+
+\textbf{Alternate Imputation Methods}\quad
 We also experiment with mean-value imputation and hot
@@
 }
-}
 
-We also calculate the number of tuples used to produce each aggregate output.
-The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples used when running the query on the imputed base table.
-This shows that when optimizing for performance, not quality, many tuples are dropped due to
-insertion of $\lambda$ operators.
-Even in cases where the SMAPE reduction from $\alpha = 1$ and $\alpha = 0$ is small
-(\Cref{q2} and \Cref{q6}) the tuple count is significantly different.
-In these cases, the aggregate value is not significantly impacted by the missing data. In
-particular, if values are missing completely at random,
-the aggregate should not be affected.
-However, if the missing data is biased then the aggregate will have a significant error.
-This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.
-\review{
-By using \ProjectName{} this cost can be lowered significantly. We can also trivially extend \ProjectName{} to warn users when the query chosen for execution has a high
-$\textsc{Penalty}$ estimate, along with the number of tuples that have been dropped (after execution), so that situations with high potential for skewed results can be identified by the user.
-
+\textbf{Pareto Frontiers}\quad
 \Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}.
@@
 and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
-can be exposed to the user to help in guiding their workflow.
-}
+can be exposed to the user to guide their workflow.
 
@@
 
+\textbf{Planning Times}\quad
 Planning times for our benchmark queries are reasonable, with mean planning times across the
@@
 
-\review{
+  \textbf{Approximate Pareto Sets}\quad
 However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
@@
 in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
-}
 
-\subsubsection{Baseline Runtime}
+\textbf{Imputation on Large Data}\quad
 In many real-world cases, applying imputation to an entire dataset is prohibitively expensive.
@@
 In contrast, \ProjectName{} executes the query \lstinline{SELECT AVG(c0) FROM acs} over the ACS dataset (also using chained-equation decision trees imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
-\review{Given the runtime difference, we can run approximately 5,000 queries before taking more time than full imputation.
+Given the runtime difference, we can run approximately 5,000 queries before taking more time than full imputation.
 The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.
@@
 An analyst could do the same, but tracking the required imputations would get increasingly complicated as the queries became more complex.
-This highlights the benefit of using our system for early data exploration.}
+This highlights the benefit of using our system for early data exploration.
+}
 

### 2017-05-15 paper/experiments.tex
Add numbers to discussion of other imputation algs

--- 2017-05-15 paper/experiments.tex
@@
 
-\review{We evaluate \ProjectName{} optimizing for quality ($\alpha=0$), optimizing for performance
+We evaluate \ProjectName{} optimizing for quality ($\alpha=0$), optimizing for performance
 ($\alpha=1$), and targeting a balance of each ($\alpha = 0.5$). As a baseline, we fully impute the tables
 used by each query and then run the query on the imputed tables, simulating the process that
-an analyst would follow to apply existing imputation techniques.}
+an analyst would follow to apply existing imputation techniques.
 
@@
 \review{
-We also experiment with mean-value imputation and hot
-deck imputation (\Cref{sec:imputation}). In our implementation of mean-value imputation,
-to facilitate non-blocking operation within the iterator
-model, we estimate the column mean from base table histograms. Therefore,
-imputations are identical no matter where the operator is placed in the query
-plan. Moreover, runtimes are essentially identical regardless of the value of
-$\alpha$ because the time cost of the imputation is relatively low independent of
-how many tuples need to be imputed. In hot deck imputation, runtimes are also
-similar across $\alpha$, because the cost of buffering
-tuples and sampling at random is negligible compared to the rest of query
-execution. \todo{be quantitative: In terms of accuracy, relative errors using hot deck are close to
-zero; since the baseline, in this case, is hot deck imputation on the base
-table, only skew in the distributions arising from filters and joins would lead
-to larger errors. In some cases, skew does arise, and errors are on the same
-magnitude as in the decision trees case. Indeed, \ProjectName{} yields larger
-benefit when using relatively sophisticated, higher quality imputation
-algorithms as opposed to simpler strategies that would even be quick to
-implement on the base table.
-}
+We also experiment with mean-value imputation and hot deck imputation
+(\Cref{sec:imputation}). In our implementation of mean-value imputation, to facilitate
+non-blocking operation within the iterator model, we estimate the column mean from base
+table histograms. Therefore, imputed values are identical no matter where the operator is
+placed in the query plan (relative errors are $0$). Runtimes are in the range of
+\runtimemeanlow{} to \runtimemeanhigh{} for queries using \ProjectName{}, as compared to
+\runtimemeanlowbase{} to \runtimemeanhighbase{} when imputing on the base table.  Here, the
+time cost of the imputation is relatively low no matter how many tuples need to be imputed.
+
+In hot deck imputation, runtimes are also similar across $\alpha$, ranging from
+\runtimehotdecklow{} to \runtimehotdeckhigh{} using dynamic imputation, as compared to
+\runtimehotdecklowbase{} to \runtimehotdeckhighbase{} when imputing on the base table. In
+this case, the cost of buffering tuples and sampling at random is negligible compared to the
+rest of query execution. Relative errors using hot deck are close to zero, ranging from
+\smapehotdecklow{}--\smapehotdeckhigh{}\%. (The exception is \Cref{q8} with $\alpha=1$, which has error
+of \smapehotdeckhighoutlier{}\% and exhibits significant skew.) Since the baseline, in these
+cases, is hot deck imputation on the base table, it is skew in the distributions arising from
+filters and joins that often leads to larger errors. Indeed, \ProjectName{} yields
+larger benefit when using relatively sophisticated, higher quality imputation algorithms as
+opposed to simpler strategies that would even be quick to implement on the base table.
 }
@@
 and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
-can be exposed to the user to guide their workflow.
+can be exposed to the user to help in guiding their workflow.
 }

### 2017-05-15 paper/queries.tex
added missing predicate to Q8 for queries table
Pass through entire paper
Add numbers to discussion of other imputation algs

@@
 \begin{lstlisting}[breaklines, basicstyle=\scriptsize\ttfamily]
-SELECT income, AVG(creatine) FROM demo, exams, labs WHERE demo.id = exams.id AND exams.id = labs.id AND income >= 13 AND income <= 15 AND weight >= 63 GROUP BY income;
+SELECT income, AVG(creatine) FROM demo, exams, labs
+WHERE demo.id = exams.id AND exams.id = labs.id
+    AND income >= 13AND income <= 15 AND weight >= 63 GROUP BY income;
 \end{lstlisting}
@@
 \begin{lstlisting}[breaklines]
-SELECT schooldegree, AVG(studentdebtowe) FROM fcc WHERE studentdebtowe > 0 GROUP BY schooldegree;
+SELECT schooldegree, AVG(studentdebtowe) FROM fcc
+WHERE studentdebtowe > 0 and schooldegree >= 0 GROUP BY schooldegree;
 \end{lstlisting}

### 2017-05-15 paper/examples.tex
paper: rewording

@@
 As an alternative strategy, the analyst can start by executing her query with $\alpha=1$ and
-incrementally reducing $\alpha$, as long as execution of the query completes within some
-acceptable time bound.
-\ProjectName{}'s planning algorithm can expose the final set of plans with optimal
+incrementally reducing $\alpha$, as long as execution of the query completes acceptably quickly.
+\ProjectName{}'s planning algorithm can also produce a set of plans with optimal
 time-quality trade-offs, which allows the user to adjust her choice of $\alpha$ to select from these plans.

### 2017-05-15 paper/algorithm.tex
reference fixes, small changes

@@
 cost-based optimizer~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
-\Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.
+\Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators. Plans produced by \ProjectName{} can mix both \textit{Impute} and \textit{Drop}
+operators.
 
@@
 
-\subsubsection{Pareto-optimal plans}
+\subsubsection{Pareto-optimal plans}\label{subsubsec:pareto}
 Given a query, \ProjectName{} produces a final set of query plans $Q$ by only keeping plans that are not dominated by others in the search space.
@@
 
-\subsubsection{Decision trees}
+\subsubsection{Decision trees}\label{subsubsec:decision-trees}
 
@@
 
-\subsubsection{Mean}
+\subsubsection{Mean}\label{subsubsec:mean}
 

### 2017-05-15 paper/examples.tex
reference fixes, small changes

--- 2017-05-15 paper/examples.tex
@@
 
-\caption{The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate, $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).
-The imputation operators are highlighted.}
+\caption{The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate, $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}), and
+can be mixed throughout a plan. The imputation operators are highlighted.}
+\label{fig:example-plans}
 \end{figure}

### 2017-05-15 paper/experiments.tex
reference fixes, small changes

--- 2017-05-15 paper/experiments.tex
@@
 \textbf{Alternate Imputation Methods}\quad
-We also experiment with mean-value imputation and hot
-deck imputation (\Cref{sec:imputation}). In our implementation of mean-value imputation,
-to facilitate non-blocking operation within the iterator
-model, we estimate the column mean from base table histograms. Therefore,
-imputations are identical no matter where the operator is placed in the query
-plan. Moreover, runtimes are essentially identical regardless of the value of
-$\alpha$ because the time cost of the imputation is relatively low independent of
-how many tuples need to be imputed. In hot deck imputation, runtimes are also
-similar across $\alpha$, because the cost of buffering
-tuples and sampling at random is negligible compared to the rest of query
-execution. \todo{be quantitative: In terms of accuracy, relative errors using hot deck are close to
-zero; since the baseline, in this case, is hot deck imputation on the base
-table, only skew in the distributions arising from filters and joins would lead
-to larger errors. In some cases, skew does arise, and errors are on the same
-magnitude as in the decision trees case. Indeed, \ProjectName{} yields larger
-benefit when using relatively sophisticated, higher quality imputation
-algorithms as opposed to simpler strategies that would even be quick to
-implement on the base table.
-}
+We also experiment with mean-value imputation and hot deck imputation
+(\Cref{sec:imputation}). In our implementation of mean-value imputation, to facilitate
+non-blocking operation within the iterator model, we estimate the column mean from base
+table histograms. Therefore, imputed values are identical no matter where the operator is
+placed in the query plan (relative errors are $0$). Runtimes are in the range of
+\runtimemeanlow{} to \runtimemeanhigh{} for queries using \ProjectName{}, as compared to
+\runtimemeanlowbase{} to \runtimemeanhighbase{} when imputing on the base table.  Here, the
+time cost of the imputation is relatively low no matter how many tuples need to be imputed.
+
+In hot deck imputation, runtimes are also similar across $\alpha$, ranging from
+\runtimehotdecklow{} to \runtimehotdeckhigh{} using dynamic imputation, as compared to
+\runtimehotdecklowbase{} to \runtimehotdeckhighbase{} when imputing on the base table. In
+this case, the cost of buffering tuples and sampling at random is negligible compared to the
+rest of query execution. Relative errors using hot deck are close to zero, ranging from
+\smapehotdecklow{}--\smapehotdeckhigh{}\%. (The exception is \Cref{q8} with $\alpha=1$, which has error
+of \smapehotdeckhighoutlier{}\% and exhibits significant skew.) Since the baseline, in these
+cases, is hot deck imputation on the base table, it is skew in the distributions arising from
+filters and joins that often leads to larger errors. Indeed, \ProjectName{} yields
+larger benefit when using relatively sophisticated, higher quality imputation algorithms as
+opposed to simpler strategies that would even be quick to implement on the base table.
 

### 2017-05-15 paper/rebuttal.tex

--- 2017-05-15 rebuttal.tex
@@
 \usepackage{enumitem}
+\usepackage{cleveref}
+\usepackage{xr}
+\externaldocument{main}
+
 \newcommand{\resp}[1]{{\color{blue}{#1}}}
@@
 
-\item \resp{We extended the original paper two additional popular imputation techniques: mean value imputation and hot-deck imputation.}
+\item \resp{We extended the original paper two additional popular imputation techniques: mean value imputation and hot-deck imputation.
 
-\item \resp{We extended experiments to include the new operators and provide more details on results for $\alpha$. We also added the Pareto frontiers for various plans and
+We added a section discussing how to construct the imputation-related functions used by our cost model, which represent the API for a new imputation operator,
+for these two new techniques and the reference technique used in the original submission (\Cref{sec:imputation})
+}
+
+\item \resp{We extended experiments to include the new operators and provide more details on results for $\alpha$ (\Cref{sec:results}).
+We also added the Pareto frontiers for various plans (\Cref{fig:pareto-frontiers} ) and
 explored planning times for more complex queries.}
 
-\item \resp{We added a section discussing how to construct the imputation-related functions used by our cost model, which represent the API for a new imputation operator.}
-We provide details on the definition of these functions for all imputation methods used.
-
-\item \resp{We included a new section that provides a roadmap for implementing \ProjectName{} on top of an existing database.}
+\item \resp{We included a new section that provides a roadmap for implementing \ProjectName{} on top of an existing database. (\Cref{sec:implementation})}
 \end{itemize}
@@
       We can trivially extend \ProjectName{} to warn users when the query chosen for execution has a high $\textsc{Penalty}$ estimate, along with the number of tuples that have
-      been dropped (after execution), so that situations such as those observed in Q8 can be identified by the user. We've added this detail to the paper.
+      been dropped (after execution), so that situations such as those observed in Q8 can be identified by the user. We've added this detail to the paper (\Cref{sec:results}).
     }
@@
    Their system helps an analyst prioritize cleaning tasks for a particular query.
-   It uses pretrained regression models as one of a number of repair strategies for missing data rather than creating a query plan which incorporates the training and running of such models.
+   It uses pre-trained regression models as one of a number of repair strategies for missing data rather than creating a query plan which incorporates the training and running of such models.
+(\Cref{subsec:missing-value})
    }
@@
 
-    \resp{We have clarified this in the paper by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and by renaming it to $\textsc{Penalty}$.}
+    \resp{We have clarified this in the paper (\Cref{sec:cost-model}) by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and by renaming it to $\textsc{Penalty}$.}
 
@@
 Our database stores tables in files, so our plans are I/O bound, but a database which uses an in-memory representation will need a different Time() function.
-We have clarified this in the paper.}
+We have clarified this in the paper (\Cref{sec:query-runtime}).}
 
@@
       to work well in our system, with errors proportional to the fraction of relevant tuples retrieved
-      and the AVG value estimated. We have noted this in the experiments section now.}
+      and the AVG value estimated. We have noted this in the experiments section now (\Cref{sec:experiments}).}
 
@@
       often does not capture the complex reasons for missingness in real world datasets.
-      We updated the discussion in S4.4.1 to emphasize this point.}
+      We updated the discussion in \Cref{sec:results} to emphasize this point.}
 
@@
     \resp{
-      We have added two new imputation methods to the experiments, to show that the planning algorithm can be used successfully with different imputation methods.
-      We have also added a subsection to the algorithm section explaining how we chose the cost functions for each imputation method that we test and giving guidelines for choosing cost functions for a new imputation method.
+      We have added two new imputation methods to the experiments (\Cref{sec:results}), to show that the planning algorithm can be used successfully with different imputation methods.
+      We have also added a subsection to the algorithm section explaining how we chose the cost functions for each imputation method that we test and giving guidelines for choosing cost functions for a new imputation method (\Cref{sec:quality} and \Cref{sec:query-runtime}).
     }
@@
    \resp{
-   We've added a section that describes in detail the components that need to be added to an existing database system to implement our technique. 
+   We've added a section (\Cref{sec:implementation}) that describes in detail the components that need to be added to an existing database system to implement our technique. 
    The details come directly from our implementation of \ProjectName{} on top of SimpleDB, a popular teaching database used at MIT, University of Washington, and Northwestern University, 
@@
 
-    \resp{Yes, we have now highlighted this for clarity in the paper as well.\todo{add specific section}}
+    \resp{Yes, we have now highlighted this for clarity in the paper as well
+    (\Cref{fig:example-plans} and \Cref{sec:search-space})}
 
@@
     \resp{We maintain the set of Pareto optimal plans at each point in the optimization. This set is significantly smaller than the set of all plans, so it represents some pruning (a point we have
-    now clarified in the paper). We found that the size of these Pareto sets is small enough that we did not need to prune further.
+    now clarified in the paper in \Cref{subsubsec:pareto}). We found that the size of these Pareto sets is small enough that we did not need to prune further.
 
@@
   we can return immediately, otherwise we return after this upper bound is reached. This is a common approach used in iterative optimization techniques, such
-  as gradient descent. We've clarified this technique in the section discussing this imputation operation.
+  as gradient descent. We've clarified this technique in the section discussing this imputation operation (\Cref{subsubsec:decision-trees}).
 }
@@
     with the average observed value. In our case, we implement a non-blocking variant that estimates the mean from the plan's histograms, avoiding the need to perform a full pass over
-    the data before imputing.}
+    the data before imputing (\Cref{subsubsec:mean}).}
 
@@
       two queries, one from each dataset used during evaluation. This frontier can be exposed to the user to help in their workflow.
+
+     The current set of results have been extended to show an additional intermediate $\alpha$ value.
 
@@
       Our cost functions are monotonic, which allows us to use a simpler optimizer than what Tummer \& Koch propose.
-    We have added a discussion of their work to the related works section.}
+    We have added a discussion of their work to the related works section (\Cref{sec:relatedwork}).}
 

### 2017-05-15 paper/algorithm.tex
paper: minor tweaks to alg. section

--- 2017-05-15 paper/algorithm.tex
@@
 algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). 
-The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
-quality of the imputations performed and the runtime performance. Our imputation planning algorithm is
-agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
-that while our algorithm is exponential in the number of joins (\Cref{sec:complexity}), in practice, planning times are not a concern for most queries.
+The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates of the 
+quality of the imputations performed and the runtime performance. Our planning algorithm is
+agnostic to the type of imputation used (\Cref{sec:imputation}).
+Finally, we show that while our algorithm is exponential in the number of joins (\Cref{sec:complexity}), planning times are not a concern in practice.
 
@@
 
-The space of query plans is similar to that considered in a canonical
-cost-based optimizer~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
+The space of query plans is similar to that considered in a canonical cost-based optimizer~\cite{blasgen1981system}, with the addition of imputation operators.
 \Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.
@@
 \label{sec:quality}
-Every instantiation of the impute-operator defines a custom penalty function $P$ for imputing \nullv{} fields.
+Each imputation method has an associated cost function $P$, which is a measure of the quality of its output.
 \ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\textsc{Penalty}_{P}(q)$, computed as follows.
@@
 Analysis of the time complexity of different imputation algorithms, especially with in-database execution, is beyond the scope of this paper.
-As with $P$, it is the ordering between queries generated by $T$ that is more important than the precision of estimates.
+As with $P$, it is the ordering between queries generated by $T$ that is more important than the precision of the estimates.
 
-We note here that for every imputation operator inserted into the query plan, a new
-statistical model is instantiated and trained. Though it is tempting to try to pre-train
-imputation models while the analyst begins to make queries, the exact rows of data, set
-of complete attributes, and set of attributes to impute are unknown until runtime, making
-this strategy infeasible. Thus, if applicable, $T$ should take into account both
+Note that for every imputation operator inserted into the query plan, a new
+statistical model is instantiated and trained.
+Though it is tempting to try to pre-train imputation models, the working data, set of complete attributes, and set of attributes to impute are unknown until runtime, making this strategy infeasible.
+Thus, if applicable, $T$ should take into account both
 training and application of the imputation model.
@@
 This set will be used to determine which attributes are imputed.
-Then, it constructs a plan cache. For each table used in the query, any available selections
-are pushed to the leaves, and
-a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}). If no selections are available, a 
+Next, it constructs a plan cache.
+For each table used in the query, any selections are pushed to the leaves, and a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}).
+If no selections are available, a 
 simple scan is added to the plan cache (\Cref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
-plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
-operations (\Cref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e.\ the
+plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is then extended for any necessary grouping and aggregation
+operations (\Cref{lst:line:group}) and for projections (\Cref{lst:line:proj}). The planner now contains the plan frontier (i.e.\ the
 best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
@@
 but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the Pareto frontiers.
-The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.
 

### 2017-05-15 paper/algorithm.tex

@@
 
-The space of query plans is similar to that considered in a canonical cost-based optimizer~\cite{blasgen1981system}, with the addition of imputation operators.
-\Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.
+The space of query plans is similar to that considered in a canonical
+cost-based optimizer~\cite{blasgen1981system}, with the addition of imputation operators.
+\Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators. Plans produced by \ProjectName{} can mix both \textit{Impute} and \textit{Drop}
+operators.
 

### 2017-05-15 paper/implementation.tex

@@
 \review{
-\section{Implementation}
+\section{Implementation}\label{sec:implementation}
 Adding imputation and our optimization algorithm to a standard SQL database requires modifications to several key database components.
@@
   In particular, this tight integration allows us to jointly choose the most effective join order and imputation placement.
-  Our goal of integrating imputation placement into the optimizer requires the other modifications that we have discussed.
 \end{itemize}

### 2017-05-15 paper/experiments.tex
paper: tweak experiments section

@@
 Our benchmark queries performed well for both cases of \texttt{AVG} and \texttt{COUNT} aggregates, 
-so we expect \ProjectName{} would provide similar performance for \texttt{SUM}, with 
+so we expect \ProjectName{} to perform similarly for \texttt{SUM}, with 
 errors proportional to the fraction of relevant tuples retrieved and the \texttt{AVG} value estimated.}
@@
 %\end{figure}
+
 \textbf{Accuracy vs.
@@
 \runtimemeanlowbase{} to \runtimemeanhighbase{} when imputing on the base table.  Here, the
-time cost of the imputation is relatively low no matter how many tuples need to be imputed.
+time cost of the imputation is low no matter how many tuples need to be imputed.
 

### 2017-05-15 paper/experiments.tex
Add sentence on tree runtimes

--- 2017-05-15 paper/experiments.tex
@@
 \textbf{Runtime vs. Base-table Imputation}\quad \Cref{fig:runtimes} summarizes the performance results. The quality-optimized
-queries are an order-of-magnitude faster than the baseline. We get another order-of-magnitude speedup when optimizing for performance.
+queries, with runtimes ranging from \runtimetreelowzero{} to \runtimetreehighzero{}, are an
+order-of-magnitude faster than the baseline, with runtimes ranging from
+\runtimetreelowbase{} to \runtimetreehighbase{}. We get another order-of-magnitude speedup when
+optimizing for performance, achieving runtimes as low as \runtimetreelowone{} to
+\runtimetreehighone{}. These speedups, ranging widely from \lowxalphazero{}x to 
+\highxalphaone{}x, depending on $\alpha$ and query, constitute significant workflow improvements
+for the analyst.
 

### 2017-05-15 paper/rebuttal.tex
rebut: squeeze

@@
 \newcommand{\ProjectName}[0]{ImputeDB}
+
+\setlist[itemize]{leftmargin=*,itemsep=-0.1em}
 
@@
 We think that the URLs add value to the references, particularly for data sets.}
-
-  \item Additional space recovered from these sections could be used to discuss the choice of techniques for imputation, and the confidence estimation idea suggested in the future work section.
-
-  \item Another interesting use of the space, or at least an idea for future work might be the role of aggregation on inference accuracy.
-In my experience aggregate queries tend to be *much* more forgiving of approximations (e.g., Online Aggregation) and data errors.
   \end{itemize}

### 2017-05-15 paper/experiments.tex
Add a=0.5 sentence to Table 4 caption
Small edits and trimming

@@
-\section{Experiments}\label{sec:experiments}
+usection{Experiments}\label{sec:experiments}
 For our experiments we plan and execute queries for three separate survey-based data sets.
@@
 \caption{Runtimes for the queries in~\Cref{tbl:queries} using reference
-    chained-equation decision trees algorithm. Using $\alpha=0$, \ProjectName{}
-    is run at the highest quality setting, while with $\alpha=1$, performance
-    is prioritized. Each query and each value of $\alpha$ was run 200 times.}
+    chained-equation decision trees algorithm, for various settings of $\alpha$.
+    Each query and each value of $\alpha$ was run 200 times.}
 \label{fig:runtimes}
@@
 \textbf{Accuracy vs.
-Base-table Imputation (AVG)}\quad \Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline, using the reference chained-equation decision trees model.
-This measures the error introduced by on-the-fly imputation as compared to full imputation on the base tables.
+Base-table Imputation (AVG)}\quad \Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) \cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline, using the reference chained-equation decision trees model.
+This measures the error introduced by dynamic imputation as compared to full imputation on the base tables.
 This comparison is the relevant one in practice --- an analyst would only be considering the trade-offs in imputation quality and time in the first place for dirty input data, for which ground truth is unavailable.
@@
     for quality ($\alpha=0$) generally achieve lower error than queries optimized for
-    efficiency ($\alpha=1$). The count fraction column shows the number of tuples used in calculating each aggregate
+    efficiency ($\alpha=1$). With ($\alpha=0.5$), an intermediate plan may be chosen if
+    available.
+    The count fraction column shows the number of tuples used in calculating each aggregate
      as a fraction of the number of tuples used when running the same query after imputing on the base table.

### 2017-05-15 paper/experiments.tex
Fix section title
paper: handle overfull hboxes

--- 2017-05-15 paper/experiments.tex
@@
 
+\begin{sloppypar}
 \textbf{Runtime vs. Base-table Imputation}\quad \Cref{fig:runtimes} summarizes the performance results. The quality-optimized
@@
 optimizing for performance, achieving runtimes as low as \runtimetreelowone{} to
-\runtimetreehighone{}. These speedups, ranging widely from \lowxalphazero{}x to 
+\runtimetreehighone{}. These speedups, ranging from \lowxalphazero{}x to 
 \highxalphaone{}x, depending on $\alpha$ and query, constitute significant workflow improvements
 for the analyst.
+\end{sloppypar}
 
@@
 
+\begin{sloppypar}
 \textbf{Accuracy vs.
-Base-table Imputation (AVG)}\quad \Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline, using the reference chained-equation decision trees model.
+Base-table Imputation (AVG)}\quad \Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline, using the chained-equation decision trees model.
 This measures the error introduced by on-the-fly imputation as compared to full imputation on the base tables.
-This comparison is the relevant one in practice --- an analyst would only be considering the trade-offs in imputation quality and time in the first place for dirty input data, for which ground truth is unavailable.
+This comparison is the relevant one in practice --- an analyst would only be considering the trade-offs in imputation quality and time for dirty input data when the ground truth is unavailable.
+\end{sloppypar}
 
@@
 
+\begin{sloppypar}
 \textbf{Planning Times}\quad
@@
 cases, the optimizer returned a query plan within \planningtimepninetynine{}. The choice of
-$\alpha$ has no impact on planning times, because query selection is done after collecting
+$\alpha$ has no impact on planning times, because query selection is performed after collecting
 the Pareto frontier.
+\end{sloppypar}
 

### 2017-05-15 paper/examples.tex

@@
 For the query in~\Cref{fig:example-query}, \ProjectName{} generates
-the plans in~\Cref{fig:quality-plan} and~\Cref{fig:fast-plan}, when 
-optimizing for quality and performance, respectively.
-
-\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operator $\mu$,
+several plans on an optimal frontier. \Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operator $\mu$,
 which employs a reference imputation strategy to fill in missing values rather than dropping tuples.
@@
 Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
-Meanwhile, \Cref{fig:fast-plan}'s performance-optimized plan uses the drop operation $\delta$ instead of filling in missing values.
+On the other hand, \Cref{fig:fast-plan} shows a performance-optimized plan that uses the drop operation $\delta$ instead of filling in missing values.
 However, it is a significant improvement over dropping all tuples with missing data in any field, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.
@@
 Tens of queries later, our epidemiologist has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.
-Knowing that there may be a need to explore this data set further, she can also incorporate her imputation model into ImputeDB for future use.
 

### 2017-05-15 paper/rebuttal.tex
Small changes to rebuttal

@@
 \begin{itemize}[leftmargin=*]
-\item Redundant figures, excessive whitespace, formatting.
-
-  \resp{Several reviewers commented on the amount of whitespace, so we've compressed many of the figures and removed whitespace to make room for additional material.}
+\item \resp{
+    Several reviewers commented on redundant figures, excessive whitespace, and
+    formatting, so we've compressed many of the figures and removed whitespace to make room for additional material.
+}
 
@@
     This has a minor impact on the experimental results.
-    It improves the ability of the regression tree model to perform imputation by making these variables available to the model.
+    It improves the ability of the decision tree model to perform imputation by making these variables available to the model.
   }
@@
     We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation.
-    These methods are in addition to the original imputation method (CART), which is based on regression trees.
-    The regression tree model incorporates correlations between attributes, so should be able to take advantage of any functional dependencies to make better predictions.
+    These methods are in addition to the original imputation method which is based on decision trees.
+    The decision tree model incorporates correlations between attributes, so should be able to take advantage of any functional dependencies to make better predictions.
     Interpolation based imputation is useful for data where the ordering of tuples is significant, like time series, but is not well suited for survey data where the ordering of tuples is arbitrary.
@@
 
- \item S1: "Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach between 0 and 24 percent" 24 percent error is pretty bad.
+ \item S1: ``Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach between 0 and 24 percent" 24 percent error is pretty bad.
 
@@
 
-  \item S3.4.4: "Drop is a special case of a sequential scan, and its time complexity can be expressed as a function of the number of heap pages read and the I/O cost per page." seems to suggest that drop is implemented as a blocking operator.
+  \item S3.4.4: ``Drop is a special case of a sequential scan, and its time complexity can be expressed as a function of the number of heap pages read and the I/O cost per page." seems to suggest that drop is implemented as a blocking operator.
 Why not just take the same approach as selection and treat it as having negligible cost?
@@
 
-    \resp{We have compacted the algorithms and tables.
+    \resp{We have compacted the algorithms, tables, and references as possible.
 Note that Tables 1-2 show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs.
@@
    The details come directly from our implementation of \ProjectName{} on top of SimpleDB, a popular teaching database used at MIT, University of Washington, and Northwestern University, 
-   amongst others. These extensions could further exploit any existing sophistication in the underlying system for tasks such as execution cost estimates.
+   among others. These extensions could further exploit any existing sophistication in the underlying system for tasks such as execution cost estimates.
 

### 2017-05-15 paper/conclusion.tex
paper: squeeze

@@
 
-We discuss the long history of dealing with missing data both in the statistical learning and database communities.
-Similar to existing work, we consider the impact of \nullv{} values in databases and develop a simple set of invariants to 
-successfully plan around them. In contrast to existing statistical learning work, our emphasis is not on the specific algorithm
+We discuss the long history of dealing with missing data both in the statistics and database communities.
+In contrast to existing work in statistics, our emphasis is not on the specific algorithm
 used to impute but rather on the placement of imputation steps in the execution of a query. In contrast to existing database work,
@@
 in some cases, and in others, operations have not been integrated into the planner. The novelty
-of our contribution lies in the formalization of missing value imputation for query planning, which results in performance
+of our contribution lies in the formalization of missing value imputation for query planning, resulting in performance
 gains over traditional approaches. This approach acknowledges that different users performing different queries on the same

### 2017-05-15 paper/rebuttal.tex
pass over rebuttal

--- 2017-05-15 paper/rebuttal.tex
@@
 
-  \resp{Several reviewers commented on the amount of whitespace, so we've compressed many of the figures and removed whitespace to make room for additional material.}
+  \resp{Several reviewers commented on the amount of whitespace, so we have compressed many of the figures and removed whitespace to make room for additional material.}
 
@@
     It improves the ability of the regression tree model to perform imputation by making these variables available to the model.
+Note that this has a small impact on the results presented in the experiments section, relative
+    to the original version submitted.
   }
 
-\item \resp{We extended the original paper two additional popular imputation techniques: mean value imputation and hot-deck imputation.
+\item \resp{We extended the original paper with two additional popular imputation techniques: mean value imputation and hot-deck imputation.
 
 We added a section discussing how to construct the imputation-related functions used by our cost model, which represent the API for a new imputation operator,
-for these two new techniques and the reference technique used in the original submission (\Cref{sec:imputation})
-}
+for these two new techniques and the reference technique used in the original submission (\Cref{sec:imputation}). }
 
@@
 
-\item \resp{We included a new section that provides a roadmap for implementing \ProjectName{} on top of an existing database. (\Cref{sec:implementation})}
+\item \resp{We included a new section that provides a roadmap for implementing \ProjectName{} on top of an existing database (\Cref{sec:implementation}).}
 \end{itemize}
@@
     The regression tree model incorporates correlations between attributes, so should be able to take advantage of any functional dependencies to make better predictions.
-    Interpolation based imputation is useful for data where the ordering of tuples is significant, like time series, but is not well suited for survey data where the ordering of tuples is arbitrary.
+    Interpolation-based imputation is useful for data where the ordering of tuples is significant, like time series, but is not well suited for survey data where the ordering of tuples is arbitrary.
 
@@
 
+      Note that after improving the categorical variable imputation in \ProjectName{}, the error for $\alpha=1.0$
+      with the reference imputation method is now slightly lower in Q8 (19.99\%).
+
       We can trivially extend \ProjectName{} to warn users when the query chosen for execution has a high $\textsc{Penalty}$ estimate, along with the number of tuples that have
@@
    Their system helps an analyst prioritize cleaning tasks for a particular query.
-   It uses pre-trained regression models as one of a number of repair strategies for missing data rather than creating a query plan which incorporates the training and running of such models.
-(\Cref{subsec:missing-value})
+   It uses pre-trained regression models as one of a number of repair strategies for missing data rather than creating a query plan which incorporates the training and running of such models (\Cref{subsec:missing-value}).
    }
@@
 
-    \resp{We have clarified this in the paper (\Cref{sec:cost-model}) by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and by renaming it to $\textsc{Penalty}$.}
+    \resp{We have clarified this in the paper (\Cref{sec:cost-model}) by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and renaming it to $\textsc{Penalty}$.}
 
@@
     \resp{This is a reasonable concern.
-Though often an appropriate proxy, optimizing for the maximum training set is flawed when the query requires the construction a large, skewed data set.
+Though often an appropriate proxy, optimizing for the maximum training set is flawed when the query constructs a large, skewed data set.
 The problem seems to be particular to queries which generate large intermediate tables.
@@
 
-    \resp{We mistakenly suggest in S3 that we treat \textit{Drop} as a blocking operator. In
-      our implementation, it has $0$ cost, and we have updated the text to reflect that.}
+    \resp{We mistakenly suggested in the original submission that we treated \textit{Drop} as a blocking operator.
+    In our implementation, it has $0$ cost, and we have updated the text to reflect that (\Cref{sec:query-runtime}).}
 
@@
     \resp{We have compacted the algorithms and tables.
-Note that Tables 1-2 show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs.
+Note that \Cref{table:nhanes-description, table:fcc-description} 1-2 show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs.
 We think that the URLs add value to the references, particularly for data sets.}
@@
   \resp{
-  We've extended the evaluation section to include more impute operators and an evaluation of planning times.
+  We've extended the evaluation section to include more impute operators and an evaluation of planning times (\Cref{sec:results}).
   }
@@
     We've broken up this section and consolidated algorithm notation. The cardinality estimation section provides more detail about the algorithm. We also
-    added intuition for how the heuristic Penalty (a renaming of Loss, for clarity) and Time should be designed. We walk through these definitions for three different
-    impute operators.
+    added intuition for how the heuristic \textsc{Penalty} (a renaming of Loss, for clarity) and \textsc{Time} should be designed. We walk through these definitions for three different
+    impute operators (\Cref{sec:quality} and \Cref{sec:query-runtime}).
     }
@@
 
-     We agree that having a way of providing an estimate on error, based on $\alpha$, would be both useful and interesting to develop.
+     We agree that having a way of providing an estimate of error, based on $\alpha$, would be both useful and interesting to develop.
      Providing error estimates further requires a reasonable notion of ground-truth. For evaluation, we've used the values imputed on the 
@@
   \resp{For the iterative regression impute operator, we set a predefined number of iterations as an upper bound on its execution. If the imputation converges,
-  we can return immediately, otherwise we return after this upper bound is reached. This is a common approach used in iterative optimization techniques, such
+  we can return immediately, otherwise we return when this upper bound is reached. This is a common approach used in iterative optimization techniques, such
   as gradient descent. We've clarified this technique in the section discussing this imputation operation (\Cref{subsubsec:decision-trees}).
@@
       a lot of the same plans for execution. We did run experiments with a range of $\alpha$ values, but instead of presenting a figure with each value of $\alpha$,
-      which can result in many duplicate plans (since different $\alpha$ values may result in the same plan), we extended the results section to include a subsection that presents the final frontiers for
-      two queries, one from each dataset used during evaluation. This frontier can be exposed to the user to help in their workflow.
+      which can result in many duplicate plans, we extended the results section to include the final frontiers for
+      two queries from our benchmarks. This frontier can be exposed to the user to help in their workflow.
 
-     The current set of results have been extended to show an additional intermediate $\alpha$ value.
+     The current set of results have also been extended to show an additional intermediate $\alpha$ value.
 
@@
       The standard algorithm presented in the original version submitted handles
-      up to 5 joins, with columns requiring imputation in all tables, with planning time of approximately one second or less. We added a small optimization to help reduce
+      up to 5 joins, with columns requiring imputation in all tables, with planning times of approximately one second or less. We added a small optimization to help reduce
       planning times for a larger number of joins. We maintain an approximation of the Pareto sets, rather than the exact set, by only adding non-dominated plans that
@@
 
-\item Could be interesting for the optimizer to pick imputation strategy based on the type of the query.
   \end{itemize}

### 2017-05-15 paper/rebuttal.tex

@@
 \begin{itemize}[leftmargin=*]
-\item Redundant figures, excessive whitespace, formatting.
-
-  \resp{Several reviewers commented on the amount of whitespace, so we have compressed many of the figures and removed whitespace to make room for additional material.}
+\item \resp{
+    Several reviewers commented on redundant figures, excessive whitespace, and
+    formatting, so we've compressed many of the figures and removed whitespace to make room for additional material.
+}
 
@@
     We have improved our system since the paper was submitted by adding imputation for categorical variables.
-    This has a minor impact on the experimental results.
-    It improves the ability of the regression tree model to perform imputation by making these variables available to the model.
-Note that this has a small impact on the results presented in the experiments section, relative
-    to the original version submitted.
+    This has a minor impact on the experimental results relative to the original submission.
+    It improves the ability of the decision tree model to perform imputation by making these variables available to the model.
   }
@@
     We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation.
-    These methods are in addition to the original imputation method (CART), which is based on regression trees.
-    The regression tree model incorporates correlations between attributes, so should be able to take advantage of any functional dependencies to make better predictions.
+    These methods are in addition to the original imputation method which is based on decision trees.
+    The decision tree model incorporates correlations between attributes, so should be able to take advantage of any functional dependencies to make better predictions.
     Interpolation-based imputation is useful for data where the ordering of tuples is significant, like time series, but is not well suited for survey data where the ordering of tuples is arbitrary.
@@
 
-    \resp{We have compacted the algorithms and tables.
-Note that \Cref{table:nhanes-description, table:fcc-description} 1-2 show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs.
+    \resp{We have compacted the algorithms, tables, and references as possible.
+Note that \Cref{table:nhanes-description, table:fcc-description} show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs.
 We think that the URLs add value to the references, particularly for data sets.}

### 2017-05-15 paper/rebuttal.tex
cref to explicit...

@@
     \resp{We have compacted the algorithms, tables, and references as possible.
-Note that \Cref{table:nhanes-description, table:fcc-description} show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs.
+Note that Tables 1 and 2 show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs.
 We think that the URLs add value to the references, particularly for data sets.}

### 2017-05-15 paper/introduction.tex
Update introduction of our baby

@@
 variety of reasons, including missing source data,
-missing columns during data integration, de-normalized databases, or as a result of outlier detection and cleaning where \nullv{} values are substituted for
-bad values~\cite{kim2003}.  Such \nullv{} values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
+missing columns during data integration, de-normalized databases, or outlier detection and
+cleaning~\cite{kim2003}.  Such \nullv{} values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
 before processing is often desirable.
@@
 This process is called \emph{imputation}.
-In this paper, we introduce \ProjectName{}, a new system designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.
-\emph{The key insight behind \ProjectName{} is that imputation
+In this paper, we introduce new methods and theory to selectively applying
+imputation to a subset of records, dynamically, during query execution.
+We instantiate these ideas in a new system, \ProjectName{}.  \emph{The key insight behind \ProjectName{} is that imputation
 only needs to be performed on the data relevant to a particular query and

### 2017-05-15 paper/rebuttal.tex
rebut: minor fix
Incorporate Nicola's comments on draft

@@
 
-    \resp{We have compacted the algorithms, tables, and references as possible.
+    \resp{We have compacted the algorithms, tables, and references.
 Note that Tables 1 and 2 show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs.
@@
 
-    \resp{We maintain the set of Pareto optimal plans at each point in the optimization. This set is significantly smaller than the set of all plans, so it represents some pruning (a point we have
+    \resp{We maintain the set of Pareto-optimal plans at each point in the optimization. This set is significantly smaller than the set of all plans, so it represents some pruning (a point we have
     now clarified in the paper in \Cref{subsubsec:pareto}). We found that the size of these Pareto sets is small enough that we did not need to prune further.
 
-      If necessary, it is possible to sample from the Pareto optimal plans during the plan enumeration.
-The final selected plan may not be Pareto optimal, but this is a potential way to control the number of plans.
+      If necessary, it is possible to sample from the Pareto-optimal plans during the plan
+      enumeration. The final selected plan may not be Pareto-optimal, but this is a potential way to control the number of plans.
       To perform this approximation, we only add plans to the frontier that are non-dominated and are sufficiently different from existing plans in the frontier.

### 2017-05-15 paper/conclusion.tex
paper: temp widow fix
Incorporate Nicola's comments on draft

@@
 \section{Conclusion}
-Our evaluation of \ProjectName{} shows that imputation of missing values can be successfully integrated into the relational algebra and
+Our implementation of \ProjectName{} and experiments show that imputation of missing values can be successfully integrated into the relational algebra and
 existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
-learning technique, as operators in the algebra and use a simple yet effective cost model to consider trade-offs in 
+learning technique, as operators in the algebra and use a simple, yet effective, cost model to consider trade-offs in 
 imputation quality and runtime. We show that different values for the trade-off parameter can yield substantially
@@
 Our experiments with the CDC NHANES, freeCodeCamp and ACS datasets
-show that \ProjectName{} can be used successfully with real-world data to run standard SQL queries over data with missing values. 
+show that \ProjectName{} can be used successfully with real-world data, running standard SQL
+queries over tuples with missing values. 
 We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
@@
 
+\pagebreak
 \subsection{Future Work}

### 2017-05-15 paper/rebuttal.tex
rebut: minor changes

--- 2017-05-15 paper/rebuttal.tex
@@
 
+    \begin{sloppypar}
 We added a section discussing how to construct the imputation-related functions used by our cost model, which represent the API for a new imputation operator,
-for these two new techniques and the reference technique used in the original submission (\Cref{sec:imputation}). }
+for these two new techniques and the reference technique used in the original submission (\Cref{sec:imputation}).
+\end{sloppypar}
+}
 
 \item \resp{We extended experiments to include the new operators and provide more details on results for $\alpha$ (\Cref{sec:results}).
-We also added the Pareto frontiers for various plans (\Cref{fig:pareto-frontiers} ) and
+We also added the Pareto frontiers for various plans (\Cref{fig:pareto-frontiers}) and
 explored planning times for more complex queries.}
@@
     \resp{
+      \begin{sloppypar}
     We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation.
@@
     Interpolation-based imputation is useful for data where the ordering of tuples is significant, like time series, but is not well suited for survey data where the ordering of tuples is arbitrary.
+  \end{sloppypar}
 
@@
      Providing error estimates further requires a reasonable notion of ground-truth. For evaluation, we've used the values imputed on the 
-     tables when applying whole-table imputation. Based on our evaluation, higher values of $\alpha$ correctly identify increased error
-     within a query, however, extending that to be predictive is not something we implemented in the current system.
+     tables when applying whole-table imputation.
+Based on our evaluation, higher values of $\alpha$ correspond to increased error, however, extending $\alpha$ to be predictive is not something we implemented in the current system.
      }
@@
 5.3.2) and we can run a simple query in 4 seconds, so we can run approximately $5,000$ queries before taking more time than full imputation.
-We've made this point explicit in the ACS experiment section now.
+We've made this point explicit in the ACS experiment section.
 

### 2017-05-15 paper/introduction.tex

@@
 optimally in query plans.   Specifically, our optimization algorithms generate the
-Pareto-optimal frontier of plans that trade imputation cost for result quality,
+Pareto-optimal trade-offs between imputation cost for result quality,
 and allow the analyst to specify their desired plan from this frontier.
@@
 --- between \lowsmapealphazero{} and \highsmapealphaone{}
-percent, in one error measure --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
+percent --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
 for details).
@@
   We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
-  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
+  It does so by maintaining multiple sets of Pareto-optimal plans according to the cost model.
   By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.

### 2017-05-15 paper/main.tex
Add sentence on tree runtimes

@@
 \renewcommand{\algorithmicensure}{\textbf{Output:}}
+
+\setlist[itemize]{leftmargin=2em,itemsep=-0.2em}
 
@@
 
+\newcommand{\runtimetreelowzero}{75 ms}
+\newcommand{\runtimetreehighzero}{1 second}
+\newcommand{\runtimetreelowone}{12 ms}
+\newcommand{\runtimetreehighone}{19 ms}
+\newcommand{\runtimetreelowbase}{6.5 seconds}
+\newcommand{\runtimetreehighbase}{27 seconds}
+
 \newcommand{\planningtimelow}{0.7 ms}
@@
 \newcommand{\planningtimepninetyninepointnine}{14 ms} % 99.9 percentile planning time of any single query
+
+\newcommand{\runtimemeanlow}{12 ms}
+\newcommand{\runtimemeanhigh}{25 ms}
+\newcommand{\runtimemeanlowbase}{24 ms}
+\newcommand{\runtimemeanhighbase}{44 ms}
+\newcommand{\runtimehotdecklow}{12 ms}
+\newcommand{\runtimehotdeckhigh}{29 ms}
+\newcommand{\runtimehotdecklowbase}{21 ms}
+\newcommand{\runtimehotdeckhighbase}{92 ms}
+
+\newcommand{\smapehotdecklow}{0}
+\newcommand{\smapehotdeckhigh}{3}
+\newcommand{\smapehotdeckhighoutlier}{24}
 
@@
   Furthermore, we show that the query results from on-the-fly imputation differ from the
-  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaone{}\%, in one error measure.
+  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaone{}\%.
   Finally, we show that while dropping tuples with missing values that fail query constraints discards 6--78\% of the data, on-the-fly imputation loses only 0--21\%.

### 2017-06-29 simpledb/src/java/simpledb/Catalog.java
feat: start cleaning up command line interface

@@ -182,7 +182,6 @@ public class Catalog {
                 TupleDesc t = new TupleDesc(typeAr, namesAr);
                 HeapFile tabHf = new HeapFile(new File(baseFolder+"/"+name + ".dat"), t);
                 addTable(tabHf,name,primaryKey);
-                System.out.println("Added table : " + name + " with schema " + t);
             }
         } catch (IOException e) {
             e.printStackTrace();

### 2017-06-29 simpledb/src/java/simpledb/Parser.java

@@ -20,7 +20,6 @@ public class Parser {
     
     public Parser(double alpha, boolean imputeAtBase) {
         this(x -> new ImputedLogicalPlan(alpha, imputeAtBase));
-        System.out.println("==> Using parser with alpha=" + alpha + " impute at base=" + imputeAtBase);
     }
 
     public Parser() {
@@ -611,97 +610,38 @@ public class Parser {
             "group by", "max(", "min(", "avg(", "count", "rollback", "commit",
             "insert", "delete", "values", "into" };
 
-    public static void main(String argv[]) throws IOException {
-        if (argv.length < 1 || argv.length > 7) {
-            System.out.println("Invalid number of arguments.\n" + usage);
-            System.exit(0);
-        }
-
-
-        List<String> cleanArgs = new ArrayList<>();
-        double alpha = 0.0;
-        boolean imputeAtBase = false;
-
-        for(int i = 0; i < argv.length; i++) {
-            String arg = argv[i];
-            if (arg.equals("--base")) {
-                imputeAtBase = true;
-            } else if (arg.equals("--alpha")) {
-                i++;
-                String alphaStr = argv[i];
-                alpha = Double.parseDouble(alphaStr);
-            } else {
-                cleanArgs.add(arg);
-            }
-        }
-
-        Parser p = new Parser(alpha, imputeAtBase);
-        p.start(cleanArgs.toArray(new String[0]));
-    }
-
     static final String usage = "Usage: parser catalogFile [-explain] [-f queryFile] [--alpha <double>] [--base]";
 
     protected void shutdown() {
         System.out.println("Bye");
     }
 
-    protected boolean interactive = true;
+    protected int start(File catalogFile, Boolean shouldExplain) {
+        explain = shouldExplain;
 
-    protected void start(String[] argv) throws IOException {
         // first add tables to database
-        Database.getCatalog().loadSchema(argv[0]);
+        Database.getCatalog().loadSchema(catalogFile.toString());
         TableStats.computeStatistics();
 
-        String queryFile = null;
-
-        if (argv.length > 1) {
-            for (int i = 1; i < argv.length; i++) {
-                if (argv[i].equals("-explain")) {
-                    explain = true;
-                    System.out.println("Explain mode enabled.");
-                } else if (argv[i].equals("-f")) {
-                    interactive = false;
-                    if (i++ == argv.length) {
-                        System.out.println("Expected file name after -f\n"
-                                + usage);
-                        System.exit(0);
-                    }
-                    queryFile = argv[i];
-
-                } else {
[…]

### 2017-06-29 simpledb/src/java/simpledb/SimpleDb.java

@@ -9,205 +9,261 @@ import java.nio.file.Files;
 import java.nio.file.Path;
 import java.nio.file.Paths;
 import java.text.ParseException;
+import java.util.ArrayList;
+import java.util.Arrays;
 import java.util.List;
+
+import org.apache.commons.cli.*;
 
 public class SimpleDb {
 	public static void main(String args[]) throws DbException, TransactionAbortedException, IOException {
-		// convert a file
-		if (args[0].equals("convert")) {
-			BufferedReader in = null;
-			BufferedOutputStream out = null;
-			try {
-				if (args.length < 3 || args.length > 5) {
-					System.err.println("Usage: simpledb convert FILE NUM_COLUMNS");
-					System.err.println("Usage: simpledb convert FILE NUM_COLUMNS TYPE_STRING [FIELD_SEP]");
-					System.exit(-1);
-				}
+		String usage =
+				"Usage: simpledb load [OPTION]...\n" +
+				"       simpledb query [OPTION]...\n" +
+				"       simpledb experiment [OPTION]...\n" +
+				"       simpledb print [OPTION]..."
+				;
 
-				File inFile = new File(args[1]);
-				if (!inFile.canRead()) {
-					System.err.format("Cannot read input file: %s", inFile);
-					System.exit(-1);
-				}
-				in = new BufferedReader(new FileReader(args[1]));
-
-				out = new BufferedOutputStream(System.out);
-
-				int numOfAttributes = Integer.parseInt(args[2]);
-				Type[] ts = new Type[numOfAttributes];
-				char fieldSeparator = ',';
-
-				if (args.length == 3)
-					for (int i = 0; i < numOfAttributes; i++)
-						ts[i] = Type.INT_TYPE;
-				else {
-					String typeString = args[3];
-					String[] typeStringAr = typeString.split(",");
-					if (typeStringAr.length != numOfAttributes) {
-						System.err.println("The number of types does not agree with the number of columns");
-						return;
-					}
-					int index = 0;
-					for (String s : typeStringAr) {
-						try {
-							ts[index++] = Type.ofString(s);
-						} catch (ParseException ex) {
-							System.err.println(ex.getMessage());
-							return;
-						}
-					}
-					if (args.length == 5)
-						fieldSeparator = args[4].charAt(0);
-				}
-
-				HeapFileEncoder.convert(in, out, BufferPool.getPageSize(), numOfAttributes, ts, fieldSeparator);
[…]

### 2017-06-29 simpledb/src/java/simpledb/TableStats.java

@@ -59,13 +59,14 @@ public class TableStats {
     public static void computeStatistics() {
         Iterator<Integer> tableIt = Database.getCatalog().tableIdIterator();
 
-        System.out.println("Computing table stats.");
+        System.err.print("Computing table stats.");
         while (tableIt.hasNext()) {
             int tableid = tableIt.next();
             TableStats s = new TableStats(tableid, IOCOSTPERPAGE);
             setTableStats(Database.getCatalog().getTableName(tableid), s);
+            System.err.print('.');
         }
-        System.out.println("Done.");
+        System.err.println(" done.");
     }
 
     /**

### 2017-06-30 paper/algorithm.tex
fig 4: fix to avoid dirty set issue

@@
 \begin{figure}
+  \cameraready{
   \begin{align*}
@@
     \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
-    \text{ where }\ P' &= \left\{q ~\Bigg|~ \begin{aligned}
-        q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
-        \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
-        \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
+    \text{ where }\\ 
+    S = \llbracket Q[T] \rrbracket \cup \{p\}\\
+    \text{ in }
+    P' &= \left\{q ~\Bigg|~ \begin{aligned}
+              q \in S,\\
+              \textsc{Dirty}(q) = \textsc{Dirty}(p) \implies\\
+              \not \exists q' \in S.\doms{q'}{q}
               \end{aligned}
@@
   \end{align*}
-  \caption{Semantics of a plan cache $Q$, where $T$ is a set of tables and $p$ is a query plan}
+  \caption{Semantics of a plan cache $Q$, where $T$ is a set of tables and $p$ is a query plan. \cameraready{When extending with a new plan $p$, the cache guarantees
+  no dominated plans with the same dirty set are kept.}}
   \label{fig:semantics-plan-cache}
+  }
 \end{figure}

### 2017-06-30 paper/experiments.tex
paper: conform to camera ready style
some changes for camera-ready, see slack post

@@
   \centering
+  \captionsetup{labelfont=bf}
+  \caption{Percentage of values missing in the CDC NHANES 2013--2014 data.}\label{table:nhanes-description}
+  \captionsetup{labelfont=up}
   \begin{subtable}{0.3\textwidth}
     \centering
+    \caption{Demographics (\texttt{demo}). \demorows{} rows.}
     \input{tables/cdc_demo}
-    \caption{Demographics (\texttt{demo}). \demorows{} rows.}
   \end{subtable}\quad\begin{subtable}{0.3\textwidth}
     \centering
+    \caption{Laboratory Results (\texttt{labs}). \labexrows{} rows.}
     \input{tables/cdc_labs}
-    \caption{Laboratory Results (\texttt{labs}). \labexrows{} rows.}
   \end{subtable}\quad\begin{subtable}{0.3\textwidth}
     \centering
+    \caption{Physical Results (\texttt{exams}). \labexrows{} rows.}
     \input{tables/cdc_exams}
-    \caption{Physical Results (\texttt{exams}). \labexrows{} rows.}
   \end{subtable}
-  \caption{Percentage of values missing in the CDC NHANES 2013--2014 data.}\label{table:nhanes-description} 
 \end{table*}
@@
   \centering
+  \captionsetup{labelfont=bf}
+  \caption{Percentage of values missing in the freeCodeCamp Survey data (\texttt{fcc}).}\label{table:fcc-description}
   \input{tables/fcc}
-  \caption{Percentage of values missing in the freeCodeCamp Survey data (\texttt{fcc}).}\label{table:fcc-description} 
 \end{table}
@@
 errors proportional to the fraction of relevant tuples retrieved and the \texttt{AVG} value estimated.}
+\cameraready{Aggregates such as \texttt{MAX} and \texttt{MIN} are unlikely to perform well with dynamic imputation,
+a difficulty carried over from traditional imputation and compounded by the dynamic nature 
+of our approach.}
 
@@
   \centering
+  \captionsetup{labelfont=bf}
+  \caption{Queries used in our experiments.}\label{tbl:queries}
   \input{queries}
-  \vspace{1ex}
-  \caption{Queries used in our experiments.}\label{tbl:queries}
 \end{table*}
@@
   \centering
-\input{tables/perf_summary.tex}
+  \captionsetup{labelfont=bf}
   \caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$ parameterizations, as compared to the baseline.
@@
     as a fraction of the number of tuples used when running the same query after imputing on the base table.
-    A lower count share reflects more potential for errors.}
-\label{table:smape}
+    A lower count share reflects more potential for errors.}\label{table:smape}
+  \input{tables/perf_summary.tex}
 \end{table}
@@
 displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
-in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
+in planning time between 14\% and 23\%, on average. \cameraready{For 6 joins, this is
+a decrease from a mean planning time of 3.3 seconds to 2.5 seconds. For 7 joins, the decrease was from 14.2 to 12.2 seconds, while for 8 joins, it was
+from 60 to 49 seconds.} Users can increase the level of approximation to reduce planning times further.
 

### 2017-07-03 paper/algorithm.tex
some changes for camera-ready, see slack post
camera ready fix: explain L_{p(q)} a bit earlier

@@
 Each imputation method has an associated cost function $P$, which is a measure of the quality of its output.
-\ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\textsc{Penalty}_{P}(q)$, computed as follows.
+\ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\textsc{Penalty}_{P}(q)$. \cameraready{Underlying
+this is a helper function $\textsc{L}_{P(q)}$, which
+is a recursive accumulator of penalties along imputation nodes instantiated
+with an impute-specific penalty operator. This accumulator is computed
+as follows.}
 
@@
 
-\subsubsection{Mean}\label{subsubsec:mean}
+\subsubsection{\cameraready{Non-Blocking Approximate Mean}}\label{subsubsec:mean}
 

### 2017-07-03 paper/algorithm.tex
paper: tweak figure 4

@@
     \llbracket Q[T] \rrbracket &= \begin{cases}
-      P & (T, P) \in Q \\
+      P & (T,\ P) \in Q \\
       \emptyset & \text{otherwise} \\
     \end{cases} \\
-    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
-    \text{ where }\\ 
-    S = \llbracket Q[T] \rrbracket \cup \{p\}\\
-    \text{ in }
-    P' &= \left\{q ~\Bigg|~ \begin{aligned}
-              q \in S,\\
-              \textsc{Dirty}(q) = \textsc{Dirty}(p) \implies\\
-              \not \exists q' \in S.\doms{q'}{q}
+    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P) \} \end{aligned} \text{ where }\\
+    P &= \left\{q \in S ~|~ \begin{aligned}
+              \textsc{Dirty}(q) \neq \textsc{Dirty}(p) \lor
+              \not \exists q' \in S:\ \doms{q'}{q}
               \end{aligned}
-            \right\}
+            \right\} \\
+    S &= \llbracket Q[T] \rrbracket \cup \{p\}
   \end{align*}
-  \caption{Semantics of a plan cache $Q$, where $T$ is a set of tables and $p$ is a query plan. \cameraready{When extending with a new plan $p$, the cache guarantees
-  no dominated plans with the same dirty set are kept.}}
+  \caption{\cameraready{Semantics of a plan cache $Q$.
+$T$ is a set of tables and $p$ is a query plan.
+The cache guaranteees that plans for the same set of tables either have distinct dirty sets or are Pareto optimal.}}
   \label{fig:semantics-plan-cache}

### 2017-07-03 paper/algorithm.tex
paper: tweak definition of alpha bound optimal

@@
 
+\cameraready{
 \begin{definition}
-Let $Q$ be a set of plans. For $q \in Q$, we say $q$ is $\alpha$-bound, if $(\textsc{Penalty}(q) - \min_{q' \in Q}(\textsc{Penalty}(q')) \leq \alpha$. We denote this as $q^\alpha$.
-We define the set of $\alpha$-bound plans in $Q$ as $Q^\alpha$.
-We say a plan $q^\alpha$ is $\alpha$-bound optimal if $\forall q' \in Q^\alpha. \textsc{Time}(q) \leq \textsc{Time}(q')$. 
+Let $Q$ be a set of plans. For $q \in Q$, we say $q$ is $\alpha$-bound if \[\textsc{Penalty}(q) - \min_{q' \in Q} \textsc{Penalty}(q') \leq \alpha.\] We denote this as $q^\alpha$.
+Let the set of $\alpha$-bound plans in $Q$ be $Q^\alpha$.
+We say a plan $q^\alpha$ is $\alpha$-bound optimal if \[\textsc{Time}(q^\alpha) = \min_{q' \in Q} \textsc{Time}(q').\]
 \end{definition}
+}
 

### 2017-07-03 paper/algorithms/histogram-transform.tex
camera ready changes: algo 1
paper: rework algorithm 1

@@
 \review{
+\cameraready{
 \begin{algorithmic}[1]
@@
   \Function{UpdateHistograms}{$H,\ op$}
+  \State Let $Cols$ be the set of attribute names in $H$.
   \If{op = $\delta_C$}\label{lst:drop-start}
-  \State Set null count to zero for $H[c]$ where $c \in C$.
-  \State Scale $H[c]$ s.t. $\Call{Card}{H[c]} = \min_d \Call{Card}{H[d]}$, $\forall c$.\label{lst:drop-end}
+  \State Set null count to zero for $H[c]$ for $c \in C$.
+  \State $m \gets \min_{d \in \mathit{Cols}} \Call{Card}{H[d]}$
+  \State Scale $H[d]$ s.t. $\Call{Card}{H[d]} = m$ for $d \in Cols$.\label{lst:drop-end}
   \ElsIf{op = $\mu_C$}\label{lst:impute-start}
-  \State Scale $H[c]$ s.t. $\Call{Card}{H[c]}$ is unchanged and null
-  \State count is zero, $\forall c$.\label{lst:impute-end}
+  \For{$c \in C$}
+  \State $k \gets \Call{Card}{H[c]}$
+  \State Set null count of $H[c]$ to zero.
+  \State Scale $H[c]$ s.t. $\Call{Card}{H[c]} = k$.
+  \EndFor\label{lst:impute-end}
   \ElsIf{op = $\sigma_\phi$}\label{lst:std-start}
-  \State Scale $H[c]$ by $\Call{Selectivity}{\phi, H}$, $\forall c$.
+  \State Scale $H[d]$ by $\Call{Selectivity}{\phi, H}$ for $d \in Cols$.
   \ElsIf{op = $\bowtie_\phi$}
-  \State Scale $H[c]$ s.t. $\Call{Card}{H[c]} = \Call{Card}{\bowtie_\phi}$, $\forall c$.
+  \State Let $k_{join}$ be the estimated cardinality of the join.
+  \State Scale $H[d]$ s.t. $\Call{Card}{H[d]} = k_{join}$ for $d \in Cols$.
   \EndIf\label{lst:std-end}
@@
 \end{algorithmic}
+}
 }

### 2017-07-08 paper/algorithm.tex
paper: fix overfull hbox and length problems
paper: tighten up inline equations
paper fix: added $ to avoid latex compilation error in scalebox
paper: a pass over

@@
 \begin{align*}
-  \textsc{Dirty}(q) = \begin{cases}
+\scalebox{0.95}[1]{
+  $\textsc{Dirty}(q) = \begin{cases}
     \textsc{Dirty}(q') \setminus C & q = \delta_C(q')\ \text{or}\ q = \mu_C(q') \\
@@
     \textsc{Dirty}(t) & q = \text{some table}\ t \\
-  \end{cases}
+  \end{cases}$
+}
 \end{align*}
@@
 Let the set of $\alpha$-bound plans in $Q$ be $Q^\alpha$.
-We say a plan $q^\alpha$ is $\alpha$-bound optimal if \[\textsc{Time}(q^\alpha) = \min_{q' \in Q} \textsc{Time}(q').\]
+We say a plan $q^\alpha$ is $\alpha$-bound optimal if \[\textsc{Time}(q^\alpha) = \min_{q' \in Q^\alpha} \textsc{Time}(q').\]
 \end{definition}
@@
 
-Cardinality estimation for standard relational operators (\Crefrange{lst:std-start}{lst:std-end}) is the same as in a normal database.
+Cardinality estimation for standard relational operators (\Crefrange{lst:std-start}{lst:std-end}) is the same as in a typical database.
 \begin{algorithm}
@@
 \ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\textsc{Penalty}_{P}(q)$. \cameraready{Underlying
-this is a helper function $\textsc{L}_{P(q)}$, which
+this is a helper function, $\textsc{L}_{P(q)}$, which
 is a recursive accumulator of penalties along imputation nodes instantiated
@@
 as follows.}
-
 \begin{align*}
@@
 The penalty should also reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
-Finally, the penalty may increase as amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.
+Finally, the penalty may increase as the amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.
 
@@
 The form of $T$ depends on the specific imputation method, but there are several considerations in choosing $T$.
-In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases and it should take into account the complexity of the imputation algorithm.
+In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases, and it should take into account the complexity of the imputation algorithm.
 Analysis of the time complexity of different imputation algorithms, especially with in-database execution, is beyond the scope of this paper.
-As with $P$, it is the ordering between queries generated by $T$ that is more important than the precision of the estimates.
+As with $P$, it is the ordering between queries imposed by $T$ that is more important than the precision of the estimates.
 
@@
 $k * |\textsc{Dirty}(q)|$ trees, we have
-
 \begin{equation*}
@@
     T(q) = k * &|\textsc{Dirty}(q)| * \textsc{Card}(q) * \\
-               &(|\textsc{Clean}(q)| + |\textsc{Dirty}(q)| - 1) * log(\textsc{Card}(q))
+               &(|\textsc{Clean}(q)| + |\textsc{Dirty}(q)| - 1) * log(\textsc{Card}(q)).
     \end{aligned}
@@
 the diminishing returns from more data.
-
 \[
@@
 have
-
 \begin{equation*}
@@
     P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
-    T(q) &\approx 0
+    T(q) &\approx 0.
 \end{aligned}
@@
 
-Random hot deck imputation is a nonparametric method in which missing values in a column are
+Random hot deck imputation is a non-parametric method in which missing values in a column are
 replaced with randomly selected complete values from the same column, preserving the
@@
 we modify $T$ to account for the fact that hot-deck is blocking in our implementation.  
-
 \begin{equation*}
@@
   \caption{\cameraready{Semantics of a plan cache $Q$.
+We use $\llbracket\ \rrbracket$ to mean evaluation.
 $T$ is a set of tables and $p$ is a query plan.
@@
 
-Finally, we note that for the queries we have examined, this exponential blowup does not affect the practical performance of our optimizer.
+Finally, we note that for the queries we have examined, the exponential blowup does not affect the practical performance of our optimizer.
 Recall that the planner maintains different plans for different dirty sets, keeping only those plans that are not dominated by others.

### 2017-07-08 paper/introduction.tex
paper: add github url
paper: a pass over

@@
 This process is called \emph{imputation}.
-In this paper, we introduce new methods and theory to selectively applying
+In this paper, we introduce new methods and theory to selectively apply
 imputation to a subset of records, dynamically, during query execution.
-We instantiate these ideas in a new system, \ProjectName{}.  \emph{The key insight behind \ProjectName{} is that imputation
+We instantiate these ideas in a new system, \ProjectName{}\footnote{\ProjectUrl{}}.  \emph{The key insight behind \ProjectName{} is that imputation
 only needs to be performed on the data relevant to a particular query and

### 2017-07-08 paper/algorithm.tex
Define k in section 3.5.1

--- 2017-07-03 paper/algorithm.tex
@@
 decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
-and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
-$k * |\textsc{Dirty}(q)|$ trees, we have
-
+and $m$ is the number of dependent attributes~\cite{witten2016data}.
+\cameraready{With $k$ cycles of the algorithm, where in each cycle there are
+$|\textsc{Dirty}(q)|$ iterations of fitting models for each dirty attribute, we have}
 \begin{equation*}

### 2017-07-08 paper/experiments.tex

@@
 We collect three data sets for our experiments.
-For all data sets, we selected a subset of the original attributes.
+For all data sets, we select a subset of the original attributes.
 We also transform all data values to an integer representation by enumerating strings and transforming floating-point values into an appropriate range.
@@
 \runtimetreehighone{}. These speedups, ranging from \lowxalphazero{}x to 
-\highxalphaone{}x, depending on $\alpha$ and query, constitute significant workflow improvements
+\highxalphaone{}x depending on $\alpha$ and query, constitute significant workflow improvements
 for the analyst.
@@
 filters and joins that often leads to larger errors. Indeed, \ProjectName{} yields
-larger benefit when using relatively sophisticated, higher quality imputation algorithms as
-opposed to simpler strategies that would even be quick to implement on the base table.
+larger benefits when using relatively sophisticated, higher quality imputation algorithms as
+opposed to simpler strategies that would be quick to implement even on the base tables.
 
@@
     \ProjectName{}'s final planner Pareto frontier for \Cref{q1} and \Cref{q6}. 
-    For clarity, we have removed from the plot plans that provide nearly indistinguishable trade-offs in either dimension.
+    For clarity, we have not shown plans that provide nearly indistinguishable trade-offs in either dimension.
 }

### 2017-07-09 paper/algorithm.tex
Clarify that attributes not being imputed are ignored if missing

--- 2017-07-08 paper/algorithm.tex
@@
 iteratively fitting decision trees for the target attributes. 
-In each iteration, the missing values of a single attribute are replaced with newly imputed values.  Values in a single column can be updated over multiple
+In each iteration, the missing values of a single attribute are replaced with newly imputed
+values. \cameraready{Attributes that contain missing values but that are not being imputed are
+ignored by the imputation.}
+Values in a single column can be updated over multiple
 iterations of the algorithm. With more iterations, the quality of the
@@
 decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
-and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
-$k * |\textsc{Dirty}(q)|$ trees, we have
+and $m$ is the number of dependent attributes~\cite{witten2016data}.
+\cameraready{With $k$ cycles of the algorithm, where in each cycle there are
+$|\textsc{Dirty}(q)|$ iterations of fitting models for each dirty attribute, we have}
 \begin{equation*}

### 2017-07-10 paper/experiments.tex
Edits

@@
 \textbf{Accuracy vs. Base-table Imputation (COUNT)}\quad
-We also calculate the number of tuples used to produce each aggregate output.
+We calculate the number of tuples used to produce each aggregate output.
 The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples used when running the query on the imputed base table.
@@
 \textbf{Alternate Imputation Methods}\quad
-We also experiment with mean-value imputation and hot deck imputation
+We experiment with mean-value imputation and hot deck imputation
 (\Cref{sec:imputation}). In our implementation of mean-value imputation, to facilitate

### 2017-07-10 paper/main.tex
paper: conform to camera ready style
add qcri acknowledgment
feat: start cleaning up command line interface
fig 4: fix to avoid dirty set issue
some changes for camera-ready, see slack post
paper: swap comment symbol
paper: fix overfull hbox and length problems
paper: add github url
paper: update github url
Update constants

@@
-\documentclass[preprint]{vldb}
+\documentclass{vldb}
 \usepackage{balance}
@@
 \usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
-\usepackage[subtle]{savetrees}
+\usepackage[subtle,bibliography]{savetrees}
 \usepackage{amsfonts,amssymb,amsmath}
@@
 \usepackage{cleveref}
+\usepackage{mathtools}
 
@@
 \renewcommand{\algorithmicensure}{\textbf{Output:}}
+\newcommand{\commentsymbol}[0]{$\blacktriangleright$}
+\renewcommand{\algorithmiccomment}[1]{\hskip2em\commentsymbol\emph{#1}}
 
@@
 \newcommand{\ProjectName}{ImputeDB\xspace}
+\newcommand{\ProjectUrl}{\url{https://github.com/mitdbg/imputedb}}
 \newcommand{\sys}{\ProjectName}
@@
 \newcommand{\todobox}[2]{{\color{red}\fbox{\parbox{\columnwidth}{\textbf{TODO:} \textit{#1}\\#2}}}}
-\newcommand{\review}[1]{{\color{blue}#1}}
+\newcommand{\review}[1]{#1}
+\newcommand{\cameraready}[1]{{\color{red}#1}}
 
@@
 \newcommand{\acsimputedboneresult}{1 second}
+
 \newcommand{\lowxalphazero}{10} % the smallest speed up, comparing a=0 to base table imputation
@@
 \newcommand{\highxalphaone}{1400} % the largest speed up, comparing a=0 OR a=1 to base table imputation
+
 \newcommand{\lowsmapealphazero}{0} % the smallest SMAPE difference, comparing a=0 to base table imputation
-\newcommand{\highsmapealphazero}{6} % the largest SMAPE difference, comparing a=0 to base table imputation
+\newcommand{\highsmapealphazero}{8} % the largest SMAPE difference, comparing a=0 to base table imputation
 \newcommand{\highsmapealphaone}{20} % the largest SMAPE difference, comparing a=1 to base table imputation
+
+\newcommand{\lowcountfractionlostalphazero}{0}
+\newcommand{\highcountfractionlostalphazero}{21}
+\newcommand{\lowcountfractionlostalphaone}{6}
+\newcommand{\highcountfractionlostalphaone}{78}
+
 %\newcommand{\highxalphaone}{16000}
@@
   traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaone{}\%.
-  Finally, we show that while dropping tuples with missing values that fail query constraints discards 6--78\% of the data, on-the-fly imputation loses only 0--21\%.
+  Finally, we show that while dropping tuples with missing values that fail query constraints discards \lowcountfractionlostalphaone--\highcountfractionlostalphaone\% of the data, on-the-fly imputation loses only \lowcountfractionlostalphazero--\highcountfractionlostalphazero\%.
 \end{abstract}
@@
 The authors would like to acknowledge \AtNextCite{\defcounter{maxnames}{3}}\citeauthor{akande2015empirical} for giving us a cleaned copy of the 2012 ACS PUMS.
+This work was supported in part by the Qatar Research Computing Institute.
 
+\emergencystretch=1em
 \bibstyle{abbrv}

### 2017-07-10 paper/introduction.tex
Rephrase results

@@
 are similar in quality to those obtained with the traditional imputation approach
---- between \lowsmapealphazero{} and \highsmapealphaone{}
+--- between \lowsmapealphazero{} and \highsmapealphazero{}
 percent --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}

### 2017-07-10 paper/main.tex
Rephrase results

@@
   Furthermore, we show that the query results from on-the-fly imputation differ from the
-  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaone{}\%.
+  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphazero{}\%.
   Finally, we show that while dropping tuples with missing values that fail query constraints discards \lowcountfractionlostalphaone--\highcountfractionlostalphaone\% of the data, on-the-fly imputation loses only \lowcountfractionlostalphazero--\highcountfractionlostalphazero\%.

### 2017-07-10 paper/introduction.tex
Add cameraready

--- 2017-07-08 paper/introduction.tex
@@
 are similar in quality to those obtained with the traditional imputation approach
---- between \lowsmapealphazero{} and \highsmapealphaone{}
+--- between \lowsmapealphazero{} and \cameraready{\highsmapealphazero{}}
 percent --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}

### 2017-07-10 paper/main.tex
Add cameraready

--- 2017-07-10 paper/main.tex
@@
   Furthermore, we show that the query results from on-the-fly imputation differ from the
-  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaone{}\%.
+  traditional base-table imputation approach by \lowsmapealphazero{}--\cameraready{\highsmapealphazero{}\%}.
   Finally, we show that while dropping tuples with missing values that fail query constraints discards \lowcountfractionlostalphaone--\highcountfractionlostalphaone\% of the data, on-the-fly imputation loses only \lowcountfractionlostalphazero--\highcountfractionlostalphazero\%.

### 2017-07-12 paper/introduction.tex
Add separate results statements

--- 2017-07-08 paper/introduction.tex
@@
 standard SQL queries over a data set, even if that data has missing values.
-These queries execute between \lowxalphazero{} and \highxalphazero{} times faster than
-imputing all missing values and then running queries (the traditional approach). Furthermore, the results obtained with this approach
-are similar in quality to those obtained with the traditional imputation approach
---- between \lowsmapealphazero{} and \highsmapealphaone{}
-percent --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
-for details).
+\cameraready{
+Using dynamic imputation, these queries execute between \lowxalphazero{} and
+\highxalphazero{} times faster than the traditional approach of first imputing all missing
+values and then running queries. Empirical results obtained using dynamic imputation
+on real-world datasets show errors within \lowsmapealphazero{} to \highsmapealphazero{} percent of
+the traditional approach (see~\Cref{sec:experiments} for more details). Alternately,
+configuring dynamic imputation to prioritize query runtime yields further speedups of
+\lowxalphaone{} to \highxalphaone{} times, with errors of \lowsmapealphaone{} to
+\highsmapealphaone{} percent.
+}
 

### 2017-07-12 paper/main.tex
Add separate results statements

@@
 
+% Compare to Figure 5
 \newcommand{\lowxalphazero}{10} % the smallest speed up, comparing a=0 to base table imputation
 \newcommand{\highxalphazero}{140} % the largest speed up, comparing a=0 to base table imputation 
-\newcommand{\highxalphaone}{1400} % the largest speed up, comparing a=0 OR a=1 to base table imputation
+\newcommand{\lowxalphaone}{TODO} % the smallest speed up, comparing a=1 to base table imputation
+\newcommand{\highxalphaone}{1400} % the largest speed up, comparing a=1 to base table imputation
 
+% Compare to Table 4
 \newcommand{\lowsmapealphazero}{0} % the smallest SMAPE difference, comparing a=0 to base table imputation
 \newcommand{\highsmapealphazero}{8} % the largest SMAPE difference, comparing a=0 to base table imputation
+\newcommand{\lowsmapealphaone}{0} % the smallest SMAPE difference, comparing a=1 to base table imputation
 \newcommand{\highsmapealphaone}{20} % the largest SMAPE difference, comparing a=1 to base table imputation
 
+% Compare to Table 4
 \newcommand{\lowcountfractionlostalphazero}{0}
@@
 
-%\newcommand{\highxalphaone}{16000}
 \newcommand{\nullv}{NULL}

### 2017-07-13 paper/algorithm.tex
Replace Dirty(q) with C in imputation algs

--- 2017-07-08 paper/algorithm.tex
@@
 iteratively fitting decision trees for the target attributes. 
-In each iteration, the missing values of a single attribute are replaced with newly imputed values.  Values in a single column can be updated over multiple
+In each iteration, the missing values of a single attribute are replaced with newly imputed
+values. \cameraready{
+Attributes that contain missing values but that are not being imputed are
+ignored by the imputation.
+}
+Values in a single column can be updated over multiple
 iterations of the algorithm. With more iterations, the quality of the
@@
 decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
-and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
-$k * |\textsc{Dirty}(q)|$ trees, we have
+and $m$ is the number of dependent attributes~\cite{witten2016data}.
+\cameraready{
+Recall that $C$ is the set of dirty attributes that are being imputed in a
+given operation, where $C \subseteq \textsc{Dirty}(q)$.
+With $k$ cycles of the algorithm, where in each cycle there are
+$|C|$ iterations of fitting models for each dirty attribute, we have
+}
 \begin{equation*}
     \begin{aligned}
-    T(q) = k * &|\textsc{Dirty}(q)| * \textsc{Card}(q) * \\
-               &(|\textsc{Clean}(q)| + |\textsc{Dirty}(q)| - 1) * log(\textsc{Card}(q)).
+    T(q) = k * &|C| * \textsc{Card}(q) * \\
+               &(|\textsc{Clean}(q)| + |C| - 1) * log(\textsc{Card}(q)).
     \end{aligned}
@@
 \[
-    P(q) = \frac{|\textsc{Dirty}(q)|}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}
+    P(q) = \frac{|C|}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}
 \]
@@
 \begin{aligned}
-    P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
+    P(q) &= \sum_{c \in C} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
     T(q) &\approx 0.
@@
 \begin{aligned}
-    P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
+    P(q) &= \sum_{c \in C} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
     T(q) &= \textsc{Card}(q)

### 2017-07-14 paper/algorithm.tex
Add cameraready
paper: reword sec 3.5.1 as discussed

--- 2017-07-09 paper/algorithm.tex
@@
 and $m$ is the number of dependent attributes~\cite{witten2016data}.
-\cameraready{With $k$ cycles of the algorithm, where in each cycle there are
-$|\textsc{Dirty}(q)|$ iterations of fitting models for each dirty attribute, we have}
+
+\cameraready{
+  We define the time function as follows.
+  Let $C$ be the set of attributes to be imputed and let $k$ be the number of cycles of the algorithm, where in each cycle there are
+  $|\textsc{Dirty}(q)|$ iterations of fitting models for each dirty attribute.
+  Then, we have
   \begin{equation*}
     \begin{aligned}
-    T(q) = k * &|\textsc{Dirty}(q)| * \textsc{Card}(q) * \\
-               &(|\textsc{Clean}(q)| + |\textsc{Dirty}(q)| - 1) * log(\textsc{Card}(q)).
+      T(q) = k &\times |C| \times \textsc{Card}(q) \\
+      &\times \left(|\textsc{Clean}(q)| + |\textsc{Dirty}(q)| - 1\right) \times \log\ \textsc{Card}(q).
     \end{aligned}
-\end{equation*}
+\end{equation*}}
 
@@
 \[
-    P(q) = \frac{|\textsc{Dirty}(q)|}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}
+    P(q) = \frac{|\textsc{Dirty}(q)|}{\sqrt{ \textsc{Attr}(q) \times \textsc{Card}(q)}}
 \]

### 2017-07-14 paper/algorithm.tex
paper: realign equation
Fix extra brace
paper: tweak wording for plan cache figure

--- 2017-07-13 paper/algorithm.tex
@@
 and $m$ is the number of dependent attributes~\cite{witten2016data}.
+
 \cameraready{
-Recall that $C$ is the set of dirty attributes that are being imputed in a
-given operation, where $C \subseteq \textsc{Dirty}(q)$.
+Recall that $C$ is the set of dirty attributes to be imputed, where $C \subseteq \textsc{Dirty}(q)$.
 With $k$ cycles of the algorithm, where in each cycle there are
@@
     \begin{aligned}
-    T(q) = k * &|C| * \textsc{Card}(q) * \\
-               &(|\textsc{Clean}(q)| + |C| - 1) * log(\textsc{Card}(q)).
+    T(q) = k \times &|\cameraready{C}| \times \textsc{Card}(q) \times \\
+               &\left(|\textsc{Clean}(q)| + |C| - 1\right) \times \log\ \textsc{Card}(q).
     \end{aligned}
-\end{equation*}
+\end{equation*}}
 
@@
 \[
-    P(q) = \frac{|C|}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}
+    P(q) = \frac{|\cameraready{C}|}{\sqrt{ \textsc{Attr}(q) \times \textsc{Card}(q)}}
 \]
@@
 \begin{aligned}
-    P(q) &= \sum_{c \in C} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
+    P(q) &= \sum_{c \in \cameraready{C}} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
     T(q) &\approx 0.
@@
 \begin{aligned}
-    P(q) &= \sum_{c \in C} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
+    P(q) &= \sum_{c \in \cameraready{C}} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
     T(q) &= \textsc{Card}(q)
@@
 \end{equation*}
-}
 
@@
   \end{align*}
-  \caption{\cameraready{Semantics of a plan cache $Q$.
+  \caption{\cameraready{Semantics of the plan cache access and insertion syntax.
 We use $\llbracket\ \rrbracket$ to mean evaluation.
-$T$ is a set of tables and $p$ is a query plan.
+$Q$ is a plan cache, $T$ is a set of tables, and $p$ is a query plan.
 The cache guaranteees that plans for the same set of tables either have distinct dirty sets or are Pareto optimal.}}

### 2017-07-14 paper/algorithm.tex
small fixes
paper: add table for notation

@@
 \section{Algorithm}
+\label{sec:algorithm}
 In order to correctly plan around missing values, we first extend the set of relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
@@
 Finally, we show that while our algorithm is exponential in the number of joins (\Cref{sec:complexity}), planning times are not a concern in practice.
+
+\begin{table}
+  \centering
+  \caption{Notation used in Section~\ref{sec:algorithm}.}
+  \begin{tabular}{ll}
+    \toprule
+    Operator & Description \\
+    \midrule
+    $\textsc{Attr}(\cdot)$ & Attributes used in a predicate. \\
+    $\textsc{Time}(\cdot)$ & Estimated time to run a plan. \\
+    $\textsc{Penalty}(\cdot)$ & Quality penalty of a query plan. \\
+    $\textsc{Card}(\cdot)$ & Estimated cardinality of a plan. \\
+    $\widehat{Var}(\cdot)$ & Estimated variance of a column. \\
+    $\textsc{Dirty}(\cdot)/\textsc{Clean}(\cdot)$ & Dirty/clean attributes in a plan. \\
+    $Q[\cdot]$ & Plan cache access operator. \\
+    $\cdot \triangleleft \cdot$ & Plan cache insertion operator. \\
+    \bottomrule
+  \end{tabular}
+\end{table}
 

### 2017-07-14 paper/algorithm.tex
minor change in table language
paper: add cameraready

@@
   \centering
-  \caption{Notation used in Section~\ref{sec:algorithm}.}
+  \caption{\cameraready{Notation used in Section~\ref{sec:algorithm}.}}
+\cameraready{
   \begin{tabular}{ll}
@@
     $\textsc{Time}(\cdot)$ & Estimated time to run a plan. \\
-    $\textsc{Penalty}(\cdot)$ & Quality penalty of a query plan. \\
+    $\textsc{Penalty}(\cdot)$ & Imputation quality penalty of a plan. \\
     $\textsc{Card}(\cdot)$ & Estimated cardinality of a plan. \\
@@
   \end{tabular}
+}
 \end{table}
@@
 \[
-    P(q) = \frac{|\cameraready{C}|}{\sqrt{ \textsc{Attr}(q) \times \textsc{Card}(q)}}
+    P(q) = \frac{|\cameraready{C}|}{\sqrt{ |\textsc{Attr}(q)| \times \textsc{Card}(q)}}
 \]
@@
 \begin{aligned}
-    P(q) &= \sum_{c \in \cameraready{C}} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
+    P(q) &= \sum\limits_{c \in \cameraready{C}} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
     T(q) &\approx 0.
@@
 \begin{aligned}
-    P(q) &= \sum_{c \in \cameraready{C}} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
+    P(q) &= \sum\limits_{c \in \cameraready{C}} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
     T(q) &= \textsc{Card}(q)
@@
 $Q$ is a plan cache, $T$ is a set of tables, and $p$ is a query plan.
-The cache guaranteees that plans for the same set of tables either have distinct dirty sets or are Pareto optimal.}}
+The cache guarantees that plans for the same set of tables either have distinct dirty sets or are Pareto optimal.}}
   \label{fig:semantics-plan-cache}

### 2017-07-14 paper/conclusion.tex
paper: fix overfull hbox and length problems

@@
 
-\pagebreak
 \subsection{Future Work}
@@
 \item \textbf{Imputation confidence:} Obtaining confidence measures for query results
-produced when missing values are imputed with a statistical model is
-another possible improvement to the system. These measures could be obtained in general
-using resampling methods like cross-validation and bootstrapping\cite{kohavi1995study}.
-In contrast to standard confidence calculations, one added complexity is that
+produced when missing values are imputed is
+a possible improvement to the system. These measures could be obtained in general
+using resampling methods like cross-validation and bootstrapping~\cite{kohavi1995study}.
+In contrast to standard confidence calculations,
 imputation takes place at various points in the plan, so the composition of results
@@
 no information is shared across queries. An intriguing direction would be to take advantage
-of imputation and execution data generated by repeated queries. Note that this goes beyond
-simply caching imputation results, and instead could entail operations such as extending
+of imputation and execution data generated by repeated queries. This goes beyond
+simply caching imputations, and instead could entail operations such as extending
 intermediate results with prior imputation values to improve accuracy, or pruning out

### 2017-07-14 paper/relatedwork.tex

@@
-\section{Related Work}
+\section{Related Work}\label{sec:relatedwork}
 
@@
 
-\subsection{Missing Values and Databases}
+\subsection{Missing Values and Databases}\label{subsec:missing-value}
 There is a long history in the database community surrounding the
@@
 a subset of standard SQL.\@ \ProjectName{}, however, is not searching the space of possible
-imputation models, but rather the space of query plans that incorporate imputation operators. Another major point
-of distinction between \ProjectName{} and Fa is that the latter doesn't consider trade-offs between accuracy and computation time, but rather returns the most accurate forecast (with some stopping criterion).
+imputation models, but rather the space of query plans that incorporate imputation operators,
+and considers the trade-offs between accuracy and computation time.
 

### 2017-07-20 paper/examples.tex
paper: bold all figure labels
small fix: italicize Impute/Drop when referring to operator, for consistency

@@
 \begin{figure}
+  \captionsetup{labelfont=bf}
 \begin{lstlisting}[language=SQL]
@@
 \begin{figure}
+  \captionsetup{labelfont=bf}
   \begin{subfigure}{\columnwidth}
@@
 For the query in~\Cref{fig:example-query}, \ProjectName{} generates
-several plans on an optimal frontier. \Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operator $\mu$,
+several plans on an optimal frontier. \Cref{fig:quality-plan} shows a quality-optimized plan that uses the \textit{Impute} operator $\mu$,
 which employs a reference imputation strategy to fill in missing values rather than dropping tuples.
@@
 Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
-On the other hand, \Cref{fig:fast-plan} shows a performance-optimized plan that uses the drop operation $\delta$ instead of filling in missing values.
+On the other hand, \Cref{fig:fast-plan} shows a performance-optimized plan that uses the \textit{Drop} operation $\delta$ instead of filling in missing values.
 However, it is a significant improvement over dropping all tuples with missing data in any field, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.

### 2017-07-20 paper/introduction.tex
small fix: italicize Impute/Drop when referring to operator, for consistency

@@
 \item \textbf{Relational algebra extended with imputation:}
-  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
+  We extend the standard relational algebra with two new operators to represent imputation operations: \textit{Impute} ($\mu$) and \textit{Drop} ($\delta$).
   The \textit{Impute} operation fills in missing data values using any statistical imputation

### 2017-07-21 paper/examples.tex
Tiny edit

@@
 Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
-On the other hand, \Cref{fig:fast-plan} shows a performance-optimized plan that uses the \textit{Drop} operation $\delta$ instead of filling in missing values.
+On the other hand, \Cref{fig:fast-plan} shows a performance-optimized plan that uses the \textit{Drop} operator $\delta$ instead of filling in missing values.
 However, it is a significant improvement over dropping all tuples with missing data in any field, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.

### 2017-10-19 simpledb/src/java/simpledb/Catalog.java
feat: print most outputs to stderr

@@ -203,7 +203,7 @@ public class Catalog {
             e.printStackTrace();
             System.exit(-1);
         } catch (IndexOutOfBoundsException e) {
-            System.out.println ("Invalid catalog entry : " + line);
+            System.err.println ("Invalid catalog entry : " + line);
             System.exit(-1);
         }
     }

### 2017-10-19 simpledb/src/java/simpledb/Field.java

@@ -51,10 +51,6 @@ public interface Field extends Serializable{
                 result = ((StringField) fo).equals((StringField) fi);
             }
         }
-        if (result==false){
-            System.out.println(fo);
-            System.out.println(fi);
-        }
         return result;
     }
 }

### 2017-10-19 simpledb/src/java/simpledb/ImputedLogicalPlan.java

@@ -396,7 +396,7 @@ public class ImputedLogicalPlan extends LogicalPlan {
 		if (tablesImputed.size() >= 7) {
 			// use approximate pareto set for large number of joins
 			// with imputation scatter across the tables
-			System.out.println("Using approximate pareto sets");
+			System.err.println("Using approximate pareto sets");
 			cache.setApproximate();
 		}
 

### 2017-10-19 simpledb/src/java/simpledb/Parser.java

@@ -213,7 +213,7 @@ public class Parser {
                                     + ") not supported.");
                 }
                 groupByField = ((ZConstant) gbe).getValue();
-                System.out.println("GROUP BY FIELD : " + groupByField);
+                System.err.println("GROUP BY FIELD : " + groupByField);
             }
 
         }
@@ -240,7 +240,7 @@ public class Parser {
                 aggField = ((ZConstant) ((ZExpression) si.getExpression())
                         .getOperand(0)).getValue();
                 aggFun = si.getAggregate();
-                System.out.println("Aggregate field is " + aggField
+                System.err.println("Aggregate field is " + aggField
                         + ", agg fun is : " + aggFun);
                 lp.addProjectField(aggField, aggFun);
             } else {
@@ -310,13 +310,13 @@ public class Parser {
                 java.lang.reflect.Method m = c.getMethod(
                         "updateOperatorCardinality", p, h, h);
 
-                System.out.println("The query plan is:");
+                System.err.println("The query plan is:");
                 m.invoke(null, (Operator) physicalPlan,
                         lp.getTableAliasToIdMapping(), TableStats.getStatsMap());
                 c = Class.forName("simpledb.QueryPlanVisualizer");
                 m = c.getMethod(
-                        "printQueryPlanTree", DbIterator.class, System.out.getClass());
-                m.invoke(c.newInstance(), physicalPlan,System.out);
+                        "printQueryPlanTree", DbIterator.class, System.err.getClass());
+                m.invoke(c.newInstance(), physicalPlan,System.err);
             } catch (ClassNotFoundException e) {
             } catch (SecurityException e) {
             } catch (NoSuchMethodException e) {
@@ -471,7 +471,7 @@ public class Parser {
             curtrans.abort();
             curtrans = null;
             inUserTrans = false;
-            System.out.println("Transaction " + curtrans.getId().getId()
+            System.err.println("Transaction " + curtrans.getId().getId()
                     + " aborted.");
 
         } else if (s.getStmtType().equals("SET TRANSACTION")) {
@@ -536,7 +536,7 @@ public class Parser {
                 if (!this.inUserTrans) {
                     curtrans = new Transaction();
                     curtrans.start();
-                    System.out.println("Started a new transaction tid = "
+                    System.err.println("Started a new transaction tid = "
                             + curtrans.getId().getId());
                 }
                 try {
@@ -550,7 +550,7 @@ public class Parser {
                         query = handleQueryStatement((ZQuery) s,
[…]

### 2017-10-19 simpledb/src/java/simpledb/ATupleFormatter.java
feat: allow printing output as CSV and add command running to cli

@@ -1,0 +1,21 @@
+package simpledb;
+
+import java.io.PrintStream;
+
+/**
+ * A tuple formatter prints tuples to some output stream.
+ *
+ * @author Jack Feser
+ */
+
+public abstract class ATupleFormatter {
+    protected final PrintStream out;
+
+    public ATupleFormatter(PrintStream out) {
+        this.out = out;
+    }
+
+    public abstract void formatHeader(TupleDesc td);
+    public abstract void formatTuple(Tuple t);
+    public abstract void formatFooter(String f);
+}

### 2017-10-19 simpledb/src/java/simpledb/CsvTupleFormatter.java

@@ -1,0 +1,35 @@
+package simpledb;
+
+import java.io.PrintStream;
+
+public class CsvTupleFormatter extends ATupleFormatter {
+    public CsvTupleFormatter(PrintStream out) {
+        super(out);
+    }
+
+    @Override
+    public void formatHeader(TupleDesc td) {
+        for (int i = 0; i < td.numFields(); i++) {
+            out.print(td.getFieldName(i));
+            if (i < td.numFields() - 1) {
+                out.print(',');
+            }
+        }
+        out.print('\n');
+    }
+
+    @Override
+    public void formatTuple(Tuple t) {
+        TupleDesc td = t.getTupleDesc();
+        for (int i = 0; i < td.numFields(); i++) {
+            out.print(t.getField(i).toString());
+            if (i < td.numFields() - 1) {
+                out.print(',');
+            }
+        }
+        out.print('\n');
+    }
+
+    @Override
+    public void formatFooter(String f) {}
+}

### 2017-10-19 simpledb/src/java/simpledb/Parser.java

@@ -11,11 +11,20 @@ import jline.ConsoleReader;
 import jline.SimpleCompletor;
 
 public class Parser {
-    static boolean explain = false;
+    private boolean explain = false;
     private final Function<Void, LogicalPlan> planFactory;
+    private Transaction curtrans = null;
+    private boolean inUserTrans = false;
+
+    // Basic SQL completions
+    public static final String[] SQL_COMMANDS = { "select", "from", "where",
+        "group by", "max(", "min(", "avg(", "count", "rollback", "commit",
+        "insert", "delete", "values", "into" };
+
+    static final String usage = "Usage: parser catalogFile [-explain] [-f queryFile] [--alpha <double>] [--base]";
     
     public Parser(Function<Void, LogicalPlan> planFactory) {
-    	this.planFactory = planFactory;
+        this.planFactory = planFactory;
     }
     
     public Parser(double alpha, boolean imputeAtBase) {
@@ -285,9 +294,6 @@ public class Parser {
         return lp;
     }
 
-    private Transaction curtrans = null;
-    private boolean inUserTrans = false;
-
     public Query handleQueryStatement(ZQuery s, TransactionId tId)
             throws TransactionAbortedException, DbException, IOException,
             simpledb.ParsingException, Zql.ParseException {
@@ -510,21 +516,20 @@ public class Parser {
     public void setTransaction(Transaction t) {
         curtrans = t;
     }
-
     public Transaction getTransaction() {
         return curtrans;
     }
 
-    public void processNextStatement(String s) {
+    public void processNextStatement(String s, ATupleFormatter formatter) {
         try {
-            processNextStatement(new ByteArrayInputStream(s.getBytes("UTF-8")));
+            processNextStatement(new ByteArrayInputStream(s.getBytes("UTF-8")), formatter);
         } catch (UnsupportedEncodingException e) {
             e.printStackTrace();
             throw new RuntimeException(e);
         }
     }
 
-    public void processNextStatement(InputStream is) {
+    public void processNextStatement(InputStream is, ATupleFormatter formatter) {
         try {
             ZqlParser p = new ZqlParser(is);
             ZStatement s = p.readStatement();
@@ -556,7 +561,7 @@ public class Parser {
                                         + "\n -- parser only handles SQL transactions, insert, delete, and select statements");
                     }
                     if (query != null)
-                        query.execute();
+                        query.execute(formatter);
 
                     if (!inUserTrans && curtrans != null) {
[…]

### 2017-10-19 simpledb/src/java/simpledb/PrintTupleFormatter.java

@@ -1,0 +1,32 @@
+package simpledb;
+
+import java.io.PrintStream;
+
+public class PrintTupleFormatter extends ATupleFormatter {
+    public PrintTupleFormatter(PrintStream out) {
+        super(out);
+    }
+
+    @Override
+    public void formatHeader(TupleDesc td) {
+        String names = "";
+        for (int i = 0; i < td.numFields(); i++) {
+            names += td.getFieldName(i) + "\t";
+        }
+        out.println(names);
+        for (int i = 0; i < names.length() + td.numFields() * 4; i++) {
+            out.print("-");
+        }
+        out.println("");
+    }
+
+    @Override
+    public void formatTuple(Tuple t) {
+        out.println(t);
+    }
+
+    @Override
+    public void formatFooter(String f) {
+        out.println(f);
+    }
+}

### 2017-10-19 simpledb/src/java/simpledb/Query.java

@@ -92,27 +92,20 @@ public class Query implements Serializable {
         started = false;
     }
 
-    public void execute() throws IOException, DbException, TransactionAbortedException {
+    public void execute(ATupleFormatter formatter) throws IOException, DbException, TransactionAbortedException {
         TupleDesc td = this.getOutputTupleDesc();
 
-        String names = "";
-        for (int i = 0; i < td.numFields(); i++) {
-            names += td.getFieldName(i) + "\t";
-        }
-        System.out.println(names);
-        for (int i = 0; i < names.length() + td.numFields() * 4; i++) {
-            System.out.print("-");
-        }
-        System.out.println("");
+        formatter.formatHeader(td);
+
+        int cnt = 0;
 
         this.start();
-        int cnt = 0;
         while (this.hasNext()) {
-            Tuple tup = this.next();
-            System.out.println(tup);
+            formatter.formatTuple(this.next());
             cnt++;
         }
-        System.out.println("\n " + cnt + " rows.");
         this.close();
+
+        formatter.formatFooter("\n " + cnt + " rows.");
     }
 }

### 2017-10-19 simpledb/src/java/simpledb/SimpleDb.java

@@ -160,15 +160,34 @@ imputationMethod);
                 .type(File.class)
                 .desc("the directory which contains the database")
                 .build());
+        options.addOption(Option.builder("c")
+                .optionalArg(true)
+                .hasArg()
+                .argName("command")
+                .type(String.class)
+                .desc("run a single query and exit")
+                .build());
+        options.addOption(Option.builder()
+                .longOpt("csv")
+                .optionalArg(true)
+                .build());
 
 	    CommandLineParser parser = new DefaultParser();
 	    try {
 	        CommandLine line = parser.parse(options, args);
             double alpha = Double.parseDouble(line.getOptionValue("alpha", defaultAlpha));
             File catalogFile = new File(line.getOptionValue("db") + "/catalog.txt");
+            String query = line.getOptionValue("c", null);
+            boolean useCsv = line.hasOption("csv");
 
             Parser sqlParser = new Parser(alpha, false);
-            return sqlParser.start(catalogFile, false);
+
+            if (query == null) {
+                return sqlParser.start(catalogFile, false);
+            } else {
+                sqlParser.runSingleQuery(catalogFile, query, false, useCsv);
+                return 0;
+            }
         } catch (org.apache.commons.cli.ParseException exp) {
 	        System.err.println(exp.getMessage());
 	        HelpFormatter formatter = new HelpFormatter();

### 2017-10-19 simpledb/test/acstest/CleanTest.java

@@ -49,6 +49,6 @@ public class CleanTest {
 		ZStatement s = p.readStatement();
 		Parser pp = new Parser();
 		Query query = pp.handleQueryStatement((ZQuery)s, new TransactionId());
-		query.execute();
+		query.execute(new PrintTupleFormatter(System.out));
 	}
 }

### 2017-10-19 simpledb/test/simpledb/systemtest/QueryTest.java

@@ -9,16 +9,7 @@ import java.util.ArrayList;
 import org.junit.Ignore;
 import org.junit.Test;
 
-import simpledb.BufferPool;
-import simpledb.Database;
-import simpledb.DbException;
-import simpledb.HeapFile;
-import simpledb.HeapFileEncoder;
-import simpledb.Parser;
-import simpledb.TableStats;
-import simpledb.Transaction;
-import simpledb.TransactionAbortedException;
-import simpledb.Utility;
+import simpledb.*;
 
 public class QueryTest {
 	
@@ -83,7 +74,8 @@ public class QueryTest {
 		// So, don't bother for now; future TODO.
 		// Regardless, each of the following should be optimized to run quickly,
 		// even though the worst case takes a very long time.
-		p.processNextStatement("SELECT * FROM emp,dept,hobbies,hobby WHERE emp.c1 = dept.c0 AND hobbies.c0 = emp.c2 AND hobbies.c1 = hobby.c0 AND emp.c3 < 1000;");
+		PrintTupleFormatter f = new PrintTupleFormatter(System.out);
+		p.processNextStatement("SELECT * FROM emp,dept,hobbies,hobby WHERE emp.c1 = dept.c0 AND hobbies.c0 = emp.c2 AND hobbies.c1 = hobby.c0 AND emp.c3 < 1000;", f);
 	}
 	
 	/**
@@ -181,6 +173,7 @@ public class QueryTest {
 		// So, don't bother for now; future TODO.
 		// Regardless, each of the following should be optimized to run quickly,
 		// even though the worst case takes a very long time.
[…]

### 2017-10-19 paper/algorithm.tex
fix: change lang in table to avoid overflow into margin
paper: bold all figure labels

@@
   \centering
+  \captionsetup{labelfont=bf}
   \caption{\cameraready{Notation used in Section~\ref{sec:algorithm}.}}
@@
     $\textsc{Time}(\cdot)$ & Estimated time to run a plan. \\
-    $\textsc{Penalty}(\cdot)$ & Imputation quality penalty of a plan. \\
+    $\textsc{Penalty}(\cdot)$ & Impute quality penalty of a plan. \\
     $\textsc{Card}(\cdot)$ & Estimated cardinality of a plan. \\
@@
 \begin{figure}
+  \captionsetup{labelfont=bf}
   \begin{minipage}[c]{0.5\columnwidth}
@@
 
-Drop operators (\Crefrange{lst:drop-start}{lst:drop-end}) reduce the cardinality of their input relation by removing tuples which contain missing values.
-For a drop $\delta_C$, for each column $c \in C$ we set its missing value count to zero, because all of these tuples will be dropped.
+\textit{Drop} operators (\Crefrange{lst:drop-start}{lst:drop-end}) reduce the cardinality of their input relation by removing tuples which contain missing values.
+For a \textit{Drop} $\delta_C$, for each column $c \in C$ we set its missing value count to zero, because all of these tuples will be dropped.
 This can result in a set of histograms which have different cardinalities if some columns have more missing values than others.
@@
 
-Impute operators (\Crefrange{lst:impute-start}{lst:impute-end}) maintain the cardinality of their inputs, but they also fill in missing values.
-For an impute $\mu_C$, we set the missing value count to zero, and rescale each histogram so that its new cardinality, with no missing values, is the same as its old cardinality.
+\textit{Impute} operators (\Crefrange{lst:impute-start}{lst:impute-end}) maintain the cardinality of their inputs, but they also fill in missing values.
+For an \textit{Impute} $\mu_C$, we set the missing value count to zero, and rescale each histogram so that its new cardinality, with no missing values, is the same as its old cardinality.
 We assume that the imputed values will not change the shape of the distribution of values in each column.
@@
 \begin{figure}
+  \captionsetup{labelfont=bf}
   \cameraready{
